### 2017-05-04 cgo18mscr.tex
first commit

\documentclass[sigplan,9pt]{acmart}

\usepackage{booktabs} % For formal tables

\usepackage[utf8]{inputenc}
\usepackage{fullpage}

\usepackage{graphics}

\usepackage{graphicx}
\usepackage{subcaption}

\usepackage{multirow}

\usepackage{color}

\usepackage{soul}

%\usepackage{flushend}

%\usepackage{indentfirst}

\usepackage{amsthm}
\usepackage{amsmath,amssymb}
%\usepackage{mathrsfs}
\usepackage{mathtools}

\usepackage[noend]{algpseudocode}
\usepackage{algorithm}

\newcommand{\etal}{et~al.}
\algrenewcommand\algorithmicindent{3ex}%

% Copyright
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
%\acmDOI{10.475/123_4}

% ISBN
%\acmISBN{123-4567-24-567/08/06}

%Conference
\acmConference[CGO'18]{International Symposium on
Code Generation and Optimization}{2018}{} 
%\acmYear{1997}
%\copyrightyear{2016}
%
%\acmPrice{15.00}

\begin{document}
%\title{A work-based metric for efficiently applying online iterative optimisation}
\title{Online Iterative Optimisation\\Guided by Low Overhead Profiling}
%\titlenote{Produces the permission block, and
%  copyright information}
%\subtitle{Extended Abstract}
%\subtitlenote{The full version of the author's guide is available as
%  \texttt{acmart.pdf} document}

\author{Rodrigo C. O. Rocha}
%\authornote{Dr.~Trovato insisted his name be first.}
%\orcid{1234-5678-9012}
\affiliation{%
  \institution{University of Edinburgh, UK}
  %\streetaddress{P.O. Box 1212}
  %\city{Dublin} 
  %\state{Ohio} 
  %\postcode{43017-6221}
}
\email{r.rocha@ed.ac.uk}

\author{Pavlos Petoumenos}
%\authornote{The secretary disavows any knowledge of this author's actions.}
\affiliation{%
  \institution{University of Edinburgh, UK}
  %\streetaddress{P.O. Box 1212}
  %\city{Dublin} 
  %\state{Ohio} 
  %\postcode{43017-6221}
}
\email{ppetoume@inf.ed.ac.uk}

\author{Lu\'is F. W. G\'oes}
%\authornote{This author is the
%  one who did all the really hard work.}
\affiliation{%
  \institution{PUC Minas, Brazil}
  %\streetaddress{1 Th{\o}rv{\"a}ld Circle}
  %\city{Hekla} 
  %\country{Iceland}
}
\email{lfwgoes@pucminas.br}

\author{Murray Cole}
\affiliation{
  \institution{University of Edinburgh, UK}
  %\streetaddress{P.O. Box 5000}
}
\email{mic@inf.ed.ac.uk}

\author{Zheng Wang}
\affiliation{%
  \institution{Lancaster University, UK}
  %\city{Moffett Field}
  %\state{California} 
  %\postcode{94035}
}
\email{z.wang@lancaster.ac.uk}

\author{Hugh Leather}
\affiliation{%
  \institution{University of Edinburgh, UK}
  %\streetaddress{8600 Datapoint Drive}
  %\city{San Antonio}
  %\state{Texas} 
  %\postcode{78229}
}
\email{hleather@inf.ed.ac.uk}

%\author{John Smith}
%\affiliation{\institution{The Th{\o}rv{\"a}ld Group}}
%\email{jsmith@affiliation.org}

%\author{Julius P.~Kumquat}
%\affiliation{\institution{The Kumquat Consortium}}
%\email{jpkumquat@consortium.net}

% The default list of authors is too long for headers}
\renewcommand{\shortauthors}{R. Rocha et al.}

\begin{abstract}
Efficiently selecting the best combination of compiler optimisations for a program
remains an open problem due to their complex interaction and the large optimisation search space.
Iterative optimisation techniques, which consist of iterating
over this large space searching for the best optimisation,
has the ability to adapt to new platforms, programs and workload while still
having a systematic and simple optimisation process.
Although some recent work have studied the impact of using multiple
input datasets for performing iterative optimisation, applying this
technique in an online scenario still poses some challenges.

In this paper, our main goal is to enable iterative optimisation in
online scenarios with the restriction of executing distinct inputs
only once.
In order to address this challenge,
we propose a the use of a work-based metric that allows for comparing different
combination of compiler optimisations even when executed with distinct inputs.
Because of its online nature, we also propose a relaxed instrumentation
that performs low overhead profiling of the amount of work executed, while
being able to trade-off between accuracy and overhead in a controlled manner.
\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011041</concept_id>
<concept_desc>Software and its engineering~Compilers</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011041.10011043</concept_id>
<concept_desc>Software and its engineering~Retargetable compilers</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

%\ccsdesc[500]{Software and its engineering~Compilers}
%\ccsdesc[500]{Software and its engineering~Retargetable compilers}

\keywords{Iterative Optimisation, Profiling, Relaxed Instrumentation}

\maketitle

\section{Introduction}

Modern optimising compilers
have reached a high level of sophistication,
providing a large number of optimisations.
Because of the unpredictability of optimisation
interactions, in addition to the growing complexity
of processor architectures and applications,
the correct choice of optimisations and their ordering
can have a significant impact on the performance of the
code being optimised~\cite{pan06,fursin07,kulkarni12}.

Each of these optimisations interacts with the code in complicated ways,
depending on all other optimisations and the order they were applied to
the code being optimised.
Understanding the interactions of optimisations is very important
in determining a good solution to the phase-ordering problem~\cite{touati06,kulkarni12}.
Because of all those dependences and interactions, although
most compiler optimisations yield significant improvements
in many programs, the potential for performance degradation in
certain program patterns is known to compiler writers and many users~\cite{pan06,zhou12,kulkarni12}.
%Traditional compilers typically apply the same set of optimisation in a fixed order to all functions in a program, without regard the code being optimised.

Efficiently selecting the best combination of compiler optimisations
for a given program or program section remains an open problem.
Compiler writers typically use a combination of experience and insight to construct
the default sequences of prearranged optimisations found in compilers.
However, these default pre-arranged set of optimisations do not include all possible optimisations
and they are always applied in the same pre-defined fixed order, without regard the code being optimised~\cite{pan06,cavazos07,zhou12,kulkarni12}.
%in hopes that a programmer will know which combination of optimisations will benefit their code~\cite{pan06,cavazos07,zhou12,kulkarni12}.

A well known compilation technique that addresses this challenge is iterative
optimisation.
Iterative optimisation has the ability to adapt to new
platforms, program and workload while still
having a systematic and simple optimisation process.
Iterative optimisations repeatedly evaluate a large number
of compiler optimisations until the best combination is
found for a particular program~\cite{fursin07,chen10}.
The main challange concerning iterative optimisation is the need for quickly
exploring such a large optimisation space~\cite{fursin07,cavazos07,zhou12}.

Until recently, most of existing work  had been focusing
on finding the best optimisation through repeated runs using a single input.
Although they demonstrate the potential of iterative optimisation,
in real scenarios the user rarely executes the same input dataset multiple times~\cite{bodin98,kisuki99,stephenson03,kulkarni04,agakov06}.
Applying iterative optimisation in light of a single input may not result in good performance when executing the optimised code with different inputs.

Most of real world applications are complex enough 
so that a single input case does not capture the whole
range of possible scenarios and program behaviour~\cite{haneda06,fursin07,chen10,chen12a}.
Because programs can exhibit behaviours that differ greatly depending on the input,
%This is something that is well known when writing test
%cases for correctness.
%We can extend the same argument when optimising the code for performance.
using a single input for iterative optimisation can produce poor performance when executed with different inputs.

Recent work have been studying
the impact of using multiple input datasets
for performing iterative optimisation.
This previous work suggests that optimising based on a representative range of input datasets
allows for selecting a robust compiler optimisation 
that produces good performance in real scenarios where
the input is expected to change constantly~\cite{haneda06,fursin07,chen10,chen12a,chen12b,fang15,mpeis16}.
Their results show that a limited number of input datasets may be sufficient to
obtain a well optimised program for a wider range of different inputs~\cite{haneda06,fursin07,chen10,chen12a}.

Finding such a robust combination of compiler optimisations by means of iterative optimisation across a large range of inputs
may be very time consuming.
In order to speed up this process we intend to reduce the number of actual executions during the exploration
of the optimisation space without much degradation of the final selected optimisation.

\begin{figure}[htb]
    \centering
    \includegraphics[width=\linewidth]{figs/diagram}
    \caption{Overview of the execution engine for applying iterative optimisation.}
    \label{fig:diagram}
\end{figure}

In this paper, our main goal is to enable iterative optimisation in
\textit{online} scenarios with the restriction of executing distinct inputs
only once, while targeting optimal performance across different inputs.
This online aspect is usually found in mobile and data centre
platforms~\cite{chen12b,fang15,mpeis16}, where the goal is to optimise
programs based on the workload of a particular user (or group of users)
between executions.

Measuring just execution time, for example, is useful only if the
amount of work is constant between executions with different inputs.
For this reason, we propose the use of a work-based metric in order
to compare the performance of different optimisations across 
multiple executions of the program with distinct inputs.

Because of the restriction of not repeating inputs, we instrument the program
for measuring the amount of work it performed during its execution.
Having a low overhead instrumentation is essential in this online scenario for
two main reasons:
$(i)$ the user is directly affected by large overheads;
$(ii)$ a highly intrusive instrumentation can havea significant impact on the
effect of the optimisations.

Our main contributions are the following:
\begin{itemize}
\item The use of a work-based metric in order to enable online iterative
optimisation by comparing different combination of compiler optimisations even
when executed with distinct inputs.
\item We propose a relaxed instrumentation for low overhead profiling, with a controlled
trade-off between accuracy and overhead.
\end{itemize}

\section{Related Work}

Chen~\etal~\cite{chen10,chen12a} evaluate the effectiveness of iterative
compilations across a large number of input test cases.
Their main motivation is to answer the question:
\textit{How data input dependent is iterative optimisation?}
Their results show that it is possible to find a combination of compiler optimisations
that achieves at least 86\% of the maximum speedup across all input test cases.
These optimal combinations are program-specific and yield average
speedups up to 3.75$\times$ over the highest optimisation level of compilers.

When optimising a program, the main method for iterative optimisation used by Chen~\etal~\cite{chen10,chen12a}
evaluates each combination of compiler optimisations across all the available inputs, i.e.,
if $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimisations,
they perform a total of $O(NM)$ runs of the program being optimised.
Furthermore, they use a pre-defined set of only 300 different combinations of compiler optimisations,
which represents a very small sample of the optimisation search space for most modern compilers, e.g.
LLVM has 56 distinct optimisation passes and GCC has about 47 high-level (SSA form) optimisation passes and
about 25 low-level (RTL) optimisation passes, which in both cases result in much more than $2^{50}$ distinct
combinations of compiler optimisations, without considering repetition.

Recent work~\cite{chen12b,fang15} have applied the same idea of performing
input-dependent iterative optimisation to distributed applications on data centres.
In summary, each worker receives a subset of the input dataset, called the evaluation dataset,
to perform an \textit{online} iterative optimisation of the code being executed.
Each worker performs the same the method for iterative optimisation used by Chen~\etal~\cite{chen10,chen12a},
i.e., they evaluate each combination of compiler optimisations across all the evaluation dataset.
However, because the optimisation is performed online, they usually consider a small evaluation dataset
and a small number of compiler optimisations.

Fursin~\etal~\cite{fursin07} addressed the problem of comparing the effect of two optimisations
on two distinct inputs. For that purpose, they proposed to use instruction per cycle (IPC) as the metric for performing such comparison.
Their result show that using IPC seems promising as a robust metric for iterative optimisation
across large input datasets.
However, 
some specific optimisation techniques may affect the use of IPC as a robust metric, and specially
IPC has been shown to provide poor performance estimation for multi-threaded programs~\cite{alameldeen06,eyerman08}.
In particular, IPC can give a skewed performance measure if threads spend time in \textit{spin-lock loops}
or other synchronisation mechanisms. 
Some existing work on performance assessment suggest that
total execution time should be used for measuring performance of multi-threaded programs~\cite{alameldeen06,eyerman08}.
Aalameldeen and Wood~\cite{alameldeen06} in particular suggest that a simple work-related metric should be used
if the unit of work is representative enough.
Work-related metrics have already been largely used for measuring performance of throughput-oriented applications,
for other applications, however, choosing an appropriate unit of work can be more challanging~\cite{alameldeen06}.

\section{Work-based Metric} \label{sec:metric}

A possible candidate for a work-based metric, borrowing from Physics,
is using the concept of \textit{average power}, i.e. the amount of
\textit{work}, $\Delta W$, performed during a period of time, $\Delta t$.
\[
   P_{avg} = \frac{\Delta W}{\Delta t}
\]

%The hypothesis is that, given two optimisations $o_1$ and $o_2$,
%a program compiled with optimisation $o_2$ would \textit{consistently} perform better
%than when it is compiled with $o_1$, i.e., $o_1 \tilde{<}_p \, o_2$,
%if it performs more \textit{work} per unit of time when compiled with $o_2$
%instead of $o_1$. 
By measuring the amount of \textit{work} done per unit of time we reduce
the impact of input-dependent aspects and focus instead on the efficiency
of the optimised program.
For this metric, the main challange is to precisely define what
represents \textit{work}.

%For that purpose, we intend to use profiling
%techniques to measure the input size and amount of computation
%performed on the execution path triggered by the input~\cite{ball94,ball96,zaparanuks12,coppa14}.

%In addition to being useful for speeding up iterative optimisation across a large
%number of different inputs, this metric based on comparing work-based efficiencies could
%potentially be used for applying iterative optimisation \textit{online}. 
%However, the overhead for estimating the amount of \textit{work} must be low
%enough for the performance gains to be beneficial.

Previous work have proposed profiling-based mechanism to estimate
input sizes~\cite{zaparanuks12,coppa14}.
Coppa~\etal~\cite{coppa14} in particular propose the concept of
\textit{read memory size} for automatically estimating the size of the input
passed to a routine, where \textit{read memory size} represents the number of
distinct memory cells first accessed by a read operation.
In other words, the \textit{read memory size} metric measures the size of the 
useful portion of the input's memory footprint.
However, because we are interested in the amount of computational work performed
in respect of a given input, the memory footprint of the input may not always
have a direct correspondence to  the amount of computational work.

Goldsmith~\etal~\cite{goldsmith07} use \textit{block frequency} as the
measure for performance for empirically describing the asymptotic behaviour
of programs, which is known as empirical computational complexity.
Block frequency is a relative metric that represents the number of times a
basic block executes~\cite{ball94,ball96}. 
They argue in favour of block frequency due to its portability, repeatability
and exactness, since it does not suffer from timer resolution problems or
non-deterministic noises.
Block frequency also has the advantage of being efficiently profiled by means
of automatic code instrumentation~\cite{knuth73,ball94}.

However, in the context of comparing different optimisations,
although block frequency would be able to capture aspects of optimisations
that simplify the control-flow graph (CFG), measuring work at the basic block
resolution would not capture effects of optimisations at the instruction level.
Because of that, we extend the idea of using basic block frequency to measure
computational work by also considering the computational cost of each
basic block. The computational cost of a basic block is given by weighing
the instructions that it contains.

We model the computational work $\Delta W$ as a linear equation based on
block frequency information and a cost-model of the instruction set.
\[
\Delta W = \varepsilon + \sum_{B} w(B)f(B)
\]
where $f(B)$ represents the frequency of basic block $B$
and $w(B)$ represents the computational work of executing $B$.
We define the work of a basic block $B$ as the sum of the cost
of its instructions, i.e.,
\[
w(B) = \sum_{i} w_i N_B(i)
\]
where $w_i$ is the cost of instruction $i$ and $N_B(i)$ is the number of
occurrences of instruction $i$ in basic block $B$.

In this simplified model, we consider that $w_i$ is constant across all
programs and executions in the given target platform.
However, $N_B(i)$ is program dependent but constant across executions,
while $f(B)$ is both program and execution dependent, since $f(B)$ can
change when executing with different inputs.
In other words, $N_B(i)$ is a static value known at compile-time and 
$f(B)$ is a dynamic value known only at run-time.
%If we define $x_i$ as
%\[
%x_i = \sum_{B\in P} N_B(i)f(B)
%\]
%we can write our definition of work $\Delta W$ as the following linear equation
%\[
%\Delta W = \varepsilon + \sum_{i\in I} w_i x_i
%\]

%\subsection{Linear regression}
%
%\begin{figure}[htb]
%    \centering
%    \includegraphics[width=\linewidth]{figs/cost-model.pdf}
%    \caption{Comparison between the naive and optimal instrumentation with no compiler optimisation.} %, i.e., compiled with \textbf{\texttt{-O0}}.}
%    \label{fig:cost-model}
%\end{figure}
%mse 0.000145373168444
%mae 0.00737261168981
%mae* 0.0033892281406
%mape 75.2924320971
%corr (0.99946441743545622, 0.0)

\section{Instrumentation}

In this section we describe how the computation of the work-metric can be
performed during runtime by means of instrumenting the code.
%with the necessary computation for the work metric, $\Delta W$, as described in Section~\ref{sec:metric}.
Because we define work as a linear equation on the block frequency counters,
it is possible to embed its computation into the execution of the program.
A naive instrumentation would consists basically of having a global counter
that starts with the interception value, $\varepsilon$, and each basic block
increments its own cost into the global counter. Although this instrumentation
is easily implemented, it inserts a large amount of overhead into the program.

In 1973, Donald Knuth~\cite{knuth73} provided an optimal algorithm for
profiling block frequency, inserting fewer probes than the previously described
naive instrumentation that was commonly used in practice~\cite{knuth71}.
Forman~\cite{forman81} proposed further overhead reductions by placing the
probes in basic blocks that are less likely to be executed based on static
heuristics.
Ball and Larus~\cite{ball94} offered a detailed discussion comparing
two approaches for optimally profiling block frequency, namely,
placing probes on the vertices or the edges of the control-flow graph (CFG).
They show that the edge-based approach produces optimal placement of probes.

Ball and Larus~\cite{ball94} show that a set of edges represents the minimum
number of probes for profiling block frequency if and only if the complementary
set of edges forms a spanning tree.
The instrumentation optimises the placement of the probes with respect to a
weighing that assigns a non-negative value to each edge in the CFG.
The cost of profiling a set of edges is proportional to the sum of the weights
of the edges.
These weights can be obtained either by empirical measurements or heuristic
estimations.
In order to minimise the profiling cost, the instrumentation computes the
maximum spanning tree for avoiding probing in frequently executed edges.
Although this instrumentation algorithm is proved to produce the optimal
placement of probes for well-structured CFGs, it may produce sub-optimal
placement for some unstructured CFGs.

Once we have the maximum spanning tree, probes are placed on any edge not in 
the spanning tree. Figure~\ref{fig:cfg-example} shows an example of a CFG
with a maximum spanning tree represented by the black edges, while the edges 
highlighted in red represent the placement of the probes.

\begin{figure}[htb]
\centering {
  \includegraphics[scale=0.9]{figs/cfg-example.pdf}\\\vspace{1ex}
  \resizebox{0.45\textwidth}{!}{
  %\scalebox{0.8}{
     \begin{minipage}{0.5\textwidth}
     Instrumented value for each probe $P_i$:
     \begin{align*}
     w(P_0) &= w(B_0) + w(B_1) + w(B_4) + w(B_6) + w(B_7) + w(B_8) + w(B_9)\\
     w(P_1) &= w(B_2) - w(B_1) - w(B_4) - w(B_6) - w(B_7) - w(B_8)\\
     w(P_2) &= w(B_5) - w(B_6)\\
     w(P_3) &= w(B_3) - w(B_4) - w(B_6) - w(B_7)
     \end{align*}
     \end{minipage}
  }
}
  \caption{Example of a CFG with its minimum spanning tree in black and the
   basic blocks highlighted in red represent the instrumented basic blocks with
   the placement of the probes.}
  \label{fig:cfg-example}
\end{figure}

In contrast to the naive instrumentation where each basic block records only
its own amount of work, with the optimal profiling, the instrumented basic blocks
need record an aggregated value of work that represents a path in the CFG.
These values are constructed with some instrumented basic blocks speculatively
assuming some paths while other probes correct when these assumptions are wrong
(see for example $w(P_0)$ and $w(P_1)$ in Figure~\ref{fig:cfg-example}).

\begin{algorithm}[h]
  \caption{Pseudocode of the data-flow analysis for assigning the values computed
  in each probe of the instrumentation for the profiling of the work metric.}
  \label{alg:populateEdgeInfo}
  \begin{algorithmic}
    \Function{\textrm{populateEdgeInfo}}{$G$}

    \For{\textbf{each} $e \in $ \textrm{instrumentedEdges}($G$) }
       \State $B_I \gets $ \textrm{instrumentedBlock}($e$)
       \State $inc[e] \gets \{ B_I \}$
       \State $dec[e] \gets \{ \}$
       \State $known[e] \gets $ \textbf{true}
    \EndFor

    \State $changed \gets $ \textbf{true}
    \While{$changed$}
       \State $changed \gets $ \textbf{false}
	   \For{\textbf{each} $B \in V(G)$}
          \State $uIn \gets $ \textrm{unknownIn}($known, G, B$)
          \State $uOut \gets $ \textrm{unknownOut}($known, G, B$)
		  \If{$uIn=0$ \textbf{and} $uOut=1$}
             \State $changed \gets $ \textbf{true}
		     \State $incOut \gets \{\}$
			 \State $decOut \gets \{\}$
		     \For{$B_p\in$ \textrm{predecessors}($G, B$)}
			    \State $incOut \gets incOut \bigcup inc[(B_p,B)]$
			    \State $decOut \gets decOut \bigcup dec[(B_p,B)]$
			 \EndFor
		     \For{$B_s\in$ \textrm{successors}($G, B$)}
			    \State $incOut \gets incOut \cup dec[(B,B_s)]$
			    \State $decOut \gets decOut \cup inc[(B,B_s)]$
			 \EndFor
		     \For{$B_s\in$ \textrm{successors}($G, B$)}
			    \If{\textbf{not} $known[(B,B_s)]$}
			       \State $inc[(B,B_s)] \gets incOut\setminus{decOut}$
			       \State $dec[(B,B_s)] \gets decOut\setminus{incOut}$
			       \State $known[(B,B_s)] \gets$ \textbf{true}
				\EndIf
			 \EndFor
		  \EndIf
		  \If{$uIn=1$ \textbf{and} $uOut=0$}
             \State $changed \gets $ \textbf{true}
		     \State $incIn \gets \{\}$
			 \State $decIn \gets \{\}$
		     \For{$B_s\in$ \textrm{successors}($G, B$)}
			    \State $incIn \gets incIn \cup inc[(B,B_s)]$
			    \State $decIn \gets decIn \cup dec[(B,B_s)]$
			 \EndFor
		     \For{$B_p\in$ \textrm{predecessors}($G, B$)}
			    \State $incIn \gets incIn \bigcup dec[(B_p,B)]$
			    \State $decIn \gets decIn \bigcup inc[(B_p,B)]$
			 \EndFor
		     \For{$B_p\in$ \textrm{predecessors}($G, B$)}
			    \If{\textbf{not} $known[(B_p,B)]$}
			       \State $inc[(B_p,B)] \gets incIn\setminus{decIn}$
			       \State $dec[(B_p,B)] \gets decIn\setminus{incIn}$
			       \State $known[(B_p,B)] \gets$ \textbf{true}
				\EndIf
			 \EndFor
		  \EndIf
	   \EndFor
    \EndWhile
    \EndFunction
  \end{algorithmic}
\end{algorithm}

Because the algorithm for the optimal placement of the probes is proved to
uniquely compute the block frequencies by propagating the probe counts,
we adapt this algorithm in order to compose the aggregated values that will
be instrumented in each probe, based on our model of computational work,
$\Delta W$, derived from the basic block frequencies (see Section~\ref{sec:metric}).
We perform a similar propagation of the probes in a symbolic fashion,
as illustrated in Figure~\ref{fig:cfg-example}.
This symbolic propagation of the probes is implemented by the data-flow analysis
described in Algorithm~\ref{alg:populateEdgeInfo}, and the final aggregated values
are extracted from the edge information as described by Algorithm~\ref{alg:instrValue}.

The data-flow analysis in Algorithm~\ref{alg:populateEdgeInfo} keeps two sets
for each edge, namely, the increment and the decrement sets.
We consider that both sets represent the \textit{edge expressions} shown in Figure~\ref{fig:cfg-example},
for which we define a \textit{symbolic sum} by computing the union of the increment
and decrement sets, respectively, with the appropriate cancellation of common elements.
This data-flow analysis is based on the invariant that the symbolic sum
of all the incoming edges must equals the symbolic sum of the outgoing
edges. For example, the symbolic sum of the incoming edges of the basic
block $B_8$ is $P_0 - P_1$, where $P_3$ is cancelled out.

%By construction, 
%the \textit{symbolic experssions} of a basic block, i.e., 
%the \textit{symbolic sum} of the incoming edges (or outgoing edges) of a basic block,
%has the following properties:
%\begin{enumerate}
%\item Probes in the increment set either (post-)dominates or is (post-)dominated by the given basic block;
%\item Probes in the decrement set are (post-)dominated by the probes in the increment set;
%\item Probes in the decrement set and the given basic block are in exclusive paths
%      from the virtual node of the CFG to the (post-)dominators in the increment set;
%\end{enumerate}
%From property (1) it follows that whenever the given basic block is executed,
%exactly one of the probes in the increment set is executed.
%From properties (2) and (3), it follows that whenever a probe in the decrement set
%is executed, although the given basic block is not executed, one of the probes
%in the decrement set is executed.

Algorithm~\ref{alg:instrValue} reads the edge information for each
basic block by computing the symbolic sum of their respective incoming
edges (or outgoing edges).
From these \textit{edge expressions}, we are able to compose the aggregated
value of the probes.
The positive terms in the edge expression of a basic block indicate that the
amount of work of this basic block will be incremented in the probes represented
by these positive terms, similarly, the negative terms indicate that the amount
of work of this basic block will be decremented in the probes represented by
these negative terms.
For example, because the edge experssion for the basic block $B_8$ is
$P_0 - P_1$, the amount of work of $B_8$, denoted by $w(B_8)$,
is incremented in probe $P_0$ and decremented in $P_1$.

\begin{algorithm}[h]
  \caption{Pseudocode that describes how the edge information is used in order to extract
           the value that will be computed in a given instrumented basic block $B_I$.
           This algorithm could equally be implemented based on the predecessors.}
  \label{alg:instrValue}
  \begin{algorithmic}
    \Function{\textrm{instrValue}}{$G, B_I, inc, dec$}
	\State $v \gets 0$
    \For{\textbf{each} $B \in V(G)$}
	   \State $inc_B \gets \{\}$
	   \State $dec_B \gets \{\}$
%	   \For{$B_p\in$ \textrm{predecessors}($G, B$)}
%	      \State $inc_B \gets inc_B \bigcup inc[(B_p,B)]$
%	      \State $dec_B \gets dec_B \bigcup dec[(B_p,B)]$
%	   \EndFor
	   \For{$B_s\in$ \textrm{successors}($G, B$)}
	      \State $inc_B \gets inc_B \bigcup inc[(B,B_s)]$
	      \State $dec_B \gets dec_B \bigcup dec[(B,B_s)]$
	   \EndFor
	   \If{ $B_I \in inc_B\setminus{dec_B}$}
	      \State $v \gets v + w(B)$
	   \EndIf
	   \If{ $B_I \in dec_B\setminus{inc_B}$}
	      \State $v \gets v - w(B)$
	   \EndIf
	\EndFor
    \Return $v$
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\subsection{Relaxed Instrumentation}

Although the optimal instrumentation significantly reduces the profiling overhead
when compared to the naive instrumentaion, from an average overhead of 79\% to 13\%,
in some critical cases, even the optimal instrumentation can have an overhead
of about 70\% (see benchmark \texttt{adpcm\_d} in Figure~\ref{fig:overhead-O3}).
In order to further reduce the overhead in these critical cases, we propose a relaxed instrumentation by trading off accuracy and overhead.

%\begin{figure*}[htb]
%    \centering
%    \includegraphics[width=\textwidth]{figs/overhead-O0.pdf}
%    \caption{Comparison between the naive and optimal instrumentation
%              with no compiler optimisation.}
%    \label{fig:overhead-O0}
%\end{figure*}

%Define an extended DAG of a CFG as a directed graph which may contain cycles
%if they have known constant trip counts.

The relaxed instrumentation performs a post-processing on the resulting instrumentation
of the optimal algorithm. 
The relaxation starts by extracting \textit{extended} DAGs (directed acyclic graphs)
from the CFG, as illustrated in Figure~\ref{fig:cfg-relax-example}.
Each loop (and the outer most region of the function) represents an extended DAG
where any inner loop is considered to never execute, i.e. only the headers of the
inner loops are included into the extended DAG.

%
%LLVM has a Scalar Evolution Analysis which is used primarily to analyse expressions
%involving induction variables in loops~\cite{pop05}.
%An induction variable is a variable that is increased or decreased by a fixed amount
%on every iteration of a loop or is a linear function of another induction variable.
%
%This Scalar Evolution Analysis provides a way to compute \textit{small} constant
%trip counts (where \textit{small} is defined by a threshold).

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.9]{figs/cfg-relax-example.pdf}
  \caption{Example of a CFG containing a loop and its decomposition into extended DAGs
           when applying the relaxation.}
  \label{fig:cfg-relax-example}
\end{figure}

For every extended DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$,
we relax the instrumentation accuracy by selecting a subset of the probes to be removed,
subject to the maximum allowed percentage error, $M$.

We model the relaxation as a 0-1 Knapsack problem:
\[
%\textrm{maximise } \sum_{i=0}^{k} f(I_i)c(I_i)x_i
\textrm{maximise } \sum_{i=0}^{k} f(P_i)x_i
\]
\[
\textrm{subject to } \sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \textrm{ and } x_i\in\{0,1\}
\]
%$c(I_i)$ is the relative overhead of the instrumentation on the instrumented basic block,
where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$
,$\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative
to the minimum work value possible to compute in the extended DAG,
and $x_i$ denotes the probes selected for removal.
Because the percentage error is computed based on the path with the minimum
amount of work, $\varepsilon(P_i)$ represents the maximum error possible
that would be incurred when removing probe $P_i$.
Furthermore, by constraining the percentage error of every extended DAG below a
given threshold we guarantee that the final error of the relaxation will always
be bounded by the threshold.

While the optimal placement of probes tries to place probes in edges that are
less likely to be executed, the relaxation focus on removing probes that are
more likely to be executed.
The necessary block frequency information for optimising both the placement of
probes can be acquired from profiles of previous executions of the program or
by a static heuristic of the CFG during compilation.

For our experiments, we implemented two solvers for the 0-1 Knapsack problem:
the optimal brute-force solver;
the greedy heuristic based on sorting the items ~\cite{dantzig57}.
We use the brute-force solver for DAGs with a small number of probes and
the greedy heuristic when the number of probes is greater than a threshold.
Some of the benchmarks have DAGs with several hundreds of probes, which could
result in a long compilation time.

\section{Experimental Evaluation}

In this section we discuss our experimental evaluation.
First we describe the benchmarks with datasets used in the experiments.
Afterwards, we discuss our results concerning the instrumentations for profiling
our work metric.
Finally we present the results of the online iterative optimisation.

We implemented the instrumentations in LLVM 4.0.
The target platform is a Linux-4.4.27 system with an Intel Core i7-4770 3.40GHz
Skylake~CPU with 16~GiB RAM.

\subsection{Benchmarks}

%Iterative optimisations is essentially based on repeatedly trying
%out a large number of compiler optimisations until the best
%combination of compiler optimisations is found for a particular program.
%Our main goal is to speed up iterative optimisation while targeting
%optimal performance across large input datasets.

For the experimental evaluation we have used a subset of the \textit{KDataSets} benchmark suit,
which is the same benchmark and dataset suit used by Chen~\etal~\cite{chen10,chen12a}.
The KDataSets contains 1000 different inputs for each one of its benchmark programs.
These benchmarks cover a broad spectrum of application scenarios, ranging from simple
embedded signal-processing tasks to common mobile-phone and desktop tasks.
The different inputs try to capture distinct characteristics in terms of workload sizes
and how these workloads exercise different control flow paths.
A summary of the benchmark and dataset suit is shown in Table~\ref{tab:kdatasets}.

\begin{table}[h]
\centering
\scalebox{.8}{
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Program} & \textbf{LOC}    & \textbf{Input file size}            & \textbf{Input description}              \\ \hline % Domain
bitcount      & 460    &  -                         & Numbers: random                \\ \hline
qsort         & 154    & 32K-1.8M                   & 3D coordinates                 \\ \hline
%dijkstra      & 163    & 0.06K-4.3M                 & Adjacency matrices             \\ \hline
%patricia      & 290    & 0.6K-1.9M                  & IP and mask pairs              \\ \hline
jpeg\_d       & 13501  & 3.6K-1.5M                  & JPEG images                    \\ \hline
jpeg\_c       & 14014  & 16K-137M                   & PPM images                     \\ \hline
tiff2bw       & 15477  & \multirow{4}{*}{9K-137M}   & \multirow{4}{*}{TIFF images}   \\ \cline{1-2}
tiff2rgba     & 15424  &                            &                                \\ \cline{1-2}
tiffdither    & 15399  &                            &                                \\ \cline{1-2}
tiffmedian    & 15870  &                            &                                \\ \hline
susan\_c      & 1376   & \multirow{3}{*}{12K-46M}   & \multirow{3}{*}{PGM images}    \\ \cline{1-2}
susan\_e      & 1376   &                            &                                \\ \cline{1-2}
susan\_s      & 1376   &                            &                                \\ \hline
%mad           & 2358   & 28K-27M                    & MP3 audios                     \\ \hline
lame          & 14491  & 167K-36M                   & WAVE audios                    \\ \hline
adpcm\_c      & 210    & 167K-36M                   & WAVE audios                    \\ \hline
adpcm\_d      & 211    & 21K-8.8M                   & ADPCM audios                   \\ \hline
%gsm           & 3806   & 83K-18M                    & Sun/NeXT audios                \\ \hline
%ghostscript   & 99869  & 11K-43M                    & Postscript files               \\ \hline
%ispell        & 6522   & \multirow{3}{*}{0.1K-42M}  & \multirow{3}{*}{Text files}    \\ \cline{1-2}
%rsynth        & 4111   &                            &                                \\ \cline{1-2}
rsynth        & 4111   &                            &  Text files                    \\ \hline %%%%%% copied line
%stringsearch  & 338    &                            &                                \\ \hline
%blowfish\_e   & 863    & 0.6K-35M                   & Files of any format            \\ \hline
%blowfish\_d   & 863    & 0.6K-35M                   & Encrypted files                \\ \hline
%pgp\_e        & 19575  & 0.6K-35M                   & Files of any format            \\ \hline
%pgp\_d        & 19575  & 0.4K-18M                   & Encrypted files                \\ \hline
%rijndael\_e   & 952    & 0.6K-35M                   & Files of any format            \\ \hline
%rijndael\_d   & 952    & 0.7K-35M                   & Encrypted files                \\ \hline
sha           & 197    & 0.6K-35M                   & Files of any format            \\ \hline
%CRC32         & 130    & 0.6K-35M                   & Files of any format            \\ \hline
%bzip2e        & 5125   & 0.7K-57M                   & Files of any format            \\ \hline
%bzip2d        & 5125   & 0.2K-25M                   & Compressed files               \\ \hline
\end{tabular}
}
\caption{Description of the KDataSets with 1000 inputs for each benchmark (Chen~\etal~\cite{chen10,chen12a}).}
\label{tab:kdatasets}
\end{table}

\subsection{Evaluation of the Instrumentation}

\begin{figure*}[htp]
    \centering
    \includegraphics[width=\textwidth]{figs/overhead-O3.pdf}
    \caption{Overhead of the instrumentations compiled with {\texttt{-O3}}.}
    \label{fig:overhead-O3}
\end{figure*}

Figure~\ref{fig:instr} shows percentage of instrumented basic blocks
for the optimal and the relaxed instrumentation with different relaxation
thresholds. The naive instrumentation always has 100\% of the basic blocks
instrumented, by definition.

\begin{figure*}[htp]
    \centering
    \includegraphics[width=\textwidth]{figs/instr.pdf}
    \caption{Percentage of instrumented basic blocks
for the optimal and the relaxed instrumentation with different relaxation
thresholds.}
    \label{fig:instr}
\end{figure*}

The \texttt{adpcm\_d} benchmark is the most critical case amongst the
used benchmarks, with a overhead of about 66\% with the optimal instrumentation.
This benchmark consists mainly of a single hot loop with several branches
inside it.
The relaxation algorithm is able to reduce this overhead down to about 52\%
(with an error threshold of 1\%) and 19\% (with an error threshold of 2\%)
by removing only one and two probes from the hot loop, respectively.
These two removed probes were placed in branches, inside the hot loop, with
a high probability of being taken, but with a small contribution to the work
measured in the loop.

%In this section we discuss the three baseline implementations that will
%be used for assessing the quality of the proposed metric.

%\vspace{1ex}
%\noindent \textbf{Iterative optimisation over a single input}

%Most of the existing iterative optimisation studies find the
%best optimisation though repeated runs on the same input.
%Although this approach will usually lead to sub-optimal performance
%across large input datasets, it provides a good baseline when considering
%the complexity and resonably low compile-time regarding
%iterative optimising compilers.
%If $M$ is the total number of combinations of compiler optimisations,
%this approach requires $O(M)$ runs of the program being optimised.
%We can consider two main scenarios:
%\textit{($i$ - best case scenario)}
%after selecting the optimisation over each individual input,
%consider the one with best performance across the whole input dataset;
%\textit{($ii$ - expected scenario)}
%after selecting the optimisation over each individual input,
%consider the average case of their performance across the whole input dataset.

%\vspace{1ex}
%\noindent \textbf{Iterative optimisation across large input datasets}

%Recent work on iterative optimisation have been targeting optimisation
%across multiple inputs~\cite{fursin07,chen10,chen12a}.
%If $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimisations,
%they perform a total of $O(NM)$ runs of the program being optimised.
%Similarly to what have been discussed previously,
%there are different ways for how to determine the optimal
%combination of compiler optimisations across multiple inputs.
%It is possible to tune the selection of the program-optimal combination
%to minimise risk or to maximise average performance.
%A compromise-based selection criterion could be to maximise average speedup
%with minimised variance.

%\vspace{1ex}
%\noindent \textbf{Iterative optimisation based on the IPC metric}

%While the previous two baselines addresses two opposite aspects of iterative optimisation,
%namely, compile-time efficiency and performance of the generated code,
%we also intend to compare against IPC as the competing baseline.
%The main reason for comparing against IPC is because it has been proposed as a metric
%for comparing the performance of two optimisations running on two distinct inputs.
%If $M$ is the total number of combinations of compiler optimisations,
%this approach requires $O(M)$ runs of the program being optimised.

\subsection{Evaluation of the Online Iterative Optimisation}

\begin{itemize}
\item Oracle-RM executes the program twice, for each input, and then measuring
the real speedup for each compiler optimisation and uses the real speedup for
selecting the best optimisation.
\item Oracle-PP represents the \textit{perfect} non-intrusive profiling by also
executing the program twice, once for estimating the amount of work and the
second for measuring its execution time.
This version uses the work-based metric, $\frac{\Delta W}{\Delta t}$, during
the iterative optimisation search.
\end{itemize}

\begin{figure*}[htb]
    \centering
    \includegraphics[width=\textwidth]{figs/speedups.pdf}
    \caption{Speedups observed with the online iterative optimisation.}
    \label{fig:speedups}
\end{figure*}

%\begin{figure*}[htb]
%    \centering
%    \includegraphics[width=\textwidth]{figs/profiled-speedups.pdf}
%    \caption{Speedups observed with the online iterative optimisation if we
%             consider the instrumentation overhead.}
%    \label{fig:profiled-speedups}
%\end{figure*}

%\subsubsection{Contribution of individual optimisation passes}
%
%Figure~\ref{fig:flagsfreq} shows an aggregated view of the final combination
%of compiler optimisations that were selected by iterative optimisation search.
%The figure presents the individual optimisation passes with at least 1\% of
%frequency in the selected combination of compiler optimisations that improved
%the performance over {\texttt{-O3}}.
%
%\begin{figure*}[htb]
%    \centering
%    \includegraphics[width=\textwidth]{figs/flagsfreq.pdf}
%    \caption{Frequency of individual optimisation passes on the final selected 
%             compiler optimisations of the iterative optimisation search over
%             all benchmarks.}
%    \label{fig:flagsfreq}
%\end{figure*}

\section{Conclusion and Future Work}

Our proposed technique for relaxed instrumentation can also be applied for similar
profiling scenarios, such as profiling empirical computational complexity~\cite{goldsmith07,zaparanuks12,coppa14}.

\section*{Acknowledgments}

%\textbf{Acknowledge the partial support by EPSRC}.

This work was supported by the UK Engineering
and Physical Sciences Research Council (EPSRC) under grants
EP/L01503X/1 for the University of Edinburgh, School
of Informatics, Centre for Doctoral Training in Pervasive
Parallelism, % (\url{http://pervasiveparallelism.inf.ed.ac.uk/}),
and also by the Institute for Computing Systems Architecture (ICSA)
in the School of Informatics at the University of Edinburgh.

\bibliographystyle{ACM-Reference-Format}
\bibliography{ref}

\end{document}

### 2017-05-08 cgo18mscr.tex
Restructuring of the section Related Work; Extract discussion of previous work from other sections and place them in as subsections of Related Work;

@@
 
+\newcommand{\itercomp}{{iterative compilation}}
+\newcommand{\Itercomp}{{Iterative compilation}}
+\newcommand{\IterComp}{{Iterative Compilation}}
+
 \begin{document}
 %\title{A work-based metric for efficiently applying online iterative optimisation}
-\title{Online Iterative Optimisation\\Guided by Low Overhead Profiling}
+\title{Online {\IterComp}\\Guided by Low Overhead Profiling}
 %\titlenote{Produces the permission block, and
@@
 remains an open problem due to their complex interaction and the large optimisation search space.
-Iterative optimisation techniques, which consist of iterating
+{\Itercomp} techniques, which consist of iterating
 over this large space searching for the best optimisation,
@@
 Although some recent work have studied the impact of using multiple
-input datasets for performing iterative optimisation, applying this
+input datasets for performing \itercomp, applying this
 technique in an online scenario still poses some challenges.
 
-In this paper, our main goal is to enable iterative optimisation in
+In this paper, our main goal is to enable \itercomp in
 online scenarios with the restriction of executing distinct inputs
@@
 
-\keywords{Iterative Optimisation, Profiling, Relaxed Instrumentation}
+\keywords{{\IterComp}, Profiling, Relaxed Instrumentation}
 
@@
 
-A well known compilation technique that addresses this challenge is iterative
-optimisation.
-Iterative optimisation has the ability to adapt to new
+A well known compilation technique that
+addresses this challenge is {\itercomp}.
+{\Itercomp} has the ability to adapt to new
 platforms, program and workload while still
 having a systematic and simple optimisation process.
-Iterative optimisations repeatedly evaluate a large number
+{\Itercomp} repeatedly evaluates a large number
 of compiler optimisations until the best combination is
 found for a particular program~\cite{fursin07,chen10}.
-The main challange concerning iterative optimisation is the need for quickly
+The main challenge concerning {\itercomp} is the need for quickly
 exploring such a large optimisation space~\cite{fursin07,cavazos07,zhou12}.
@@
 on finding the best optimisation through repeated runs using a single input.
-Although they demonstrate the potential of iterative optimisation,
+Although they demonstrate the potential of {\itercomp},
 in real scenarios the user rarely executes the same input dataset multiple times~\cite{bodin98,kisuki99,stephenson03,kulkarni04,agakov06}.
-Applying iterative optimisation in light of a single input may not result in good performance when executing the optimised code with different inputs.
+Applying {\itercomp} in light of a single input may not result in good performance when executing the optimised code with different inputs.
 
@@
 %We can extend the same argument when optimising the code for performance.
-using a single input for iterative optimisation can produce poor performance when executed with different inputs.
+using a single input for {\itercomp} can produce poor performance when executed with different inputs.
 
@@
 the impact of using multiple input datasets
-for performing iterative optimisation.
+for performing {\itercomp}.
 This previous work suggests that optimising based on a representative range of input datasets
@@
 
-Finding such a robust combination of compiler optimisations by means of iterative optimisation across a large range of inputs
+Finding such a robust combination of compiler optimisations by means of {\itercomp} across a large range of inputs
 may be very time consuming.
@@
     \includegraphics[width=\linewidth]{figs/diagram}
-    \caption{Overview of the execution engine for applying iterative optimisation.}
+    \caption{Overview of the execution engine for applying {\itercomp}.}
     \label{fig:diagram}
@@
 
-In this paper, our main goal is to enable iterative optimisation in
+In this paper, our main goal is to enable {\itercomp} in
 \textit{online} scenarios with the restriction of executing distinct inputs
@@
 \begin{itemize}
-\item The use of a work-based metric in order to enable online iterative
-optimisation by comparing different combination of compiler optimisations even
-when executed with distinct inputs.
+\item The use of a work-based performance metric in order to enable
+\textit{online} {\itercomp} by comparing different combination of
+compiler optimisations even when executed with distinct inputs.
 \item We propose a relaxed instrumentation for low overhead profiling, with a controlled
@@
 
-\section{Related Work}
-
-Chen~\etal~\cite{chen10,chen12a} evaluate the effectiveness of iterative
-compilations across a large number of input test cases.
-Their main motivation is to answer the question:
-\textit{How data input dependent is iterative optimisation?}
-Their results show that it is possible to find a combination of compiler optimisations
-that achieves at least 86\% of the maximum speedup across all input test cases.
-These optimal combinations are program-specific and yield average
-speedups up to 3.75$\times$ over the highest optimisation level of compilers.
-
-When optimising a program, the main method for iterative optimisation used by Chen~\etal~\cite{chen10,chen12a}
-evaluates each combination of compiler optimisations across all the available inputs, i.e.,
-if $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimisations,
-they perform a total of $O(NM)$ runs of the program being optimised.
-Furthermore, they use a pre-defined set of only 300 different combinations of compiler optimisations,
-which represents a very small sample of the optimisation search space for most modern compilers, e.g.
-LLVM has 56 distinct optimisation passes and GCC has about 47 high-level (SSA form) optimisation passes and
-about 25 low-level (RTL) optimisation passes, which in both cases result in much more than $2^{50}$ distinct
-combinations of compiler optimisations, without considering repetition.
-
-Recent work~\cite{chen12b,fang15} have applied the same idea of performing
-input-dependent iterative optimisation to distributed applications on data centres.
-In summary, each worker receives a subset of the input dataset, called the evaluation dataset,
-to perform an \textit{online} iterative optimisation of the code being executed.
-Each worker performs the same the method for iterative optimisation used by Chen~\etal~\cite{chen10,chen12a},
-i.e., they evaluate each combination of compiler optimisations across all the evaluation dataset.
-However, because the optimisation is performed online, they usually consider a small evaluation dataset
-and a small number of compiler optimisations.
-
-Fursin~\etal~\cite{fursin07} addressed the problem of comparing the effect of two optimisations
-on two distinct inputs. For that purpose, they proposed to use instruction per cycle (IPC) as the metric for performing such comparison.
-Their result show that using IPC seems promising as a robust metric for iterative optimisation
-across large input datasets.
-However, 
-some specific optimisation techniques may affect the use of IPC as a robust metric, and specially
-IPC has been shown to provide poor performance estimation for multi-threaded programs~\cite{alameldeen06,eyerman08}.
-In particular, IPC can give a skewed performance measure if threads spend time in \textit{spin-lock loops}
-or other synchronisation mechanisms. 
-Some existing work on performance assessment suggest that
-total execution time should be used for measuring performance of multi-threaded programs~\cite{alameldeen06,eyerman08}.
-Aalameldeen and Wood~\cite{alameldeen06} in particular suggest that a simple work-related metric should be used
-if the unit of work is representative enough.
-Work-related metrics have already been largely used for measuring performance of throughput-oriented applications,
-for other applications, however, choosing an appropriate unit of work can be more challanging~\cite{alameldeen06}.
-
 \section{Work-based Metric} \label{sec:metric}
 
-A possible candidate for a work-based metric, borrowing from Physics,
-is using the concept of \textit{average power}, i.e. the amount of
+In this section we define the work-based performance metric used in this paper.
+We define the performance metric as the ratio between the amount of
 \textit{work}, $\Delta W$, performed during a period of time, $\Delta t$.
 \[
-   P_{avg} = \frac{\Delta W}{\Delta t}
+   P = \frac{\Delta W}{\Delta t}
 \]
@@
 of the optimised program.
-For this metric, the main challange is to precisely define what
+For this metric, the main challenge is to precisely define what
 represents \textit{work}.
@@
 
-%In addition to being useful for speeding up iterative optimisation across a large
+%In addition to being useful for speeding up {\itercomp} across a large
 %number of different inputs, this metric based on comparing work-based efficiencies could
-%potentially be used for applying iterative optimisation \textit{online}. 
+%potentially be used for applying {\itercomp} \textit{online}. 
 %However, the overhead for estimating the amount of \textit{work} must be low
 %enough for the performance gains to be beneficial.
-
-Previous work have proposed profiling-based mechanism to estimate
-input sizes~\cite{zaparanuks12,coppa14}.
-Coppa~\etal~\cite{coppa14} in particular propose the concept of
-\textit{read memory size} for automatically estimating the size of the input
-passed to a routine, where \textit{read memory size} represents the number of
-distinct memory cells first accessed by a read operation.
-In other words, the \textit{read memory size} metric measures the size of the 
-useful portion of the input's memory footprint.
-However, because we are interested in the amount of computational work performed
-in respect of a given input, the memory footprint of the input may not always
-have a direct correspondence to  the amount of computational work.
-
-Goldsmith~\etal~\cite{goldsmith07} use \textit{block frequency} as the
-measure for performance for empirically describing the asymptotic behaviour
-of programs, which is known as empirical computational complexity.
-Block frequency is a relative metric that represents the number of times a
-basic block executes~\cite{ball94,ball96}. 
-They argue in favour of block frequency due to its portability, repeatability
-and exactness, since it does not suffer from timer resolution problems or
-non-deterministic noises.
-Block frequency also has the advantage of being efficiently profiled by means
-of automatic code instrumentation~\cite{knuth73,ball94}.
-
-However, in the context of comparing different optimisations,
-although block frequency would be able to capture aspects of optimisations
-that simplify the control-flow graph (CFG), measuring work at the basic block
-resolution would not capture effects of optimisations at the instruction level.
-Because of that, we extend the idea of using basic block frequency to measure
-computational work by also considering the computational cost of each
-basic block. The computational cost of a basic block is given by weighing
-the instructions that it contains.
 
@@
 
+Similarly to previous work~\cite{giusto01,powell09,brandolese11},
+we derive the cost model for the instruction set by modeling the
+problem as a multi-variable linear regression, where the
+\textit{regression coefficients} are the costs of the instructions
+and the \textit{regressors} (or \textit{explanatory variables})
+are computed as $\sum_B N_B(i)f(B)$ for each instruction $i$.
+
 \section{Instrumentation}
@@
 performed during runtime by means of instrumenting the code.
-%with the necessary computation for the work metric, $\Delta W$, as described in Section~\ref{sec:metric}.
+In particular, we adapt the optimal algorithm proposed originally 
+for profiling block frequency~\cite{knuth73,ball94}.
+Afterwards, we propose a relaxed instrumentation that focus on
+further reducing the overhead by considering the trade-off between
+profiling accuracy and instrumentation overhead.
+
 Because we define work as a linear equation on the block frequency counters,
@@
 is easily implemented, it inserts a large amount of overhead into the program.
-
-In 1973, Donald Knuth~\cite{knuth73} provided an optimal algorithm for
-profiling block frequency, inserting fewer probes than the previously described
-naive instrumentation that was commonly used in practice~\cite{knuth71}.
-Forman~\cite{forman81} proposed further overhead reductions by placing the
-probes in basic blocks that are less likely to be executed based on static
-heuristics.
-Ball and Larus~\cite{ball94} offered a detailed discussion comparing
-two approaches for optimally profiling block frequency, namely,
-placing probes on the vertices or the edges of the control-flow graph (CFG).
-They show that the edge-based approach produces optimal placement of probes.
+However, it is possible to insert fewer probes by carefully placing the probes
+in a way that is possible to reconstruct
+the complete profiling information~\cite{knuth73,ball94}.
 
@@
 our work metric.
-Finally we present the results of the online iterative optimisation.
+Finally we present the results of the online {\itercomp}.
 
@@
 
-%Iterative optimisations is essentially based on repeatedly trying
+%{\Itercomp} is essentially based on repeatedly trying
 %out a large number of compiler optimisations until the best
 %combination of compiler optimisations is found for a particular program.
-%Our main goal is to speed up iterative optimisation while targeting
+%Our main goal is to speed up {\itercomp} while targeting
 %optimal performance across large input datasets.
@@
 %\vspace{1ex}
-%\noindent \textbf{Iterative optimisation over a single input}
+%\noindent \textbf{{\Itercomp} over a single input}
 
-%Most of the existing iterative optimisation studies find the
+%Most of the existing {\itercomp} studies find the
 %best optimisation though repeated runs on the same input.
@@
 %\vspace{1ex}
-%\noindent \textbf{Iterative optimisation across large input datasets}
+%\noindent \textbf{{\Itercomp} across large input datasets}
 
-%Recent work on iterative optimisation have been targeting optimisation
+%Recent work on {\itercomp} have been targeting optimisation
 %across multiple inputs~\cite{fursin07,chen10,chen12a}.
@@
 %\vspace{1ex}
-%\noindent \textbf{Iterative optimisation based on the IPC metric}
+%\noindent \textbf{{\Itercomp} based on the IPC metric}
 
-%While the previous two baselines addresses two opposite aspects of iterative optimisation,
+%While the previous two baselines addresses two opposite aspects of {\itercomp},
 %namely, compile-time efficiency and performance of the generated code,
@@
 
-\subsection{Evaluation of the Online Iterative Optimisation}
+\subsection{Evaluation of the Online {\IterComp}}
 
@@
 This version uses the work-based metric, $\frac{\Delta W}{\Delta t}$, during
-the iterative optimisation search.
+the {\itercomp} search.
 \end{itemize}
@@
     \includegraphics[width=\textwidth]{figs/speedups.pdf}
-    \caption{Speedups observed with the online iterative optimisation.}
+    \caption{Speedups observed with the online {\itercomp}.}
     \label{fig:speedups}
@@
 %    \includegraphics[width=\textwidth]{figs/profiled-speedups.pdf}
-%    \caption{Speedups observed with the online iterative optimisation if we
+%    \caption{Speedups observed with the online {\itercomp} if we
 %             consider the instrumentation overhead.}
@@
 %Figure~\ref{fig:flagsfreq} shows an aggregated view of the final combination
-%of compiler optimisations that were selected by iterative optimisation search.
+%of compiler optimisations that were selected by {\itercomp} search.
 %The figure presents the individual optimisation passes with at least 1\% of
@@
 %    \caption{Frequency of individual optimisation passes on the final selected 
-%             compiler optimisations of the iterative optimisation search over
+%             compiler optimisations of the {\itercomp} search over
 %             all benchmarks.}
@@
 %\end{figure*}
+
+\section{Related Work}
+
+\subsection{{\IterComp}}
+
+Chen~\etal~\cite{chen10,chen12a} evaluate the effectiveness of iterative
+compilations across a large number of input test cases.
+Their main motivation is to answer the question:
+\textit{How data input dependent is {\itercomp}?}
+Their results show that it is possible to find a combination of compiler optimisations
+that achieves at least 86\% of the maximum speedup across all input test cases.
+These optimal combinations are program-specific and yield average
+speedups up to 3.75$\times$ over the highest optimisation level of compilers.
+
+When optimising a program, the main method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a}
+evaluates each combination of compiler optimisations across all the available inputs, i.e.,
+if $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimisations,
+they perform a total of $O(NM)$ runs of the program being optimised.
+Furthermore, they use a pre-defined set of only 300 different combinations of compiler optimisations,
+which represents a very small sample of the optimisation search space for most modern compilers, e.g.
+LLVM has 56 distinct optimisation passes and GCC has about 47 high-level (SSA form) optimisation passes and
+about 25 low-level (RTL) optimisation passes, which in both cases result in much more than $2^{50}$ distinct
+combinations of compiler optimisations, without considering repetition.
+
+Recent work~\cite{chen12b,fang15} have applied the same idea of performing
+input-dependent {\itercomp} to distributed applications on data centres.
+In summary, each worker receives a subset of the input dataset, called the evaluation dataset,
+to perform an \textit{online} {\itercomp} of the code being executed.
+Each worker performs the same the method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a},
+i.e., they evaluate each combination of compiler optimisations across all the evaluation dataset.
+However, because the optimisation is performed online, they usually consider a small evaluation dataset
+and a small number of compiler optimisations.
+
+Fursin~\etal~\cite{fursin07} addressed the problem of comparing the effect of two optimisations
+on two distinct inputs. For that purpose, they proposed to use instruction per cycle (IPC) as the metric for performing such comparison.
+Their result show that using IPC seems promising as a robust metric for {\itercomp}
+across large input datasets.
+However, 
+some specific optimisation techniques may affect the use of IPC as a robust metric, and specially
+IPC has been shown to provide poor performance estimation for multi-threaded programs~\cite{alameldeen06,eyerman08}.
+In particular, IPC can give a skewed performance measure if threads spend time in \textit{spin-lock loops}
+or other synchronisation mechanisms. 
+Some existing work on performance assessment suggest that
+total execution time should be used for measuring performance of multi-threaded programs~\cite{alameldeen06,eyerman08}.
+Aalameldeen and Wood~\cite{alameldeen06} in particular suggest that a simple work-related metric should be used
+if the unit of work is representative enough.
+Work-related metrics have already been largely used for measuring performance of throughput-oriented applications,
+for other applications, however, choosing an appropriate unit of work can be more challanging~\cite{alameldeen06}.
+
+\subsection{Work and Input Size Metrics}
+
+Previous work have proposed profiling-based mechanism to estimate
+input sizes~\cite{zaparanuks12,coppa14}.
+Coppa~\etal~\cite{coppa14} in particular propose the concept of
+\textit{read memory size} for automatically estimating the size of the input
+passed to a routine, where \textit{read memory size} represents the number of
+distinct memory cells first accessed by a read operation.
+In other words, the \textit{read memory size} metric measures the size of the 
+useful portion of the input's memory footprint.
+However, because we are interested in the amount of computational work performed
+in respect of a given input, the memory footprint of the input may not always
+have a direct correspondence to  the amount of computational work.
+
+Goldsmith~\etal~\cite{goldsmith07} use \textit{block frequency} as the
+measure for performance for empirically describing the asymptotic behaviour
+of programs, which is known as empirical computational complexity.
+Block frequency is a relative metric that represents the number of times a
+basic block executes~\cite{ball94,ball96}. 
+They argue in favour of block frequency due to its portability, repeatability
+and exactness, since it does not suffer from timer resolution problems or
+non-deterministic noises.
+Block frequency also has the advantage of being efficiently profiled by means
+of automatic code instrumentation~\cite{knuth73,ball94}.
+
+However, in the context of comparing different optimisations,
+although block frequency would be able to capture aspects of optimisations
+that simplify the control-flow graph (CFG), measuring work at the basic block
+resolution would not capture effects of optimisations at the instruction level.
+Because of that, we extend the idea of using basic block frequency to measure
+computational work by also considering the computational cost of each
+basic block. The computational cost of a basic block is given by weighing
+the instructions that it contains.
+
+\subsection{Optimal Instrumentation}
+
+In 1973, Donald Knuth~\cite{knuth73} provided an optimal algorithm for
+profiling block frequency, inserting fewer probes than the previously described
+naive instrumentation that was commonly used in practice~\cite{knuth71}.
+Forman~\cite{forman81} proposed further overhead reductions by placing the
+probes in basic blocks that are less likely to be executed based on static
+heuristics.
+
+Ball and Larus~\cite{ball94} offered a detailed discussion comparing
+two approaches for optimally profiling block frequency, namely,
+placing probes on the vertices or the edges of the control-flow graph (CFG).
+They show that the edge-based approach produces optimal placement of probes.
 

### 2017-05-09 cgo18mscr.tex
minor change: organising line breaks

@@
 \begin{document}
-%\title{A work-based metric for efficiently applying online iterative optimisation}
 \title{Online {\IterComp}\\Guided by Low Overhead Profiling}
@@
 
-Modern optimising compilers
-have reached a high level of sophistication,
-providing a large number of optimisations.
-Because of the unpredictability of optimisation
-interactions, in addition to the growing complexity
-of processor architectures and applications,
-the correct choice of optimisations and their ordering
-can have a significant impact on the performance of the
-code being optimised~\cite{pan06,fursin07,kulkarni12}.
+Modern optimising compilers have reached a high level of sophistication, providing a large number of optimisations.
+Because of the unpredictability of optimisation interactions, in addition to the growing complexity of processor architectures and applications, the correct choice of optimisations and their ordering can have a significant impact on the performance of the code being optimised~\cite{pan06,fursin07,kulkarni12}.
 
-Each of these optimisations interacts with the code in complicated ways,
-depending on all other optimisations and the order they were applied to
-the code being optimised.
-Understanding the interactions of optimisations is very important
-in determining a good solution to the phase-ordering problem~\cite{touati06,kulkarni12}.
-Because of all those dependences and interactions, although
-most compiler optimisations yield significant improvements
-in many programs, the potential for performance degradation in
-certain program patterns is known to compiler writers and many users~\cite{pan06,zhou12,kulkarni12}.
+Each of these optimisations interacts with the code in complicated ways, depending on all other optimisations and the order they were applied to the code being optimised.
+Understanding the interactions of optimisations is very important in determining a good solution to the phase-ordering problem~\cite{touati06,kulkarni12}.
+Because of all those dependences and interactions, although most compiler optimisations yield significant improvements in many programs, the potential for performance degradation in certain program patterns is known to compiler writers and many users~\cite{pan06,zhou12,kulkarni12}.
 %Traditional compilers typically apply the same set of optimisation in a fixed order to all functions in a program, without regard the code being optimised.
 
-Efficiently selecting the best combination of compiler optimisations
-for a given program or program section remains an open problem.
-Compiler writers typically use a combination of experience and insight to construct
-the default sequences of prearranged optimisations found in compilers.
-However, these default pre-arranged set of optimisations do not include all possible optimisations
-and they are always applied in the same pre-defined fixed order, without regard the code being optimised~\cite{pan06,cavazos07,zhou12,kulkarni12}.
+Efficiently selecting the best combination of compiler optimisations for a given program or program section remains an open problem.
+Compiler writers typically use a combination of experience and insight to construct the default sequences of prearranged optimisations found in compilers.
+However, these default pre-arranged set of optimisations do not include all possible optimisations and they are always applied in the same pre-defined fixed order, without regard the code being optimised~\cite{pan06,cavazos07,zhou12,kulkarni12}.
 %in hopes that a programmer will know which combination of optimisations will benefit their code~\cite{pan06,cavazos07,zhou12,kulkarni12}.
 
-A well known compilation technique that
-addresses this challenge is {\itercomp}.
-{\Itercomp} has the ability to adapt to new
-platforms, program and workload while still
-having a systematic and simple optimisation process.
-{\Itercomp} repeatedly evaluates a large number
-of compiler optimisations until the best combination is
-found for a particular program~\cite{fursin07,chen10}.
-The main challenge concerning {\itercomp} is the need for quickly
-exploring such a large optimisation space~\cite{fursin07,cavazos07,zhou12}.
+A well known compilation technique that addresses this challenge is {\itercomp}.
+{\Itercomp} has the ability to adapt to new platforms, program and workload while still having a systematic and simple optimisation process.
+It works by repeatedly evaluating a large number of compiler optimisations until the best combination is found for a particular program~\cite{fursin07,chen10}.
+The main challenge concerning {\itercomp} is the need for efficiently exploring such a large optimisation space~\cite{fursin07,cavazos07,zhou12}.
 
-Until recently, most of existing work  had been focusing
-on finding the best optimisation through repeated runs using a single input.
-Although they demonstrate the potential of {\itercomp},
-in real scenarios the user rarely executes the same input dataset multiple times~\cite{bodin98,kisuki99,stephenson03,kulkarni04,agakov06}.
+Until recently, most of existing work  had been focusing on finding the best optimisation through repeated runs using a single input.
+Although they demonstrate the potential of {\itercomp}, in real scenarios the user rarely executes the same input dataset multiple times~\cite{bodin98,kisuki99,stephenson03,kulkarni04,agakov06}.
 Applying {\itercomp} in light of a single input may not result in good performance when executing the optimised code with different inputs.
 
-Most of real world applications are complex enough 
-so that a single input case does not capture the whole
-range of possible scenarios and program behaviour~\cite{haneda06,fursin07,chen10,chen12a}.
+Most of real world applications are complex enough so that a single input case does not capture the whole range of possible scenarios and program behaviour~\cite{haneda06,fursin07,chen10,chen12a}.
 Because programs can exhibit behaviours that differ greatly depending on the input,
-%This is something that is well known when writing test
-%cases for correctness.
+%This is something that is well known when writing test cases for correctness.
 %We can extend the same argument when optimising the code for performance.
@@
 
-Recent work have been studying
-the impact of using multiple input datasets
+Recent work have been studying the impact of using multiple input datasets
 for performing {\itercomp}.
-This previous work suggests that optimising based on a representative range of input datasets
-allows for selecting a robust compiler optimisation 
-that produces good performance in real scenarios where
-the input is expected to change constantly~\cite{haneda06,fursin07,chen10,chen12a,chen12b,fang15,mpeis16}.
-Their results show that a limited number of input datasets may be sufficient to
-obtain a well optimised program for a wider range of different inputs~\cite{haneda06,fursin07,chen10,chen12a}.
+This previous work suggests that optimising based on a representative range of input datasets allows for selecting a robust compiler optimisation that produces good performance in real scenarios where the input is expected to change constantly~\cite{haneda06,fursin07,chen10,chen12a,chen12b,fang15,mpeis16}.
+Their results show that a limited number of input datasets may be sufficient to obtain a well optimised program for a wider range of different inputs~\cite{haneda06,fursin07,chen10,chen12a}.
 
-Finding such a robust combination of compiler optimisations by means of {\itercomp} across a large range of inputs
-may be very time consuming.
-In order to speed up this process we intend to reduce the number of actual executions during the exploration
-of the optimisation space without much degradation of the final selected optimisation.
+Finding such a robust combination of compiler optimisations by means of {\itercomp} across a large range of inputs may be very time consuming.
+In order to speed up this process we intend to reduce the number of actual executions during the exploration of the optimisation space without much degradation of the final selected optimisation.
 
@@
 
-In this paper, our main goal is to enable {\itercomp} in
-\textit{online} scenarios with the restriction of executing distinct inputs
-only once, while targeting optimal performance across different inputs.
-This online aspect is usually found in mobile and data centre
-platforms~\cite{chen12b,fang15,mpeis16}, where the goal is to optimise
-programs based on the workload of a particular user (or group of users)
-between executions.
+In this paper, our main goal is to enable {\itercomp} in \textit{online} scenarios with the restriction of executing distinct inputs only once, while targeting optimal performance across different inputs.
+This online aspect is usually found in mobile and data centre platforms~\cite{chen12b,fang15,mpeis16}, where the goal is to optimise programs based on the workload of a particular user (or group of users) between executions.
 
-Measuring just execution time, for example, is useful only if the
-amount of work is constant between executions with different inputs.
-For this reason, we propose the use of a work-based metric in order
-to compare the performance of different optimisations across 
-multiple executions of the program with distinct inputs.
+Measuring just execution time, for example, is useful only if the amount of work is constant between executions with different inputs.
+For this reason, we propose the use of a work-based metric in order to compare the performance of different optimisations across multiple executions of the program with distinct inputs.
 
-Because of the restriction of not repeating inputs, we instrument the program
-for measuring the amount of work it performed during its execution.
-Having a low overhead instrumentation is essential in this online scenario for
-two main reasons:
+Because of the restriction of not repeating inputs, we instrument the program for measuring the amount of work it performed during its execution.
+Having a low overhead instrumentation is essential in this online scenario for two main reasons:
 $(i)$ the user is directly affected by large overheads;
-$(ii)$ a highly intrusive instrumentation can havea significant impact on the
-effect of the optimisations.
+$(ii)$ a highly intrusive instrumentation can havea significant impact on the effect of the optimisations.
 
@@
 \begin{itemize}
-\item The use of a work-based performance metric in order to enable
-\textit{online} {\itercomp} by comparing different combination of
-compiler optimisations even when executed with distinct inputs.
-\item We propose a relaxed instrumentation for low overhead profiling, with a controlled
-trade-off between accuracy and overhead.
+\item The use of a work-based performance metric in order to enable \textit{online} {\itercomp} by comparing different combination of compiler optimisations even when executed with distinct inputs.
+\item We propose a relaxed instrumentation for low overhead profiling, with a controlled trade-off between accuracy and overhead.
 \end{itemize}
@@
 In this section we define the work-based performance metric used in this paper.
-We define the performance metric as the ratio between the amount of
-\textit{work}, $\Delta W$, performed during a period of time, $\Delta t$.
+We define the performance metric as the ratio between the amount of \textit{work}, $\Delta W$, performed during a period of time, $\Delta t$.
 \[
@@
 
-%The hypothesis is that, given two optimisations $o_1$ and $o_2$,
-%a program compiled with optimisation $o_2$ would \textit{consistently} perform better
-%than when it is compiled with $o_1$, i.e., $o_1 \tilde{<}_p \, o_2$,
-%if it performs more \textit{work} per unit of time when compiled with $o_2$
-%instead of $o_1$. 
-By measuring the amount of \textit{work} done per unit of time we reduce
-the impact of input-dependent aspects and focus instead on the efficiency
-of the optimised program.
-For this metric, the main challenge is to precisely define what
-represents \textit{work}.
+%The hypothesis is that, given two optimisations $o_1$ and $o_2$, a program compiled with optimisation $o_2$ would \textit{consistently} perform better than when it is compiled with $o_1$, i.e., $o_1 \tilde{<}_p \, o_2$, if it performs more \textit{work} per unit of time when compiled with $o_2$ instead of $o_1$.
+By measuring the amount of \textit{work} done per unit of time we reduce the impact of input-dependent aspects and focus instead on the efficiency of the optimised program.
+For this metric, the main challenge is to precisely define what represents \textit{work}.
 
-%For that purpose, we intend to use profiling
-%techniques to measure the input size and amount of computation
-%performed on the execution path triggered by the input~\cite{ball94,ball96,zaparanuks12,coppa14}.
+%For that purpose, we intend to use profiling techniques to measure the input size and amount of computation performed on the execution path triggered by the input~\cite{ball94,ball96,zaparanuks12,coppa14}.
 
-%In addition to being useful for speeding up {\itercomp} across a large
-%number of different inputs, this metric based on comparing work-based efficiencies could
+%In addition to being useful for speeding up {\itercomp} across a large number of different inputs, this metric based on comparing work-based efficiencies could
 %potentially be used for applying {\itercomp} \textit{online}.
-%However, the overhead for estimating the amount of \textit{work} must be low
-%enough for the performance gains to be beneficial.
+%However, the overhead for estimating the amount of \textit{work} must be low enough for the performance gains to be beneficial.
 
-We model the computational work $\Delta W$ as a linear equation based on
-block frequency information and a cost-model of the instruction set.
+We model the computational work $\Delta W$ as a linear equation based on block frequency information and a cost-model of the instruction set.
 \[
@@
 \]
-where $f(B)$ represents the frequency of basic block $B$
-and $w(B)$ represents the computational work of executing $B$.
-We define the work of a basic block $B$ as the sum of the cost
-of its instructions, i.e.,
+where $f(B)$ represents the frequency of basic block $B$ and $w(B)$ represents the computational work of executing $B$.
+We define the work of a basic block $B$ as the sum of the cost of its instructions, i.e.,
 \[
@@
 \]
-where $w_i$ is the cost of instruction $i$ and $N_B(i)$ is the number of
-occurrences of instruction $i$ in basic block $B$.
+where $w_i$ is the cost of instruction $i$ and $N_B(i)$ is the number of occurrences of instruction $i$ in basic block $B$.
 
-In this simplified model, we consider that $w_i$ is constant across all
-programs and executions in the given target platform.
-However, $N_B(i)$ is program dependent but constant across executions,
-while $f(B)$ is both program and execution dependent, since $f(B)$ can
-change when executing with different inputs.
-In other words, $N_B(i)$ is a static value known at compile-time and 
-$f(B)$ is a dynamic value known only at run-time.
+In this simplified model, we consider that $w_i$ is constant across all programs and executions in the given target platform.
+However, $N_B(i)$ is program dependent but constant across executions, while $f(B)$ is both program and execution dependent, since $f(B)$ can change when executing with different inputs.
+In other words, $N_B(i)$ is a static value known at compile-time and $f(B)$ is a dynamic value known only at run-time.
 %If we define $x_i$ as
@@
 
-Similarly to previous work~\cite{giusto01,powell09,brandolese11},
-we derive the cost model for the instruction set by modeling the
-problem as a multi-variable linear regression, where the
-\textit{regression coefficients} are the costs of the instructions
-and the \textit{regressors} (or \textit{explanatory variables})
-are computed as $\sum_B N_B(i)f(B)$ for each instruction $i$.
+Similarly to previous work~\cite{giusto01,powell09,brandolese11}, we derive the cost model for the instruction set by modeling the problem as a multi-variable linear regression, where the \textit{regression coefficients} are the costs of the instructions and the \textit{regressors} (or \textit{explanatory variables}) are computed as $\sum_B N_B(i)f(B)$ for each instruction $i$.
 
@@
 
-In this section we describe how the computation of the work-metric can be
-performed during runtime by means of instrumenting the code.
-In particular, we adapt the optimal algorithm proposed originally 
-for profiling block frequency~\cite{knuth73,ball94}.
-Afterwards, we propose a relaxed instrumentation that focus on
-further reducing the overhead by considering the trade-off between
-profiling accuracy and instrumentation overhead.
+In this section we describe how the computation of the work-metric can be performed during runtime by means of instrumenting the code.
+In particular, we adapt the optimal algorithm proposed originally for profiling block frequency~\cite{knuth73,ball94}.
+Afterwards, we propose a relaxed instrumentation that focus on further reducing the overhead by considering the trade-off between profiling accuracy and instrumentation overhead.
 
-Because we define work as a linear equation on the block frequency counters,
-it is possible to embed its computation into the execution of the program.
-A naive instrumentation would consists basically of having a global counter
-that starts with the interception value, $\varepsilon$, and each basic block
-increments its own cost into the global counter. Although this instrumentation
-is easily implemented, it inserts a large amount of overhead into the program.
-However, it is possible to insert fewer probes by carefully placing the probes
-in a way that is possible to reconstruct
-the complete profiling information~\cite{knuth73,ball94}.
+Because we define work as a linear equation on the block frequency counters, it is possible to embed its computation into the execution of the program.
+A naive instrumentation would consists basically of having a global counter that starts with the interception value, $\varepsilon$, and each basic block increments its own cost into the global counter.
+Although this instrumentation is easily implemented, it inserts a large amount of overhead into the program.
+However, it is possible to insert fewer probes by carefully placing the probes in a way that is possible to reconstruct the complete profiling information~\cite{knuth73,ball94}.
 
-Ball and Larus~\cite{ball94} show that a set of edges represents the minimum
-number of probes for profiling block frequency if and only if the complementary
-set of edges forms a spanning tree.
-The instrumentation optimises the placement of the probes with respect to a
-weighing that assigns a non-negative value to each edge in the CFG.
-The cost of profiling a set of edges is proportional to the sum of the weights
-of the edges.
-These weights can be obtained either by empirical measurements or heuristic
-estimations.
-In order to minimise the profiling cost, the instrumentation computes the
-maximum spanning tree for avoiding probing in frequently executed edges.
-Although this instrumentation algorithm is proved to produce the optimal
-placement of probes for well-structured CFGs, it may produce sub-optimal
-placement for some unstructured CFGs.
+Ball and Larus~\cite{ball94} show that a set of edges represents the minimum number of probes for profiling block frequency if and only if the complementary set of edges forms a spanning tree.
+The instrumentation optimises the placement of the probes with respect to a weighing that assigns a non-negative value to each edge in the CFG.
+The cost of profiling a set of edges is proportional to the sum of the weights of the edges.
+These weights can be obtained either by empirical measurements or heuristic estimations.
+In order to minimise the profiling cost, the instrumentation computes the maximum spanning tree for avoiding probing in frequently executed edges.
+Although this instrumentation algorithm is proved to produce the optimal placement of probes for well-structured CFGs, it may produce sub-optimal placement for some unstructured CFGs.
 
-Once we have the maximum spanning tree, probes are placed on any edge not in 
-the spanning tree. Figure~\ref{fig:cfg-example} shows an example of a CFG
-with a maximum spanning tree represented by the black edges, while the edges 
-highlighted in red represent the placement of the probes.
+Once we have the maximum spanning tree, probes are placed on any edge not in the spanning tree.
+Figure~\ref{fig:cfg-example} shows an example of a CFG with a maximum spanning tree represented by the black edges, while the edges highlighted in red represent the placement of the probes.
 
@@
 
-In contrast to the naive instrumentation where each basic block records only
-its own amount of work, with the optimal profiling, the instrumented basic blocks
-need record an aggregated value of work that represents a path in the CFG.
-These values are constructed with some instrumented basic blocks speculatively
-assuming some paths while other probes correct when these assumptions are wrong
-(see for example $w(P_0)$ and $w(P_1)$ in Figure~\ref{fig:cfg-example}).
+In contrast to the naive instrumentation where each basic block records only its own amount of work, with the optimal profiling, the instrumented basic blocks need record an aggregated value of work that represents a path in the CFG.
+These values are constructed with some instrumented basic blocks speculatively assuming some paths while other probes correct when these assumptions are wrong (see for example $w(P_0)$ and $w(P_1)$ in Figure~\ref{fig:cfg-example}).
 
@@
 
-Because the algorithm for the optimal placement of the probes is proved to
-uniquely compute the block frequencies by propagating the probe counts,
-we adapt this algorithm in order to compose the aggregated values that will
-be instrumented in each probe, based on our model of computational work,
-$\Delta W$, derived from the basic block frequencies (see Section~\ref{sec:metric}).
-We perform a similar propagation of the probes in a symbolic fashion,
-as illustrated in Figure~\ref{fig:cfg-example}.
-This symbolic propagation of the probes is implemented by the data-flow analysis
-described in Algorithm~\ref{alg:populateEdgeInfo}, and the final aggregated values
-are extracted from the edge information as described by Algorithm~\ref{alg:instrValue}.
+Because the algorithm for the optimal placement of the probes is proved to uniquely compute the block frequencies by propagating the probe counts, we adapt this algorithm in order to compose the aggregated values that will be instrumented in each probe, based on our model of computational work, $\Delta W$, derived from the basic block frequencies (see Section~\ref{sec:metric}).
+We perform a similar propagation of the probes in a symbolic fashion, as illustrated in Figure~\ref{fig:cfg-example}.
+This symbolic propagation of the probes is implemented by the data-flow analysis described in Algorithm~\ref{alg:populateEdgeInfo}, and the final aggregated values are extracted from the edge information as described by Algorithm~\ref{alg:instrValue}.
 
-The data-flow analysis in Algorithm~\ref{alg:populateEdgeInfo} keeps two sets
-for each edge, namely, the increment and the decrement sets.
-We consider that both sets represent the \textit{edge expressions} shown in Figure~\ref{fig:cfg-example},
-for which we define a \textit{symbolic sum} by computing the union of the increment
-and decrement sets, respectively, with the appropriate cancellation of common elements.
-This data-flow analysis is based on the invariant that the symbolic sum
-of all the incoming edges must equals the symbolic sum of the outgoing
-edges. For example, the symbolic sum of the incoming edges of the basic
-block $B_8$ is $P_0 - P_1$, where $P_3$ is cancelled out.
+The data-flow analysis in Algorithm~\ref{alg:populateEdgeInfo} keeps two sets for each edge, namely, the increment and the decrement sets.
+We consider that both sets represent the \textit{edge expressions} shown in Figure~\ref{fig:cfg-example}, for which we define a \textit{symbolic sum} by computing the union of the increment and decrement sets, respectively, with the appropriate cancellation of common elements.
+This data-flow analysis is based on the invariant that the symbolic sum of all the incoming edges must equals the symbolic sum of the outgoing edges.
+For example, the symbolic sum of the incoming edges of the basic block $B_8$ is $P_0 - P_1$, where $P_3$ is cancelled out.
 
-%By construction, 
-%the \textit{symbolic experssions} of a basic block, i.e., 
-%the \textit{symbolic sum} of the incoming edges (or outgoing edges) of a basic block,
-%has the following properties:
+%By construction, the \textit{symbolic experssions} of a basic block, i.e., the \textit{symbolic sum} of the incoming edges (or outgoing edges) of a basic block, has the following properties:
 %\begin{enumerate}
@@
 %\item Probes in the decrement set are (post-)dominated by the probes in the increment set;
-%\item Probes in the decrement set and the given basic block are in exclusive paths
-%      from the virtual node of the CFG to the (post-)dominators in the increment set;
+%\item Probes in the decrement set and the given basic block are in exclusive paths from the virtual node of the CFG to the (post-)dominators in the increment set;
 %\end{enumerate}
-%From property (1) it follows that whenever the given basic block is executed,
-%exactly one of the probes in the increment set is executed.
-%From properties (2) and (3), it follows that whenever a probe in the decrement set
-%is executed, although the given basic block is not executed, one of the probes
-%in the decrement set is executed.
+%From property (1) it follows that whenever the given basic block is executed, exactly one of the probes in the increment set is executed.
+%From properties (2) and (3), it follows that whenever a probe in the decrement set is executed, although the given basic block is not executed, one of the probes in the decrement set is executed.
 
-Algorithm~\ref{alg:instrValue} reads the edge information for each
-basic block by computing the symbolic sum of their respective incoming
-edges (or outgoing edges).
-From these \textit{edge expressions}, we are able to compose the aggregated
-value of the probes.
-The positive terms in the edge expression of a basic block indicate that the
-amount of work of this basic block will be incremented in the probes represented
-by these positive terms, similarly, the negative terms indicate that the amount
-of work of this basic block will be decremented in the probes represented by
-these negative terms.
-For example, because the edge experssion for the basic block $B_8$ is
-$P_0 - P_1$, the amount of work of $B_8$, denoted by $w(B_8)$,
-is incremented in probe $P_0$ and decremented in $P_1$.
+Algorithm~\ref{alg:instrValue} reads the edge information for each basic block by computing the symbolic sum of their respective incoming edges (or outgoing edges).
+From these \textit{edge expressions}, we are able to compose the aggregated value of the probes.
+The positive terms in the edge expression of a basic block indicate that the amount of work of this basic block will be incremented in the probes represented by these positive terms, similarly, the negative terms indicate that the amount of work of this basic block will be decremented in the probes represented by these negative terms.
+For example, because the edge experssion for the basic block $B_8$ is $P_0 - P_1$, the amount of work of $B_8$, denoted by $w(B_8)$, is incremented in probe $P_0$ and decremented in $P_1$.
 
 \begin{algorithm}[h]
-  \caption{Pseudocode that describes how the edge information is used in order to extract
-           the value that will be computed in a given instrumented basic block $B_I$.
+  \caption{Pseudocode that describes how the edge information is used in order to extract the value that will be computed in a given instrumented basic block $B_I$.
 This algorithm could equally be implemented based on the predecessors.}
@@
 
-Although the optimal instrumentation significantly reduces the profiling overhead
-when compared to the naive instrumentaion, from an average overhead of 79\% to 13\%,
-in some critical cases, even the optimal instrumentation can have an overhead
-of about 70\% (see benchmark \texttt{adpcm\_d} in Figure~\ref{fig:overhead-O3}).
+Although the optimal instrumentation significantly reduces the profiling overhead when compared to the naive instrumentaion, from an average overhead of 79\% to 13\%, in some critical cases, even the optimal instrumentation can have an overhead of about 70\% (see benchmark \texttt{adpcm\_d} in Figure~\ref{fig:overhead-O3}).
 In order to further reduce the overhead in these critical cases, we propose a relaxed instrumentation by trading off accuracy and overhead.
@@
 
-%Define an extended DAG of a CFG as a directed graph which may contain cycles
-%if they have known constant trip counts.
+%Define an extended DAG of a CFG as a directed graph which may contain cycles if they have known constant trip counts.
 
-The relaxed instrumentation performs a post-processing on the resulting instrumentation
-of the optimal algorithm. 
-The relaxation starts by extracting \textit{extended} DAGs (directed acyclic graphs)
-from the CFG, as illustrated in Figure~\ref{fig:cfg-relax-example}.
-Each loop (and the outer most region of the function) represents an extended DAG
-where any inner loop is considered to never execute, i.e. only the headers of the
-inner loops are included into the extended DAG.
+The relaxed instrumentation performs a post-processing on the resulting instrumentation of the optimal algorithm.
+The relaxation starts by extracting \textit{extended} DAGs (directed acyclic graphs) from the CFG, as illustrated in Figure~\ref{fig:cfg-relax-example}.
+Each loop (and the outer most region of the function) represents an extended DAG where any inner loop is considered to never execute, i.e.
+only the headers of the inner loops are included into the extended DAG.
 
 %
-%LLVM has a Scalar Evolution Analysis which is used primarily to analyse expressions
-%involving induction variables in loops~\cite{pop05}.
-%An induction variable is a variable that is increased or decreased by a fixed amount
-%on every iteration of a loop or is a linear function of another induction variable.
+%LLVM has a Scalar Evolution Analysis which is used primarily to analyse expressions involving induction variables in loops~\cite{pop05}.
+%An induction variable is a variable that is increased or decreased by a fixed amount on every iteration of a loop or is a linear function of another induction variable.
 %
-%This Scalar Evolution Analysis provides a way to compute \textit{small} constant
-%trip counts (where \textit{small} is defined by a threshold).
+%This Scalar Evolution Analysis provides a way to compute \textit{small} constant trip counts (where \textit{small} is defined by a threshold).
 
@@
   \includegraphics[scale=0.9]{figs/cfg-relax-example.pdf}
-  \caption{Example of a CFG containing a loop and its decomposition into extended DAGs
-           when applying the relaxation.}
+  \caption{Example of a CFG containing a loop and its decomposition into extended DAGs when applying the relaxation.}
   \label{fig:cfg-relax-example}
@@
 
-For every extended DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$,
-we relax the instrumentation accuracy by selecting a subset of the probes to be removed,
-subject to the maximum allowed percentage error, $M$.
+For every extended DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we relax the instrumentation accuracy by selecting a subset of the probes to be removed, subject to the maximum allowed percentage error, $M$.
 
@@
 %$c(I_i)$ is the relative overhead of the instrumentation on the instrumented basic block,
-where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$
-,$\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative
-to the minimum work value possible to compute in the extended DAG,
-and $x_i$ denotes the probes selected for removal.
-Because the percentage error is computed based on the path with the minimum
-amount of work, $\varepsilon(P_i)$ represents the maximum error possible
-that would be incurred when removing probe $P_i$.
-Furthermore, by constraining the percentage error of every extended DAG below a
-given threshold we guarantee that the final error of the relaxation will always
-be bounded by the threshold.
+where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$, $\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to the minimum work value possible to compute in the extended DAG, and $x_i$ denotes the probes selected for removal.
+Because the percentage error is computed based on the path with the minimum amount of work, $\varepsilon(P_i)$ represents the maximum error possible that would be incurred when removing probe $P_i$.
+Furthermore, by constraining the percentage error of every extended DAG below a given threshold we guarantee that the final error of the relaxation will always be bounded by the threshold.
 
-While the optimal placement of probes tries to place probes in edges that are
-less likely to be executed, the relaxation focus on removing probes that are
-more likely to be executed.
-The necessary block frequency information for optimising both the placement of
-probes can be acquired from profiles of previous executions of the program or
-by a static heuristic of the CFG during compilation.
+While the optimal placement of probes tries to place probes in edges that are less likely to be executed, the relaxation focus on removing probes that are more likely to be executed.
+The necessary block frequency information for optimising both the placement of probes can be acquired from profiles of previous executions of the program or by a static heuristic of the CFG during compilation.
 
@@
 the greedy heuristic based on sorting the items ~\cite{dantzig57}.
-We use the brute-force solver for DAGs with a small number of probes and
-the greedy heuristic when the number of probes is greater than a threshold.
-Some of the benchmarks have DAGs with several hundreds of probes, which could
-result in a long compilation time.
+We use the brute-force solver for DAGs with a small number of probes and the greedy heuristic when the number of probes is greater than a threshold.
+Some of the benchmarks have DAGs with several hundreds of probes, which could result in a long compilation time.
 
@@
 First we describe the benchmarks with datasets used in the experiments.
-Afterwards, we discuss our results concerning the instrumentations for profiling
-our work metric.
+Afterwards, we discuss our results concerning the instrumentations for profiling our work metric.
 Finally we present the results of the online {\itercomp}.
@@
 We implemented the instrumentations in LLVM 4.0.
-The target platform is a Linux-4.4.27 system with an Intel Core i7-4770 3.40GHz
-Skylake~CPU with 16~GiB RAM.
+The target platform is a Linux-4.4.27 system with an Intel Core i7-4770 3.40GHz Skylake~CPU with 16~GiB RAM.
 
@@
 
-%{\Itercomp} is essentially based on repeatedly trying
-%out a large number of compiler optimisations until the best
-%combination of compiler optimisations is found for a particular program.
-%Our main goal is to speed up {\itercomp} while targeting
-%optimal performance across large input datasets.
+%{\Itercomp} is essentially based on repeatedly trying out a large number of compiler optimisations until the best combination of compiler optimisations is found for a particular program.
+%Our main goal is to speed up {\itercomp} while targeting optimal performance across large input datasets.
 
-For the experimental evaluation we have used a subset of the \textit{KDataSets} benchmark suit,
-which is the same benchmark and dataset suit used by Chen~\etal~\cite{chen10,chen12a}.
+For the experimental evaluation we have used a subset of the \textit{KDataSets} benchmark suit, which is the same benchmark and dataset suit used by Chen~\etal~\cite{chen10,chen12a}.
 The KDataSets contains 1000 different inputs for each one of its benchmark programs.
-These benchmarks cover a broad spectrum of application scenarios, ranging from simple
-embedded signal-processing tasks to common mobile-phone and desktop tasks.
-The different inputs try to capture distinct characteristics in terms of workload sizes
-and how these workloads exercise different control flow paths.
+These benchmarks cover a broad spectrum of application scenarios, ranging from simple embedded signal-processing tasks to common mobile-phone and desktop tasks.
+The different inputs try to capture distinct characteristics in terms of workload sizes and how these workloads exercise different control flow paths.
 A summary of the benchmark and dataset suit is shown in Table~\ref{tab:kdatasets}.
@@
 \textbf{Program} & \textbf{LOC}    & \textbf{Input file size}            & \textbf{Input description}              \\ \hline % Domain
-bitcount      & 460    &  -                         & Numbers: random                \\ \hline
+%bitcount      & 460    &  -                         & Numbers: random                \\ \hline
 qsort         & 154    & 32K-1.8M                   & 3D coordinates                 \\ \hline
@@
 %mad           & 2358   & 28K-27M                    & MP3 audios                     \\ \hline
-lame          & 14491  & 167K-36M                   & WAVE audios                    \\ \hline
+%lame          & 14491  & 167K-36M                   & WAVE audios                    \\ \hline
 adpcm\_c      & 210    & 167K-36M                   & WAVE audios                    \\ \hline
@@
 
-Figure~\ref{fig:instr} shows percentage of instrumented basic blocks
-for the optimal and the relaxed instrumentation with different relaxation
-thresholds. The naive instrumentation always has 100\% of the basic blocks
-instrumented, by definition.
+Figure~\ref{fig:instr} shows percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.
+The naive instrumentation always has 100\% of the basic blocks instrumented, by definition.
 
@@
     \includegraphics[width=\textwidth]{figs/instr.pdf}
-    \caption{Percentage of instrumented basic blocks
-for the optimal and the relaxed instrumentation with different relaxation
-thresholds.}
+    \caption{Percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.}
     \label{fig:instr}
@@
 
-The \texttt{adpcm\_d} benchmark is the most critical case amongst the
-used benchmarks, with a overhead of about 66\% with the optimal instrumentation.
-This benchmark consists mainly of a single hot loop with several branches
-inside it.
-The relaxation algorithm is able to reduce this overhead down to about 52\%
-(with an error threshold of 1\%) and 19\% (with an error threshold of 2\%)
-by removing only one and two probes from the hot loop, respectively.
-These two removed probes were placed in branches, inside the hot loop, with
-a high probability of being taken, but with a small contribution to the work
-measured in the loop.
+The \texttt{adpcm\_d} benchmark is the most critical case amongst the used benchmarks, with a overhead of about 66\% with the optimal instrumentation.
+This benchmark consists mainly of a single hot loop with several branches inside it.
+The relaxation algorithm is able to reduce this overhead down to about 52\% (with an error threshold of 1\%) and 19\% (with an error threshold of 2\%) by removing only one and two probes from the hot loop, respectively.
+These two removed probes were placed in branches, inside the hot loop, with a high probability of being taken, but with a small contribution to the work measured in the loop.
 
-%In this section we discuss the three baseline implementations that will
-%be used for assessing the quality of the proposed metric.
+%In this section we discuss the three baseline implementations that will be used for assessing the quality of the proposed metric.
 
@@
 
-%Most of the existing {\itercomp} studies find the
-%best optimisation though repeated runs on the same input.
-%Although this approach will usually lead to sub-optimal performance
-%across large input datasets, it provides a good baseline when considering
-%the complexity and resonably low compile-time regarding
-%iterative optimising compilers.
-%If $M$ is the total number of combinations of compiler optimisations,
-%this approach requires $O(M)$ runs of the program being optimised.
+%Most of the existing {\itercomp} studies find the best optimisation though repeated runs on the same input.
+%Although this approach will usually lead to sub-optimal performance across large input datasets, it provides a good baseline when considering the complexity and resonably low compile-time regarding iterative optimising compilers.
+%If $M$ is the total number of combinations of compiler optimisations, this approach requires $O(M)$ runs of the program being optimised.
 %We can consider two main scenarios:
-%\textit{($i$ - best case scenario)}
-%after selecting the optimisation over each individual input,
-%consider the one with best performance across the whole input dataset;
-%\textit{($ii$ - expected scenario)}
-%after selecting the optimisation over each individual input,
-%consider the average case of their performance across the whole input dataset.
+%\textit{($i$ - best case scenario)} after selecting the optimisation over each individual input, consider the one with best performance across the whole input dataset;
+%\textit{($ii$ - expected scenario)} after selecting the optimisation over each individual input, consider the average case of their performance across the whole input dataset.
 
@@
 
-%Recent work on {\itercomp} have been targeting optimisation
-%across multiple inputs~\cite{fursin07,chen10,chen12a}.
-%If $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimisations,
-%they perform a total of $O(NM)$ runs of the program being optimised.
-%Similarly to what have been discussed previously,
-%there are different ways for how to determine the optimal
-%combination of compiler optimisations across multiple inputs.
-%It is possible to tune the selection of the program-optimal combination
-%to minimise risk or to maximise average performance.
-%A compromise-based selection criterion could be to maximise average speedup
-%with minimised variance.
+%Recent work on {\itercomp} have been targeting optimisation across multiple inputs~\cite{fursin07,chen10,chen12a}.
+%If $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimisations, they perform a total of $O(NM)$ runs of the program being optimised.
+%Similarly to what have been discussed previously, there are different ways for how to determine the optimal combination of compiler optimisations across multiple inputs.
+%It is possible to tune the selection of the program-optimal combination to minimise risk or to maximise average performance.
+%A compromise-based selection criterion could be to maximise average speedup with minimised variance.
 
@@
 
-%While the previous two baselines addresses two opposite aspects of {\itercomp},
-%namely, compile-time efficiency and performance of the generated code,
-%we also intend to compare against IPC as the competing baseline.
-%The main reason for comparing against IPC is because it has been proposed as a metric
-%for comparing the performance of two optimisations running on two distinct inputs.
-%If $M$ is the total number of combinations of compiler optimisations,
-%this approach requires $O(M)$ runs of the program being optimised.
+%While the previous two baselines addresses two opposite aspects of {\itercomp}, namely, compile-time efficiency and performance of the generated code, we also intend to compare against IPC as the competing baseline.
+%The main reason for comparing against IPC is because it has been proposed as a metric for comparing the performance of two optimisations running on two distinct inputs.
+%If $M$ is the total number of combinations of compiler optimisations, this approach requires $O(M)$ runs of the program being optimised.
 
@@
 \begin{itemize}
-\item Oracle-RM executes the program twice, for each input, and then measuring
-the real speedup for each compiler optimisation and uses the real speedup for
-selecting the best optimisation.
-\item Oracle-PP represents the \textit{perfect} non-intrusive profiling by also
-executing the program twice, once for estimating the amount of work and the
-second for measuring its execution time.
-This version uses the work-based metric, $\frac{\Delta W}{\Delta t}$, during
-the {\itercomp} search.
+\item Oracle-RM executes the program twice, for each input, and then measuring the real speedup for each compiler optimisation and uses the real speedup for selecting the best optimisation.
+\item Oracle-PP represents the \textit{perfect} non-intrusive profiling by also executing the program twice, once for estimating the amount of work and the second for measuring its execution time.
+This version uses the work-based metric, $\frac{\Delta W}{\Delta t}$, during the {\itercomp} search.
 \end{itemize}
@@
 %
-%Figure~\ref{fig:flagsfreq} shows an aggregated view of the final combination
-%of compiler optimisations that were selected by {\itercomp} search.
-%The figure presents the individual optimisation passes with at least 1\% of
-%frequency in the selected combination of compiler optimisations that improved
-%the performance over {\texttt{-O3}}.
+%Figure~\ref{fig:flagsfreq} shows an aggregated view of the final combination of compiler optimisations that were selected by {\itercomp} search.
+%The figure presents the individual optimisation passes with at least 1\% of frequency in the selected combination of compiler optimisations that improved the performance over {\texttt{-O3}}.
 %
@@
 
-Chen~\etal~\cite{chen10,chen12a} evaluate the effectiveness of iterative
-compilations across a large number of input test cases.
+Chen~\etal~\cite{chen10,chen12a} evaluate the effectiveness of iterative compilations across a large number of input test cases.
 Their main motivation is to answer the question:
 \textit{How data input dependent is {\itercomp}?}
-Their results show that it is possible to find a combination of compiler optimisations
-that achieves at least 86\% of the maximum speedup across all input test cases.
-These optimal combinations are program-specific and yield average
-speedups up to 3.75$\times$ over the highest optimisation level of compilers.
+Their results show that it is possible to find a combination of compiler optimisations that achieves at least 86\% of the maximum speedup across all input test cases.
+These optimal combinations are program-specific and yield average speedups up to 3.75$\times$ over the highest optimisation level of compilers.
 
-When optimising a program, the main method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a}
-evaluates each combination of compiler optimisations across all the available inputs, i.e.,
-if $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimisations,
-they perform a total of $O(NM)$ runs of the program being optimised.
-Furthermore, they use a pre-defined set of only 300 different combinations of compiler optimisations,
-which represents a very small sample of the optimisation search space for most modern compilers, e.g.
-LLVM has 56 distinct optimisation passes and GCC has about 47 high-level (SSA form) optimisation passes and
-about 25 low-level (RTL) optimisation passes, which in both cases result in much more than $2^{50}$ distinct
-combinations of compiler optimisations, without considering repetition.
+When optimising a program, the main method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a} evaluates each combination of compiler optimisations across all the available inputs, i.e., if $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimisations, they perform a total of $O(NM)$ runs of the program being optimised.
+Furthermore, they use a pre-defined set of only 300 different combinations of compiler optimisations, which represents a very small sample of the optimisation search space for most modern compilers, e.g.
+LLVM has 56 distinct optimisation passes and GCC has about 47 high-level (SSA form) optimisation passes and about 25 low-level (RTL) optimisation passes, which in both cases result in much more than $2^{50}$ distinct combinations of compiler optimisations, without considering repetition.
 
-Recent work~\cite{chen12b,fang15} have applied the same idea of performing
-input-dependent {\itercomp} to distributed applications on data centres.
-In summary, each worker receives a subset of the input dataset, called the evaluation dataset,
-to perform an \textit{online} {\itercomp} of the code being executed.
-Each worker performs the same the method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a},
-i.e., they evaluate each combination of compiler optimisations across all the evaluation dataset.
-However, because the optimisation is performed online, they usually consider a small evaluation dataset
-and a small number of compiler optimisations.
+Recent work~\cite{chen12b,fang15} have applied the same idea of performing input-dependent {\itercomp} to distributed applications on data centres.
+In summary, each worker receives a subset of the input dataset, called the evaluation dataset, to perform an \textit{online} {\itercomp} of the code being executed.
+Each worker performs the same the method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a}, i.e., they evaluate each combination of compiler optimisations across all the evaluation dataset.
+However, because the optimisation is performed online, they usually consider a small evaluation dataset and a small number of compiler optimisations.
 
-Fursin~\etal~\cite{fursin07} addressed the problem of comparing the effect of two optimisations
-on two distinct inputs. For that purpose, they proposed to use instruction per cycle (IPC) as the metric for performing such comparison.
-Their result show that using IPC seems promising as a robust metric for {\itercomp}
-across large input datasets.
-However, 
-some specific optimisation techniques may affect the use of IPC as a robust metric, and specially
-IPC has been shown to provide poor performance estimation for multi-threaded programs~\cite{alameldeen06,eyerman08}.
-In particular, IPC can give a skewed performance measure if threads spend time in \textit{spin-lock loops}
-or other synchronisation mechanisms. 
-Some existing work on performance assessment suggest that
-total execution time should be used for measuring performance of multi-threaded programs~\cite{alameldeen06,eyerman08}.
-Aalameldeen and Wood~\cite{alameldeen06} in particular suggest that a simple work-related metric should be used
-if the unit of work is representative enough.
-Work-related metrics have already been largely used for measuring performance of throughput-oriented applications,
-for other applications, however, choosing an appropriate unit of work can be more challanging~\cite{alameldeen06}.
+Fursin~\etal~\cite{fursin07} addressed the problem of comparing the effect of two optimisations on two distinct inputs.
+For that purpose, they proposed to use instruction per cycle (IPC) as the metric for performing such comparison.
+Their result show that using IPC seems promising as a robust metric for {\itercomp} across large input datasets.
+However, some specific optimisation techniques may affect the use of IPC as a robust metric, and specially IPC has been shown to provide poor performance estimation for multi-threaded programs~\cite{alameldeen06,eyerman08}.
+In particular, IPC can give a skewed performance measure if threads spend time in \textit{spin-lock loops} or other synchronisation mechanisms. 
+Some existing work on performance assessment suggest that total execution time should be used for measuring performance of multi-threaded programs~\cite{alameldeen06,eyerman08}.
+Aalameldeen and Wood~\cite{alameldeen06} in particular suggest that a simple work-related metric should be used if the unit of work is representative enough.
+Work-related metrics have already been largely used for measuring performance of throughput-oriented applications, for other applications, however, choosing an appropriate unit of work can be more challanging~\cite{alameldeen06}.
 
@@
 
-Previous work have proposed profiling-based mechanism to estimate
-input sizes~\cite{zaparanuks12,coppa14}.
-Coppa~\etal~\cite{coppa14} in particular propose the concept of
-\textit{read memory size} for automatically estimating the size of the input
-passed to a routine, where \textit{read memory size} represents the number of
-distinct memory cells first accessed by a read operation.
-In other words, the \textit{read memory size} metric measures the size of the 
-useful portion of the input's memory footprint.
-However, because we are interested in the amount of computational work performed
-in respect of a given input, the memory footprint of the input may not always
-have a direct correspondence to  the amount of computational work.
+Previous work have proposed profiling-based mechanism to estimate input sizes~\cite{zaparanuks12,coppa14}.
+Coppa~\etal~\cite{coppa14} in particular propose the concept of \textit{read memory size} for automatically estimating the size of the input passed to a routine, where \textit{read memory size} represents the number of distinct memory cells first accessed by a read operation.
+In other words, the \textit{read memory size} metric measures the size of the useful portion of the input's memory footprint.
+However, because we are interested in the amount of computational work performed in respect of a given input, the memory footprint of the input may not always have a direct correspondence to  the amount of computational work.
 
-Goldsmith~\etal~\cite{goldsmith07} use \textit{block frequency} as the
-measure for performance for empirically describing the asymptotic behaviour
-of programs, which is known as empirical computational complexity.
-Block frequency is a relative metric that represents the number of times a
-basic block executes~\cite{ball94,ball96}. 
-They argue in favour of block frequency due to its portability, repeatability
-and exactness, since it does not suffer from timer resolution problems or
-non-deterministic noises.
-Block frequency also has the advantage of being efficiently profiled by means
-of automatic code instrumentation~\cite{knuth73,ball94}.
+Goldsmith~\etal~\cite{goldsmith07} use \textit{block frequency} as the measure for performance for empirically describing the asymptotic behaviour of programs, which is known as empirical computational complexity.
+Block frequency is a relative metric that represents the number of times a basic block executes~\cite{ball94,ball96}.
+They argue in favour of block frequency due to its portability, repeatability and exactness, since it does not suffer from timer resolution problems or non-deterministic noises.
+Block frequency also has the advantage of being efficiently profiled by means of automatic code instrumentation~\cite{knuth73,ball94}.
 
-However, in the context of comparing different optimisations,
-although block frequency would be able to capture aspects of optimisations
-that simplify the control-flow graph (CFG), measuring work at the basic block
-resolution would not capture effects of optimisations at the instruction level.
-Because of that, we extend the idea of using basic block frequency to measure
-computational work by also considering the computational cost of each
-basic block. The computational cost of a basic block is given by weighing
-the instructions that it contains.
+However, in the context of comparing different optimisations, although block frequency would be able to capture aspects of optimisations that simplify the control-flow graph (CFG), measuring work at the basic block resolution would not capture effects of optimisations at the instruction level.
+Because of that, we extend the idea of using basic block frequency to measure computational work by also considering the computational cost of each basic block.
+The computational cost of a basic block is given by weighing the instructions that it contains.
 
@@
 
-In 1973, Donald Knuth~\cite{knuth73} provided an optimal algorithm for
-profiling block frequency, inserting fewer probes than the previously described
-naive instrumentation that was commonly used in practice~\cite{knuth71}.
-Forman~\cite{forman81} proposed further overhead reductions by placing the
-probes in basic blocks that are less likely to be executed based on static
-heuristics.
+In 1973, Donald Knuth~\cite{knuth73} provided an optimal algorithm for profiling block frequency, inserting fewer probes than the previously described naive instrumentation that was commonly used in practice~\cite{knuth71}.
+Forman~\cite{forman81} proposed further overhead reductions by placing the probes in basic blocks that are less likely to be executed based on static heuristics.
 
-Ball and Larus~\cite{ball94} offered a detailed discussion comparing
-two approaches for optimally profiling block frequency, namely,
-placing probes on the vertices or the edges of the control-flow graph (CFG).
+Ball and Larus~\cite{ball94} offered a detailed discussion comparing two approaches for optimally profiling block frequency, namely, placing probes on the vertices or the edges of the control-flow graph (CFG).
 They show that the edge-based approach produces optimal placement of probes.
@@
 
-Our proposed technique for relaxed instrumentation can also be applied for similar
-profiling scenarios, such as profiling empirical computational complexity~\cite{goldsmith07,zaparanuks12,coppa14}.
+Our proposed technique for relaxed instrumentation can also be applied for similar profiling scenarios, such as profiling empirical computational complexity~\cite{goldsmith07,zaparanuks12,coppa14}.
 

### 2017-08-11 cgo18mscr.tex
updates with MScR Thesis

@@
-\documentclass[sigplan,9pt]{acmart}
+\PassOptionsToPackage{table}{xcolor}
+\documentclass[sigplan,10pt]{acmart}
 
@@
 
-\usepackage{color}
+%\usepackage{setspace}
+\usepackage{ragged2e}
 
@@
 
-\usepackage[noend]{algpseudocode}
-\usepackage{algorithm}
+%% Listing related stuff
+
+\usepackage{listings}
+\usepackage{courier}            %Required for pretty listings (more dense)
+
+%% EMACS colors for listing:
+
+\usepackage{color}
+\definecolor{sh_comment}{rgb}{0.65, 0.00, 0.00 } % red-ish
+\definecolor{sh_keyword}{rgb}{0.37, 0.69, 0.69}  % blue-green
+\definecolor{sh_string}{rgb}{0.08, 0.69, 0.08} % green
+
+%% This manipulates the listing numbering. It decreases it by one at the point used.
+\newcommand*\lstDNumber{\addtocounter{lstnumber}{-1}}
+
+%% Change from Listing->Algorithm
+%% \renewcommand\lstlistingname{Algorithm} % Fix listing caption name: Listing -> Algorithm
+%% \renewcommand\lstlistlistingname{Algorithms}
+%% \def\lstlistingautorefname{Alg.}
+
+%% \renewcommand{\thelstlisting}{\thesection-\arabic{lstlisting}}
+%\renewcommand{\thelstlisting}{\arabic{lstlisting}} % Fix listing numbering Algorithm 1.1 -> Algorithm 1.
+
+%% \renewcommand{\labelenumi}{\roman{enumi}.}
+
+\lstset{
+float=[*],
+language=C,                % choose the language of the code
+%% basicstyle=\linespread{0.9}
+%% basicstyle=\scriptsize\ttfamily,
+ basicstyle={\linespread{0.97} \scriptsize\ttfamily},
+ stringstyle=\color{sh_string},
+ keywordstyle = \color{sh_keyword}\bfseries,
+ commentstyle=\color{sh_comment}\itshape,
+numbers=left,                   % where to put the line-numbers
+%% numberstyle=\scriptsize,        % the size of the fonts that are used for the line-numbers
+numberstyle=\scriptsize,        % the size of the fonts that are used for the line-numbers
+stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
+numbersep=5pt,                  % how far the line-numbers are from the code
+backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
+showspaces=false,               % show spaces adding particular underscores
+showstringspaces=false,         % underline spaces within strings
+showtabs=false,                 % show tabs within strings adding particular underscores
+xleftmargin=2em,                % Left margin
+frame=lines,                   % adds a frame around the code. none|leftline|topline|bottomline|lines|single|shadowbox
+framexleftmargin=1.5em,         % Margin from frame to line numbers
+framexbottommargin=0em,         % Distance from last text to frame
+morekeywords={in,true,false,and,or,set,is,not},
+%% prebreak=\mbox{\tiny$\searrow$},
+prebreak=\space,                % Line break: insert this character at the end of the top line
+%% postbreak=\mbox{{\color{blue}\scriptsize$\hookrightarrow$}}, % Line break: blue arrow at the beginning of the bottom line
+postbreak=\mbox{{\color{blue}\scriptsize$\hookrightarrow$}}, % Line break: blue arrow at the beginning of the bottom line
+breaklines=true,                % sets automatic line breaking
+breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
+tabsize=2,                      % sets default tabsize to 2 spaces
+captionpos=b,                   % sets the caption-position to top
+escapeinside={\@}{\@}             % Escape sequence \@ESCAPED STUFF\@
+}
+
+\usepackage{nasm/lang}
+\usepackage{nasm/style}
+\usepackage{llvm/lang}
 
 \newcommand{\etal}{et~al.}
-\algrenewcommand\algorithmicindent{3ex}%
 
@@
 \title{Online {\IterComp}\\Guided by Low Overhead Profiling}
-%\titlenote{Produces the permission block, and
-%  copyright information}
-%\subtitle{Extended Abstract}
-%\subtitlenote{The full version of the author's guide is available as
-%  \texttt{acmart.pdf} document}
+%\titlenote{...}
+%\subtitle{...}
+%\subtitlenote{...}
 
-\author{Rodrigo C. O. Rocha}
-%\authornote{Dr.~Trovato insisted his name be first.}
-%\orcid{1234-5678-9012}
+\author{ }
 \affiliation{%
-  \institution{University of Edinburgh, UK}
-  %\streetaddress{P.O. Box 1212}
-  %\city{Dublin} 
-  %\state{Ohio} 
-  %\postcode{43017-6221}
+  \institution{ }
 }
-\email{r.rocha@ed.ac.uk}
+\email{ }
 
-\author{Pavlos Petoumenos}
-%\authornote{The secretary disavows any knowledge of this author's actions.}
-\affiliation{%
-  \institution{University of Edinburgh, UK}
-  %\streetaddress{P.O. Box 1212}
-  %\city{Dublin} 
-  %\state{Ohio} 
-  %\postcode{43017-6221}
-}
-\email{ppetoume@inf.ed.ac.uk}
+%\author{Rodrigo C. O. Rocha}
+%\affiliation{%
+%  \institution{University of Edinburgh, UK}
+%}
+%\email{r.rocha@ed.ac.uk}
 
-\author{Lu\'is F. W. G\'oes}
-%\authornote{This author is the
-%  one who did all the really hard work.}
-\affiliation{%
-  \institution{PUC Minas, Brazil}
-  %\streetaddress{1 Th{\o}rv{\"a}ld Circle}
-  %\city{Hekla} 
-  %\country{Iceland}
-}
-\email{lfwgoes@pucminas.br}
+%\author{Pavlos Petoumenos}
+%\affiliation{%
+%  \institution{University of Edinburgh, UK}
+%}
+%\email{ppetoume@inf.ed.ac.uk}
 
-\author{Murray Cole}
-\affiliation{
-  \institution{University of Edinburgh, UK}
-  %\streetaddress{P.O. Box 5000}
-}
-\email{mic@inf.ed.ac.uk}
+%\author{Lu\'is F. W. G\'oes}
+%\affiliation{%
+%  \institution{PUC Minas, Brazil}
+%}
+%\email{lfwgoes@pucminas.br}
 
-\author{Zheng Wang}
-\affiliation{%
-  \institution{Lancaster University, UK}
-  %\city{Moffett Field}
-  %\state{California} 
-  %\postcode{94035}
-}
-\email{z.wang@lancaster.ac.uk}
+%\author{Murray Cole}
+%\affiliation{
+%  \institution{University of Edinburgh, UK}
+%}
+%\email{mic@inf.ed.ac.uk}
 
-\author{Hugh Leather}
-\affiliation{%
-  \institution{University of Edinburgh, UK}
-  %\streetaddress{8600 Datapoint Drive}
-  %\city{San Antonio}
-  %\state{Texas} 
-  %\postcode{78229}
-}
-\email{hleather@inf.ed.ac.uk}
+%\author{Zheng Wang}
+%\affiliation{%
+%  \institution{Lancaster University, UK}
+%}
+%\email{z.wang@lancaster.ac.uk}
 
-%\author{John Smith}
-%\affiliation{\institution{The Th{\o}rv{\"a}ld Group}}
-%\email{jsmith@affiliation.org}
+%\author{Hugh Leather}
+%\affiliation{%
+%  \institution{University of Edinburgh, UK}
+%}
+%\email{hleather@inf.ed.ac.uk}
 
-%\author{Julius P.~Kumquat}
-%\affiliation{\institution{The Kumquat Consortium}}
-%\email{jpkumquat@consortium.net}
-
-% The default list of authors is too long for headers}
 \renewcommand{\shortauthors}{R. Rocha et al.}
@@
 Understanding the interactions of optimisations is very important in determining a good solution to the phase-ordering problem~\cite{touati06,kulkarni12}.
-Because of all those dependences and interactions, although most compiler optimisations yield significant improvements in many programs, the potential for performance degradation in certain program patterns is known to compiler writers and many users~\cite{pan06,zhou12,kulkarni12}.
+Because of all those dependencies and interactions, although most compiler optimisations yield significant improvements in many programs, the potential for performance degradation in certain program patterns is known to compiler writers and many users~\cite{pan06,zhou12,kulkarni12}.
 %Traditional compilers typically apply the same set of optimisation in a fixed order to all functions in a program, without regard the code being optimised.
@@
     \centering
-    \includegraphics[width=\linewidth]{figs/diagram}
+    \includegraphics[width=\linewidth]{figs/infra-diagram}
     \caption{Overview of the execution engine for applying {\itercomp}.}
-    \label{fig:diagram}
+    \label{fig:infra-diagram}
 \end{figure}
@@
 
-Similarly to previous work~\cite{giusto01,powell09,brandolese11}, we derive the cost model for the instruction set by modeling the problem as a multi-variable linear regression, where the \textit{regression coefficients} are the costs of the instructions and the \textit{regressors} (or \textit{explanatory variables}) are computed as $\sum_B N_B(i)f(B)$ for each instruction $i$.
+Similarly to previous work~\cite{giusto01,powell09,brandolese11}, we derive the cost model for the instruction set by modelling the problem as a multi-variable linear regression, where the \textit{regression coefficients} are the costs of the instructions and the \textit{regressors} (or \textit{explanatory variables}) are computed as $\sum_B N_B(i)f(B)$ for each instruction $i$.
+\[
+\Delta W = \varepsilon + \sum_{i} \left(w_i \sum_{B} N_B(i)f(B)\right)
+\]
+
+By having some empirical data after executing several benchmarks with different inputs, we can fit the linear model with this empirical data in order to obtain the costs of the instructions.
 
@@
 
-\begin{algorithm}[h]
-  \caption{Pseudocode of the data-flow analysis for assigning the values computed
-  in each probe of the instrumentation for the profiling of the work metric.}
-  \label{alg:populateEdgeInfo}
-  \begin{algorithmic}
-    \Function{\textrm{populateEdgeInfo}}{$G$}
-
-    \For{\textbf{each} $e \in $ \textrm{instrumentedEdges}($G$) }
-       \State $B_I \gets $ \textrm{instrumentedBlock}($e$)
-       \State $inc[e] \gets \{ B_I \}$
-       \State $dec[e] \gets \{ \}$
-       \State $known[e] \gets $ \textbf{true}
-    \EndFor
-
-    \State $changed \gets $ \textbf{true}
-    \While{$changed$}
-       \State $changed \gets $ \textbf{false}
-	   \For{\textbf{each} $B \in V(G)$}
-          \State $uIn \gets $ \textrm{unknownIn}($known, G, B$)
-          \State $uOut \gets $ \textrm{unknownOut}($known, G, B$)
-		  \If{$uIn=0$ \textbf{and} $uOut=1$}
-             \State $changed \gets $ \textbf{true}
-		     \State $incOut \gets \{\}$
-			 \State $decOut \gets \{\}$
-		     \For{$B_p\in$ \textrm{predecessors}($G, B$)}
-			    \State $incOut \gets incOut \bigcup inc[(B_p,B)]$
-			    \State $decOut \gets decOut \bigcup dec[(B_p,B)]$
-			 \EndFor
-		     \For{$B_s\in$ \textrm{successors}($G, B$)}
-			    \State $incOut \gets incOut \cup dec[(B,B_s)]$
-			    \State $decOut \gets decOut \cup inc[(B,B_s)]$
-			 \EndFor
-		     \For{$B_s\in$ \textrm{successors}($G, B$)}
-			    \If{\textbf{not} $known[(B,B_s)]$}
-			       \State $inc[(B,B_s)] \gets incOut\setminus{decOut}$
-			       \State $dec[(B,B_s)] \gets decOut\setminus{incOut}$
-			       \State $known[(B,B_s)] \gets$ \textbf{true}
-				\EndIf
-			 \EndFor
-		  \EndIf
-		  \If{$uIn=1$ \textbf{and} $uOut=0$}
-             \State $changed \gets $ \textbf{true}
-		     \State $incIn \gets \{\}$
-			 \State $decIn \gets \{\}$
-		     \For{$B_s\in$ \textrm{successors}($G, B$)}
-			    \State $incIn \gets incIn \cup inc[(B,B_s)]$
-			    \State $decIn \gets decIn \cup dec[(B,B_s)]$
-			 \EndFor
-		     \For{$B_p\in$ \textrm{predecessors}($G, B$)}
-			    \State $incIn \gets incIn \bigcup dec[(B_p,B)]$
-			    \State $decIn \gets decIn \bigcup inc[(B_p,B)]$
-			 \EndFor
-		     \For{$B_p\in$ \textrm{predecessors}($G, B$)}
-			    \If{\textbf{not} $known[(B_p,B)]$}
-			       \State $inc[(B_p,B)] \gets incIn\setminus{decIn}$
-			       \State $dec[(B_p,B)] \gets decIn\setminus{incIn}$
-			       \State $known[(B_p,B)] \gets$ \textbf{true}
-				\EndIf
-			 \EndFor
-		  \EndIf
-	   \EndFor
-    \EndWhile
-    \EndFunction
-  \end{algorithmic}
-\end{algorithm}
+\begin{lstlisting}[caption={Post-processing of the CFG for populating all edge flows based on the collected probes.}, label={lst:populateEdgeFlows}, float]
+// Inputs: CFG with the known edges flows from
+//         the cotree (collected probes).
+// Output: Updated CFG with all edge flows.
+populateEdgeFlows(G) {
+  changed = true
+  while changed:
+    changed = false
+    for B in G.vertices():
+      unIN = count( G.unknownIncomingEdges(B) )
+      unOUT = count( G.unknownOutgoingEdges(B) )
+      if unIN==0 and unOUT==1:
+        //sum known incoming and outgoing edges
+        sIN = sum( G.incomingEdges(B) )
+        sOUT = sum( G.outgoingEdges(B) )
+        //update unknown outgoing edge in B
+        //with (sIN-sOUT)
+        G.setUnknownOutgoingEdge(B, (sIN-sOUT))
+        changed = true
+      if unIN==1 and unOUT==0:
+        //sum known incoming and outgoing edges
+        sIN = sum( G.incomingEdges(B) )
+        sOUT = sum( G.outgoingEdges(B) )
+        //update unknown incoming edge in B
+        //with (sOUT-sIN)
+        G.setUnknownIncomingEdge(B, (sOUT-sIN))
+        changed = true
+}
+\end{lstlisting}
 
@@
 The positive terms in the edge expression of a basic block indicate that the amount of work of this basic block will be incremented in the probes represented by these positive terms, similarly, the negative terms indicate that the amount of work of this basic block will be decremented in the probes represented by these negative terms.
-For example, because the edge experssion for the basic block $B_8$ is $P_0 - P_1$, the amount of work of $B_8$, denoted by $w(B_8)$, is incremented in probe $P_0$ and decremented in $P_1$.
+For example, because the edge expression for the basic block $B_8$ is $P_0 - P_1$, the amount of work of $B_8$, denoted by $w(B_8)$, is incremented in probe $P_0$ and decremented in $P_1$.
 
-\begin{algorithm}[h]
-  \caption{Pseudocode that describes how the edge information is used in order to extract the value that will be computed in a given instrumented basic block $B_I$.
-This algorithm could equally be implemented based on the predecessors.}
-  \label{alg:instrValue}
-  \begin{algorithmic}
-    \Function{\textrm{instrValue}}{$G, B_I, inc, dec$}
-	\State $v \gets 0$
-    \For{\textbf{each} $B \in V(G)$}
-	   \State $inc_B \gets \{\}$
-	   \State $dec_B \gets \{\}$
-%	   \For{$B_p\in$ \textrm{predecessors}($G, B$)}
-%	      \State $inc_B \gets inc_B \bigcup inc[(B_p,B)]$
-%	      \State $dec_B \gets dec_B \bigcup dec[(B_p,B)]$
-%	   \EndFor
-	   \For{$B_s\in$ \textrm{successors}($G, B$)}
-	      \State $inc_B \gets inc_B \bigcup inc[(B,B_s)]$
-	      \State $dec_B \gets dec_B \bigcup dec[(B,B_s)]$
-	   \EndFor
-	   \If{ $B_I \in inc_B\setminus{dec_B}$}
-	      \State $v \gets v + w(B)$
-	   \EndIf
-	   \If{ $B_I \in dec_B\setminus{inc_B}$}
-	      \State $v \gets v - w(B)$
-	   \EndIf
-	\EndFor
-    \Return $v$
-    \EndFunction
-  \end{algorithmic}
-\end{algorithm}
+\begin{lstlisting}[caption={Pseudocode that describes how the edge information is used in order to extract the value that will be computed in a given instrumented basic block $B_I$.
+This algorithm could equally be implemented based on the predecessors.}, label={lst:instrValue}, float]
+// Inputs: 1) CFG with the known edges flows
+//            of the chords.
+//         2) Basic block targeted for probing.
+// Output: Work value to be incremented by the
+//         given probe B.
+instrValue(G, B, inc, dec) {
+  value = 0
+  for B in G.vertices():
+    incB = set()
+    decB = set()
+    for succB in G.successor(B):
+      incB = incB union inc[G.getEdge(B, succB)]
+      decB = decB union dec[G.getEdge(B, succB)]
+    if B in (incB - decB):
+      value = value + w(B)
+    if B in (decB - incB):
+      value = value - w(B)
+  return value
+}
+\end{lstlisting}
 
@@
 
-Although the optimal instrumentation significantly reduces the profiling overhead when compared to the naive instrumentaion, from an average overhead of 79\% to 13\%, in some critical cases, even the optimal instrumentation can have an overhead of about 70\% (see benchmark \texttt{adpcm\_d} in Figure~\ref{fig:overhead-O3}).
+Although the optimal instrumentation significantly reduces the profiling overhead when compared to the naive instrumentation, from an average overhead of 79\% to 13\%, in some critical cases, even the optimal instrumentation can have an overhead of about 70\% (see benchmark \texttt{adpcm\_d} in Figure~\ref{fig:overhead-O3}).
 In order to further reduce the overhead in these critical cases, we propose a relaxed instrumentation by trading off accuracy and overhead.
@@
 
-%{\Itercomp} is essentially based on repeatedly trying out a large number of compiler optimisations until the best combination of compiler optimisations is found for a particular program.
-%Our main goal is to speed up {\itercomp} while targeting optimal performance across large input datasets.
-
 For the experimental evaluation we have used a subset of the \textit{KDataSets} benchmark suit, which is the same benchmark and dataset suit used by Chen~\etal~\cite{chen10,chen12a}.
@@
 The different inputs try to capture distinct characteristics in terms of workload sizes and how these workloads exercise different control flow paths.
-A summary of the benchmark and dataset suit is shown in Table~\ref{tab:kdatasets}.
+A summary of the benchmarks and dataset suit is shown in Table~\ref{tab:kdatasets}.
 
@@
 \textbf{Program} & \textbf{LOC}    & \textbf{Input file size}            & \textbf{Input description}              \\ \hline % Domain
-%bitcount      & 460    &  -                         & Numbers: random                \\ \hline
 qsort         & 154    & 32K-1.8M                   & 3D coordinates                 \\ \hline
-%dijkstra      & 163    & 0.06K-4.3M                 & Adjacency matrices             \\ \hline
-%patricia      & 290    & 0.6K-1.9M                  & IP and mask pairs              \\ \hline
 jpeg\_d       & 13501  & 3.6K-1.5M                  & JPEG images                    \\ \hline
@@
 susan\_s      & 1376   &                            &                                \\ \hline
-%mad           & 2358   & 28K-27M                    & MP3 audios                     \\ \hline
-%lame          & 14491  & 167K-36M                   & WAVE audios                    \\ \hline
 adpcm\_c      & 210    & 167K-36M                   & WAVE audios                    \\ \hline
 adpcm\_d      & 211    & 21K-8.8M                   & ADPCM audios                   \\ \hline
+lame          & 14491  & 167K-36M                   & WAVE audios                    \\ \hline
+rsynth        & 4111   & 0.1K-42M                   &  Text files                    \\ \hline
+sha           & 197    & 0.6K-35M                   & Files of any format            \\ \hline
+\rowcolor{gray!20}
+bitcount      & 460    &  -                         & Numbers: random                \\ \hline
+\rowcolor{gray!20}
+dijkstra      & 163    & 0.06K-4.3M                 & Adjacency matrices             \\ \hline
+\rowcolor{gray!20}
+patricia      & 290    & 0.6K-1.9M                  & IP and mask pairs              \\ \hline
+\rowcolor{gray!20}
+mad           & 2358   & 28K-27M                    & MP3 audios                     \\ \hline
+\rowcolor{gray!20}
 %gsm           & 3806   & 83K-18M                    & Sun/NeXT audios                \\ \hline
-%ghostscript   & 99869  & 11K-43M                    & Postscript files               \\ \hline
+ghostscript   & 99869  & 11K-43M                    & Postscript files               \\ \hline
+\rowcolor{gray!20}
 %ispell        & 6522   & \multirow{3}{*}{0.1K-42M}  & \multirow{3}{*}{Text files}    \\ \cline{1-2}
 %rsynth        & 4111   &                            &                                \\ \cline{1-2}
-rsynth        & 4111   &                            &  Text files                    \\ \hline %%%%%% copied line
-%stringsearch  & 338    &                            &                                \\ \hline
+stringsearch  & 338    &  0.1K-42M                 &  Text files                     \\ \hline
+\rowcolor{gray!20}
 %blowfish\_e   & 863    & 0.6K-35M                   & Files of any format            \\ \hline
@@
 %rijndael\_d   & 952    & 0.7K-35M                   & Encrypted files                \\ \hline
-sha           & 197    & 0.6K-35M                   & Files of any format            \\ \hline
-%CRC32         & 130    & 0.6K-35M                   & Files of any format            \\ \hline
-%bzip2e        & 5125   & 0.7K-57M                   & Files of any format            \\ \hline
-%bzip2d        & 5125   & 0.2K-25M                   & Compressed files               \\ \hline
+CRC32         & 130    & 0.6K-35M                   & Files of any format            \\ \hline
+\rowcolor{gray!20}
+bzip2e        & 5125   & 0.7K-57M                   & Files of any format            \\ \hline
+\rowcolor{gray!20}
+bzip2d        & 5125   & 0.2K-25M                   & Compressed files               \\ \hline
 \end{tabular}
@@
 \end{table}
+
+The shaded (grey) benchmarks in Table~\ref{tab:kdatasets} represent the benchmarks used for training the cost model used for computing the weight of the instructions for the work metric.
+These same training benchmarks were also used for collecting a fixed set of optimisations for the {\itercomp}.
+The remaining (white) benchmarks are used for the experimental evaluation.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{figs/instr.pdf}
+    \includegraphics[width=\textwidth]{figs/num-probes.pdf}
     \caption{Percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.}
-    \label{fig:instr}
+    \label{fig:num-probes}
 \end{figure*}
@@
 
+\begin{figure*}[h]
+    \centering
+    \includegraphics[width=\textwidth]{figs/ipc-vs-work.pdf}
+    \caption{Comparison between {\itercomp} using IPC or the WP metric.}
+    \label{fig:ipc-vs-work}
+\end{figure*}
+
+Figure~\ref{fig:ipc-vs-work} presents a comparison between IPC and the WP metric regarding their correlation with the actual speedup over \verb|-O0| observed during {\itercomp}.
+Each point in the figure correspond to each metric averaged over input windows collected during multiple executions of the {\itercomp}.
+The vertical lines correspond to the optimisation sequence selected by the {\itercomp} search, which either represents the highest IPC or the highest value of WP.
+In all cases, the optimisation sequence selected based on the WP is much closer to the best speedup than the optimisation sequence selected based on IPC, which can be attributed to the fact that the WP have a stronger correlation with the speedup.
+In all cases, IPC shows little to no correlation with the speedup, which corroborates the argument given in Section~\ref{sec:ipc-vs-work-metric}.
+
+For some cases, namely for the tiff-related benchmarks, the WP shows a worse correlation with speedup.
+We believe these cases could be improved by improving the cost model used to compute the work metric.
+However, even in these cases, the optimisation sequence selected based on the WP outperforms those selected based on IPC.
+
 \section{Related Work}
@@
 Aalameldeen and Wood~\cite{alameldeen06} in particular suggest that a simple work-related metric should be used if the unit of work is representative enough.
-Work-related metrics have already been largely used for measuring performance of throughput-oriented applications, for other applications, however, choosing an appropriate unit of work can be more challanging~\cite{alameldeen06}.
+Work-related metrics have already been largely used for measuring performance of throughput-oriented applications, for other applications, however, choosing an appropriate unit of work can be more challenging~\cite{alameldeen06}.
 
@@
 
-%\textbf{Acknowledge the partial support by EPSRC}.
-
 This work was supported by the UK Engineering
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{ref}
+\bibliography{bibliography}
 

### 2017-08-22 cgo18mscr.tex
minor change
updates

@@
 \usepackage{mathtools}
+
+\theoremstyle{definition}
+\newtheorem{prop}{Proposition}[section]
 
@@
 
-\newcommand{\etal}{et~al.}
-
 % Copyright
@@
 
+\newcommand{\etal}{et~al.}
+
 \newcommand{\itercomp}{{iterative compilation}}
@@
 \newcommand{\IterComp}{{Iterative Compilation}}
+
+\newcommand{\flagstype}{\usefont{T1}{cmr}{m}{n}}
 
@@
 
-\author{ }
+\author{ \vspace{10em} }
 \affiliation{%
@@
 
-In this section we define the work-based performance metric used in this paper.
+In this section we define the work-based performance (WP) metric proposed for comparing different optimised versions of a program when executing with different inputs.
 We define the performance metric as the ratio between the amount of \textit{work}, $\Delta W$, performed during a period of time, $\Delta t$.
@@
 %The hypothesis is that, given two optimisations $o_1$ and $o_2$, a program compiled with optimisation $o_2$ would \textit{consistently} perform better than when it is compiled with $o_1$, i.e., $o_1 \tilde{<}_p \, o_2$, if it performs more \textit{work} per unit of time when compiled with $o_2$ instead of $o_1$.
-By measuring the amount of \textit{work} done per unit of time we reduce the impact of input-dependent aspects and focus instead on the efficiency of the optimised program.
+By measuring the amount of \textit{work} done per unit of time, we reduce the impact of input-dependent aspects and focus instead on the efficiency of the optimised program.
 For this metric, the main challenge is to precisely define what represents \textit{work}.
+As discussed in Section~\ref{sec:related-metrics}, many work metrics have been presented in the literature, such as block frequency and others.
+However, for comparing the performance benefits of different compiler optimisations on a program, although block frequency would be able to capture aspects of optimisations that simplify the control-flow graph (CFG), we argue that measuring work only at the basic block resolution would not be enough for capturing the effects of optimisations at the instruction level.
+For this reason, we extend the idea of using basic block frequency to measure computational work by also considering the computational cost of each basic block.
 
 %For that purpose, we intend to use profiling techniques to measure the input size and amount of computation performed on the execution path triggered by the input~\cite{ball94,ball96,zaparanuks12,coppa14}.
-
 %In addition to being useful for speeding up {\itercomp} across a large number of different inputs, this metric based on comparing work-based efficiencies could
@@
 %However, the overhead for estimating the amount of \textit{work} must be low enough for the performance gains to be beneficial.
-
-We model the computational work $\Delta W$ as a linear equation based on block frequency information and a cost-model of the instruction set.
+To this end, we model the computational work $\Delta W$ as a linear equation based on block frequency information and a cost model of the instruction set.
+Formally,
 \[
@@
 
-In this simplified model, we consider that $w_i$ is constant across all programs and executions in the given target platform.
-However, $N_B(i)$ is program dependent but constant across executions, while $f(B)$ is both program and execution dependent, since $f(B)$ can change when executing with different inputs.
-In other words, $N_B(i)$ is a static value known at compile-time and $f(B)$ is a dynamic value known only at run-time.
+In this simplified model, we consider that $w_i$ is constant across all programs and executions, varying only between target architectures.
+On the other hand, $N_B(i)$ is a program-dependent static value which is known at compile-time and $f(B)$ is a dynamic value known only at run-time,
+since $f(B)$ is both program and execution dependent as the execution frequency of a basic block can change when executing with different inputs.
+%However, $N_B(i)$ is program dependent but constant across executions, while $f(B)$ is both program and execution dependent, since $f(B)$ can change when executing with different inputs.
+%In other words, $N_B(i)$ is a static value known at compile-time and $f(B)$ is a dynamic value known only at run-time.
+
 %If we define $x_i$ as
@@
 
-Similarly to previous work~\cite{giusto01,powell09,brandolese11}, we derive the cost model for the instruction set by modelling the problem as a multi-variable linear regression, where the \textit{regression coefficients} are the costs of the instructions and the \textit{regressors} (or \textit{explanatory variables}) are computed as $\sum_B N_B(i)f(B)$ for each instruction $i$.
-\[
+\subsection{Estimating a Cost Model of the Instructions}
+
+Similarly to previous work~\citep{giusto01,powell09,brandolese11}, we derive the cost model of the instruction set by modelling the problem as a multi-variable linear regression, where the \textit{regression coefficients} are the costs of the instructions and the \textit{regressors} are computed as $\sum_B N_B(i)f(B)$ for each instruction.
+\begin{equation}\label{eq:linear-work-expression}
 \Delta W = \varepsilon + \sum_{i} \left(w_i \sum_{B} N_B(i)f(B)\right)
-\]
+\end{equation}
+By having some empirical data after executing several benchmarks with different inputs, we can fit the linear model with this empirical data in order to obtain estimate costs of the instructions.
+In order to fit the linear model, we measure the wall-clock time when executing the training benchmarks described in Section\ref{sec:benchmarks} with their respective 1000 input datasets.
+For these measurements, the training benchmarks are compiled without optimisation.
+The reason for using no optimisation, as discussed in Section\ref{sec:oic-infra}, is because the amount of work must be only input dependent and consistent between different optimisation sequences.
+This aspect is crucial for the work-based performance metric to enable a consistent comparison between the various optimisation sequences when executing with distinct inputs.
+%it allows to use the estimated cost model for computing a work metric which is independent of optimisations, as we explained in Section~\ref{}.
 
-By having some empirical data after executing several benchmarks with different inputs, we can fit the linear model with this empirical data in order to obtain the costs of the instructions.
+Our cost model has a total of 52 LLVM instructions\footnote{We do not model all LLVM instructions because some instructions are more common in optimised programs, such as the vector instructions.}.
+Every program, in the set of training benchmarks, was compiled twice: once including the necessary instrumentation for the block frequency profiling, which is required for deriving the linear expression defined in Equation~\ref{eq:linear-work-expression};
+and a standard compilation without any optimisation or instrumentation.
+For each input in the training dataset, the benchmarks were executed once with the instrumented version for collecting the block frequency profiling, and multiple times with the standard compilation just for measuring the wall-clock execution time, until the confidence interval was no larger than 1\% for a 99\% confidence.
+After collecting these measurements, this data can be used to estimate the unknown parameters in the linear model.
+Because we fit the linear model based on the wall-clock execution time, the derived cost model can be interpreted as an estimate of the execution time when the program is compiled without optimisations.
+
+\begin{figure}[htb]
+    \centering
+    \includegraphics[width=0.9\linewidth]{figs/cost-model.pdf}
+    \caption{Linear model fitted from empirical data. The mean absolute error (MAE) for the fitted curve is seven milliseconds.}
+    \label{fig:cost-model}
+\end{figure}
+
+Figure~\ref{fig:cost-model} compare the work metric with the corresponding execution time for some instances of the test benchmarks.
+Notice how the fitted model has a higher relative error for the instances with very short execution time, namely those that run for less than one tenth of a second.
+The mean absolute error (MAE) for the fitted curve is seven milliseconds.
+
+\subsection{Comparison with Instructions Per Cycle} \label{sec:ipc-vs-work-metric}
+
+The IPC metric has been widely used for studying performance benefits of hardware optimisations.
+Although previous work has suggested the use of IPC for guiding {\itercomp}, in this section we argue in favour of the WP metric over IPC.
+
+The IPC metric differs from the proposed WP metric in a key aspect:
+the IPC metric is computed solely based on the final optimised program.
+When executing different optimised versions of a program with the same input, both the number of instructions and the number of clock cycles can change.
+For this reason, higher IPC does not necessarily translate to shorter execution time (or even fewer clock cycles).
+
+We can illustrate this fact with a very small example as shown in Table~\ref{tab:ipc-example}.
+This example shows two versions of a program, namely P1 and P2, and their respective measurements related to IPC.
+Although version P1 has twice the IPC of P2, P1 is one cycle slower than P2.
+As this example illustrates, the IPC metric can be misleading when compared different versions of the same program.
+This problem is only intensified when both the optimisation sequence and the input changes.
+In Chapter~\ref{chap:eval} we show empirical evidence for this argument.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{|c|c|c|c|}
+\hline
+                       & P1 & P2  \\
+\hline
+Number of Instructions & 5  & 2   \\
+Number of Cycles       & 5  & 4   \\
+IPC                    & 1  & 0.5 \\
+\hline
+\end{tabular}
+\caption{Example that illustrates that a higher IPC does not necessarily translate to shorter execution time.}
+\label{tab:ipc-example}
+\end{table}
+
+On the other hand, WP computes the amount of work based on the unoptimised version of the program, which means that it always measures the same amount of work for the same input, regardless of the optimisations applied on the program.
+This is a key aspect that enables the use of the WP metric for guiding {\itercomp}, because a higher WP, which represents work per unit time, naturally translate to shorter execution time.
+Moreover, previous work has also presented other arguments against the use of IPC in similar use-case scenarios, as discussed in Chapter~\ref{chap:related}.
+
+\section{Online {\IterComp} Infrastructure} \label{sec:oic-infra}
+
+Although {\itercomp} had been originally proposed as an \textit{offline} optimisation strategy, it can also be adapted to work in online scenarios.
+Instead of selecting the best optimisation sequence as part of the development time (pre-shipping) of a program, a first version of the program is shipped together with an {\itercomp} mechanism.
+In the online scenario, the program is shipped with an initial optimisation sequence and different optimisation sequences are evaluated as the end-user executes the program.
+This optimisation strategy is also known as idle-time optimisation, as the re-compilation happens between runs of the program.
+
+LLVM is particularly suitable for iterative compilation as it makes possible to cache a pre-compiled, but still unoptimised, version of the input program in the bitcode format of the LLVM IR.  
+This caching allows to speedup the time required for re-compilation as it is able to bypass the frontend phase.
+If re-compilation time is critical, it would also be possible to keep only the hot portion of the code in the LLVM bitcode format, while the remaining portion of the code is already compiled to the final object code.
+However, this is out of the scope of this thesis and we always re-compile the whole program.
+
+\begin{figure}[htb]
+    \centering
+    \includegraphics[width=\linewidth]{figs/infra-diagram}
+    \caption{Overview of the execution engine for applying {\itercomp}.}
+    \label{fig:infra-diagram}
+\end{figure}
+%\textbf{Describe the process step by step.}
+
+Figure~\ref{fig:infra-diagram} shows an overview of the infrastructure required for applying online {\itercomp}
+using WP as the metric of choice for evaluating optimisation sequences.
+The online {\itercomp} follows as described bellow:
+\begin{enumerate}
+\item The program is pre-compiled to the LLVM bitcode format without optimisation.
+\item The unoptimised program is instrumented for work profiling.
+\item Execution-based optimisation search:
+ \begin{enumerate}
+   \item The current optimisation sequence is used for the re-compilation of the program.
+   \item The program is executed with any input provided by the end-user.
+         During the execution of the program, wall-clock time and the work metric are recorded by profiling instrumentation.
+   \item If the recorded profiling for the current optimisation can be used to compute an average performance measurement within a small confidence interval,
+         then a new optimisation sequence is generated.
+         Otherwise, the same optimisation sequence is used for the next execution.
+ \end{enumerate}
+\end{enumerate}
+
+In this work, we focus mainly on the highlighted components.
+The \textit{Work Instrumentation} phase focuses on providing a low-overhead instrumentation for profiling the work metric.
+The instrumentation consists of adding a global counter to the program which is used to accumulate the amount of work computed during the program's execution, using the cost model of the instruction set.
+A detailed description of the work instrumentation is presented in Section~\ref{chap:instr}.
+
+Notice how the \textit{Work Instrumentation} phase is executed before performing any optimisation to the program.
+It is critical so that the work profiling always measure the same amount of work for a given input, regardless of the optimisation sequence applied to the program.
+Because the instrumentation is performed before optimising the program, it means that the work profiling derives the linear expression defined in Equation~\ref{eq:linear-work-expression}
+based on the unoptimised program.
+In other words, the basic blocks and the number of occurrences of the instruction in the basic blocks reflect the unoptimised program.
+This particular sequence of compilation guarantees that the amount of work must is only input dependent, but consistent between different optimisation sequences.
+
+%In both phases the goal is to improve the performance of the execution phase, but from different perspectives.
+%\textit{(i.)} In the compilation phase we focused on providing a low-overhead instrumentation for profiling the work metric.
+This phase is responsible to improve performance by lowering the overhead of the work profiling.
+%\textit{(ii.)} In the evaluation phase we proposed the use of the WP metric in order to enable {\itercomp} in an online scenario.
+%This phase is responsible to improve performance by being able to properly assess different optimisations such that the best optimisation can be selected.
+%Chapter~\ref{chap:instr} describes our work profiling strategy and Section~\ref{sec:metric} describes the work-based metric.
+
+%\textbf{Talk about the process of keeping the same optimization for a given input-window and how the optimizations as selected based on the evaluation of these "evidences".
+Perhaps we could suggest using the Theory of Evidences at this point?!}
+
+As the name suggests, the component called \textit{Optimisation Selector} is responsible for selecting which optimisation sequence to use for the next execution of the program.
+It can either keep the same optimisation sequence used in the previous execution or start monitoring the performance of a new optimisation sequence.
+An optimisation sequence can be kept for multiple executions of the program in order to gather enough measurement to compute an average performance with small statistical deviations, i.e., for which the confidence interval has a small range.
+We call by \textit{Input-Window Size} the number of executions performed using the same optimisation sequence.
+
+\subsection{Real Online Scenarios}
+
+In most online scenarios, it is common for periods of peak usage and idle periods.
+For example, mobile devices are usually intensely used during the day, with some idle periods usually when the battery is being re-charged~\citep{mpeis16}.
+Similarly, many authors have also considered peak and idle (or underutilised) periods in the context of data centres~\citep{armbrust10,chen12b}.
+
+The proposed infrastructure is very well suited for these real online scenarios\footnote{
+However, if idle time is almost non-existent, the proposed infrastructure can still be used by re-compiling the program with a different optimisation while multiple runs of the program are being executed.}.
+In particular, periods of peak usage could be used to monitor multiple runs of the program using the same optimisation sequence, collecting the work profiling and measuring its execution time,
+while periods of idleness or underutilisation could be leveraged to use the profiling statistics for selecting better optimisations and re-compiling the program.
 
@@
 In order to further reduce the overhead in these critical cases, we propose a relaxed instrumentation by trading off accuracy and overhead.
+While the optimal placement of probes tries to place probes in edges that are less likely to be executed, the relaxation focus on removing probes that are more likely to be executed but add little to the work metric.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[scale=0.9]{figs/relax-instr-diagram.pdf}
+  \caption{Overview of the work instrumentation algorithm, including the relaxation technique.}
+  \label{fig:relax-instr-diagram}
+\end{figure}
 
@@
 
-%Define an extended DAG of a CFG as a directed graph which may contain cycles if they have known constant trip counts.
+%Define an DAG of a CFG as a directed graph which may contain cycles if they have known constant trip counts.
+Figure~\ref{fig:relax-instr-diagram} shows an overview of the relaxed instrumentation algorithm.
+The highlighted step is introduced by the relaxed instrumentation on top of the previously defined optimal profiling.
+The relaxed instrumentation performs a post-processing on the resulting instrumentation of the optimal algorithm.
+This post-processing identifies probes that add little to the work metric, and removes their instrumentation.
+In order to guarantee an upper bound for the dynamic error in the profiling measurement, the relaxation algorithm applies a constrained post-processing on a per DAG (directed acyclic graph) basis.
+By constraining the relaxation within each DAG by a maximum allowed static error, it guarantees that the overall relaxation will also be constrained by the same bound.
 
-The relaxed instrumentation performs a post-processing on the resulting instrumentation of the optimal algorithm.
-The relaxation starts by extracting \textit{extended} DAGs (directed acyclic graphs) from the CFG, as illustrated in Figure~\ref{fig:cfg-relax-example}.
-Each loop (and the outer most region of the function) represents an extended DAG where any inner loop is considered to never execute, i.e.
-only the headers of the inner loops are included into the extended DAG.
-
+The relaxation starts by extracting DAGs (directed acyclic graphs) from the CFG, as illustrated in Figure~\ref{fig:cfg-relax-example}.
+First, the algorithm extracts all the subgraphs that represent a loop or the outer most region of the function.
+Section~\ref{sec:natural-loops} explains how loops can be identified in the CFG.
+Afterwards, these subgraphs are transformed into DAGs by ignoring the backedge and also by considering that any loop within the subgraph is never executed, i.e., only the headers of the inner loops are actually included into the DAG.
+Figure~\ref{fig:cfg-relax-example} shows a CFG partitioned into two DAGs (consider only basic blocks and edges completely inside the yellow and green boundaries).
 %
@@
   \centering
-  \includegraphics[scale=0.9]{figs/cfg-relax-example.pdf}
-  \caption{Example of a CFG containing a loop and its decomposition into extended DAGs when applying the relaxation.}
+  \includegraphics[scale=1]{figs/cfg-relax-example.pdf}
+  \caption{Example of a CFG containing a loop and its decomposition into DAGs when applying the relaxation.
+           The DAGs are the subgraphs within the dashed boundaries.}
   \label{fig:cfg-relax-example}
@@
 
-For every extended DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we relax the instrumentation accuracy by selecting a subset of the probes to be removed, subject to the maximum allowed percentage error, $M$.
+\newpage
+\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
+// Input: CFG	
+relaxInstrumentation(G) {
+  for loop in G:
+     DAG = colapseInnerLoops(loop)
+     relaxInstrumentedDAG(DAG)
+  DAG = colapseInnerLoops(G)
+  relaxInstrumentedDAG(DAG)
+}
+\end{lstlisting}
+
+For every DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we relax the profiling accuracy by selecting a subset of the probes to be removed, subject to the maximum allowed percentage error, $M$.
 
 We model the relaxation as a 0-1 Knapsack problem:
+\begin{equation*}
+\begin{aligned}
+& \textrm{maximise }\quad \sum_{i=0}^{k} f(P_i)x_i \\
+& \textrm{subject to }\quad \sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M
+\end{aligned}
+\end{equation*}
 \[
-%\textrm{maximise } \sum_{i=0}^{k} f(I_i)c(I_i)x_i
-\textrm{maximise } \sum_{i=0}^{k} f(P_i)x_i
+x_i\in\{0,1\}, i\in\{0,\ldots,k\}
 \]
+where $f(P_i)$ is the execution frequency of probe $P_i$, $x_i$ denotes the probes selected for removal, and $\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to the minimum work value possible to compute in the DAG, i.e., if $m$ is the minimum amount of work possible to be computed when executing the DAG, then $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$.
+Because the percentage error is computed based on the path with the minimum amount of work, $\varepsilon(P_i)$ represents the maximum error possible that would be incurred when removing probe $P_i$.
+Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold, as demonstrated by Proposition~\ref{prop:relax-bound}.
+
+\begin{prop}\label{prop:relax-bound}
+Let $n_i$ be the number of times a given DAG $i$ is executed, $r_i$ be the total relaxation (amount of work removed) in DAG $i$, and $m_i$ be its minimum amount of work.
+If $\frac{r_i}{m_i} \leq M$ for every $i$,
+then the final error of the relaxation will always be bounded by the same threshold.
+\end{prop}
+\begin{proof}
+We can model the overall error of the relaxation as:
 \[
-\textrm{subject to } \sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \textrm{ and } x_i\in\{0,1\}
+1 - \frac{n_1(m_1 - r_1) + n_2(m_2 - r_2) + \ldots + n_k(m_k - r_k) + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
 \]
-%$c(I_i)$ is the relative overhead of the instrumentation on the instrumented basic block,
-where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$, $\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to the minimum work value possible to compute in the extended DAG, and $x_i$ denotes the probes selected for removal.
-Because the percentage error is computed based on the path with the minimum amount of work, $\varepsilon(P_i)$ represents the maximum error possible that would be incurred when removing probe $P_i$.
-Furthermore, by constraining the percentage error of every extended DAG below a given threshold we guarantee that the final error of the relaxation will always be bounded by the threshold.
+That is,
+%\begin{equation*}
+\begin{gather*}
+ 1 - \frac{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} + \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} = \\
+ \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
+\end{gather*}
+%\end{equation*}
+If $\frac{r_j}{m_j}$ is the maximum ratio $\frac{r_i}{m_i}$ for every $i$, then
+\begin{equation*}
+\begin{aligned}
+ \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} &\leq\\
+ \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j + c} &\leq\\
+ \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} & 
+\end{aligned}
+\end{equation*}
+If $N = max\{n_i$ for every $i\}$, then
+\begin{equation*}
+\begin{aligned}
+ \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &\leq\\
+ \frac{Nr_j + Nr_j + \ldots + Nr_j}{Nm_j + Nm_j + \ldots + Nm_j} &=\\
+ \frac{Nkr_j}{Nkm_j} = \frac{r_j}{m_j} &\leq M
+\end{aligned}
+\end{equation*}
+\end{proof}
 
-While the optimal placement of probes tries to place probes in edges that are less likely to be executed, the relaxation focus on removing probes that are more likely to be executed.
-The necessary block frequency information for optimising both the placement of probes can be acquired from profiles of previous executions of the program or by a static heuristic of the CFG during compilation.
+\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
+// Input: CFG	
+relaxInstrumentedDAG(DAG){
+   P = ProbesIn(DAG)
+   m = minWork(DAG,P)
+   K = createKnapsackModel(P,m)
+   Bag = solveKnapsack(K)
+   for B in (P-Bag):
+     removeProbe(B)
+}
+\end{lstlisting}
+
+The necessary block-frequency information for optimising both the placement of probes can be acquired from profiles of previous executions of the program or by a static heuristic of the CFG during compilation.
 
@@
 the optimal brute-force solver;
-the greedy heuristic based on sorting the items ~\cite{dantzig57}.
+the greedy heuristic based on sorting the items~\citep{dantzig57}.
 We use the brute-force solver for DAGs with a small number of probes and the greedy heuristic when the number of probes is greater than a threshold.
 Some of the benchmarks have DAGs with several hundreds of probes, which could result in a long compilation time.
+
+\subsection{Whole Program Relaxation}
+
+In some cases, the proposed relaxation can be very conservative, because it considers the static error of removing probe $P_i$ relative to the minimum work possible of a DAG, in order to be able to guarantee that the dynamic error will be bounded by a given threshold.
+This conservatism can be overly restrictive in some cases, resulting in a negligible overhead reduction but also causing just a negligible dynamic error to the work profiling.
+
+In some cases, this overly restrictive conservatism may be unnecessary.
+For these cases, we propose an adapted version of the relaxation algorithm that operates on the whole program.
+Traditionally, compilers optimisations are performed on the function-level, or at best on a per module basis.
+Whole program optimisation (WPO) means that the compiler considers all compilation units of the program and optimises them using the combined knowledge of how they are used together.
+
+The whole program relaxation works by using block-frequency profiling from previous executions.
+By having this profiling information, the \textit{whole program relaxation} is able to compute the error of removing a given probe in terms of the whole program's execution,
+and then use this error values for selecting a subset of all the probes to be removed.
+
+For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we model the whole program the relaxation as the following 0-1 Knapsack problem:
+\begin{equation*}
+\begin{aligned}
+& \textrm{maximise }\quad \sum_{i=0}^{k} f(P_i)x_i \\
+& \textrm{subject to }\quad \sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M
+\end{aligned}
+\end{equation*}
+\[
+x_i\in\{0,1\}, i\in\{0,\ldots,k\}
+\]
+where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$, $x_i$ denotes the probes selected for removal, $M$ is the error threshold, and $\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to the profiled global work, i.e.,
+if $\Delta W$ is the work value for the whole program's execution, computed from the basic block frequencies profiled from previous executions, the error for a given probe $P_i$ is
+\[
+\varepsilon(P_i) = \frac{\omega(P_i)f(P_i)}{\Delta W}.
+\]
+
+Contrary to the per DAG relaxation, the whole program relaxation is not guaranteed to be bounded by the error threshold $M$,
+as it depends on the representativity % representativeness
+ of the profiling information provided to the whole program relaxation.
 
@@
 
+For comparison, we use four configurations\footnote{A configuration using the WPO relaxation was not used due to the time necessary for executing the experiments.}.
+In all configurations, the same optimisation sequence is used for multiple inputs, using a dynamic input-window size, as explained in Section~\ref{sec:oic-infra}.
+The average performance over the input window provides an estimate for the overall performance of the optimisation sequence across distinct inputs.
+The optimisation sequences are ranked based on their average performance, and the best optimisation sequence is selected.
+%When selecting the best optimisation sequence, they are ranked by the average of their performance measurement, using the WP metric, except for the Oracle-RM which uses the real speedup over \texttt{-O0}.
 \begin{itemize}
-\item Oracle-RM executes the program twice, for each input, and then measuring the real speedup for each compiler optimisation and uses the real speedup for selecting the best optimisation.
-\item Oracle-PP represents the \textit{perfect} non-intrusive profiling by also executing the program twice, once for estimating the amount of work and the second for measuring its execution time.
-This version uses the work-based metric, $\frac{\Delta W}{\Delta t}$, during the {\itercomp} search.
+\item \textbf{Oracle-RM} executes the program multiple times for each input, measuring the real speedup for each optimisation sequence, and then uses the real speedup over {\flagstype -O0} for comparing the optimisation sequences.
+  The speedups are computed based on the wall-clock time.
+  In order to reduce noise, the program is executed several times for the same input, until the confidence interval was no larger than 1\% for a 99\% confidence.
+\newpage
+\item \textbf{Oracle-PP} represents an oracle with a \textit{perfect} non-intrusive profiling.
+  Although it uses the WP metric for comparing optimisation sequences, this oracle also avoids noise in its measurements by also executing the program multiple times for each input.
+  The first execution is used for profiling the work metric.
+  The remaining executions are used for measuring the wall-clock time without using the work profiling.
+\item \textbf{Real-OP} corresponds to the online {\itercomp} as it would be applied in a real online scenario.
+  For each optimisation, a random sample of inputs is selected, and the program is executed only once with each input.
+  When executing each input, it uses the optimal work instrumentation for profiling the work metric.
+  The average of the WP for the sample of inputs is then used for selecting the best optimisation.
+\item \textbf{Real-R5} is similar to the {Real-OP}.
+  It also corresponds to the online {\itercomp} as it would be applied in a real online scenario.
+  For each optimisation, a random sample of inputs is selected, and the program is executed only once with each input.
+  When executing each input, it uses the relaxed work instrumentation, with a 5\% threshold, for profiling the work metric.
+  The average of the WP for the sample of inputs is then used for selecting the best optimisation.
+\item \textbf{Real-IPC} corresponds to the online {\itercomp} based on the IPC metric.
+  For each optimisation, a random sample of inputs is selected, and the program is executed only once with each input.
+  The average of the IPC metric for the sample of inputs is then used for selecting the best optimisation.
 \end{itemize}
+The comparison between Oracle-RM and Oracle-PP is useful for validating the use of the WP metric for guiding {\itercomp}, while the other configurations demonstrate the viability of applying the online {\itercomp} in a real scenario.
+
+Figure~\ref{fig:window-size} shows the average input-window size for each benchmark and configuration.
+It shows that we can have a statistically sound measurement of the performance metric using just a small number of inputs.
+
+\begin{figure}[htb]
+    \centering
+    \includegraphics[width=\textwidth]{figs/window-size.pdf}
+    \caption{Average input-window sizes observed during the online {\itercomp}.}
+    \label{fig:window-size}
+\end{figure}
+
+\subsection{The Set of Optimisation Sequences}
+
+For the purpose of evaluating the use of the WP metric with {\itercomp}, we collected in advance a fixed set of optimisation sequences.
+The reason for using this fixed set, as explained in Section~\ref{sec:oic-infra}, is because this thesis is focused on other components of the infrastructure for performing {\itercomp} and we assume that a good generator of optimisation sequences will be used in a real online scenario.
+This set contains 500 optimisation sequences collected in a random search using the training benchmarks.
+These optimisation sequences contain an average of 40 individual optimisation passes, including repetitions, with a maximum of 119 optimisation passes, but it also contains optimisation sequences which consist of a single flag, such as the default optimisations {\flagstype -O1}, {\flagstype -O2}, {\flagstype -O3}, {\flagstype -Os}, and {\flagstype -Oz}.
+
+  \begin{minipage}{0.9\textwidth}
+     \vspace{1em}
+     \noindent\textbf{Example of a short optimisation sequence:}\vspace{-1ex}
+     \justify{\flagstype -mem2reg -simplifycfg -constprop -dce}
+  \end{minipage}
+
+  \begin{minipage}{0.9\textwidth}
+     \vspace{1em}
+     \noindent\textbf{Example of a long optimisation sequence:}\vspace{-1ex}
+     \justify{\flagstype -globalopt -reassociate -instcombine -loop-rotate -block-freq -deadargelim -early-cse -sroa -argpromotion -sccp -tbaa -barrier -constmerge \mbox{-loop-vectorize} -domtree -basicaa -memdep -basiccg -memcpyopt \mbox{-constprop} -adce -globaldce -mem2reg -constmerge \mbox{-globaldce} -constprop -instsimplify -dse -dce -simplifycfg -loop-unroll -reassociate -constprop \mbox{-globaldce} -instsimplify -adce -constmerge -bb-vectorize -dce -mergefunc -simplifycfg -dse -loop-unroll -globaldce}
+  \end{minipage}
+
+  \begin{minipage}{0.9\textwidth}
+     \vspace{1em}
+     \noindent\textbf{Example of an optimisation sequence which includes {-O3}:}\vspace{-1ex}
+     \justify{\flagstype -O3 -adce -globaldce -simplifycfg -memcpyopt -reassociate -mergefunc \mbox{-dce} -dse}
+     \vspace{2em}
+  \end{minipage}
+
+Repeating the same optimisation pass can be beneficial and usually expected by other passes.
+For example, the {\flagstype -loop-simplify} pass is used for transforming loops into a canonical form by inserting pre-header and exit basic blocks.
+Although this pass inserts jumps due to redundant basic blocks, this canonical form can be favourable to other loop optimisations.
+Because of the redundant basic blocks, this optimisation pass expects that the {\flagstype -simplifycfg} will eventually be executor later on the optimisation pipeline.
+Another example of such inter-relation between transformations concerns the {\flagstype -licm} and {\flagstype -mem2reg} passes.
+The {\flagstype -licm} pass is responsible for moving invariant code out from the loop body.
+It usually creates new local variables, using memory access operations, for assisting with the code manipulation, which means that the executing the {\flagstype -mem2reg} pass afterwards would be useful as a cleanup pass for removing the extra memory accesses generated.
+However, many of the analysis required for identifying loop invariant also benefit from the transformations performed by the {\flagstype -mem2reg} pass.
+These examples illustrate the importance of repeating optimisation passes.
+Moreover, they illustrate the intricate relation amongst several transformations.
+
+However, all optimisation sequences in the set of optimisations were generated completely at random, without using any knowledge of individual transformations.
+Each optimisation sequence was generated in two steps: \textit{(1)} randomly selects the number of flags; \textit{(2)} randomly selects the flags, allowing repetitions.
+Afterwards, this randomly generated optimisation sequence would be included in the set of optimisation sequences only if it was able to improve the performance of a training benchmark, also selected at random, in respect of the {\flagstype -O3} default optimisation.
+This process was repeated until we obtained all the 500 distinct optimisation sequences.
+
+\subsection{Performance Evaluation}
+
+In order to evaluate the quality of the final optimisation sequences selected by the {\itercomp} search, we compare their speedup by measuring wall-clock time of the benchmarks when compiled with the standard {\flagstype -O3} optimisation.
+For each benchmark, after selecting the final optimisation sequence, we compute the average speedup over all the 1000 input dataset.
+When measuring the wall-clock time for each input, to reduce noise, we execute the same input until we have a statistically sound measurement, i.e.
+we execute until we have an interval no larger than 1\% with 99\% confidence.
+Figure~\ref{fig:speedups} shows these average speedups over all the 1000 input dataset.
+This figure shows that the best optimisation sequence selected with the Oracle-PP is very close to the performance of the best optimisation sequence selected with the Oracle-RM.
+This result is important for demonstrating that a work-based metric has the potential to produce good results in a real online scenario, where there is the restriction that programs execute distinct inputs only once.
+Moreover, the evaluation also indicates that the use of relaxation algorithms does not degrade the optimisation search,
+it might even benefit the {\itercomp} as it tends to insert smaller interferences in the performance measurement.
+Finally, Figure~\ref{fig:speedups} also presents the results of {\itercomp} guided by IPC, which is futher discussed in Figure~\ref{fig:ipc-vs-work}.
 
@@
     \includegraphics[width=\textwidth]{figs/speedups.pdf}
-    \caption{Speedups observed with the online {\itercomp}.}
+    \caption{Speedups obtained from the final optimisation sequence selected by the online {\itercomp}.
+	         The speedups reported for each benchmark represents the average speedup across their complete 1000 input datasets.}
     \label{fig:speedups}
 \end{figure*}
-
-%\begin{figure*}[htb]
-%    \centering
-%    \includegraphics[width=\textwidth]{figs/profiled-speedups.pdf}
-%    \caption{Speedups observed with the online {\itercomp} if we
-%             consider the instrumentation overhead.}
-%    \label{fig:profiled-speedups}
-%\end{figure*}
-
-%\subsubsection{Contribution of individual optimisation passes}
-%
-%Figure~\ref{fig:flagsfreq} shows an aggregated view of the final combination of compiler optimisations that were selected by {\itercomp} search.
-%The figure presents the individual optimisation passes with at least 1\% of frequency in the selected combination of compiler optimisations that improved the performance over {\texttt{-O3}}.
-%
-%\begin{figure*}[htb]
-%    \centering
-%    \includegraphics[width=\textwidth]{figs/flagsfreq.pdf}
-%    \caption{Frequency of individual optimisation passes on the final selected 
-%             compiler optimisations of the {\itercomp} search over
-%             all benchmarks.}
-%    \label{fig:flagsfreq}
-%\end{figure*}
 
@@
 Figure~\ref{fig:ipc-vs-work} presents a comparison between IPC and the WP metric regarding their correlation with the actual speedup over \verb|-O0| observed during {\itercomp}.
-Each point in the figure correspond to each metric averaged over input windows collected during multiple executions of the {\itercomp}.
+Each point in the figure corresponds to each metric averaged over input windows collected during multiple executions of the {\itercomp}.
 The vertical lines correspond to the optimisation sequence selected by the {\itercomp} search, which either represents the highest IPC or the highest value of WP.
-In all cases, the optimisation sequence selected based on the WP is much closer to the best speedup than the optimisation sequence selected based on IPC, which can be attributed to the fact that the WP have a stronger correlation with the speedup.
+In all cases, the optimisation sequence selected based on the WP is much closer to the best speedup than the optimisation sequence selected based on IPC, which can be attributed to the fact that the WP has a stronger correlation with the speedup.
 In all cases, IPC shows little to no correlation with the speedup, which corroborates the argument given in Section~\ref{sec:ipc-vs-work-metric}.
@@
 We believe these cases could be improved by improving the cost model used to compute the work metric.
-However, even in these cases, the optimisation sequence selected based on the WP outperforms those selected based on IPC.
+However, even in these cases, the optimisation sequences selected based on the WP outperform those selected based on the IPC metric.
 

### 2017-08-30 cgo18mscr.tex
minor updates
updates

@@
 
-\author{ \vspace{10em} }
+\author{ \vspace{1em} }
 \affiliation{%
@@
 
-Each of these optimisations interacts with the code in complicated ways, depending on all other optimisations and the order they were applied to the code being optimised.
-Understanding the interactions of optimisations is very important in determining a good solution to the phase-ordering problem~\cite{touati06,kulkarni12}.
+Each of these optimisations interacts with the code in complicated ways, depending on all other optimisations and the order they are applied to the code being optimised.
+However, understanding the interactions of optimisations is very important in determining a good solution to the phase-ordering problem~\cite{touati06,kulkarni12}.
 Because of all those dependencies and interactions, although most compiler optimisations yield significant improvements in many programs, the potential for performance degradation in certain program patterns is known to compiler writers and many users~\cite{pan06,zhou12,kulkarni12}.
@@
 
-Efficiently selecting the best combination of compiler optimisations for a given program or program section remains an open problem.
+Efficiently selecting the best optimisation sequence for a given program, or program section, remains an open problem.
 Compiler writers typically use a combination of experience and insight to construct the default sequences of prearranged optimisations found in compilers.
-However, these default pre-arranged set of optimisations do not include all possible optimisations and they are always applied in the same pre-defined fixed order, without regard the code being optimised~\cite{pan06,cavazos07,zhou12,kulkarni12}.
+However, these pre-arranged optimisation sequences do not include all possible optimisations and they are always applied in the same pre-defined order, without regard the code being optimised~\cite{pan06,cavazos07,zhou12,kulkarni12}.
 %in hopes that a programmer will know which combination of optimisations will benefit their code~\cite{pan06,cavazos07,zhou12,kulkarni12}.
@@
 {\Itercomp} has the ability to adapt to new platforms, program and workload while still having a systematic and simple optimisation process.
-It works by repeatedly evaluating a large number of compiler optimisations until the best combination is found for a particular program~\cite{fursin07,chen10}.
+It works by repeatedly evaluating a large number of optimisation sequences until the best combination is found for a particular program~\cite{fursin07,chen10}.
 The main challenge concerning {\itercomp} is the need for efficiently exploring such a large optimisation space~\cite{fursin07,cavazos07,zhou12}.
@@
 Until recently, most of existing work  had been focusing on finding the best optimisation through repeated runs using a single input.
-Although they demonstrate the potential of {\itercomp}, in real scenarios the user rarely executes the same input dataset multiple times~\cite{bodin98,kisuki99,stephenson03,kulkarni04,agakov06}.
-Applying {\itercomp} in light of a single input may not result in good performance when executing the optimised code with different inputs.
-
-Most of real world applications are complex enough so that a single input case does not capture the whole range of possible scenarios and program behaviour~\cite{haneda06,fursin07,chen10,chen12a}.
+Although they demonstrate the potential of {\itercomp}, in real scenarios the user rarely executes a program with the same input multiple times~\cite{bodin98,kisuki99,stephenson03,kulkarni04,agakov06}.
+Furthermore, most of real world applications are complex enough so that a single input case does not capture the whole range of possible scenarios and program behaviour~\cite{haneda06,fursin07,chen10,chen12a}.
 Because programs can exhibit behaviours that differ greatly depending on the input,
+%applying {\itercomp} in light of a single input may not result in a good performance when executing the optimised code with different inputs.
+using a single input for {\itercomp} may not result in a good performance when executing the optimised code with different inputs.
+
 %This is something that is well known when writing test cases for correctness.
 %We can extend the same argument when optimising the code for performance.
-using a single input for {\itercomp} can produce poor performance when executed with different inputs.
-
-Recent work have been studying the impact of using multiple input datasets
-for performing {\itercomp}.
-This previous work suggests that optimising based on a representative range of input datasets allows for selecting a robust compiler optimisation that produces good performance in real scenarios where the input is expected to change constantly~\cite{haneda06,fursin07,chen10,chen12a,chen12b,fang15,mpeis16}.
-Their results show that a limited number of input datasets may be sufficient to obtain a well optimised program for a wider range of different inputs~\cite{haneda06,fursin07,chen10,chen12a}.
-
-Finding such a robust combination of compiler optimisations by means of {\itercomp} across a large range of inputs may be very time consuming.
-In order to speed up this process we intend to reduce the number of actual executions during the exploration of the optimisation space without much degradation of the final selected optimisation.
-
-\begin{figure}[htb]
-    \centering
-    \includegraphics[width=\linewidth]{figs/infra-diagram}
-    \caption{Overview of the execution engine for applying {\itercomp}.}
-    \label{fig:infra-diagram}
-\end{figure}
-
-In this paper, our main goal is to enable {\itercomp} in \textit{online} scenarios with the restriction of executing distinct inputs only once, while targeting optimal performance across different inputs.
-This online aspect is usually found in mobile and data centre platforms~\cite{chen12b,fang15,mpeis16}, where the goal is to optimise programs based on the workload of a particular user (or group of users) between executions.
-
-Measuring just execution time, for example, is useful only if the amount of work is constant between executions with different inputs.
-For this reason, we propose the use of a work-based metric in order to compare the performance of different optimisations across multiple executions of the program with distinct inputs.
-
-Because of the restriction of not repeating inputs, we instrument the program for measuring the amount of work it performed during its execution.
+%using a single input for {\itercomp} can produce poor performance when executed with different inputs.
+
+Recent work have been studying the impact of using multiple inputs for
+performing {\itercomp}.
+%This previous work suggests that optimising based on a representative range of
+%inputs allows for selecting a robust optimisation sequence that produces
+%good performance in real scenarios where the input is expected to change
+%constantly~\cite{haneda06,fursin07,chen10,chen12a,chen12b,fang15,mpeis16}.
+Their results suggest that optimising based on a relatively small set of
+representative inputs delivers good performance across a wide range of different
+inputs~\cite{haneda06,fursin07,chen10,chen12a,chen12b,fang15,mpeis16}.
+%Their results show that a limited number of inputs may be sufficient to obtain
+%a well-optimised program for a wide range of different
+%inputs~\cite{haneda06,fursin07,chen10,chen12a}.
+
+In this work, our main goal is to enable {\itercomp} in \textit{online} scenarios.
+For the purposes of this paper, we define the online scenario as having the restriction that programs execute multiple inputs and distinct inputs are executed only once.
+This online aspect is usually found in mobile and data centre platforms~\citep{chen12b,fang15,mpeis16}, where the goal is to optimise programs based on the workload of a particular user (or group of users) between executions.
+Online {\itercomp} must target optimal performance across different inputs.
+
+Because of the restriction of having a single execution per input, it is not possible to measure speedup for comparing optimisations.
+On the other hand, measuring just execution time, for example, is useful only if the amount of work is constant between executions with different inputs.
+While previous work have suggested using \textit{instructions per cycle} (IPC) for performing {\itercomp} in \textit{online} scenarios, IPC have no correlation with speedup~\citep{fursin07}.
+
+In order to address this challenge, we propose the use of a work-based metric to compare the performance of different optimisations across multiple executions of the program with distinct inputs.
+We instrument the program for measuring the amount of work it performed during its execution.
 Having a low overhead instrumentation is essential in this online scenario for two main reasons:
 $(i)$ the user is directly affected by large overheads;
-$(ii)$ a highly intrusive instrumentation can havea significant impact on the effect of the optimisations.
-
-Our main contributions are the following:
+$(ii)$ a highly intrusive instrumentation can have a significant impact on the effect of the optimisations.
+With the purpose of reducing profiling overhead, we propose two relaxation algorithms which provide a trade-off between measurement accuracy and overhead.
+The first is a relaxation algorithm that operates on the level of regions of functions, while the second performs the relaxation considering the whole program at the same time.
+
+Our experimental evaluation shows that performing online {\itercomp} guided by the work-based performance (WP) metric good results compared to the oracle, which is allowed to execute each input multiple times in order to use the actual speedup for guiding the {\itercomp}.
+Online {\itercomp} guided by the WP metric is able to achieve an average of 7.5\% and a maximum of 33\% improvement over the standard {\flagstype -O3} optimisation.
+Moreover, the experiments regarding the work profiling show that both relaxation algorithms are able to significantly reduce the profiling overhead while incurring a dynamic error of less than 5\% in the work measurement.
+The whole program relaxation achieves an average of $2\times$ reduction in the overhead compared with the optimal profiling technique, while the more conservative relaxation that operates per region achieves an average improvement of 40\% over the optimal profiling.
+
+%Our main contributions are the following:
+%\begin{itemize}
+%\item The use of a work-based performance metric in order to enable \textit{online} {\itercomp} by comparing different combination of compiler optimisations even when executed with distinct inputs.
+%\item We propose a relaxed instrumentation for low overhead profiling, with a controlled trade-off between accuracy and overhead.
+%\end{itemize}
+
+To summarise, the main contributions of this paper are the following:
 \begin{itemize}
-\item The use of a work-based performance metric in order to enable \textit{online} {\itercomp} by comparing different combination of compiler optimisations even when executed with distinct inputs.
-\item We propose a relaxed instrumentation for low overhead profiling, with a controlled trade-off between accuracy and overhead.
+\item We propose the use of a work-based performance (WP) metric for comparing different optimised versions of a program.
+      In particular, we use the WP metric to guide {\itercomp} in an online scenario, where the program is expected to execute only once for distinct inputs.
+
+\item Contrary to what previous work has suggested, we show that instructions per cycle (IPC) is not a good metric for online {\itercomp}.
+\item We adapt the block frequency profiling in order to measure the WP metric.
+\item A relaxation algorithm is proposed for lowering the overhead of the work profiling, with a controlled trade-off between accuracy and overhead.
+\item A whole program relaxation is proposed in order to further reduce the overhead of the work profiling.
 \end{itemize}
 
-\section{Work-based Metric} \label{sec:metric}
+\section{Work-based Performance Metric} \label{sec:metric}
 
@@
 
-%The hypothesis is that, given two optimisations $o_1$ and $o_2$, a program compiled with optimisation $o_2$ would \textit{consistently} perform better than when it is compiled with $o_1$, i.e., $o_1 \tilde{<}_p \, o_2$, if it performs more \textit{work} per unit of time when compiled with $o_2$ instead of $o_1$.
 By measuring the amount of \textit{work} done per unit of time, we reduce the impact of input-dependent aspects and focus instead on the efficiency of the optimised program.
@@
 
-%For that purpose, we intend to use profiling techniques to measure the input size and amount of computation performed on the execution path triggered by the input~\cite{ball94,ball96,zaparanuks12,coppa14}.
-%In addition to being useful for speeding up {\itercomp} across a large number of different inputs, this metric based on comparing work-based efficiencies could
-%potentially be used for applying {\itercomp} \textit{online}.
-%However, the overhead for estimating the amount of \textit{work} must be low enough for the performance gains to be beneficial.
 To this end, we model the computational work $\Delta W$ as a linear equation based on block frequency information and a cost model of the instruction set.
@@
 since $f(B)$ is both program and execution dependent as the execution frequency of a basic block can change when executing with different inputs.
-%However, $N_B(i)$ is program dependent but constant across executions, while $f(B)$ is both program and execution dependent, since $f(B)$ can change when executing with different inputs.
-%In other words, $N_B(i)$ is a static value known at compile-time and $f(B)$ is a dynamic value known only at run-time.
-
-%If we define $x_i$ as
-%\[
-%x_i = \sum_{B\in P} N_B(i)f(B)
-%\]
-%we can write our definition of work $\Delta W$ as the following linear equation
-%\[
-%\Delta W = \varepsilon + \sum_{i\in I} w_i x_i
-%\]
-
-%\subsection{Linear regression}
-%
-%\begin{figure}[htb]
-%    \centering
-%    \includegraphics[width=\linewidth]{figs/cost-model.pdf}
-%    \caption{Comparison between the naive and optimal instrumentation with no compiler optimisation.} %, i.e., compiled with \textbf{\texttt{-O0}}.}
-%    \label{fig:cost-model}
-%\end{figure}
-%mse 0.000145373168444
-%mae 0.00737261168981
-%mae* 0.0033892281406
-%mape 75.2924320971
-%corr (0.99946441743545622, 0.0)
 
@@
 
-We can illustrate this fact with a very small example as shown in Table~\ref{tab:ipc-example}.
-This example shows two versions of a program, namely P1 and P2, and their respective measurements related to IPC.
-Although version P1 has twice the IPC of P2, P1 is one cycle slower than P2.
-As this example illustrates, the IPC metric can be misleading when compared different versions of the same program.
-This problem is only intensified when both the optimisation sequence and the input changes.
-In Chapter~\ref{chap:eval} we show empirical evidence for this argument.
-
-\begin{table}[h]
-\centering
-\begin{tabular}{|c|c|c|c|}
-\hline
-                       & P1 & P2  \\
-\hline
-Number of Instructions & 5  & 2   \\
-Number of Cycles       & 5  & 4   \\
-IPC                    & 1  & 0.5 \\
-\hline
-\end{tabular}
-\caption{Example that illustrates that a higher IPC does not necessarily translate to shorter execution time.}
-\label{tab:ipc-example}
-\end{table}
+% We can illustrate this fact with a very small example as shown in Table~\ref{tab:ipc-example}.
+% This example shows two versions of a program, namely P1 and P2, and their respective measurements related to IPC.
+% Although version P1 has twice the IPC of P2, P1 is one cycle slower than P2.
+% As this example illustrates, the IPC metric can be misleading when compared different versions of the same program.
+% This problem is only intensified when both the optimisation sequence and the input changes.
+% In Chapter~\ref{chap:eval} we show empirical evidence for this argument.
+%
+% \begin{table}[h]
+% \centering
+% \begin{tabular}{|c|c|c|c|}
+% \hline
+%                        & P1 & P2  \\
+% \hline
+% Number of Instructions & 5  & 2   \\
+% Number of Cycles       & 5  & 4   \\
+% IPC                    & 1  & 0.5 \\
+% \hline
+% \end{tabular}
+% \caption{Example that illustrates that a higher IPC does not necessarily translate to shorter execution time.}
+% \label{tab:ipc-example}
+% \end{table}
 
@@
 If re-compilation time is critical, it would also be possible to keep only the hot portion of the code in the LLVM bitcode format, while the remaining portion of the code is already compiled to the final object code.
-However, this is out of the scope of this thesis and we always re-compile the whole program.
+However, this is out of the scope of this paper and we always re-compile the whole program.
 
@@
 
-%In both phases the goal is to improve the performance of the execution phase, but from different perspectives.
-%\textit{(i.)} In the compilation phase we focused on providing a low-overhead instrumentation for profiling the work metric.
-This phase is responsible to improve performance by lowering the overhead of the work profiling.
-%\textit{(ii.)} In the evaluation phase we proposed the use of the WP metric in order to enable {\itercomp} in an online scenario.
-%This phase is responsible to improve performance by being able to properly assess different optimisations such that the best optimisation can be selected.
-%Chapter~\ref{chap:instr} describes our work profiling strategy and Section~\ref{sec:metric} describes the work-based metric.
-
-%\textbf{Talk about the process of keeping the same optimization for a given input-window and how the optimizations as selected based on the evaluation of these "evidences".
-Perhaps we could suggest using the Theory of Evidences at this point?!}
-
 As the name suggests, the component called \textit{Optimisation Selector} is responsible for selecting which optimisation sequence to use for the next execution of the program.
@@
 \centering {
-  \includegraphics[scale=0.9]{figs/cfg-example.pdf}\\\vspace{1ex}
+  \includegraphics[scale=0.85]{figs/cfg-example.pdf}\\\vspace{1ex}
   \resizebox{0.45\textwidth}{!}{
@@
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[scale=0.9]{figs/relax-instr-diagram.pdf}
-  \caption{Overview of the work instrumentation algorithm, including the relaxation technique.}
-  \label{fig:relax-instr-diagram}
-\end{figure}
-
-%\begin{figure*}[htb]
+% \begin{figure}[h]
 %   \centering
-%    \includegraphics[width=\textwidth]{figs/overhead-O0.pdf}
-%    \caption{Comparison between the naive and optimal instrumentation
-%              with no compiler optimisation.}
-%    \label{fig:overhead-O0}
-%\end{figure*}
-
-%Define an DAG of a CFG as a directed graph which may contain cycles if they have known constant trip counts.
-Figure~\ref{fig:relax-instr-diagram} shows an overview of the relaxed instrumentation algorithm.
-The highlighted step is introduced by the relaxed instrumentation on top of the previously defined optimal profiling.
+%   \includegraphics[scale=0.85]{figs/relax-instr-diagram.pdf}
+%   \caption{Overview of the work instrumentation algorithm, including the relaxation technique.}
+%   \label{fig:relax-instr-diagram}
+% \end{figure}
+
+%Figure~\ref{fig:relax-instr-diagram} shows an overview of the relaxed instrumentation algorithm.
+%The highlighted step is introduced by the relaxed instrumentation on top of the previously defined optimal profiling.
 The relaxed instrumentation performs a post-processing on the resulting instrumentation of the optimal algorithm.
@@
 Figure~\ref{fig:cfg-relax-example} shows a CFG partitioned into two DAGs (consider only basic blocks and edges completely inside the yellow and green boundaries).
-%
-%LLVM has a Scalar Evolution Analysis which is used primarily to analyse expressions involving induction variables in loops~\cite{pop05}.
-%An induction variable is a variable that is increased or decreased by a fixed amount on every iteration of a loop or is a linear function of another induction variable.
-%
-%This Scalar Evolution Analysis provides a way to compute \textit{small} constant trip counts (where \textit{small} is defined by a threshold).
 
@@
   \centering
-  \includegraphics[scale=1]{figs/cfg-relax-example.pdf}
+  \includegraphics[scale=0.85]{figs/cfg-relax-example.pdf}
   \caption{Example of a CFG containing a loop and its decomposition into DAGs when applying the relaxation.
@@
 
-\newpage
-\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
-// Input: CFG	
-relaxInstrumentation(G) {
-  for loop in G:
-     DAG = colapseInnerLoops(loop)
-     relaxInstrumentedDAG(DAG)
-  DAG = colapseInnerLoops(G)
-  relaxInstrumentedDAG(DAG)
-}
-\end{lstlisting}
+%\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
+%// Input: CFG
+%relaxInstrumentation(G) {
+%  for loop in G:
+%     DAG = colapseInnerLoops(loop)
+%     relaxInstrumentedDAG(DAG)
+%  DAG = colapseInnerLoops(G)
+%  relaxInstrumentedDAG(DAG)
+%}
+%\end{lstlisting}
 
@@
 Because the percentage error is computed based on the path with the minimum amount of work, $\varepsilon(P_i)$ represents the maximum error possible that would be incurred when removing probe $P_i$.
-Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold, as demonstrated by Proposition~\ref{prop:relax-bound}.
-
-\begin{prop}\label{prop:relax-bound}
-Let $n_i$ be the number of times a given DAG $i$ is executed, $r_i$ be the total relaxation (amount of work removed) in DAG $i$, and $m_i$ be its minimum amount of work.
-If $\frac{r_i}{m_i} \leq M$ for every $i$,
-then the final error of the relaxation will always be bounded by the same threshold.
-\end{prop}
-\begin{proof}
-We can model the overall error of the relaxation as:
-\[
-1 - \frac{n_1(m_1 - r_1) + n_2(m_2 - r_2) + \ldots + n_k(m_k - r_k) + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
-\]
-That is,
-%\begin{equation*}
-\begin{gather*}
- 1 - \frac{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} + \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} = \\
- \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
-\end{gather*}
-%\end{equation*}
-If $\frac{r_j}{m_j}$ is the maximum ratio $\frac{r_i}{m_i}$ for every $i$, then
-\begin{equation*}
-\begin{aligned}
- \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} &\leq\\
- \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j + c} &\leq\\
- \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} & 
-\end{aligned}
-\end{equation*}
-If $N = max\{n_i$ for every $i\}$, then
-\begin{equation*}
-\begin{aligned}
- \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &\leq\\
- \frac{Nr_j + Nr_j + \ldots + Nr_j}{Nm_j + Nm_j + \ldots + Nm_j} &=\\
- \frac{Nkr_j}{Nkm_j} = \frac{r_j}{m_j} &\leq M
-\end{aligned}
-\end{equation*}
-\end{proof}
-
-\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
-// Input: CFG	
-relaxInstrumentedDAG(DAG){
-   P = ProbesIn(DAG)
-   m = minWork(DAG,P)
-   K = createKnapsackModel(P,m)
-   Bag = solveKnapsack(K)
-   for B in (P-Bag):
-     removeProbe(B)
-}
-\end{lstlisting}
+ Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold.
+% Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold, as demonstrated by Proposition~\ref{prop:relax-bound}.
+%
+% \begin{prop}\label{prop:relax-bound}
+% Let $n_i$ be the number of times a given DAG $i$ is executed, $r_i$ be the total relaxation (amount of work removed) in DAG $i$, and $m_i$ be its minimum amount of work.
+% If $\frac{r_i}{m_i} \leq M$ for every $i$,
+% then the final error of the relaxation will always be bounded by the same threshold.
+% \end{prop}
+% \begin{proof}
+% We can model the overall error of the relaxation as:
+% \[
+% 1 - \frac{n_1(m_1 - r_1) + n_2(m_2 - r_2) + \ldots + n_k(m_k - r_k) + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
+% \]
+% That is,
+% %\begin{equation*}
+% \begin{gather*}
+%  1 - \frac{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} + \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} = \\
+%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
+% \end{gather*}
+% %\end{equation*}
+% If $\frac{r_j}{m_j}$ is the maximum ratio $\frac{r_i}{m_i}$ for every $i$, then
+% \begin{equation*}
+% \begin{aligned}
+%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} &\leq\\
+%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j + c} &\leq\\
+%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &
+% \end{aligned}
+% \end{equation*}
+% If $N = max\{n_i$ for every $i\}$, then
+% \begin{equation*}
+% \begin{aligned}
+%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &\leq\\
+%  \frac{Nr_j + Nr_j + \ldots + Nr_j}{Nm_j + Nm_j + \ldots + Nm_j} &=\\
+%  \frac{Nkr_j}{Nkm_j} = \frac{r_j}{m_j} &\leq M
+% \end{aligned}
+% \end{equation*}
+% \end{proof}
+
+%\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
+%// Input: CFG
+%relaxInstrumentedDAG(DAG){
+%   P = ProbesIn(DAG)
+%   m = minWork(DAG,P)
+%   K = createKnapsackModel(P,m)
+%   Bag = solveKnapsack(K)
+%   for B in (P-Bag):
+%     removeProbe(B)
+%}
+%\end{lstlisting}
 
@@
 
-%In this section we discuss the three baseline implementations that will be used for assessing the quality of the proposed metric.
-
-%\vspace{1ex}
-%\noindent \textbf{{\Itercomp} over a single input}
-
-%Most of the existing {\itercomp} studies find the best optimisation though repeated runs on the same input.
-%Although this approach will usually lead to sub-optimal performance across large input datasets, it provides a good baseline when considering the complexity and resonably low compile-time regarding iterative optimising compilers.
-%If $M$ is the total number of combinations of compiler optimisations, this approach requires $O(M)$ runs of the program being optimised.
-%We can consider two main scenarios:
-%\textit{($i$ - best case scenario)} after selecting the optimisation over each individual input, consider the one with best performance across the whole input dataset;
-%\textit{($ii$ - expected scenario)} after selecting the optimisation over each individual input, consider the average case of their performance across the whole input dataset.
-
-%\vspace{1ex}
-%\noindent \textbf{{\Itercomp} across large input datasets}
-
-%Recent work on {\itercomp} have been targeting optimisation across multiple inputs~\cite{fursin07,chen10,chen12a}.
-%If $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimisations, they perform a total of $O(NM)$ runs of the program being optimised.
-%Similarly to what have been discussed previously, there are different ways for how to determine the optimal combination of compiler optimisations across multiple inputs.
-%It is possible to tune the selection of the program-optimal combination to minimise risk or to maximise average performance.
-%A compromise-based selection criterion could be to maximise average speedup with minimised variance.
-
-%\vspace{1ex}
-%\noindent \textbf{{\Itercomp} based on the IPC metric}
-
-%While the previous two baselines addresses two opposite aspects of {\itercomp}, namely, compile-time efficiency and performance of the generated code, we also intend to compare against IPC as the competing baseline.
-%The main reason for comparing against IPC is because it has been proposed as a metric for comparing the performance of two optimisations running on two distinct inputs.
-%If $M$ is the total number of combinations of compiler optimisations, this approach requires $O(M)$ runs of the program being optimised.
-
 \subsection{Evaluation of the Online {\IterComp}}
@@
   In order to reduce noise, the program is executed several times for the same input, until the confidence interval was no larger than 1\% for a 99\% confidence.
-\newpage
 \item \textbf{Oracle-PP} represents an oracle with a \textit{perfect} non-intrusive profiling.
@@
 
-\begin{figure}[htb]
+\begin{figure*}[htb]
     \centering
@@
     \label{fig:window-size}
-\end{figure}
+\end{figure*}
 
@@
 For the purpose of evaluating the use of the WP metric with {\itercomp}, we collected in advance a fixed set of optimisation sequences.
-The reason for using this fixed set, as explained in Section~\ref{sec:oic-infra}, is because this thesis is focused on other components of the infrastructure for performing {\itercomp} and we assume that a good generator of optimisation sequences will be used in a real online scenario.
+The reason for using this fixed set, as explained in Section~\ref{sec:oic-infra}, is because this paper is focused on other components of the infrastructure for performing {\itercomp} and we assume that a good generator of optimisation sequences will be used in a real online scenario.
 This set contains 500 optimisation sequences collected in a random search using the training benchmarks.
@@
 
-  \begin{minipage}{0.9\textwidth}
-     \vspace{1em}
-     \noindent\textbf{Example of a short optimisation sequence:}\vspace{-1ex}
-     \justify{\flagstype -mem2reg -simplifycfg -constprop -dce}
-  \end{minipage}
-
-  \begin{minipage}{0.9\textwidth}
-     \vspace{1em}
-     \noindent\textbf{Example of a long optimisation sequence:}\vspace{-1ex}
-     \justify{\flagstype -globalopt -reassociate -instcombine -loop-rotate -block-freq -deadargelim -early-cse -sroa -argpromotion -sccp -tbaa -barrier -constmerge \mbox{-loop-vectorize} -domtree -basicaa -memdep -basiccg -memcpyopt \mbox{-constprop} -adce -globaldce -mem2reg -constmerge \mbox{-globaldce} -constprop -instsimplify -dse -dce -simplifycfg -loop-unroll -reassociate -constprop \mbox{-globaldce} -instsimplify -adce -constmerge -bb-vectorize -dce -mergefunc -simplifycfg -dse -loop-unroll -globaldce}
-  \end{minipage}
-
-  \begin{minipage}{0.9\textwidth}
-     \vspace{1em}
-     \noindent\textbf{Example of an optimisation sequence which includes {-O3}:}\vspace{-1ex}
-     \justify{\flagstype -O3 -adce -globaldce -simplifycfg -memcpyopt -reassociate -mergefunc \mbox{-dce} -dse}
-     \vspace{2em}
-  \end{minipage}
+  % \begin{minipage}{0.9\textwidth}
+  %    \vspace{1em}
+  %    \noindent\textbf{Example of a short optimisation sequence:}\vspace{-1ex}
+  %    \justify{\flagstype -mem2reg -simplifycfg -constprop -dce}
+  % \end{minipage}
+  %
+  % \begin{minipage}{0.9\textwidth}
+  %    \vspace{1em}
+  %    \noindent\textbf{Example of a long optimisation sequence:}\vspace{-1ex}
+  %    \justify{\flagstype -globalopt -reassociate -instcombine -loop-rotate -block-freq -deadargelim -early-cse -sroa -argpromotion -sccp -tbaa -barrier -constmerge \mbox{-loop-vectorize} -domtree -basicaa -memdep -basiccg -memcpyopt \mbox{-constprop} -adce -globaldce -mem2reg -constmerge \mbox{-globaldce} -constprop -instsimplify -dse -dce -simplifycfg -loop-unroll -reassociate -constprop \mbox{-globaldce} -instsimplify -adce -constmerge -bb-vectorize -dce -mergefunc -simplifycfg -dse -loop-unroll -globaldce}
+  % \end{minipage}
+  %
+  % \begin{minipage}{0.9\textwidth}
+  %    \vspace{1em}
+  %    \noindent\textbf{Example of an optimisation sequence which includes {-O3}:}\vspace{-1ex}
+  %    \justify{\flagstype -O3 -adce -globaldce -simplifycfg -memcpyopt -reassociate -mergefunc \mbox{-dce} -dse}
+  %    \vspace{2em}
+  % \end{minipage}
 

### 2017-09-01 cgo18mscr.tex
updates paper

@@
 \begin{abstract}
-Efficiently selecting the best combination of compiler optimisations for a program
-remains an open problem due to their complex interaction and the large optimisation search space.
-{\Itercomp} techniques, which consist of iterating
-over this large space searching for the best optimisation,
-has the ability to adapt to new platforms, programs and workload while still
-having a systematic and simple optimisation process.
-Although some recent work have studied the impact of using multiple
-input datasets for performing \itercomp, applying this
-technique in an online scenario still poses some challenges.
-
-In this paper, our main goal is to enable \itercomp in
-online scenarios with the restriction of executing distinct inputs
-only once.
-In order to address this challenge,
-we propose a the use of a work-based metric that allows for comparing different
-combination of compiler optimisations even when executed with distinct inputs.
-Because of its online nature, we also propose a relaxed instrumentation
-that performs low overhead profiling of the amount of work executed, while
-being able to trade-off between accuracy and overhead in a controlled manner.
+%Efficiently selecting the best combination of compiler optimisations for a program
+%remains an open problem due to their complex interaction and the large optimisation search space.
+%{\Itercomp} techniques, which consist of iterating
+%over this large space searching for the best optimisation,
+%has the ability to adapt to new platforms, programs and workload while still
+%having a systematic and simple optimisation process.
+%Although some recent work have studied the impact of using multiple
+%input datasets for performing \itercomp, applying this
+%technique in an online scenario still poses some challenges.
+%
+%In this paper, our main goal is to enable \itercomp in
+%online scenarios with the restriction of executing distinct inputs
+%only once.
+%In order to address this challenge,
+%we propose a the use of a work-based metric that allows for comparing different
+%combination of compiler optimisations even when executed with distinct inputs.
+%Because of its online nature, we also propose a relaxed instrumentation
+%that performs low overhead profiling of the amount of work executed, while
+%being able to trade-off between accuracy and overhead in a controlled manner.
+In modern optimising compilers,  the correct choice of an optimisation sequence can have a significant impact on the performance of the final optimised program.
+%Each of these optimisation passes interacts with the code in complicated ways, depending on all other optimisations and the order they were applied to the code being optimised.
+Because of the unpredictability of optimisation interactions and their complexity,
+efficiently selecting the best optimisation sequence for a given program, or program section, remains an open problem.
+A well-known compilation technique that addresses this challenge is {\itercomp}.
+
+In this thesis, we focus on enabling {\itercomp} in \textit{online} scenarios.
+%We define the online scenario as having the restriction that programs execute multiple inputs and distinct inputs are executed only once.
+%Because of these restrictions, standard metrics for comparing optimisations, such as speedups or execution time alone, are not viable.
+We propose the use of a work-based performance metric for guiding {\itercomp},
+where we instrument the program for measuring the amount of work it performed during its execution.
+In order to reduce the profiling overhead, we also propose two relaxation algorithms which provide a trade-off between measurement accuracy and overhead.
+Our experimental evaluation shows that the two relaxation algorithms can reduce
+the average overhead by 43\% and $2.1\times$ over the optimal work profiling.
+%In our experimental evaluation, the first relaxation algorithm that operates on regions of functions is able to reduce the average overhead by about 43\% over the optimal work profiling, while incurring in a dynamic measurement error of much less than 1\%.
+%Similarly, the second relaxation algorithm which operates on the whole program, which also makes use of profiling from previous execution, was able to reduce even further this overhead by a factor of $2.1\times$ compared to the optimal profiling, while incurring in a dynamic measurement error of less than 3\%.
+
+Our evaluation also corroborates the use of the work-based performance metric for guiding online {\itercomp}.
+Under real conditions, it was able to achieve good performance improvements when compared to the oracles, which were allowed to execute the program with the same input multiple times,
+achieving an average improvement of 5.4\% and a maximum of about 20\%.
+Contrary to what previous work has suggested, we have also shown that instructions per cycle (IPC) is not a good metric for guiding online {\itercomp}.
 \end{abstract}
@@
 This is a key aspect that enables the use of the WP metric for guiding {\itercomp}, because a higher WP, which represents work per unit time, naturally translate to shorter execution time.
-Moreover, previous work has also presented other arguments against the use of IPC in similar use-case scenarios, as discussed in Chapter~\ref{chap:related}.
+Moreover, previous work has also presented other arguments against the use of IPC in similar use-case scenarios. %, as discussed in Chapter~\ref{chap:related}.
 
@@
 
-\begin{figure*}[htp]
-    \centering
-    \includegraphics[width=\textwidth]{figs/overhead-O3.pdf}
-    \caption{Overhead of the instrumentations compiled with {\texttt{-O3}}.}
-    \label{fig:overhead-O3}
-\end{figure*}
-
-Figure~\ref{fig:instr} shows percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.
+We first compare static aspects of the instrumentation algorithms.
 The naive instrumentation always has 100\% of the basic blocks instrumented, by definition.
-
-\begin{figure*}[htp]
+First, we evaluate the optimal work profiling compared to the relaxation algorithm applied per DAG.
+Afterwards, we compare these profiling techniques with the whole program relaxation.
+
+Figure~\ref{fig:num-probes} shows the percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.
+While Figure~\ref{fig:num-probes} compares the various instrumentation algorithms in respect of the naive instrumentation, Figure~\ref{fig:num-probes-improvement} shows the improvement of the relaxation algorithm  over the optimal instrumentation.
+
+\begin{figure*}[ht]
     \centering
@@
 
-The \texttt{adpcm\_d} benchmark is the most critical case amongst the used benchmarks, with a overhead of about 66\% with the optimal instrumentation.
-This benchmark consists mainly of a single hot loop with several branches inside it.
-The relaxation algorithm is able to reduce this overhead down to about 52\% (with an error threshold of 1\%) and 19\% (with an error threshold of 2\%) by removing only one and two probes from the hot loop, respectively.
-These two removed probes were placed in branches, inside the hot loop, with a high probability of being taken, but with a small contribution to the work measured in the loop.
+For the evaluation of the performance overhead that the instrumentation incur to the benchmarks, we measure the wall-clock time of the benchmarks when compiled with the default {\flagstype -O3} optimisation.
+For each benchmark, we compute the average overhead over all its 1000 input dataset.
+When measuring the wall-clock time for each input, in order to reduce noise, we execute the same input until we have a statistically sound measurement, i.e.
+we execute until we have an interval no larger than 1\% with 99\% confidence.
+Figure~\ref{fig:overhead-O3} shows the performance overhead imposed by the work instrumentation on the benchmarks when compared to their non-instrumented counterparts.
+Figure~\ref{fig:overhead-improvement-O3} indicates the improvement of the relaxation algorithm over the optimal instrumentation, regarding the reduction in overhead.
+
+\begin{figure*}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{figs/overhead-O3.pdf}
+    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.}
+    \label{fig:overhead-O3}
+\end{figure*}
+
+The WPO relaxation was performed using profiling information for every specific input, which provides the perfect information to perform the WPO relaxation.
+This means that the experiments show its best performance enabled by having perfect profiling information.
+
+%\begin{figure*}[ht]
+%    \centering
+%    \includegraphics[width=\textwidth]{figs/overhead-improvement-O3.pdf}
+%    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.
+%             Notice that the average of the ratios does not equal the ratio of the averages.}
+%    \label{fig:overhead-improvement-O3}
+%\end{figure*}
+
+\begin{figure*}[h!]
+    \centering
+    \includegraphics[width=\textwidth]{figs/error-O3.pdf}
+    \caption{Dynamic error of the work profiling averaged over the 1000 inputs, after relaxing the number of probes.}
+    \label{fig:error-O3}
+\end{figure*}
+
+%Figure~\ref{fig:overhead-improvement-O3} shows that the relaxation algorithm, with 5\% threshold, is able to improve an average of 48\% over the optimal instrumentation, 
+%and the WPO relaxation achieves an average of $2\times$ improvement over the optimal instrumentation.
+%While the optimal profiling has a maximum overhead of almost 60\%, the work profiling with a relaxation of 5\% threshold has an overhead of less than 20\% for all benchmarks,
+%with the WPO relaxation reaching an average overhead of only 5\%, compared to an average of 12\% for the optimal profiling.
+
+Furthermore, these improvements can be obtained while incurring very little dynamic error, as shown in Figure~\ref{fig:error-O3}.
+This result confirms the overly conservative aspect of the relaxation algorithm applied per DAG,
+while the WPO relaxation is significantly less conservative.
+Because we performed the WPO relaxation having perfect profiling information, its dynamic error was always below the 5\% thresholds.
+Although it is not guaranteed by the algorithm, profiling representative inputs should also result in small dynamic errors.
 

### 2017-09-01 cgo18mscr.tex
adds ToDo's

@@
 
-\begin{lstlisting}[caption={Post-processing of the CFG for populating all edge flows based on the collected probes.}, label={lst:populateEdgeFlows}, float]
+\begin{lstlisting}[caption={Post-processing of the CFG for populating all edge flows based on the collected probes.
+\textbf{ToDo: Replace the algorithm with a formal data-flow equation.}}, label={lst:populateEdgeFlows}, float]
 // Inputs: CFG with the known edges flows from
@@
 \begin{lstlisting}[caption={Pseudocode that describes how the edge information is used in order to extract the value that will be computed in a given instrumented basic block $B_I$.
-This algorithm could equally be implemented based on the predecessors.}, label={lst:instrValue}, float]
+This algorithm could equally be implemented based on the predecessors.
+\textbf{ToDo: Using the vector format, we could replace this algorithm with a single equation. Let $V$ be the vector of $\{-1,0,1\}$ and $W$ the weight for each basic block.
+This algorithm could be re-written as $reduce (+) (map (*) (zip V W))$.}
+}, label={lst:instrValue}, float]
 // Inputs: 1) CFG with the known edges flows
@@
 
+\textbf{ToDo: Include results using Genetic Algorithm search.}
+
 \section{Related Work}

### 2017-09-06 cgo18mscr.tex
update abstract

@@
 \PassOptionsToPackage{table}{xcolor}
+
 \documentclass[sigplan,10pt]{acmart}
@@
 %Conference
-\acmConference[CGO'18]{International Symposium on
-Code Generation and Optimization}{2018}{}
+\acmConference[CGO'18]{International Symposium on Code Generation and Optimization}{2018}{}
 %\acmYear{1997}
@@
 \begin{abstract}
-%Efficiently selecting the best combination of compiler optimisations for a program
-%remains an open problem due to their complex interaction and the large optimisation search space.
-%{\Itercomp} techniques, which consist of iterating
-%over this large space searching for the best optimisation,
-%has the ability to adapt to new platforms, programs and workload while still
-%having a systematic and simple optimisation process.
-%Although some recent work have studied the impact of using multiple
-%input datasets for performing \itercomp, applying this
-%technique in an online scenario still poses some challenges.
-%
-%In this paper, our main goal is to enable \itercomp in
-%online scenarios with the restriction of executing distinct inputs
-%only once.
-%In order to address this challenge,
-%we propose a the use of a work-based metric that allows for comparing different
-%combination of compiler optimisations even when executed with distinct inputs.
-%Because of its online nature, we also propose a relaxed instrumentation
-%that performs low overhead profiling of the amount of work executed, while
-%being able to trade-off between accuracy and overhead in a controlled manner.
-In modern optimising compilers,  the correct choice of an optimisation sequence can have a significant impact on the performance of the final optimised program.
-%Each of these optimisation passes interacts with the code in complicated ways, depending on all other optimisations and the order they were applied to the code being optimised.
-Because of the unpredictability of optimisation interactions and their complexity,
-efficiently selecting the best optimisation sequence for a given program, or program section, remains an open problem.
-A well-known compilation technique that addresses this challenge is {\itercomp}.
+Until recently, most of the existing work in {\itercomp} had been focusing on
+finding the best optimisation through repeated runs using a single input.
+However, in real-world online scenarios, the user rarely executes a program with
+the same input multiple times and most of the programs are complex enough that a
+single input case does not capture the whole range of possible scenarios and
+program behaviours.
 
 In this thesis, we focus on enabling {\itercomp} in \textit{online} scenarios.
-%We define the online scenario as having the restriction that programs execute multiple inputs and distinct inputs are executed only once.
-%Because of these restrictions, standard metrics for comparing optimisations, such as speedups or execution time alone, are not viable.
-We propose the use of a work-based performance metric for guiding {\itercomp},
-where we instrument the program for measuring the amount of work it performed during its execution.
-In order to reduce the profiling overhead, we also propose two relaxation algorithms which provide a trade-off between measurement accuracy and overhead.
-Our experimental evaluation shows that the two relaxation algorithms can reduce
-the average overhead by 43\% and $2.1\times$ over the optimal work profiling.
-%In our experimental evaluation, the first relaxation algorithm that operates on regions of functions is able to reduce the average overhead by about 43\% over the optimal work profiling, while incurring in a dynamic measurement error of much less than 1\%.
-%Similarly, the second relaxation algorithm which operates on the whole program, which also makes use of profiling from previous execution, was able to reduce even further this overhead by a factor of $2.1\times$ compared to the optimal profiling, while incurring in a dynamic measurement error of less than 3\%.
-
-Our evaluation also corroborates the use of the work-based performance metric for guiding online {\itercomp}.
-Under real conditions, it was able to achieve good performance improvements when compared to the oracles, which were allowed to execute the program with the same input multiple times,
-achieving an average improvement of 5.4\% and a maximum of about 20\%.
-Contrary to what previous work has suggested, we have also shown that instructions per cycle (IPC) is not a good metric for guiding online {\itercomp}.
+Because of the restriction on repeating the execution with same input, it is
+not always feasible to measure the real speedup when comparing optimisations.
+To this end, we propose a \textit{work efficiency} metric to guide 
+online {\itercomp}, which is able to achieve about 80\% of the performance 
+of the offline oracle, with improvements of up to about 20\%.
+
+Moreover, because we need to instrument the program for measuring work
+efficiency, providing a low-overhead profiling is essential.
+In order to reduce the profiling overhead, we also propose two relaxation
+algorithms which provide a trade-off between measurement accuracy and overhead.
+Our results show that the two relaxation algorithms can reduce the average
+overhead by 43\% and $2.1\times$ over the optimal work profiling, observing
+up to about $5\times$ of overhead reduction.
 \end{abstract}
@@
 %
-\begin{CCSXML}
-<ccs2012>
-<concept>
-<concept_id>10011007.10011006.10011041</concept_id>
-<concept_desc>Software and its engineering~Compilers</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-<concept>
-<concept_id>10011007.10011006.10011041.10011043</concept_id>
-<concept_desc>Software and its engineering~Retargetable compilers</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-</ccs2012>
-\end{CCSXML}
+%\begin{CCSXML}
+%<ccs2012>
+%<concept>
+%<concept_id>10011007.10011006.10011041</concept_id>
+%<concept_desc>Software and its engineering~Compilers</concept_desc>
+%<concept_significance>500</concept_significance>
+%</concept>
+%<concept>
+%<concept_id>10011007.10011006.10011041.10011043</concept_id>
+%<concept_desc>Software and its engineering~Retargetable compilers</concept_desc>
+%<concept_significance>500</concept_significance>
+%</concept>
+%</ccs2012>
+%\end{CCSXML}
 
@@
 \section{Introduction}
-
-Modern optimising compilers have reached a high level of sophistication, providing a large number of optimisations.
-Because of the unpredictability of optimisation interactions, in addition to the growing complexity of processor architectures and applications, the correct choice of optimisations and their ordering can have a significant impact on the performance of the code being optimised~\cite{pan06,fursin07,kulkarni12}.
-
-Each of these optimisations interacts with the code in complicated ways, depending on all other optimisations and the order they are applied to the code being optimised.
-However, understanding the interactions of optimisations is very important in determining a good solution to the phase-ordering problem~\cite{touati06,kulkarni12}.
-Because of all those dependencies and interactions, although most compiler optimisations yield significant improvements in many programs, the potential for performance degradation in certain program patterns is known to compiler writers and many users~\cite{pan06,zhou12,kulkarni12}.
-%Traditional compilers typically apply the same set of optimisation in a fixed order to all functions in a program, without regard the code being optimised.
-
-Efficiently selecting the best optimisation sequence for a given program, or program section, remains an open problem.
-Compiler writers typically use a combination of experience and insight to construct the default sequences of prearranged optimisations found in compilers.
-However, these pre-arranged optimisation sequences do not include all possible optimisations and they are always applied in the same pre-defined order, without regard the code being optimised~\cite{pan06,cavazos07,zhou12,kulkarni12}.
-%in hopes that a programmer will know which combination of optimisations will benefit their code~\cite{pan06,cavazos07,zhou12,kulkarni12}.
-
-A well known compilation technique that addresses this challenge is {\itercomp}.
-{\Itercomp} has the ability to adapt to new platforms, program and workload while still having a systematic and simple optimisation process.
-It works by repeatedly evaluating a large number of optimisation sequences until the best combination is found for a particular program~\cite{fursin07,chen10}.
-The main challenge concerning {\itercomp} is the need for efficiently exploring such a large optimisation space~\cite{fursin07,cavazos07,zhou12}.
-
-Until recently, most of existing work  had been focusing on finding the best optimisation through repeated runs using a single input.
-Although they demonstrate the potential of {\itercomp}, in real scenarios the user rarely executes a program with the same input multiple times~\cite{bodin98,kisuki99,stephenson03,kulkarni04,agakov06}.
-Furthermore, most of real world applications are complex enough so that a single input case does not capture the whole range of possible scenarios and program behaviour~\cite{haneda06,fursin07,chen10,chen12a}.
-Because programs can exhibit behaviours that differ greatly depending on the input,
-%applying {\itercomp} in light of a single input may not result in a good performance when executing the optimised code with different inputs.
-using a single input for {\itercomp} may not result in a good performance when executing the optimised code with different inputs.
-
-%This is something that is well known when writing test cases for correctness.
-%We can extend the same argument when optimising the code for performance.
-%using a single input for {\itercomp} can produce poor performance when executed with different inputs.
-
-Recent work have been studying the impact of using multiple inputs for
-performing {\itercomp}.
-%This previous work suggests that optimising based on a representative range of
-%inputs allows for selecting a robust optimisation sequence that produces
-%good performance in real scenarios where the input is expected to change
-%constantly~\cite{haneda06,fursin07,chen10,chen12a,chen12b,fang15,mpeis16}.
-Their results suggest that optimising based on a relatively small set of
-representative inputs delivers good performance across a wide range of different
-inputs~\cite{haneda06,fursin07,chen10,chen12a,chen12b,fang15,mpeis16}.
-%Their results show that a limited number of inputs may be sufficient to obtain
-%a well-optimised program for a wide range of different
-%inputs~\cite{haneda06,fursin07,chen10,chen12a}.
-
-In this work, our main goal is to enable {\itercomp} in \textit{online} scenarios.
-For the purposes of this paper, we define the online scenario as having the restriction that programs execute multiple inputs and distinct inputs are executed only once.
-This online aspect is usually found in mobile and data centre platforms~\citep{chen12b,fang15,mpeis16}, where the goal is to optimise programs based on the workload of a particular user (or group of users) between executions.
-Online {\itercomp} must target optimal performance across different inputs.
+Modern optimising compilers have reached a high level of sophistication,
+providing a large number of optimisations, where the correct choice of
+optimisations and their ordering can have a significant impact on the
+performance of the code being optimised.
+Although compilers offer a set of prearranged optimisation sequences that are
+expected to yield reasonable improvements in many programs, there is still the
+potential for performance degradation for certain programs, as these
+optimisation sequences do not include all possible optimisations and are always
+applied in the same pre-defined order, without regard the code being
+optimised~\cite{pan06,cavazos07,zhou12,kulkarni12}.
+A well known compilation technique that addresses this challenge is {\itercomp},
+which has the ability to adapt to new platforms, program and workload while
+still having a systematic and simple optimisation process.
+It works by repeatedly evaluating a large number of optimisation sequences until
+the best combination is found for a particular
+program~\cite{kisuki99,fursin07,chen10}.
+
+%{\Itercomp} is a well-known compilation technique that addresses the problem of
+%efficiently selecting the best optimisation sequence for a given program.
+%Although compilers offer a set of prearranged optimisation sequences that are
+%expected to yield reasonable improvements in many programs, there is still the
+%potential for performance degradation for certain programs, as these
+%optimisation sequences do not include all possible optimisations and are always
+%applied in the same pre-defined order, without regard the code being
+%optimised~\cite{pan06,cavazos07,zhou12,kulkarni12}.
+%{\Itercomp} have a systematic and simple optimisation process, where it works
+%by repeatedly evaluating a large number of optimisation sequences until the best
+%combination is found for a particular program~\cite{kisuki99,fursin07,chen10}.
+
+Until recently, most of the existing work in {\itercomp} had been focusing on
+finding the best optimisation through repeated runs using a single input.
+Although they demonstrate the potential of {\itercomp}, in real scenarios the
+user rarely executes a program with the same input multiple
+times~\cite{bodin98,kisuki99,stephenson03,kulkarni04,agakov06}.
+Furthermore, most of real world applications are complex enough so that a single
+input case does not capture the whole range of possible scenarios and program
+behaviours~\cite{haneda06,fursin07,chen10,chen12a}.
+Because programs can exhibit behaviours that differ greatly depending on the
+input, using a single input for {\itercomp} may not result in a good performance
+when executing the optimised code with different inputs.
+
+The main goal of this paper is to enable {\itercomp} in \textit{online} scenarios.
+We define the online scenario as having the restriction that programs execute multiple inputs and distinct inputs are executed only once.
+This online aspect is usually found in mobile and data centre platforms~\citep{chen12b,fang15,mpeis16}, where the goal is to optimise programs based on the workload of a particular user, or group of users.
+Online {\itercomp} must also target optimal performance across different inputs.
 
@@
 For this metric, the main challenge is to precisely define what represents \textit{work}.
-As discussed in Section~\ref{sec:related-metrics}, many work metrics have been presented in the literature, such as block frequency and others.
+As discussed in Section~\ref{sec:relatedwork}, many work metrics have been presented in the literature, such as block frequency and others.
 However, for comparing the performance benefits of different compiler optimisations on a program, although block frequency would be able to capture aspects of optimisations that simplify the control-flow graph (CFG), we argue that measuring work only at the basic block resolution would not be enough for capturing the effects of optimisations at the instruction level.
@@
 
-\section{Related Work}
+\section{Related Work}\label{sec:relatedwork}
 

### 2017-09-06 abstract.tex
moving staff

--- 2017-09-06 cgo18mscr.tex
@@
-\PassOptionsToPackage{table}{xcolor}
-
-\documentclass[sigplan,10pt]{acmart}
-
-\usepackage{booktabs} % For formal tables
-
-\usepackage[utf8]{inputenc}
-\usepackage{fullpage}
-
-\usepackage{graphics}
-
-\usepackage{graphicx}
-\usepackage{subcaption}
-
-\usepackage{multirow}
-
-%\usepackage{setspace}
-\usepackage{ragged2e}
-
-	\usepackage{soul}
-
-%\usepackage{flushend}
-
-%\usepackage{indentfirst}
-
-\usepackage{amsthm}
-\usepackage{amsmath,amssymb}
-%\usepackage{mathrsfs}
-\usepackage{mathtools}
-
-\theoremstyle{definition}
-\newtheorem{prop}{Proposition}[section]
-
-%% Listing related stuff
-
-\usepackage{listings}
-\usepackage{courier}            %Required for pretty listings (more dense)
-
-%% EMACS colors for listing:
-
-\usepackage{color}
-\definecolor{sh_comment}{rgb}{0.65, 0.00, 0.00 } % red-ish
-\definecolor{sh_keyword}{rgb}{0.37, 0.69, 0.69}  % blue-green
-\definecolor{sh_string}{rgb}{0.08, 0.69, 0.08} % green
-
-%% This manipulates the listing numbering. It decreases it by one at the point used.
-\newcommand*\lstDNumber{\addtocounter{lstnumber}{-1}}
-
-%% Change from Listing->Algorithm
-%% \renewcommand\lstlistingname{Algorithm} % Fix listing caption name: Listing -> Algorithm
-%% \renewcommand\lstlistlistingname{Algorithms}
-%% \def\lstlistingautorefname{Alg.}
-
-%% \renewcommand{\thelstlisting}{\thesection-\arabic{lstlisting}}
-%\renewcommand{\thelstlisting}{\arabic{lstlisting}} % Fix listing numbering Algorithm 1.1 -> Algorithm 1.
-
-%% \renewcommand{\labelenumi}{\roman{enumi}.}
-
-\lstset{
-float=[*],
-language=C,                % choose the language of the code
-%% basicstyle=\linespread{0.9}
-%% basicstyle=\scriptsize\ttfamily,
- basicstyle={\linespread{0.97} \scriptsize\ttfamily},
- stringstyle=\color{sh_string},
- keywordstyle = \color{sh_keyword}\bfseries,
- commentstyle=\color{sh_comment}\itshape,
-numbers=left,                   % where to put the line-numbers
-%% numberstyle=\scriptsize,        % the size of the fonts that are used for the line-numbers
-numberstyle=\scriptsize,        % the size of the fonts that are used for the line-numbers
-stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
-numbersep=5pt,                  % how far the line-numbers are from the code
-backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
-showspaces=false,               % show spaces adding particular underscores
-showstringspaces=false,         % underline spaces within strings
-showtabs=false,                 % show tabs within strings adding particular underscores
-xleftmargin=2em,                % Left margin
-frame=lines,                   % adds a frame around the code. none|leftline|topline|bottomline|lines|single|shadowbox
-framexleftmargin=1.5em,         % Margin from frame to line numbers
-framexbottommargin=0em,         % Distance from last text to frame
-morekeywords={in,true,false,and,or,set,is,not},
-%% prebreak=\mbox{\tiny$\searrow$},
-prebreak=\space,                % Line break: insert this character at the end of the top line
-%% postbreak=\mbox{{\color{blue}\scriptsize$\hookrightarrow$}}, % Line break: blue arrow at the beginning of the bottom line
-postbreak=\mbox{{\color{blue}\scriptsize$\hookrightarrow$}}, % Line break: blue arrow at the beginning of the bottom line
-breaklines=true,                % sets automatic line breaking
-breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
-tabsize=2,                      % sets default tabsize to 2 spaces
-captionpos=b,                   % sets the caption-position to top
-escapeinside={\@}{\@}             % Escape sequence \@ESCAPED STUFF\@
-}
-
-\usepackage{nasm/lang}
-\usepackage{nasm/style}
-\usepackage{llvm/lang}
-
-% Copyright
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\setcopyright{usgov}
-%\setcopyright{usgovmixed}
-%\setcopyright{cagov}
-%\setcopyright{cagovmixed}
-
-% DOI
-%\acmDOI{10.475/123_4}
-
-% ISBN
-%\acmISBN{123-4567-24-567/08/06}
-
-%Conference
-\acmConference[CGO'18]{International Symposium on Code Generation and Optimization}{2018}{}
-%\acmYear{1997}
-%\copyrightyear{2016}
-%
-%\acmPrice{15.00}
-
-\newcommand{\etal}{et~al.}
-
-\newcommand{\itercomp}{{iterative compilation}}
-\newcommand{\Itercomp}{{Iterative compilation}}
-\newcommand{\IterComp}{{Iterative Compilation}}
-
-\newcommand{\flagstype}{\usefont{T1}{cmr}{m}{n}}
-
-\begin{document}
-\title{Online {\IterComp}\\Guided by Low Overhead Profiling}
-%\titlenote{...}
-%\subtitle{...}
-%\subtitlenote{...}
-
-\author{ \vspace{1em} }
-\affiliation{%
-  \institution{ }
-}
-\email{ }
-
-%\author{Rodrigo C. O. Rocha}
-%\affiliation{%
-%  \institution{University of Edinburgh, UK}
-%}
-%\email{r.rocha@ed.ac.uk}
-
-%\author{Pavlos Petoumenos}
-%\affiliation{%
-%  \institution{University of Edinburgh, UK}
-%}
-%\email{ppetoume@inf.ed.ac.uk}
-
-%\author{Lu\'is F. W. G\'oes}
-%\affiliation{%
-%  \institution{PUC Minas, Brazil}
-%}
-%\email{lfwgoes@pucminas.br}
-
-%\author{Murray Cole}
-%\affiliation{
-%  \institution{University of Edinburgh, UK}
-%}
-%\email{mic@inf.ed.ac.uk}
-
-%\author{Zheng Wang}
-%\affiliation{%
-%  \institution{Lancaster University, UK}
-%}
-%\email{z.wang@lancaster.ac.uk}
-
-%\author{Hugh Leather}
-%\affiliation{%
-%  \institution{University of Edinburgh, UK}
-%}
-%\email{hleather@inf.ed.ac.uk}
-
-\renewcommand{\shortauthors}{R. Rocha et al.}
-
 \begin{abstract}
@@
 \end{abstract}
-
-%
-% The code below should be generated by the tool at
-% http://dl.acm.org/ccs.cfm
-% Please copy and paste the code instead of the example below.
-%
-%\begin{CCSXML}
-%<ccs2012>
-%<concept>
-%<concept_id>10011007.10011006.10011041</concept_id>
-%<concept_desc>Software and its engineering~Compilers</concept_desc>
-%<concept_significance>500</concept_significance>
-%</concept>
-%<concept>
-%<concept_id>10011007.10011006.10011041.10011043</concept_id>
-%<concept_desc>Software and its engineering~Retargetable compilers</concept_desc>
-%<concept_significance>500</concept_significance>
-%</concept>
-%</ccs2012>
-%\end{CCSXML}
-
-%\ccsdesc[500]{Software and its engineering~Compilers}
-%\ccsdesc[500]{Software and its engineering~Retargetable compilers}
-
-\keywords{{\IterComp}, Profiling, Relaxed Instrumentation}
-
-\maketitle
-
-\section{Introduction}
-Modern optimising compilers have reached a high level of sophistication,
-providing a large number of optimisations, where the correct choice of
-optimisations and their ordering can have a significant impact on the
-performance of the code being optimised.
-Although compilers offer a set of prearranged optimisation sequences that are
-expected to yield reasonable improvements in many programs, there is still the
-potential for performance degradation for certain programs, as these
-optimisation sequences do not include all possible optimisations and are always
-applied in the same pre-defined order, without regard the code being
-optimised~\cite{pan06,cavazos07,zhou12,kulkarni12}.
-A well known compilation technique that addresses this challenge is {\itercomp},
-which has the ability to adapt to new platforms, program and workload while
-still having a systematic and simple optimisation process.
-It works by repeatedly evaluating a large number of optimisation sequences until
-the best combination is found for a particular
-program~\cite{kisuki99,fursin07,chen10}.
-
-%{\Itercomp} is a well-known compilation technique that addresses the problem of
-%efficiently selecting the best optimisation sequence for a given program.
-%Although compilers offer a set of prearranged optimisation sequences that are
-%expected to yield reasonable improvements in many programs, there is still the
-%potential for performance degradation for certain programs, as these
-%optimisation sequences do not include all possible optimisations and are always
-%applied in the same pre-defined order, without regard the code being
-%optimised~\cite{pan06,cavazos07,zhou12,kulkarni12}.
-%{\Itercomp} have a systematic and simple optimisation process, where it works
-%by repeatedly evaluating a large number of optimisation sequences until the best
-%combination is found for a particular program~\cite{kisuki99,fursin07,chen10}.
-
-Until recently, most of the existing work in {\itercomp} had been focusing on
-finding the best optimisation through repeated runs using a single input.
-Although they demonstrate the potential of {\itercomp}, in real scenarios the
-user rarely executes a program with the same input multiple
-times~\cite{bodin98,kisuki99,stephenson03,kulkarni04,agakov06}.
-Furthermore, most of real world applications are complex enough so that a single
-input case does not capture the whole range of possible scenarios and program
-behaviours~\cite{haneda06,fursin07,chen10,chen12a}.
-Because programs can exhibit behaviours that differ greatly depending on the
-input, using a single input for {\itercomp} may not result in a good performance
-when executing the optimised code with different inputs.
-
-The main goal of this paper is to enable {\itercomp} in \textit{online} scenarios.
-We define the online scenario as having the restriction that programs execute multiple inputs and distinct inputs are executed only once.
-This online aspect is usually found in mobile and data centre platforms~\citep{chen12b,fang15,mpeis16}, where the goal is to optimise programs based on the workload of a particular user, or group of users.
-Online {\itercomp} must also target optimal performance across different inputs.
-
-Because of the restriction of having a single execution per input, it is not possible to measure speedup for comparing optimisations.
-On the other hand, measuring just execution time, for example, is useful only if the amount of work is constant between executions with different inputs.
-While previous work have suggested using \textit{instructions per cycle} (IPC) for performing {\itercomp} in \textit{online} scenarios, IPC have no correlation with speedup~\citep{fursin07}.
-
-In order to address this challenge, we propose the use of a work-based metric to compare the performance of different optimisations across multiple executions of the program with distinct inputs.
-We instrument the program for measuring the amount of work it performed during its execution.
-Having a low overhead instrumentation is essential in this online scenario for two main reasons:
-$(i)$ the user is directly affected by large overheads;
-$(ii)$ a highly intrusive instrumentation can have a significant impact on the effect of the optimisations.
-With the purpose of reducing profiling overhead, we propose two relaxation algorithms which provide a trade-off between measurement accuracy and overhead.
-The first is a relaxation algorithm that operates on the level of regions of functions, while the second performs the relaxation considering the whole program at the same time.
-
-Our experimental evaluation shows that performing online {\itercomp} guided by the work-based performance (WP) metric good results compared to the oracle, which is allowed to execute each input multiple times in order to use the actual speedup for guiding the {\itercomp}.
-Online {\itercomp} guided by the WP metric is able to achieve an average of 7.5\% and a maximum of 33\% improvement over the standard {\flagstype -O3} optimisation.
-Moreover, the experiments regarding the work profiling show that both relaxation algorithms are able to significantly reduce the profiling overhead while incurring a dynamic error of less than 5\% in the work measurement.
-The whole program relaxation achieves an average of $2\times$ reduction in the overhead compared with the optimal profiling technique, while the more conservative relaxation that operates per region achieves an average improvement of 40\% over the optimal profiling.
-
-%Our main contributions are the following:
-%\begin{itemize}
-%\item The use of a work-based performance metric in order to enable \textit{online} {\itercomp} by comparing different combination of compiler optimisations even when executed with distinct inputs.
-%\item We propose a relaxed instrumentation for low overhead profiling, with a controlled trade-off between accuracy and overhead.
-%\end{itemize}
-
-To summarise, the main contributions of this paper are the following:
-\begin{itemize}
-\item We propose the use of a work-based performance (WP) metric for comparing different optimised versions of a program.
-      In particular, we use the WP metric to guide {\itercomp} in an online scenario, where the program is expected to execute only once for distinct inputs.
-
-\item Contrary to what previous work has suggested, we show that instructions per cycle (IPC) is not a good metric for online {\itercomp}.
-\item We adapt the block frequency profiling in order to measure the WP metric.
-\item A relaxation algorithm is proposed for lowering the overhead of the work profiling, with a controlled trade-off between accuracy and overhead.
-\item A whole program relaxation is proposed in order to further reduce the overhead of the work profiling.
-\end{itemize}
-
-\section{Work-based Performance Metric} \label{sec:metric}
-
-In this section we define the work-based performance (WP) metric proposed for comparing different optimised versions of a program when executing with different inputs.
-We define the performance metric as the ratio between the amount of \textit{work}, $\Delta W$, performed during a period of time, $\Delta t$.
-\[
-   P = \frac{\Delta W}{\Delta t}
-\]
-
-By measuring the amount of \textit{work} done per unit of time, we reduce the impact of input-dependent aspects and focus instead on the efficiency of the optimised program.
-For this metric, the main challenge is to precisely define what represents \textit{work}.
-As discussed in Section~\ref{sec:relatedwork}, many work metrics have been presented in the literature, such as block frequency and others.
-However, for comparing the performance benefits of different compiler optimisations on a program, although block frequency would be able to capture aspects of optimisations that simplify the control-flow graph (CFG), we argue that measuring work only at the basic block resolution would not be enough for capturing the effects of optimisations at the instruction level.
-For this reason, we extend the idea of using basic block frequency to measure computational work by also considering the computational cost of each basic block.
-
-To this end, we model the computational work $\Delta W$ as a linear equation based on block frequency information and a cost model of the instruction set.
-Formally,
-\[
-\Delta W = \varepsilon + \sum_{B} w(B)f(B)
-\]
-where $f(B)$ represents the frequency of basic block $B$ and $w(B)$ represents the computational work of executing $B$.
-We define the work of a basic block $B$ as the sum of the cost of its instructions, i.e.,
-\[
-w(B) = \sum_{i} w_i N_B(i)
-\]
-where $w_i$ is the cost of instruction $i$ and $N_B(i)$ is the number of occurrences of instruction $i$ in basic block $B$.
-
-In this simplified model, we consider that $w_i$ is constant across all programs and executions, varying only between target architectures.
-On the other hand, $N_B(i)$ is a program-dependent static value which is known at compile-time and $f(B)$ is a dynamic value known only at run-time,
-since $f(B)$ is both program and execution dependent as the execution frequency of a basic block can change when executing with different inputs.
-
-\subsection{Estimating a Cost Model of the Instructions}
-
-Similarly to previous work~\citep{giusto01,powell09,brandolese11}, we derive the cost model of the instruction set by modelling the problem as a multi-variable linear regression, where the \textit{regression coefficients} are the costs of the instructions and the \textit{regressors} are computed as $\sum_B N_B(i)f(B)$ for each instruction.
-\begin{equation}\label{eq:linear-work-expression}
-\Delta W = \varepsilon + \sum_{i} \left(w_i \sum_{B} N_B(i)f(B)\right)
-\end{equation}
-By having some empirical data after executing several benchmarks with different inputs, we can fit the linear model with this empirical data in order to obtain estimate costs of the instructions.
-In order to fit the linear model, we measure the wall-clock time when executing the training benchmarks described in Section\ref{sec:benchmarks} with their respective 1000 input datasets.
-For these measurements, the training benchmarks are compiled without optimisation.
-The reason for using no optimisation, as discussed in Section\ref{sec:oic-infra}, is because the amount of work must be only input dependent and consistent between different optimisation sequences.
-This aspect is crucial for the work-based performance metric to enable a consistent comparison between the various optimisation sequences when executing with distinct inputs.
-%it allows to use the estimated cost model for computing a work metric which is independent of optimisations, as we explained in Section~\ref{}.
-
-Our cost model has a total of 52 LLVM instructions\footnote{We do not model all LLVM instructions because some instructions are more common in optimised programs, such as the vector instructions.}.
-Every program, in the set of training benchmarks, was compiled twice: once including the necessary instrumentation for the block frequency profiling, which is required for deriving the linear expression defined in Equation~\ref{eq:linear-work-expression};
-and a standard compilation without any optimisation or instrumentation.
-For each input in the training dataset, the benchmarks were executed once with the instrumented version for collecting the block frequency profiling, and multiple times with the standard compilation just for measuring the wall-clock execution time, until the confidence interval was no larger than 1\% for a 99\% confidence.
-After collecting these measurements, this data can be used to estimate the unknown parameters in the linear model.
-Because we fit the linear model based on the wall-clock execution time, the derived cost model can be interpreted as an estimate of the execution time when the program is compiled without optimisations.
-
-\begin{figure}[htb]
-    \centering
-    \includegraphics[width=0.9\linewidth]{figs/cost-model.pdf}
-    \caption{Linear model fitted from empirical data. The mean absolute error (MAE) for the fitted curve is seven milliseconds.}
-    \label{fig:cost-model}
-\end{figure}
-
-Figure~\ref{fig:cost-model} compare the work metric with the corresponding execution time for some instances of the test benchmarks.
-Notice how the fitted model has a higher relative error for the instances with very short execution time, namely those that run for less than one tenth of a second.
-The mean absolute error (MAE) for the fitted curve is seven milliseconds.
-
-\subsection{Comparison with Instructions Per Cycle} \label{sec:ipc-vs-work-metric}
-
-The IPC metric has been widely used for studying performance benefits of hardware optimisations.
-Although previous work has suggested the use of IPC for guiding {\itercomp}, in this section we argue in favour of the WP metric over IPC.
-
-The IPC metric differs from the proposed WP metric in a key aspect:
-the IPC metric is computed solely based on the final optimised program.
-When executing different optimised versions of a program with the same input, both the number of instructions and the number of clock cycles can change.
-For this reason, higher IPC does not necessarily translate to shorter execution time (or even fewer clock cycles).
-
-% We can illustrate this fact with a very small example as shown in Table~\ref{tab:ipc-example}.
-% This example shows two versions of a program, namely P1 and P2, and their respective measurements related to IPC.
-% Although version P1 has twice the IPC of P2, P1 is one cycle slower than P2.
-% As this example illustrates, the IPC metric can be misleading when compared different versions of the same program.
-% This problem is only intensified when both the optimisation sequence and the input changes.
-% In Chapter~\ref{chap:eval} we show empirical evidence for this argument.
-%
-% \begin{table}[h]
-% \centering
-% \begin{tabular}{|c|c|c|c|}
-% \hline
-%                        & P1 & P2  \\
-% \hline
-% Number of Instructions & 5  & 2   \\
-% Number of Cycles       & 5  & 4   \\
-% IPC                    & 1  & 0.5 \\
-% \hline
-% \end{tabular}
-% \caption{Example that illustrates that a higher IPC does not necessarily translate to shorter execution time.}
-% \label{tab:ipc-example}
-% \end{table}
-
-On the other hand, WP computes the amount of work based on the unoptimised version of the program, which means that it always measures the same amount of work for the same input, regardless of the optimisations applied on the program.
-This is a key aspect that enables the use of the WP metric for guiding {\itercomp}, because a higher WP, which represents work per unit time, naturally translate to shorter execution time.
-Moreover, previous work has also presented other arguments against the use of IPC in similar use-case scenarios. %, as discussed in Chapter~\ref{chap:related}.
-
-\section{Online {\IterComp} Infrastructure} \label{sec:oic-infra}
-
-Although {\itercomp} had been originally proposed as an \textit{offline} optimisation strategy, it can also be adapted to work in online scenarios.
-Instead of selecting the best optimisation sequence as part of the development time (pre-shipping) of a program, a first version of the program is shipped together with an {\itercomp} mechanism.
-In the online scenario, the program is shipped with an initial optimisation sequence and different optimisation sequences are evaluated as the end-user executes the program.
-This optimisation strategy is also known as idle-time optimisation, as the re-compilation happens between runs of the program.
-
-LLVM is particularly suitable for iterative compilation as it makes possible to cache a pre-compiled, but still unoptimised, version of the input program in the bitcode format of the LLVM IR.
-This caching allows to speedup the time required for re-compilation as it is able to bypass the frontend phase.
-If re-compilation time is critical, it would also be possible to keep only the hot portion of the code in the LLVM bitcode format, while the remaining portion of the code is already compiled to the final object code.
-However, this is out of the scope of this paper and we always re-compile the whole program.
-
-\begin{figure}[htb]
-    \centering
-    \includegraphics[width=\linewidth]{figs/infra-diagram}
-    \caption{Overview of the execution engine for applying {\itercomp}.}
-    \label{fig:infra-diagram}
-\end{figure}
-%\textbf{Describe the process step by step.}
-
-Figure~\ref{fig:infra-diagram} shows an overview of the infrastructure required for applying online {\itercomp}
-using WP as the metric of choice for evaluating optimisation sequences.
-The online {\itercomp} follows as described bellow:
-\begin{enumerate}
-\item The program is pre-compiled to the LLVM bitcode format without optimisation.
-\item The unoptimised program is instrumented for work profiling.
-\item Execution-based optimisation search:
- \begin{enumerate}
-   \item The current optimisation sequence is used for the re-compilation of the program.
-   \item The program is executed with any input provided by the end-user.
-         During the execution of the program, wall-clock time and the work metric are recorded by profiling instrumentation.
-   \item If the recorded profiling for the current optimisation can be used to compute an average performance measurement within a small confidence interval,
-         then a new optimisation sequence is generated.
-         Otherwise, the same optimisation sequence is used for the next execution.
- \end{enumerate}
-\end{enumerate}
-
-In this work, we focus mainly on the highlighted components.
-The \textit{Work Instrumentation} phase focuses on providing a low-overhead instrumentation for profiling the work metric.
-The instrumentation consists of adding a global counter to the program which is used to accumulate the amount of work computed during the program's execution, using the cost model of the instruction set.
-A detailed description of the work instrumentation is presented in Section~\ref{chap:instr}.
-
-Notice how the \textit{Work Instrumentation} phase is executed before performing any optimisation to the program.
-It is critical so that the work profiling always measure the same amount of work for a given input, regardless of the optimisation sequence applied to the program.
-Because the instrumentation is performed before optimising the program, it means that the work profiling derives the linear expression defined in Equation~\ref{eq:linear-work-expression}
-based on the unoptimised program.
-In other words, the basic blocks and the number of occurrences of the instruction in the basic blocks reflect the unoptimised program.
-This particular sequence of compilation guarantees that the amount of work must is only input dependent, but consistent between different optimisation sequences.
-
-As the name suggests, the component called \textit{Optimisation Selector} is responsible for selecting which optimisation sequence to use for the next execution of the program.
-It can either keep the same optimisation sequence used in the previous execution or start monitoring the performance of a new optimisation sequence.
-An optimisation sequence can be kept for multiple executions of the program in order to gather enough measurement to compute an average performance with small statistical deviations, i.e., for which the confidence interval has a small range.
-We call by \textit{Input-Window Size} the number of executions performed using the same optimisation sequence.
-
-\subsection{Real Online Scenarios}
-
-In most online scenarios, it is common for periods of peak usage and idle periods.
-For example, mobile devices are usually intensely used during the day, with some idle periods usually when the battery is being re-charged~\citep{mpeis16}.
-Similarly, many authors have also considered peak and idle (or underutilised) periods in the context of data centres~\citep{armbrust10,chen12b}.
-
-The proposed infrastructure is very well suited for these real online scenarios\footnote{
-However, if idle time is almost non-existent, the proposed infrastructure can still be used by re-compiling the program with a different optimisation while multiple runs of the program are being executed.}.
-In particular, periods of peak usage could be used to monitor multiple runs of the program using the same optimisation sequence, collecting the work profiling and measuring its execution time,
-while periods of idleness or underutilisation could be leveraged to use the profiling statistics for selecting better optimisations and re-compiling the program.
-
-\section{Instrumentation}
-
-In this section we describe how the computation of the work-metric can be performed during runtime by means of instrumenting the code.
-In particular, we adapt the optimal algorithm proposed originally for profiling block frequency~\cite{knuth73,ball94}.
-Afterwards, we propose a relaxed instrumentation that focus on further reducing the overhead by considering the trade-off between profiling accuracy and instrumentation overhead.
-
-Because we define work as a linear equation on the block frequency counters, it is possible to embed its computation into the execution of the program.
-A naive instrumentation would consists basically of having a global counter that starts with the interception value, $\varepsilon$, and each basic block increments its own cost into the global counter.
-Although this instrumentation is easily implemented, it inserts a large amount of overhead into the program.
-However, it is possible to insert fewer probes by carefully placing the probes in a way that is possible to reconstruct the complete profiling information~\cite{knuth73,ball94}.
-
-Ball and Larus~\cite{ball94} show that a set of edges represents the minimum number of probes for profiling block frequency if and only if the complementary set of edges forms a spanning tree.
-The instrumentation optimises the placement of the probes with respect to a weighing that assigns a non-negative value to each edge in the CFG.
-The cost of profiling a set of edges is proportional to the sum of the weights of the edges.
-These weights can be obtained either by empirical measurements or heuristic estimations.
-In order to minimise the profiling cost, the instrumentation computes the maximum spanning tree for avoiding probing in frequently executed edges.
-Although this instrumentation algorithm is proved to produce the optimal placement of probes for well-structured CFGs, it may produce sub-optimal placement for some unstructured CFGs.
-
-Once we have the maximum spanning tree, probes are placed on any edge not in the spanning tree.
-Figure~\ref{fig:cfg-example} shows an example of a CFG with a maximum spanning tree represented by the black edges, while the edges highlighted in red represent the placement of the probes.
-
-\begin{figure}[htb]
-\centering {
-  \includegraphics[scale=0.85]{figs/cfg-example.pdf}\\\vspace{1ex}
-  \resizebox{0.45\textwidth}{!}{
-  %\scalebox{0.8}{
-     \begin{minipage}{0.5\textwidth}
-     Instrumented value for each probe $P_i$:
-     \begin{align*}
-     w(P_0) &= w(B_0) + w(B_1) + w(B_4) + w(B_6) + w(B_7) + w(B_8) + w(B_9)\\
-     w(P_1) &= w(B_2) - w(B_1) - w(B_4) - w(B_6) - w(B_7) - w(B_8)\\
-     w(P_2) &= w(B_5) - w(B_6)\\
-     w(P_3) &= w(B_3) - w(B_4) - w(B_6) - w(B_7)
-     \end{align*}
-     \end{minipage}
-  }
-}
-  \caption{Example of a CFG with its minimum spanning tree in black and the
-   basic blocks highlighted in red represent the instrumented basic blocks with
-   the placement of the probes.}
-  \label{fig:cfg-example}
-\end{figure}
-
-In contrast to the naive instrumentation where each basic block records only its own amount of work, with the optimal profiling, the instrumented basic blocks need record an aggregated value of work that represents a path in the CFG.
-These values are constructed with some instrumented basic blocks speculatively assuming some paths while other probes correct when these assumptions are wrong (see for example $w(P_0)$ and $w(P_1)$ in Figure~\ref{fig:cfg-example}).
-
-\begin{lstlisting}[caption={Post-processing of the CFG for populating all edge flows based on the collected probes.
-\textbf{ToDo: Replace the algorithm with a formal data-flow equation.}}, label={lst:populateEdgeFlows}, float]
-// Inputs: CFG with the known edges flows from
-//         the cotree (collected probes).
-// Output: Updated CFG with all edge flows.
-populateEdgeFlows(G) {
-  changed = true
-  while changed:
-    changed = false
-    for B in G.vertices():
-      unIN = count( G.unknownIncomingEdges(B) )
-      unOUT = count( G.unknownOutgoingEdges(B) )
-      if unIN==0 and unOUT==1:
-        //sum known incoming and outgoing edges
-        sIN = sum( G.incomingEdges(B) )
-        sOUT = sum( G.outgoingEdges(B) )
-        //update unknown outgoing edge in B
-        //with (sIN-sOUT)
-        G.setUnknownOutgoingEdge(B, (sIN-sOUT))
-        changed = true
-      if unIN==1 and unOUT==0:
-        //sum known incoming and outgoing edges
-        sIN = sum( G.incomingEdges(B) )
-        sOUT = sum( G.outgoingEdges(B) )
-        //update unknown incoming edge in B
-        //with (sOUT-sIN)
-        G.setUnknownIncomingEdge(B, (sOUT-sIN))
-        changed = true
-}
-\end{lstlisting}
-
-Because the algorithm for the optimal placement of the probes is proved to uniquely compute the block frequencies by propagating the probe counts, we adapt this algorithm in order to compose the aggregated values that will be instrumented in each probe, based on our model of computational work, $\Delta W$, derived from the basic block frequencies (see Section~\ref{sec:metric}).
-We perform a similar propagation of the probes in a symbolic fashion, as illustrated in Figure~\ref{fig:cfg-example}.
-This symbolic propagation of the probes is implemented by the data-flow analysis described in Algorithm~\ref{alg:populateEdgeInfo}, and the final aggregated values are extracted from the edge information as described by Algorithm~\ref{alg:instrValue}.
-
-The data-flow analysis in Algorithm~\ref{alg:populateEdgeInfo} keeps two sets for each edge, namely, the increment and the decrement sets.
-We consider that both sets represent the \textit{edge expressions} shown in Figure~\ref{fig:cfg-example}, for which we define a \textit{symbolic sum} by computing the union of the increment and decrement sets, respectively, with the appropriate cancellation of common elements.
-This data-flow analysis is based on the invariant that the symbolic sum of all the incoming edges must equals the symbolic sum of the outgoing edges.
-For example, the symbolic sum of the incoming edges of the basic block $B_8$ is $P_0 - P_1$, where $P_3$ is cancelled out.
-
-%By construction, the \textit{symbolic experssions} of a basic block, i.e., the \textit{symbolic sum} of the incoming edges (or outgoing edges) of a basic block, has the following properties:
-%\begin{enumerate}
-%\item Probes in the increment set either (post-)dominates or is (post-)dominated by the given basic block;
-%\item Probes in the decrement set are (post-)dominated by the probes in the increment set;
-%\item Probes in the decrement set and the given basic block are in exclusive paths from the virtual node of the CFG to the (post-)dominators in the increment set;
-%\end{enumerate}
-%From property (1) it follows that whenever the given basic block is executed, exactly one of the probes in the increment set is executed.
-%From properties (2) and (3), it follows that whenever a probe in the decrement set is executed, although the given basic block is not executed, one of the probes in the decrement set is executed.
-
-Algorithm~\ref{alg:instrValue} reads the edge information for each basic block by computing the symbolic sum of their respective incoming edges (or outgoing edges).
-From these \textit{edge expressions}, we are able to compose the aggregated value of the probes.
-The positive terms in the edge expression of a basic block indicate that the amount of work of this basic block will be incremented in the probes represented by these positive terms, similarly, the negative terms indicate that the amount of work of this basic block will be decremented in the probes represented by these negative terms.
-For example, because the edge expression for the basic block $B_8$ is $P_0 - P_1$, the amount of work of $B_8$, denoted by $w(B_8)$, is incremented in probe $P_0$ and decremented in $P_1$.
-
-\begin{lstlisting}[caption={Pseudocode that describes how the edge information is used in order to extract the value that will be computed in a given instrumented basic block $B_I$.
-This algorithm could equally be implemented based on the predecessors.
-\textbf{ToDo: Using the vector format, we could replace this algorithm with a single equation. Let $V$ be the vector of $\{-1,0,1\}$ and $W$ the weight for each basic block.
-This algorithm could be re-written as $reduce (+) (map (*) (zip V W))$.}
-}, label={lst:instrValue}, float]
-// Inputs: 1) CFG with the known edges flows
-//            of the chords.
-//         2) Basic block targeted for probing.
-// Output: Work value to be incremented by the
-//         given probe B.
-instrValue(G, B, inc, dec) {
-  value = 0
-  for B in G.vertices():
-    incB = set()
-    decB = set()
-    for succB in G.successor(B):
-      incB = incB union inc[G.getEdge(B, succB)]
-      decB = decB union dec[G.getEdge(B, succB)]
-    if B in (incB - decB):
-      value = value + w(B)
-    if B in (decB - incB):
-      value = value - w(B)
-  return value
-}
-\end{lstlisting}
-
-\subsection{Relaxed Instrumentation}
-
-Although the optimal instrumentation significantly reduces the profiling overhead when compared to the naive instrumentation, from an average overhead of 79\% to 13\%, in some critical cases, even the optimal instrumentation can have an overhead of about 70\% (see benchmark \texttt{adpcm\_d} in Figure~\ref{fig:overhead-O3}).
-In order to further reduce the overhead in these critical cases, we propose a relaxed instrumentation by trading off accuracy and overhead.
-While the optimal placement of probes tries to place probes in edges that are less likely to be executed, the relaxation focus on removing probes that are more likely to be executed but add little to the work metric.
-
-% \begin{figure}[h]
-%   \centering
-%   \includegraphics[scale=0.85]{figs/relax-instr-diagram.pdf}
-%   \caption{Overview of the work instrumentation algorithm, including the relaxation technique.}
-%   \label{fig:relax-instr-diagram}
-% \end{figure}
-
-%Figure~\ref{fig:relax-instr-diagram} shows an overview of the relaxed instrumentation algorithm.
-%The highlighted step is introduced by the relaxed instrumentation on top of the previously defined optimal profiling.
-The relaxed instrumentation performs a post-processing on the resulting instrumentation of the optimal algorithm.
-This post-processing identifies probes that add little to the work metric, and removes their instrumentation.
-In order to guarantee an upper bound for the dynamic error in the profiling measurement, the relaxation algorithm applies a constrained post-processing on a per DAG (directed acyclic graph) basis.
-By constraining the relaxation within each DAG by a maximum allowed static error, it guarantees that the overall relaxation will also be constrained by the same bound.
-
-The relaxation starts by extracting DAGs (directed acyclic graphs) from the CFG, as illustrated in Figure~\ref{fig:cfg-relax-example}.
-First, the algorithm extracts all the subgraphs that represent a loop or the outer most region of the function.
-Section~\ref{sec:natural-loops} explains how loops can be identified in the CFG.
-Afterwards, these subgraphs are transformed into DAGs by ignoring the backedge and also by considering that any loop within the subgraph is never executed, i.e., only the headers of the inner loops are actually included into the DAG.
-Figure~\ref{fig:cfg-relax-example} shows a CFG partitioned into two DAGs (consider only basic blocks and edges completely inside the yellow and green boundaries).
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[scale=0.85]{figs/cfg-relax-example.pdf}
-  \caption{Example of a CFG containing a loop and its decomposition into DAGs when applying the relaxation.
-           The DAGs are the subgraphs within the dashed boundaries.}
-  \label{fig:cfg-relax-example}
-\end{figure}
-
-%\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
-%// Input: CFG
-%relaxInstrumentation(G) {
-%  for loop in G:
-%     DAG = colapseInnerLoops(loop)
-%     relaxInstrumentedDAG(DAG)
-%  DAG = colapseInnerLoops(G)
-%  relaxInstrumentedDAG(DAG)
-%}
-%\end{lstlisting}
-
-For every DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we relax the profiling accuracy by selecting a subset of the probes to be removed, subject to the maximum allowed percentage error, $M$.
-
-We model the relaxation as a 0-1 Knapsack problem:
-\begin{equation*}
-\begin{aligned}
-& \textrm{maximise }\quad \sum_{i=0}^{k} f(P_i)x_i \\
-& \textrm{subject to }\quad \sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M
-\end{aligned}
-\end{equation*}
-\[
-x_i\in\{0,1\}, i\in\{0,\ldots,k\}
-\]
-where $f(P_i)$ is the execution frequency of probe $P_i$, $x_i$ denotes the probes selected for removal, and $\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to the minimum work value possible to compute in the DAG, i.e., if $m$ is the minimum amount of work possible to be computed when executing the DAG, then $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$.
-Because the percentage error is computed based on the path with the minimum amount of work, $\varepsilon(P_i)$ represents the maximum error possible that would be incurred when removing probe $P_i$.
- Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold.
-% Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold, as demonstrated by Proposition~\ref{prop:relax-bound}.
-%
-% \begin{prop}\label{prop:relax-bound}
-% Let $n_i$ be the number of times a given DAG $i$ is executed, $r_i$ be the total relaxation (amount of work removed) in DAG $i$, and $m_i$ be its minimum amount of work.
-% If $\frac{r_i}{m_i} \leq M$ for every $i$,
-% then the final error of the relaxation will always be bounded by the same threshold.
-% \end{prop}
-% \begin{proof}
-% We can model the overall error of the relaxation as:
-% \[
-% 1 - \frac{n_1(m_1 - r_1) + n_2(m_2 - r_2) + \ldots + n_k(m_k - r_k) + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
-% \]
-% That is,
-% %\begin{equation*}
-% \begin{gather*}
-%  1 - \frac{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} + \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} = \\
-%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
-% \end{gather*}
-% %\end{equation*}
-% If $\frac{r_j}{m_j}$ is the maximum ratio $\frac{r_i}{m_i}$ for every $i$, then
-% \begin{equation*}
-% \begin{aligned}
-%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} &\leq\\
-%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j + c} &\leq\\
-%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &
-% \end{aligned}
-% \end{equation*}
-% If $N = max\{n_i$ for every $i\}$, then
-% \begin{equation*}
-% \begin{aligned}
-%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &\leq\\
-%  \frac{Nr_j + Nr_j + \ldots + Nr_j}{Nm_j + Nm_j + \ldots + Nm_j} &=\\
-%  \frac{Nkr_j}{Nkm_j} = \frac{r_j}{m_j} &\leq M
-% \end{aligned}
-% \end{equation*}
-% \end{proof}
-
-%\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
-%// Input: CFG
-%relaxInstrumentedDAG(DAG){
-%   P = ProbesIn(DAG)
-%   m = minWork(DAG,P)
-%   K = createKnapsackModel(P,m)
-%   Bag = solveKnapsack(K)
-%   for B in (P-Bag):
-%     removeProbe(B)
-%}
-%\end{lstlisting}
-
-The necessary block-frequency information for optimising both the placement of probes can be acquired from profiles of previous executions of the program or by a static heuristic of the CFG during compilation.
-
-For our experiments, we implemented two solvers for the 0-1 Knapsack problem:
-the optimal brute-force solver;
-the greedy heuristic based on sorting the items~\citep{dantzig57}.
-We use the brute-force solver for DAGs with a small number of probes and the greedy heuristic when the number of probes is greater than a threshold.
-Some of the benchmarks have DAGs with several hundreds of probes, which could result in a long compilation time.
-
-\subsection{Whole Program Relaxation}
-
-In some cases, the proposed relaxation can be very conservative, because it considers the static error of removing probe $P_i$ relative to the minimum work possible of a DAG, in order to be able to guarantee that the dynamic error will be bounded by a given threshold.
-This conservatism can be overly restrictive in some cases, resulting in a negligible overhead reduction but also causing just a negligible dynamic error to the work profiling.
-
-In some cases, this overly restrictive conservatism may be unnecessary.
-For these cases, we propose an adapted version of the relaxation algorithm that operates on the whole program.
-Traditionally, compilers optimisations are performed on the function-level, or at best on a per module basis.
-Whole program optimisation (WPO) means that the compiler considers all compilation units of the program and optimises them using the combined knowledge of how they are used together.
-
-The whole program relaxation works by using block-frequency profiling from previous executions.
-By having this profiling information, the \textit{whole program relaxation} is able to compute the error of removing a given probe in terms of the whole program's execution,
-and then use this error values for selecting a subset of all the probes to be removed.
-
-For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we model the whole program the relaxation as the following 0-1 Knapsack problem:
-\begin{equation*}
-\begin{aligned}
-& \textrm{maximise }\quad \sum_{i=0}^{k} f(P_i)x_i \\
-& \textrm{subject to }\quad \sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M
-\end{aligned}
-\end{equation*}
-\[
-x_i\in\{0,1\}, i\in\{0,\ldots,k\}
-\]
-where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$, $x_i$ denotes the probes selected for removal, $M$ is the error threshold, and $\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to the profiled global work, i.e.,
-if $\Delta W$ is the work value for the whole program's execution, computed from the basic block frequencies profiled from previous executions, the error for a given probe $P_i$ is
-\[
-\varepsilon(P_i) = \frac{\omega(P_i)f(P_i)}{\Delta W}.
-\]
-
-Contrary to the per DAG relaxation, the whole program relaxation is not guaranteed to be bounded by the error threshold $M$,
-as it depends on the representativity % representativeness
- of the profiling information provided to the whole program relaxation.
-
-\section{Experimental Evaluation}
-
-In this section we discuss our experimental evaluation.
-First we describe the benchmarks with datasets used in the experiments.
-Afterwards, we discuss our results concerning the instrumentations for profiling our work metric.
-Finally we present the results of the online {\itercomp}.
-
-We implemented the instrumentations in LLVM 4.0.
-The target platform is a Linux-4.4.27 system with an Intel Core i7-4770 3.40GHz Skylake~CPU with 16~GiB RAM.
-
-\subsection{Benchmarks}
-
-For the experimental evaluation we have used a subset of the \textit{KDataSets} benchmark suit, which is the same benchmark and dataset suit used by Chen~\etal~\cite{chen10,chen12a}.
-The KDataSets contains 1000 different inputs for each one of its benchmark programs.
-These benchmarks cover a broad spectrum of application scenarios, ranging from simple embedded signal-processing tasks to common mobile-phone and desktop tasks.
-The different inputs try to capture distinct characteristics in terms of workload sizes and how these workloads exercise different control flow paths.
-A summary of the benchmarks and dataset suit is shown in Table~\ref{tab:kdatasets}.
-
-\begin{table}[h]
-\centering
-\scalebox{.8}{
-\begin{tabular}{|c|c|c|c|}
-\hline
-\textbf{Program} & \textbf{LOC}    & \textbf{Input file size}            & \textbf{Input description}              \\ \hline % Domain
-qsort         & 154    & 32K-1.8M                   & 3D coordinates                 \\ \hline
-jpeg\_d       & 13501  & 3.6K-1.5M                  & JPEG images                    \\ \hline
-jpeg\_c       & 14014  & 16K-137M                   & PPM images                     \\ \hline
-tiff2bw       & 15477  & \multirow{4}{*}{9K-137M}   & \multirow{4}{*}{TIFF images}   \\ \cline{1-2}
-tiff2rgba     & 15424  &                            &                                \\ \cline{1-2}
-tiffdither    & 15399  &                            &                                \\ \cline{1-2}
-tiffmedian    & 15870  &                            &                                \\ \hline
-susan\_c      & 1376   & \multirow{3}{*}{12K-46M}   & \multirow{3}{*}{PGM images}    \\ \cline{1-2}
-susan\_e      & 1376   &                            &                                \\ \cline{1-2}
-susan\_s      & 1376   &                            &                                \\ \hline
-adpcm\_c      & 210    & 167K-36M                   & WAVE audios                    \\ \hline
-adpcm\_d      & 211    & 21K-8.8M                   & ADPCM audios                   \\ \hline
-lame          & 14491  & 167K-36M                   & WAVE audios                    \\ \hline
-rsynth        & 4111   & 0.1K-42M                   &  Text files                    \\ \hline
-sha           & 197    & 0.6K-35M                   & Files of any format            \\ \hline
-\rowcolor{gray!20}
-bitcount      & 460    &  -                         & Numbers: random                \\ \hline
-\rowcolor{gray!20}
-dijkstra      & 163    & 0.06K-4.3M                 & Adjacency matrices             \\ \hline
-\rowcolor{gray!20}
-patricia      & 290    & 0.6K-1.9M                  & IP and mask pairs              \\ \hline
-\rowcolor{gray!20}
-mad           & 2358   & 28K-27M                    & MP3 audios                     \\ \hline
-\rowcolor{gray!20}
-%gsm           & 3806   & 83K-18M                    & Sun/NeXT audios                \\ \hline
-ghostscript   & 99869  & 11K-43M                    & Postscript files               \\ \hline
-\rowcolor{gray!20}
-%ispell        & 6522   & \multirow{3}{*}{0.1K-42M}  & \multirow{3}{*}{Text files}    \\ \cline{1-2}
-%rsynth        & 4111   &                            &                                \\ \cline{1-2}
-stringsearch  & 338    &  0.1K-42M                 &  Text files                     \\ \hline
-\rowcolor{gray!20}
-%blowfish\_e   & 863    & 0.6K-35M                   & Files of any format            \\ \hline
-%blowfish\_d   & 863    & 0.6K-35M                   & Encrypted files                \\ \hline
-%pgp\_e        & 19575  & 0.6K-35M                   & Files of any format            \\ \hline
-%pgp\_d        & 19575  & 0.4K-18M                   & Encrypted files                \\ \hline
-%rijndael\_e   & 952    & 0.6K-35M                   & Files of any format            \\ \hline
-%rijndael\_d   & 952    & 0.7K-35M                   & Encrypted files                \\ \hline
-CRC32         & 130    & 0.6K-35M                   & Files of any format            \\ \hline
-\rowcolor{gray!20}
-bzip2e        & 5125   & 0.7K-57M                   & Files of any format            \\ \hline
-\rowcolor{gray!20}
-bzip2d        & 5125   & 0.2K-25M                   & Compressed files               \\ \hline
-\end{tabular}
-}
-\caption{Description of the KDataSets with 1000 inputs for each benchmark (Chen~\etal~\cite{chen10,chen12a}).}
-\label{tab:kdatasets}
-\end{table}
-
-The shaded (grey) benchmarks in Table~\ref{tab:kdatasets} represent the benchmarks used for training the cost model used for computing the weight of the instructions for the work metric.
-These same training benchmarks were also used for collecting a fixed set of optimisations for the {\itercomp}.
-The remaining (white) benchmarks are used for the experimental evaluation.
-
-\subsection{Evaluation of the Instrumentation}
-
-We first compare static aspects of the instrumentation algorithms.
-The naive instrumentation always has 100\% of the basic blocks instrumented, by definition.
-First, we evaluate the optimal work profiling compared to the relaxation algorithm applied per DAG.
-Afterwards, we compare these profiling techniques with the whole program relaxation.
-
-Figure~\ref{fig:num-probes} shows the percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.
-While Figure~\ref{fig:num-probes} compares the various instrumentation algorithms in respect of the naive instrumentation, Figure~\ref{fig:num-probes-improvement} shows the improvement of the relaxation algorithm  over the optimal instrumentation.
-
-\begin{figure*}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{figs/num-probes.pdf}
-    \caption{Percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.}
-    \label{fig:num-probes}
-\end{figure*}
-
-For the evaluation of the performance overhead that the instrumentation incur to the benchmarks, we measure the wall-clock time of the benchmarks when compiled with the default {\flagstype -O3} optimisation.
-For each benchmark, we compute the average overhead over all its 1000 input dataset.
-When measuring the wall-clock time for each input, in order to reduce noise, we execute the same input until we have a statistically sound measurement, i.e.
-we execute until we have an interval no larger than 1\% with 99\% confidence.
-Figure~\ref{fig:overhead-O3} shows the performance overhead imposed by the work instrumentation on the benchmarks when compared to their non-instrumented counterparts.
-Figure~\ref{fig:overhead-improvement-O3} indicates the improvement of the relaxation algorithm over the optimal instrumentation, regarding the reduction in overhead.
-
-\begin{figure*}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{figs/overhead-O3.pdf}
-    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.}
-    \label{fig:overhead-O3}
-\end{figure*}
-
-The WPO relaxation was performed using profiling information for every specific input, which provides the perfect information to perform the WPO relaxation.
-This means that the experiments show its best performance enabled by having perfect profiling information.
-
-%\begin{figure*}[ht]
-%    \centering
-%    \includegraphics[width=\textwidth]{figs/overhead-improvement-O3.pdf}
-%    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.
-%             Notice that the average of the ratios does not equal the ratio of the averages.}
-%    \label{fig:overhead-improvement-O3}
-%\end{figure*}
-
-\begin{figure*}[h!]
-    \centering
-    \includegraphics[width=\textwidth]{figs/error-O3.pdf}
-    \caption{Dynamic error of the work profiling averaged over the 1000 inputs, after relaxing the number of probes.}
-    \label{fig:error-O3}
-\end{figure*}
-
-%Figure~\ref{fig:overhead-improvement-O3} shows that the relaxation algorithm, with 5\% threshold, is able to improve an average of 48\% over the optimal instrumentation, 
-%and the WPO relaxation achieves an average of $2\times$ improvement over the optimal instrumentation.
-%While the optimal profiling has a maximum overhead of almost 60\%, the work profiling with a relaxation of 5\% threshold has an overhead of less than 20\% for all benchmarks,
-%with the WPO relaxation reaching an average overhead of only 5\%, compared to an average of 12\% for the optimal profiling.
-
-Furthermore, these improvements can be obtained while incurring very little dynamic error, as shown in Figure~\ref{fig:error-O3}.
-This result confirms the overly conservative aspect of the relaxation algorithm applied per DAG,
-while the WPO relaxation is significantly less conservative.
-Because we performed the WPO relaxation having perfect profiling information, its dynamic error was always below the 5\% thresholds.
-Although it is not guaranteed by the algorithm, profiling representative inputs should also result in small dynamic errors.
-
-\subsection{Evaluation of the Online {\IterComp}}
-
-For comparison, we use four configurations\footnote{A configuration using the WPO relaxation was not used due to the time necessary for executing the experiments.}.
-In all configurations, the same optimisation sequence is used for multiple inputs, using a dynamic input-window size, as explained in Section~\ref{sec:oic-infra}.
-The average performance over the input window provides an estimate for the overall performance of the optimisation sequence across distinct inputs.
-The optimisation sequences are ranked based on their average performance, and the best optimisation sequence is selected.
-%When selecting the best optimisation sequence, they are ranked by the average of their performance measurement, using the WP metric, except for the Oracle-RM which uses the real speedup over \texttt{-O0}.
-\begin{itemize}
-\item \textbf{Oracle-RM} executes the program multiple times for each input, measuring the real speedup for each optimisation sequence, and then uses the real speedup over {\flagstype -O0} for comparing the optimisation sequences.
-  The speedups are computed based on the wall-clock time.
-  In order to reduce noise, the program is executed several times for the same input, until the confidence interval was no larger than 1\% for a 99\% confidence.
-\item \textbf{Oracle-PP} represents an oracle with a \textit{perfect} non-intrusive profiling.
-  Although it uses the WP metric for comparing optimisation sequences, this oracle also avoids noise in its measurements by also executing the program multiple times for each input.
-  The first execution is used for profiling the work metric.
-  The remaining executions are used for measuring the wall-clock time without using the work profiling.
-\item \textbf{Real-OP} corresponds to the online {\itercomp} as it would be applied in a real online scenario.
-  For each optimisation, a random sample of inputs is selected, and the program is executed only once with each input.
-  When executing each input, it uses the optimal work instrumentation for profiling the work metric.
-  The average of the WP for the sample of inputs is then used for selecting the best optimisation.
-\item \textbf{Real-R5} is similar to the {Real-OP}.
-  It also corresponds to the online {\itercomp} as it would be applied in a real online scenario.
-  For each optimisation, a random sample of inputs is selected, and the program is executed only once with each input.
-  When executing each input, it uses the relaxed work instrumentation, with a 5\% threshold, for profiling the work metric.
-  The average of the WP for the sample of inputs is then used for selecting the best optimisation.
-\item \textbf{Real-IPC} corresponds to the online {\itercomp} based on the IPC metric.
-  For each optimisation, a random sample of inputs is selected, and the program is executed only once with each input.
-  The average of the IPC metric for the sample of inputs is then used for selecting the best optimisation.
-\end{itemize}
-The comparison between Oracle-RM and Oracle-PP is useful for validating the use of the WP metric for guiding {\itercomp}, while the other configurations demonstrate the viability of applying the online {\itercomp} in a real scenario.
-
-Figure~\ref{fig:window-size} shows the average input-window size for each benchmark and configuration.
-It shows that we can have a statistically sound measurement of the performance metric using just a small number of inputs.
-
-\begin{figure*}[htb]
-    \centering
-    \includegraphics[width=\textwidth]{figs/window-size.pdf}
-    \caption{Average input-window sizes observed during the online {\itercomp}.}
-    \label{fig:window-size}
-\end{figure*}
-
-\subsection{The Set of Optimisation Sequences}
-
-For the purpose of evaluating the use of the WP metric with {\itercomp}, we collected in advance a fixed set of optimisation sequences.
-The reason for using this fixed set, as explained in Section~\ref{sec:oic-infra}, is because this paper is focused on other components of the infrastructure for performing {\itercomp} and we assume that a good generator of optimisation sequences will be used in a real online scenario.
-This set contains 500 optimisation sequences collected in a random search using the training benchmarks.
-These optimisation sequences contain an average of 40 individual optimisation passes, including repetitions, with a maximum of 119 optimisation passes, but it also contains optimisation sequences which consist of a single flag, such as the default optimisations {\flagstype -O1}, {\flagstype -O2}, {\flagstype -O3}, {\flagstype -Os}, and {\flagstype -Oz}.
-
-  % \begin{minipage}{0.9\textwidth}
-  %    \vspace{1em}
-  %    \noindent\textbf{Example of a short optimisation sequence:}\vspace{-1ex}
-  %    \justify{\flagstype -mem2reg -simplifycfg -constprop -dce}
-  % \end{minipage}
-  %
-  % \begin{minipage}{0.9\textwidth}
-  %    \vspace{1em}
-  %    \noindent\textbf{Example of a long optimisation sequence:}\vspace{-1ex}
-  %    \justify{\flagstype -globalopt -reassociate -instcombine -loop-rotate -block-freq -deadargelim -early-cse -sroa -argpromotion -sccp -tbaa -barrier -constmerge \mbox{-loop-vectorize} -domtree -basicaa -memdep -basiccg -memcpyopt \mbox{-constprop} -adce -globaldce -mem2reg -constmerge \mbox{-globaldce} -constprop -instsimplify -dse -dce -simplifycfg -loop-unroll -reassociate -constprop \mbox{-globaldce} -instsimplify -adce -constmerge -bb-vectorize -dce -mergefunc -simplifycfg -dse -loop-unroll -globaldce}
-  % \end{minipage}
-  %
-  % \begin{minipage}{0.9\textwidth}
-  %    \vspace{1em}
-  %    \noindent\textbf{Example of an optimisation sequence which includes {-O3}:}\vspace{-1ex}
-  %    \justify{\flagstype -O3 -adce -globaldce -simplifycfg -memcpyopt -reassociate -mergefunc \mbox{-dce} -dse}
-  %    \vspace{2em}
-  % \end{minipage}
-
-Repeating the same optimisation pass can be beneficial and usually expected by other passes.
-For example, the {\flagstype -loop-simplify} pass is used for transforming loops into a canonical form by inserting pre-header and exit basic blocks.
-Although this pass inserts jumps due to redundant basic blocks, this canonical form can be favourable to other loop optimisations.
-Because of the redundant basic blocks, this optimisation pass expects that the {\flagstype -simplifycfg} will eventually be executor later on the optimisation pipeline.
-Another example of such inter-relation between transformations concerns the {\flagstype -licm} and {\flagstype -mem2reg} passes.
-The {\flagstype -licm} pass is responsible for moving invariant code out from the loop body.
-It usually creates new local variables, using memory access operations, for assisting with the code manipulation, which means that the executing the {\flagstype -mem2reg} pass afterwards would be useful as a cleanup pass for removing the extra memory accesses generated.
-However, many of the analysis required for identifying loop invariant also benefit from the transformations performed by the {\flagstype -mem2reg} pass.
-These examples illustrate the importance of repeating optimisation passes.
-Moreover, they illustrate the intricate relation amongst several transformations.
-
-However, all optimisation sequences in the set of optimisations were generated completely at random, without using any knowledge of individual transformations.
-Each optimisation sequence was generated in two steps: \textit{(1)} randomly selects the number of flags; \textit{(2)} randomly selects the flags, allowing repetitions.
-Afterwards, this randomly generated optimisation sequence would be included in the set of optimisation sequences only if it was able to improve the performance of a training benchmark, also selected at random, in respect of the {\flagstype -O3} default optimisation.
-This process was repeated until we obtained all the 500 distinct optimisation sequences.
-
-\subsection{Performance Evaluation}
-
-In order to evaluate the quality of the final optimisation sequences selected by the {\itercomp} search, we compare their speedup by measuring wall-clock time of the benchmarks when compiled with the standard {\flagstype -O3} optimisation.
-For each benchmark, after selecting the final optimisation sequence, we compute the average speedup over all the 1000 input dataset.
-When measuring the wall-clock time for each input, to reduce noise, we execute the same input until we have a statistically sound measurement, i.e.
-we execute until we have an interval no larger than 1\% with 99\% confidence.
-Figure~\ref{fig:speedups} shows these average speedups over all the 1000 input dataset.
-This figure shows that the best optimisation sequence selected with the Oracle-PP is very close to the performance of the best optimisation sequence selected with the Oracle-RM.
-This result is important for demonstrating that a work-based metric has the potential to produce good results in a real online scenario, where there is the restriction that programs execute distinct inputs only once.
-Moreover, the evaluation also indicates that the use of relaxation algorithms does not degrade the optimisation search,
-it might even benefit the {\itercomp} as it tends to insert smaller interferences in the performance measurement.
-Finally, Figure~\ref{fig:speedups} also presents the results of {\itercomp} guided by IPC, which is futher discussed in Figure~\ref{fig:ipc-vs-work}.
-
-\begin{figure*}[htb]
-    \centering
-    \includegraphics[width=\textwidth]{figs/speedups.pdf}
-    \caption{Speedups obtained from the final optimisation sequence selected by the online {\itercomp}.
-	         The speedups reported for each benchmark represents the average speedup across their complete 1000 input datasets.}
-    \label{fig:speedups}
-\end{figure*}
-
-\begin{figure*}[h]
-    \centering
-    \includegraphics[width=\textwidth]{figs/ipc-vs-work.pdf}
-    \caption{Comparison between {\itercomp} using IPC or the WP metric.}
-    \label{fig:ipc-vs-work}
-\end{figure*}
-
-Figure~\ref{fig:ipc-vs-work} presents a comparison between IPC and the WP metric regarding their correlation with the actual speedup over \verb|-O0| observed during {\itercomp}.
-Each point in the figure corresponds to each metric averaged over input windows collected during multiple executions of the {\itercomp}.
-The vertical lines correspond to the optimisation sequence selected by the {\itercomp} search, which either represents the highest IPC or the highest value of WP.
-In all cases, the optimisation sequence selected based on the WP is much closer to the best speedup than the optimisation sequence selected based on IPC, which can be attributed to the fact that the WP has a stronger correlation with the speedup.
-In all cases, IPC shows little to no correlation with the speedup, which corroborates the argument given in Section~\ref{sec:ipc-vs-work-metric}.
-
-For some cases, namely for the tiff-related benchmarks, the WP shows a worse correlation with speedup.
-We believe these cases could be improved by improving the cost model used to compute the work metric.
-However, even in these cases, the optimisation sequences selected based on the WP outperform those selected based on the IPC metric.
-
-\textbf{ToDo: Include results using Genetic Algorithm search.}
-
-\section{Related Work}\label{sec:relatedwork}
-
-\subsection{{\IterComp}}
-
-Chen~\etal~\cite{chen10,chen12a} evaluate the effectiveness of iterative compilations across a large number of input test cases.
-Their main motivation is to answer the question:
-\textit{How data input dependent is {\itercomp}?}
-Their results show that it is possible to find a combination of compiler optimisations that achieves at least 86\% of the maximum speedup across all input test cases.
-These optimal combinations are program-specific and yield average speedups up to 3.75$\times$ over the highest optimisation level of compilers.
-
-When optimising a program, the main method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a} evaluates each combination of compiler optimisations across all the available inputs, i.e., if $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimisations, they perform a total of $O(NM)$ runs of the program being optimised.
-Furthermore, they use a pre-defined set of only 300 different combinations of compiler optimisations, which represents a very small sample of the optimisation search space for most modern compilers, e.g.
-LLVM has 56 distinct optimisation passes and GCC has about 47 high-level (SSA form) optimisation passes and about 25 low-level (RTL) optimisation passes, which in both cases result in much more than $2^{50}$ distinct combinations of compiler optimisations, without considering repetition.
-
-Recent work~\cite{chen12b,fang15} have applied the same idea of performing input-dependent {\itercomp} to distributed applications on data centres.
-In summary, each worker receives a subset of the input dataset, called the evaluation dataset, to perform an \textit{online} {\itercomp} of the code being executed.
-Each worker performs the same the method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a}, i.e., they evaluate each combination of compiler optimisations across all the evaluation dataset.
-However, because the optimisation is performed online, they usually consider a small evaluation dataset and a small number of compiler optimisations.
-
-Fursin~\etal~\cite{fursin07} addressed the problem of comparing the effect of two optimisations on two distinct inputs.
-For that purpose, they proposed to use instruction per cycle (IPC) as the metric for performing such comparison.
-Their result show that using IPC seems promising as a robust metric for {\itercomp} across large input datasets.
-However, some specific optimisation techniques may affect the use of IPC as a robust metric, and specially IPC has been shown to provide poor performance estimation for multi-threaded programs~\cite{alameldeen06,eyerman08}.
-In particular, IPC can give a skewed performance measure if threads spend time in \textit{spin-lock loops} or other synchronisation mechanisms.
-Some existing work on performance assessment suggest that total execution time should be used for measuring performance of multi-threaded programs~\cite{alameldeen06,eyerman08}.
-Aalameldeen and Wood~\cite{alameldeen06} in particular suggest that a simple work-related metric should be used if the unit of work is representative enough.
-Work-related metrics have already been largely used for measuring performance of throughput-oriented applications, for other applications, however, choosing an appropriate unit of work can be more challenging~\cite{alameldeen06}.
-
-\subsection{Work and Input Size Metrics}
-
-Previous work have proposed profiling-based mechanism to estimate input sizes~\cite{zaparanuks12,coppa14}.
-Coppa~\etal~\cite{coppa14} in particular propose the concept of \textit{read memory size} for automatically estimating the size of the input passed to a routine, where \textit{read memory size} represents the number of distinct memory cells first accessed by a read operation.
-In other words, the \textit{read memory size} metric measures the size of the useful portion of the input's memory footprint.
-However, because we are interested in the amount of computational work performed in respect of a given input, the memory footprint of the input may not always have a direct correspondence to  the amount of computational work.
-
-Goldsmith~\etal~\cite{goldsmith07} use \textit{block frequency} as the measure for performance for empirically describing the asymptotic behaviour of programs, which is known as empirical computational complexity.
-Block frequency is a relative metric that represents the number of times a basic block executes~\cite{ball94,ball96}.
-They argue in favour of block frequency due to its portability, repeatability and exactness, since it does not suffer from timer resolution problems or non-deterministic noises.
-Block frequency also has the advantage of being efficiently profiled by means of automatic code instrumentation~\cite{knuth73,ball94}.
-
-However, in the context of comparing different optimisations, although block frequency would be able to capture aspects of optimisations that simplify the control-flow graph (CFG), measuring work at the basic block resolution would not capture effects of optimisations at the instruction level.
-Because of that, we extend the idea of using basic block frequency to measure computational work by also considering the computational cost of each basic block.
-The computational cost of a basic block is given by weighing the instructions that it contains.
-
-\subsection{Optimal Instrumentation}
-
-In 1973, Donald Knuth~\cite{knuth73} provided an optimal algorithm for profiling block frequency, inserting fewer probes than the previously described naive instrumentation that was commonly used in practice~\cite{knuth71}.
-Forman~\cite{forman81} proposed further overhead reductions by placing the probes in basic blocks that are less likely to be executed based on static heuristics.
-
-Ball and Larus~\cite{ball94} offered a detailed discussion comparing two approaches for optimally profiling block frequency, namely, placing probes on the vertices or the edges of the control-flow graph (CFG).
-They show that the edge-based approach produces optimal placement of probes.
-
-\section{Conclusion and Future Work}
-
-Our proposed technique for relaxed instrumentation can also be applied for similar profiling scenarios, such as profiling empirical computational complexity~\cite{goldsmith07,zaparanuks12,coppa14}.
-
-\section*{Acknowledgments}
-
-This work was supported by the UK Engineering
-and Physical Sciences Research Council (EPSRC) under grants
-EP/L01503X/1 for the University of Edinburgh, School
-of Informatics, Centre for Doctoral Training in Pervasive
-Parallelism, % (\url{http://pervasiveparallelism.inf.ed.ac.uk/}),
-and also by the Institute for Computing Systems Architecture (ICSA)
-in the School of Informatics at the University of Edinburgh.
-
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{bibliography}
-
-\end{document}

### 2017-09-06 cgo18mscr.tex
moving staff

@@
 \documentclass[sigplan,10pt]{acmart}
-
-\usepackage{booktabs} % For formal tables
-
-\usepackage[utf8]{inputenc}
-\usepackage{fullpage}
-
-\usepackage{graphics}
-
-\usepackage{graphicx}
-\usepackage{subcaption}
-
-\usepackage{multirow}
-
-%\usepackage{setspace}
-\usepackage{ragged2e}
-
-	\usepackage{soul}
-
-%\usepackage{flushend}
-
-%\usepackage{indentfirst}
-
-\usepackage{amsthm}
-\usepackage{amsmath,amssymb}
-%\usepackage{mathrsfs}
-\usepackage{mathtools}
-
-\theoremstyle{definition}
-\newtheorem{prop}{Proposition}[section]
-
-%% Listing related stuff
-
-\usepackage{listings}
-\usepackage{courier}            %Required for pretty listings (more dense)
-
-%% EMACS colors for listing:
-
-\usepackage{color}
-\definecolor{sh_comment}{rgb}{0.65, 0.00, 0.00 } % red-ish
-\definecolor{sh_keyword}{rgb}{0.37, 0.69, 0.69}  % blue-green
-\definecolor{sh_string}{rgb}{0.08, 0.69, 0.08} % green
-
-%% This manipulates the listing numbering. It decreases it by one at the point used.
-\newcommand*\lstDNumber{\addtocounter{lstnumber}{-1}}
-
-%% Change from Listing->Algorithm
-%% \renewcommand\lstlistingname{Algorithm} % Fix listing caption name: Listing -> Algorithm
-%% \renewcommand\lstlistlistingname{Algorithms}
-%% \def\lstlistingautorefname{Alg.}
-
-%% \renewcommand{\thelstlisting}{\thesection-\arabic{lstlisting}}
-%\renewcommand{\thelstlisting}{\arabic{lstlisting}} % Fix listing numbering Algorithm 1.1 -> Algorithm 1.
-
-%% \renewcommand{\labelenumi}{\roman{enumi}.}
-
-\lstset{
-float=[*],
-language=C,                % choose the language of the code
-%% basicstyle=\linespread{0.9}
-%% basicstyle=\scriptsize\ttfamily,
- basicstyle={\linespread{0.97} \scriptsize\ttfamily},
- stringstyle=\color{sh_string},
- keywordstyle = \color{sh_keyword}\bfseries,
- commentstyle=\color{sh_comment}\itshape,
-numbers=left,                   % where to put the line-numbers
-%% numberstyle=\scriptsize,        % the size of the fonts that are used for the line-numbers
-numberstyle=\scriptsize,        % the size of the fonts that are used for the line-numbers
-stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
-numbersep=5pt,                  % how far the line-numbers are from the code
-backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
-showspaces=false,               % show spaces adding particular underscores
-showstringspaces=false,         % underline spaces within strings
-showtabs=false,                 % show tabs within strings adding particular underscores
-xleftmargin=2em,                % Left margin
-frame=lines,                   % adds a frame around the code. none|leftline|topline|bottomline|lines|single|shadowbox
-framexleftmargin=1.5em,         % Margin from frame to line numbers
-framexbottommargin=0em,         % Distance from last text to frame
-morekeywords={in,true,false,and,or,set,is,not},
-%% prebreak=\mbox{\tiny$\searrow$},
-prebreak=\space,                % Line break: insert this character at the end of the top line
-%% postbreak=\mbox{{\color{blue}\scriptsize$\hookrightarrow$}}, % Line break: blue arrow at the beginning of the bottom line
-postbreak=\mbox{{\color{blue}\scriptsize$\hookrightarrow$}}, % Line break: blue arrow at the beginning of the bottom line
-breaklines=true,                % sets automatic line breaking
-breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
-tabsize=2,                      % sets default tabsize to 2 spaces
-captionpos=b,                   % sets the caption-position to top
-escapeinside={\@}{\@}             % Escape sequence \@ESCAPED STUFF\@
-}
-
-\usepackage{nasm/lang}
-\usepackage{nasm/style}
-\usepackage{llvm/lang}
-
-% Copyright
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\setcopyright{usgov}
-%\setcopyright{usgovmixed}
-%\setcopyright{cagov}
-%\setcopyright{cagovmixed}
-
-% DOI
-%\acmDOI{10.475/123_4}
-
-% ISBN
-%\acmISBN{123-4567-24-567/08/06}
-
-%Conference
-\acmConference[CGO'18]{International Symposium on Code Generation and Optimization}{2018}{}
-%\acmYear{1997}
-%\copyrightyear{2016}
-%
-%\acmPrice{15.00}
-
-\newcommand{\etal}{et~al.}
-
-\newcommand{\itercomp}{{iterative compilation}}
-\newcommand{\Itercomp}{{Iterative compilation}}
-\newcommand{\IterComp}{{Iterative Compilation}}
-
-\newcommand{\flagstype}{\usefont{T1}{cmr}{m}{n}}
+\input{premable}
 
@@
 \title{Online {\IterComp}\\Guided by Low Overhead Profiling}
-%\titlenote{...}
-%\subtitle{...}
-%\subtitlenote{...}
 
-\author{ \vspace{1em} }
-\affiliation{%
-  \institution{ }
-}
-\email{ }
+\author{  }
 
@@
 
-\renewcommand{\shortauthors}{R. Rocha et al.}
+%\renewcommand{\shortauthors}{R. Rocha et al.}
 
-\begin{abstract}
-Until recently, most of the existing work in {\itercomp} had been focusing on
-finding the best optimisation through repeated runs using a single input.
-However, in real-world online scenarios, the user rarely executes a program with
-the same input multiple times and most of the programs are complex enough that a
-single input case does not capture the whole range of possible scenarios and
-program behaviours.
-
-In this thesis, we focus on enabling {\itercomp} in \textit{online} scenarios.
-Because of the restriction on repeating the execution with same input, it is
-not always feasible to measure the real speedup when comparing optimisations.
-To this end, we propose a \textit{work efficiency} metric to guide 
-online {\itercomp}, which is able to achieve about 80\% of the performance 
-of the offline oracle, with improvements of up to about 20\%.
-
-Moreover, because we need to instrument the program for measuring work
-efficiency, providing a low-overhead profiling is essential.
-In order to reduce the profiling overhead, we also propose two relaxation
-algorithms which provide a trade-off between measurement accuracy and overhead.
-Our results show that the two relaxation algorithms can reduce the average
-overhead by 43\% and $2.1\times$ over the optimal work profiling, observing
-up to about $5\times$ of overhead reduction.
-\end{abstract}
+\input{abstract}
 
@@
 %\ccsdesc[500]{Software and its engineering~Retargetable compilers}
-
 \keywords{{\IterComp}, Profiling, Relaxed Instrumentation}
@@
 
-\section{Introduction}
-Modern optimising compilers have reached a high level of sophistication,
-providing a large number of optimisations, where the correct choice of
-optimisations and their ordering can have a significant impact on the
-performance of the code being optimised.
-Although compilers offer a set of prearranged optimisation sequences that are
-expected to yield reasonable improvements in many programs, there is still the
-potential for performance degradation for certain programs, as these
-optimisation sequences do not include all possible optimisations and are always
-applied in the same pre-defined order, without regard the code being
-optimised~\cite{pan06,cavazos07,zhou12,kulkarni12}.
-A well known compilation technique that addresses this challenge is {\itercomp},
-which has the ability to adapt to new platforms, program and workload while
-still having a systematic and simple optimisation process.
-It works by repeatedly evaluating a large number of optimisation sequences until
-the best combination is found for a particular
-program~\cite{kisuki99,fursin07,chen10}.
-
-%{\Itercomp} is a well-known compilation technique that addresses the problem of
-%efficiently selecting the best optimisation sequence for a given program.
-%Although compilers offer a set of prearranged optimisation sequences that are
-%expected to yield reasonable improvements in many programs, there is still the
-%potential for performance degradation for certain programs, as these
-%optimisation sequences do not include all possible optimisations and are always
-%applied in the same pre-defined order, without regard the code being
-%optimised~\cite{pan06,cavazos07,zhou12,kulkarni12}.
-%{\Itercomp} have a systematic and simple optimisation process, where it works
-%by repeatedly evaluating a large number of optimisation sequences until the best
-%combination is found for a particular program~\cite{kisuki99,fursin07,chen10}.
-
-Until recently, most of the existing work in {\itercomp} had been focusing on
-finding the best optimisation through repeated runs using a single input.
-Although they demonstrate the potential of {\itercomp}, in real scenarios the
-user rarely executes a program with the same input multiple
-times~\cite{bodin98,kisuki99,stephenson03,kulkarni04,agakov06}.
-Furthermore, most of real world applications are complex enough so that a single
-input case does not capture the whole range of possible scenarios and program
-behaviours~\cite{haneda06,fursin07,chen10,chen12a}.
-Because programs can exhibit behaviours that differ greatly depending on the
-input, using a single input for {\itercomp} may not result in a good performance
-when executing the optimised code with different inputs.
-
-The main goal of this paper is to enable {\itercomp} in \textit{online} scenarios.
-We define the online scenario as having the restriction that programs execute multiple inputs and distinct inputs are executed only once.
-This online aspect is usually found in mobile and data centre platforms~\citep{chen12b,fang15,mpeis16}, where the goal is to optimise programs based on the workload of a particular user, or group of users.
-Online {\itercomp} must also target optimal performance across different inputs.
-
-Because of the restriction of having a single execution per input, it is not possible to measure speedup for comparing optimisations.
-On the other hand, measuring just execution time, for example, is useful only if the amount of work is constant between executions with different inputs.
-While previous work have suggested using \textit{instructions per cycle} (IPC) for performing {\itercomp} in \textit{online} scenarios, IPC have no correlation with speedup~\citep{fursin07}.
-
-In order to address this challenge, we propose the use of a work-based metric to compare the performance of different optimisations across multiple executions of the program with distinct inputs.
-We instrument the program for measuring the amount of work it performed during its execution.
-Having a low overhead instrumentation is essential in this online scenario for two main reasons:
-$(i)$ the user is directly affected by large overheads;
-$(ii)$ a highly intrusive instrumentation can have a significant impact on the effect of the optimisations.
-With the purpose of reducing profiling overhead, we propose two relaxation algorithms which provide a trade-off between measurement accuracy and overhead.
-The first is a relaxation algorithm that operates on the level of regions of functions, while the second performs the relaxation considering the whole program at the same time.
-
-Our experimental evaluation shows that performing online {\itercomp} guided by the work-based performance (WP) metric good results compared to the oracle, which is allowed to execute each input multiple times in order to use the actual speedup for guiding the {\itercomp}.
-Online {\itercomp} guided by the WP metric is able to achieve an average of 7.5\% and a maximum of 33\% improvement over the standard {\flagstype -O3} optimisation.
-Moreover, the experiments regarding the work profiling show that both relaxation algorithms are able to significantly reduce the profiling overhead while incurring a dynamic error of less than 5\% in the work measurement.
-The whole program relaxation achieves an average of $2\times$ reduction in the overhead compared with the optimal profiling technique, while the more conservative relaxation that operates per region achieves an average improvement of 40\% over the optimal profiling.
-
-%Our main contributions are the following:
-%\begin{itemize}
-%\item The use of a work-based performance metric in order to enable \textit{online} {\itercomp} by comparing different combination of compiler optimisations even when executed with distinct inputs.
-%\item We propose a relaxed instrumentation for low overhead profiling, with a controlled trade-off between accuracy and overhead.
-%\end{itemize}
-
-To summarise, the main contributions of this paper are the following:
-\begin{itemize}
-\item We propose the use of a work-based performance (WP) metric for comparing different optimised versions of a program.
-      In particular, we use the WP metric to guide {\itercomp} in an online scenario, where the program is expected to execute only once for distinct inputs.
-
-\item Contrary to what previous work has suggested, we show that instructions per cycle (IPC) is not a good metric for online {\itercomp}.
-\item We adapt the block frequency profiling in order to measure the WP metric.
-\item A relaxation algorithm is proposed for lowering the overhead of the work profiling, with a controlled trade-off between accuracy and overhead.
-\item A whole program relaxation is proposed in order to further reduce the overhead of the work profiling.
-\end{itemize}
+\input{intro}
 
@@
 
-\section{Related Work}\label{sec:relatedwork}
+\input{related}
 
-\subsection{{\IterComp}}
+\input{conclusions}
 
-Chen~\etal~\cite{chen10,chen12a} evaluate the effectiveness of iterative compilations across a large number of input test cases.
-Their main motivation is to answer the question:
-\textit{How data input dependent is {\itercomp}?}
-Their results show that it is possible to find a combination of compiler optimisations that achieves at least 86\% of the maximum speedup across all input test cases.
-These optimal combinations are program-specific and yield average speedups up to 3.75$\times$ over the highest optimisation level of compilers.
-
-When optimising a program, the main method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a} evaluates each combination of compiler optimisations across all the available inputs, i.e., if $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimisations, they perform a total of $O(NM)$ runs of the program being optimised.
-Furthermore, they use a pre-defined set of only 300 different combinations of compiler optimisations, which represents a very small sample of the optimisation search space for most modern compilers, e.g.
-LLVM has 56 distinct optimisation passes and GCC has about 47 high-level (SSA form) optimisation passes and about 25 low-level (RTL) optimisation passes, which in both cases result in much more than $2^{50}$ distinct combinations of compiler optimisations, without considering repetition.
-
-Recent work~\cite{chen12b,fang15} have applied the same idea of performing input-dependent {\itercomp} to distributed applications on data centres.
-In summary, each worker receives a subset of the input dataset, called the evaluation dataset, to perform an \textit{online} {\itercomp} of the code being executed.
-Each worker performs the same the method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a}, i.e., they evaluate each combination of compiler optimisations across all the evaluation dataset.
-However, because the optimisation is performed online, they usually consider a small evaluation dataset and a small number of compiler optimisations.
-
-Fursin~\etal~\cite{fursin07} addressed the problem of comparing the effect of two optimisations on two distinct inputs.
-For that purpose, they proposed to use instruction per cycle (IPC) as the metric for performing such comparison.
-Their result show that using IPC seems promising as a robust metric for {\itercomp} across large input datasets.
-However, some specific optimisation techniques may affect the use of IPC as a robust metric, and specially IPC has been shown to provide poor performance estimation for multi-threaded programs~\cite{alameldeen06,eyerman08}.
-In particular, IPC can give a skewed performance measure if threads spend time in \textit{spin-lock loops} or other synchronisation mechanisms.
-Some existing work on performance assessment suggest that total execution time should be used for measuring performance of multi-threaded programs~\cite{alameldeen06,eyerman08}.
-Aalameldeen and Wood~\cite{alameldeen06} in particular suggest that a simple work-related metric should be used if the unit of work is representative enough.
-Work-related metrics have already been largely used for measuring performance of throughput-oriented applications, for other applications, however, choosing an appropriate unit of work can be more challenging~\cite{alameldeen06}.
-
-\subsection{Work and Input Size Metrics}
-
-Previous work have proposed profiling-based mechanism to estimate input sizes~\cite{zaparanuks12,coppa14}.
-Coppa~\etal~\cite{coppa14} in particular propose the concept of \textit{read memory size} for automatically estimating the size of the input passed to a routine, where \textit{read memory size} represents the number of distinct memory cells first accessed by a read operation.
-In other words, the \textit{read memory size} metric measures the size of the useful portion of the input's memory footprint.
-However, because we are interested in the amount of computational work performed in respect of a given input, the memory footprint of the input may not always have a direct correspondence to  the amount of computational work.
-
-Goldsmith~\etal~\cite{goldsmith07} use \textit{block frequency} as the measure for performance for empirically describing the asymptotic behaviour of programs, which is known as empirical computational complexity.
-Block frequency is a relative metric that represents the number of times a basic block executes~\cite{ball94,ball96}.
-They argue in favour of block frequency due to its portability, repeatability and exactness, since it does not suffer from timer resolution problems or non-deterministic noises.
-Block frequency also has the advantage of being efficiently profiled by means of automatic code instrumentation~\cite{knuth73,ball94}.
-
-However, in the context of comparing different optimisations, although block frequency would be able to capture aspects of optimisations that simplify the control-flow graph (CFG), measuring work at the basic block resolution would not capture effects of optimisations at the instruction level.
-Because of that, we extend the idea of using basic block frequency to measure computational work by also considering the computational cost of each basic block.
-The computational cost of a basic block is given by weighing the instructions that it contains.
-
-\subsection{Optimal Instrumentation}
-
-In 1973, Donald Knuth~\cite{knuth73} provided an optimal algorithm for profiling block frequency, inserting fewer probes than the previously described naive instrumentation that was commonly used in practice~\cite{knuth71}.
-Forman~\cite{forman81} proposed further overhead reductions by placing the probes in basic blocks that are less likely to be executed based on static heuristics.
-
-Ball and Larus~\cite{ball94} offered a detailed discussion comparing two approaches for optimally profiling block frequency, namely, placing probes on the vertices or the edges of the control-flow graph (CFG).
-They show that the edge-based approach produces optimal placement of probes.
-
-\section{Conclusion and Future Work}
-
-Our proposed technique for relaxed instrumentation can also be applied for similar profiling scenarios, such as profiling empirical computational complexity~\cite{goldsmith07,zaparanuks12,coppa14}.
-
-\section*{Acknowledgments}
-
-This work was supported by the UK Engineering
-and Physical Sciences Research Council (EPSRC) under grants
-EP/L01503X/1 for the University of Edinburgh, School
-of Informatics, Centre for Doctoral Training in Pervasive
-Parallelism, % (\url{http://pervasiveparallelism.inf.ed.ac.uk/}),
-and also by the Institute for Computing Systems Architecture (ICSA)
-in the School of Informatics at the University of Edinburgh.
+%
+%\section*{Acknowledgments}
+%
+%This work was supported by the UK Engineering
+%and Physical Sciences Research Council (EPSRC) under grants
+%EP/L01503X/1 for the University of Edinburgh, School
+%of Informatics, Centre for Doctoral Training in Pervasive
+%Parallelism, % (\url{http://pervasiveparallelism.inf.ed.ac.uk/}),
+%and also by the Institute for Computing Systems Architecture (ICSA)
+%in the School of Informatics at the University of Edinburgh.
 

### 2017-09-06 conclusions.tex
moving staff

\section{Conclusion and Future Work}

Our proposed technique for relaxed instrumentation can also be applied for similar profiling scenarios, such as profiling empirical computational complexity~\cite{goldsmith07,zaparanuks12,coppa14}.

### 2017-09-06 premable.tex
moving staff

\usepackage{booktabs} % For formal tables

\usepackage[utf8]{inputenc}
\usepackage{fullpage}

\usepackage{graphics}

\usepackage{graphicx}
\usepackage{subcaption}

\usepackage{multirow}

%\usepackage{setspace}
\usepackage{ragged2e}

	\usepackage{soul}

%\usepackage{flushend}

%\usepackage{indentfirst}

\usepackage{amsthm}
\usepackage{amsmath,amssymb}
%\usepackage{mathrsfs}
\usepackage{mathtools}

\theoremstyle{definition}
\newtheorem{prop}{Proposition}[section]

%% Listing related stuff

\usepackage{listings}
\usepackage{courier}            %Required for pretty listings (more dense)

%% EMACS colors for listing:

\usepackage{color}
\definecolor{sh_comment}{rgb}{0.65, 0.00, 0.00 } % red-ish
\definecolor{sh_keyword}{rgb}{0.37, 0.69, 0.69}  % blue-green
\definecolor{sh_string}{rgb}{0.08, 0.69, 0.08} % green

%% This manipulates the listing numbering. It decreases it by one at the point used.
\newcommand*\lstDNumber{\addtocounter{lstnumber}{-1}}

%% Change from Listing->Algorithm
%% \renewcommand\lstlistingname{Algorithm} % Fix listing caption name: Listing -> Algorithm
%% \renewcommand\lstlistlistingname{Algorithms}
%% \def\lstlistingautorefname{Alg.}

%% \renewcommand{\thelstlisting}{\thesection-\arabic{lstlisting}}
%\renewcommand{\thelstlisting}{\arabic{lstlisting}} % Fix listing numbering Algorithm 1.1 -> Algorithm 1.

%% \renewcommand{\labelenumi}{\roman{enumi}.}

\lstset{
float=[*],
language=C,                % choose the language of the code
%% basicstyle=\linespread{0.9}
%% basicstyle=\scriptsize\ttfamily,
 basicstyle={\linespread{0.97} \scriptsize\ttfamily},
 stringstyle=\color{sh_string},
 keywordstyle = \color{sh_keyword}\bfseries,
 commentstyle=\color{sh_comment}\itshape,
numbers=left,                   % where to put the line-numbers
%% numberstyle=\scriptsize,        % the size of the fonts that are used for the line-numbers
numberstyle=\scriptsize,        % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
xleftmargin=2em,                % Left margin
frame=lines,                   % adds a frame around the code. none|leftline|topline|bottomline|lines|single|shadowbox
framexleftmargin=1.5em,         % Margin from frame to line numbers
framexbottommargin=0em,         % Distance from last text to frame
morekeywords={in,true,false,and,or,set,is,not},
%% prebreak=\mbox{\tiny$\searrow$},
prebreak=\space,                % Line break: insert this character at the end of the top line
%% postbreak=\mbox{{\color{blue}\scriptsize$\hookrightarrow$}}, % Line break: blue arrow at the beginning of the bottom line
postbreak=\mbox{{\color{blue}\scriptsize$\hookrightarrow$}}, % Line break: blue arrow at the beginning of the bottom line
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
tabsize=2,                      % sets default tabsize to 2 spaces
captionpos=b,                   % sets the caption-position to top
escapeinside={\@}{\@}             % Escape sequence \@ESCAPED STUFF\@
}

\usepackage{nasm/lang}
\usepackage{nasm/style}
\usepackage{llvm/lang}

% Copyright
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
%\acmDOI{10.475/123_4}

% ISBN
%\acmISBN{123-4567-24-567/08/06}

%Conference
\acmConference[CGO'18]{International Symposium on Code Generation and Optimization}{2018}{}
%\acmYear{1997}
%\copyrightyear{2016}
%
%\acmPrice{15.00}

\newcommand{\etal}{et~al.}

\newcommand{\itercomp}{{iterative compilation}}
\newcommand{\Itercomp}{{Iterative compilation}}
\newcommand{\IterComp}{{Iterative Compilation}}
\newcommand{\flagstype}{\usefont{T1}{cmr}{m}{n}}

### 2017-09-06 related.tex
moving staff

\section{Related Work}\label{sec:relatedwork}

\subsection{{\IterComp}}

Chen~\etal~\cite{chen10,chen12a} evaluate the effectiveness of iterative compilations across a large number of input test cases.
Their main motivation is to answer the question:
\textit{How data input dependent is {\itercomp}?}
Their results show that it is possible to find a combination of compiler optimisations that achieves at least 86\% of the maximum speedup across all input test cases.
These optimal combinations are program-specific and yield average speedups up to 3.75$\times$ over the highest optimisation level of compilers.

When optimising a program, the main method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a} evaluates each combination of compiler optimisations across all the available inputs, i.e., if $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimisations, they perform a total of $O(NM)$ runs of the program being optimised.
Furthermore, they use a pre-defined set of only 300 different combinations of compiler optimisations, which represents a very small sample of the optimisation search space for most modern compilers, e.g.
LLVM has 56 distinct optimisation passes and GCC has about 47 high-level (SSA form) optimisation passes and about 25 low-level (RTL) optimisation passes, which in both cases result in much more than $2^{50}$ distinct combinations of compiler optimisations, without considering repetition.

Recent work~\cite{chen12b,fang15} have applied the same idea of performing input-dependent {\itercomp} to distributed applications on data centres.
In summary, each worker receives a subset of the input dataset, called the evaluation dataset, to perform an \textit{online} {\itercomp} of the code being executed.
Each worker performs the same the method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a}, i.e., they evaluate each combination of compiler optimisations across all the evaluation dataset.
However, because the optimisation is performed online, they usually consider a small evaluation dataset and a small number of compiler optimisations.

Fursin~\etal~\cite{fursin07} addressed the problem of comparing the effect of two optimisations on two distinct inputs. For that purpose, they proposed to use instruction per cycle (IPC) as the metric for performing such comparison.
Their result show that using IPC seems promising as a robust metric for {\itercomp} across large input datasets.
However, some specific optimisation techniques may affect the use of IPC as a robust metric, and specially IPC has been shown to provide poor performance estimation for multi-threaded programs~\cite{alameldeen06,eyerman08}.
In particular, IPC can give a skewed performance measure if threads spend time in \textit{spin-lock loops} or other synchronisation mechanisms.
Some existing work on performance assessment suggest that total execution time should be used for measuring performance of multi-threaded programs~\cite{alameldeen06,eyerman08}.
Aalameldeen and Wood~\cite{alameldeen06} in particular suggest that a simple work-related metric should be used if the unit of work is representative enough.
Work-related metrics have already been largely used for measuring performance of throughput-oriented applications, for other applications, however, choosing an appropriate unit of work can be more challenging~\cite{alameldeen06}.

\subsection{Work and Input Size Metrics}

Previous work have proposed profiling-based mechanism to estimate input sizes~\cite{zaparanuks12,coppa14}.
Coppa~\etal~\cite{coppa14} in particular propose the concept of \textit{read memory size} for automatically estimating the size of the input passed to a routine, where \textit{read memory size} represents the number of distinct memory cells first accessed by a read operation.
In other words, the \textit{read memory size} metric measures the size of the useful portion of the input's memory footprint.
However, because we are interested in the amount of computational work performed in respect of a given input, the memory footprint of the input may not always have a direct correspondence to  the amount of computational work.

Goldsmith~\etal~\cite{goldsmith07} use \textit{block frequency} as the measure for performance for empirically describing the asymptotic behaviour of programs, which is known as empirical computational complexity.
Block frequency is a relative metric that represents the number of times a basic block executes~\cite{ball94,ball96}.
They argue in favour of block frequency due to its portability, repeatability and exactness, since it does not suffer from timer resolution problems or non-deterministic noises.
Block frequency also has the advantage of being efficiently profiled by means of automatic code instrumentation~\cite{knuth73,ball94}.

However, in the context of comparing different optimisations, although block frequency would be able to capture aspects of optimisations that simplify the control-flow graph (CFG), measuring work at the basic block resolution would not capture effects of optimisations at the instruction level.
Because of that, we extend the idea of using basic block frequency to measure computational work by also considering the computational cost of each basic block.
The computational cost of a basic block is given by weighing the instructions that it contains.

\subsection{Optimal Instrumentation}

In 1973, Donald Knuth~\cite{knuth73} provided an optimal algorithm for profiling block frequency, inserting fewer probes than the previously described naive instrumentation that was commonly used in practice~\cite{knuth71}.
Forman~\cite{forman81} proposed further overhead reductions by placing the probes in basic blocks that are less likely to be executed based on static heuristics.

Ball and Larus~\cite{ball94} offered a detailed discussion comparing two approaches for optimally profiling block frequency, namely, placing probes on the vertices or the edges of the control-flow graph (CFG).
They show that the edge-based approach produces optimal placement of probes.

### 2017-09-06 cgo18mscr.tex
moving files

@@
 
-\section{Experimental Evaluation}
+\input{setup}
 
-In this section we discuss our experimental evaluation.
-First we describe the benchmarks with datasets used in the experiments.
-Afterwards, we discuss our results concerning the instrumentations for profiling our work metric.
-Finally we present the results of the online {\itercomp}.
-
-We implemented the instrumentations in LLVM 4.0.
-The target platform is a Linux-4.4.27 system with an Intel Core i7-4770 3.40GHz Skylake~CPU with 16~GiB RAM.
-
-\subsection{Benchmarks}
-
-For the experimental evaluation we have used a subset of the \textit{KDataSets} benchmark suit, which is the same benchmark and dataset suit used by Chen~\etal~\cite{chen10,chen12a}.
-The KDataSets contains 1000 different inputs for each one of its benchmark programs.
-These benchmarks cover a broad spectrum of application scenarios, ranging from simple embedded signal-processing tasks to common mobile-phone and desktop tasks.
-The different inputs try to capture distinct characteristics in terms of workload sizes and how these workloads exercise different control flow paths.
-A summary of the benchmarks and dataset suit is shown in Table~\ref{tab:kdatasets}.
-
-\begin{table}[h]
-\centering
-\scalebox{.8}{
-\begin{tabular}{|c|c|c|c|}
-\hline
-\textbf{Program} & \textbf{LOC}    & \textbf{Input file size}            & \textbf{Input description}              \\ \hline % Domain
-qsort         & 154    & 32K-1.8M                   & 3D coordinates                 \\ \hline
-jpeg\_d       & 13501  & 3.6K-1.5M                  & JPEG images                    \\ \hline
-jpeg\_c       & 14014  & 16K-137M                   & PPM images                     \\ \hline
-tiff2bw       & 15477  & \multirow{4}{*}{9K-137M}   & \multirow{4}{*}{TIFF images}   \\ \cline{1-2}
-tiff2rgba     & 15424  &                            &                                \\ \cline{1-2}
-tiffdither    & 15399  &                            &                                \\ \cline{1-2}
-tiffmedian    & 15870  &                            &                                \\ \hline
-susan\_c      & 1376   & \multirow{3}{*}{12K-46M}   & \multirow{3}{*}{PGM images}    \\ \cline{1-2}
-susan\_e      & 1376   &                            &                                \\ \cline{1-2}
-susan\_s      & 1376   &                            &                                \\ \hline
-adpcm\_c      & 210    & 167K-36M                   & WAVE audios                    \\ \hline
-adpcm\_d      & 211    & 21K-8.8M                   & ADPCM audios                   \\ \hline
-lame          & 14491  & 167K-36M                   & WAVE audios                    \\ \hline
-rsynth        & 4111   & 0.1K-42M                   &  Text files                    \\ \hline
-sha           & 197    & 0.6K-35M                   & Files of any format            \\ \hline
-\rowcolor{gray!20}
-bitcount      & 460    &  -                         & Numbers: random                \\ \hline
-\rowcolor{gray!20}
-dijkstra      & 163    & 0.06K-4.3M                 & Adjacency matrices             \\ \hline
-\rowcolor{gray!20}
-patricia      & 290    & 0.6K-1.9M                  & IP and mask pairs              \\ \hline
-\rowcolor{gray!20}
-mad           & 2358   & 28K-27M                    & MP3 audios                     \\ \hline
-\rowcolor{gray!20}
-%gsm           & 3806   & 83K-18M                    & Sun/NeXT audios                \\ \hline
-ghostscript   & 99869  & 11K-43M                    & Postscript files               \\ \hline
-\rowcolor{gray!20}
-%ispell        & 6522   & \multirow{3}{*}{0.1K-42M}  & \multirow{3}{*}{Text files}    \\ \cline{1-2}
-%rsynth        & 4111   &                            &                                \\ \cline{1-2}
-stringsearch  & 338    &  0.1K-42M                 &  Text files                     \\ \hline
-\rowcolor{gray!20}
-%blowfish\_e   & 863    & 0.6K-35M                   & Files of any format            \\ \hline
-%blowfish\_d   & 863    & 0.6K-35M                   & Encrypted files                \\ \hline
-%pgp\_e        & 19575  & 0.6K-35M                   & Files of any format            \\ \hline
-%pgp\_d        & 19575  & 0.4K-18M                   & Encrypted files                \\ \hline
-%rijndael\_e   & 952    & 0.6K-35M                   & Files of any format            \\ \hline
-%rijndael\_d   & 952    & 0.7K-35M                   & Encrypted files                \\ \hline
-CRC32         & 130    & 0.6K-35M                   & Files of any format            \\ \hline
-\rowcolor{gray!20}
-bzip2e        & 5125   & 0.7K-57M                   & Files of any format            \\ \hline
-\rowcolor{gray!20}
-bzip2d        & 5125   & 0.2K-25M                   & Compressed files               \\ \hline
-\end{tabular}
-}
-\caption{Description of the KDataSets with 1000 inputs for each benchmark (Chen~\etal~\cite{chen10,chen12a}).}
-\label{tab:kdatasets}
-\end{table}
-
-The shaded (grey) benchmarks in Table~\ref{tab:kdatasets} represent the benchmarks used for training the cost model used for computing the weight of the instructions for the work metric.
-These same training benchmarks were also used for collecting a fixed set of optimisations for the {\itercomp}.
-The remaining (white) benchmarks are used for the experimental evaluation.
-
-\subsection{Evaluation of the Instrumentation}
-
-We first compare static aspects of the instrumentation algorithms.
-The naive instrumentation always has 100\% of the basic blocks instrumented, by definition.
-First, we evaluate the optimal work profiling compared to the relaxation algorithm applied per DAG.
-Afterwards, we compare these profiling techniques with the whole program relaxation.
-
-Figure~\ref{fig:num-probes} shows the percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.
-While Figure~\ref{fig:num-probes} compares the various instrumentation algorithms in respect of the naive instrumentation, Figure~\ref{fig:num-probes-improvement} shows the improvement of the relaxation algorithm  over the optimal instrumentation.
-
-\begin{figure*}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{figs/num-probes.pdf}
-    \caption{Percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.}
-    \label{fig:num-probes}
-\end{figure*}
-
-For the evaluation of the performance overhead that the instrumentation incur to the benchmarks, we measure the wall-clock time of the benchmarks when compiled with the default {\flagstype -O3} optimisation.
-For each benchmark, we compute the average overhead over all its 1000 input dataset.
-When measuring the wall-clock time for each input, in order to reduce noise, we execute the same input until we have a statistically sound measurement, i.e.
-we execute until we have an interval no larger than 1\% with 99\% confidence.
-Figure~\ref{fig:overhead-O3} shows the performance overhead imposed by the work instrumentation on the benchmarks when compared to their non-instrumented counterparts.
-Figure~\ref{fig:overhead-improvement-O3} indicates the improvement of the relaxation algorithm over the optimal instrumentation, regarding the reduction in overhead.
-
-\begin{figure*}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{figs/overhead-O3.pdf}
-    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.}
-    \label{fig:overhead-O3}
-\end{figure*}
-
-The WPO relaxation was performed using profiling information for every specific input, which provides the perfect information to perform the WPO relaxation.
-This means that the experiments show its best performance enabled by having perfect profiling information.
-
-%\begin{figure*}[ht]
-%    \centering
-%    \includegraphics[width=\textwidth]{figs/overhead-improvement-O3.pdf}
-%    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.
-%             Notice that the average of the ratios does not equal the ratio of the averages.}
-%    \label{fig:overhead-improvement-O3}
-%\end{figure*}
-
-\begin{figure*}[h!]
-    \centering
-    \includegraphics[width=\textwidth]{figs/error-O3.pdf}
-    \caption{Dynamic error of the work profiling averaged over the 1000 inputs, after relaxing the number of probes.}
-    \label{fig:error-O3}
-\end{figure*}
-
-%Figure~\ref{fig:overhead-improvement-O3} shows that the relaxation algorithm, with 5\% threshold, is able to improve an average of 48\% over the optimal instrumentation,
-%and the WPO relaxation achieves an average of $2\times$ improvement over the optimal instrumentation.
-%While the optimal profiling has a maximum overhead of almost 60\%, the work profiling with a relaxation of 5\% threshold has an overhead of less than 20\% for all benchmarks,
-%with the WPO relaxation reaching an average overhead of only 5\%, compared to an average of 12\% for the optimal profiling.
-
-Furthermore, these improvements can be obtained while incurring very little dynamic error, as shown in Figure~\ref{fig:error-O3}.
-This result confirms the overly conservative aspect of the relaxation algorithm applied per DAG,
-while the WPO relaxation is significantly less conservative.
-Because we performed the WPO relaxation having perfect profiling information, its dynamic error was always below the 5\% thresholds.
-Although it is not guaranteed by the algorithm, profiling representative inputs should also result in small dynamic errors.
-
-\subsection{Evaluation of the Online {\IterComp}}
-
-For comparison, we use four configurations\footnote{A configuration using the WPO relaxation was not used due to the time necessary for executing the experiments.}.
-In all configurations, the same optimisation sequence is used for multiple inputs, using a dynamic input-window size, as explained in Section~\ref{sec:oic-infra}.
-The average performance over the input window provides an estimate for the overall performance of the optimisation sequence across distinct inputs.
-The optimisation sequences are ranked based on their average performance, and the best optimisation sequence is selected.
-%When selecting the best optimisation sequence, they are ranked by the average of their performance measurement, using the WP metric, except for the Oracle-RM which uses the real speedup over \texttt{-O0}.
-\begin{itemize}
-\item \textbf{Oracle-RM} executes the program multiple times for each input, measuring the real speedup for each optimisation sequence, and then uses the real speedup over {\flagstype -O0} for comparing the optimisation sequences.
-  The speedups are computed based on the wall-clock time.
-  In order to reduce noise, the program is executed several times for the same input, until the confidence interval was no larger than 1\% for a 99\% confidence.
-\item \textbf{Oracle-PP} represents an oracle with a \textit{perfect} non-intrusive profiling.
-  Although it uses the WP metric for comparing optimisation sequences, this oracle also avoids noise in its measurements by also executing the program multiple times for each input.
-  The first execution is used for profiling the work metric.
-  The remaining executions are used for measuring the wall-clock time without using the work profiling.
-\item \textbf{Real-OP} corresponds to the online {\itercomp} as it would be applied in a real online scenario.
-  For each optimisation, a random sample of inputs is selected, and the program is executed only once with each input.
-  When executing each input, it uses the optimal work instrumentation for profiling the work metric.
-  The average of the WP for the sample of inputs is then used for selecting the best optimisation.
-\item \textbf{Real-R5} is similar to the {Real-OP}.
-  It also corresponds to the online {\itercomp} as it would be applied in a real online scenario.
-  For each optimisation, a random sample of inputs is selected, and the program is executed only once with each input.
-  When executing each input, it uses the relaxed work instrumentation, with a 5\% threshold, for profiling the work metric.
-  The average of the WP for the sample of inputs is then used for selecting the best optimisation.
-\item \textbf{Real-IPC} corresponds to the online {\itercomp} based on the IPC metric.
-  For each optimisation, a random sample of inputs is selected, and the program is executed only once with each input.
-  The average of the IPC metric for the sample of inputs is then used for selecting the best optimisation.
-\end{itemize}
-The comparison between Oracle-RM and Oracle-PP is useful for validating the use of the WP metric for guiding {\itercomp}, while the other configurations demonstrate the viability of applying the online {\itercomp} in a real scenario.
-
-Figure~\ref{fig:window-size} shows the average input-window size for each benchmark and configuration.
-It shows that we can have a statistically sound measurement of the performance metric using just a small number of inputs.
-
-\begin{figure*}[htb]
-    \centering
-    \includegraphics[width=\textwidth]{figs/window-size.pdf}
-    \caption{Average input-window sizes observed during the online {\itercomp}.}
-    \label{fig:window-size}
-\end{figure*}
-
-\subsection{The Set of Optimisation Sequences}
-
-For the purpose of evaluating the use of the WP metric with {\itercomp}, we collected in advance a fixed set of optimisation sequences.
-The reason for using this fixed set, as explained in Section~\ref{sec:oic-infra}, is because this paper is focused on other components of the infrastructure for performing {\itercomp} and we assume that a good generator of optimisation sequences will be used in a real online scenario.
-This set contains 500 optimisation sequences collected in a random search using the training benchmarks.
-These optimisation sequences contain an average of 40 individual optimisation passes, including repetitions, with a maximum of 119 optimisation passes, but it also contains optimisation sequences which consist of a single flag, such as the default optimisations {\flagstype -O1}, {\flagstype -O2}, {\flagstype -O3}, {\flagstype -Os}, and {\flagstype -Oz}.
-
-  % \begin{minipage}{0.9\textwidth}
-  %    \vspace{1em}
-  %    \noindent\textbf{Example of a short optimisation sequence:}\vspace{-1ex}
-  %    \justify{\flagstype -mem2reg -simplifycfg -constprop -dce}
-  % \end{minipage}
-  %
-  % \begin{minipage}{0.9\textwidth}
-  %    \vspace{1em}
-  %    \noindent\textbf{Example of a long optimisation sequence:}\vspace{-1ex}
-  %    \justify{\flagstype -globalopt -reassociate -instcombine -loop-rotate -block-freq -deadargelim -early-cse -sroa -argpromotion -sccp -tbaa -barrier -constmerge \mbox{-loop-vectorize} -domtree -basicaa -memdep -basiccg -memcpyopt \mbox{-constprop} -adce -globaldce -mem2reg -constmerge \mbox{-globaldce} -constprop -instsimplify -dse -dce -simplifycfg -loop-unroll -reassociate -constprop \mbox{-globaldce} -instsimplify -adce -constmerge -bb-vectorize -dce -mergefunc -simplifycfg -dse -loop-unroll -globaldce}
-  % \end{minipage}
-  %
-  % \begin{minipage}{0.9\textwidth}
-  %    \vspace{1em}
-  %    \noindent\textbf{Example of an optimisation sequence which includes {-O3}:}\vspace{-1ex}
-  %    \justify{\flagstype -O3 -adce -globaldce -simplifycfg -memcpyopt -reassociate -mergefunc \mbox{-dce} -dse}
-  %    \vspace{2em}
-  % \end{minipage}
-
-Repeating the same optimisation pass can be beneficial and usually expected by other passes.
-For example, the {\flagstype -loop-simplify} pass is used for transforming loops into a canonical form by inserting pre-header and exit basic blocks.
-Although this pass inserts jumps due to redundant basic blocks, this canonical form can be favourable to other loop optimisations.
-Because of the redundant basic blocks, this optimisation pass expects that the {\flagstype -simplifycfg} will eventually be executor later on the optimisation pipeline.
-Another example of such inter-relation between transformations concerns the {\flagstype -licm} and {\flagstype -mem2reg} passes.
-The {\flagstype -licm} pass is responsible for moving invariant code out from the loop body.
-It usually creates new local variables, using memory access operations, for assisting with the code manipulation, which means that the executing the {\flagstype -mem2reg} pass afterwards would be useful as a cleanup pass for removing the extra memory accesses generated.
-However, many of the analysis required for identifying loop invariant also benefit from the transformations performed by the {\flagstype -mem2reg} pass.
-These examples illustrate the importance of repeating optimisation passes.
-Moreover, they illustrate the intricate relation amongst several transformations.
-
-However, all optimisation sequences in the set of optimisations were generated completely at random, without using any knowledge of individual transformations.
-Each optimisation sequence was generated in two steps: \textit{(1)} randomly selects the number of flags; \textit{(2)} randomly selects the flags, allowing repetitions.
-Afterwards, this randomly generated optimisation sequence would be included in the set of optimisation sequences only if it was able to improve the performance of a training benchmark, also selected at random, in respect of the {\flagstype -O3} default optimisation.
-This process was repeated until we obtained all the 500 distinct optimisation sequences.
-
-\subsection{Performance Evaluation}
-
-In order to evaluate the quality of the final optimisation sequences selected by the {\itercomp} search, we compare their speedup by measuring wall-clock time of the benchmarks when compiled with the standard {\flagstype -O3} optimisation.
-For each benchmark, after selecting the final optimisation sequence, we compute the average speedup over all the 1000 input dataset.
-When measuring the wall-clock time for each input, to reduce noise, we execute the same input until we have a statistically sound measurement, i.e.
-we execute until we have an interval no larger than 1\% with 99\% confidence.
-Figure~\ref{fig:speedups} shows these average speedups over all the 1000 input dataset.
-This figure shows that the best optimisation sequence selected with the Oracle-PP is very close to the performance of the best optimisation sequence selected with the Oracle-RM.
-This result is important for demonstrating that a work-based metric has the potential to produce good results in a real online scenario, where there is the restriction that programs execute distinct inputs only once.
-Moreover, the evaluation also indicates that the use of relaxation algorithms does not degrade the optimisation search,
-it might even benefit the {\itercomp} as it tends to insert smaller interferences in the performance measurement.
-Finally, Figure~\ref{fig:speedups} also presents the results of {\itercomp} guided by IPC, which is futher discussed in Figure~\ref{fig:ipc-vs-work}.
-
-\begin{figure*}[htb]
-    \centering
-    \includegraphics[width=\textwidth]{figs/speedups.pdf}
-    \caption{Speedups obtained from the final optimisation sequence selected by the online {\itercomp}.
-	         The speedups reported for each benchmark represents the average speedup across their complete 1000 input datasets.}
-    \label{fig:speedups}
-\end{figure*}
-
-\begin{figure*}[h]
-    \centering
-    \includegraphics[width=\textwidth]{figs/ipc-vs-work.pdf}
-    \caption{Comparison between {\itercomp} using IPC or the WP metric.}
-    \label{fig:ipc-vs-work}
-\end{figure*}
-
-Figure~\ref{fig:ipc-vs-work} presents a comparison between IPC and the WP metric regarding their correlation with the actual speedup over \verb|-O0| observed during {\itercomp}.
-Each point in the figure corresponds to each metric averaged over input windows collected during multiple executions of the {\itercomp}.
-The vertical lines correspond to the optimisation sequence selected by the {\itercomp} search, which either represents the highest IPC or the highest value of WP.
-In all cases, the optimisation sequence selected based on the WP is much closer to the best speedup than the optimisation sequence selected based on IPC, which can be attributed to the fact that the WP has a stronger correlation with the speedup.
-In all cases, IPC shows little to no correlation with the speedup, which corroborates the argument given in Section~\ref{sec:ipc-vs-work-metric}.
-
-For some cases, namely for the tiff-related benchmarks, the WP shows a worse correlation with speedup.
-We believe these cases could be improved by improving the cost model used to compute the work metric.
-However, even in these cases, the optimisation sequences selected based on the WP outperform those selected based on the IPC metric.
-
-\textbf{ToDo: Include results using Genetic Algorithm search.}
+\input{results}
 

### 2017-09-06 results.tex
moving files

\section{Experimental Results}

In this section we discuss our experimental evaluation.
First we describe the benchmarks with datasets used in the experiments.
Afterwards, we discuss our results concerning the instrumentations for profiling our work metric.
Finally we present the results of the online {\itercomp}.

\subsection{Evaluation of the Instrumentation}

We first compare static aspects of the instrumentation algorithms.
The naive instrumentation always has 100\% of the basic blocks instrumented, by definition.
First, we evaluate the optimal work profiling compared to the relaxation algorithm applied per DAG.
Afterwards, we compare these profiling techniques with the whole program relaxation.

Figure~\ref{fig:num-probes} shows the percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.
While Figure~\ref{fig:num-probes} compares the various instrumentation algorithms in respect of the naive instrumentation, Figure~\ref{fig:num-probes-improvement} shows the improvement of the relaxation algorithm  over the optimal instrumentation.

\begin{figure*}[ht]
    \centering
    \includegraphics[width=\textwidth]{figs/num-probes.pdf}
    \caption{Percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.}
    \label{fig:num-probes}
\end{figure*}

For the evaluation of the performance overhead that the instrumentation incur to the benchmarks, we measure the wall-clock time of the benchmarks when compiled with the default {\flagstype -O3} optimisation.
For each benchmark, we compute the average overhead over all its 1000 input dataset.
When measuring the wall-clock time for each input, in order to reduce noise, we execute the same input until we have a statistically sound measurement, i.e. we execute until we have an interval no larger than 1\% with 99\% confidence.
Figure~\ref{fig:overhead-O3} shows the performance overhead imposed by the work instrumentation on the benchmarks when compared to their non-instrumented counterparts.
Figure~\ref{fig:overhead-improvement-O3} indicates the improvement of the relaxation algorithm over the optimal instrumentation, regarding the reduction in overhead.

\begin{figure*}[ht]
    \centering
    \includegraphics[width=\textwidth]{figs/overhead-O3.pdf}
    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.}
    \label{fig:overhead-O3}
\end{figure*}

The WPO relaxation was performed using profiling information for every specific input, which provides the perfect information to perform the WPO relaxation.
This means that the experiments show its best performance enabled by having perfect profiling information.

%\begin{figure*}[ht]
%    \centering
%    \includegraphics[width=\textwidth]{figs/overhead-improvement-O3.pdf}
%    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.
%             Notice that the average of the ratios does not equal the ratio of the averages.}
%    \label{fig:overhead-improvement-O3}
%\end{figure*}

\begin{figure*}[h!]
    \centering
    \includegraphics[width=\textwidth]{figs/error-O3.pdf}
    \caption{Dynamic error of the work profiling averaged over the 1000 inputs, after relaxing the number of probes.}
    \label{fig:error-O3}
\end{figure*}

%Figure~\ref{fig:overhead-improvement-O3} shows that the relaxation algorithm, with 5\% threshold, is able to improve an average of 48\% over the optimal instrumentation,
%and the WPO relaxation achieves an average of $2\times$ improvement over the optimal instrumentation.
%While the optimal profiling has a maximum overhead of almost 60\%, the work profiling with a relaxation of 5\% threshold has an overhead of less than 20\% for all benchmarks,
%with the WPO relaxation reaching an average overhead of only 5\%, compared to an average of 12\% for the optimal profiling.

Furthermore, these improvements can be obtained while incurring very little dynamic error, as shown in Figure~\ref{fig:error-O3}.
This result confirms the overly conservative aspect of the relaxation algorithm applied per DAG,
while the WPO relaxation is significantly less conservative.
Because we performed the WPO relaxation having perfect profiling information, its dynamic error was always below the 5\% thresholds.
Although it is not guaranteed by the algorithm, profiling representative inputs should also result in small dynamic errors.

\subsection{Evaluation of the Online {\IterComp}}

For comparison, we use four configurations\footnote{A configuration using the WPO relaxation was not used due to the time necessary for executing the experiments.}.
In all configurations, the same optimisation sequence is used for multiple inputs, using a dynamic input-window size, as explained in Section~\ref{sec:oic-infra}.
The average performance over the input window provides an estimate for the overall performance of the optimisation sequence across distinct inputs.
The optimisation sequences are ranked based on their average performance, and the best optimisation sequence is selected.
%When selecting the best optimisation sequence, they are ranked by the average of their performance measurement, using the WP metric, except for the Oracle-RM which uses the real speedup over \texttt{-O0}.
\begin{itemize}
\item \textbf{Oracle-RM} executes the program multiple times for each input, measuring the real speedup for each optimisation sequence, and then uses the real speedup over {\flagstype -O0} for comparing the optimisation sequences.
  The speedups are computed based on the wall-clock time.
  In order to reduce noise, the program is executed several times for the same input, until the confidence interval was no larger than 1\% for a 99\% confidence.
\item \textbf{Oracle-PP} represents an oracle with a \textit{perfect} non-intrusive profiling.
  Although it uses the WP metric for comparing optimisation sequences, this oracle also avoids noise in its measurements by also executing the program multiple times for each input.
  The first execution is used for profiling the work metric.
  The remaining executions are used for measuring the wall-clock time without using the work profiling.
\item \textbf{Real-OP} corresponds to the online {\itercomp} as it would be applied in a real online scenario.
  For each optimisation, a random sample of inputs is selected, and the program is executed only once with each input.
  When executing each input, it uses the optimal work instrumentation for profiling the work metric.
  The average of the WP for the sample of inputs is then used for selecting the best optimisation.
\item \textbf{Real-R5} is similar to the {Real-OP}.
  It also corresponds to the online {\itercomp} as it would be applied in a real online scenario.
  For each optimisation, a random sample of inputs is selected, and the program is executed only once with each input.
  When executing each input, it uses the relaxed work instrumentation, with a 5\% threshold, for profiling the work metric.
  The average of the WP for the sample of inputs is then used for selecting the best optimisation.
\item \textbf{Real-IPC} corresponds to the online {\itercomp} based on the IPC metric.
  For each optimisation, a random sample of inputs is selected, and the program is executed only once with each input.
  The average of the IPC metric for the sample of inputs is then used for selecting the best optimisation.
\end{itemize}
The comparison between Oracle-RM and Oracle-PP is useful for validating the use of the WP metric for guiding {\itercomp}, while the other configurations demonstrate the viability of applying the online {\itercomp} in a real scenario.

Figure~\ref{fig:window-size} shows the average input-window size for each benchmark and configuration.
It shows that we can have a statistically sound measurement of the performance metric using just a small number of inputs.

\begin{figure*}[htb]
    \centering
    \includegraphics[width=\textwidth]{figs/window-size.pdf}
    \caption{Average input-window sizes observed during the online {\itercomp}.}
    \label{fig:window-size}
\end{figure*}

\subsection{The Set of Optimisation Sequences}

For the purpose of evaluating the use of the WP metric with {\itercomp}, we collected in advance a fixed set of optimisation sequences.
The reason for using this fixed set, as explained in Section~\ref{sec:oic-infra}, is because this paper is focused on other components of the infrastructure for performing {\itercomp} and we assume that a good generator of optimisation sequences will be used in a real online scenario.
This set contains 500 optimisation sequences collected in a random search using the training benchmarks.
These optimisation sequences contain an average of 40 individual optimisation passes, including repetitions, with a maximum of 119 optimisation passes, but it also contains optimisation sequences which consist of a single flag, such as the default optimisations {\flagstype -O1}, {\flagstype -O2}, {\flagstype -O3}, {\flagstype -Os}, and {\flagstype -Oz}.

  % \begin{minipage}{0.9\textwidth}
  %    \vspace{1em}
  %    \noindent\textbf{Example of a short optimisation sequence:}\vspace{-1ex}
  %    \justify{\flagstype -mem2reg -simplifycfg -constprop -dce}
  % \end{minipage}
  %
  % \begin{minipage}{0.9\textwidth}
  %    \vspace{1em}
  %    \noindent\textbf{Example of a long optimisation sequence:}\vspace{-1ex}
  %    \justify{\flagstype -globalopt -reassociate -instcombine -loop-rotate -block-freq -deadargelim -early-cse -sroa -argpromotion -sccp -tbaa -barrier -constmerge \mbox{-loop-vectorize} -domtree -basicaa -memdep -basiccg -memcpyopt \mbox{-constprop} -adce -globaldce -mem2reg -constmerge \mbox{-globaldce} -constprop -instsimplify -dse -dce -simplifycfg -loop-unroll -reassociate -constprop \mbox{-globaldce} -instsimplify -adce -constmerge -bb-vectorize -dce -mergefunc -simplifycfg -dse -loop-unroll -globaldce}
  % \end{minipage}
  %
  % \begin{minipage}{0.9\textwidth}
  %    \vspace{1em}
  %    \noindent\textbf{Example of an optimisation sequence which includes {-O3}:}\vspace{-1ex}
  %    \justify{\flagstype -O3 -adce -globaldce -simplifycfg -memcpyopt -reassociate -mergefunc \mbox{-dce} -dse}
  %    \vspace{2em}
  % \end{minipage}

Repeating the same optimisation pass can be beneficial and usually expected by other passes.
For example, the {\flagstype -loop-simplify} pass is used for transforming loops into a canonical form by inserting pre-header and exit basic blocks.
Although this pass inserts jumps due to redundant basic blocks, this canonical form can be favourable to other loop optimisations.
Because of the redundant basic blocks, this optimisation pass expects that the {\flagstype -simplifycfg} will eventually be executor later on the optimisation pipeline.
Another example of such inter-relation between transformations concerns the {\flagstype -licm} and {\flagstype -mem2reg} passes.
The {\flagstype -licm} pass is responsible for moving invariant code out from the loop body.
It usually creates new local variables, using memory access operations, for assisting with the code manipulation, which means that the executing the {\flagstype -mem2reg} pass afterwards would be useful as a cleanup pass for removing the extra memory accesses generated.
However, many of the analysis required for identifying loop invariant also benefit from the transformations performed by the {\flagstype -mem2reg} pass.
These examples illustrate the importance of repeating optimisation passes.
Moreover, they illustrate the intricate relation amongst several transformations.

However, all optimisation sequences in the set of optimisations were generated completely at random, without using any knowledge of individual transformations.
Each optimisation sequence was generated in two steps: \textit{(1)} randomly selects the number of flags; \textit{(2)} randomly selects the flags, allowing repetitions.
Afterwards, this randomly generated optimisation sequence would be included in the set of optimisation sequences only if it was able to improve the performance of a training benchmark, also selected at random, in respect of the {\flagstype -O3} default optimisation.
This process was repeated until we obtained all the 500 distinct optimisation sequences.

\subsection{Performance Evaluation}

In order to evaluate the quality of the final optimisation sequences selected by the {\itercomp} search, we compare their speedup by measuring wall-clock time of the benchmarks when compiled with the standard {\flagstype -O3} optimisation.
For each benchmark, after selecting the final optimisation sequence, we compute the average speedup over all the 1000 input dataset.
When measuring the wall-clock time for each input, to reduce noise, we execute the same input until we have a statistically sound measurement, i.e. we execute until we have an interval no larger than 1\% with 99\% confidence.
Figure~\ref{fig:speedups} shows these average speedups over all the 1000 input dataset.
This figure shows that the best optimisation sequence selected with the Oracle-PP is very close to the performance of the best optimisation sequence selected with the Oracle-RM.
This result is important for demonstrating that a work-based metric has the potential to produce good results in a real online scenario, where there is the restriction that programs execute distinct inputs only once.
Moreover, the evaluation also indicates that the use of relaxation algorithms does not degrade the optimisation search,
it might even benefit the {\itercomp} as it tends to insert smaller interferences in the performance measurement.
Finally, Figure~\ref{fig:speedups} also presents the results of {\itercomp} guided by IPC, which is futher discussed in Figure~\ref{fig:ipc-vs-work}.

\begin{figure*}[htb]
    \centering
    \includegraphics[width=\textwidth]{figs/speedups.pdf}
    \caption{Speedups obtained from the final optimisation sequence selected by the online {\itercomp}.
	         The speedups reported for each benchmark represents the average speedup across their complete 1000 input datasets.}
    \label{fig:speedups}
\end{figure*}

\begin{figure*}[h]
    \centering
    \includegraphics[width=\textwidth]{figs/ipc-vs-work.pdf}
    \caption{Comparison between {\itercomp} using IPC or the WP metric.}
    \label{fig:ipc-vs-work}
\end{figure*}

Figure~\ref{fig:ipc-vs-work} presents a comparison between IPC and the WP metric regarding their correlation with the actual speedup over \verb|-O0| observed during {\itercomp}.
Each point in the figure corresponds to each metric averaged over input windows collected during multiple executions of the {\itercomp}.
The vertical lines correspond to the optimisation sequence selected by the {\itercomp} search, which either represents the highest IPC or the highest value of WP.
In all cases, the optimisation sequence selected based on the WP is much closer to the best speedup than the optimisation sequence selected based on IPC, which can be attributed to the fact that the WP has a stronger correlation with the speedup.
In all cases, IPC shows little to no correlation with the speedup, which corroborates the argument given in Section~\ref{sec:ipc-vs-work-metric}.

For some cases, namely for the tiff-related benchmarks, the WP shows a worse correlation with speedup.
We believe these cases could be improved by improving the cost model used to compute the work metric.
However, even in these cases, the optimisation sequences selected based on the WP outperform those selected based on the IPC metric.

\textbf{ToDo: Include results using Genetic Algorithm search.}

### 2017-09-06 abstract.tex
updates intro

@@
 online {\itercomp}, which is able to achieve about 80\% of the performance
-of the offline oracle, with improvements of up to about 20\%.
+of an offline oracle, with improvements of up to 20\%.
 

### 2017-09-07 motivation.tex
motivation and overview

\section{Motivation}

As a motivation example, consider...

Here we give an example to motivate our work. E.g. assuming we have a zero-cost profiling based work metric, 
how can we use it to perform iterative compilation across runs? Then we talk about while it is impossible to 
build a zero-overhead metric, we can develop one with low overhead.

### 2017-09-08 cgo18mscr.tex
updates intro
motivation and overview
motivation figure
Merge remote-tracking branch 'origin/master'
# Conflicts:
#	cgo18mscr.tex
#	motivation.tex
updates motivation

@@
 \begin{document}
-\title{Online {\IterComp}\\Guided by Low Overhead Profiling}
+\title{Online {\IterComp}}
 
@@
 %\ccsdesc[500]{Software and its engineering~Retargetable compilers}
-\keywords{{\IterComp}, Profiling, Relaxed Instrumentation}
+\keywords{{\IterComp}, Profiling} %, Relaxed Instrumentation}
 
@@
 \input{intro}
+\input{motivation}
 
@@
 
-In this section we define the work-based performance (WP) metric proposed for comparing different optimised versions of a program when executing with different inputs.
+In this section we define the work-based performance (WP) metric proposed for comparing different optimized versions of a program when executing with different inputs.
 We define the performance metric as the ratio between the amount of \textit{work}, $\Delta W$, performed during a period of time, $\Delta t$.
@@
 
-By measuring the amount of \textit{work} done per unit of time, we reduce the impact of input-dependent aspects and focus instead on the efficiency of the optimised program.
+By measuring the amount of \textit{work} done per unit of time, we reduce the impact of input-dependent aspects and focus instead on the efficiency of the optimized program.
 For this metric, the main challenge is to precisely define what represents \textit{work}.
 As discussed in Section~\ref{sec:relatedwork}, many work metrics have been presented in the literature, such as block frequency and others.
-However, for comparing the performance benefits of different compiler optimisations on a program, although block frequency would be able to capture aspects of optimisations that simplify the control-flow graph (CFG), we argue that measuring work only at the basic block resolution would not be enough for capturing the effects of optimisations at the instruction level.
+However, for comparing the performance benefits of different compiler optimizations on a program, although block frequency would be able to capture aspects of optimizations that simplify the control-flow graph (CFG), we argue that measuring work only at the basic block resolution would not be enough for capturing the effects of optimizations at the instruction level.
 For this reason, we extend the idea of using basic block frequency to measure computational work by also considering the computational cost of each basic block.
@@
 In order to fit the linear model, we measure the wall-clock time when executing the training benchmarks described in Section\ref{sec:benchmarks} with their respective 1000 input datasets.
-For these measurements, the training benchmarks are compiled without optimisation.
-The reason for using no optimisation, as discussed in Section\ref{sec:oic-infra}, is because the amount of work must be only input dependent and consistent between different optimisation sequences.
-This aspect is crucial for the work-based performance metric to enable a consistent comparison between the various optimisation sequences when executing with distinct inputs.
-%it allows to use the estimated cost model for computing a work metric which is independent of optimisations, as we explained in Section~\ref{}.
+For these measurements, the training benchmarks are compiled without optimization.
+The reason for using no optimization, as discussed in Section\ref{sec:oic-infra}, is because the amount of work must be only input dependent and consistent between different optimization sequences.
+This aspect is crucial for the work-based performance metric to enable a consistent comparison between the various optimization sequences when executing with distinct inputs.
+%it allows to use the estimated cost model for computing a work metric which is independent of optimizations, as we explained in Section~\ref{}.
 
-Our cost model has a total of 52 LLVM instructions\footnote{We do not model all LLVM instructions because some instructions are more common in optimised programs, such as the vector instructions.}.
+Our cost model has a total of 52 LLVM instructions\footnote{We do not model all LLVM instructions because some instructions are more common in optimized programs, such as the vector instructions.}.
 Every program, in the set of training benchmarks, was compiled twice: once including the necessary instrumentation for the block frequency profiling, which is required for deriving the linear expression defined in Equation~\ref{eq:linear-work-expression};
-and a standard compilation without any optimisation or instrumentation.
+and a standard compilation without any optimization or instrumentation.
 For each input in the training dataset, the benchmarks were executed once with the instrumented version for collecting the block frequency profiling, and multiple times with the standard compilation just for measuring the wall-clock execution time, until the confidence interval was no larger than 1\% for a 99\% confidence.
 After collecting these measurements, this data can be used to estimate the unknown parameters in the linear model.
-Because we fit the linear model based on the wall-clock execution time, the derived cost model can be interpreted as an estimate of the execution time when the program is compiled without optimisations.
+Because we fit the linear model based on the wall-clock execution time, the derived cost model can be interpreted as an estimate of the execution time when the program is compiled without optimizations.
 
@@
 
-The IPC metric has been widely used for studying performance benefits of hardware optimisations.
+The IPC metric has been widely used for studying performance benefits of hardware optimizations.
 Although previous work has suggested the use of IPC for guiding {\itercomp}, in this section we argue in favour of the WP metric over IPC.
@@
 The IPC metric differs from the proposed WP metric in a key aspect:
-the IPC metric is computed solely based on the final optimised program.
-When executing different optimised versions of a program with the same input, both the number of instructions and the number of clock cycles can change.
+the IPC metric is computed solely based on the final optimized program.
+When executing different optimized versions of a program with the same input, both the number of instructions and the number of clock cycles can change.
 For this reason, higher IPC does not necessarily translate to shorter execution time (or even fewer clock cycles).
@@
 % As this example illustrates, the IPC metric can be misleading when compared different versions of the same program.
-% This problem is only intensified when both the optimisation sequence and the input changes.
+% This problem is only intensified when both the optimization sequence and the input changes.
 % In Chapter~\ref{chap:eval} we show empirical evidence for this argument.
@@
 
-On the other hand, WP computes the amount of work based on the unoptimised version of the program, which means that it always measures the same amount of work for the same input, regardless of the optimisations applied on the program.
+On the other hand, WP computes the amount of work based on the unoptimized version of the program, which means that it always measures the same amount of work for the same input, regardless of the optimizations applied on the program.
 This is a key aspect that enables the use of the WP metric for guiding {\itercomp}, because a higher WP, which represents work per unit time, naturally translate to shorter execution time.
@@
 
-Although {\itercomp} had been originally proposed as an \textit{offline} optimisation strategy, it can also be adapted to work in online scenarios.
-Instead of selecting the best optimisation sequence as part of the development time (pre-shipping) of a program, a first version of the program is shipped together with an {\itercomp} mechanism.
-In the online scenario, the program is shipped with an initial optimisation sequence and different optimisation sequences are evaluated as the end-user executes the program.
-This optimisation strategy is also known as idle-time optimisation, as the re-compilation happens between runs of the program.
+Although {\itercomp} had been originally proposed as an \textit{offline} optimization strategy, it can also be adapted to work in online scenarios.
+Instead of selecting the best optimization sequence as part of the development time (pre-shipping) of a program, a first version of the program is shipped together with an {\itercomp} mechanism.
+In the online scenario, the program is shipped with an initial optimization sequence and different optimization sequences are evaluated as the end-user executes the program.
+This optimization strategy is also known as idle-time optimization, as the re-compilation happens between runs of the program.
 
-LLVM is particularly suitable for iterative compilation as it makes possible to cache a pre-compiled, but still unoptimised, version of the input program in the bitcode format of the LLVM IR.
+LLVM is particularly suitable for iterative compilation as it makes possible to cache a pre-compiled, but still unoptimized, version of the input program in the bitcode format of the LLVM IR.
 This caching allows to speedup the time required for re-compilation as it is able to bypass the frontend phase.
@@
 Figure~\ref{fig:infra-diagram} shows an overview of the infrastructure required for applying online {\itercomp}
-using WP as the metric of choice for evaluating optimisation sequences.
+using WP as the metric of choice for evaluating optimization sequences.
 The online {\itercomp} follows as described bellow:
 \begin{enumerate}
-\item The program is pre-compiled to the LLVM bitcode format without optimisation.
-\item The unoptimised program is instrumented for work profiling.
-\item Execution-based optimisation search:
+\item The program is pre-compiled to the LLVM bitcode format without optimization.
+\item The unoptimized program is instrumented for work profiling.
+\item Execution-based optimization search:
  \begin{enumerate}
-   \item The current optimisation sequence is used for the re-compilation of the program.
+   \item The current optimization sequence is used for the re-compilation of the program.
    \item The program is executed with any input provided by the end-user.
          During the execution of the program, wall-clock time and the work metric are recorded by profiling instrumentation.
-   \item If the recorded profiling for the current optimisation can be used to compute an average performance measurement within a small confidence interval,
-         then a new optimisation sequence is generated.
-         Otherwise, the same optimisation sequence is used for the next execution.
+   \item If the recorded profiling for the current optimization can be used to compute an average performance measurement within a small confidence interval,
+         then a new optimization sequence is generated.
+         Otherwise, the same optimization sequence is used for the next execution.
  \end{enumerate}
@@
 
-Notice how the \textit{Work Instrumentation} phase is executed before performing any optimisation to the program.
-It is critical so that the work profiling always measure the same amount of work for a given input, regardless of the optimisation sequence applied to the program.
+Notice how the \textit{Work Instrumentation} phase is executed before performing any optimization to the program.
+It is critical so that the work profiling always measure the same amount of work for a given input, regardless of the optimization sequence applied to the program.
 Because the instrumentation is performed before optimising the program, it means that the work profiling derives the linear expression defined in Equation~\ref{eq:linear-work-expression}
-based on the unoptimised program.
-In other words, the basic blocks and the number of occurrences of the instruction in the basic blocks reflect the unoptimised program.
-This particular sequence of compilation guarantees that the amount of work must is only input dependent, but consistent between different optimisation sequences.
+based on the unoptimized program.
+In other words, the basic blocks and the number of occurrences of the instruction in the basic blocks reflect the unoptimized program.
+This particular sequence of compilation guarantees that the amount of work must is only input dependent, but consistent between different optimization sequences.
 
-As the name suggests, the component called \textit{Optimisation Selector} is responsible for selecting which optimisation sequence to use for the next execution of the program.
-It can either keep the same optimisation sequence used in the previous execution or start monitoring the performance of a new optimisation sequence.
-An optimisation sequence can be kept for multiple executions of the program in order to gather enough measurement to compute an average performance with small statistical deviations, i.e., for which the confidence interval has a small range.
-We call by \textit{Input-Window Size} the number of executions performed using the same optimisation sequence.
+As the name suggests, the component called \textit{Optimization Selector} is responsible for selecting which optimization sequence to use for the next execution of the program.
+It can either keep the same optimization sequence used in the previous execution or start monitoring the performance of a new optimization sequence.
+An optimization sequence can be kept for multiple executions of the program in order to gather enough measurement to compute an average performance with small statistical deviations, i.e., for which the confidence interval has a small range.
+We call by \textit{Input-Window Size} the number of executions performed using the same optimization sequence.
 
@@
 For example, mobile devices are usually intensely used during the day, with some idle periods usually when the battery is being re-charged~\citep{mpeis16}.
-Similarly, many authors have also considered peak and idle (or underutilised) periods in the context of data centres~\citep{armbrust10,chen12b}.
+Similarly, many authors have also considered peak and idle (or underutilized) periods in the context of data centres~\citep{armbrust10,chen12b}.
 
 The proposed infrastructure is very well suited for these real online scenarios\footnote{
-However, if idle time is almost non-existent, the proposed infrastructure can still be used by re-compiling the program with a different optimisation while multiple runs of the program are being executed.}.
-In particular, periods of peak usage could be used to monitor multiple runs of the program using the same optimisation sequence, collecting the work profiling and measuring its execution time,
-while periods of idleness or underutilisation could be leveraged to use the profiling statistics for selecting better optimisations and re-compiling the program.
+However, if idle time is almost non-existent, the proposed infrastructure can still be used by re-compiling the program with a different optimization while multiple runs of the program are being executed.}.
+In particular, periods of peak usage could be used to monitor multiple runs of the program using the same optimization sequence, collecting the work profiling and measuring its execution time,
+while periods of idleness or underutilisation could be leveraged to use the profiling statistics for selecting better optimizations and re-compiling the program.
 
@@
 Ball and Larus~\cite{ball94} show that a set of edges represents the minimum number of probes for profiling block frequency if and only if the complementary set of edges forms a spanning tree.
-The instrumentation optimises the placement of the probes with respect to a weighing that assigns a non-negative value to each edge in the CFG.
+The instrumentation optimizes the placement of the probes with respect to a weighing that assigns a non-negative value to each edge in the CFG.
 The cost of profiling a set of edges is proportional to the sum of the weights of the edges.
 These weights can be obtained either by empirical measurements or heuristic estimations.
-In order to minimise the profiling cost, the instrumentation computes the maximum spanning tree for avoiding probing in frequently executed edges.
+In order to minimize the profiling cost, the instrumentation computes the maximum spanning tree for avoiding probing in frequently executed edges.
 Although this instrumentation algorithm is proved to produce the optimal placement of probes for well-structured CFGs, it may produce sub-optimal placement for some unstructured CFGs.
@@
 \begin{aligned}
-& \textrm{maximise }\quad \sum_{i=0}^{k} f(P_i)x_i \\
+& \textrm{maximize }\quad \sum_{i=0}^{k} f(P_i)x_i \\
 & \textrm{subject to }\quad \sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M
@@
 For these cases, we propose an adapted version of the relaxation algorithm that operates on the whole program.
-Traditionally, compilers optimisations are performed on the function-level, or at best on a per module basis.
-Whole program optimisation (WPO) means that the compiler considers all compilation units of the program and optimises them using the combined knowledge of how they are used together.
+Traditionally, compilers optimizations are performed on the function-level, or at best on a per module basis.
+Whole program optimization (WPO) means that the compiler considers all compilation units of the program and optimizes them using the combined knowledge of how they are used together.
 

### 2017-09-08 motivation.tex
motivation figure
Merge remote-tracking branch 'origin/master'
# Conflicts:
#	cgo18mscr.tex
#	motivation.tex
updates motivation

@@
 
-As a motivation example, consider...
+%As a motivation example, consider...
 
-Here we give an example to motivate our work. E.g. assuming we have a zero-cost profiling based work metric, 
-how can we use it to perform iterative compilation across runs? Then we talk about while it is impossible to 
-build a zero-overhead metric, we can develop one with low overhead.
+%Here we give an example to motivate our work. E.g. assuming we have a zero-cost profiling based work metric,
+%how can we use it to perform iterative compilation across runs? Then we talk about while it is impossible to
+%build a zero-overhead metric, we can develop one with low overhead.
+
+This section provides a comparison between the offline and online approaches for
+{\itercomp}.
+The goal is to motivate the need for an online {\itercomp}, highlighting
+weaknesses of using offline {\itercomp}.
+
+In the offline approach, a fixed set of inputs is selected for which the best
+optimization sequence is selected during the development time (pre-shipping) of
+a program.
+Although this fixed set of inputs is expected to be representative of the user's
+workload,
+in most cases, it might be infeasible to identify and collect representative
+workloads for every group of users.
+Furthermore, a user can dynamically change its workload pattern, rendering
+a fixed optimization sequence innapropriate for his performance requirements.
+
+In the online approach, the program is shipped with an initial optimization
+sequence and different optimisation sequences are evaluated as the end-user
+executes the program.
+This approach frees developers from the need to collect representative
+workloads for every group of users so that the program can be optimized
+accordingly.
+More importantly, even in scenarios where the user's workload can dynamically
+change its pattern, the online {\itercomp} would still be able to optimize the
+program accordingly.
+
+Figure~\ref{fig:motivation} compares the offline {\itercomp} with a potential
+implementation of the online {\itercomp}.
+This figure shows the performance gains of the best optimization 
+sequence found by each approach, compared with the standard
+compiler's optimization (i.e.,\texttt{-O3}), executed with 1,000 inputs.
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/motivation.pdf}
+    \caption{.}
+    \label{fig:motivation}
+\end{figure}

### 2017-09-08 related.tex
updates motivation

@@
 \textit{How data input dependent is {\itercomp}?}
-Their results show that it is possible to find a combination of compiler optimisations that achieves at least 86\% of the maximum speedup across all input test cases.
-These optimal combinations are program-specific and yield average speedups up to 3.75$\times$ over the highest optimisation level of compilers.
+Their results show that it is possible to find a combination of compiler optimizations that achieves at least 86\% of the maximum speedup across all input test cases.
+These optimal combinations are program-specific and yield average speedups up to 3.75$\times$ over the highest optimization level of compilers.
 
-When optimising a program, the main method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a} evaluates each combination of compiler optimisations across all the available inputs, i.e., if $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimisations, they perform a total of $O(NM)$ runs of the program being optimised.
-Furthermore, they use a pre-defined set of only 300 different combinations of compiler optimisations, which represents a very small sample of the optimisation search space for most modern compilers, e.g.
-LLVM has 56 distinct optimisation passes and GCC has about 47 high-level (SSA form) optimisation passes and about 25 low-level (RTL) optimisation passes, which in both cases result in much more than $2^{50}$ distinct combinations of compiler optimisations, without considering repetition.
+When optimising a program, the main method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a} evaluates each combination of compiler optimizations across all the available inputs, i.e., if $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimizations, they perform a total of $O(NM)$ runs of the program being optimized.
+Furthermore, they use a pre-defined set of only 300 different combinations of compiler optimizations, which represents a very small sample of the optimization search space for most modern compilers, e.g.
+LLVM has 56 distinct optimization passes and GCC has about 47 high-level (SSA form) optimization passes and about 25 low-level (RTL) optimization passes, which in both cases result in much more than $2^{50}$ distinct combinations of compiler optimizations, without considering repetition.
 
@@
 In summary, each worker receives a subset of the input dataset, called the evaluation dataset, to perform an \textit{online} {\itercomp} of the code being executed.
-Each worker performs the same the method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a}, i.e., they evaluate each combination of compiler optimisations across all the evaluation dataset.
-However, because the optimisation is performed online, they usually consider a small evaluation dataset and a small number of compiler optimisations.
+Each worker performs the same the method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a}, i.e., they evaluate each combination of compiler optimizations across all the evaluation dataset.
+However, because the optimization is performed online, they usually consider a small evaluation dataset and a small number of compiler optimizations.
 
-Fursin~\etal~\cite{fursin07} addressed the problem of comparing the effect of two optimisations on two distinct inputs.
+Fursin~\etal~\cite{fursin07} addressed the problem of comparing the effect of two optimizations on two distinct inputs.
 For that purpose, they proposed to use instruction per cycle (IPC) as the metric for performing such comparison.
 Their result show that using IPC seems promising as a robust metric for {\itercomp} across large input datasets.
-However, some specific optimisation techniques may affect the use of IPC as a robust metric, and specially IPC has been shown to provide poor performance estimation for multi-threaded programs~\cite{alameldeen06,eyerman08}.
+However, some specific optimization techniques may affect the use of IPC as a robust metric, and specially IPC has been shown to provide poor performance estimation for multi-threaded programs~\cite{alameldeen06,eyerman08}.
 In particular, IPC can give a skewed performance measure if threads spend time in \textit{spin-lock loops} or other synchronisation mechanisms.
@@
 
-Goldsmith~\etal~\cite{goldsmith07} use \textit{block frequency} as the measure for performance for empirically describing the asymptotic behaviour of programs, which is known as empirical computational complexity.
+Goldsmith~\etal~\cite{goldsmith07} use \textit{block frequency} as the measure for performance for empirically describing the asymptotic behavior of programs, which is known as empirical computational complexity.
 Block frequency is a relative metric that represents the number of times a basic block executes~\cite{ball94,ball96}.
-They argue in favour of block frequency due to its portability, repeatability and exactness, since it does not suffer from timer resolution problems or non-deterministic noises.
+They argue in favor of block frequency due to its portability, repeatability and exactness, since it does not suffer from timer resolution problems or non-deterministic noises.
 Block frequency also has the advantage of being efficiently profiled by means of automatic code instrumentation~\cite{knuth73,ball94}.
 
-However, in the context of comparing different optimisations, although block frequency would be able to capture aspects of optimisations that simplify the control-flow graph (CFG), measuring work at the basic block resolution would not capture effects of optimisations at the instruction level.
+However, in the context of comparing different optimizations, although block frequency would be able to capture aspects of optimizations that simplify the control-flow graph (CFG), measuring work at the basic block resolution would not capture effects of optimizations at the instruction level.
 Because of that, we extend the idea of using basic block frequency to measure computational work by also considering the computational cost of each basic block.

### 2017-09-08 results.tex
updates motivation

@@
 
-For the evaluation of the performance overhead that the instrumentation incur to the benchmarks, we measure the wall-clock time of the benchmarks when compiled with the default {\flagstype -O3} optimisation.
+For the evaluation of the performance overhead that the instrumentation incur to the benchmarks, we measure the wall-clock time of the benchmarks when compiled with the default {\flagstype -O3} optimization.
 For each benchmark, we compute the average overhead over all its 1000 input dataset.
@@
 For comparison, we use four configurations\footnote{A configuration using the WPO relaxation was not used due to the time necessary for executing the experiments.}.
-In all configurations, the same optimisation sequence is used for multiple inputs, using a dynamic input-window size, as explained in Section~\ref{sec:oic-infra}.
-The average performance over the input window provides an estimate for the overall performance of the optimisation sequence across distinct inputs.
-The optimisation sequences are ranked based on their average performance, and the best optimisation sequence is selected.
-%When selecting the best optimisation sequence, they are ranked by the average of their performance measurement, using the WP metric, except for the Oracle-RM which uses the real speedup over \texttt{-O0}.
+In all configurations, the same optimization sequence is used for multiple inputs, using a dynamic input-window size, as explained in Section~\ref{sec:oic-infra}.
+The average performance over the input window provides an estimate for the overall performance of the optimization sequence across distinct inputs.
+The optimization sequences are ranked based on their average performance, and the best optimization sequence is selected.
+%When selecting the best optimization sequence, they are ranked by the average of their performance measurement, using the WP metric, except for the Oracle-RM which uses the real speedup over \texttt{-O0}.
 \begin{itemize}
-\item \textbf{Oracle-RM} executes the program multiple times for each input, measuring the real speedup for each optimisation sequence, and then uses the real speedup over {\flagstype -O0} for comparing the optimisation sequences.
+\item \textbf{Oracle-RM} executes the program multiple times for each input, measuring the real speedup for each optimization sequence, and then uses the real speedup over {\flagstype -O0} for comparing the optimization sequences.
   The speedups are computed based on the wall-clock time.
@@
 \item \textbf{Oracle-PP} represents an oracle with a \textit{perfect} non-intrusive profiling.
-  Although it uses the WP metric for comparing optimisation sequences, this oracle also avoids noise in its measurements by also executing the program multiple times for each input.
+  Although it uses the WP metric for comparing optimization sequences, this oracle also avoids noise in its measurements by also executing the program multiple times for each input.
   The first execution is used for profiling the work metric.
@@
 \item \textbf{Real-OP} corresponds to the online {\itercomp} as it would be applied in a real online scenario.
-  For each optimisation, a random sample of inputs is selected, and the program is executed only once with each input.
+  For each optimization, a random sample of inputs is selected, and the program is executed only once with each input.
   When executing each input, it uses the optimal work instrumentation for profiling the work metric.
-  The average of the WP for the sample of inputs is then used for selecting the best optimisation.
+  The average of the WP for the sample of inputs is then used for selecting the best optimization.
 \item \textbf{Real-R5} is similar to the {Real-OP}.
   It also corresponds to the online {\itercomp} as it would be applied in a real online scenario.
-  For each optimisation, a random sample of inputs is selected, and the program is executed only once with each input.
+  For each optimization, a random sample of inputs is selected, and the program is executed only once with each input.
   When executing each input, it uses the relaxed work instrumentation, with a 5\% threshold, for profiling the work metric.
-  The average of the WP for the sample of inputs is then used for selecting the best optimisation.
+  The average of the WP for the sample of inputs is then used for selecting the best optimization.
 \item \textbf{Real-IPC} corresponds to the online {\itercomp} based on the IPC metric.
-  For each optimisation, a random sample of inputs is selected, and the program is executed only once with each input.
-  The average of the IPC metric for the sample of inputs is then used for selecting the best optimisation.
+  For each optimization, a random sample of inputs is selected, and the program is executed only once with each input.
+  The average of the IPC metric for the sample of inputs is then used for selecting the best optimization.
 \end{itemize}
@@
 
-\subsection{The Set of Optimisation Sequences}
+\subsection{The Set of Optimization Sequences}
 
-For the purpose of evaluating the use of the WP metric with {\itercomp}, we collected in advance a fixed set of optimisation sequences.
-The reason for using this fixed set, as explained in Section~\ref{sec:oic-infra}, is because this paper is focused on other components of the infrastructure for performing {\itercomp} and we assume that a good generator of optimisation sequences will be used in a real online scenario.
-This set contains 500 optimisation sequences collected in a random search using the training benchmarks.
-These optimisation sequences contain an average of 40 individual optimisation passes, including repetitions, with a maximum of 119 optimisation passes, but it also contains optimisation sequences which consist of a single flag, such as the default optimisations {\flagstype -O1}, {\flagstype -O2}, {\flagstype -O3}, {\flagstype -Os}, and {\flagstype -Oz}.
+For the purpose of evaluating the use of the WP metric with {\itercomp}, we collected in advance a fixed set of optimization sequences.
+The reason for using this fixed set, as explained in Section~\ref{sec:oic-infra}, is because this paper is focused on other components of the infrastructure for performing {\itercomp} and we assume that a good generator of optimization sequences will be used in a real online scenario.
+This set contains 500 optimization sequences collected in a random search using the training benchmarks.
+These optimization sequences contain an average of 40 individual optimization passes, including repetitions, with a maximum of 119 optimization passes, but it also contains optimization sequences which consist of a single flag, such as the default optimizations {\flagstype -O1}, {\flagstype -O2}, {\flagstype -O3}, {\flagstype -Os}, and {\flagstype -Oz}.
 
@@
   %    \vspace{1em}
-  %    \noindent\textbf{Example of a short optimisation sequence:}\vspace{-1ex}
+  %    \noindent\textbf{Example of a short optimization sequence:}\vspace{-1ex}
   %    \justify{\flagstype -mem2reg -simplifycfg -constprop -dce}
@@
   %    \vspace{1em}
-  %    \noindent\textbf{Example of a long optimisation sequence:}\vspace{-1ex}
+  %    \noindent\textbf{Example of a long optimization sequence:}\vspace{-1ex}
   %    \justify{\flagstype -globalopt -reassociate -instcombine -loop-rotate -block-freq -deadargelim -early-cse -sroa -argpromotion -sccp -tbaa -barrier -constmerge \mbox{-loop-vectorize} -domtree -basicaa -memdep -basiccg -memcpyopt \mbox{-constprop} -adce -globaldce -mem2reg -constmerge \mbox{-globaldce} -constprop -instsimplify -dse -dce -simplifycfg -loop-unroll -reassociate -constprop \mbox{-globaldce} -instsimplify -adce -constmerge -bb-vectorize -dce -mergefunc -simplifycfg -dse -loop-unroll -globaldce}
@@
   %    \vspace{1em}
-  %    \noindent\textbf{Example of an optimisation sequence which includes {-O3}:}\vspace{-1ex}
+  %    \noindent\textbf{Example of an optimization sequence which includes {-O3}:}\vspace{-1ex}
   %    \justify{\flagstype -O3 -adce -globaldce -simplifycfg -memcpyopt -reassociate -mergefunc \mbox{-dce} -dse}
@@
 
-Repeating the same optimisation pass can be beneficial and usually expected by other passes.
+Repeating the same optimization pass can be beneficial and usually expected by other passes.
 For example, the {\flagstype -loop-simplify} pass is used for transforming loops into a canonical form by inserting pre-header and exit basic blocks.
-Although this pass inserts jumps due to redundant basic blocks, this canonical form can be favourable to other loop optimisations.
-Because of the redundant basic blocks, this optimisation pass expects that the {\flagstype -simplifycfg} will eventually be executor later on the optimisation pipeline.
+Although this pass inserts jumps due to redundant basic blocks, this canonical form can be favourable to other loop optimizations.
+Because of the redundant basic blocks, this optimization pass expects that the {\flagstype -simplifycfg} will eventually be executor later on the optimization pipeline.
 Another example of such inter-relation between transformations concerns the {\flagstype -licm} and {\flagstype -mem2reg} passes.
@@
 However, many of the analysis required for identifying loop invariant also benefit from the transformations performed by the {\flagstype -mem2reg} pass.
-These examples illustrate the importance of repeating optimisation passes.
+These examples illustrate the importance of repeating optimization passes.
 Moreover, they illustrate the intricate relation amongst several transformations.
 
-However, all optimisation sequences in the set of optimisations were generated completely at random, without using any knowledge of individual transformations.
-Each optimisation sequence was generated in two steps: \textit{(1)} randomly selects the number of flags; \textit{(2)} randomly selects the flags, allowing repetitions.
-Afterwards, this randomly generated optimisation sequence would be included in the set of optimisation sequences only if it was able to improve the performance of a training benchmark, also selected at random, in respect of the {\flagstype -O3} default optimisation.
-This process was repeated until we obtained all the 500 distinct optimisation sequences.
+However, all optimization sequences in the set of optimizations were generated completely at random, without using any knowledge of individual transformations.
+Each optimization sequence was generated in two steps: \textit{(1)} randomly selects the number of flags; \textit{(2)} randomly selects the flags, allowing repetitions.
+Afterwards, this randomly generated optimization sequence would be included in the set of optimization sequences only if it was able to improve the performance of a training benchmark, also selected at random, in respect of the {\flagstype -O3} default optimization.
+This process was repeated until we obtained all the 500 distinct optimization sequences.
 
@@
 
-In order to evaluate the quality of the final optimisation sequences selected by the {\itercomp} search, we compare their speedup by measuring wall-clock time of the benchmarks when compiled with the standard {\flagstype -O3} optimisation.
-For each benchmark, after selecting the final optimisation sequence, we compute the average speedup over all the 1000 input dataset.
+In order to evaluate the quality of the final optimization sequences selected by the {\itercomp} search, we compare their speedup by measuring wall-clock time of the benchmarks when compiled with the standard {\flagstype -O3} optimization.
+For each benchmark, after selecting the final optimization sequence, we compute the average speedup over all the 1000 input dataset.
 When measuring the wall-clock time for each input, to reduce noise, we execute the same input until we have a statistically sound measurement, i.e.
@@
 Figure~\ref{fig:speedups} shows these average speedups over all the 1000 input dataset.
-This figure shows that the best optimisation sequence selected with the Oracle-PP is very close to the performance of the best optimisation sequence selected with the Oracle-RM.
+This figure shows that the best optimization sequence selected with the Oracle-PP is very close to the performance of the best optimization sequence selected with the Oracle-RM.
 This result is important for demonstrating that a work-based metric has the potential to produce good results in a real online scenario, where there is the restriction that programs execute distinct inputs only once.
-Moreover, the evaluation also indicates that the use of relaxation algorithms does not degrade the optimisation search,
+Moreover, the evaluation also indicates that the use of relaxation algorithms does not degrade the optimization search,
 it might even benefit the {\itercomp} as it tends to insert smaller interferences in the performance measurement.
@@
     \includegraphics[width=\textwidth]{figs/speedups.pdf}
-    \caption{Speedups obtained from the final optimisation sequence selected by the online {\itercomp}.
+    \caption{Speedups obtained from the final optimization sequence selected by the online {\itercomp}.
 	         The speedups reported for each benchmark represents the average speedup across their complete 1000 input datasets.}
@@
 
-\begin{figure*}[h]
-    \centering
-    \includegraphics[width=\textwidth]{figs/ipc-vs-work.pdf}
-    \caption{Comparison between {\itercomp} using IPC or the WP metric.}
-    \label{fig:ipc-vs-work}
-\end{figure*}
+%\begin{figure*}[h]
+%    \centering
+%    \includegraphics[width=\textwidth]{figs/ipc-vs-work.pdf}
+%    \caption{Comparison between {\itercomp} using IPC or the WP metric.}
+%    \label{fig:ipc-vs-work}
+%\end{figure*}
 
-Figure~\ref{fig:ipc-vs-work} presents a comparison between IPC and the WP metric regarding their correlation with the actual speedup over \verb|-O0| observed during {\itercomp}.
-Each point in the figure corresponds to each metric averaged over input windows collected during multiple executions of the {\itercomp}.
-The vertical lines correspond to the optimisation sequence selected by the {\itercomp} search, which either represents the highest IPC or the highest value of WP.
-In all cases, the optimisation sequence selected based on the WP is much closer to the best speedup than the optimisation sequence selected based on IPC, which can be attributed to the fact that the WP has a stronger correlation with the speedup.
-In all cases, IPC shows little to no correlation with the speedup, which corroborates the argument given in Section~\ref{sec:ipc-vs-work-metric}.
+%Figure~\ref{fig:ipc-vs-work} presents a comparison between IPC and the WP metric regarding their correlation with the actual speedup over \verb|-O0| observed during {\itercomp}.
+%Each point in the figure corresponds to each metric averaged over input windows collected during multiple executions of the {\itercomp}.
+%The vertical lines correspond to the optimization sequence selected by the {\itercomp} search, which either represents the highest IPC or the highest value of WP.
+%In all cases, the optimization sequence selected based on the WP is much closer to the best speedup than the optimization sequence selected based on IPC, which can be attributed to the fact that the WP has a stronger correlation with the speedup.
+%In all cases, IPC shows little to no correlation with the speedup, which corroborates the argument given in Section~\ref{sec:ipc-vs-work-metric}.
 
-For some cases, namely for the tiff-related benchmarks, the WP shows a worse correlation with speedup.
-We believe these cases could be improved by improving the cost model used to compute the work metric.
-However, even in these cases, the optimisation sequences selected based on the WP outperform those selected based on the IPC metric.
+%For some cases, namely for the tiff-related benchmarks, the WP shows a worse correlation with speedup.
+%We believe these cases could be improved by improving the cost model used to compute the work metric.
+%However, even in these cases, the optimization sequences selected based on the WP outperform those selected based on the IPC metric.
 

### 2017-09-10 cgo18mscr.tex
overall updates

@@
 
-\section{Work-based Performance Metric} \label{sec:metric}
+\section{The Work-Efficiency Performance Metric} \label{sec:metric}
 
-In this section we define the work-based performance (WP) metric proposed for comparing different optimized versions of a program when executing with different inputs.
-We define the performance metric as the ratio between the amount of \textit{work}, $\Delta W$, performed during a period of time, $\Delta t$.
+In this section we define work efficiency, a metric proposed for comparing different optimized versions of a program when executing with different inputs.
+We define it as the ratio between the amount of \textit{work}, $\Delta W$, performed during a period of time, $\Delta t$.
 \[
@@
 
-By measuring the amount of \textit{work} done per unit of time, we reduce the impact of input-dependent aspects and focus instead on the efficiency of the optimized program.
-For this metric, the main challenge is to precisely define what represents \textit{work}.
-As discussed in Section~\ref{sec:relatedwork}, many work metrics have been presented in the literature, such as block frequency and others.
-However, for comparing the performance benefits of different compiler optimizations on a program, although block frequency would be able to capture aspects of optimizations that simplify the control-flow graph (CFG), we argue that measuring work only at the basic block resolution would not be enough for capturing the effects of optimizations at the instruction level.
-For this reason, we extend the idea of using basic block frequency to measure computational work by also considering the computational cost of each basic block.
+By measuring the amount of \textit{computational work} done per unit of time,
+it reduces the impact of input-dependent aspects and focus instead on the
+efficiency of the optimized program.
+For this metric, the main challenge is to precisely define what represents
+\textit{work}.
+In the context of experimental algorithmics, previous work has suggested using
+\textit{block frequency} as a metric for empirical computational
+complexity~\cite{goldsmith07}.
+%Block frequency is a relative metric that represents the number of times a basic
+%block executes~\citep{ball94,ball96}.
+
+However, in the context of comparing different optimization sequences,
+although block frequency would be able to capture aspects of optimizations that
+simplify the control-flow graph (CFG), measuring work at the basic block
+resolution would not capture effects of optimisations at the instruction level.
+Because of that, in this thesis, we extend the idea of using basic block frequency to measure computational work by also considering the computational cost of each basic block.
+The computational cost of a basic block is given by weighting the instructions that it contains.
 
@@
     \centering
-    \includegraphics[width=0.9\linewidth]{figs/cost-model.pdf}
+    \includegraphics[width=0.6\linewidth]{figs/cost-model.pdf}
     \caption{Linear model fitted from empirical data. The mean absolute error (MAE) for the fitted curve is seven milliseconds.}
@@
 \centering {
-  \includegraphics[scale=0.85]{figs/cfg-example.pdf}\\\vspace{1ex}
+  \includegraphics[scale=0.75]{figs/cfg-example.pdf}\\\vspace{1ex}
   \resizebox{0.45\textwidth}{!}{
@@
   \centering
-  \includegraphics[scale=0.85]{figs/cfg-relax-example.pdf}
+  \includegraphics[scale=0.75]{figs/cfg-relax-example.pdf}
   \caption{Example of a CFG containing a loop and its decomposition into DAGs when applying the relaxation.
@@
 We model the relaxation as a 0-1 Knapsack problem:
-\begin{equation*}
-\begin{aligned}
-& \textrm{maximize }\quad \sum_{i=0}^{k} f(P_i)x_i \\
-& \textrm{subject to }\quad \sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M
-\end{aligned}
-\end{equation*}
-\[
+\begin{gather*}
+\textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
+\textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
 x_i\in\{0,1\}, i\in\{0,\ldots,k\}
-\]
+\end{gather*}
 where $f(P_i)$ is the execution frequency of probe $P_i$, $x_i$ denotes the probes selected for removal, and $\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to the minimum work value possible to compute in the DAG, i.e., if $m$ is the minimum amount of work possible to be computed when executing the DAG, then $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$.
@@
 For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we model the whole program the relaxation as the following 0-1 Knapsack problem:
-\begin{equation*}
-\begin{aligned}
-& \textrm{maximize }\quad \sum_{i=0}^{k} f(P_i)x_i \\
-& \textrm{subject to }\quad \sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M
-\end{aligned}
-\end{equation*}
-\[
+\begin{gather*}
+\textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
+\textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
 x_i\in\{0,1\}, i\in\{0,\ldots,k\}
-\]
+\end{gather*}
 where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$, $x_i$ denotes the probes selected for removal, $M$ is the error threshold, and $\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to the profiled global work, i.e.,

### 2017-09-10 motivation.tex
overall updates

@@
 In the online approach, the program is shipped with an initial optimization
-sequence and different optimisation sequences are evaluated as the end-user
+sequence and different optimization sequences are evaluated as the end-user
 executes the program.
@@
 implementation of the online {\itercomp}.
+For this experiment, we have executed the program twice for each input
+in order to compute the speedup of the optimization sequence being evaluated.
 This figure shows the performance gains of the best optimization 
@@
 compiler's optimization (i.e.,\texttt{-O3}), executed with 1,000 inputs.
+Both approaches used the same set of optimization sequences, but because
+the online {\itercomp} compares optimization sequences across inputs,
+it is better informed throughout the search.
 
@@
     \includegraphics[width=0.5\textwidth]{figs/motivation.pdf}
-    \caption{.}
+    \caption{Comparison between the offline and online {\itercomp}.
+             The vertical lines represent the average speedups over
+             all the 1,000 inputs.
+             }
     \label{fig:motivation}
 \end{figure}
+
+However, because of the restriction of having a single execution per input,
+it is not possible to measure speedup in a real-world online scenario.
+Moreover, measuring just execution time, for example, is also not viable since
+different inputs often mean different amounts of work,
+rendering it meaningless to compare optimizations only by execution time.
+
+In the remaining of this paper we develop a work efficiency metric, suitable
+for guiding online {\itercomp}, which can be profiled with a single execution
+of the program with the optimization sequence being evaluated.

### 2017-09-10 related.tex
overall updates

@@
 
-\subsection{Work and Input Size Metrics}
+%\subsection{Work and Input Size Metrics}
 
-Previous work have proposed profiling-based mechanism to estimate input sizes~\cite{zaparanuks12,coppa14}.
-Coppa~\etal~\cite{coppa14} in particular propose the concept of \textit{read memory size} for automatically estimating the size of the input passed to a routine, where \textit{read memory size} represents the number of distinct memory cells first accessed by a read operation.
-In other words, the \textit{read memory size} metric measures the size of the useful portion of the input's memory footprint.
-However, because we are interested in the amount of computational work performed in respect of a given input, the memory footprint of the input may not always have a direct correspondence to  the amount of computational work.
+%Previous work have proposed profiling-based mechanism to estimate input sizes~\cite{zaparanuks12,coppa14}.
+%Coppa~\etal~\cite{coppa14} in particular propose the concept of \textit{read memory size} for automatically estimating the size of the input passed to a routine, where \textit{read memory size} represents the number of distinct memory cells first accessed by a read operation.
+%In other words, the \textit{read memory size} metric measures the size of the useful portion of the input's memory footprint.
+%However, because we are interested in the amount of computational work performed in respect of a given input, the memory footprint of the input may not always have a direct correspondence to  the amount of computational work.
 
-Goldsmith~\etal~\cite{goldsmith07} use \textit{block frequency} as the measure for performance for empirically describing the asymptotic behavior of programs, which is known as empirical computational complexity.
-Block frequency is a relative metric that represents the number of times a basic block executes~\cite{ball94,ball96}.
-They argue in favor of block frequency due to its portability, repeatability and exactness, since it does not suffer from timer resolution problems or non-deterministic noises.
-Block frequency also has the advantage of being efficiently profiled by means of automatic code instrumentation~\cite{knuth73,ball94}.
+%Goldsmith~\etal~\cite{goldsmith07} use \textit{block frequency} as the measure for performance for empirically describing the asymptotic behavior of programs, which is known as empirical computational complexity.
+%Block frequency is a relative metric that represents the number of times a basic block executes~\cite{ball94,ball96}.
+%They argue in favor of block frequency due to its portability, repeatability and exactness, since it does not suffer from timer resolution problems or non-deterministic noises.
+%Block frequency also has the advantage of being efficiently profiled by means of automatic code instrumentation~\cite{knuth73,ball94}.
 
-However, in the context of comparing different optimizations, although block frequency would be able to capture aspects of optimizations that simplify the control-flow graph (CFG), measuring work at the basic block resolution would not capture effects of optimizations at the instruction level.
-Because of that, we extend the idea of using basic block frequency to measure computational work by also considering the computational cost of each basic block.
-The computational cost of a basic block is given by weighing the instructions that it contains.
+%However, in the context of comparing different optimizations, although block frequency would be able to capture aspects of optimizations that simplify the control-flow graph (CFG), measuring work at the basic block resolution would not capture effects of optimizations at the instruction level.
+%Because of that, we extend the idea of using basic block frequency to measure computational work by also considering the computational cost of each basic block.
+%The computational cost of a basic block is given by weighing the instructions that it contains.
 

### 2017-09-11 results.tex
overall updates
minor changes in results

@@
 
-\begin{figure*}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{figs/num-probes.pdf}
-    \caption{Percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.}
-    \label{fig:num-probes}
-\end{figure*}
+%\begin{figure*}[ht]
+%    \centering
+%    \includegraphics[width=\textwidth]{figs/num-probes.pdf}
+%    \caption{Percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.}
+%    \label{fig:num-probes}
+%\end{figure*}
 
@@
 \begin{itemize}
-\item \textbf{Oracle-RM} executes the program multiple times for each input, measuring the real speedup for each optimization sequence, and then uses the real speedup over {\flagstype -O0} for comparing the optimization sequences.
-  The speedups are computed based on the wall-clock time.
-  In order to reduce noise, the program is executed several times for the same input, until the confidence interval was no larger than 1\% for a 99\% confidence.
+\item \textbf{Oracle-RM} executes the program twice for each input, measuring the real speedup for each optimization sequence, and then uses the real speedup over {\flagstype -O0} for guiding online {\itercomp}.
+This oracle provides a reference results for assessing the effectiveness work-efficiency metric.
 \item \textbf{Oracle-PP} represents an oracle with a \textit{perfect} non-intrusive profiling.
-  Although it uses the WP metric for comparing optimization sequences, this oracle also avoids noise in its measurements by also executing the program multiple times for each input.
-  The first execution is used for profiling the work metric.
-  The remaining executions are used for measuring the wall-clock time without using the work profiling.
-\item \textbf{Real-OP} corresponds to the online {\itercomp} as it would be applied in a real online scenario.
-  For each optimization, a random sample of inputs is selected, and the program is executed only once with each input.
-  When executing each input, it uses the optimal work instrumentation for profiling the work metric.
-  The average of the WP for the sample of inputs is then used for selecting the best optimization.
-\item \textbf{Real-R5} is similar to the {Real-OP}.
-  It also corresponds to the online {\itercomp} as it would be applied in a real online scenario.
-  For each optimization, a random sample of inputs is selected, and the program is executed only once with each input.
-  When executing each input, it uses the relaxed work instrumentation, with a 5\% threshold, for profiling the work metric.
-  The average of the WP for the sample of inputs is then used for selecting the best optimization.
-\item \textbf{Real-IPC} corresponds to the online {\itercomp} based on the IPC metric.
-  For each optimization, a random sample of inputs is selected, and the program is executed only once with each input.
-  The average of the IPC metric for the sample of inputs is then used for selecting the best optimization.
+  It uses the work-efficiency metric for guiding online {\itercomp}.
+  Because this version simulates a \textit{perfect} profiling, with zero overhead,
+  it allows us isolate and evaluate the effectiveness work-efficiency metric.
+\item \textbf{Real-OP} corresponds to the online {\itercomp} as it would be applied in a real-world online scenario.
+  The program is executed only once with each input.
+  It uses the optimal work profiling for computing the work efficiency.
+\item \textbf{Real-R5} is similar to the {Real-OP}, but with a relaxation of 5\% in the work profiling.
+\item \textbf{Real-IPC} uses IPC for guiding the online {\itercomp}.
 \end{itemize}

### 2017-09-11 motivation.tex
updates to motivation

--- 2017-09-07 motivation.tex
@@
 
-As a motivation example, consider...
+%As a motivation example, consider...
 
-Here we give an example to motivate our work. E.g. assuming we have a zero-cost profiling based work metric, 
-how can we use it to perform iterative compilation across runs? Then we talk about while it is impossible to 
-build a zero-overhead metric, we can develop one with low overhead.
+%Here we give an example to motivate our work. E.g. assuming we have a zero-cost profiling based work metric,
+%how can we use it to perform iterative compilation across runs? Then we talk about while it is impossible to
+%build a zero-overhead metric, we can develop one with low overhead.
+
+This section provides a comparison between the offline and online approaches for
+{\itercomp}.
+The goal is to motivate the need for an online {\itercomp}, highlighting
+weaknesses of using offline {\itercomp}.
+
+%In the offline approach, a fixed set of inputs is selected for which the best
+%optimization sequence is selected during the development time (pre-shipping) of
+%a program.
+%Although this fixed set of inputs is expected to be representative of the user's
+%workload,
+%in most cases, it might be infeasible to identify and collect representative
+%workloads for every group of users.
+%Furthermore, a user can dynamically change its workload pattern, rendering
+%a fixed optimization sequence innapropriate for his performance requirements.
+
+%In the online approach, the program is shipped with an initial optimization
+%sequence and different optimization sequences are evaluated as the end-user
+%executes the program.
+%This approach frees developers from the need to collect representative
+%workloads for every group of users so that the program can be optimized
+%accordingly.
+%More importantly, even in scenarios where the user's workload can dynamically
+%change its pattern, the online {\itercomp} would still be able to optimize the
+%program accordingly.
+
+In the offline approach, a small fixed set of inputs is selected for which all
+optimization sequences are evaluated.
+In the online approach, each optimization sequence is evaluated based on a,
+dynamically selected, distinct set of inputs.
+In both cases, the optimization sequences are ranked based on their average
+speedup, and the best optimization sequence is selected.
+
+In order to assess the quality of the optimization sequence selected by each
+approach, we evaluate both optimization sequences against all available inputs
+in our dataset.
+Figure~\ref{fig:motivation-online} shows this comparison between both the offline
+and online approaches for {\itercomp}.
+It shows the performance gains of both optimization sequences compared with the
+standard compiler's optimization (i.e.,\texttt{-O3}).
+The online {\itercomp} achieves an average speedup which is about 13\% better
+than the solution selected by the offline approach.
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/motivation-online.pdf}
+    \caption{Comparison between the offline and online {\itercomp}.
+             The vertical lines represent the average speedups over
+             all the 1,000 inputs.
+             }
+    \label{fig:motivation-online}
+\end{figure}
+
+It is worth mentioning that our goal is to improve the average performance
+across the user's workload, and not only for a specific input.
+This means that the online {\itercomp} is better informed throughout the search,
+as it compares optimization sequences across different sets inputs.
+
+However, because of the restriction of having a single execution per input,
+it is not possible to measure speedup in a real-world online scenario.
+Moreover, measuring just execution time, for example, is also not viable since
+different inputs often mean different amounts of work,
+rendering it meaningless to compare optimizations only by execution time.
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/motivation-metric.pdf}
+    \caption{Comparison between the offline and online {\itercomp}.
+             The vertical lines represent the average speedups over
+             all the 1,000 inputs.
+             }
+    \label{fig:motivation-metric}
+\end{figure}
+
+In the remaining of this paper we develop a work efficiency metric, suitable
+for guiding online {\itercomp}, which can be profiled with a single execution
+of the program with the optimization sequence being evaluated.

### 2017-09-12 abstract.tex
Abstract

@@
 \begin{abstract}
-Until recently, most of the existing work in {\itercomp} had been focusing on
-finding the best optimisation through repeated runs using a single input.
-However, in real-world online scenarios, the user rarely executes a program with
-the same input multiple times and most of the programs are complex enough that a
-single input case does not capture the whole range of possible scenarios and
-program behaviours.
 
-In this thesis, we focus on enabling {\itercomp} in \textit{online} scenarios.
-Because of the restriction on repeating the execution with same input, it is
-not always feasible to measure the real speedup when comparing optimisations.
-To this end, we propose a \textit{work efficiency} metric to guide
-online {\itercomp}, which is able to achieve about 80\% of the performance
-of an offline oracle, with improvements of up to 20\%.
+    Iterative or adaptive compilation can dramatically improve the performance of a program by searching for the best optimisation
+    settings. However, it relies upon repeatedly evaluating and comparing optimisations off-line on representative inputs. If the
+    developers fail to guess what is a representative input or usage patterns change over time, then the wrong optimisations may be
+    selected. Until this paper, it has been impossible to search online over actual user inputs. Program side effects, interactivity and
+    difficulties capturing the system state prevent repeated execution. Since each input may be executed only once and each input may
+    entail a different amount of work, the search cannot use runtime to compare optimisations. Unless we find a way to compare
+    optimisations even when inputs are executed only once, we will be unable to use real user input for online iterative compilation, users
+    will not have the best optimised programs, and performance and energy efficiency will suffer.
 
-Moreover, because we need to instrument the program for measuring work
-efficiency, providing a low-overhead profiling is essential.
-In order to reduce the profiling overhead, we also propose two relaxation
-algorithms which provide a trade-off between measurement accuracy and overhead.
-Our results show that the two relaxation algorithms can reduce the average
-overhead by 43\% and $2.1\times$ over the optimal work profiling, observing
-up to about $5\times$ of overhead reduction.
+    This paper demonstrates online iterative compilation. We compare optimisations by work efficiency, or work divided by time, executing
+    each input only once. We develop an instrumented, low overhead, work efficiency metric with optimal probe placement. Then we further
+    reduce the overhead by removing high-cost probes whose absence introduces only small errors to the work metric, without significantly
+    affecting the outcome of the iterative compilation. We give two variants for probe removal optimisation: the first offers strong
+    guarantees about the maximum possible error, while the second targets the average, whole program error.
+
+    Our online method yields programs that get 80\% of the speedup an offline oracle achieves but without executing any input more than
+    once, which the oracle cannot do. Through probe removal we bring the instrumentation overhead down to only 4\% on average, making our
+    approach suitable for regular use. We show practical online iterative compilation for the first time, optimising programs according to
+    the range of inputs users actually present.
+
 \end{abstract}

### 2017-09-12 cgo18mscr.tex
minor changes in results
updates motivation and other minor changes

@@
 \maketitle
+\vspace{-1em}
 
@@
 
-In contrast to the naive instrumentation where each basic block records only its own amount of work, with the optimal profiling, the instrumented basic blocks need record an aggregated value of work that represents a path in the CFG.
+In contrast to the naive instrumentation where each basic block records only its own amount of work, with the optimal profiling, the instrumented basic blocks need to record an aggregated value of work that represents a path in the CFG.
 These values are constructed with some instrumented basic blocks speculatively assuming some paths while other probes correct when these assumptions are wrong (see for example $w(P_0)$ and $w(P_1)$ in Figure~\ref{fig:cfg-example}).
 
-\begin{lstlisting}[caption={Post-processing of the CFG for populating all edge flows based on the collected probes.
-\textbf{ToDo: Replace the algorithm with a formal data-flow equation.}}, label={lst:populateEdgeFlows}, float]
+\begin{lstlisting}[caption={Pseudocode of the data-flow analysis necessary for assigning the values computed in each probe.},
+label={lst:populateEdgeInfo}, float]
 // Inputs: CFG with the known edges flows from
@@
 
-Because the algorithm for the optimal placement of the probes is proved to uniquely compute the block frequencies by propagating the probe counts, we adapt this algorithm in order to compose the aggregated values that will be instrumented in each probe, based on our model of computational work, $\Delta W$, derived from the basic block frequencies (see Section~\ref{sec:metric}).
+Because the algorithm for the optimal placement of the probes is proved to uniquely compute the block frequencies by propagating the probe counts, we adapt this algorithm (see \lstlistingname~\ref{lst:populateEdgeFlows}) in order to compose the aggregated values that will be instrumented in each probe, based on our model of computational work, $\Delta W$, derived from the basic block frequencies (see Section~\ref{sec:metric}).
 We perform a similar propagation of the probes in a symbolic fashion, as illustrated in Figure~\ref{fig:cfg-example}.
-This symbolic propagation of the probes is implemented by the data-flow analysis described in Algorithm~\ref{alg:populateEdgeInfo}, and the final aggregated values are extracted from the edge information as described by Algorithm~\ref{alg:instrValue}.
+This symbolic propagation of the probes is implemented by the data-flow analysis described in \lstlistingname~\ref{lst:populateEdgeInfo}, and the final aggregated values are extracted from the edge information as described by \lstlistingname~\ref{lst:instrValue}.
 
-The data-flow analysis in Algorithm~\ref{alg:populateEdgeInfo} keeps two sets for each edge, namely, the increment and the decrement sets.
+The data-flow analysis in \lstlistingname~\ref{lst:populateEdgeInfo} keeps two sets for each edge, namely, the increment and the decrement sets.
 We consider that both sets represent the \textit{edge expressions} shown in Figure~\ref{fig:cfg-example}, for which we define a \textit{symbolic sum} by computing the union of the increment and decrement sets, respectively, with the appropriate cancellation of common elements.
@@
 
-Algorithm~\ref{alg:instrValue} reads the edge information for each basic block by computing the symbolic sum of their respective incoming edges (or outgoing edges).
+\lstlistingname~\ref{lst:instrValue} reads the edge information for each basic block by computing the symbolic sum of their respective incoming edges (or outgoing edges).
 From these \textit{edge expressions}, we are able to compose the aggregated value of the probes.
-The positive terms in the edge expression of a basic block indicate that the amount of work of this basic block will be incremented in the probes represented by these positive terms, similarly, the negative terms indicate that the amount of work of this basic block will be decremented in the probes represented by these negative terms.
+The positive terms in the edge expression of a basic block indicate that the amount of work of this basic block will be incremented in the probes represented by these positive terms.
+Similarly, the negative terms indicate that the amount of work of this basic block will be decremented in the probes represented by these negative terms.
 For example, because the edge expression for the basic block $B_8$ is $P_0 - P_1$, the amount of work of $B_8$, denoted by $w(B_8)$, is incremented in probe $P_0$ and decremented in $P_1$.
 
+%Let $V$ be a vector of $\{-1,0,1\}$ with the sum of the edge information from the incoming edges of a given basic block.
+%Let $W$ be the 
+%\[
+%\mathrm{reduce}\,\,(+)\,\,(\mathrm{map}\,\,(\times W)\,\,E^+(B))
+%\]
+
 \begin{lstlisting}[caption={Pseudocode that describes how the edge information is used in order to extract the value that will be computed in a given instrumented basic block $B_I$.
-This algorithm could equally be implemented based on the predecessors.
-\textbf{ToDo: Using the vector format, we could replace this algorithm with a single equation. Let $V$ be the vector of $\{-1,0,1\}$ and $W$ the weight for each basic block.
-This algorithm could be re-written as $reduce (+) (map (*) (zip V W))$.}
-}, label={lst:instrValue}, float]
+This algorithm could equally be implemented based on the outgoing edges.}, label={lst:instrValue}, float]
 // Inputs: 1) CFG with the known edges flows
 //            of the chords.
-//         2) Basic block targeted for probing.
+//         2) Probe to be instrumented.
 // Output: Work value to be incremented by the
-//         given probe B.
-instrValue(G, B, inc, dec) {
+//         given probe P.
+instrValue(G, P) {
   value = 0
   for B in G.vertices():
-    incB = set()
-    decB = set()
-    for succB in G.successor(B):
-      incB = incB union inc[G.getEdge(B, succB)]
-      decB = decB union dec[G.getEdge(B, succB)]
-    if B in (incB - decB):
-      value = value + w(B)
-    if B in (decB - incB):
-      value = value - w(B)
+    expr = sum( G.incomingEdges(B) ) 
+    //coef(P,expr) is the coefficient
+    //-1,0, or 1 of P in expr
+    value = value + coef(P, expr)*w(B)
   return value
@@
 %The highlighted step is introduced by the relaxed instrumentation on top of the previously defined optimal profiling.
-The relaxed instrumentation performs a post-processing on the resulting instrumentation of the optimal algorithm.
+The relaxed instrumentation performs a post processing on the resulting instrumentation of the optimal algorithm.
 This post-processing identifies probes that add little to the work metric, and removes their instrumentation.

### 2017-09-12 motivation.tex
updates motivation and other minor changes

@@
 rendering it meaningless to compare optimizations only by execution time.
+Therefore, we need a performance metric which can be profiled from a single
+execution of the program.
+Although a previous work has suggested that \textit{instructions per cycle}~(IPC)
+could be used with that goal, IPC have no correlation with speedup~\citep{fursin07}.
+In a case where we have two distinct versions of the program, due to different
+optimizations, which is also executed with distinct inputs, higher IPC does not
+necessarily mean lower execution time (higher performance).
 
@@
 
-In the remaining of this paper we develop a work efficiency metric, suitable
+Figure~\ref{fig:motivation-metric} shows a comparison among these three metrics,
+namely speedup, execution time, and IPC, and their correlation with the actual
+speedup.
+Each point in the figure corresponds to each metric averaged over 
+a set of inputs, collected during the {\itercomp}.
+The vertical lines correspond to the optimization sequence 
+with the best score for each one of the metrics.
+This figure shows that both execution time and IPC are unsuitable for
+online {\itercomp}, as they present no direct correlation with the actual
+speedup obtained by the optimization sequence being evaluated.
+
+In the remaining of this paper, we develop a work efficiency metric, suitable
 for guiding online {\itercomp}, which can be profiled with a single execution

### 2017-09-12 results.tex
updates to motivation
updates motivation and other minor changes

@@
 
+\begin{figure*}[t]
+    \centering
+    \includegraphics[width=\textwidth]{figs/overhead-O3.pdf}
+    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.}
+    \label{fig:overhead-O3}
+\end{figure*}
+
+\begin{figure*}[t]
+    \centering
+    \includegraphics[width=\textwidth]{figs/error-O3.pdf}
+    \caption{Dynamic error of the work profiling averaged over the 1000 inputs, after relaxing the number of probes.}
+    \label{fig:error-O3}
+\end{figure*}
+
 For the evaluation of the performance overhead that the instrumentation incur to the benchmarks, we measure the wall-clock time of the benchmarks when compiled with the default {\flagstype -O3} optimization.
@@
 Figure~\ref{fig:overhead-improvement-O3} indicates the improvement of the relaxation algorithm over the optimal instrumentation, regarding the reduction in overhead.
-
-\begin{figure*}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{figs/overhead-O3.pdf}
-    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.}
-    \label{fig:overhead-O3}
-\end{figure*}
 
@@
 %\end{figure*}
-
-\begin{figure*}[h!]
-    \centering
-    \includegraphics[width=\textwidth]{figs/error-O3.pdf}
-    \caption{Dynamic error of the work profiling averaged over the 1000 inputs, after relaxing the number of probes.}
-    \label{fig:error-O3}
-\end{figure*}
 
@@
 \item \textbf{Real-R5} is similar to the {Real-OP}, but with a relaxation of 5\% in the work profiling.
-\item \textbf{Real-IPC} uses IPC for guiding the online {\itercomp}.
+\item \textbf{Real-WPO-R5} is similar to the {Real-OP}, but with a whole-program relaxation of 5\% in the work profiling.
+%\item \textbf{Real-IPC} uses IPC for guiding the online {\itercomp}.
 \end{itemize}
@@
 
-\begin{figure*}[htb]
+\begin{figure*}[t]
     \centering
@@
 
-Repeating the same optimization pass can be beneficial and usually expected by other passes.
-For example, the {\flagstype -loop-simplify} pass is used for transforming loops into a canonical form by inserting pre-header and exit basic blocks.
-Although this pass inserts jumps due to redundant basic blocks, this canonical form can be favourable to other loop optimizations.
-Because of the redundant basic blocks, this optimization pass expects that the {\flagstype -simplifycfg} will eventually be executor later on the optimization pipeline.
-Another example of such inter-relation between transformations concerns the {\flagstype -licm} and {\flagstype -mem2reg} passes.
-The {\flagstype -licm} pass is responsible for moving invariant code out from the loop body.
-It usually creates new local variables, using memory access operations, for assisting with the code manipulation, which means that the executing the {\flagstype -mem2reg} pass afterwards would be useful as a cleanup pass for removing the extra memory accesses generated.
-However, many of the analysis required for identifying loop invariant also benefit from the transformations performed by the {\flagstype -mem2reg} pass.
-These examples illustrate the importance of repeating optimization passes.
-Moreover, they illustrate the intricate relation amongst several transformations.
+%Repeating the same optimization pass can be beneficial and usually expected by other passes.
+%For example, the {\flagstype -loop-simplify} pass is used for transforming loops into a canonical form by inserting pre-header and exit basic blocks.
+%Although this pass inserts jumps due to redundant basic blocks, this canonical form can be favourable to other loop optimizations.
+%Because of the redundant basic blocks, this optimization pass expects that the {\flagstype -simplifycfg} will eventually be executor later on the optimization pipeline.
+%Another example of such inter-relation between transformations concerns the {\flagstype -licm} and {\flagstype -mem2reg} passes.
+%The {\flagstype -licm} pass is responsible for moving invariant code out from the loop body.
+%It usually creates new local variables, using memory access operations, for assisting with the code manipulation, which means that the executing the {\flagstype -mem2reg} pass afterwards would be useful as a cleanup pass for removing the extra memory accesses generated.
+%However, many of the analysis required for identifying loop invariant also benefit from the transformations performed by the {\flagstype -mem2reg} pass.
+%These examples illustrate the importance of repeating optimization passes.
+%Moreover, they illustrate the intricate relation amongst several transformations.
 
-However, all optimization sequences in the set of optimizations were generated completely at random, without using any knowledge of individual transformations.
+All optimization sequences in the set of optimizations were generated completely at random, without using any knowledge of individual transformations.
 Each optimization sequence was generated in two steps: \textit{(1)} randomly selects the number of flags; \textit{(2)} randomly selects the flags, allowing repetitions.
@@
 it might even benefit the {\itercomp} as it tends to insert smaller interferences in the performance measurement.
-Finally, Figure~\ref{fig:speedups} also presents the results of {\itercomp} guided by IPC, which is futher discussed in Figure~\ref{fig:ipc-vs-work}.
+%Finally, Figure~\ref{fig:speedups} also presents the results of {\itercomp} guided by IPC, which is futher discussed in Figure~\ref{fig:ipc-vs-work}.
 
-\begin{figure*}[htb]
+\begin{figure*}[t]
     \centering
@@
 
-\textbf{ToDo: Include results using Genetic Algorithm search.}
+%\textbf{ToDo: Include results using Genetic Algorithm search.}

### 2017-09-12 cgo18mscr.tex
adds formal description of instrumentation

@@
 
-\section{The Work-Efficiency Performance Metric} \label{sec:metric}
+\section{The Work Efficiency Metric} \label{sec:metric}
 
-In this section we define work efficiency, a metric proposed for comparing different optimized versions of a program when executing with different inputs.
-We define it as the ratio between the amount of \textit{work}, $\Delta W$, performed during a period of time, $\Delta t$.
+In this section we define work efficiency, a metric proposed for comparing
+different optimized versions of a program when executing with different inputs.
+We define it as the ratio between the amount of \textit{work}, $\Delta W$,
+performed during a period of time, $\Delta t$.
 \[
@@
 \textit{work}.
-In the context of experimental algorithmics, previous work has suggested using
+In the context of experimental algorithmics, previous works have suggested using
 \textit{block frequency} as a metric for empirical computational
@@
 resolution would not capture effects of optimisations at the instruction level.
-Because of that, in this thesis, we extend the idea of using basic block frequency to measure computational work by also considering the computational cost of each basic block.
-The computational cost of a basic block is given by weighting the instructions that it contains.
+%Because of that, in this thesis, we extend the idea of using basic block
+%frequency to measure computational work by also considering the computational
+%cost of each basic block.
+Because of that, we adopt a work metric that takes into account the
+computational cost within each basic block.
+The computational cost of a basic block is given by weighting the instructions
+that it contains.
 
@@
 
+We first initialize the edge information $D_{u,v}$, for each edge $(u,v)$ in the CFG.
+\[
+D_{u,v} \gets 
+\begin{cases}
+    P_{u,v} & \quad \text{if edge $(u,v)$ has a probe}\\
+    0       & \quad \text{otherwise}
+\end{cases}   
+\]
+Afetwards, for each vertex $u$, in post-order of the spanning tree, we apply the following operations:
+\[
+\Sigma^+_u = \sum_{v\in N^+(u)} D_{u,v}
+\]
+\[
+\Sigma^-_u = \sum_{v\in N^-(u)} D_{v,u}
+\]
+\[
+\forall v\in N^+(u):  D_{u,v} \gets 
+\begin{cases}
+    D_{u,v} & \quad \text{if $D_{u,v}\neq 0$}\\
+    \Sigma^-_u - \Sigma^+_u       & \quad \text{otherwise}
+\end{cases}   
+\]
+\[
+\forall v\in N^-(u):  D_{v,u} \gets 
+\begin{cases}
+    D_{v,u} & \quad \text{if $D_{v,u}\neq 0$}\\
+    \Sigma^+_u - \Sigma^-_u       & \quad \text{otherwise}
+\end{cases}   
+\]
+
 %By construction, the \textit{symbolic experssions} of a basic block, i.e., the \textit{symbolic sum} of the incoming edges (or outgoing edges) of a basic block, has the following properties:
@@
 \end{lstlisting}
+
+Let $x$ be a symbolic expression, and $\kappa_P(x)$ be the coefficient of the term $P$ in $x$.
+In our case, $\kappa_P(x)$ will belong to the set $\{-1,0,1\}$.
+Then, for all probes $P$, we have that
+\[
+w(P) = \sum_u \kappa_P(\Sigma^+_u)w(u)
+\]
 

### 2017-09-12 motivation.tex
minor change
adds formal description of instrumentation

@@
              The vertical lines represent the average speedups over
-             all the 1,000 inputs.
-             }
+             all the 1,000 inputs.}
     \label{fig:motivation-online}
@@
     \includegraphics[width=0.5\textwidth]{figs/motivation-metric.pdf}
-    \caption{Comparison between the offline and online {\itercomp}.
-             The vertical lines represent the average speedups over
-             all the 1,000 inputs.
-             }
+    \caption{Comparison among different performance metrics for performing
+             online {\itercomp}.
+             The vertical lines represent the best score for each metric.}
     \label{fig:motivation-metric}
@@
 
-Figure~\ref{fig:motivation-metric} shows a comparison among these three metrics,
-namely speedup, execution time, and IPC, and their correlation with the actual
-speedup.
+Figure~\ref{fig:motivation-metric} shows a comparison between execution time and IPC,
+and their correlation with the actual speedup.
 Each point in the figure corresponds to each metric averaged over 

### 2017-09-13 abstract.tex
tweaked abstract

@@
     Iterative or adaptive compilation can dramatically improve the performance of a program by searching for the best optimisation
-    settings. However, it relies upon repeatedly evaluating and comparing optimisations off-line on representative inputs. If the
-    developers fail to guess what is a representative input or usage patterns change over time, then the wrong optimisations may be
-    selected. Until this paper, it has been impossible to search online over actual user inputs. Program side effects, interactivity and
-    difficulties capturing the system state prevent repeated execution. Since each input may be executed only once and each input may
-    entail a different amount of work, the search cannot use runtime to compare optimisations. Unless we find a way to compare
-    optimisations even when inputs are executed only once, we will be unable to use real user input for online iterative compilation, users
-    will not have the best optimised programs, and performance and energy efficiency will suffer.
+    settings. However, it relies upon repeatedly evaluating and comparing optimisations off-line on predefined
+    representative inputs. If the developers fail to guess what is a representative input or usage patterns change over time, 
+    then the wrong optimisations may be selected. 
+    Searching the optimisation space with actual user inputs, during deployment, has previously seemed infeasible. 
+    Program side effects, interactivity and difficulties capturing the system state mean that any given input can 
+    only be executed once. Since each input may entail a different amount of work, the search cannot use runtime to compare 
+    optimisations, and there is no baseline against which to define speedup.
+    Unless we find a way to compare optimisations even when inputs are executed only once, we will be unable to use real user 
+    input for iterative compilation, users will not have the best optimised programs, and performance and energy efficiency will suffer.
 
-    This paper demonstrates online iterative compilation. We compare optimisations by work efficiency, or work divided by time, executing
-    each input only once. We develop an instrumented, low overhead, work efficiency metric with optimal probe placement. Then we further
-    reduce the overhead by removing high-cost probes whose absence introduces only small errors to the work metric, without significantly
-    affecting the outcome of the iterative compilation. We give two variants for probe removal optimisation: the first offers strong
+    We present metrics and techniques which, for the first time,  enable the use of true {\em online iterative compilation}.
+    %Until this paper, it has been impossible to search online over actual user inputs. Program side effects, interactivity and
+    %difficulties capturing the system state prevent repeated execution. Since each input may be executed only once and each input may
+    %entail a different amount of work, the search cannot use runtime to compare optimisations. 
+    %This paper demonstrates online iterative compilation. 
+    We compare optimisations by a new work efficiency metric, which requires only a single execution with any one input.
+    We develop an instrumented, low overhead, mechanism for gather the requireed data, using optimal probe placement in the source
+    code. We show that overhead can be further reduced by removing high-cost probes whose absence introduces only small errors to the 
+    work metric, without significantly affecting the outcome of the iterative compilation. 
+    We give two variants for probe removal optimisation: the first offers strong
     guarantees about the maximum possible error, while the second targets the average, whole program error.
 
-    Our online method yields programs that get 80\% of the speedup an offline oracle achieves but without executing any input more than
-    once, which the oracle cannot do. Through probe removal we bring the instrumentation overhead down to only 4\% on average, making our
-    approach suitable for regular use. We show practical online iterative compilation for the first time, optimising programs according to
-    the range of inputs users actually present.
+    Our online iterative compilation method yields programs that get 80\% of the speedup achievable by an offline oracle, 
+    but without executing any input more than once, which the oracle cannot do. 
+    Through probe removal we bring the instrumentation overhead down to only 4\% on average, making our
+    approach suitable for regular use. We show practical online iterative compilation for the first time, 
+    optimising programs according to the range of inputs users actually present.
 

### 2017-09-13 cgo18mscr.tex
minor
updates the instrumentation

@@
 
-\section{The Work Efficiency Metric} \label{sec:metric}
+\section{Work Efficiency Metric} \label{sec:metric}
 
@@
 
-\begin{figure}[htb]
+\begin{figure}[t]
     \centering
@@
 
-In this section we describe how the computation of the work-metric can be performed during runtime by means of instrumenting the code.
-In particular, we adapt the optimal algorithm proposed originally for profiling block frequency~\cite{knuth73,ball94}.
-Afterwards, we propose a relaxed instrumentation that focus on further reducing the overhead by considering the trade-off between profiling accuracy and instrumentation overhead.
+In this section, we describe how the computation of the work metric can be performed during runtime by means of instrumenting the code.
+%In particular, we adapt the optimal algorithm proposed originally for block-frequency profiling~\citep{nahapetian73,knuth73,ball94}.
+Because we define work as a linear equation on the block-frequency counters, it is possible to embed its computation into the execution of the program.
+A naive instrumentation would consist basically of having a global counter that starts with the interception value, $\varepsilon$, and each basic block increments its own cost into the global counter.
+Although this instrumentation is easily implemented, it imposes a significant overhead on the instrumented program.
+However, it is possible to insert fewer probes by carefully placing the probes in a way that is possible to reconstruct the complete profiling information~\citep{knuth73,ball94}.
 
-Because we define work as a linear equation on the block frequency counters, it is possible to embed its computation into the execution of the program.
-A naive instrumentation would consists basically of having a global counter that starts with the interception value, $\varepsilon$, and each basic block increments its own cost into the global counter.
-Although this instrumentation is easily implemented, it inserts a large amount of overhead into the program.
-However, it is possible to insert fewer probes by carefully placing the probes in a way that is possible to reconstruct the complete profiling information~\cite{knuth73,ball94}.
+We adapt the optimal block-frequency instrumentation in order to perform the work profiling efficiently.
+The proposed work instrumentation differs from the optimal block-frequency instrumentation as the latter occurs in two stages:
+\textit{(i.) Before execution:} The code is instrumented with counters for each probe.
+\textit{(ii.) After execution:} The information from the recorded probes is propagated in the CFGs of the program.
+In contrast, our work instrumentation has a single counter, and it only requires the instrumentation before execution, without any post-processing of the recorded profiling.
 
-Ball and Larus~\cite{ball94} show that a set of edges represents the minimum number of probes for profiling block frequency if and only if the complementary set of edges forms a spanning tree.
-The instrumentation optimizes the placement of the probes with respect to a weighing that assigns a non-negative value to each edge in the CFG.
-The cost of profiling a set of edges is proportional to the sum of the weights of the edges.
-These weights can be obtained either by empirical measurements or heuristic estimations.
-In order to minimize the profiling cost, the instrumentation computes the maximum spanning tree for avoiding probing in frequently executed edges.
-Although this instrumentation algorithm is proved to produce the optimal placement of probes for well-structured CFGs, it may produce sub-optimal placement for some unstructured CFGs.
+The instrumented code is assumed to be generated before optimising the code.
+This assumption is based on three key points:
+\textit{(i.)} it guarantees that the work metric is independent of optimisation, i.e., the same input is always mapped to the same amount of work, regardless of the optimisation;
+\textit{(ii.)} it simplifies the code generator;
+\textit{(iii.)} it leverages from the optimisations to further improve the instrumentation code.
+Having a work metric that is independent of optimisation is the most important reason for which we instrument the code before optimisations.
 
-Once we have the maximum spanning tree, probes are placed on any edge not in the spanning tree.
-Figure~\ref{fig:cfg-example} shows an example of a CFG with a maximum spanning tree represented by the black edges, while the edges highlighted in red represent the placement of the probes.
+The optimal placement of the probes works by first computing the maximum spanning
+tree based on a weighing that assigns a non-negative value to each edge in the CFG.
+These weights can be obtained either by empirical measurements or heuristic estimations,
+and their goal is to avoid probing in frequently executed edges.
+Once we have the maximum spanning tree, probes are placed on every edge not in the spanning tree.
+Figure~\ref{fig:cfg-example} shows an example of a CFG with a maximum spanning tree.
+%represented by the black edges, while the edges highlighted in red represent the placement of the probes.
 
-\begin{figure}[htb]
+\begin{figure}[t]
 \centering {
@@
 
-In contrast to the naive instrumentation where each basic block records only its own amount of work, with the optimal profiling, the instrumented basic blocks need to record an aggregated value of work that represents a path in the CFG.
-These values are constructed with some instrumented basic blocks speculatively assuming some paths while other probes correct when these assumptions are wrong (see for example $w(P_0)$ and $w(P_1)$ in Figure~\ref{fig:cfg-example}).
+In contrast to the naive instrumentation where each basic block records only its
+own amount of work, with the optimal profiling, the instrumented basic blocks need
+to record an aggregated value of work that represents a path in the CFG.
+These values are constructed with some instrumented basic blocks speculatively
+assuming some paths while other probes correct when these assumptions are wrong
+(see for example $w(P_0)$ and $w(P_1)$ in Figure~\ref{fig:cfg-example}).
 
-\begin{lstlisting}[caption={Pseudocode of the data-flow analysis necessary for assigning the values computed in each probe.},
-label={lst:populateEdgeInfo}, float]
-// Inputs: CFG with the known edges flows from
-//         the cotree (collected probes).
-// Output: Updated CFG with all edge flows.
-populateEdgeFlows(G) {
-  changed = true
-  while changed:
-    changed = false
-    for B in G.vertices():
-      unIN = count( G.unknownIncomingEdges(B) )
-      unOUT = count( G.unknownOutgoingEdges(B) )
-      if unIN==0 and unOUT==1:
-        //sum known incoming and outgoing edges
-        sIN = sum( G.incomingEdges(B) )
-        sOUT = sum( G.outgoingEdges(B) )
-        //update unknown outgoing edge in B
-        //with (sIN-sOUT)
-        G.setUnknownOutgoingEdge(B, (sIN-sOUT))
-        changed = true
-      if unIN==1 and unOUT==0:
-        //sum known incoming and outgoing edges
-        sIN = sum( G.incomingEdges(B) )
-        sOUT = sum( G.outgoingEdges(B) )
-        //update unknown incoming edge in B
-        //with (sOUT-sIN)
-        G.setUnknownIncomingEdge(B, (sOUT-sIN))
-        changed = true
-}
-\end{lstlisting}
+We compute the aggregated value of work for each probe in two steps:
+\textit{(i.)} we propagate information about the probes through the edges of the CFG,
+as symbolic expressions, and then \textit{(ii.)} we use this edge information
+to compose the aggregated value of the probes.
 
-Because the algorithm for the optimal placement of the probes is proved to uniquely compute the block frequencies by propagating the probe counts, we adapt this algorithm (see \lstlistingname~\ref{lst:populateEdgeFlows}) in order to compose the aggregated values that will be instrumented in each probe, based on our model of computational work, $\Delta W$, derived from the basic block frequencies (see Section~\ref{sec:metric}).
-We perform a similar propagation of the probes in a symbolic fashion, as illustrated in Figure~\ref{fig:cfg-example}.
-This symbolic propagation of the probes is implemented by the data-flow analysis described in \lstlistingname~\ref{lst:populateEdgeInfo}, and the final aggregated values are extracted from the edge information as described by \lstlistingname~\ref{lst:instrValue}.
+The symbolic expressions in the edges can then be used to compose the aggregated
+value of the probes.
+These edge expressions describe the probes that are in the same path of each
+basic block (positive terms) and those that are in complementary paths (negative
+terms).
+Therefore, the positive terms in the edge expression of a basic block indicate
+that the amount of work of this basic block will be incremented in the probes
+represented by these positive terms.
+Similarly, the negative terms indicate that the amount of work of this basic block
+will be decremented in the probes represented by these negative terms.
+For example, because the edge expression for the basic block $B_8$ is $P_0 - P_1$,
+the amount of work of $B_8$, denoted by $w(B_8)$, is incremented in probe $P_0$
+and decremented in $P_1$.
 
-The data-flow analysis in \lstlistingname~\ref{lst:populateEdgeInfo} keeps two sets for each edge, namely, the increment and the decrement sets.
-We consider that both sets represent the \textit{edge expressions} shown in Figure~\ref{fig:cfg-example}, for which we define a \textit{symbolic sum} by computing the union of the increment and decrement sets, respectively, with the appropriate cancellation of common elements.
-This data-flow analysis is based on the invariant that the symbolic sum of all the incoming edges must equals the symbolic sum of the outgoing edges.
-For example, the symbolic sum of the incoming edges of the basic block $B_8$ is $P_0 - P_1$, where $P_3$ is cancelled out.
-
-We first initialize the edge information $D_{u,v}$, for each edge $(u,v)$ in the CFG.
+\noindent\textbf{Populating the edge information:}
+Intuitively, if all the edge flows are known for the complement of a spanning tree
+then at any leaf of the spanning tree there is only one unknown edge flow.
+This unknown edge flow can be calculated by Kirchhoff's first law.
+This process repeats as a bottom-up propagation until all the unknown edge flows
+have been calculated.
+This algorithm first initializes the edge information $D_{(u,v)}$, for each edge
+$(u,v)$ in the CFG, as follows:
 \[
-D_{u,v} \gets 
+D_{(u,v)} \gets
 \begin{cases}
@@
 \]
-Afetwards, for each vertex $u$, in post-order of the spanning tree, we apply the following operations:
+Afetwards, for each vertex $u$, in post-order traversal of the spanning tree,
+we can use the Kirchhoff's first law and apply the following operations
+in a symbolic fashion:
 \[
-\Sigma^+_u = \sum_{v\in N^+(u)} D_{u,v}
+\Sigma^+_u = \sum_{v\in N^+(u)} D_{(u,v)}
 \]
 \[
-\Sigma^-_u = \sum_{v\in N^-(u)} D_{v,u}
+\Sigma^-_u = \sum_{v\in N^-(u)} D_{(v,u)}
 \]
 \[
-\forall v\in N^+(u):  D_{u,v} \gets 
+\forall v\in N^+(u):  D_{(u,v)} \gets
 \begin{cases}
-    D_{u,v} & \quad \text{if $D_{u,v}\neq 0$}\\
+    D_{(u,v)} & \quad \text{if $D_{(u,v)}\neq 0$}\\
     \Sigma^-_u - \Sigma^+_u       & \quad \text{otherwise}
@@
 \[
-\forall v\in N^-(u):  D_{v,u} \gets 
+\forall v\in N^-(u):  D_{(v,u)} \gets
 \begin{cases}
-    D_{v,u} & \quad \text{if $D_{v,u}\neq 0$}\\
+    D_{(v,u)} & \quad \text{if $D_{(v,u)}\neq 0$}\\
     \Sigma^+_u - \Sigma^-_u       & \quad \text{otherwise}
@@
 
-%By construction, the \textit{symbolic experssions} of a basic block, i.e., the \textit{symbolic sum} of the incoming edges (or outgoing edges) of a basic block, has the following properties:
-%\begin{enumerate}
-%\item Probes in the increment set either (post-)dominates or is (post-)dominated by the given basic block;
-%\item Probes in the decrement set are (post-)dominated by the probes in the increment set;
-%\item Probes in the decrement set and the given basic block are in exclusive paths from the virtual node of the CFG to the (post-)dominators in the increment set;
-%\end{enumerate}
-%From property (1) it follows that whenever the given basic block is executed, exactly one of the probes in the increment set is executed.
-%From properties (2) and (3), it follows that whenever a probe in the decrement set is executed, although the given basic block is not executed, one of the probes in the decrement set is executed.
-
-\lstlistingname~\ref{lst:instrValue} reads the edge information for each basic block by computing the symbolic sum of their respective incoming edges (or outgoing edges).
-From these \textit{edge expressions}, we are able to compose the aggregated value of the probes.
-The positive terms in the edge expression of a basic block indicate that the amount of work of this basic block will be incremented in the probes represented by these positive terms.
-Similarly, the negative terms indicate that the amount of work of this basic block will be decremented in the probes represented by these negative terms.
-For example, because the edge expression for the basic block $B_8$ is $P_0 - P_1$, the amount of work of $B_8$, denoted by $w(B_8)$, is incremented in probe $P_0$ and decremented in $P_1$.
-
-%Let $V$ be a vector of $\{-1,0,1\}$ with the sum of the edge information from the incoming edges of a given basic block.
-%Let $W$ be the 
-%\[
-%\mathrm{reduce}\,\,(+)\,\,(\mathrm{map}\,\,(\times W)\,\,E^+(B))
-%\]
-
-\begin{lstlisting}[caption={Pseudocode that describes how the edge information is used in order to extract the value that will be computed in a given instrumented basic block $B_I$.
-This algorithm could equally be implemented based on the outgoing edges.}, label={lst:instrValue}, float]
-// Inputs: 1) CFG with the known edges flows
-//            of the chords.
-//         2) Probe to be instrumented.
-// Output: Work value to be incremented by the
-//         given probe P.
-instrValue(G, P) {
-  value = 0
-  for B in G.vertices():
-    expr = sum( G.incomingEdges(B) ) 
-    //coef(P,expr) is the coefficient
-    //-1,0, or 1 of P in expr
-    value = value + coef(P, expr)*w(B)
-  return value
-}
-\end{lstlisting}
-
-Let $x$ be a symbolic expression, and $\kappa_P(x)$ be the coefficient of the term $P$ in $x$.
+\noindent\textbf{Composing the aggregated value of the probes:}
+Let $x$ be a symbolic expression, and $\kappa_P(x)$ be the coefficient of the
+term $P$ in $x$.
 In our case, $\kappa_P(x)$ will belong to the set $\{-1,0,1\}$.

### 2017-09-13 cgo18mscr.tex
change motivation-metric figure; minor updates to text;

@@
      \begin{align*}
-     w(P_0) &= w(B_0) + w(B_1) + w(B_4) + w(B_6) + w(B_7) + w(B_8) + w(B_9)\\
-     w(P_1) &= w(B_2) - w(B_1) - w(B_4) - w(B_6) - w(B_7) - w(B_8)\\
-     w(P_2) &= w(B_5) - w(B_6)\\
-     w(P_3) &= w(B_3) - w(B_4) - w(B_6) - w(B_7)
+     \omega(P_0) &= w(B_0) + w(B_1) + w(B_4) + w(B_6) + w(B_7) + w(B_8) + w(B_9)\\
+     \omega(P_1) &= w(B_2) - w(B_1) - w(B_4) - w(B_6) - w(B_7) - w(B_8)\\
+     \omega(P_2) &= w(B_5) - w(B_6)\\
+     \omega(P_3) &= w(B_3) - w(B_4) - w(B_6) - w(B_7)
      \end{align*}
@@
 have been calculated.
+Let $G$ be the CFG.
 This algorithm first initializes the edge information $D_{(u,v)}$, for each edge
@@
 \begin{cases}
-    P_{u,v} & \quad \text{if edge $(u,v)$ has a probe}\\
+    P_{(u,v)} & \quad \text{if edge $(u,v)$ has a probe $P_{(u,v)}$}\\
     0       & \quad \text{otherwise}
@@
 \]
-Afetwards, for each vertex $u$, in post-order traversal of the spanning tree,
+Afetwards, for each vertex $u\in G$, in post-order traversal of the spanning tree,
 we can use the Kirchhoff's first law and apply the following operations
@@
 \noindent\textbf{Composing the aggregated value of the probes:}
-Let $x$ be a symbolic expression, and $\kappa_P(x)$ be the coefficient of the
-term $P$ in $x$.
-In our case, $\kappa_P(x)$ will belong to the set $\{-1,0,1\}$.
+If $x$ is a symbolic expression, then $\kappa_P(x)$ represents the coefficient
+of the term $P$ in $x$.
+In our case, $\kappa_P(x)$ will belong to the set $\{-1,0,1\}$, i.e.,
+
 Then, for all probes $P$, we have that
 \[
-w(P) = \sum_u \kappa_P(\Sigma^+_u)w(u)
+\omega(P) = \sum_{u\in G} \kappa_P(\Sigma^+_u)w(u)
 \]
@@
 First, the algorithm extracts all the subgraphs that represent a loop or the outer most region of the function.
-Section~\ref{sec:natural-loops} explains how loops can be identified in the CFG.
 Afterwards, these subgraphs are transformed into DAGs by ignoring the backedge and also by considering that any loop within the subgraph is never executed, i.e., only the headers of the inner loops are actually included into the DAG.

### 2017-09-13 cgo18mscr.tex
minor updates to the Section: Work Profiling.

@@
 
-\section{Instrumentation}
+\section{Work Profiling}
 
@@
 
-We adapt the optimal block-frequency instrumentation in order to perform the work profiling efficiently.
-The proposed work instrumentation differs from the optimal block-frequency instrumentation as the latter occurs in two stages:
+We adapt the optimal block-frequency profiling in order to perform the work profiling efficiently.
+The proposed work profiling differs from the optimal block-frequency profiling as the latter occurs in two stages:
 \textit{(i.) Before execution:} The code is instrumented with counters for each probe.
 \textit{(ii.) After execution:} The information from the recorded probes is propagated in the CFGs of the program.
-In contrast, our work instrumentation has a single counter, and it only requires the instrumentation before execution, without any post-processing of the recorded profiling.
+In contrast, our work profiling has a single counter, and it only requires the instrumentation before execution, without any post-processing of the recorded profiling.
 
@@
 The optimal placement of the probes works by first computing the maximum spanning
-tree based on a weighing that assigns a non-negative value to each edge in the CFG.
+tree based on a weighing that assigns a value to each edge in the CFG.
 These weights can be obtained either by empirical measurements or heuristic estimations,
@@
 own amount of work, with the optimal profiling, the instrumented basic blocks need
-to record an aggregated value of work that represents a path in the CFG.
-These values are constructed with some instrumented basic blocks speculatively
-assuming some paths while other probes correct when these assumptions are wrong
+to record an aggregated value of work that also takes other basic blocks into
+account.
+Some probes speculatively assume a path of execution,
+while other probes correct when these assumptions are wrong
 (see for example $w(P_0)$ and $w(P_1)$ in Figure~\ref{fig:cfg-example}).
@@
 \textit{(i.)} we propagate information about the probes through the edges of the CFG,
-as symbolic expressions, and then \textit{(ii.)} we use this edge information
+as symbolic expressions, and then \textit{(ii.)} we use this edge flows
 to compose the aggregated value of the probes.
@@
 value of the probes.
-These edge expressions describe the probes that are in the same path of each
-basic block (positive terms) and those that are in complementary paths (negative
+These symbolic expressions describe the probes that are in the same path of the
+edge (positive terms) and those that are in complementary paths (negative
 terms).
-Therefore, the positive terms in the edge expression of a basic block indicate
-that the amount of work of this basic block will be incremented in the probes
-represented by these positive terms.
+Therefore, the positive terms in the summed symbolic expression of a
+basic block indicate that the amount of work of this basic block will be
+incremented in the probes represented by these positive terms.
 Similarly, the negative terms indicate that the amount of work of this basic block
 will be decremented in the probes represented by these negative terms.
-For example, because the edge expression for the basic block $B_8$ is $P_0 - P_1$,
+For example, because the symbolic expression for the basic block $B_8$ is $P_0 - P_1$,
 the amount of work of $B_8$, denoted by $w(B_8)$, is incremented in probe $P_0$
@@
 
-\noindent\textbf{Populating the edge information:}
+\noindent\textbf{Populating the edge flows:}
 Intuitively, if all the edge flows are known for the complement of a spanning tree
@@
 have been calculated.
+This algorithm can be formally defined as a post-order traversal on the spanning
+tree.
 Let $G$ be the CFG.
-This algorithm first initializes the edge information $D_{(u,v)}$, for each edge
+This algorithm first initializes the edge flows $D_{(u,v)}$, for each edge
 $(u,v)$ in the CFG, as follows:
@@
 \]
-Afetwards, for each vertex $u\in G$, in post-order traversal of the spanning tree,
+Afterwards, for each vertex $u\in G$, in a post-order traversal of the spanning tree,
 we can use the Kirchhoff's first law and apply the following operations

### 2017-09-13 motivation.tex
minor
motivation first part

@@
 
-%As a motivation example, consider...
+Offline \itercomp can be very sensitive on having representative inputs for the program under optimization. If the
+developers cannot predict what will be the typical usage of the application or that typical usage changes over time,
+then \itercomp might fail making the application faster for the user.
 
-%Here we give an example to motivate our work. E.g. assuming we have a zero-cost profiling based work metric,
-%how can we use it to perform iterative compilation across runs? Then we talk about while it is impossible to
-%build a zero-overhead metric, we can develop one with low overhead.
+To show this, we randomly generated multiple
+input subsets out of the full set of XXX inputs. Working with each subset, \itercomp found a ``best'' optimization
+sequence which maximized the speedup for the given input subset. We evaluated the true quality of each such sequence
+by measuring the average speedup against \texttt{-O3} for all the XXX inputs. The green bars in
+Figure~\ref{fig:motivation-online} show the histogram of those speedups. Depending on the selection of inputs,
+performance can degrade by 18\% compared to \texttt{-O3}, improve by 46\%, or anything in between. Overall, almost half
+of the optimization sequences discovered fail to improve performance. This means that without intelligent selection of
+inputs, offline \itercomp might be a waste of time.
 
-This section provides a comparison between the offline and online approaches for
-{\itercomp}.
-The goal is to motivate the need for an online {\itercomp}, highlighting
-weaknesses of using offline {\itercomp}.
+The blue bars in Figure~\ref{fig:motivation-online} show a different approach where during \itercomp the speedup of
+each optimization sequence is again measured on a small input subset but this subset changes for every optimization
+sequence. This is similar to how online \itercomp would have to work, using each input only once. This might introduce
+errors, since the speedups we measure for each optimization sequence are not strictly comparable, they are measured on
+different workloads. But in the end the histogram shows that being able to test more inputs allows us to make better
+informed decisions. This approach never degrades performance compared to \textt{-O3}, performance varies less, and on
+average is 13\% better than the offline approach.
 
-%In the offline approach, a fixed set of inputs is selected for which the best
-%optimization sequence is selected during the development time (pre-shipping) of
-%a program.
-%Although this fixed set of inputs is expected to be representative of the user's
-%workload,
-%in most cases, it might be infeasible to identify and collect representative
-%workloads for every group of users.
-%Furthermore, a user can dynamically change its workload pattern, rendering
-%a fixed optimization sequence innapropriate for his performance requirements.
+% %As a motivation example, consider...
+% 
+% %Here we give an example to motivate our work. E.g. assuming we have a zero-cost profiling based work metric,
+% %how can we use it to perform iterative compilation across runs? Then we talk about while it is impossible to
+% %build a zero-overhead metric, we can develop one with low overhead.
+% 
+% This section provides a comparison between the offline and online approaches for
+% {\itercomp}.
+% The goal is to motivate the need for an online {\itercomp}, highlighting
+% weaknesses of using offline {\itercomp}.
+% 
+% %In the offline approach, a fixed set of inputs is selected for which the best
+% %optimization sequence is selected during the development time (pre-shipping) of
+% %a program.
+% %Although this fixed set of inputs is expected to be representative of the user's
+% %workload,
+% %in most cases, it might be infeasible to identify and collect representative
+% %workloads for every group of users.
+% %Furthermore, a user can dynamically change its workload pattern, rendering
+% %a fixed optimization sequence innapropriate for his performance requirements.
+% 
+% %In the online approach, the program is shipped with an initial optimization
+% %sequence and different optimization sequences are evaluated as the end-user
+% %executes the program.
+% %This approach frees developers from the need to collect representative
+% %workloads for every group of users so that the program can be optimized
+% %accordingly.
+% %More importantly, even in scenarios where the user's workload can dynamically
+% %change its pattern, the online {\itercomp} would still be able to optimize the
+% %program accordingly.
+% 
+% In the offline approach, a small fixed set of inputs is selected for which all
+% optimization sequences are evaluated.
+% In the online approach, each optimization sequence is evaluated based on a,
+% dynamically selected, distinct set of inputs.
+% In both cases, the optimization sequences are ranked based on their average
+% speedup, and the best optimization sequence is selected.
+% 
+% In order to assess the quality of the optimization sequence selected by each
+% approach, we evaluate both optimization sequences against all available inputs
+% in our dataset.
+% Figure~\ref{fig:motivation-online} shows this comparison between both the offline
+% and online approaches for {\itercomp}.
+% It shows the performance gains of both optimization sequences compared with the
+% standard compiler's optimization (i.e.,\texttt{-O3}).
+% The online {\itercomp} achieves an average speedup which is about 13\% better
+% than the solution selected by the offline approach.
 
-%In the online approach, the program is shipped with an initial optimization
-%sequence and different optimization sequences are evaluated as the end-user
-%executes the program.
-%This approach frees developers from the need to collect representative
-%workloads for every group of users so that the program can be optimized
-%accordingly.
-%More importantly, even in scenarios where the user's workload can dynamically
-%change its pattern, the online {\itercomp} would still be able to optimize the
-%program accordingly.
-
-In the offline approach, a small fixed set of inputs is selected for which all
-optimization sequences are evaluated.
-In the online approach, each optimization sequence is evaluated based on a,
-dynamically selected, distinct set of inputs.
-In both cases, the optimization sequences are ranked based on their average
-speedup, and the best optimization sequence is selected.
-
-In order to assess the quality of the optimization sequence selected by each
-approach, we evaluate both optimization sequences against all available inputs
-in our dataset.
-Figure~\ref{fig:motivation-online} shows this comparison between both the offline
-and online approaches for {\itercomp}.
-It shows the performance gains of both optimization sequences compared with the
-standard compiler's optimization (i.e.,\texttt{-O3}).
-The online {\itercomp} achieves an average speedup which is about 13\% better
-than the solution selected by the offline approach.
-
-\begin{figure}[ht]
+\begin{figure}[t!]
     \centering
@@
 
-It is worth mentioning that our goal is to improve the average performance
-across the user's workload, and not only for a specific input.
-This means that the online {\itercomp} is better informed throughout the search,
-as it compares optimization sequences across different sets inputs.
+% It is worth mentioning that our goal is to improve the average performance
+% across the user's workload, and not only for a specific input.
+% This means that the online {\itercomp} is better informed throughout the search,
+% as it compares optimization sequences across different sets inputs.
 
@@
 
-\begin{figure}[ht]
+\begin{figure}[t]
     \centering

### 2017-09-13 motivation.tex
fix

@@
 
-The blue bars in Figure~\ref{fig:motivation-online} show a different approach where during \itercomp the speedup of
-each optimization sequence is again measured on a small input subset but this subset changes for every optimization
-sequence. This is similar to how online \itercomp would have to work, using each input only once. This might introduce
-errors, since the speedups we measure for each optimization sequence are not strictly comparable, they are measured on
-different workloads. But in the end the histogram shows that being able to test more inputs allows us to make better
-informed decisions. This approach never degrades performance compared to \textt{-O3}, performance varies less, and on
-average is 13\% better than the offline approach.
+The blue bars in Figure~\ref{fig:motivation-online} show a different approach where during \itercomp the speedup of each optimization
+sequence is again measured on a small input subset but this subset changes for every optimization sequence. This is similar to how online
+\itercomp would have to work, using each input only once. This might introduce errors, since the speedups we measure for each optimization
+sequence are not strictly comparable, they are measured on different workloads. But in the end the histogram shows that being able to test
+more inputs allows us to make better informed decisions. This approach never degrades performance compared to \texttt{-O3}, performance
+varies less, and on average is 13\% better than the offline approach.
 

### 2017-09-13 motivation.tex
motivation fix

@@
 
-Offline \itercomp can be very sensitive on having representative inputs for the program under optimization. If the
-developers cannot predict what will be the typical usage of the application or that typical usage changes over time,
-then \itercomp might fail making the application faster for the user.
+Offline \itercomp can be very sensitive on having representative inputs for the program under optimization. If the developers cannot
+predict what will be the typical usage of the application or that typical usage changes over time, then \itercomp might fail making the
+application faster for the user.
 
-To show this, we randomly generated multiple
-input subsets out of the full set of XXX inputs. Working with each subset, \itercomp found a ``best'' optimization
-sequence which maximized the speedup for the given input subset. We evaluated the true quality of each such sequence
-by measuring the average speedup against \texttt{-O3} for all the XXX inputs. The green bars in
-Figure~\ref{fig:motivation-online} show the histogram of those speedups. Depending on the selection of inputs,
-performance can degrade by 18\% compared to \texttt{-O3}, improve by 46\%, or anything in between. Overall, almost half
-of the optimization sequences discovered fail to improve performance. This means that without intelligent selection of
-inputs, offline \itercomp might be a waste of time.
+To show this, we performed offline \itercomp on XXX using a small subset of its inputs and we identified the optimization sequence that
+maximizes performance. In Figure~\ref{fig:motivation-online}, the green bars are the histogram of speedups compared to \texttt{-O3} for
+every individual input, while the dark green line shows the speedup averaged over all inputs. Despite the selected optimization sequence
+being the best for some inputs, we see that on average it improves performance by only 6\% and hurts performance for almost half the
+inputs. Just randomly choosing an input set for offline \itercomp can be counterproductive.
 
@@
 sequence are not strictly comparable, they are measured on different workloads. But in the end the histogram shows that being able to test
-more inputs allows us to make better informed decisions. This approach never degrades performance compared to \texttt{-O3}, performance
-varies less, and on average is 13\% better than the offline approach.
+more inputs allows us to make better informed decisions. The optimization sequence we selected never degrades performance compared to
+\texttt{-O3} and is on average 13\% better than the offline approach.
 

### 2017-09-13 motivation.tex
motivation done

@@
 application faster for the user.
+
+\begin{figure}[t!]
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/motivation-online.pdf}
+    \caption{Histograms of speedup over \texttt{-O3} for each of the 1,000 inputs of XXX. The green histogram shows the speedup when using
+    the optimization sequence selected by offline \itercomp, the blue one is the speedup using the optimization sequence selected by an
+    online-like approach.}
+    \label{fig:motivation-online}
+\end{figure}
 
@@
 \texttt{-O3} and is on average 13\% better than the offline approach.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/motivation-metric.pdf}
+    \caption{Relationship between IPC and execution time vs optimization speedup for XXX different binaries of XXX. Each point represents
+	averages over all inputs.}
+    \label{fig:motivation-metric}
+\end{figure}
+
+This is not an entirely surprising result. Evaluating our optimization decisions on more inputs decreases the chances of optimizing for an
+unrepresentative input. Despite that offline \itercomp is still the preferred way. Measuring the speedup caused by an optimization requires
+running the same input \textit{at least twice}, one of the optimized and one for the baseline binary. This is either infeasible or
+undesirable in a realistic online \itercomp scenario. There are two existing workarounds to this problem: directly comparing the
+\textit{execution time} of each run against previous runs or directly comparing the \textit{instructions per cycle}~(IPC)~\citep{fursin07}.
+Neither metric is a good proxy for speedup. Figure~\ref{fig:motivation-metric} plots the measured IPC and execution time versus the real
+speedup for XXX different binaries of the XXX benchmark. IPC, execution time, and speedup are averaged over all inputs. Both metrics
+correlate little with speedup. Different inputs require a different amount of work which dominates the changes in execution time. IPC on
+the other hand might improve with a more efficient binary but the opposite can also be true. For example, inserting \texttt{NOP} statements
+will usually improve IPC.
+
+In the remaining of this paper, we examine our work efficiency metric, a novel way for approximating optimization speedup through a single
+run which can be used to drive an online \itercomp methodology.
 
@@
 
-\begin{figure}[t!]
-    \centering
-    \includegraphics[width=0.5\textwidth]{figs/motivation-online.pdf}
-    \caption{Comparison between the offline and online {\itercomp}.
-             The vertical lines represent the average speedups over
-             all the 1,000 inputs.}
-    \label{fig:motivation-online}
-\end{figure}
+% \begin{figure}[t!]
+%     \centering
+%     \includegraphics[width=0.5\textwidth]{figs/motivation-online.pdf}
+%     \caption{Comparison between the offline and online {\itercomp}.
+%              The vertical lines represent the average speedups over
+%              all the 1,000 inputs.}
+%     \label{fig:motivation-online}
+% \end{figure}
 
@@
 
-However, because of the restriction of having a single execution per input,
-it is not possible to measure speedup in a real-world online scenario.
-Moreover, measuring just execution time, for example, is also not viable since
-different inputs often mean different amounts of work,
-rendering it meaningless to compare optimizations only by execution time.
-Therefore, we need a performance metric which can be profiled from a single
-execution of the program.
-Although a previous work has suggested that \textit{instructions per cycle}~(IPC)
-could be used with that goal, IPC have no correlation with speedup~\citep{fursin07}.
-In a case where we have two distinct versions of the program, due to different
-optimizations, which is also executed with distinct inputs, higher IPC does not
-necessarily mean lower execution time (higher performance).
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.5\textwidth]{figs/motivation-metric.pdf}
-    \caption{Comparison among different performance metrics for performing
-             online {\itercomp}.
-             The vertical lines represent the best score for each metric.}
-    \label{fig:motivation-metric}
-\end{figure}
-
-Figure~\ref{fig:motivation-metric} shows a comparison between execution time and IPC,
-and their correlation with the actual speedup.
-Each point in the figure corresponds to each metric averaged over
-a set of inputs, collected during the {\itercomp}.
-The vertical lines correspond to the optimization sequence
-with the best score for each one of the metrics.
-This figure shows that both execution time and IPC are unsuitable for
-online {\itercomp}, as they present no direct correlation with the actual
-speedup obtained by the optimization sequence being evaluated.
-
-In the remaining of this paper, we develop a work efficiency metric, suitable
-for guiding online {\itercomp}, which can be profiled with a single execution
-of the program with the optimization sequence being evaluated.
+% However, because of the restriction of having a single execution per input,
+% it is not possible to measure speedup in a real-world online scenario.
+% Moreover, measuring just execution time, for example, is also not viable since
+% different inputs often mean different amounts of work,
+% rendering it meaningless to compare optimizations only by execution time.
+% Therefore, we need a performance metric which can be profiled from a single
+% execution of the program.
+% Although a previous work has suggested that \textit{instructions per cycle}~(IPC)
+% could be used with that goal, IPC have no correlation with speedup~\citep{fursin07}.
+% In a case where we have two distinct versions of the program, due to different
+% optimizations, which is also executed with distinct inputs, higher IPC does not
+% necessarily mean lower execution time (higher performance).
+%
+% \begin{figure}[t]
+%     \centering
+%     \includegraphics[width=0.5\textwidth]{figs/motivation-metric.pdf}
+%     \caption{Comparison among different performance metrics for performing
+%              online {\itercomp}.
+%              The vertical lines represent the best score for each metric.}
+%     \label{fig:motivation-metric}
+% \end{figure}
+%
+% Figure~\ref{fig:motivation-metric} shows a comparison between execution time and IPC,
+% and their correlation with the actual speedup.
+% Each point in the figure corresponds to each metric averaged over
+% a set of inputs, collected during the {\itercomp}.
+% The vertical lines correspond to the optimization sequence
+% with the best score for each one of the metrics.
+% This figure shows that both execution time and IPC are unsuitable for
+% online {\itercomp}, as they present no direct correlation with the actual
+% speedup obtained by the optimization sequence being evaluated.
+% 
+% In the remaining of this paper, we develop a work efficiency metric, suitable
+% for guiding online {\itercomp}, which can be profiled with a single execution
+% of the program with the optimization sequence being evaluated.

### 2017-09-13 cgo18mscr.tex
minor fix
updates section on the relaxation strategy

@@
 of the term $P$ in $x$.
-In our case, $\kappa_P(x)$ will belong to the set $\{-1,0,1\}$, i.e.,
-
+In our case, $\kappa_P(x)$ will belong to the set $\{-1,0,1\}$.
 Then, for all probes $P$, we have that
@@
 
-Although the optimal instrumentation significantly reduces the profiling overhead when compared to the naive instrumentation, from an average overhead of 79\% to 13\%, in some critical cases, even the optimal instrumentation can have an overhead of about 70\% (see benchmark \texttt{adpcm\_d} in Figure~\ref{fig:overhead-O3}).
-In order to further reduce the overhead in these critical cases, we propose a relaxed instrumentation by trading off accuracy and overhead.
-While the optimal placement of probes tries to place probes in edges that are less likely to be executed, the relaxation focus on removing probes that are more likely to be executed but add little to the work metric.
+Although the optimal instrumentation significantly reduces the profiling
+overhead when compared to the naive instrumentation, from an average overhead
+of 79\% to 13\%, in some critical cases, even the optimal instrumentation can
+result in overheads of up to 60\% (see benchmark \texttt{adpcm\_d} in Figure~\ref{fig:overhead-O3}).
+In order to further reduce the overhead in these critical cases, we propose a
+relaxation strategy that offers a trade-off between accuracy and overhead.
+While the optimal placement of probes tries to place probes in edges that are
+less likely to be executed, the relaxation focus on removing probes that are
+more likely to be executed but add little to the final work metric.
 
@@
 %The highlighted step is introduced by the relaxed instrumentation on top of the previously defined optimal profiling.
-The relaxed instrumentation performs a post processing on the resulting instrumentation of the optimal algorithm.
-This post-processing identifies probes that add little to the work metric, and removes their instrumentation.
-In order to guarantee an upper bound for the dynamic error in the profiling measurement, the relaxation algorithm applies a constrained post-processing on a per DAG (directed acyclic graph) basis.
-By constraining the relaxation within each DAG by a maximum allowed static error, it guarantees that the overall relaxation will also be constrained by the same bound.
+The relaxation strategy performs a post processing on the resulting instrumentation of the optimal algorithm.
+This post processing identifies probes that add little to the work metric, and removes their instrumentation.
+In order to guarantee an upper bound for the dynamic error in the profiling measurement,
+the relaxation algorithm applies a constrained post-processing on a per DAG (directed acyclic graph) basis.
+By constraining the relaxation within each DAG by a maximum allowed static error,
+it guarantees that the overall relaxation will also be constrained by the same bound.
 
-The relaxation starts by extracting DAGs (directed acyclic graphs) from the CFG, as illustrated in Figure~\ref{fig:cfg-relax-example}.
+The relaxation starts by extracting DAGs from the CFG, as illustrated in Figure~\ref{fig:cfg-relax-example}.
 First, the algorithm extracts all the subgraphs that represent a loop or the outer most region of the function.
-Afterwards, these subgraphs are transformed into DAGs by ignoring the backedge and also by considering that any loop within the subgraph is never executed, i.e., only the headers of the inner loops are actually included into the DAG.
-Figure~\ref{fig:cfg-relax-example} shows a CFG partitioned into two DAGs (consider only basic blocks and edges completely inside the yellow and green boundaries).
+Afterwards, these subgraphs are transformed into DAGs by ignoring the backedge
+and also by considering that any loop within the subgraph is never executed,
+i.e., only the headers of the inner loops are actually included into the DAG.
+Figure~\ref{fig:cfg-relax-example} shows a CFG partitioned into two DAGs (consider
+only basic blocks and edges completely inside the yellow and green boundaries).
 
@@
 
-For every DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we relax the profiling accuracy by selecting a subset of the probes to be removed, subject to the maximum allowed percentage error, $M$.
-
+For every DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we relax the
+profiling accuracy by selecting a subset of the probes to be removed, subject to
+the maximum allowed percentage error, $M$.
 We model the relaxation as a 0-1 Knapsack problem:
@@
 \end{gather*}
-where $f(P_i)$ is the execution frequency of probe $P_i$, $x_i$ denotes the probes selected for removal, and $\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to the minimum work value possible to compute in the DAG, i.e., if $m$ is the minimum amount of work possible to be computed when executing the DAG, then $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$.
-Because the percentage error is computed based on the path with the minimum amount of work, $\varepsilon(P_i)$ represents the maximum error possible that would be incurred when removing probe $P_i$.
- Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold.
+where $f(P_i)$ is the execution frequency of probe $P_i$, $x_i$ denotes the
+probes selected for removal, and $\varepsilon(P_i)$ is the percentage error of
+removing probe $P_i$ relative to the minimum work value possible to compute in
+the DAG, i.e., if $m$ is the minimum amount of work possible to be computed when
+executing the DAG, then $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$.
+Because the percentage error is computed based on the path with the minimum
+amount of work, $\varepsilon(P_i)$ represents the maximum error possible that
+would be incurred when removing probe $P_i$.
+Furthermore, by constraining the percentage error of every DAG below a given
+threshold, we guarantee that the final error of the relaxation will always be
+bounded by the threshold.
 % Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold, as demonstrated by Proposition~\ref{prop:relax-bound}.
@@
 
-The necessary block-frequency information for optimising both the placement of probes can be acquired from profiles of previous executions of the program or by a static heuristic of the CFG during compilation.
+%The necessary block-frequency information for optimising both the placement of
+%probes can be acquired from profiles of previous executions of the program or by
+%a static heuristic of the CFG during compilation.
 
 For our experiments, we implemented two solvers for the 0-1 Knapsack problem:
-the optimal brute-force solver;
+the optimal brute-force solver and
 the greedy heuristic based on sorting the items~\citep{dantzig57}.
-We use the brute-force solver for DAGs with a small number of probes and the greedy heuristic when the number of probes is greater than a threshold.
-Some of the benchmarks have DAGs with several hundreds of probes, which could result in a long compilation time.
+We use the brute-force solver for DAGs with a small number of probes and the
+greedy heuristic when the number of probes is greater than a threshold.
+Some of the benchmarks have DAGs with several hundreds of probes, which could
+result in a long compilation time.
 

### 2017-09-13 abstract.tex
Americanizzzation and Spling

@@
 
-    Iterative or adaptive compilation can dramatically improve the performance of a program by searching for the best optimisation
-    settings. However, it relies upon repeatedly evaluating and comparing optimisations off-line on predefined
+    Iterative or adaptive compilation can dramatically improve the performance of a program by searching for the best optimization
+    settings. However, it relies upon repeatedly evaluating and comparing optimizations off-line on predefined
     representative inputs. If the developers fail to guess what is a representative input or usage patterns change over time, 
-    then the wrong optimisations may be selected. 
-    Searching the optimisation space with actual user inputs, during deployment, has previously seemed infeasible. 
+    then the wrong optimizations may be selected. 
+    Searching the optimization space with actual user inputs, during deployment, has previously seemed infeasible. 
     Program side effects, interactivity and difficulties capturing the system state mean that any given input can 
     only be executed once. Since each input may entail a different amount of work, the search cannot use runtime to compare 
-    optimisations, and there is no baseline against which to define speedup.
-    Unless we find a way to compare optimisations even when inputs are executed only once, we will be unable to use real user 
-    input for iterative compilation, users will not have the best optimised programs, and performance and energy efficiency will suffer.
+    optimizations, and there is no baseline against which to define speedup.
+    Unless we find a way to compare optimizations even when inputs are executed only once, we will be unable to use real user 
+    input for iterative compilation, users will not have the best optimized programs, and performance and energy efficiency will suffer.
 
     We present metrics and techniques which, for the first time,  enable the use of true {\em online iterative compilation}.
-    %Until this paper, it has been impossible to search online over actual user inputs. Program side effects, interactivity and
-    %difficulties capturing the system state prevent repeated execution. Since each input may be executed only once and each input may
-    %entail a different amount of work, the search cannot use runtime to compare optimisations. 
-    %This paper demonstrates online iterative compilation. 
-    We compare optimisations by a new work efficiency metric, which requires only a single execution with any one input.
-    We develop an instrumented, low overhead, mechanism for gather the requireed data, using optimal probe placement in the source
+    We compare optimizations by a new work efficiency metric, which requires only a single execution with any one input.
+    We develop an instrumented, low overhead, mechanism for gathering the required data, using optimal probe placement in the source
     code. We show that overhead can be further reduced by removing high-cost probes whose absence introduces only small errors to the 
     work metric, without significantly affecting the outcome of the iterative compilation. 
-    We give two variants for probe removal optimisation: the first offers strong
+    We give two variants for probe removal optimization: the first offers strong
     guarantees about the maximum possible error, while the second targets the average, whole program error.
@@
     approach suitable for regular use. We show practical online iterative compilation for the first time, 
-    optimising programs according to the range of inputs users actually present.
+    optimizing programs according to the range of inputs users actually present.
 

### 2017-09-13 abstract.tex
And some things to fix

@@
 
-    Our online iterative compilation method yields programs that get 80\% of the speedup achievable by an offline oracle, 
+    Our online iterative compilation method yields programs that get \FIXME{80\%} of the speedup achievable by an offline oracle, 
     but without executing any input more than once, which the oracle cannot do. 
-    Through probe removal we bring the instrumentation overhead down to only 4\% on average, making our
+    Through probe removal we bring the instrumentation overhead down to only \FIXME{4\%} on average, making our
     approach suitable for regular use. We show practical online iterative compilation for the first time, 

### 2017-09-13 cgo18mscr.tex
minor changes

@@
 In our case, $\kappa_P(x)$ will belong to the set $\{-1,0,1\}$.
-Then, for all probes $P$, we have that
+Then, for all probes $P$:
 \[
@@
 relaxation strategy that offers a trade-off between accuracy and overhead.
-While the optimal placement of probes tries to place probes in edges that are
-less likely to be executed, the relaxation focus on removing probes that are
+While the optimal instrumentation places probes in edges that are
+less likely to be executed, the relaxation removes probes that are
 more likely to be executed but add little to the final work metric.
@@
 
-The relaxation starts by extracting DAGs from the CFG, as illustrated in Figure~\ref{fig:cfg-relax-example}.
+The relaxation starts by extracting DAGs from the CFG.
 First, the algorithm extracts all the subgraphs that represent a loop or the outer most region of the function.
@@
   \includegraphics[scale=0.75]{figs/cfg-relax-example.pdf}
-  \caption{Example of a CFG containing a loop and its decomposition into DAGs when applying the relaxation.
+  \caption{Example of a CFG containing a loop and its decomposition into DAGs.
            The DAGs are the subgraphs within the dashed boundaries.}
@@
 
-For every DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we relax the
-profiling accuracy by selecting a subset of the probes to be removed, subject to
+For every DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$,
+we select a subset of the probes to be removed, subject to
 the maximum allowed percentage error, $M$.
-We model the relaxation as a 0-1 Knapsack problem:
+This strategy can be modeled as a 0-1 Knapsack problem:
 \begin{gather*}
@@
 
-In some cases, the proposed relaxation can be very conservative, because it considers the static error of removing probe $P_i$ relative to the minimum work possible of a DAG, in order to be able to guarantee that the dynamic error will be bounded by a given threshold.
-This conservatism can be overly restrictive in some cases, resulting in a negligible overhead reduction but also causing just a negligible dynamic error to the work profiling.
+In some cases, the proposed relaxation can be very conservative, because it
+considers the static error of removing probe $P_i$ relative to the minimum work
+possible of a DAG, in order to be able to guarantee that the dynamic error will
+be bounded by a given threshold.
+This conservatism can be overly restrictive in some cases, resulting in a
+negligible overhead reduction but also causing just a negligible dynamic error
+to the work profiling.
 

### 2017-09-13 cgo18mscr.tex
updates sections about the relaxation strategies

@@
 
-In some cases, the proposed relaxation can be very conservative, because it
-considers the static error of removing probe $P_i$ relative to the minimum work
-possible of a DAG, in order to be able to guarantee that the dynamic error will
-be bounded by a given threshold.
+In this section we go one step even further.
+In some cases, even the proposed relaxation strategy can still be conservative.
 This conservatism can be overly restrictive in some cases, resulting in a
@@
 to the work profiling.
+For these cases, we propose an adapted version of the relaxation algorithm that
+operates on the whole program.
+Traditionally, compiler optimizations are performed on the function-level, or at
+best on a per module basis.
+Whole program optimization (WPO) means that the compiler considers all
+compilation units of the program and optimizes them using the combined knowledge
+of how they are used together.
 
-In some cases, this overly restrictive conservatism may be unnecessary.
-For these cases, we propose an adapted version of the relaxation algorithm that operates on the whole program.
-Traditionally, compilers optimizations are performed on the function-level, or at best on a per module basis.
-Whole program optimization (WPO) means that the compiler considers all compilation units of the program and optimizes them using the combined knowledge of how they are used together.
+The \textit{whole program relaxation} works by using block-frequency profiling
+from previous executions.
+By having this profiling information, the whole program relaxation is able to
+compute the error of removing a given probe in terms of the whole program's execution,
+and then select a subset of all the probes to be removed.
 
-The whole program relaxation works by using block-frequency profiling from previous executions.
-By having this profiling information, the \textit{whole program relaxation} is able to compute the error of removing a given probe in terms of the whole program's execution,
-and then use this error values for selecting a subset of all the probes to be removed.
-
-For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we model the whole program the relaxation as the following 0-1 Knapsack problem:
+For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we model the
+whole program relaxation as the following 0-1 Knapsack problem:
 \begin{gather*}
@@
 \end{gather*}
-where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$, $x_i$ denotes the probes selected for removal, $M$ is the error threshold, and $\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to the profiled global work, i.e.,
-if $\Delta W$ is the work value for the whole program's execution, computed from the basic block frequencies profiled from previous executions, the error for a given probe $P_i$ is
+where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$,
+$x_i$ denotes the probes selected for removal, $M$ is the error threshold, and
+$\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to
+the profiled global work, i.e.,
+if $\Delta W$ is the work value for the whole program's execution, computed from
+the basic block frequencies profiled from previous executions, the error for a
+given probe $P_i$ is
 \[
@@
 
-Contrary to the per DAG relaxation, the whole program relaxation is not guaranteed to be bounded by the error threshold $M$,
+Contrary to the per DAG relaxation, the whole program relaxation is not
+guaranteed to be bounded by the error threshold $M$,
 as it depends on the representativity % representativeness

### 2017-09-13 cgo18mscr.tex
minor change

@@
 
-\noindent\textbf{Populating the edge flows:}
+\paragraph{Populating the edge flows:}
 Intuitively, if all the edge flows are known for the complement of a spanning tree
@@
 
-\noindent\textbf{Composing the aggregated value of the probes:}
+\paragraph{Composing the aggregated value of the probes:}
 If $x$ is a symbolic expression, then $\kappa_P(x)$ represents the coefficient

### 2017-09-13 motivation.tex
updates figure with WPO-R2

@@
     \includegraphics[width=0.5\textwidth]{figs/motivation-online.pdf}
-    \caption{Histograms of speedup over \texttt{-O3} for each of the 1,000 inputs of XXX. The green histogram shows the speedup when using
+    \caption{Histograms of speedup over \texttt{-O3} for each of the 1,000 inputs of \texttt{susan\_c}. The green histogram shows the speedup when using
     the optimization sequence selected by offline \itercomp, the blue one is the speedup using the optimization sequence selected by an
@@
 
-To show this, we performed offline \itercomp on XXX using a small subset of its inputs and we identified the optimization sequence that
+To show this, we performed offline \itercomp on \texttt{susan\_c} using a small subset of its inputs and we identified the optimization sequence that
 maximizes performance. In Figure~\ref{fig:motivation-online}, the green bars are the histogram of speedups compared to \texttt{-O3} for
@@
     \includegraphics[width=0.5\textwidth]{figs/motivation-metric.pdf}
-    \caption{Relationship between IPC and execution time vs optimization speedup for XXX different binaries of XXX. Each point represents
-	averages over all inputs.}
+    \caption{Relationship between IPC and execution time vs optimization speedup for 500 different binaries of \texttt{susan\_c}. Each point represents
+	averages over a distinct subsets of its inputs.}
     \label{fig:motivation-metric}
@@
 Neither metric is a good proxy for speedup. Figure~\ref{fig:motivation-metric} plots the measured IPC and execution time versus the real
-speedup for XXX different binaries of the XXX benchmark. IPC, execution time, and speedup are averaged over all inputs. Both metrics
-correlate little with speedup. Different inputs require a different amount of work which dominates the changes in execution time. IPC on
-the other hand might improve with a more efficient binary but the opposite can also be true. For example, inserting \texttt{NOP} statements
-will usually improve IPC.
+speedup for 500 different binaries of the \texttt{susan\_c} benchmark. IPC, execution time, and speedup are averaged over distinct subsets of its inputs.
+Both metrics have no direct correlation with speedup. Different inputs require a different amount of work which dominates the changes in execution time. IPC on
+the other hand might improve with a more efficient binary but the opposite can also be true. For example,
+replacing many one-cycle instructions by fewer multi-cycle instructions.
+%inserting \texttt{NOP} statements will usually improve IPC.
 

### 2017-09-14 conclusions.tex
first attack on conclusions

@@
-\section{Conclusion and Future Work}
+\section{Conclusions}
+This paper has presented a novel work efficiency metric to enable one to compare the performance of any two different compiled versions of
+a program without executing a given input more than once. This ability allows us to implement, for the first time, a true online iterative
+compilation  to search for the best compilation options  based on actual user inputs seen in the deployment environment.
 
-Our proposed technique for relaxed instrumentation can also be applied for similar profiling scenarios, such as profiling empirical computational complexity~\cite{goldsmith07,zaparanuks12,coppa14}.
+We develop a compiler-based, low-overhead instrumentation mechanism to gather the require data to evaluate the work metric. Our mechanism
+does this without any human involvement. We show that by carefully selecting where to place the probes in the source code, the
+instrumentation overhead can be negligible. We give two probe optimization strategies, allowing developers to balance the instrumentation
+overhead and the performance error.
+
+We  evaluate our approach using the KDataSets benchmark suit. Experimental results show that our approach delivers \red{80\%} of the
+speedup achieved by an offline  oracle, but without executing any input more than once. We show that our probe optimization strategy is
+highly effective, which brings the instrumentation overhead down to only \red{4\%} on average, making our approach practical for regular
+use.
+
+%Our proposed technique for relaxed instrumentation can also be applied for similar profiling scenarios, such as profiling empirical
+%computational complexity~\cite{goldsmith07,zaparanuks12,coppa14}.

### 2017-09-14 motivation.tex
first attack on conclusions

@@
 
-Offline \itercomp can be very sensitive on having representative inputs for the program under optimization. If the developers cannot
-predict what will be the typical usage of the application or that typical usage changes over time, then \itercomp might fail making the
-application faster for the user.
+The success of \itercomp relies on having representative inputs for the target program to be optimized. If the developers cannot predict
+what will be the typical usage of the application or that typical usage changes over time, then \itercomp might fail making the application
+faster for the user.
 
@@
 
-To show this, we performed offline \itercomp on \texttt{susan\_c} using a small subset of its inputs and we identified the optimization sequence that
-maximizes performance. In Figure~\ref{fig:motivation-online}, the green bars are the histogram of speedups compared to \texttt{-O3} for
-every individual input, while the dark green line shows the speedup averaged over all inputs. Despite the selected optimization sequence
-being the best for some inputs, we see that on average it improves performance by only 6\% and hurts performance for almost half the
-inputs. Just randomly choosing an input set for offline \itercomp can be counterproductive.
+To show this, we performed offline \itercomp on \texttt{susan\_c} from the \textit{KDataSets} benchmark suit~\cite{chen10,chen12a} using a
+small subset of its inputs to identify the best-performing optimization sequence. In Figure~\ref{fig:motivation-online}, the green bars are
+the histogram of speedups compared to \red{LLVM?} \texttt{-O3} for each individual input, while the dark green line shows the speedup
+averaged over all inputs. Despite the selected optimization sequence being the best for some inputs, we see that it delivers, on average,
+6\% of performance improvement, but gives slows down performance for nearly of the inputs. This example shows that randomly choosing an
+input set to use for offline \itercomp can be counterproductive.
 
-The blue bars in Figure~\ref{fig:motivation-online} show a different approach where during \itercomp the speedup of each optimization
-sequence is again measured on a small input subset but this subset changes for every optimization sequence. This is similar to how online
-\itercomp would have to work, using each input only once. This might introduce errors, since the speedups we measure for each optimization
-sequence are not strictly comparable, they are measured on different workloads. But in the end the histogram shows that being able to test
-more inputs allows us to make better informed decisions. The optimization sequence we selected never degrades performance compared to
-\texttt{-O3} and is on average 13\% better than the offline approach.
+The blue bars \red{zw: the reviewer won't be able to see the difference if the paper is printed in black and white.} in
+Figure~\ref{fig:motivation-online} show a different approach where during \itercomp the speedup of each optimization sequence is again
+measured on a small input subset but this subset changes for every optimization sequence. This is similar to how online \itercomp would
+have to work, using each input only once. This might introduce errors, since the speedups we measure for each optimization sequence are not
+strictly comparable, they are measured on different workloads. But in the end the histogram shows that being able to test more inputs
+allows us to make better informed decisions. The optimization sequence we selected never degrades performance compared to \texttt{-O3} and
+is on average 13\% better than the offline approach.
 
@@
     \includegraphics[width=0.5\textwidth]{figs/motivation-metric.pdf}
-    \caption{Relationship between IPC and execution time vs optimization speedup for 500 different binaries of \texttt{susan\_c}. Each point represents
+    \caption{Relationship between IPC and execution time vs optimization speedup for 500 different binaries of \texttt{susan\_c}.
+The binaries are compiled using different optimization sequences.
+Each point represents
 	averages over a distinct subsets of its inputs.}
+    \red{ZW: Perhaps take out the blue line (execution time). It is rather confusing. }
     \label{fig:motivation-metric}
@@
 
-In the remaining of this paper, we examine our work efficiency metric, a novel way for approximating optimization speedup through a single
-run which can be used to drive an online \itercomp methodology.
+In the remaining of this paper, we describe our work efficiency metric, which provides a new way for approximating optimization speedup
+through a single run. We show that by utilizing this metric, a highly effective online \itercomp methodology can be built.
 

### 2017-09-14 motivation.tex
minor fixes in motivation

@@
 
-To show this, we performed offline \itercomp on \texttt{susan\_c} from the \textit{KDataSets} benchmark suit~\cite{chen10,chen12a} using a
+To show this, we performed offline \itercomp on \texttt{susan\_c} from the \textit{KDataSets} benchmark suite~\cite{chen10,chen12a} using a
 small subset of its inputs to identify the best-performing optimization sequence. In Figure~\ref{fig:motivation-online}, the green bars are
-the histogram of speedups compared to \red{LLVM?} \texttt{-O3} for each individual input, while the dark green line shows the speedup
+the histogram of speedups over LLVM \texttt{-O3} for each individual input, when using this ``best'' optimization sequence. The dark green line shows the same speedup
 averaged over all inputs. Despite the selected optimization sequence being the best for some inputs, we see that it delivers, on average,
-6\% of performance improvement, but gives slows down performance for nearly of the inputs. This example shows that randomly choosing an
+only 6\% performance improvement and slows down performance for nearly half of the inputs. This example shows that randomly choosing an
 input set to use for offline \itercomp can be counterproductive.
@@
     \includegraphics[width=0.5\textwidth]{figs/motivation-metric.pdf}
-    \caption{Relationship between IPC and execution time vs optimization speedup for 500 different binaries of \texttt{susan\_c}.
-The binaries are compiled using different optimization sequences.
-Each point represents
-	averages over a distinct subsets of its inputs.}
-    \red{ZW: Perhaps take out the blue line (execution time). It is rather confusing. }
+    \caption{Relationship between IPC and execution time vs optimization speedup for 500 different binaries of \texttt{susan\_c}. The
+    binaries are compiled using different optimization sequences. Each point represents averages over a distinct subset of its inputs.}
+    \red{ZW: Perhaps take out the blue line (execution time). It is rather confusing.}
     \label{fig:motivation-metric}
@@
 Neither metric is a good proxy for speedup. Figure~\ref{fig:motivation-metric} plots the measured IPC and execution time versus the real
-speedup for 500 different binaries of the \texttt{susan\_c} benchmark. IPC, execution time, and speedup are averaged over distinct subsets of its inputs.
+speedup for 500 different binaries of the \texttt{susan\_c} benchmark. IPC, execution time, and speedup are averages over distinct subsets of its inputs.
 Both metrics have no direct correlation with speedup. Different inputs require a different amount of work which dominates the changes in execution time. IPC on
 the other hand might improve with a more efficient binary but the opposite can also be true. For example,
-replacing many one-cycle instructions by fewer multi-cycle instructions.
-%inserting \texttt{NOP} statements will usually improve IPC.
+% replacing many one-cycle instructions by fewer multi-cycle instructions.
+inserting \texttt{NOP} statements will usually improve IPC but hurt performance.
 

### 2017-09-14 motivation.tex
Remotivation - surely that's a word?

--- 2017-09-13 motivation.tex
@@
 
-Offline \itercomp can be very sensitive on having representative inputs for the program under optimization. If the developers cannot
-predict what will be the typical usage of the application or that typical usage changes over time, then \itercomp might fail making the
-application faster for the user.
+    In this section we first motivate online iterative compilation by showing the deleterious effects of choosing the wrong inputs for the
+    offline case. Then we go on to show the unsuitability of previously proposed program version comparison metrics.
 
@@
         \includegraphics[width=0.5\textwidth]{figs/motivation-online.pdf}
-    \caption{Histograms of speedup over \texttt{-O3} for each of the 1,000 inputs of \texttt{susan\_c}. The green histogram shows the speedup when using
-    the optimization sequence selected by offline \itercomp, the blue one is the speedup using the optimization sequence selected by an
-    online-like approach.}
+        \caption{
+            Histograms of speedup over \texttt{-O3} for each of the 1,000 inputs of \texttt{susan\_c}. The green histogram shows the
+            speedup when using the optimization sequence selected by offline \itercomp, the blue one is the speedup using the optimization
+            sequence selected by an online-like approach. \FIXME{Beware Daltonism!}
+        }
         \label{fig:motivation-online}
@@
 
-To show this, we performed offline \itercomp on \texttt{susan\_c} using a small subset of its inputs and we identified the optimization sequence that
-maximizes performance. In Figure~\ref{fig:motivation-online}, the green bars are the histogram of speedups compared to \texttt{-O3} for
-every individual input, while the dark green line shows the speedup averaged over all inputs. Despite the selected optimization sequence
-being the best for some inputs, we see that on average it improves performance by only 6\% and hurts performance for almost half the
-inputs. Just randomly choosing an input set for offline \itercomp can be counterproductive.
+    We performed offline iterative compilation on \texttt{susan\_c} to find the optimization settings that maximize the average performance
+    over \FIXME{five} randomly chosen inputs from a set of \FIXME{1,000} provided by \textit{KDataSets} benchmark
+    suite~\cite{chen10,chen12a}. The green histogram in Figure~\ref{fig:motivation-online} shows the speedups for all \FIXME{1,000} inputs
+    of that chosen optimization setting versus \texttt{-O3}. The dark green line shows the average speedup for all those inputs. The
+    \FIXME{five} training inputs are all in the top \FIXME{twelve} performing inputs. On the other hand, the chosen optimization does
+    poorly on other inputs, often causing a slowdown and yielding only a slight speedup of 6\% on average.
 
-The blue bars in Figure~\ref{fig:motivation-online} show a different approach where during \itercomp the speedup of each optimization
-sequence is again measured on a small input subset but this subset changes for every optimization sequence. This is similar to how online
-\itercomp would have to work, using each input only once. This might introduce errors, since the speedups we measure for each optimization
-sequence are not strictly comparable, they are measured on different workloads. But in the end the histogram shows that being able to test
-more inputs allows us to make better informed decisions. The optimization sequence we selected never degrades performance compared to
-\texttt{-O3} and is on average 13\% better than the offline approach.
+    The blue bars in Figure~\ref{fig:motivation-online} show a different approach. Here the iterative compilation proceeds by evaluating
+    each optimization on \FIXME{five} inputs. Each input is evaluated only once during the search, but the search is provided with oracle
+    knowledge about the unoptimized runtime for the inputs, giving a perfect work metric. The environment of the search constantly changes
+    in the same way as for online iterative compilation. The best optimization for this case never degrades performance compared to
+    \texttt{-O3} and is on average 13\% better than the offline approach. The online case, considering truly representative inputs, leads
+    to superior performance, if a suitable work metric can be found that comes close to the oracle.
 
@@
         \includegraphics[width=0.5\textwidth]{figs/motivation-metric.pdf}
-    \caption{Relationship between IPC and execution time vs optimization speedup for 500 different binaries of \texttt{susan\_c}. Each point represents
-	averages over a distinct subsets of its inputs.}
+        \caption{
+            Relationship between IPC and execution time vs optimization speedup for 500 different binaries of \texttt{susan\_c}.
+            Each point represents averages over all inputs.
+        }
         \label{fig:motivation-metric}
@@
 
-This is not an entirely surprising result. Evaluating our optimization decisions on more inputs decreases the chances of optimizing for an
-unrepresentative input. Despite that offline \itercomp is still the preferred way. Measuring the speedup caused by an optimization requires
-running the same input \textit{at least twice}, one of the optimized and one for the baseline binary. This is either infeasible or
-undesirable in a realistic online \itercomp scenario. There are two existing workarounds to this problem: directly comparing the
-\textit{execution time} of each run against previous runs or directly comparing the \textit{instructions per cycle}~(IPC)~\citep{fursin07}.
-Neither metric is a good proxy for speedup. Figure~\ref{fig:motivation-metric} plots the measured IPC and execution time versus the real
-speedup for 500 different binaries of the \texttt{susan\_c} benchmark. IPC, execution time, and speedup are averaged over distinct subsets of its inputs.
-Both metrics have no direct correlation with speedup. Different inputs require a different amount of work which dominates the changes in execution time. IPC on
-the other hand might improve with a more efficient binary but the opposite can also be true. For example,
-replacing many one-cycle instructions by fewer multi-cycle instructions.
-%inserting \texttt{NOP} statements will usually improve IPC.
+    Prior to this paper there were two proposed metrics to correlate with program efficiency. The first is simply to use runtime, ignoring
+    the work aspect. This functions well when the inputs all require similar effort. When that is not the case it bears little or no
+    correlation to efficiency. The second approach is \textit{instructions per cycle}~(IPC), suggested by~\citep{fursin07}. While IPC seems
+    promising in the sense that faster instruction execution is a good thing, this is belied by the simple observation that adding
+    redundant NOPs to a program will increase IPC while also slowing the program down. Figure~\ref{fig:motivation-metric} shows the
+    relationship between execution time, IPC and speedup for different optimizations of \texttt{susan\_c}, averaged across all inputs.
+    Neither IPC nor execution time correlate well, and so neither can be used for iterative compilation. We need a work efficiency metric
+    which can approximate speedup instead, using only a single execution. \FIXME{HJL: exec time avg over all inputs, why not good?}
 
-In the remaining of this paper, we examine our work efficiency metric, a novel way for approximating optimization speedup through a single
-run which can be used to drive an online \itercomp methodology.
+    In the remainder of this paper we present our work efficiency metric and demonstrate its utility for iterative compilation.
 
-% %As a motivation example, consider...
+%<<<<<<< Updated upstream
+%The success of \itercomp relies on having representative inputs for the target program to be optimized. If the developers cannot predict
+%what will be the typical usage of the application or that typical usage changes over time, then \itercomp might fail making the application
+%faster for the user.
 %
-% %Here we give an example to motivate our work. E.g. assuming we have a zero-cost profiling based work metric,
-% %how can we use it to perform iterative compilation across runs? Then we talk about while it is impossible to
-% %build a zero-overhead metric, we can develop one with low overhead.
-%
-% This section provides a comparison between the offline and online approaches for
-% {\itercomp}.
-% The goal is to motivate the need for an online {\itercomp}, highlighting
-% weaknesses of using offline {\itercomp}.
-%
-% %In the offline approach, a fixed set of inputs is selected for which the best
-% %optimization sequence is selected during the development time (pre-shipping) of
-% %a program.
-% %Although this fixed set of inputs is expected to be representative of the user's
-% %workload,
-% %in most cases, it might be infeasible to identify and collect representative
-% %workloads for every group of users.
-% %Furthermore, a user can dynamically change its workload pattern, rendering
-% %a fixed optimization sequence innapropriate for his performance requirements.
-%
-% %In the online approach, the program is shipped with an initial optimization
-% %sequence and different optimization sequences are evaluated as the end-user
-% %executes the program.
-% %This approach frees developers from the need to collect representative
-% %workloads for every group of users so that the program can be optimized
-% %accordingly.
-% %More importantly, even in scenarios where the user's workload can dynamically
-% %change its pattern, the online {\itercomp} would still be able to optimize the
-% %program accordingly.
-%
-% In the offline approach, a small fixed set of inputs is selected for which all
-% optimization sequences are evaluated.
-% In the online approach, each optimization sequence is evaluated based on a,
-% dynamically selected, distinct set of inputs.
-% In both cases, the optimization sequences are ranked based on their average
-% speedup, and the best optimization sequence is selected.
-%
-% In order to assess the quality of the optimization sequence selected by each
-% approach, we evaluate both optimization sequences against all available inputs
-% in our dataset.
-% Figure~\ref{fig:motivation-online} shows this comparison between both the offline
-% and online approaches for {\itercomp}.
-% It shows the performance gains of both optimization sequences compared with the
-% standard compiler's optimization (i.e.,\texttt{-O3}).
-% The online {\itercomp} achieves an average speedup which is about 13\% better
-% than the solution selected by the offline approach.
-
-% \begin{figure}[t!]
+%\begin{figure}[t!]
 %    \centering
 %    \includegraphics[width=0.5\textwidth]{figs/motivation-online.pdf}
-%     \caption{Comparison between the offline and online {\itercomp}.
-%              The vertical lines represent the average speedups over
-%              all the 1,000 inputs.}
+%    \caption{Histograms of speedup over \texttt{-O3} for each of the 1,000 inputs of \texttt{susan\_c}. The green histogram shows the speedup when using
+%    the optimization sequence selected by offline \itercomp, the blue one is the speedup using the optimization sequence selected by an
+%    online-like approach.}
 %    \label{fig:motivation-online}
-% \end{figure}
-
-% It is worth mentioning that our goal is to improve the average performance
-% across the user's workload, and not only for a specific input.
-% This means that the online {\itercomp} is better informed throughout the search,
-% as it compares optimization sequences across different sets inputs.
-
-% However, because of the restriction of having a single execution per input,
-% it is not possible to measure speedup in a real-world online scenario.
-% Moreover, measuring just execution time, for example, is also not viable since
-% different inputs often mean different amounts of work,
-% rendering it meaningless to compare optimizations only by execution time.
-% Therefore, we need a performance metric which can be profiled from a single
-% execution of the program.
-% Although a previous work has suggested that \textit{instructions per cycle}~(IPC)
-% could be used with that goal, IPC have no correlation with speedup~\citep{fursin07}.
-% In a case where we have two distinct versions of the program, due to different
-% optimizations, which is also executed with distinct inputs, higher IPC does not
-% necessarily mean lower execution time (higher performance).
+%\end{figure}
 %
-% \begin{figure}[t]
+%To show this, we performed offline \itercomp on \texttt{susan\_c} from the \textit{KDataSets} benchmark suit~\cite{chen10,chen12a} using a
+%small subset of its inputs to identify the best-performing optimization sequence. In Figure~\ref{fig:motivation-online}, the green bars are
+%the histogram of speedups compared to \red{LLVM?} \texttt{-O3} for each individual input, while the dark green line shows the speedup
+%averaged over all inputs. Despite the selected optimization sequence being the best for some inputs, we see that it delivers, on average,
+%6\% of performance improvement, but gives slows down performance for nearly of the inputs. This example shows that randomly choosing an
+%input set to use for offline \itercomp can be counterproductive.
+%
+%The blue bars \red{zw: the reviewer won't be able to see the difference if the paper is printed in black and white.} in
+%Figure~\ref{fig:motivation-online} show a different approach where during \itercomp the speedup of each optimization sequence is again
+%measured on a small input subset but this subset changes for every optimization sequence. This is similar to how online \itercomp would
+%have to work, using each input only once. This might introduce errors, since the speedups we measure for each optimization sequence are not
+%strictly comparable, they are measured on different workloads. But in the end the histogram shows that being able to test more inputs
+%allows us to make better informed decisions. The optimization sequence we selected never degrades performance compared to \texttt{-O3} and
+%is on average 13\% better than the offline approach.
+%
+%\begin{figure}[t]
 %    \centering
 %    \includegraphics[width=0.5\textwidth]{figs/motivation-metric.pdf}
-%     \caption{Comparison among different performance metrics for performing
-%              online {\itercomp}.
-%              The vertical lines represent the best score for each metric.}
+%    \caption{Relationship between IPC and execution time vs optimization speedup for 500 different binaries of \texttt{susan\_c}.
+The binaries are compiled using different optimization sequences.
+Each point represents
+%	averages over a distinct subsets of its inputs.}
+%    \red{ZW: Perhaps take out the blue line (execution time). It is rather confusing. }
 %    \label{fig:motivation-metric}
-% \end{figure}
+%\end{figure}
 %
-% Figure~\ref{fig:motivation-metric} shows a comparison between execution time and IPC,
-% and their correlation with the actual speedup.
-% Each point in the figure corresponds to each metric averaged over
-% a set of inputs, collected during the {\itercomp}.
-% The vertical lines correspond to the optimization sequence
-% with the best score for each one of the metrics.
-% This figure shows that both execution time and IPC are unsuitable for
-% online {\itercomp}, as they present no direct correlation with the actual
-% speedup obtained by the optimization sequence being evaluated.
+%This is not an entirely surprising result. Evaluating our optimization decisions on more inputs decreases the chances of optimizing for an
+%unrepresentative input. Despite that offline \itercomp is still the preferred way. Measuring the speedup caused by an optimization requires
+%running the same input \textit{at least twice}, one of the optimized and one for the baseline binary. This is either infeasible or
+%undesirable in a realistic online \itercomp scenario. There are two existing workarounds to this problem: directly comparing the
+%\textit{execution time} of each run against previous runs or directly comparing the \textit{instructions per cycle}~(IPC)~\citep{fursin07}.
+%Neither metric is a good proxy for speedup. Figure~\ref{fig:motivation-metric} plots the measured IPC and execution time versus the real
+%speedup for 500 different binaries of the \texttt{susan\_c} benchmark. IPC, execution time, and speedup are averaged over distinct subsets of its inputs.
+%Both metrics have no direct correlation with speedup. Different inputs require a different amount of work which dominates the changes in execution time. IPC on
+%the other hand might improve with a more efficient binary but the opposite can also be true. For example,
+%replacing many one-cycle instructions by fewer multi-cycle instructions.
+%%inserting \texttt{NOP} statements will usually improve IPC.
 %
-% In the remaining of this paper, we develop a work efficiency metric, suitable
-% for guiding online {\itercomp}, which can be profiled with a single execution
-% of the program with the optimization sequence being evaluated.
+%In the remaining of this paper, we describe our work efficiency metric, which provides a new way for approximating optimization speedup
+%through a single run. We show that by utilizing this metric, a highly effective online \itercomp methodology can be built.

### 2017-09-14 motivation.tex
changes in Hugh's motivation

@@
     We performed offline iterative compilation on \texttt{susan\_c} to find the optimization settings that maximize the average performance
-    over \FIXME{five} randomly chosen inputs from a set of \FIXME{1,000} provided by \textit{KDataSets} benchmark
-    suite~\cite{chen10,chen12a}. The green histogram in Figure~\ref{fig:motivation-online} shows the speedups for all \FIXME{1,000} inputs
-    of that chosen optimization setting versus \texttt{-O3}. The dark green line shows the average speedup for all those inputs. The
+    over \FIXME{five} randomly chosen inputs from a set of 1,000 provided by the \textit{KDataSets} benchmark
+	suite~\cite{chen10,chen12a}. The green histogram in Figure~\ref{fig:motivation-online} shows for each of the 1,0000 inputs the speedup versus \texttt{-O3}
+    of that chosen optimization setting. The dark green line shows the average speedup for all those inputs. The
     \FIXME{five} training inputs are all in the top \FIXME{twelve} performing inputs. On the other hand, the chosen optimization does
@@
     each optimization on \FIXME{five} inputs. Each input is evaluated only once during the search, but the search is provided with oracle
-    knowledge about the unoptimized runtime for the inputs, giving a perfect work metric. The environment of the search constantly changes
+    knowledge about the unoptimized runtime for the inputs, giving a perfect speedup metric. The environment of the search constantly changes
     in the same way as for online iterative compilation. The best optimization for this case never degrades performance compared to
     \texttt{-O3} and is on average 13\% better than the offline approach. The online case, considering truly representative inputs, leads
-    to superior performance, if a suitable work metric can be found that comes close to the oracle.
+	to superior performance, if the speedup caused by optimizing can be estimated.
 
@@
             Relationship between IPC and execution time vs optimization speedup for 500 different binaries of \texttt{susan\_c}.
-            Each point represents averages over all inputs.
+            Each point represents averages over a subset of all inputs.
         }
@@
     redundant NOPs to a program will increase IPC while also slowing the program down. Figure~\ref{fig:motivation-metric} shows the
-    relationship between execution time, IPC and speedup for different optimizations of \texttt{susan\_c}, averaged across all inputs.
+    relationship between execution time, IPC and speedup for different optimizations of \texttt{susan\_c}, averaged across a subset of the inputs.
     Neither IPC nor execution time correlate well, and so neither can be used for iterative compilation. We need a work efficiency metric
-    which can approximate speedup instead, using only a single execution. \FIXME{HJL: exec time avg over all inputs, why not good?}
+    which can approximate speedup instead, using only a single execution.
 
     In the remainder of this paper we present our work efficiency metric and demonstrate its utility for iterative compilation.
-
-%<<<<<<< Updated upstream
-%The success of \itercomp relies on having representative inputs for the target program to be optimized. If the developers cannot predict
-%what will be the typical usage of the application or that typical usage changes over time, then \itercomp might fail making the application
-%faster for the user.
-%
-%\begin{figure}[t!]
-%    \centering
-%    \includegraphics[width=0.5\textwidth]{figs/motivation-online.pdf}
-%    \caption{Histograms of speedup over \texttt{-O3} for each of the 1,000 inputs of \texttt{susan\_c}. The green histogram shows the speedup when using
-%    the optimization sequence selected by offline \itercomp, the blue one is the speedup using the optimization sequence selected by an
-%    online-like approach.}
-%    \label{fig:motivation-online}
-%\end{figure}
-%
-%To show this, we performed offline \itercomp on \texttt{susan\_c} from the \textit{KDataSets} benchmark suit~\cite{chen10,chen12a} using a
-%small subset of its inputs to identify the best-performing optimization sequence. In Figure~\ref{fig:motivation-online}, the green bars are
-%the histogram of speedups compared to \red{LLVM?} \texttt{-O3} for each individual input, while the dark green line shows the speedup
-%averaged over all inputs. Despite the selected optimization sequence being the best for some inputs, we see that it delivers, on average,
-%6\% of performance improvement, but gives slows down performance for nearly of the inputs. This example shows that randomly choosing an
-%input set to use for offline \itercomp can be counterproductive.
-%
-%The blue bars \red{zw: the reviewer won't be able to see the difference if the paper is printed in black and white.} in
-%Figure~\ref{fig:motivation-online} show a different approach where during \itercomp the speedup of each optimization sequence is again
-%measured on a small input subset but this subset changes for every optimization sequence. This is similar to how online \itercomp would
-%have to work, using each input only once. This might introduce errors, since the speedups we measure for each optimization sequence are not
-%strictly comparable, they are measured on different workloads. But in the end the histogram shows that being able to test more inputs
-%allows us to make better informed decisions. The optimization sequence we selected never degrades performance compared to \texttt{-O3} and
-%is on average 13\% better than the offline approach.
-%
-%\begin{figure}[t]
-%    \centering
-%    \includegraphics[width=0.5\textwidth]{figs/motivation-metric.pdf}
-%    \caption{Relationship between IPC and execution time vs optimization speedup for 500 different binaries of \texttt{susan\_c}.
-The binaries are compiled using different optimization sequences.
-Each point represents
-%	averages over a distinct subsets of its inputs.}
-%    \red{ZW: Perhaps take out the blue line (execution time). It is rather confusing. }
-%    \label{fig:motivation-metric}
-%\end{figure}
-%
-%This is not an entirely surprising result. Evaluating our optimization decisions on more inputs decreases the chances of optimizing for an
-%unrepresentative input. Despite that offline \itercomp is still the preferred way. Measuring the speedup caused by an optimization requires
-%running the same input \textit{at least twice}, one of the optimized and one for the baseline binary. This is either infeasible or
-%undesirable in a realistic online \itercomp scenario. There are two existing workarounds to this problem: directly comparing the
-%\textit{execution time} of each run against previous runs or directly comparing the \textit{instructions per cycle}~(IPC)~\citep{fursin07}.
-%Neither metric is a good proxy for speedup. Figure~\ref{fig:motivation-metric} plots the measured IPC and execution time versus the real
-%speedup for 500 different binaries of the \texttt{susan\_c} benchmark. IPC, execution time, and speedup are averaged over distinct subsets of its inputs.
-%Both metrics have no direct correlation with speedup. Different inputs require a different amount of work which dominates the changes in execution time. IPC on
-%the other hand might improve with a more efficient binary but the opposite can also be true. For example,
-%replacing many one-cycle instructions by fewer multi-cycle instructions.
-%%inserting \texttt{NOP} statements will usually improve IPC.
-%
-%In the remaining of this paper, we describe our work efficiency metric, which provides a new way for approximating optimization speedup
-%through a single run. We show that by utilizing this metric, a highly effective online \itercomp methodology can be built.

### 2017-09-14 cgo18mscr.tex
first attack on conclusions
Frac
Splitting

@@
 \input{motivation}
-
-\section{Work Efficiency Metric} \label{sec:metric}
-
-In this section we define work efficiency, a metric proposed for comparing
-different optimized versions of a program when executing with different inputs.
-We define it as the ratio between the amount of \textit{work}, $\Delta W$,
-performed during a period of time, $\Delta t$.
-\[
-   P = \frac{\Delta W}{\Delta t}
-\]
-
-By measuring the amount of \textit{computational work} done per unit of time,
-it reduces the impact of input-dependent aspects and focus instead on the
-efficiency of the optimized program.
-For this metric, the main challenge is to precisely define what represents
-\textit{work}.
-In the context of experimental algorithmics, previous works have suggested using
-\textit{block frequency} as a metric for empirical computational
-complexity~\cite{goldsmith07}.
-%Block frequency is a relative metric that represents the number of times a basic
-%block executes~\citep{ball94,ball96}.
-
-However, in the context of comparing different optimization sequences,
-although block frequency would be able to capture aspects of optimizations that
-simplify the control-flow graph (CFG), measuring work at the basic block
-resolution would not capture effects of optimisations at the instruction level.
-%Because of that, in this thesis, we extend the idea of using basic block
-%frequency to measure computational work by also considering the computational
-%cost of each basic block.
-Because of that, we adopt a work metric that takes into account the
-computational cost within each basic block.
-The computational cost of a basic block is given by weighting the instructions
-that it contains.
-
-To this end, we model the computational work $\Delta W$ as a linear equation based on block frequency information and a cost model of the instruction set.
-Formally,
-\[
-\Delta W = \varepsilon + \sum_{B} w(B)f(B)
-\]
-where $f(B)$ represents the frequency of basic block $B$ and $w(B)$ represents the computational work of executing $B$.
-We define the work of a basic block $B$ as the sum of the cost of its instructions, i.e.,
-\[
-w(B) = \sum_{i} w_i N_B(i)
-\]
-where $w_i$ is the cost of instruction $i$ and $N_B(i)$ is the number of occurrences of instruction $i$ in basic block $B$.
-
-In this simplified model, we consider that $w_i$ is constant across all programs and executions, varying only between target architectures.
-On the other hand, $N_B(i)$ is a program-dependent static value which is known at compile-time and $f(B)$ is a dynamic value known only at run-time,
-since $f(B)$ is both program and execution dependent as the execution frequency of a basic block can change when executing with different inputs.
-
-\subsection{Estimating a Cost Model of the Instructions}
-
-Similarly to previous work~\citep{giusto01,powell09,brandolese11}, we derive the cost model of the instruction set by modelling the problem as a multi-variable linear regression, where the \textit{regression coefficients} are the costs of the instructions and the \textit{regressors} are computed as $\sum_B N_B(i)f(B)$ for each instruction.
-\begin{equation}\label{eq:linear-work-expression}
-\Delta W = \varepsilon + \sum_{i} \left(w_i \sum_{B} N_B(i)f(B)\right)
-\end{equation}
-By having some empirical data after executing several benchmarks with different inputs, we can fit the linear model with this empirical data in order to obtain estimate costs of the instructions.
-In order to fit the linear model, we measure the wall-clock time when executing the training benchmarks described in Section\ref{sec:benchmarks} with their respective 1000 input datasets.
-For these measurements, the training benchmarks are compiled without optimization.
-The reason for using no optimization, as discussed in Section\ref{sec:oic-infra}, is because the amount of work must be only input dependent and consistent between different optimization sequences.
-This aspect is crucial for the work-based performance metric to enable a consistent comparison between the various optimization sequences when executing with distinct inputs.
-%it allows to use the estimated cost model for computing a work metric which is independent of optimizations, as we explained in Section~\ref{}.
-
-Our cost model has a total of 52 LLVM instructions\footnote{We do not model all LLVM instructions because some instructions are more common in optimized programs, such as the vector instructions.}.
-Every program, in the set of training benchmarks, was compiled twice: once including the necessary instrumentation for the block frequency profiling, which is required for deriving the linear expression defined in Equation~\ref{eq:linear-work-expression};
-and a standard compilation without any optimization or instrumentation.
-For each input in the training dataset, the benchmarks were executed once with the instrumented version for collecting the block frequency profiling, and multiple times with the standard compilation just for measuring the wall-clock execution time, until the confidence interval was no larger than 1\% for a 99\% confidence.
-After collecting these measurements, this data can be used to estimate the unknown parameters in the linear model.
-Because we fit the linear model based on the wall-clock execution time, the derived cost model can be interpreted as an estimate of the execution time when the program is compiled without optimizations.
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.6\linewidth]{figs/cost-model.pdf}
-    \caption{Linear model fitted from empirical data. The mean absolute error (MAE) for the fitted curve is seven milliseconds.}
-    \label{fig:cost-model}
-\end{figure}
-
-Figure~\ref{fig:cost-model} compare the work metric with the corresponding execution time for some instances of the test benchmarks.
-Notice how the fitted model has a higher relative error for the instances with very short execution time, namely those that run for less than one tenth of a second.
-The mean absolute error (MAE) for the fitted curve is seven milliseconds.
-
-\subsection{Comparison with Instructions Per Cycle} \label{sec:ipc-vs-work-metric}
-
-The IPC metric has been widely used for studying performance benefits of hardware optimizations.
-Although previous work has suggested the use of IPC for guiding {\itercomp}, in this section we argue in favour of the WP metric over IPC.
-
-The IPC metric differs from the proposed WP metric in a key aspect:
-the IPC metric is computed solely based on the final optimized program.
-When executing different optimized versions of a program with the same input, both the number of instructions and the number of clock cycles can change.
-For this reason, higher IPC does not necessarily translate to shorter execution time (or even fewer clock cycles).
-
-% We can illustrate this fact with a very small example as shown in Table~\ref{tab:ipc-example}.
-% This example shows two versions of a program, namely P1 and P2, and their respective measurements related to IPC.
-% Although version P1 has twice the IPC of P2, P1 is one cycle slower than P2.
-% As this example illustrates, the IPC metric can be misleading when compared different versions of the same program.
-% This problem is only intensified when both the optimization sequence and the input changes.
-% In Chapter~\ref{chap:eval} we show empirical evidence for this argument.
-%
-% \begin{table}[h]
-% \centering
-% \begin{tabular}{|c|c|c|c|}
-% \hline
-%                        & P1 & P2  \\
-% \hline
-% Number of Instructions & 5  & 2   \\
-% Number of Cycles       & 5  & 4   \\
-% IPC                    & 1  & 0.5 \\
-% \hline
-% \end{tabular}
-% \caption{Example that illustrates that a higher IPC does not necessarily translate to shorter execution time.}
-% \label{tab:ipc-example}
-% \end{table}
-
-On the other hand, WP computes the amount of work based on the unoptimized version of the program, which means that it always measures the same amount of work for the same input, regardless of the optimizations applied on the program.
-This is a key aspect that enables the use of the WP metric for guiding {\itercomp}, because a higher WP, which represents work per unit time, naturally translate to shorter execution time.
-Moreover, previous work has also presented other arguments against the use of IPC in similar use-case scenarios. %, as discussed in Chapter~\ref{chap:related}.
-
-\section{Online {\IterComp} Infrastructure} \label{sec:oic-infra}
-
-Although {\itercomp} had been originally proposed as an \textit{offline} optimization strategy, it can also be adapted to work in online scenarios.
-Instead of selecting the best optimization sequence as part of the development time (pre-shipping) of a program, a first version of the program is shipped together with an {\itercomp} mechanism.
-In the online scenario, the program is shipped with an initial optimization sequence and different optimization sequences are evaluated as the end-user executes the program.
-This optimization strategy is also known as idle-time optimization, as the re-compilation happens between runs of the program.
-
-LLVM is particularly suitable for iterative compilation as it makes possible to cache a pre-compiled, but still unoptimized, version of the input program in the bitcode format of the LLVM IR.
-This caching allows to speedup the time required for re-compilation as it is able to bypass the frontend phase.
-If re-compilation time is critical, it would also be possible to keep only the hot portion of the code in the LLVM bitcode format, while the remaining portion of the code is already compiled to the final object code.
-However, this is out of the scope of this paper and we always re-compile the whole program.
-
-\begin{figure}[htb]
-    \centering
-    \includegraphics[width=\linewidth]{figs/infra-diagram}
-    \caption{Overview of the execution engine for applying {\itercomp}.}
-    \label{fig:infra-diagram}
-\end{figure}
-%\textbf{Describe the process step by step.}
-
-Figure~\ref{fig:infra-diagram} shows an overview of the infrastructure required for applying online {\itercomp}
-using WP as the metric of choice for evaluating optimization sequences.
-The online {\itercomp} follows as described bellow:
-\begin{enumerate}
-\item The program is pre-compiled to the LLVM bitcode format without optimization.
-\item The unoptimized program is instrumented for work profiling.
-\item Execution-based optimization search:
- \begin{enumerate}
-   \item The current optimization sequence is used for the re-compilation of the program.
-   \item The program is executed with any input provided by the end-user.
-         During the execution of the program, wall-clock time and the work metric are recorded by profiling instrumentation.
-   \item If the recorded profiling for the current optimization can be used to compute an average performance measurement within a small confidence interval,
-         then a new optimization sequence is generated.
-         Otherwise, the same optimization sequence is used for the next execution.
- \end{enumerate}
-\end{enumerate}
-
-In this work, we focus mainly on the highlighted components.
-The \textit{Work Instrumentation} phase focuses on providing a low-overhead instrumentation for profiling the work metric.
-The instrumentation consists of adding a global counter to the program which is used to accumulate the amount of work computed during the program's execution, using the cost model of the instruction set.
-A detailed description of the work instrumentation is presented in Section~\ref{chap:instr}.
-
-Notice how the \textit{Work Instrumentation} phase is executed before performing any optimization to the program.
-It is critical so that the work profiling always measure the same amount of work for a given input, regardless of the optimization sequence applied to the program.
-Because the instrumentation is performed before optimising the program, it means that the work profiling derives the linear expression defined in Equation~\ref{eq:linear-work-expression}
-based on the unoptimized program.
-In other words, the basic blocks and the number of occurrences of the instruction in the basic blocks reflect the unoptimized program.
-This particular sequence of compilation guarantees that the amount of work must is only input dependent, but consistent between different optimization sequences.
-
-As the name suggests, the component called \textit{Optimization Selector} is responsible for selecting which optimization sequence to use for the next execution of the program.
-It can either keep the same optimization sequence used in the previous execution or start monitoring the performance of a new optimization sequence.
-An optimization sequence can be kept for multiple executions of the program in order to gather enough measurement to compute an average performance with small statistical deviations, i.e., for which the confidence interval has a small range.
-We call by \textit{Input-Window Size} the number of executions performed using the same optimization sequence.
-
-\subsection{Real Online Scenarios}
-
-In most online scenarios, it is common for periods of peak usage and idle periods.
-For example, mobile devices are usually intensely used during the day, with some idle periods usually when the battery is being re-charged~\citep{mpeis16}.
-Similarly, many authors have also considered peak and idle (or underutilized) periods in the context of data centres~\citep{armbrust10,chen12b}.
-
-The proposed infrastructure is very well suited for these real online scenarios\footnote{
-However, if idle time is almost non-existent, the proposed infrastructure can still be used by re-compiling the program with a different optimization while multiple runs of the program are being executed.}.
-In particular, periods of peak usage could be used to monitor multiple runs of the program using the same optimization sequence, collecting the work profiling and measuring its execution time,
-while periods of idleness or underutilisation could be leveraged to use the profiling statistics for selecting better optimizations and re-compiling the program.
-
-\section{Work Profiling}
-
-In this section, we describe how the computation of the work metric can be performed during runtime by means of instrumenting the code.
-%In particular, we adapt the optimal algorithm proposed originally for block-frequency profiling~\citep{nahapetian73,knuth73,ball94}.
-Because we define work as a linear equation on the block-frequency counters, it is possible to embed its computation into the execution of the program.
-A naive instrumentation would consist basically of having a global counter that starts with the interception value, $\varepsilon$, and each basic block increments its own cost into the global counter.
-Although this instrumentation is easily implemented, it imposes a significant overhead on the instrumented program.
-However, it is possible to insert fewer probes by carefully placing the probes in a way that is possible to reconstruct the complete profiling information~\citep{knuth73,ball94}.
-
-We adapt the optimal block-frequency profiling in order to perform the work profiling efficiently.
-The proposed work profiling differs from the optimal block-frequency profiling as the latter occurs in two stages:
-\textit{(i.) Before execution:} The code is instrumented with counters for each probe.
-\textit{(ii.) After execution:} The information from the recorded probes is propagated in the CFGs of the program.
-In contrast, our work profiling has a single counter, and it only requires the instrumentation before execution, without any post-processing of the recorded profiling.
-
-The instrumented code is assumed to be generated before optimising the code.
-This assumption is based on three key points:
-\textit{(i.)} it guarantees that the work metric is independent of optimisation, i.e., the same input is always mapped to the same amount of work, regardless of the optimisation;
-\textit{(ii.)} it simplifies the code generator;
-\textit{(iii.)} it leverages from the optimisations to further improve the instrumentation code.
-Having a work metric that is independent of optimisation is the most important reason for which we instrument the code before optimisations.
-
-The optimal placement of the probes works by first computing the maximum spanning
-tree based on a weighing that assigns a value to each edge in the CFG.
-These weights can be obtained either by empirical measurements or heuristic estimations,
-and their goal is to avoid probing in frequently executed edges.
-Once we have the maximum spanning tree, probes are placed on every edge not in the spanning tree.
-Figure~\ref{fig:cfg-example} shows an example of a CFG with a maximum spanning tree.
-%represented by the black edges, while the edges highlighted in red represent the placement of the probes.
-
-\begin{figure}[t]
-\centering {
-  \includegraphics[scale=0.75]{figs/cfg-example.pdf}\\\vspace{1ex}
-  \resizebox{0.45\textwidth}{!}{
-  %\scalebox{0.8}{
-     \begin{minipage}{0.5\textwidth}
-     Instrumented value for each probe $P_i$:
-     \begin{align*}
-     \omega(P_0) &= w(B_0) + w(B_1) + w(B_4) + w(B_6) + w(B_7) + w(B_8) + w(B_9)\\
-     \omega(P_1) &= w(B_2) - w(B_1) - w(B_4) - w(B_6) - w(B_7) - w(B_8)\\
-     \omega(P_2) &= w(B_5) - w(B_6)\\
-     \omega(P_3) &= w(B_3) - w(B_4) - w(B_6) - w(B_7)
-     \end{align*}
-     \end{minipage}
-  }
-}
-  \caption{Example of a CFG with its minimum spanning tree in black and the
-   basic blocks highlighted in red represent the instrumented basic blocks with
-   the placement of the probes.}
-  \label{fig:cfg-example}
-\end{figure}
-
-In contrast to the naive instrumentation where each basic block records only its
-own amount of work, with the optimal profiling, the instrumented basic blocks need
-to record an aggregated value of work that also takes other basic blocks into
-account.
-Some probes speculatively assume a path of execution,
-while other probes correct when these assumptions are wrong
-(see for example $w(P_0)$ and $w(P_1)$ in Figure~\ref{fig:cfg-example}).
-
-We compute the aggregated value of work for each probe in two steps:
-\textit{(i.)} we propagate information about the probes through the edges of the CFG,
-as symbolic expressions, and then \textit{(ii.)} we use this edge flows
-to compose the aggregated value of the probes.
-
-The symbolic expressions in the edges can then be used to compose the aggregated
-value of the probes.
-These symbolic expressions describe the probes that are in the same path of the
-edge (positive terms) and those that are in complementary paths (negative
-terms).
-Therefore, the positive terms in the summed symbolic expression of a
-basic block indicate that the amount of work of this basic block will be
-incremented in the probes represented by these positive terms.
-Similarly, the negative terms indicate that the amount of work of this basic block
-will be decremented in the probes represented by these negative terms.
-For example, because the symbolic expression for the basic block $B_8$ is $P_0 - P_1$,
-the amount of work of $B_8$, denoted by $w(B_8)$, is incremented in probe $P_0$
-and decremented in $P_1$.
-
-\paragraph{Populating the edge flows:}
-Intuitively, if all the edge flows are known for the complement of a spanning tree
-then at any leaf of the spanning tree there is only one unknown edge flow.
-This unknown edge flow can be calculated by Kirchhoff's first law.
-This process repeats as a bottom-up propagation until all the unknown edge flows
-have been calculated.
-This algorithm can be formally defined as a post-order traversal on the spanning
-tree.
-Let $G$ be the CFG.
-This algorithm first initializes the edge flows $D_{(u,v)}$, for each edge
-$(u,v)$ in the CFG, as follows:
-\[
-D_{(u,v)} \gets
-\begin{cases}
-    P_{(u,v)} & \quad \text{if edge $(u,v)$ has a probe $P_{(u,v)}$}\\
-    0       & \quad \text{otherwise}
-\end{cases}
-\]
-Afterwards, for each vertex $u\in G$, in a post-order traversal of the spanning tree,
-we can use the Kirchhoff's first law and apply the following operations
-in a symbolic fashion:
-\[
-\Sigma^+_u = \sum_{v\in N^+(u)} D_{(u,v)}
-\]
-\[
-\Sigma^-_u = \sum_{v\in N^-(u)} D_{(v,u)}
-\]
-\[
-\forall v\in N^+(u):  D_{(u,v)} \gets
-\begin{cases}
-    D_{(u,v)} & \quad \text{if $D_{(u,v)}\neq 0$}\\
-    \Sigma^-_u - \Sigma^+_u       & \quad \text{otherwise}
-\end{cases}
-\]
-\[
-\forall v\in N^-(u):  D_{(v,u)} \gets
-\begin{cases}
-    D_{(v,u)} & \quad \text{if $D_{(v,u)}\neq 0$}\\
-    \Sigma^+_u - \Sigma^-_u       & \quad \text{otherwise}
-\end{cases}
-\]
-
-\paragraph{Composing the aggregated value of the probes:}
-If $x$ is a symbolic expression, then $\kappa_P(x)$ represents the coefficient
-of the term $P$ in $x$.
-In our case, $\kappa_P(x)$ will belong to the set $\{-1,0,1\}$.
-Then, for all probes $P$:
-\[
-\omega(P) = \sum_{u\in G} \kappa_P(\Sigma^+_u)w(u)
-\]
-
-\subsection{Relaxed Instrumentation}
-
-Although the optimal instrumentation significantly reduces the profiling
-overhead when compared to the naive instrumentation, from an average overhead
-of 79\% to 13\%, in some critical cases, even the optimal instrumentation can
-result in overheads of up to 60\% (see benchmark \texttt{adpcm\_d} in Figure~\ref{fig:overhead-O3}).
-In order to further reduce the overhead in these critical cases, we propose a
-relaxation strategy that offers a trade-off between accuracy and overhead.
-While the optimal instrumentation places probes in edges that are
-less likely to be executed, the relaxation removes probes that are
-more likely to be executed but add little to the final work metric.
-
-% \begin{figure}[h]
-%   \centering
-%   \includegraphics[scale=0.85]{figs/relax-instr-diagram.pdf}
-%   \caption{Overview of the work instrumentation algorithm, including the relaxation technique.}
-%   \label{fig:relax-instr-diagram}
-% \end{figure}
-
-%Figure~\ref{fig:relax-instr-diagram} shows an overview of the relaxed instrumentation algorithm.
-%The highlighted step is introduced by the relaxed instrumentation on top of the previously defined optimal profiling.
-The relaxation strategy performs a post processing on the resulting instrumentation of the optimal algorithm.
-This post processing identifies probes that add little to the work metric, and removes their instrumentation.
-In order to guarantee an upper bound for the dynamic error in the profiling measurement,
-the relaxation algorithm applies a constrained post-processing on a per DAG (directed acyclic graph) basis.
-By constraining the relaxation within each DAG by a maximum allowed static error,
-it guarantees that the overall relaxation will also be constrained by the same bound.
-
-The relaxation starts by extracting DAGs from the CFG.
-First, the algorithm extracts all the subgraphs that represent a loop or the outer most region of the function.
-Afterwards, these subgraphs are transformed into DAGs by ignoring the backedge
-and also by considering that any loop within the subgraph is never executed,
-i.e., only the headers of the inner loops are actually included into the DAG.
-Figure~\ref{fig:cfg-relax-example} shows a CFG partitioned into two DAGs (consider
-only basic blocks and edges completely inside the yellow and green boundaries).
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[scale=0.75]{figs/cfg-relax-example.pdf}
-  \caption{Example of a CFG containing a loop and its decomposition into DAGs.
-           The DAGs are the subgraphs within the dashed boundaries.}
-  \label{fig:cfg-relax-example}
-\end{figure}
-
-%\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
-%// Input: CFG
-%relaxInstrumentation(G) {
-%  for loop in G:
-%     DAG = colapseInnerLoops(loop)
-%     relaxInstrumentedDAG(DAG)
-%  DAG = colapseInnerLoops(G)
-%  relaxInstrumentedDAG(DAG)
-%}
-%\end{lstlisting}
-
-For every DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$,
-we select a subset of the probes to be removed, subject to
-the maximum allowed percentage error, $M$.
-This strategy can be modeled as a 0-1 Knapsack problem:
-\begin{gather*}
-\textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
-\textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
-x_i\in\{0,1\}, i\in\{0,\ldots,k\}
-\end{gather*}
-where $f(P_i)$ is the execution frequency of probe $P_i$, $x_i$ denotes the
-probes selected for removal, and $\varepsilon(P_i)$ is the percentage error of
-removing probe $P_i$ relative to the minimum work value possible to compute in
-the DAG, i.e., if $m$ is the minimum amount of work possible to be computed when
-executing the DAG, then $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$.
-Because the percentage error is computed based on the path with the minimum
-amount of work, $\varepsilon(P_i)$ represents the maximum error possible that
-would be incurred when removing probe $P_i$.
-Furthermore, by constraining the percentage error of every DAG below a given
-threshold, we guarantee that the final error of the relaxation will always be
-bounded by the threshold.
-% Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold, as demonstrated by Proposition~\ref{prop:relax-bound}.
-%
-% \begin{prop}\label{prop:relax-bound}
-% Let $n_i$ be the number of times a given DAG $i$ is executed, $r_i$ be the total relaxation (amount of work removed) in DAG $i$, and $m_i$ be its minimum amount of work.
-% If $\frac{r_i}{m_i} \leq M$ for every $i$,
-% then the final error of the relaxation will always be bounded by the same threshold.
-% \end{prop}
-% \begin{proof}
-% We can model the overall error of the relaxation as:
-% \[
-% 1 - \frac{n_1(m_1 - r_1) + n_2(m_2 - r_2) + \ldots + n_k(m_k - r_k) + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
-% \]
-% That is,
-% %\begin{equation*}
-% \begin{gather*}
-%  1 - \frac{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} + \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} = \\
-%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
-% \end{gather*}
-% %\end{equation*}
-% If $\frac{r_j}{m_j}$ is the maximum ratio $\frac{r_i}{m_i}$ for every $i$, then
-% \begin{equation*}
-% \begin{aligned}
-%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} &\leq\\
-%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j + c} &\leq\\
-%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &
-% \end{aligned}
-% \end{equation*}
-% If $N = max\{n_i$ for every $i\}$, then
-% \begin{equation*}
-% \begin{aligned}
-%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &\leq\\
-%  \frac{Nr_j + Nr_j + \ldots + Nr_j}{Nm_j + Nm_j + \ldots + Nm_j} &=\\
-%  \frac{Nkr_j}{Nkm_j} = \frac{r_j}{m_j} &\leq M
-% \end{aligned}
-% \end{equation*}
-% \end{proof}
-
-%\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
-%// Input: CFG
-%relaxInstrumentedDAG(DAG){
-%   P = ProbesIn(DAG)
-%   m = minWork(DAG,P)
-%   K = createKnapsackModel(P,m)
-%   Bag = solveKnapsack(K)
-%   for B in (P-Bag):
-%     removeProbe(B)
-%}
-%\end{lstlisting}
-
-%The necessary block-frequency information for optimising both the placement of
-%probes can be acquired from profiles of previous executions of the program or by
-%a static heuristic of the CFG during compilation.
-
-For our experiments, we implemented two solvers for the 0-1 Knapsack problem:
-the optimal brute-force solver and
-the greedy heuristic based on sorting the items~\citep{dantzig57}.
-We use the brute-force solver for DAGs with a small number of probes and the
-greedy heuristic when the number of probes is greater than a threshold.
-Some of the benchmarks have DAGs with several hundreds of probes, which could
-result in a long compilation time.
-
-\subsection{Whole Program Relaxation}
-
-In this section we go one step even further.
-In some cases, even the proposed relaxation strategy can still be conservative.
-This conservatism can be overly restrictive in some cases, resulting in a
-negligible overhead reduction but also causing just a negligible dynamic error
-to the work profiling.
-For these cases, we propose an adapted version of the relaxation algorithm that
-operates on the whole program.
-Traditionally, compiler optimizations are performed on the function-level, or at
-best on a per module basis.
-Whole program optimization (WPO) means that the compiler considers all
-compilation units of the program and optimizes them using the combined knowledge
-of how they are used together.
-
-The \textit{whole program relaxation} works by using block-frequency profiling
-from previous executions.
-By having this profiling information, the whole program relaxation is able to
-compute the error of removing a given probe in terms of the whole program's execution,
-and then select a subset of all the probes to be removed.
-
-For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we model the
-whole program relaxation as the following 0-1 Knapsack problem:
-\begin{gather*}
-\textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
-\textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
-x_i\in\{0,1\}, i\in\{0,\ldots,k\}
-\end{gather*}
-where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$,
-$x_i$ denotes the probes selected for removal, $M$ is the error threshold, and
-$\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to
-the profiled global work, i.e.,
-if $\Delta W$ is the work value for the whole program's execution, computed from
-the basic block frequencies profiled from previous executions, the error for a
-given probe $P_i$ is
-\[
-\varepsilon(P_i) = \frac{\omega(P_i)f(P_i)}{\Delta W}.
-\]
-
-Contrary to the per DAG relaxation, the whole program relaxation is not
-guaranteed to be bounded by the error threshold $M$,
-as it depends on the representativity % representativeness
-of the profiling information provided to the whole program relaxation.
-
+\input{work}
+\input{infra}
+\input{prof}
 \input{setup}
-
 \input{results}
-
 \input{related}
-
 \input{conclusions}

### 2017-09-14 infra.tex
Splitting

\section{Online {\IterComp} Infrastructure} \label{sec:oic-infra}

Although {\itercomp} had been originally proposed as an \textit{offline} optimization strategy, it can also be adapted to work in online scenarios.
Instead of selecting the best optimization sequence as part of the development time (pre-shipping) of a program, a first version of the program is shipped together with an {\itercomp} mechanism.
In the online scenario, the program is shipped with an initial optimization sequence and different optimization sequences are evaluated as the end-user executes the program.
This optimization strategy is also known as idle-time optimization, as the re-compilation happens between runs of the program.

LLVM is particularly suitable for iterative compilation as it makes possible to cache a pre-compiled, but still unoptimized, version of the input program in the bitcode format of the LLVM IR.
This caching allows to speedup the time required for re-compilation as it is able to bypass the frontend phase.
If re-compilation time is critical, it would also be possible to keep only the hot portion of the code in the LLVM bitcode format, while the remaining portion of the code is already compiled to the final object code.
However, this is out of the scope of this paper and we always re-compile the whole program.

\begin{figure}[htb]
    \centering
    \includegraphics[width=\linewidth]{figs/infra-diagram}
    \caption{Overview of the execution engine for applying {\itercomp}.}
    \label{fig:infra-diagram}
\end{figure}
%\textbf{Describe the process step by step.}

Figure~\ref{fig:infra-diagram} shows an overview of the infrastructure required for applying online {\itercomp}
using WP as the metric of choice for evaluating optimization sequences.
The online {\itercomp} follows as described bellow:
\begin{enumerate}
\item The program is pre-compiled to the LLVM bitcode format without optimization.
\item The unoptimized program is instrumented for work profiling.
\item Execution-based optimization search:
 \begin{enumerate}
   \item The current optimization sequence is used for the re-compilation of the program.
   \item The program is executed with any input provided by the end-user.
         During the execution of the program, wall-clock time and the work metric are recorded by profiling instrumentation.
   \item If the recorded profiling for the current optimization can be used to compute an average performance measurement within a small confidence interval,
         then a new optimization sequence is generated.
         Otherwise, the same optimization sequence is used for the next execution.
 \end{enumerate}
\end{enumerate}

In this work, we focus mainly on the highlighted components.
The \textit{Work Instrumentation} phase focuses on providing a low-overhead instrumentation for profiling the work metric.
The instrumentation consists of adding a global counter to the program which is used to accumulate the amount of work computed during the program's execution, using the cost model of the instruction set.
A detailed description of the work instrumentation is presented in Section~\ref{chap:instr}.

Notice how the \textit{Work Instrumentation} phase is executed before performing any optimization to the program.
It is critical so that the work profiling always measure the same amount of work for a given input, regardless of the optimization sequence applied to the program.
Because the instrumentation is performed before optimising the program, it means that the work profiling derives the linear expression defined in Equation~\ref{eq:linear-work-expression}
based on the unoptimized program.
In other words, the basic blocks and the number of occurrences of the instruction in the basic blocks reflect the unoptimized program.
This particular sequence of compilation guarantees that the amount of work must is only input dependent, but consistent between different optimization sequences.

As the name suggests, the component called \textit{Optimization Selector} is responsible for selecting which optimization sequence to use for the next execution of the program.
It can either keep the same optimization sequence used in the previous execution or start monitoring the performance of a new optimization sequence.
An optimization sequence can be kept for multiple executions of the program in order to gather enough measurement to compute an average performance with small statistical deviations, i.e., for which the confidence interval has a small range.
We call by \textit{Input-Window Size} the number of executions performed using the same optimization sequence.

\subsection{Real Online Scenarios}

In most online scenarios, it is common for periods of peak usage and idle periods.
For example, mobile devices are usually intensely used during the day, with some idle periods usually when the battery is being re-charged~\citep{mpeis16}.
Similarly, many authors have also considered peak and idle (or underutilized) periods in the context of data centres~\citep{armbrust10,chen12b}.

The proposed infrastructure is very well suited for these real online scenarios\footnote{
However, if idle time is almost non-existent, the proposed infrastructure can still be used by re-compiling the program with a different optimization while multiple runs of the program are being executed.}.
In particular, periods of peak usage could be used to monitor multiple runs of the program using the same optimization sequence, collecting the work profiling and measuring its execution time,
while periods of idleness or underutilisation could be leveraged to use the profiling statistics for selecting better optimizations and re-compiling the program.

### 2017-09-14 prof.tex
Splitting

\section{Work Profiling}

In this section, we describe how the computation of the work metric can be performed during runtime by means of instrumenting the code.
%In particular, we adapt the optimal algorithm proposed originally for block-frequency profiling~\citep{nahapetian73,knuth73,ball94}.
Because we define work as a linear equation on the block-frequency counters, it is possible to embed its computation into the execution of the program.
A naive instrumentation would consist basically of having a global counter that starts with the interception value, $\varepsilon$, and each basic block increments its own cost into the global counter.
Although this instrumentation is easily implemented, it imposes a significant overhead on the instrumented program.
However, it is possible to insert fewer probes by carefully placing the probes in a way that is possible to reconstruct the complete profiling information~\citep{knuth73,ball94}.

We adapt the optimal block-frequency profiling in order to perform the work profiling efficiently.
The proposed work profiling differs from the optimal block-frequency profiling as the latter occurs in two stages:
\textit{(i.) Before execution:} The code is instrumented with counters for each probe.
\textit{(ii.) After execution:} The information from the recorded probes is propagated in the CFGs of the program.
In contrast, our work profiling has a single counter, and it only requires the instrumentation before execution, without any post-processing of the recorded profiling.

The instrumented code is assumed to be generated before optimising the code.
This assumption is based on three key points:
\textit{(i.)} it guarantees that the work metric is independent of optimisation, i.e., the same input is always mapped to the same amount of work, regardless of the optimisation;
\textit{(ii.)} it simplifies the code generator;
\textit{(iii.)} it leverages from the optimisations to further improve the instrumentation code.
Having a work metric that is independent of optimisation is the most important reason for which we instrument the code before optimisations.

The optimal placement of the probes works by first computing the maximum spanning
tree based on a weighing that assigns a value to each edge in the CFG.
These weights can be obtained either by empirical measurements or heuristic estimations,
and their goal is to avoid probing in frequently executed edges.
Once we have the maximum spanning tree, probes are placed on every edge not in the spanning tree.
Figure~\ref{fig:cfg-example} shows an example of a CFG with a maximum spanning tree.
%represented by the black edges, while the edges highlighted in red represent the placement of the probes.

\begin{figure}[t]
\centering {
  \includegraphics[scale=0.75]{figs/cfg-example.pdf}\\\vspace{1ex}
  \resizebox{0.45\textwidth}{!}{
  %\scalebox{0.8}{
     \begin{minipage}{0.5\textwidth}
     Instrumented value for each probe $P_i$:
     \begin{align*}
     \omega(P_0) &= w(B_0) + w(B_1) + w(B_4) + w(B_6) + w(B_7) + w(B_8) + w(B_9)\\
     \omega(P_1) &= w(B_2) - w(B_1) - w(B_4) - w(B_6) - w(B_7) - w(B_8)\\
     \omega(P_2) &= w(B_5) - w(B_6)\\
     \omega(P_3) &= w(B_3) - w(B_4) - w(B_6) - w(B_7)
     \end{align*}
     \end{minipage}
  }
}
  \caption{Example of a CFG with its minimum spanning tree in black and the
   basic blocks highlighted in red represent the instrumented basic blocks with
   the placement of the probes.}
  \label{fig:cfg-example}
\end{figure}

In contrast to the naive instrumentation where each basic block records only its
own amount of work, with the optimal profiling, the instrumented basic blocks need
to record an aggregated value of work that also takes other basic blocks into
account.
Some probes speculatively assume a path of execution,
while other probes correct when these assumptions are wrong
(see for example $w(P_0)$ and $w(P_1)$ in Figure~\ref{fig:cfg-example}).

We compute the aggregated value of work for each probe in two steps:
\textit{(i.)} we propagate information about the probes through the edges of the CFG,
as symbolic expressions, and then \textit{(ii.)} we use this edge flows
to compose the aggregated value of the probes.

The symbolic expressions in the edges can then be used to compose the aggregated
value of the probes.
These symbolic expressions describe the probes that are in the same path of the
edge (positive terms) and those that are in complementary paths (negative
terms).
Therefore, the positive terms in the summed symbolic expression of a
basic block indicate that the amount of work of this basic block will be
incremented in the probes represented by these positive terms.
Similarly, the negative terms indicate that the amount of work of this basic block
will be decremented in the probes represented by these negative terms.
For example, because the symbolic expression for the basic block $B_8$ is $P_0 - P_1$,
the amount of work of $B_8$, denoted by $w(B_8)$, is incremented in probe $P_0$
and decremented in $P_1$.

\paragraph{Populating the edge flows:}
Intuitively, if all the edge flows are known for the complement of a spanning tree
then at any leaf of the spanning tree there is only one unknown edge flow.
This unknown edge flow can be calculated by Kirchhoff's first law.
This process repeats as a bottom-up propagation until all the unknown edge flows
have been calculated.
This algorithm can be formally defined as a post-order traversal on the spanning
tree.
Let $G$ be the CFG.
This algorithm first initializes the edge flows $D_{(u,v)}$, for each edge
$(u,v)$ in the CFG, as follows:
\[
D_{(u,v)} \gets
\begin{cases}
    P_{(u,v)} & \quad \text{if edge $(u,v)$ has a probe $P_{(u,v)}$}\\
    0       & \quad \text{otherwise}
\end{cases}
\]
Afterwards, for each vertex $u\in G$, in a post-order traversal of the spanning tree,
we can use the Kirchhoff's first law and apply the following operations
in a symbolic fashion:
\[
\Sigma^+_u = \sum_{v\in N^+(u)} D_{(u,v)}
\]
\[
\Sigma^-_u = \sum_{v\in N^-(u)} D_{(v,u)}
\]
\[
\forall v\in N^+(u):  D_{(u,v)} \gets
\begin{cases}
    D_{(u,v)} & \quad \text{if $D_{(u,v)}\neq 0$}\\
    \Sigma^-_u - \Sigma^+_u       & \quad \text{otherwise}
\end{cases}
\]
\[
\forall v\in N^-(u):  D_{(v,u)} \gets
\begin{cases}
    D_{(v,u)} & \quad \text{if $D_{(v,u)}\neq 0$}\\
    \Sigma^+_u - \Sigma^-_u       & \quad \text{otherwise}
\end{cases}
\]

\paragraph{Composing the aggregated value of the probes:}
If $x$ is a symbolic expression, then $\kappa_P(x)$ represents the coefficient
of the term $P$ in $x$.
In our case, $\kappa_P(x)$ will belong to the set $\{-1,0,1\}$.
Then, for all probes $P$:
\[
\omega(P) = \sum_{u\in G} \kappa_P(\Sigma^+_u)w(u)
\]

\subsection{Relaxed Instrumentation}

Although the optimal instrumentation significantly reduces the profiling
overhead when compared to the naive instrumentation, from an average overhead
of 79\% to 13\%, in some critical cases, even the optimal instrumentation can
result in overheads of up to 60\% (see benchmark \texttt{adpcm\_d} in Figure~\ref{fig:overhead-O3}).
In order to further reduce the overhead in these critical cases, we propose a
relaxation strategy that offers a trade-off between accuracy and overhead.
While the optimal instrumentation places probes in edges that are
less likely to be executed, the relaxation removes probes that are
more likely to be executed but add little to the final work metric.

% \begin{figure}[h]
%   \centering
%   \includegraphics[scale=0.85]{figs/relax-instr-diagram.pdf}
%   \caption{Overview of the work instrumentation algorithm, including the relaxation technique.}
%   \label{fig:relax-instr-diagram}
% \end{figure}

%Figure~\ref{fig:relax-instr-diagram} shows an overview of the relaxed instrumentation algorithm.
%The highlighted step is introduced by the relaxed instrumentation on top of the previously defined optimal profiling.
The relaxation strategy performs a post processing on the resulting instrumentation of the optimal algorithm.
This post processing identifies probes that add little to the work metric, and removes their instrumentation.
In order to guarantee an upper bound for the dynamic error in the profiling measurement,
the relaxation algorithm applies a constrained post-processing on a per DAG (directed acyclic graph) basis.
By constraining the relaxation within each DAG by a maximum allowed static error,
it guarantees that the overall relaxation will also be constrained by the same bound.

The relaxation starts by extracting DAGs from the CFG.
First, the algorithm extracts all the subgraphs that represent a loop or the outer most region of the function.
Afterwards, these subgraphs are transformed into DAGs by ignoring the backedge
and also by considering that any loop within the subgraph is never executed,
i.e., only the headers of the inner loops are actually included into the DAG.
Figure~\ref{fig:cfg-relax-example} shows a CFG partitioned into two DAGs (consider
only basic blocks and edges completely inside the yellow and green boundaries).

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.75]{figs/cfg-relax-example.pdf}
  \caption{Example of a CFG containing a loop and its decomposition into DAGs.
           The DAGs are the subgraphs within the dashed boundaries.}
  \label{fig:cfg-relax-example}
\end{figure}

%\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
%// Input: CFG
%relaxInstrumentation(G) {
%  for loop in G:
%     DAG = colapseInnerLoops(loop)
%     relaxInstrumentedDAG(DAG)
%  DAG = colapseInnerLoops(G)
%  relaxInstrumentedDAG(DAG)
%}
%\end{lstlisting}

For every DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$,
we select a subset of the probes to be removed, subject to
the maximum allowed percentage error, $M$.
This strategy can be modeled as a 0-1 Knapsack problem:
\begin{gather*}
\textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
\textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
x_i\in\{0,1\}, i\in\{0,\ldots,k\}
\end{gather*}
where $f(P_i)$ is the execution frequency of probe $P_i$, $x_i$ denotes the
probes selected for removal, and $\varepsilon(P_i)$ is the percentage error of
removing probe $P_i$ relative to the minimum work value possible to compute in
the DAG, i.e., if $m$ is the minimum amount of work possible to be computed when
executing the DAG, then $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$.
Because the percentage error is computed based on the path with the minimum
amount of work, $\varepsilon(P_i)$ represents the maximum error possible that
would be incurred when removing probe $P_i$.
Furthermore, by constraining the percentage error of every DAG below a given
threshold, we guarantee that the final error of the relaxation will always be
bounded by the threshold.
% Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold, as demonstrated by Proposition~\ref{prop:relax-bound}.
%
% \begin{prop}\label{prop:relax-bound}
% Let $n_i$ be the number of times a given DAG $i$ is executed, $r_i$ be the total relaxation (amount of work removed) in DAG $i$, and $m_i$ be its minimum amount of work.
% If $\frac{r_i}{m_i} \leq M$ for every $i$,
% then the final error of the relaxation will always be bounded by the same threshold.
% \end{prop}
% \begin{proof}
% We can model the overall error of the relaxation as:
% \[
% 1 - \frac{n_1(m_1 - r_1) + n_2(m_2 - r_2) + \ldots + n_k(m_k - r_k) + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
% \]
% That is,
% %\begin{equation*}
% \begin{gather*}
%  1 - \frac{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} + \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} = \\
%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
% \end{gather*}
% %\end{equation*}
% If $\frac{r_j}{m_j}$ is the maximum ratio $\frac{r_i}{m_i}$ for every $i$, then
% \begin{equation*}
% \begin{aligned}
%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} &\leq\\
%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j + c} &\leq\\
%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &
% \end{aligned}
% \end{equation*}
% If $N = max\{n_i$ for every $i\}$, then
% \begin{equation*}
% \begin{aligned}
%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &\leq\\
%  \frac{Nr_j + Nr_j + \ldots + Nr_j}{Nm_j + Nm_j + \ldots + Nm_j} &=\\
%  \frac{Nkr_j}{Nkm_j} = \frac{r_j}{m_j} &\leq M
% \end{aligned}
% \end{equation*}
% \end{proof}

%\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
%// Input: CFG
%relaxInstrumentedDAG(DAG){
%   P = ProbesIn(DAG)
%   m = minWork(DAG,P)
%   K = createKnapsackModel(P,m)
%   Bag = solveKnapsack(K)
%   for B in (P-Bag):
%     removeProbe(B)
%}
%\end{lstlisting}

%The necessary block-frequency information for optimising both the placement of
%probes can be acquired from profiles of previous executions of the program or by
%a static heuristic of the CFG during compilation.

For our experiments, we implemented two solvers for the 0-1 Knapsack problem:
the optimal brute-force solver and
the greedy heuristic based on sorting the items~\citep{dantzig57}.
We use the brute-force solver for DAGs with a small number of probes and the
greedy heuristic when the number of probes is greater than a threshold.
Some of the benchmarks have DAGs with several hundreds of probes, which could
result in a long compilation time.

\subsection{Whole Program Relaxation}

In this section we go one step even further.
In some cases, even the proposed relaxation strategy can still be conservative.
This conservatism can be overly restrictive in some cases, resulting in a
negligible overhead reduction but also causing just a negligible dynamic error
to the work profiling.
For these cases, we propose an adapted version of the relaxation algorithm that
operates on the whole program.
Traditionally, compiler optimizations are performed on the function-level, or at
best on a per module basis.
Whole program optimization (WPO) means that the compiler considers all
compilation units of the program and optimizes them using the combined knowledge
of how they are used together.

The \textit{whole program relaxation} works by using block-frequency profiling
from previous executions.
By having this profiling information, the whole program relaxation is able to
compute the error of removing a given probe in terms of the whole program's execution,
and then select a subset of all the probes to be removed.

For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we model the
whole program relaxation as the following 0-1 Knapsack problem:
\begin{gather*}
\textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
\textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
x_i\in\{0,1\}, i\in\{0,\ldots,k\}
\end{gather*}
where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$,
$x_i$ denotes the probes selected for removal, $M$ is the error threshold, and
$\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to
the profiled global work, i.e.,
if $\Delta W$ is the work value for the whole program's execution, computed from
the basic block frequencies profiled from previous executions, the error for a
given probe $P_i$ is
\[
\varepsilon(P_i) = \frac{\omega(P_i)f(P_i)}{\Delta W}.
\]

Contrary to the per DAG relaxation, the whole program relaxation is not
guaranteed to be bounded by the error threshold $M$,
as it depends on the representativity % representativeness
of the profiling information provided to the whole program relaxation.

### 2017-09-14 work.tex
Splitting

\section{Work Efficiency Metric} \label{sec:metric}

In this section we define work efficiency, a metric for comparing different optimized versions of a program when executing with different
inputs. We define it as the ratio between the amount of \textit{work}, $\Delta W$, performed during a period of time, $\Delta t$.

\[
   P = \nicefrac{\Delta W}{\Delta t}
\]

The main challenge is to precisely define what represents \textit{work}. 
We define the work done by a program on an input to be the time taken by the unoptimized program to process that input.
We estimate the runtime of each instruction type via a linear regression and then define the work done by a program as the sum of the

We adopt a work metric that takes into account the
computational cost within each basic block.
The computational cost of a basic block is given by weighting the instructions
that it contains.

To this end, we model the computational work $\Delta W$ as a linear equation based on block frequency information and a cost model of the instruction set.
Formally,
\[
\Delta W = \varepsilon + \sum_{B} w(B)f(B)
\]
where $f(B)$ represents the frequency of basic block $B$ and $w(B)$ represents the computational work of executing $B$.
We define the work of a basic block $B$ as the sum of the cost of its instructions, i.e.,
\[
w(B) = \sum_{i} w_i N_B(i)
\]
where $w_i$ is the cost of instruction $i$ and $N_B(i)$ is the number of occurrences of instruction $i$ in basic block $B$.

In this simplified model, we consider that $w_i$ is constant across all programs and executions, varying only between target architectures.
On the other hand, $N_B(i)$ is a program-dependent static value which is known at compile-time and $f(B)$ is a dynamic value known only at run-time,
since $f(B)$ is both program and execution dependent as the execution frequency of a basic block can change when executing with different inputs.

\subsection{Estimating a Cost Model of the Instructions}

Similarly to previous work~\citep{giusto01,powell09,brandolese11}, we derive the cost model of the instruction set by modelling the problem as a multi-variable linear regression, where the \textit{regression coefficients} are the costs of the instructions and the \textit{regressors} are computed as $\sum_B N_B(i)f(B)$ for each instruction.
\begin{equation}\label{eq:linear-work-expression}
\Delta W = \varepsilon + \sum_{i} \left(w_i \sum_{B} N_B(i)f(B)\right)
\end{equation}
By having some empirical data after executing several benchmarks with different inputs, we can fit the linear model with this empirical data in order to obtain estimate costs of the instructions.
In order to fit the linear model, we measure the wall-clock time when executing the training benchmarks described in Section\ref{sec:benchmarks} with their respective 1000 input datasets.
For these measurements, the training benchmarks are compiled without optimization.
The reason for using no optimization, as discussed in Section\ref{sec:oic-infra}, is because the amount of work must be only input dependent and consistent between different optimization sequences.
This aspect is crucial for the work-based performance metric to enable a consistent comparison between the various optimization sequences when executing with distinct inputs.
%it allows to use the estimated cost model for computing a work metric which is independent of optimizations, as we explained in Section~\ref{}.

Our cost model has a total of 52 LLVM instructions\footnote{We do not model all LLVM instructions because some instructions are more common in optimized programs, such as the vector instructions.}.
Every program, in the set of training benchmarks, was compiled twice: once including the necessary instrumentation for the block frequency profiling, which is required for deriving the linear expression defined in Equation~\ref{eq:linear-work-expression};
and a standard compilation without any optimization or instrumentation.
For each input in the training dataset, the benchmarks were executed once with the instrumented version for collecting the block frequency profiling, and multiple times with the standard compilation just for measuring the wall-clock execution time, until the confidence interval was no larger than 1\% for a 99\% confidence.
After collecting these measurements, this data can be used to estimate the unknown parameters in the linear model.
Because we fit the linear model based on the wall-clock execution time, the derived cost model can be interpreted as an estimate of the execution time when the program is compiled without optimizations.

\begin{figure}[t]
    \centering
    \includegraphics[width=0.6\linewidth]{figs/cost-model.pdf}
    \caption{Linear model fitted from empirical data. The mean absolute error (MAE) for the fitted curve is seven milliseconds.}
    \red{zw: bigger font size.}
    \label{fig:cost-model}
\end{figure}

Figure~\ref{fig:cost-model} compare the work metric with the corresponding execution time for some instances of the test benchmarks.
Notice how the fitted model has a higher relative error for the instances with very short execution time, namely those that run for less than one tenth of a second.
The mean absolute error (MAE) for the fitted curve is seven milliseconds.

\subsection{Comparison with Instructions Per Cycle} \label{sec:ipc-vs-work-metric}

The IPC metric has been widely used for studying performance benefits of hardware optimizations.
Although previous work has suggested the use of IPC for guiding {\itercomp}, in this section we argue in favour of the WP metric over IPC.

The IPC metric differs from the proposed WP metric in a key aspect:
the IPC metric is computed solely based on the final optimized program.
When executing different optimized versions of a program with the same input, both the number of instructions and the number of clock cycles can change.
For this reason, higher IPC does not necessarily translate to shorter execution time (or even fewer clock cycles).

% We can illustrate this fact with a very small example as shown in Table~\ref{tab:ipc-example}.
% This example shows two versions of a program, namely P1 and P2, and their respective measurements related to IPC.
% Although version P1 has twice the IPC of P2, P1 is one cycle slower than P2.
% As this example illustrates, the IPC metric can be misleading when compared different versions of the same program.
% This problem is only intensified when both the optimization sequence and the input changes.
% In Chapter~\ref{chap:eval} we show empirical evidence for this argument.
%
% \begin{table}[h]
% \centering
% \begin{tabular}{|c|c|c|c|}
% \hline
%                        & P1 & P2  \\
% \hline
% Number of Instructions & 5  & 2   \\
% Number of Cycles       & 5  & 4   \\
% IPC                    & 1  & 0.5 \\
% \hline
% \end{tabular}
% \caption{Example that illustrates that a higher IPC does not necessarily translate to shorter execution time.}
% \label{tab:ipc-example}
% \end{table}

On the other hand, WP computes the amount of work based on the unoptimized version of the program, which means that it always measures the same amount of work for the same input, regardless of the optimizations applied on the program.
This is a key aspect that enables the use of the WP metric for guiding {\itercomp}, because a higher WP, which represents work per unit time, naturally translate to shorter execution time.
Moreover, previous work has also presented other arguments against the use of IPC in similar use-case scenarios. %, as discussed in Chapter~\ref{chap:related}.

### 2017-09-14 infra.tex
itercomp overview

@@
-\section{Online {\IterComp} Infrastructure} \label{sec:oic-infra}
+\section{Online {\IterComp} Overview} \label{sec:oic-infra}
 
-Although {\itercomp} had been originally proposed as an \textit{offline} optimization strategy, it can also be adapted to work in online scenarios.
-Instead of selecting the best optimization sequence as part of the development time (pre-shipping) of a program, a first version of the program is shipped together with an {\itercomp} mechanism.
-In the online scenario, the program is shipped with an initial optimization sequence and different optimization sequences are evaluated as the end-user executes the program.
-This optimization strategy is also known as idle-time optimization, as the re-compilation happens between runs of the program.
-
-LLVM is particularly suitable for iterative compilation as it makes possible to cache a pre-compiled, but still unoptimized, version of the input program in the bitcode format of the LLVM IR.
-This caching allows to speedup the time required for re-compilation as it is able to bypass the frontend phase.
-If re-compilation time is critical, it would also be possible to keep only the hot portion of the code in the LLVM bitcode format, while the remaining portion of the code is already compiled to the final object code.
-However, this is out of the scope of this paper and we always re-compile the whole program.
+Although {\itercomp} had been originally proposed as an \textit{offline} optimization strategy, it can be adapted to work in
+\textit{online} scenarios. Instead of selecting the best optimization sequence at development time, a first version of the program is
+shipped bundled with the {\itercomp} mechanism. Our mechanism produces new application binaries using different optimization sequences.
+When the user starts the application, we load and execute one of these binaries to evaluate the corresponding sequence online. Our
+mechanism operates on the application's Intermediate Representation (IR) produced by LLVM. This saves us the effort of repeatedly parsing
+the code through the frontend and removes the need for shipping the full application code.
 
@@
     \includegraphics[width=\linewidth]{figs/infra-diagram}
-    \caption{Overview of the execution engine for applying {\itercomp}.}
+    \caption{Overview of the online \itercomp mechanism.}
     \label{fig:infra-diagram}
 \end{figure}
-%\textbf{Describe the process step by step.}
 
-Figure~\ref{fig:infra-diagram} shows an overview of the infrastructure required for applying online {\itercomp}
-using WP as the metric of choice for evaluating optimization sequences.
-The online {\itercomp} follows as described bellow:
-\begin{enumerate}
-\item The program is pre-compiled to the LLVM bitcode format without optimization.
-\item The unoptimized program is instrumented for work profiling.
-\item Execution-based optimization search:
- \begin{enumerate}
-   \item The current optimization sequence is used for the re-compilation of the program.
-   \item The program is executed with any input provided by the end-user.
-         During the execution of the program, wall-clock time and the work metric are recorded by profiling instrumentation.
-   \item If the recorded profiling for the current optimization can be used to compute an average performance measurement within a small confidence interval,
-         then a new optimization sequence is generated.
-         Otherwise, the same optimization sequence is used for the next execution.
- \end{enumerate}
-\end{enumerate}
+Figure~\ref{fig:infra-diagram} shows an overview of how our online {\itercomp} mechanism works. On the developer's side, the program is
+pre-compiled to IR without optimizations and is instrumented to measure the amount of work performed and the wall-clock time. On the user's
+side, we select an optimization sequence and then compile the instrumented IR to binary using this sequence. Next time the user runs the
+application, we measure the work and the execution time. We accumulate multiple such measurements until we can estimate the efficiency
+metric P of this optimization sequence with high confidence. Finally, we use this information to help us select the next optimization
+sequence.
 
-In this work, we focus mainly on the highlighted components.
-The \textit{Work Instrumentation} phase focuses on providing a low-overhead instrumentation for profiling the work metric.
-The instrumentation consists of adding a global counter to the program which is used to accumulate the amount of work computed during the program's execution, using the cost model of the instruction set.
-A detailed description of the work instrumentation is presented in Section~\ref{chap:instr}.
+This work focuses on the highlighted components of Figure~\ref{fig:infra-diagram}. The first one, \textit{Work Instrumentation} allows
+measuring the amount of work performed. Conceptually, we use the cost model of Section~\ref{sec:metric} to estimate a cost for each basic
+block of the IR and then add instrumentation to increase a global work counter by that cost every time a basic block is executed. The
+measured work should depend only on the program and the input, not the applied optimizations, so we instrument the unoptimized IR before
+any optimizations modify it. We present a detailed description of work instrumentation and how to keep it lightweight in
+Section~\ref{chap:instr}.
 
-Notice how the \textit{Work Instrumentation} phase is executed before performing any optimization to the program.
-It is critical so that the work profiling always measure the same amount of work for a given input, regardless of the optimization sequence applied to the program.
-Because the instrumentation is performed before optimising the program, it means that the work profiling derives the linear expression defined in Equation~\ref{eq:linear-work-expression}
-based on the unoptimized program.
-In other words, the basic blocks and the number of occurrences of the instruction in the basic blocks reflect the unoptimized program.
-This particular sequence of compilation guarantees that the amount of work must is only input dependent, but consistent between different optimization sequences.
+\textit{Optimization Selector} is responsible for selecting which optimization sequence to use for the next execution of the program. Each
+sequence is used for multiple executions of the program in order to gather enough work efficiency measurements. When we are able to
+estimate this work efficiency with a confidence interval narrower than a certain threshold, we can start evaluating a different
+optimization sequence. We call the number of executions performed using the same optimization sequence the \textit{Input-Window Size}.
 
-As the name suggests, the component called \textit{Optimization Selector} is responsible for selecting which optimization sequence to use for the next execution of the program.
-It can either keep the same optimization sequence used in the previous execution or start monitoring the performance of a new optimization sequence.
-An optimization sequence can be kept for multiple executions of the program in order to gather enough measurement to compute an average performance with small statistical deviations, i.e., for which the confidence interval has a small range.
-We call by \textit{Input-Window Size} the number of executions performed using the same optimization sequence.
+\FIXME{WHAT IS THE OPTIMIZATION EVALUATOR? WHY IS IT HIGHLIGHTED? DO WE HAVE ANYTHING TO SAY ABOUT IT?}
 
-\subsection{Real Online Scenarios}
+\subsection{Efficient Optimization Selection and Compilation}
 
-In most online scenarios, it is common for periods of peak usage and idle periods.
-For example, mobile devices are usually intensely used during the day, with some idle periods usually when the battery is being re-charged~\citep{mpeis16}.
-Similarly, many authors have also considered peak and idle (or underutilized) periods in the context of data centres~\citep{armbrust10,chen12b}.
-
-The proposed infrastructure is very well suited for these real online scenarios\footnote{
-However, if idle time is almost non-existent, the proposed infrastructure can still be used by re-compiling the program with a different optimization while multiple runs of the program are being executed.}.
-In particular, periods of peak usage could be used to monitor multiple runs of the program using the same optimization sequence, collecting the work profiling and measuring its execution time,
-while periods of idleness or underutilisation could be leveraged to use the profiling statistics for selecting better optimizations and re-compiling the program.
+Most devices have distinct idle and peak usage periods. For example, mobile devices are used intensely during the day and not at all while
+the user is sleeping and the battery is charging~\citep{mpeis16}. Many authors have identified similar idle and peak usage periods in the
+context of data centers~\citep{armbrust10,chen12b}. The proposed \itercomp infrastructure is very well suited to such scenarios. In
+particular, we can use periods of idleness or underutilization to select the next few optimization sequences to try and compile the
+corresponding binaries. This approach effectively hides the compilation overhead from the user. During peak periods, we only have to select
+which of the existing binaries to evaluate which introduces almost no overhead.

### 2017-09-14 work.tex
fix intro; add figure about work-efficiency 'motivation'.

@@
 
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.5\textwidth]{figs/motivation-work-efficiency.pdf}
+    \caption{Relationship between work efficiency vs speedup for 500 different binaries of \texttt{susan\_c}.
+            Each point represents averages over a subset of all inputs.}
+    \label{fig:motivation-work-efficiency}
+\end{figure}
+
+\FIXME{Rodrigo: I think we could show how the work efficiency metric performs compared with speedup, the same way we showed the other metrics does not work well.}
+
 \subsection{Comparison with Instructions Per Cycle} \label{sec:ipc-vs-work-metric}
+
+\FIXME{Rodrigo: Perhaps we could remove this section, since we have already discussed the IPC in the Motivation.}
 

### 2017-09-14 results.tex
arrrrrrrrrrrrrrrrrrrrrrrrrrrrrrrgh

@@
 
-In this section we discuss our experimental evaluation.
-First we describe the benchmarks with datasets used in the experiments.
-Afterwards, we discuss our results concerning the instrumentations for profiling our work metric.
-Finally we present the results of the online {\itercomp}.
+%In this section we discuss our experimental evaluation. Afterwards, we discuss our results concerning the instrumentations for profiling
+%our work metric. Finally we present the results of the online {\itercomp}.
 
@@
 
-We first compare static aspects of the instrumentation algorithms.
-The naive instrumentation always has 100\% of the basic blocks instrumented, by definition.
-First, we evaluate the optimal work profiling compared to the relaxation algorithm applied per DAG.
-Afterwards, we compare these profiling techniques with the whole program relaxation.
+We first compare static aspects of the instrumentation algorithms. The naive instrumentation instrument all basic blocks, by definition.
+First, we evaluate the optimal work profiling compared to the relaxation algorithm applied per DAG. Afterwards, we compare these profiling
+techniques with the whole program relaxation.
 

### 2017-09-14 work.tex
arrrrrrrrrrrrrrrrrrrrrrrrrrrrrrrgh

@@
 
-Similarly to previous work~\citep{giusto01,powell09,brandolese11}, we derive the cost model of the instruction set by modelling the problem as a multi-variable linear regression, where the \textit{regression coefficients} are the costs of the instructions and the \textit{regressors} are computed as $\sum_B N_B(i)f(B)$ for each instruction.
+Similarly to previous work~\citep{giusto01,powell09,brandolese11}, we derive the cost model of the instruction set by modeling the problem
+as a multi-variable linear regression. The \textit{regression coefficients} are the costs of the instructions and the \textit{regressors}
+are computed as $\sum_B N_B(i)f(B)$ for each instruction.
 \begin{equation}\label{eq:linear-work-expression}
@@
 \end{equation}
-By having some empirical data after executing several benchmarks with different inputs, we can fit the linear model with this empirical data in order to obtain estimate costs of the instructions.
-In order to fit the linear model, we measure the wall-clock time when executing the training benchmarks described in Section\ref{sec:benchmarks} with their respective 1000 input datasets.
-For these measurements, the training benchmarks are compiled without optimization.
-The reason for using no optimization, as discussed in Section\ref{sec:oic-infra}, is because the amount of work must be only input dependent and consistent between different optimization sequences.
-This aspect is crucial for the work-based performance metric to enable a consistent comparison between the various optimization sequences when executing with distinct inputs.
+
+We use a set of training programs to construct the regression model. Our training program set is given in
+Table~\ref{tab:kdatasets:training}. To determine the model coefficients, we profile each training benchmark with 1000 input datasets
+provided by the benchmark suite, and we measure the wall-clock time. It is to note that we turn off all compiler optimisations during this
+training stage, to ensure that the amount of work we measured only depends on the program input but not compiler optimization. This aspect
+is crucial for the work-based performance metric to allow a consistent comparison between the various optimization sequences when executing
+with distinct inputs.
+
 %it allows to use the estimated cost model for computing a work metric which is independent of optimizations, as we explained in Section~\ref{}.

### 2017-09-14 prof.tex
hmm
prof

@@
 
-In this section, we describe how the computation of the work metric can be performed during runtime by means of instrumenting the code.
-%In particular, we adapt the optimal algorithm proposed originally for block-frequency profiling~\citep{nahapetian73,knuth73,ball94}.
-Because we define work as a linear equation on the block-frequency counters, it is possible to embed its computation into the execution of the program.
-A naive instrumentation would consist basically of having a global counter that starts with the interception value, $\varepsilon$, and each basic block increments its own cost into the global counter.
-Although this instrumentation is easily implemented, it imposes a significant overhead on the instrumented program.
-However, it is possible to insert fewer probes by carefully placing the probes in a way that is possible to reconstruct the complete profiling information~\citep{knuth73,ball94}.
+In this section, we describe how to measure work online. Total work is just the sum of the costs of all executed basic blocks, so it is
+possible to embed its computation into the execution of the program. A naive instrumentation would use a global counter initialized at the
+interception value, $\varepsilon$. Each basic block would then have only to increment this counter with the total cost of its instructions
+every time it is executed. Although easy to implement, the naive approach introduces a significant overhead.
 
-We adapt the optimal block-frequency profiling in order to perform the work profiling efficiently.
-The proposed work profiling differs from the optimal block-frequency profiling as the latter occurs in two stages:
-\textit{(i.) Before execution:} The code is instrumented with counters for each probe.
-\textit{(ii.) After execution:} The information from the recorded probes is propagated in the CFGs of the program.
-In contrast, our work profiling has a single counter, and it only requires the instrumentation before execution, without any post-processing of the recorded profiling.
+Previous work on basic block profiling has proposed optimal ways for instrumenting the code with as little overhead as possible without
+losing any profiling information~\citep{knuth73,ball94}. The underlying idea is to instrument only a subset of the basic blocks of a
+function and use the function's CFG to calculate execution counts for the rest. We can use the same idea for our purposes. Work profiling
+and basic block profiling are similar, both need to count how many times each basic block was executed. The only important difference is
+that we do not have to store the execution count of each basic block, only the total work.
 
-The instrumented code is assumed to be generated before optimising the code.
-This assumption is based on three key points:
-\textit{(i.)} it guarantees that the work metric is independent of optimisation, i.e., the same input is always mapped to the same amount of work, regardless of the optimisation;
-\textit{(ii.)} it simplifies the code generator;
-\textit{(iii.)} it leverages from the optimisations to further improve the instrumentation code.
-Having a work metric that is independent of optimisation is the most important reason for which we instrument the code before optimisations.
-
-The optimal placement of the probes works by first computing the maximum spanning
-tree based on a weighing that assigns a value to each edge in the CFG.
-These weights can be obtained either by empirical measurements or heuristic estimations,
-and their goal is to avoid probing in frequently executed edges.
-Once we have the maximum spanning tree, probes are placed on every edge not in the spanning tree.
-Figure~\ref{fig:cfg-example} shows an example of a CFG with a maximum spanning tree.
-%represented by the black edges, while the edges highlighted in red represent the placement of the probes.
+Graph theory shows that we can calculate precisely the frequencies of all edges between basic blocks if we choose a spanning tree of the
+function CFG and instrument the edges not in the tree. To make the instrumentation probe placement optimal, we choose the maximum spanning
+tree, with the edge weights representing edge frequency estimates. As many high frequency edges as possible will be in the maximum spanning
+tree and they will not be instrumented.
 
@@
 }
-  \caption{Example of a CFG with its minimum spanning tree in black and the
-   basic blocks highlighted in red represent the instrumented basic blocks with
-   the placement of the probes.}
+  \caption{Example of a CFG with its maximum spanning tree in black. Basic blocks and edges highlighted in red are instrumented.
+    Instrumenting them is enough for calculating the total work performed by the whole CFG.}
   \label{fig:cfg-example}
@@
 
-In contrast to the naive instrumentation where each basic block records only its
-own amount of work, with the optimal profiling, the instrumented basic blocks need
-to record an aggregated value of work that also takes other basic blocks into
-account.
-Some probes speculatively assume a path of execution,
-while other probes correct when these assumptions are wrong
-(see for example $w(P_0)$ and $w(P_1)$ in Figure~\ref{fig:cfg-example}).
-
-We compute the aggregated value of work for each probe in two steps:
-\textit{(i.)} we propagate information about the probes through the edges of the CFG,
-as symbolic expressions, and then \textit{(ii.)} we use this edge flows
-to compose the aggregated value of the probes.
+In contrast to the naive instrumentation where each basic block records only its own amount of work, with the optimal profiling the work
+counter increments needs to take other basic blocks into account. Figure~\ref{fig:cfg-example} shows an example CFG, where the instrumented
+blocks are marked with red. In this example, executing block $B_3$ implies also executing $B_1$ and not executing $B_4$. We compute the
+aggregated value of work for each probe in two steps: \textit{(i.)} we propagate information about the probes through the edges of the CFG,
+as symbolic expressions, and then \textit{(ii.)} we use these edge flows to compose the aggregated value of the probes.
 
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
   \centering

### 2017-09-14 infra.tex
arrrrrrrrrrrrrrrrrrrrrrrrrrrrrrrgh
minor fix

@@
 
-\begin{figure}[htb]
+\begin{figure}[t]
     \centering
     \includegraphics[width=\linewidth]{figs/infra-diagram}
-    \caption{Overview of the online \itercomp mechanism.}
+    \caption{Overview of the online \itercomp mechanism. \red{zw: mark the boundary of online and offline.}}
     \label{fig:infra-diagram}
@@
 
-This work focuses on the highlighted components of Figure~\ref{fig:infra-diagram}. The first one, \textit{Work Instrumentation} allows
-measuring the amount of work performed. Conceptually, we use the cost model of Section~\ref{sec:metric} to estimate a cost for each basic
-block of the IR and then add instrumentation to increase a global work counter by that cost every time a basic block is executed. The
-measured work should depend only on the program and the input, not the applied optimizations, so we instrument the unoptimized IR before
-any optimizations modify it. We present a detailed description of work instrumentation and how to keep it lightweight in
-Section~\ref{chap:instr}.
+This work focuses on the highlighted components of Figure~\ref{fig:infra-diagram}. The first one, \textit{Work Instrumentation} \blue{is
+performed once at offline compile time}. It allows measuring the amount of work performed. Conceptually, we use the cost model of
+Section~\ref{sec:metric} to estimate a cost for each basic block of the IR and then add instrumentation to increase a global work counter
+by that cost every time a basic block is executed. The measured work should depend only on the program and the input, not the applied
+optimizations, so we instrument the unoptimized IR before any optimizations modify it. We present a detailed description of work
+instrumentation and how to keep it lightweight in Section~\ref{chap:instr}.
 
@@
 
-\FIXME{WHAT IS THE OPTIMIZATION EVALUATOR? WHY IS IT HIGHLIGHTED? DO WE HAVE ANYTHING TO SAY ABOUT IT?}
+%\FIXME{WHAT IS THE OPTIMIZATION EVALUATOR? WHY IS IT HIGHLIGHTED? DO WE HAVE ANYTHING TO SAY ABOUT IT?}
 
@@
 Most devices have distinct idle and peak usage periods. For example, mobile devices are used intensely during the day and not at all while
-the user is sleeping and the battery is charging~\citep{mpeis16}. Many authors have identified similar idle and peak usage periods in the
-context of data centers~\citep{armbrust10,chen12b}. The proposed \itercomp infrastructure is very well suited to such scenarios. In
+the user is sleeping and the battery is charging~\citep{mpeis16}. Many researchers have identified similar idle and peak usage periods in
+the context of data centers~\citep{armbrust10,chen12b}. The proposed \itercomp infrastructure is very well suited to such scenarios. In
 particular, we can use periods of idleness or underutilization to select the next few optimization sequences to try and compile the

### 2017-09-14 prof.tex
minor fix

@@
 In this section, we describe how to measure work online. Total work is just the sum of the costs of all executed basic blocks, so it is
-possible to embed its computation into the execution of the program. A naive instrumentation would use a global counter initialized at the
-interception value, $\varepsilon$. Each basic block would then have only to increment this counter with the total cost of its instructions
-every time it is executed. Although easy to implement, the naive approach introduces a significant overhead.
+possible to embed its computation into the execution of the program. A na\"{\i}ve instrumentation would use a global counter initialized at
+the interception value, $\varepsilon$. Each basic block would then have only to increment this counter with the total cost of its
+instructions every time it is executed. Although easy to implement, the na\"{\i}ve approach introduces a significant overhead.
 
@@
 losing any profiling information~\citep{knuth73,ball94}. The underlying idea is to instrument only a subset of the basic blocks of a
-function and use the function's CFG to calculate execution counts for the rest. We can use the same idea for our purposes. Work profiling
-and basic block profiling are similar, both need to count how many times each basic block was executed. The only important difference is
-that we do not have to store the execution count of each basic block, only the total work.
+function and use the function's control flow graph (CFG) to calculate execution counts for the rest. We can use the same idea for our
+purposes. Work profiling and basic block profiling are similar, both need to count how many times each basic block was executed. The only
+important difference is that we do not have to store the execution count of each basic block, only the total work.
 
@@
 function CFG and instrument the edges not in the tree. To make the instrumentation probe placement optimal, we choose the maximum spanning
-tree, with the edge weights representing edge frequency estimates. As many high frequency edges as possible will be in the maximum spanning
-tree and they will not be instrumented.
+tree, with the edge weights representing edge frequency estimates. By doing so, we leave as many high frequency edges as possible in the
+maximum spanning tree and they will not be instrumented, allowing us to minimize the instrumentation overhead.
 
@@
 \paragraph{Populating the edge flows:}
-Intuitively, if all the edge flows are known for the complement of a spanning tree
-then at any leaf of the spanning tree there is only one unknown edge flow.
-This unknown edge flow can be calculated by Kirchhoff's first law.
-This process repeats as a bottom-up propagation until all the unknown edge flows
-have been calculated.
-This algorithm can be formally defined as a post-order traversal on the spanning
-tree.
-Let $G$ be the CFG.
-This algorithm first initializes the edge flows $D_{(u,v)}$, for each edge
-$(u,v)$ in the CFG, as follows:
+Intuitively, if all the edge flows are known for the complement of a spanning tree then at any leaf of the spanning tree there is only one
+unknown edge flow. This unknown edge flow can be calculated by Kirchhoff's first law\FIXME{\cite{}}. This process repeats as a bottom-up
+propagation until all the unknown edge flows have been calculated. This algorithm can be formally defined as a post-order traversal on the
+spanning tree. Let $G$ be the CFG. This algorithm first initializes the edge flows $D_{(u,v)}$, for each edge $(u,v)$ in the CFG, as
+follows:
 \[
@@
 
-In this section we go one step even further.
-In some cases, even the proposed relaxation strategy can still be conservative.
-This conservatism can be overly restrictive in some cases, resulting in a
-negligible overhead reduction but also causing just a negligible dynamic error
-to the work profiling.
-For these cases, we propose an adapted version of the relaxation algorithm that
-operates on the whole program.
-Traditionally, compiler optimizations are performed on the function-level, or at
-best on a per module basis.
-Whole program optimization (WPO) means that the compiler considers all
-compilation units of the program and optimizes them using the combined knowledge
-of how they are used together.
+In this section we go one step further. In some cases, even the proposed relaxation strategy can still be conservative. This conservatism
+can be overly restrictive in some cases, resulting in a negligible overhead reduction but also causing just a negligible dynamic error to
+the work profiling. For these cases, we propose an adapted version of the relaxation algorithm that operates on the whole program.
+Traditionally, compiler optimizations are performed on the function-level, or at best on a per module basis. Whole program optimization
+(WPO) means that the compiler considers all compilation units of the program and optimizes them using the combined knowledge of how they
+are used together.
 

### 2017-09-14 prof.tex
prof again

@@
 counter increments needs to take other basic blocks into account. Figure~\ref{fig:cfg-example} shows an example CFG, where the instrumented
-blocks are marked with red. In this example, executing block $B_3$ implies also executing $B_1$ and not executing $B_4$. We compute the
-aggregated value of work for each probe in two steps: \textit{(i.)} we propagate information about the probes through the edges of the CFG,
-as symbolic expressions, and then \textit{(ii.)} we use these edge flows to compose the aggregated value of the probes.
-
-The symbolic expressions in the edges can then be used to compose the aggregated
-value of the probes.
-These symbolic expressions describe the probes that are in the same path of the
-edge (positive terms) and those that are in complementary paths (negative
-terms).
-Therefore, the positive terms in the summed symbolic expression of a
-basic block indicate that the amount of work of this basic block will be
-incremented in the probes represented by these positive terms.
-Similarly, the negative terms indicate that the amount of work of this basic block
-will be decremented in the probes represented by these negative terms.
-For example, because the symbolic expression for the basic block $B_8$ is $P_0 - P_1$,
-the amount of work of $B_8$, denoted by $w(B_8)$, is incremented in probe $P_0$
-and decremented in $P_1$.
+blocks are marked with red. In this example, executing block $B_5$ implies not executing $B_6$. To compute the amount of work associated
+with each instrumentation probe, we first need to determine how reaching a probe relates with executing or not a basic block. Starting from
+instrumented edges, we build symbolic expressions for each edge of their frequency counts as a linear function of probe counts, as seen in
+Figure~\ref{fig:cfg-example}. If a symbolic expression for an edge and the corresponding block contains a positive term for a probe, this
+means that the block is executed when we reach the probe, so the block's cost should be added at that probe. Conversely, if the expression
+contains a negative term for a probe, then the block is executed when we do not reach the probe, so the block's cost should be subtracted.
+In our example CFG, the symbolic expression for the basic block $B_8$ is $P_0 - P_1$, so the amount of work of $B_8$, denoted by $w(B_8)$,
+is incremented in probe $P_0$ and decremented in $P_1$.
 
@@
 
-Although the optimal instrumentation significantly reduces the profiling
-overhead when compared to the naive instrumentation, from an average overhead
-of 79\% to 13\%, in some critical cases, even the optimal instrumentation can
-result in overheads of up to 60\% (see benchmark \texttt{adpcm\_d} in Figure~\ref{fig:overhead-O3}).
-In order to further reduce the overhead in these critical cases, we propose a
-relaxation strategy that offers a trade-off between accuracy and overhead.
-While the optimal instrumentation places probes in edges that are
-less likely to be executed, the relaxation removes probes that are
-more likely to be executed but add little to the final work metric.
+Optimal instrumentation significantly reduces the profiling overhead when compared to naive instrumentation, from an average overhead of
+79\% to 13\%. In some extreme cases though optimal instrumentation can result in overheads of up to 60\% (see benchmark \texttt{adpcm\_d}
+in Figure~\ref{fig:overhead-O3}). To further reduce the overhead in these cases, we propose a relaxation strategy that offers a trade-off
+between accuracy and overhead. While in traditional basic block profiling the aim is to accurately measure the frequency counts of every
+single block, in our case we only care about the total amount of work. Our insight was that if the work associated with a probe is small,
+then ignoring the probe will have little effect on the total work estimation. By removing such probes, especially if they are frequently
+executed, we can reduce the instrumentation overhead dramatically.
 
-% \begin{figure}[h]
-%   \centering
-%   \includegraphics[scale=0.85]{figs/relax-instr-diagram.pdf}
-%   \caption{Overview of the work instrumentation algorithm, including the relaxation technique.}
-%   \label{fig:relax-instr-diagram}
-% \end{figure}
+Our relaxation strategy starts from the probes and their weights as decided by the optimal algorithm. Removing low weight probes is easy,
+but we need a way to guarantee that the error in our work estimation will not grow too big. Our conservative approach, decides on an upper
+bound for the error and applies it on every recursively on every loop of the CFG. Specifically, our algorithm extracts all subgraphs
+representing loops and transforms them into DAGs (Directed Acyclic Graphs) by removing the backedge and moving inner loops into separate
+DAGs. \FIXME{PP: NOT DONE....}
 
-%Figure~\ref{fig:relax-instr-diagram} shows an overview of the relaxed instrumentation algorithm.
-%The highlighted step is introduced by the relaxed instrumentation on top of the previously defined optimal profiling.
-The relaxation strategy performs a post processing on the resulting instrumentation of the optimal algorithm.
-This post processing identifies probes that add little to the work metric, and removes their instrumentation.
 In order to guarantee an upper bound for the dynamic error in the profiling measurement,

### 2017-09-14 work.tex
Work

@@
 
-In this section we define work efficiency, a metric for comparing different optimized versions of a program when executing with different
-inputs. We define it as the ratio between the amount of \textit{work}, $\Delta W$, performed during a period of time, $\Delta t$.
+    In this section we define work efficiency, a metric for comparing different optimized versions of a program when executing with
+    different inputs. We define it as the ratio between the amount of \textit{work}, $\Delta W$, performed during a period of time, $\Delta
+    t$.
 
@@
 
-The main challenge is to precisely define what represents \textit{work}.
-We define the work done by a program on an input to be the time taken by the unoptimized program to process that input.
-We estimate the runtime of each instruction type via a linear regression and then define the work done by a program as the sum of the
+    The main challenge is to precisely define what represents \textit{work}. We define the work done by a program on an input to be the
+    time taken by the unoptimized program to process that input. We estimate the runtime of each instruction type and then work is the sum
+    of the runtime for each instruction type multiplied by its dynamic count.
 
-We adopt a work metric that takes into account the
-computational cost within each basic block.
-The computational cost of a basic block is given by weighting the instructions
-that it contains.
+    \[ \Delta W = \sum_i w_i.n_i \]
 
-To this end, we model the computational work $\Delta W$ as a linear equation based on block frequency information and a cost model of the instruction set.
-Formally,
-\[
-\Delta W = \varepsilon + \sum_{B} w(B)f(B)
-\]
-where $f(B)$ represents the frequency of basic block $B$ and $w(B)$ represents the computational work of executing $B$.
-We define the work of a basic block $B$ as the sum of the cost of its instructions, i.e.,
-\[
-w(B) = \sum_{i} w_i N_B(i)
-\]
-where $w_i$ is the cost of instruction $i$ and $N_B(i)$ is the number of occurrences of instruction $i$ in basic block $B$.
+    Where $i$ is the instruction type, $w_i$ is the estimated runtime for instruction type $i$, and $n_i$ is the number of observed
+    executions of instruction type $i$.
 
-In this simplified model, we consider that $w_i$ is constant across all programs and executions, varying only between target architectures.
-On the other hand, $N_B(i)$ is a program-dependent static value which is known at compile-time and $f(B)$ is a dynamic value known only at run-time,
-since $f(B)$ is both program and execution dependent as the execution frequency of a basic block can change when executing with different inputs.
+    Similarly to previous work~\citep{giusto01,powell09,brandolese11}, the coefficients, $w_i$, are found by linear regression. We create
+    one data point for the regressor for a set of program and input pairs. The programs are given in Table~\ref{tab:kdatasets:training},
+    and each has 1,000 inputs. The response variable is the runtime of the unoptimized program as it processes the input. To ensure
+    statistically sound results, the runtime is measured by repeated execution until the 99\% confidence interval is no larger than 1\% of
+    the mean. The explanatory variables are the dynamic instruction counts for each program and input. They are recorded by a separate,
+    instrumented run of the program on the input.
 
-\subsection{Estimating a Cost Model of the Instructions}
-
-Similarly to previous work~\citep{giusto01,powell09,brandolese11}, we derive the cost model of the instruction set by modeling the problem
-as a multi-variable linear regression. The \textit{regression coefficients} are the costs of the instructions and the \textit{regressors}
-are computed as $\sum_B N_B(i)f(B)$ for each instruction.
-\begin{equation}\label{eq:linear-work-expression}
-\Delta W = \varepsilon + \sum_{i} \left(w_i \sum_{B} N_B(i)f(B)\right)
-\end{equation}
-
-We use a set of training programs to construct the regression model. Our training program set is given in
-Table~\ref{tab:kdatasets:training}. To determine the model coefficients, we profile each training benchmark with 1000 input datasets
-provided by the benchmark suite, and we measure the wall-clock time. It is to note that we turn off all compiler optimisations during this
-training stage, to ensure that the amount of work we measured only depends on the program input but not compiler optimization. This aspect
-is crucial for the work-based performance metric to allow a consistent comparison between the various optimization sequences when executing
-with distinct inputs.
-
-%it allows to use the estimated cost model for computing a work metric which is independent of optimizations, as we explained in Section~\ref{}.
-
-Our cost model has a total of 52 LLVM instructions\footnote{We do not model all LLVM instructions because some instructions are more common in optimized programs, such as the vector instructions.}.
-Every program, in the set of training benchmarks, was compiled twice: once including the necessary instrumentation for the block frequency profiling, which is required for deriving the linear expression defined in Equation~\ref{eq:linear-work-expression};
-and a standard compilation without any optimization or instrumentation.
-For each input in the training dataset, the benchmarks were executed once with the instrumented version for collecting the block frequency profiling, and multiple times with the standard compilation just for measuring the wall-clock execution time, until the confidence interval was no larger than 1\% for a 99\% confidence.
-After collecting these measurements, this data can be used to estimate the unknown parameters in the linear model.
-Because we fit the linear model based on the wall-clock execution time, the derived cost model can be interpreted as an estimate of the execution time when the program is compiled without optimizations.
+    Figure~\ref{fig:cost-model} shows the correlation of the model with runtime on instances of test benchmarks. The correlation
+    coefficient is \FIXME{XXX}, and the mean absolute percentage error (MAPE) is \FIXME{XXX}. Figure~\ref{fig:motivation-work-efficiency}
+    shows a similar plot to Figure~\FIXME{\ref{fig:???}}, of work efficiency versus speedup for 500 different binaries of \texttt{susan\_c}.
+Here the new metric correlates much better than either performance or IPC.
 
@@
         \includegraphics[width=0.6\linewidth]{figs/cost-model.pdf}
-    \caption{Linear model fitted from empirical data. The mean absolute error (MAE) for the fitted curve is seven milliseconds.}
+        \caption{
+            Linear model for work metric fitted from empirical data.
             \red{zw: bigger font size.}
+        }
         \label{fig:cost-model}
     \end{figure}
-
-Figure~\ref{fig:cost-model} compare the work metric with the corresponding execution time for some instances of the test benchmarks.
-Notice how the fitted model has a higher relative error for the instances with very short execution time, namely those that run for less than one tenth of a second.
-The mean absolute error (MAE) for the fitted curve is seven milliseconds.
 
@@
     \end{figure}
-
-\FIXME{Rodrigo: I think we could show how the work efficiency metric performs compared with speedup, the same way we showed the other metrics does not work well.}
-
-\subsection{Comparison with Instructions Per Cycle} \label{sec:ipc-vs-work-metric}
-
-\FIXME{Rodrigo: Perhaps we could remove this section, since we have already discussed the IPC in the Motivation.}
-
-The IPC metric has been widely used for studying performance benefits of hardware optimizations.
-Although previous work has suggested the use of IPC for guiding {\itercomp}, in this section we argue in favour of the WP metric over IPC.
-
-The IPC metric differs from the proposed WP metric in a key aspect:
-the IPC metric is computed solely based on the final optimized program.
-When executing different optimized versions of a program with the same input, both the number of instructions and the number of clock cycles can change.
-For this reason, higher IPC does not necessarily translate to shorter execution time (or even fewer clock cycles).
-
-% We can illustrate this fact with a very small example as shown in Table~\ref{tab:ipc-example}.
-% This example shows two versions of a program, namely P1 and P2, and their respective measurements related to IPC.
-% Although version P1 has twice the IPC of P2, P1 is one cycle slower than P2.
-% As this example illustrates, the IPC metric can be misleading when compared different versions of the same program.
-% This problem is only intensified when both the optimization sequence and the input changes.
-% In Chapter~\ref{chap:eval} we show empirical evidence for this argument.
-%
-% \begin{table}[h]
-% \centering
-% \begin{tabular}{|c|c|c|c|}
-% \hline
-%                        & P1 & P2  \\
-% \hline
-% Number of Instructions & 5  & 2   \\
-% Number of Cycles       & 5  & 4   \\
-% IPC                    & 1  & 0.5 \\
-% \hline
-% \end{tabular}
-% \caption{Example that illustrates that a higher IPC does not necessarily translate to shorter execution time.}
-% \label{tab:ipc-example}
-% \end{table}
-
-On the other hand, WP computes the amount of work based on the unoptimized version of the program, which means that it always measures the same amount of work for the same input, regardless of the optimizations applied on the program.
-This is a key aspect that enables the use of the WP metric for guiding {\itercomp}, because a higher WP, which represents work per unit time, naturally translate to shorter execution time.
-Moreover, previous work has also presented other arguments against the use of IPC in similar use-case scenarios. %, as discussed in Chapter~\ref{chap:related}.

### 2017-09-14 work.tex
Small fixes

@@
 
-    Where $i$ is the instruction type, $w_i$ is the estimated runtime for instruction type $i$, and $n_i$ is the number of observed
+    Where $i$ is the instruction type; $w_i$ is the estimated runtime for instruction type $i$; and $n_i$ is the number of observed
     executions of instruction type $i$.
@@
     Similarly to previous work~\citep{giusto01,powell09,brandolese11}, the coefficients, $w_i$, are found by linear regression. We create
-    one data point for the regressor for a set of program and input pairs. The programs are given in Table~\ref{tab:kdatasets:training},
+    one regressor data point for each of a set of program and input pairs. The programs are given in Table~\ref{tab:kdatasets:training},
     and each has 1,000 inputs. The response variable is the runtime of the unoptimized program as it processes the input. To ensure
     statistically sound results, the runtime is measured by repeated execution until the 99\% confidence interval is no larger than 1\% of
-    the mean. The explanatory variables are the dynamic instruction counts for each program and input. They are recorded by a separate,
+    the mean. The explanatory variables are the dynamic instruction counts for each program and input pair. They are recorded by a separate,
     instrumented run of the program on the input.

### 2017-09-14 results.tex
List margin
fix minor layout things in result figures; add WPO speedup results. other text restructuring

@@
 
-For comparison, we use four configurations\footnote{A configuration using the WPO relaxation was not used due to the time necessary for executing the experiments.}.
+For comparison, we use five configurations.
 In all configurations, the same optimization sequence is used for multiple inputs, using a dynamic input-window size, as explained in Section~\ref{sec:oic-infra}.
@@
 %When selecting the best optimization sequence, they are ranked by the average of their performance measurement, using the WP metric, except for the Oracle-RM which uses the real speedup over \texttt{-O0}.
-\begin{itemize}
+\begin{itemize}[leftmargin=3mm]
 \item \textbf{Oracle-RM} executes the program twice for each input, measuring the real speedup for each optimization sequence, and then uses the real speedup over {\flagstype -O0} for guiding online {\itercomp}.
@@
 
-Figure~\ref{fig:window-size} shows the average input-window size for each benchmark and configuration.
-It shows that we can have a statistically sound measurement of the performance metric using just a small number of inputs.
+%Figure~\ref{fig:window-size} shows the average input-window size for each benchmark and configuration.
+%It shows that we can have a statistically sound measurement of the performance metric using just a small number of inputs.
 
-\begin{figure*}[t]
-    \centering
-    \includegraphics[width=\textwidth]{figs/window-size.pdf}
-    \caption{Average input-window sizes observed during the online {\itercomp}.}
-    \label{fig:window-size}
-\end{figure*}
+%\begin{figure*}[t]
+%    \centering
+%    \includegraphics[width=\textwidth]{figs/window-size.pdf}
+%    \caption{Average input-window sizes observed during the online {\itercomp}.}
+%    \label{fig:window-size}
+%\end{figure*}
 
@@
 
+\begin{figure*}[t]
+    \centering
+    \includegraphics[width=\textwidth]{figs/speedups.pdf}
+    \caption{Speedups obtained from the final optimization sequence selected by the online {\itercomp}.
+	         The speedups reported for each benchmark represents the average speedup across their complete 1000 input datasets.}
+    \label{fig:speedups}
+\end{figure*}
+
 In order to evaluate the quality of the final optimization sequences selected by the {\itercomp} search, we compare their speedup by measuring wall-clock time of the benchmarks when compiled with the standard {\flagstype -O3} optimization.
@@
 %Finally, Figure~\ref{fig:speedups} also presents the results of {\itercomp} guided by IPC, which is futher discussed in Figure~\ref{fig:ipc-vs-work}.
-
-\begin{figure*}[t]
-    \centering
-    \includegraphics[width=\textwidth]{figs/speedups.pdf}
-    \caption{Speedups obtained from the final optimization sequence selected by the online {\itercomp}.
-	         The speedups reported for each benchmark represents the average speedup across their complete 1000 input datasets.}
-    \label{fig:speedups}
-\end{figure*}
 

### 2017-09-14 prof.tex
done with prof. probably. perhaps. we will see

@@
 
-Our relaxation strategy starts from the probes and their weights as decided by the optimal algorithm. Removing low weight probes is easy,
-but we need a way to guarantee that the error in our work estimation will not grow too big. Our conservative approach, decides on an upper
-bound for the error and applies it on every recursively on every loop of the CFG. Specifically, our algorithm extracts all subgraphs
+Our relaxation strategy operates on selection of probes and weights as decided by the optimal algorithm. Removing low weight probes is
+easy, but we need a way to guarantee that the error in our work estimation will not grow too big. Our conservative approach, decides on an
+upper bound for the error and applies it recursively on every loop of the CFG. Specifically, our algorithm extracts all subgraphs
 representing loops and transforms them into DAGs (Directed Acyclic Graphs) by removing the backedge and moving inner loops into separate
-DAGs. \FIXME{PP: NOT DONE....}
+DAGs. An example of this is shown in Figure~\ref{fig:cfg-relax-example}, where the CFG is partitioned into two DAGs.
 
-In order to guarantee an upper bound for the dynamic error in the profiling measurement,
-the relaxation algorithm applies a constrained post-processing on a per DAG (directed acyclic graph) basis.
-By constraining the relaxation within each DAG by a maximum allowed static error,
-it guarantees that the overall relaxation will also be constrained by the same bound.
-
-The relaxation starts by extracting DAGs from the CFG.
-First, the algorithm extracts all the subgraphs that represent a loop or the outer most region of the function.
-Afterwards, these subgraphs are transformed into DAGs by ignoring the backedge
-and also by considering that any loop within the subgraph is never executed,
-i.e., only the headers of the inner loops are actually included into the DAG.
-Figure~\ref{fig:cfg-relax-example} shows a CFG partitioned into two DAGs (consider
-only basic blocks and edges completely inside the yellow and green boundaries).
+%The relaxation starts by extracting DAGs from the CFG.
+%First, the algorithm extracts all the subgraphs that represent a loop or the outer most region of the function.
+%Afterwards, these subgraphs are transformed into DAGs by ignoring the backedge
+%and also by considering that any loop within the subgraph is never executed,
+%i.e., only the headers of the inner loops are actually included into the DAG.
+%Figure~\ref{fig:cfg-relax-example} shows a CFG partitioned into two DAGs (consider
+%only basic blocks and edges completely inside the yellow and green boundaries).
 
@@
 
-For every DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$,
-we select a subset of the probes to be removed, subject to
-the maximum allowed percentage error, $M$.
-This strategy can be modeled as a 0-1 Knapsack problem:
+For every DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we calculate $m$, the minimum amount of work that can be performed in the
+DAG. Given this, the worst case relative error caused by removing a probe $P_i$ is $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$. Reducing the
+instrumentation overhead as much as possible while keeping the total worst case error below a threshold can be modeled as a 0-1 Knapsack
+problem:
 \begin{gather*}
@@
 \end{gather*}
-where $f(P_i)$ is the execution frequency of probe $P_i$, $x_i$ denotes the
-probes selected for removal, and $\varepsilon(P_i)$ is the percentage error of
-removing probe $P_i$ relative to the minimum work value possible to compute in
-the DAG, i.e., if $m$ is the minimum amount of work possible to be computed when
-executing the DAG, then $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$.
-Because the percentage error is computed based on the path with the minimum
-amount of work, $\varepsilon(P_i)$ represents the maximum error possible that
-would be incurred when removing probe $P_i$.
-Furthermore, by constraining the percentage error of every DAG below a given
-threshold, we guarantee that the final error of the relaxation will always be
-bounded by the threshold.
+where $f(P_i)$ is the execution frequency of probe $P_i$ and $x_i$ denotes the probes selected for removal. By constraining the worst case
+relative error in every single DAG of the CFG, we guarantee that the final error introduced by relaxation will always be below the chosen
+threshold.
 % Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold, as demonstrated by Proposition~\ref{prop:relax-bound}.
@@
 
-For our experiments, we implemented two solvers for the 0-1 Knapsack problem:
-the optimal brute-force solver and
-the greedy heuristic based on sorting the items~\citep{dantzig57}.
-We use the brute-force solver for DAGs with a small number of probes and the
-greedy heuristic when the number of probes is greater than a threshold.
-Some of the benchmarks have DAGs with several hundreds of probes, which could
-result in a long compilation time.
+For our experiments, we implemented two solvers for the 0-1 Knapsack problem: an optimal but potentially slow brute-force solver and a
+greedy heuristic based on sorting the items~\citep{dantzig57}. We use the brute-force solver for DAGs with a small number of probes, the
+greedy heuristic otherwise.
 
@@
 
-In this section we go one step further. In some cases, even the proposed relaxation strategy can still be conservative. This conservatism
-can be overly restrictive in some cases, resulting in a negligible overhead reduction but also causing just a negligible dynamic error to
-the work profiling. For these cases, we propose an adapted version of the relaxation algorithm that operates on the whole program.
-Traditionally, compiler optimizations are performed on the function-level, or at best on a per module basis. Whole program optimization
-(WPO) means that the compiler considers all compilation units of the program and optimizes them using the combined knowledge of how they
-are used together.
+The proposed relaxation strategy provides hard guarantees that error will remain below the chosen threshold but in many cases these
+guarantees are too conservative. Our worst case scenario approach assumes that each loop will always perform the minimum amount of work
+possible which might be orders of magnitude less than the work actually performed. In such cases, we remove too few probes and have little
+effect on the instrumentation overhead, while the actual relative error in our work estimation is negligible. To avoid this problem, we
+propose an adapted version of the relaxation algorithm that operates on the whole program.
 
-The \textit{whole program relaxation} works by using block-frequency profiling
-from previous executions.
-By having this profiling information, the whole program relaxation is able to
-compute the error of removing a given probe in terms of the whole program's execution,
-and then select a subset of all the probes to be removed.
+\textit{Whole program relaxation} works by using block-frequency profiling from previous executions. With this information, we are able to
+compute the error of removing a given probe in terms of the whole program's execution, and then select a subset of all the probes to be
+removed.
 
-For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we model the
-whole program relaxation as the following 0-1 Knapsack problem:
+For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we model the whole program relaxation as the following 0-1 Knapsack
+problem:
 \begin{gather*}
@@
 \end{gather*}
-where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$,
-$x_i$ denotes the probes selected for removal, $M$ is the error threshold, and
-$\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to
-the profiled global work, i.e.,
-if $\Delta W$ is the work value for the whole program's execution, computed from
-the basic block frequencies profiled from previous executions, the error for a
-given probe $P_i$ is
-\[
-\varepsilon(P_i) = \frac{\omega(P_i)f(P_i)}{\Delta W}.
-\]
+where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$, $x_i$ denotes the probes selected for removal, $M$ is the
+error threshold, and $\varepsilon(P_i)$ is the error of removing probe $P_i$ relative to the total work. 
+%if $\Delta W$ is the work
+%value for the whole program's execution, computed from the basic block frequencies profiled from previous executions, the error for a
+%given probe $P_i$ is
+%\[
+%\varepsilon(P_i) = \frac{\omega(P_i)f(P_i)}{\Delta W}.
+%\]
 
-Contrary to the per DAG relaxation, the whole program relaxation is not
-guaranteed to be bounded by the error threshold $M$,
-as it depends on the representativity % representativeness
-of the profiling information provided to the whole program relaxation.
+Contrary to the per DAG relaxation, the error introduced by whole program relaxation is not guaranteed to be bounded by the error threshold
+$M$. Still, we expect it to remain close to the threshold as long as the program's behavior does not deviate significantly from the
+behavior captured in profile we used.

### 2017-09-14 results.tex
updates setup with 1-column table

@@
 
-\subsection{The Set of Optimization Sequences}
-
-For the purpose of evaluating the use of the WP metric with {\itercomp}, we collected in advance a fixed set of optimization sequences.
-The reason for using this fixed set, as explained in Section~\ref{sec:oic-infra}, is because this paper is focused on other components of the infrastructure for performing {\itercomp} and we assume that a good generator of optimization sequences will be used in a real online scenario.
-This set contains 500 optimization sequences collected in a random search using the training benchmarks.
-These optimization sequences contain an average of 40 individual optimization passes, including repetitions, with a maximum of 119 optimization passes, but it also contains optimization sequences which consist of a single flag, such as the default optimizations {\flagstype -O1}, {\flagstype -O2}, {\flagstype -O3}, {\flagstype -Os}, and {\flagstype -Oz}.
-
-  % \begin{minipage}{0.9\textwidth}
-  %    \vspace{1em}
-  %    \noindent\textbf{Example of a short optimization sequence:}\vspace{-1ex}
-  %    \justify{\flagstype -mem2reg -simplifycfg -constprop -dce}
-  % \end{minipage}
-  %
-  % \begin{minipage}{0.9\textwidth}
-  %    \vspace{1em}
-  %    \noindent\textbf{Example of a long optimization sequence:}\vspace{-1ex}
-  %    \justify{\flagstype -globalopt -reassociate -instcombine -loop-rotate -block-freq -deadargelim -early-cse -sroa -argpromotion -sccp -tbaa -barrier -constmerge \mbox{-loop-vectorize} -domtree -basicaa -memdep -basiccg -memcpyopt \mbox{-constprop} -adce -globaldce -mem2reg -constmerge \mbox{-globaldce} -constprop -instsimplify -dse -dce -simplifycfg -loop-unroll -reassociate -constprop \mbox{-globaldce} -instsimplify -adce -constmerge -bb-vectorize -dce -mergefunc -simplifycfg -dse -loop-unroll -globaldce}
-  % \end{minipage}
-  %
-  % \begin{minipage}{0.9\textwidth}
-  %    \vspace{1em}
-  %    \noindent\textbf{Example of an optimization sequence which includes {-O3}:}\vspace{-1ex}
-  %    \justify{\flagstype -O3 -adce -globaldce -simplifycfg -memcpyopt -reassociate -mergefunc \mbox{-dce} -dse}
-  %    \vspace{2em}
-  % \end{minipage}
-
-%Repeating the same optimization pass can be beneficial and usually expected by other passes.
-%For example, the {\flagstype -loop-simplify} pass is used for transforming loops into a canonical form by inserting pre-header and exit basic blocks.
-%Although this pass inserts jumps due to redundant basic blocks, this canonical form can be favourable to other loop optimizations.
-%Because of the redundant basic blocks, this optimization pass expects that the {\flagstype -simplifycfg} will eventually be executor later on the optimization pipeline.
-%Another example of such inter-relation between transformations concerns the {\flagstype -licm} and {\flagstype -mem2reg} passes.
-%The {\flagstype -licm} pass is responsible for moving invariant code out from the loop body.
-%It usually creates new local variables, using memory access operations, for assisting with the code manipulation, which means that the executing the {\flagstype -mem2reg} pass afterwards would be useful as a cleanup pass for removing the extra memory accesses generated.
-%However, many of the analysis required for identifying loop invariant also benefit from the transformations performed by the {\flagstype -mem2reg} pass.
-%These examples illustrate the importance of repeating optimization passes.
-%Moreover, they illustrate the intricate relation amongst several transformations.
-
-All optimization sequences in the set of optimizations were generated completely at random, without using any knowledge of individual transformations.
-Each optimization sequence was generated in two steps: \textit{(1)} randomly selects the number of flags; \textit{(2)} randomly selects the flags, allowing repetitions.
-Afterwards, this randomly generated optimization sequence would be included in the set of optimization sequences only if it was able to improve the performance of a training benchmark, also selected at random, in respect of the {\flagstype -O3} default optimization.
-This process was repeated until we obtained all the 500 distinct optimization sequences.
-
 \subsection{Performance Evaluation}

### 2017-09-15 results.tex
changes to the result section; and other minor changes

@@
 
-We first compare static aspects of the instrumentation algorithms. The naive instrumentation instrument all basic blocks, by definition.
-First, we evaluate the optimal work profiling compared to the relaxation algorithm applied per DAG. Afterwards, we compare these profiling
-techniques with the whole program relaxation.
+In this section we evaluate the performance overhead incurred by the work efficiency profiling.
+We measure the wall-clock time of the benchmarks when compiled with the {\flagstype -O3} optimization.
+For each benchmark, we compute the average overhead over all its 1,000 inputs.
+Figure~\ref{fig:overhead-O3} shows the performance overhead imposed by the work
+profiling with each one of the instrumentation strategies, compared to their
+non-instrumented counterparts.
+The WPO relaxation was performed using profiling information for every specific
+input, which provides the perfect information to perform the WPO relaxation.
+This means that the experiments show its best performance enabled by having
+perfect profiling information.
 
-Figure~\ref{fig:num-probes} shows the percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.
-While Figure~\ref{fig:num-probes} compares the various instrumentation algorithms in respect of the naive instrumentation, Figure~\ref{fig:num-probes-improvement} shows the improvement of the relaxation algorithm  over the optimal instrumentation.
+%\begin{figure}[t]
+%    \centering
+%    \includegraphics[width=0.395\textwidth]{figs/adpcm_d-cfg-instr.pdf}
+%    \caption{CFG of the function that contains the hot loop of the {\flagstype adpcm\_d} benchmark.}
+%    \label{fig:adpcm_d-cfg-instr}
+%\end{figure}
 
-%\begin{figure*}[ht]
-%    \centering
-%    \includegraphics[width=\textwidth]{figs/num-probes.pdf}
-%    \caption{Percentage of instrumented basic blocks for the optimal and the relaxed instrumentation with different relaxation thresholds.}
-%    \label{fig:num-probes}
-%\end{figure*}
+While the optimal profiling has a maximum overhead of almost 60\%,
+the work profiling with a relaxation of 5\% threshold has an overhead of less
+than 16\% for all benchmarks,
+with the WPO relaxation reaching an average overhead of only 5\%,
+compared to an average of 12\% for the optimal profiling.
 
-\begin{figure*}[t]
+\begin{figure*}[t!]
     \centering
@@
 \end{figure*}
-
-\begin{figure*}[t]
+\begin{figure*}[t!]
     \centering
@@
 
-For the evaluation of the performance overhead that the instrumentation incur to the benchmarks, we measure the wall-clock time of the benchmarks when compiled with the default {\flagstype -O3} optimization.
-For each benchmark, we compute the average overhead over all its 1000 input dataset.
-When measuring the wall-clock time for each input, in order to reduce noise, we execute the same input until we have a statistically sound measurement, i.e.
-we execute until we have an interval no larger than 1\% with 99\% confidence.
-Figure~\ref{fig:overhead-O3} shows the performance overhead imposed by the work instrumentation on the benchmarks when compared to their non-instrumented counterparts.
-Figure~\ref{fig:overhead-improvement-O3} indicates the improvement of the relaxation algorithm over the optimal instrumentation, regarding the reduction in overhead.
+The most critical case, amongst the evaluated benchmarks, happens with the
+\texttt{adpcm\_d} program, with an overhead of about 59\% for the optimal
+instrumentation.
+With a threshold of only 5\%, the relaxation reduces
+the overhead by a factor of $4.5\times$ over the optimal instrumentation,
+and the WPO relaxation by a factor of $4.8\times$.
+This benchmark has a single hot function
+which consists mainly of a single hot loop with several branches inside it. %, as depicted by Figure~\ref{fig:adpcm_d-cfg-instr}.
+Most of the improvement achieved by both strategies comes from removing only two
+probes from the hot loop.
+The two relaxed probes were placed in branches with a high probability of being
+taken (above 55\% probability), but with a small contribution to the total amount
+of work measured in DAG for this loop (both probes with an static error of
+about 1.3\%).
+Because these probes were placed in basic blocks with a considerable execution
+frequency, their removal resulted in a significant reduction in performance
+overhead.
 
-The WPO relaxation was performed using profiling information for every specific input, which provides the perfect information to perform the WPO relaxation.
-This means that the experiments show its best performance enabled by having perfect profiling information.
-
-%\begin{figure*}[ht]
-%    \centering
-%    \includegraphics[width=\textwidth]{figs/overhead-improvement-O3.pdf}
-%    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.
-%             Notice that the average of the ratios does not equal the ratio of the averages.}
-%    \label{fig:overhead-improvement-O3}
-%\end{figure*}
-
-%Figure~\ref{fig:overhead-improvement-O3} shows that the relaxation algorithm, with 5\% threshold, is able to improve an average of 48\% over the optimal instrumentation,
-%and the WPO relaxation achieves an average of $2\times$ improvement over the optimal instrumentation.
-%While the optimal profiling has a maximum overhead of almost 60\%, the work profiling with a relaxation of 5\% threshold has an overhead of less than 20\% for all benchmarks,
-%with the WPO relaxation reaching an average overhead of only 5\%, compared to an average of 12\% for the optimal profiling.
+A few of the benchmark present abnormal regressions in their execution time.
+This increase in the profiling overhead is counter-intuitive because the
+relaxation algorithms only reduces the number of instrumented probes,
+without changing their placement.
+We suggest some reasons for this abnormal regressions.
+By analysing these cases, we noticed that several of the frequently executed
+probes could not be removed by the relaxation.
+Moreover, another important aspect to point out is that we also observed
+differences among the final CFGs after applying the \texttt{-O3} optimisation.
+Although relaxation does not directly alter the CFG, it is worth mentioning that
+many optimisation passes make use of LLVM's built-in target-based cost model,
+which means that adding or removing instructions from basic blocks may affect
+decisions during transformations of the code.
+For example, some of the optimizations that make use of this built-in cost model
+and also alter the CFG are: loop-unrolling, function inlining, and
+simplifications of the CFG.
+These changes in the program's CFGs may affect other compiler optimisations and
+also possibly in hardware performance, for example affecting the instruction
+caching.
 
@@
 while the WPO relaxation is significantly less conservative.
-Because we performed the WPO relaxation having perfect profiling information, its dynamic error was always below the 5\% thresholds.
-Although it is not guaranteed by the algorithm, profiling representative inputs should also result in small dynamic errors.
+Although the error bound is not guaranteed by the WPO relaxation,
+its dynamic error was always below the 5\% thresholds due to the use
+of perfect profiling information.
 
 \subsection{Evaluation of the Online {\IterComp}}
+
+\begin{figure*}[t]
+    \centering
+    \includegraphics[width=\textwidth]{figs/speedups.pdf}
+    \caption{Speedups obtained from the final optimization sequence selected by the online {\itercomp}.
+	         The speedups reported for each benchmark represents the average speedup across their complete 1000 input datasets.}
+    \label{fig:speedups}
+\end{figure*}
 
@@
 \end{itemize}
-The comparison between Oracle-RM and Oracle-PP is useful for validating the use of the WP metric for guiding {\itercomp}, while the other configurations demonstrate the viability of applying the online {\itercomp} in a real scenario.
+
+The comparison between Oracle-RM and Oracle-PP is crucial for validating the
+effectiveness of the work efficiency metric for guiding online {\itercomp},
+while the other configurations demonstrate the viability of applying online
+{\itercomp} in real-world scenarios.
+For the relaxation strategies, we evaluate only those with a 5\% threshold, since
+they offered better overhead reductions for little dynamic error in the measurement.
 
@@
 
-\begin{figure*}[t]
-    \centering
-    \includegraphics[width=\textwidth]{figs/speedups.pdf}
-    \caption{Speedups obtained from the final optimization sequence selected by the online {\itercomp}.
-	         The speedups reported for each benchmark represents the average speedup across their complete 1000 input datasets.}
-    \label{fig:speedups}
-\end{figure*}
+In order to evaluate the quality of the final optimization sequences selected by
+each configuration of the online {\itercomp}, we compare their speedup over the
+{\flagstype -O3} optimization across all the 1,000 inputs of the benchmark
+being optimized.
+%When measuring the wall-clock time for each input, to reduce noise, we execute
+%the same input until we have a statistically sound measurement, i.e. we execute
+%until we have an interval no larger than 1\% with 99\% confidence.
+Figure~\ref{fig:speedups} shows these average speedups for all test benchmarks.
+This figure shows that the best optimization sequence selected with the Oracle-PP
+is very close to the performance of the best optimization sequence selected with
+the Oracle-RM, where Oracle-PP achieves on average about 80\% of the performance
+improved obtained by Oracle-RM.
+This result is important for demonstrating that a work-based metric has the
+potential to produce good results in a real-world online scenario.
 
-In order to evaluate the quality of the final optimization sequences selected by the {\itercomp} search, we compare their speedup by measuring wall-clock time of the benchmarks when compiled with the standard {\flagstype -O3} optimization.
-For each benchmark, after selecting the final optimization sequence, we compute the average speedup over all the 1000 input dataset.
-When measuring the wall-clock time for each input, to reduce noise, we execute the same input until we have a statistically sound measurement, i.e.
-we execute until we have an interval no larger than 1\% with 99\% confidence.
-Figure~\ref{fig:speedups} shows these average speedups over all the 1000 input dataset.
-This figure shows that the best optimization sequence selected with the Oracle-PP is very close to the performance of the best optimization sequence selected with the Oracle-RM.
-This result is important for demonstrating that a work-based metric has the potential to produce good results in a real online scenario, where there is the restriction that programs execute distinct inputs only once.
-Moreover, the evaluation also indicates that the use of relaxation algorithms does not degrade the optimization search,
-it might even benefit the {\itercomp} as it tends to insert smaller interferences in the performance measurement.
+Moreover, the evaluation also indicates that the use of relaxation algorithms
+does not degrade the optimization search, it might even benefit the {\itercomp}
+as it tends to insert smaller interferences in the performance measurement.
 %Finally, Figure~\ref{fig:speedups} also presents the results of {\itercomp} guided by IPC, which is futher discussed in Figure~\ref{fig:ipc-vs-work}.

### 2017-09-15 motivation.tex
taaaabs. sorry
Errrgh
changes to the result section; and other minor changes
results

@@
     We performed offline iterative compilation on \texttt{susan\_c} to find the optimization settings that maximize the average performance
-    over \FIXME{five} randomly chosen inputs from a set of 1,000 provided by the \textit{KDataSets} benchmark
-	suite~\cite{chen10,chen12a}. The green histogram in Figure~\ref{fig:motivation-online} shows for each of the 1,0000 inputs the speedup versus \texttt{-O3}
-    of that chosen optimization setting. The dark green line shows the average speedup for all those inputs. The
-    \FIXME{five} training inputs are all in the top \FIXME{twelve} performing inputs. On the other hand, the chosen optimization does
-    poorly on other inputs, often causing a slowdown and yielding only a slight speedup of 6\% on average.
+    over five randomly chosen inputs from a set of 1,000 provided by the \textit{KDataSets} benchmark suite~\cite{chen10,chen12a}.
+    The green histogram in Figure~\ref{fig:motivation-online} shows for each of the 1,0000 inputs the speedup versus \texttt{-O3} of that
+    chosen optimization setting. The dark green line shows the average speedup for all those inputs. The five training inputs are
+    all among the top performing inputs. On the other hand, the chosen optimization does poorly on other inputs, often causing
+    a slowdown and yielding only a slight speedup of 4\% on average.
 
     The blue bars in Figure~\ref{fig:motivation-online} show a different approach. Here the iterative compilation proceeds by evaluating
-    each optimization on \FIXME{five} inputs. Each input is evaluated only once during the search, but the search is provided with oracle
-    knowledge about the unoptimized runtime for the inputs, giving a perfect speedup metric. The environment of the search constantly changes
-    in the same way as for online iterative compilation. The best optimization for this case never degrades performance compared to
+    each optimization on five inputs. Each input is evaluated only once during the search, but the search is provided with oracle
+    knowledge about the unoptimized runtime for the inputs, giving a perfect speedup metric. The environment of the search constantly
+    changes in the same way as for online iterative compilation. The best optimization for this case never degrades performance compared to
     \texttt{-O3} and is on average 13\% better than the offline approach. The online case, considering truly representative inputs, leads
@@
             Each point represents averages over a subset of all inputs.
+            The IPC shows a correlation coefficient of $-0.14$.
         }
@@
     redundant NOPs to a program will increase IPC while also slowing the program down. Figure~\ref{fig:motivation-metric} shows the
-    relationship between execution time, IPC and speedup for different optimizations of \texttt{susan\_c}, averaged across a subset of the inputs.
-    Neither IPC nor execution time correlate well, and so neither can be used for iterative compilation. We need a work efficiency metric
-    which can approximate speedup instead, using only a single execution.
+    relationship between execution time, IPC and speedup for different optimizations of \texttt{susan\_c}, averaged across a subset of the
+    inputs. Neither IPC nor execution time correlate well, and so neither can be used for iterative compilation. We need a work efficiency
+    metric which can approximate speedup instead, using only a single execution.
 

### 2017-09-15 prof.tex
results

@@
 Graph theory shows that we can calculate precisely the frequencies of all edges between basic blocks if we choose a spanning tree of the
-function CFG and instrument the edges not in the tree. To make the instrumentation probe placement optimal, we choose the maximum spanning
-tree, with the edge weights representing edge frequency estimates. By doing so, we leave as many high frequency edges as possible in the
-maximum spanning tree and they will not be instrumented, allowing us to minimize the instrumentation overhead.
+function CFG and instrument the edges in the complement of the spanning tree~\cite{nahapetian73,forman81}.
+To make the instrumentation probe placement optimal, we choose the maximum spanning tree,
+with the edge weights representing edge frequency estimates. By doing so, we leave as many high frequency edges as possible in the
+maximum spanning tree and they will not be instrumented, allowing us to minimize the instrumentation overhead~\cite{forman81,ball94}.
 
@@
 Intuitively, if all the edge flows are known for the complement of a spanning tree then at any leaf of the spanning tree there is only one
-unknown edge flow. This unknown edge flow can be calculated by Kirchhoff's first law\FIXME{\cite{}}. This process repeats as a bottom-up
+unknown edge flow. This unknown edge flow can be calculated by Kirchhoff's first law\cite{knuth73,ball94}. This process repeats as a bottom-up
 propagation until all the unknown edge flows have been calculated. This algorithm can be formally defined as a post-order traversal on the

### 2017-09-15 results.tex
results

@@
 Moreover, another important aspect to point out is that we also observed
-differences among the final CFGs after applying the \texttt{-O3} optimisation.
+differences among the final CFGs after applying the \texttt{-O3} optimization.
 Although relaxation does not directly alter the CFG, it is worth mentioning that
-many optimisation passes make use of LLVM's built-in target-based cost model,
+many optimization passes make use of LLVM's built-in target-based cost model,
 which means that adding or removing instructions from basic blocks may affect
@@
 simplifications of the CFG.
-These changes in the program's CFGs may affect other compiler optimisations and
+These changes in the program's CFGs may affect other compiler optimizations and
 also possibly in hardware performance, for example affecting the instruction
@@
 each configuration of the online {\itercomp}, we compare their speedup over the
-{\flagstype -O3} optimization across all the 1,000 inputs of the benchmark
-being optimized.
+\texttt{-O3} optimization across all the 1,000 inputs of the benchmark being
+optimized.
+None of the configurations degrades the performance over \texttt{-O3} (with
+statistical significance).
+
 %When measuring the wall-clock time for each input, to reduce noise, we execute
@@
 the Oracle-RM, where Oracle-PP achieves on average about 80\% of the performance
-improved obtained by Oracle-RM.
-This result is important for demonstrating that a work-based metric has the
-potential to produce good results in a real-world online scenario.
+improvement obtained by the Oracle-RM.
+This result is important for demonstrating that our work efficiency metric has
+the potential to produce good results in real-world online scenarios.
 
-Moreover, the evaluation also indicates that the use of relaxation algorithms
-does not degrade the optimization search, it might even benefit the {\itercomp}
-as it tends to insert smaller interferences in the performance measurement.
-%Finally, Figure~\ref{fig:speedups} also presents the results of {\itercomp} guided by IPC, which is futher discussed in Figure~\ref{fig:ipc-vs-work}.
+The Real-OP achieves 4\% improvement on average, which represents 45\% of the
+performance improvement obtained by the Oracle-RM.
+This difference to Oracle-PP is explained by the real (intrusive) profiling used
+in the Real-OP configuration.
+The profiling affects the search in two key ways:
+\textit{(i.)} the overhead incurred by the profiling affects the execution time,
+and therefore also affects the work efficiency metric,
+\textit{(ii.)} the instrumentation code may affect many of the optimizations, e.g.,
+due to the use of a global variable or in decisions taken based on a cost-model
+for the instructions.
+However, even in the Real-OP configuration, which has a highly intrusive instrumentation,
+it is still possible to obtain performance improvements in a realistic online
+{\itercomp}.
 
-%\begin{figure*}[h]
-%    \centering
-%    \includegraphics[width=\textwidth]{figs/ipc-vs-work.pdf}
-%    \caption{Comparison between {\itercomp} using IPC or the WP metric.}
-%    \label{fig:ipc-vs-work}
-%\end{figure*}
-
-%Figure~\ref{fig:ipc-vs-work} presents a comparison between IPC and the WP metric regarding their correlation with the actual speedup over \verb|-O0| observed during {\itercomp}.
-%Each point in the figure corresponds to each metric averaged over input windows collected during multiple executions of the {\itercomp}.
-%The vertical lines correspond to the optimization sequence selected by the {\itercomp} search, which either represents the highest IPC or the highest value of WP.
-%In all cases, the optimization sequence selected based on the WP is much closer to the best speedup than the optimization sequence selected based on IPC, which can be attributed to the fact that the WP has a stronger correlation with the speedup.
-%In all cases, IPC shows little to no correlation with the speedup, which corroborates the argument given in Section~\ref{sec:ipc-vs-work-metric}.
-
-%For some cases, namely for the tiff-related benchmarks, the WP shows a worse correlation with speedup.
-%We believe these cases could be improved by improving the cost model used to compute the work metric.
-%However, even in these cases, the optimization sequences selected based on the WP outperform those selected based on the IPC metric.
-
-%\textbf{ToDo: Include results using Genetic Algorithm search.}
+Finally, the evaluation also indicates that the use of relaxation algorithms
+is beneficial not only for its reduction in overhead, which is directly observed
+by the user, by it also tends to improve the quality of the search.
+This improvement has to do with the aforementioned effects of an intrusive
+instrumentation on the search.
+The relaxation has the potential to reduce both effects of the profiling.
+Both the Real-R5 and Real-WPO-R5 achieve about 60\% of the
+performance improvement obtained by the Oracle-RM.

### 2017-09-15 work.tex
changes to the result section; and other minor changes
results

@@
     \[
-       P = \nicefrac{\Delta W}{\Delta t}
+       \textrm{Work Efficiency} = \frac{\Delta W}{\Delta t}
     \]
@@
     Figure~\ref{fig:cost-model} shows the correlation of the model with runtime on instances of test benchmarks. The correlation
-    coefficient is \FIXME{XXX}, and the mean absolute percentage error (MAPE) is \FIXME{XXX}. Figure~\ref{fig:motivation-work-efficiency}
-    shows a similar plot to Figure~\FIXME{\ref{fig:???}}, of work efficiency versus speedup for 500 different binaries of \texttt{susan\_c}.
-Here the new metric correlates much better than either performance or IPC.
+    coefficient is $0.99$, and the mean absolute percentage error (MAPE) is about 24\%.
+    Figure~\ref{fig:motivation-work-efficiency} shows a similar plot to Figure~\FIXME{\ref{fig:motivation-metric}},
+    of work efficiency versus speedup for 500 different binaries of \texttt{susan\_c}.
+    Here the new metric correlates much better than either execution time or IPC,
+    exhibiting a correlation coefficient of $0.93$ with speedup.
 
@@
         \caption{Relationship between work efficiency vs speedup for 500 different binaries of \texttt{susan\_c}.
-                Each point represents averages over a subset of all inputs.}
+                Each point represents averages over a subset of all inputs.
+                Correlation coefficient of $0.93$.}
         \label{fig:motivation-work-efficiency}

### 2017-09-15 motivation.tex
updates figures

@@
         \centering
+        %\begin{tabular}{cc}
+        %\includegraphics[width=0.25\textwidth]{figs/motivation-metric-ipc.pdf} &
+        %\includegraphics[width=0.25\textwidth]{figs/motivation-metric-runtime.pdf}
+        %\end{tabular}
         \includegraphics[width=0.5\textwidth]{figs/motivation-metric.pdf}
         \caption{
-            Relationship between IPC and execution time vs optimization speedup for 500 different binaries of \texttt{susan\_c}.
-            Each point represents averages over a subset of all inputs.
-            The IPC shows a correlation coefficient of $-0.14$.
+            Relationship between IPC and execution time \textit{vs} speedup over the unoptimized program
+            for 500 different binaries of \texttt{susan\_c}.
+            Each point represents the average over a subset of all inputs.
+            The IPC and execution time show correlation coefficients of $-0.14$ and $-0.16$, respectively.
         }

### 2017-09-15 work.tex
updates figures

@@
         \includegraphics[width=0.5\textwidth]{figs/motivation-work-efficiency.pdf}
-        \caption{Relationship between work efficiency vs speedup for 500 different binaries of \texttt{susan\_c}.
-                Each point represents averages over a subset of all inputs.
-                Correlation coefficient of $0.93$.}
+        \caption{Relationship between work efficiency \textit{vs} speedup for 500 different binaries of \texttt{susan\_c}.
+                Each point represents the average over a subset of all inputs.
+                The correlation coefficient is $0.93$.}
         \label{fig:motivation-work-efficiency}

### 2017-09-15 abstract.tex
changes to abstract

--- 2017-09-13 abstract.tex
@@
 
-    Iterative or adaptive compilation can dramatically improve the performance of a program by searching for the best optimization
-    settings. However, it relies upon repeatedly evaluating and comparing optimizations off-line on predefined
-    representative inputs. If the developers fail to guess what is a representative input or usage patterns change over time, 
+    Iterative compilation can dramatically improve the performance of a program by searching for the best optimization.
+    However, it relies upon repeatedly evaluating optimizations on predefined representative inputs.
+    If the developers fail to guess what is a representative input or usage patterns change over time,
     then the wrong optimizations may be selected.
     Searching the optimization space with actual user inputs, during deployment, has previously seemed infeasible.
-    Program side effects, interactivity and difficulties capturing the system state mean that any given input can 
-    only be executed once. Since each input may entail a different amount of work, the search cannot use runtime to compare 
-    optimizations, and there is no baseline against which to define speedup.
+    Program side effects and difficulties capturing the system state mean that any given input can
+    only be executed once. %Since each input may entail a different amount of work, the search cannot use runtime to compare
+    %optimizations, and there is no baseline against which to define speedup.
     Unless we find a way to compare optimizations even when inputs are executed only once, we will be unable to use real user
-    input for iterative compilation, users will not have the best optimized programs, and performance and energy efficiency will suffer.
+    input for iterative compilation, users will not have the best optimized programs, and performance will suffer.
 
@@
     We give two variants for probe removal optimization: the first offers strong
-    guarantees about the maximum possible error, while the second targets the average, whole program error.
+    guarantees about the maximum possible error, while the second targets the average whole program error.
 
@@
     but without executing any input more than once, which the oracle cannot do.
-    Through probe removal we bring the instrumentation overhead down to only 4\% on average, making our
-    approach suitable for regular use. We show practical online iterative compilation for the first time, 
-    optimizing programs according to the range of inputs users actually present.
-
+    Through probe removal we bring the instrumentation overhead down to only 5\% on average, making our
+    approach suitable for real-world scenarios. %We show practical online iterative compilation for the first time,
+    %optimizing programs according to the range of inputs users actually present.
 \end{abstract}

### 2017-09-15 results.tex
changes to abstract

@@
 
-A few of the benchmark present abnormal regressions in their execution time.
+A few of the benchmarks present abnormal regressions in their execution time.
 This increase in the profiling overhead is counter-intuitive because the
@@
 without changing their placement.
-We suggest some reasons for this abnormal regressions.
-By analysing these cases, we noticed that several of the frequently executed
-probes could not be removed by the relaxation.
-Moreover, another important aspect to point out is that we also observed
+We suggest some reasons for these abnormal regressions.
+By analyzing these cases, we noticed that several of the frequently executed
+probes could not be removed by the relaxation strategies with small thresholds.
+Another important aspect to point out is that we also observed
 differences among the final CFGs after applying the \texttt{-O3} optimization.
@@
 
-Furthermore, these improvements can be obtained while incurring very little dynamic error, as shown in Figure~\ref{fig:error-O3}.
-This result confirms the overly conservative aspect of the relaxation algorithm applied per DAG,
-while the WPO relaxation is significantly less conservative.
+Finally, as shown in Figure~\ref{fig:error-O3}, these improvements can be
+obtained while incurring very little dynamic error.
+This result confirms the overly conservative aspect of the relaxation algorithm
+applied per DAG, while the WPO relaxation is significantly less conservative.
 Although the error bound is not guaranteed by the WPO relaxation,
@@
 The comparison between Oracle-RM and Oracle-PP is crucial for validating the
-effectiveness of the work efficiency metric for guiding online {\itercomp},
+effectiveness of the work efficiency metric in guiding online {\itercomp},
 while the other configurations demonstrate the viability of applying online
@@
 For the relaxation strategies, we evaluate only those with a 5\% threshold, since
-they offered better overhead reductions for little dynamic error in the measurement.
+they offered better overhead reductions for little dynamic errors in the measurement.
 

### 2017-09-15 motivation.tex
motivation and setup

@@
     relationship between execution time, IPC and speedup for different optimizations of \texttt{susan\_c}, averaged across a subset of the
-    inputs. Neither IPC nor execution time correlate well, and so neither can be used for iterative compilation. We need a work efficiency
+    inputs. We use correlation coefficient~\cite{Bishop:2006:PRM:1162264} to quantify if  IPC and execution time is strongly correlated to
+    speedup. This correlation coefficient takes a value between -1 and 1, the closer the coefficient is to +/-1, the stronger the
+    correlation between the variables. In our case, the IPC and execution time show correlation coefficients of $-0.14$ and $-0.16$,
+    respectively, to speedup. These values are far from 1, suggesting that there is a weak correlation between speedup and IPC or
+    execution time. Clearly, neither IPC nor execution time correlate well, and so neither can be used for iterative compilation. We need a work efficiency
     metric which can approximate speedup instead, using only a single execution.

### 2017-09-15 results.tex
experimental results part 1

@@
 
-In this section we evaluate the performance overhead incurred by the work efficiency profiling.
-We measure the wall-clock time of the benchmarks when compiled with the {\flagstype -O3} optimization.
-For each benchmark, we compute the average overhead over all its 1,000 inputs.
-Figure~\ref{fig:overhead-O3} shows the performance overhead imposed by the work
-profiling with each one of the instrumentation strategies, compared to their
-non-instrumented counterparts.
-The WPO relaxation was performed using profiling information for every specific
-input, which provides the perfect information to perform the WPO relaxation.
-This means that the experiments show its best performance enabled by having
-perfect profiling information.
+In this experiment we evaluate the runtime overhead introduced by the work efficiency profiling. We compile the program using the
+{\flagstype -O3}. We then measure the wall-clock time of the instrumented and the native versions to calculate the runtime overhead. For
+each benchmark, we compute the average overhead over all its 1,000 inputs.
+
+Figure~\ref{fig:overhead-O3} shows the runtime overhead imposed by the work profiling using each of the instrumentation strategies over
+their non-instrumented counterparts. The WPO relaxation was performed using profiling information for every specific input, which provides
+the perfect information to perform the WPO relaxation. This means that the experiments show its best performance enabled by having perfect
+profiling information.
 
@@
 
-While the optimal profiling has a maximum overhead of almost 60\%,
-the work profiling with a relaxation of 5\% threshold has an overhead of less
-than 16\% for all benchmarks,
-with the WPO relaxation reaching an average overhead of only 5\%,
-compared to an average of 12\% for the optimal profiling.
+While the optimal profiling has a maximum overhead of near 60\%, the work profiling with a relaxation of 5\% threshold has an overhead of
+less than 16\% for all benchmarks, with the WPO relaxation reaching an average overhead of only 5\%, compared to an average of 12\% for the
+optimal profiling.
 
@@
 which consists mainly of a single hot loop with several branches inside it. %, as depicted by Figure~\ref{fig:adpcm_d-cfg-instr}.
-Most of the improvement achieved by both strategies comes from removing only two
-probes from the hot loop.
-The two relaxed probes were placed in branches with a high probability of being
-taken (above 55\% probability), but with a small contribution to the total amount
-of work measured in DAG for this loop (both probes with an static error of
-about 1.3\%).
-Because these probes were placed in basic blocks with a considerable execution
-frequency, their removal resulted in a significant reduction in performance
-overhead.
+Most of the improvement achieved by both strategies comes from removing only two probes from the hot loop. The two relaxed probes were
+placed in branches with a high probability of being taken (above 55\% probability), but with a small contribution to the total amount of
+work measured in DAG for this loop (both probes with an static error of about 1.3\%). Because these probes were placed in basic blocks with
+a considerable execution frequency, removing the probes resulted in a significant reduction in runtime overhead.
 
@@
 
-\subsection{Evaluation of the Online {\IterComp}}
+\subsection{Evaluation of Online {\IterComp}}
 

### 2017-09-15 results.tex
small tweaks on experimental results

@@
 
-For comparison, we use five configurations.
-In all configurations, the same optimization sequence is used for multiple inputs, using a dynamic input-window size, as explained in Section~\ref{sec:oic-infra}.
-The average performance over the input window provides an estimate for the overall performance of the optimization sequence across distinct inputs.
-The optimization sequences are ranked based on their average performance, and the best optimization sequence is selected.
+\subsubsection{\Itercomp Strategies}
+In this experiment, we use evaluate our \itercomp approach against an oracle. We consider five \itercomp strategies:
 %When selecting the best optimization sequence, they are ranked by the average of their performance measurement, using the WP metric, except for the Oracle-RM which uses the real speedup over \texttt{-O0}.
@@
 \end{itemize}
+
+In all configurations, the same optimization sequence is used for multiple inputs, using a dynamic input-window size, as explained in
+Section~\ref{sec:oic-infra}. The average performance over the input window provides an estimate for the overall performance of the
+optimization sequence across distinct inputs. The optimization sequences are ranked based on their average performance, and the best
+optimization sequence is selected.
 
@@
 
-\subsection{Performance Evaluation}
+In order to evaluate the quality of the final optimization sequences selected by each configuration of the online {\itercomp}, we compare
+their speedup over the \texttt{-O3} optimization across all the 1,000 inputs of the benchmark being optimized. None of the configurations
+degrades the performance over \texttt{-O3} (with statistical significance).
 
-In order to evaluate the quality of the final optimization sequences selected by
-each configuration of the online {\itercomp}, we compare their speedup over the
-\texttt{-O3} optimization across all the 1,000 inputs of the benchmark being
-optimized.
-None of the configurations degrades the performance over \texttt{-O3} (with
-statistical significance).
+\subsubsection{Results}
 

### 2017-09-15 related.tex
related work rewrite

@@
 \subsection{{\IterComp}}
+Multiple different adaptive compilation techniques have been proposed for finding a near optimal combination of optimization
+settings~\cite{agakov06,hoste2008cole,kulkarni04,stephenson03} by probabilistically searching the space of possible combinations. But with
+all of them there is uncertainty about whether the quality of their results depends on the data set used. Chen~\etal~\cite{chen10,chen12a}
+examined this issue and showed that indeed the best optimization options differ across inputs. Still, in most cases it is possible to find
+a combination of optimizations that produces good results across all inputs, by using a few tens of inputs to evaluate our optimizations
+on. While a positive result, this is still impractical. Producing tens of realistic input sets might be difficult. Even if not, each
+combination of optimization options has to be evaluated on every input set, increasing the time \itercomp takes by at least an order of
+magnitude.
 
-Chen~\etal~\cite{chen10,chen12a} evaluate the effectiveness of iterative compilations across a large number of input test cases.
-Their main motivation is to answer the question:
-\textit{How data input dependent is {\itercomp}?}
-Their results show that it is possible to find a combination of compiler optimizations that achieves at least 86\% of the maximum speedup across all input test cases.
-These optimal combinations are program-specific and yield average speedups up to 3.75$\times$ over the highest optimization level of compilers.
+Recent research~\cite{chen12b,fang15} has used real data for performing online {\itercomp} on distributed data center applications. Each
+worker receives a subset of the input dataset to evaluate a small set of optimization settings on. The best such setting from each round
+is used for subsequent executions of the same code. By testing new settings and re-evaluating old ones on new datasets this approach makes
+it more likely to select an optimization setting that works well across inputs. Like in~\cite{chen10} every setting is tested on the same
+set of inputs, wasting time and energy. This approach only works well with MapReduce-like workloads, since it relies on the MapReduce
+framework for repeating the same computation multiple times without side-effects or corrupting the input.
 
-When optimising a program, the main method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a} evaluates each combination of compiler optimizations across all the available inputs, i.e., if $N$ is the number of input test cases and $M$ is the total number of combinations of compiler optimizations, they perform a total of $O(NM)$ runs of the program being optimized.
-Furthermore, they use a pre-defined set of only 300 different combinations of compiler optimizations, which represents a very small sample of the optimization search space for most modern compilers, e.g.
-LLVM has 56 distinct optimization passes and GCC has about 47 high-level (SSA form) optimization passes and about 25 low-level (RTL) optimization passes, which in both cases result in much more than $2^{50}$ distinct combinations of compiler optimizations, without considering repetition.
-
-Recent work~\cite{chen12b,fang15} have applied the same idea of performing input-dependent {\itercomp} to distributed applications on data centres.
-In summary, each worker receives a subset of the input dataset, called the evaluation dataset, to perform an \textit{online} {\itercomp} of the code being executed.
-Each worker performs the same the method for {\itercomp} used by Chen~\etal~\cite{chen10,chen12a}, i.e., they evaluate each combination of compiler optimizations across all the evaluation dataset.
-However, because the optimization is performed online, they usually consider a small evaluation dataset and a small number of compiler optimizations.
-
-Fursin~\etal~\cite{fursin07} addressed the problem of comparing the effect of two optimizations on two distinct inputs.
-For that purpose, they proposed to use instruction per cycle (IPC) as the metric for performing such comparison.
-Their result show that using IPC seems promising as a robust metric for {\itercomp} across large input datasets.
-However, some specific optimization techniques may affect the use of IPC as a robust metric, and specially IPC has been shown to provide poor performance estimation for multi-threaded programs~\cite{alameldeen06,eyerman08}.
-In particular, IPC can give a skewed performance measure if threads spend time in \textit{spin-lock loops} or other synchronisation mechanisms.
-Some existing work on performance assessment suggest that total execution time should be used for measuring performance of multi-threaded programs~\cite{alameldeen06,eyerman08}.
-Aalameldeen and Wood~\cite{alameldeen06} in particular suggest that a simple work-related metric should be used if the unit of work is representative enough.
-Work-related metrics have already been largely used for measuring performance of throughput-oriented applications, for other applications, however, choosing an appropriate unit of work can be more challenging~\cite{alameldeen06}.
+Fursin~\etal~\cite{fursin07} attempted to find a more general approach for online \itercomp. They proposed instructions per cycle (IPC) as
+a metric for comparing runs of the same program with different optimization settings and different inputs: application binaries displaying
+higher IPC are more efficient than lower IPC binaries. Their results show that IPC seems promising as a robust metric for {\itercomp}.
+However, some optimization techniques affect IPC in the opposite way, making the program more efficient while lowering IPC. In the context
+of multithreaded application, IPC cannot be used to estimate performance~\cite{alameldeen06,eyerman08}. Aalameldeen and
+Wood~\cite{alameldeen06} suggest a work metric for quantifying efficiency, but choosing a unit of work is program specific and challenging.
 
@@
 
-In 1973, Donald Knuth~\cite{knuth73} provided an optimal algorithm for profiling block frequency, inserting fewer probes than the previously described naive instrumentation that was commonly used in practice~\cite{knuth71}.
-Forman~\cite{forman81} proposed further overhead reductions by placing the probes in basic blocks that are less likely to be executed based on static heuristics.
-
-Ball and Larus~\cite{ball94} offered a detailed discussion comparing two approaches for optimally profiling block frequency, namely, placing probes on the vertices or the edges of the control-flow graph (CFG).
-They show that the edge-based approach produces optimal placement of probes.
+Donald Knuth~\cite{knuth73} introduced an optimal algorithm for profiling block frequency, inserting much fewer probes than naive
+instrumentation approaches that were used in practice~\cite{knuth71}. Forman~\cite{forman81} proposed further overhead reductions by
+placing the probes in basic blocks that are less likely to be executed based on static heuristics. Ball and Larus~\cite{ball94} offered a
+detailed discussion comparing two approaches for optimally profiling block frequency, namely, placing probes on the vertices or the edges
+of the control-flow graph (CFG). They show that the edge-based approach produces optimal placement of probes. Theirs is the algorithm we
+used for our initial selection of probes before relaxation.

### 2017-09-15 motivation.tex
overall changes

--- 2017-09-15 motivation.tex
@@
             speedup when using the optimization sequence selected by offline \itercomp, the blue one is the speedup using the optimization
-            sequence selected by an online-like approach. \FIXME{Beware Daltonism!}
+            sequence selected by an online-like approach.
         }
@@
 
-    We performed offline iterative compilation on \texttt{susan\_c} to find the optimization settings that maximize the average performance
+    We performed offline iterative compilation on the \texttt{susan\_c} benchmark to find the optimization settings that maximize the average performance
     over five randomly chosen inputs from a set of 1,000 provided by the \textit{KDataSets} benchmark suite~\cite{chen10,chen12a}.
-    The green histogram in Figure~\ref{fig:motivation-online} shows for each of the 1,0000 inputs the speedup versus \texttt{-O3} of that
+    The green histogram in Figure~\ref{fig:motivation-online} shows, for each of the 1,000 inputs, the speedup versus \texttt{-O3} of that
     chosen optimization setting. The dark green line shows the average speedup for all those inputs. The five training inputs are
@@
     relationship between execution time, IPC and speedup for different optimizations of \texttt{susan\_c}, averaged across a subset of the
-    inputs. Neither IPC nor execution time correlate well, and so neither can be used for iterative compilation. We need a work efficiency
+    inputs. We use correlation coefficient~\cite{Bishop:2006:PRM:1162264} to quantify if  IPC and execution time are strongly correlated to
+    speedup.
+    %This correlation coefficient takes a value between -1 and 1, the closer the coefficient is to +/-1, the stronger the
+    %correlation between the variables.
+    In our case, the IPC and execution time show correlation coefficients of $-0.14$ and $-0.16$,
+    respectively, to speedup.
+    %These values are far from 1, suggesting that there is a weak correlation between speedup and IPC or execution time.
+    Neither IPC nor execution time correlate well, and so neither can be used for iterative compilation. We need a work efficiency
     metric which can approximate speedup instead, using only a single execution.

### 2017-09-15 prof.tex
overall changes

@@
-\section{Work Profiling}
+\section{Work Profiling}\label{sec:prof}
 
 In this section, we describe how to measure work online. Total work is just the sum of the costs of all executed basic blocks, so it is
-possible to embed its computation into the execution of the program. A na\"{\i}ve instrumentation would use a global counter initialized at
-the interception value, $\varepsilon$. Each basic block would then have only to increment this counter with the total cost of its
+possible to embed its computation into the execution of the program. A na\"{\i}ve instrumentation would use a global counter initialized 
+to zero. Each basic block would then have only to increment this counter with the total cost of its
 instructions every time it is executed. Although easy to implement, the na\"{\i}ve approach introduces a significant overhead.
@@
 Intuitively, if all the edge flows are known for the complement of a spanning tree then at any leaf of the spanning tree there is only one
-unknown edge flow. This unknown edge flow can be calculated by Kirchhoff's first law\cite{knuth73,ball94}. This process repeats as a bottom-up
-propagation until all the unknown edge flows have been calculated. This algorithm can be formally defined as a post-order traversal on the
+unknown edge flow. This unknown edge flow can be calculated by Kirchhoff's first law\cite{knuth73,ball94},
+also known as the principle of \textit{conservation of flow}, which states that
+the amount of flow into a vertex equals the amount of output flow.
+This process repeats as a bottom-up propagation until all the unknown edge flows have been calculated. This algorithm can be formally defined as a post-order traversal on the
 spanning tree. Let $G$ be the CFG. This algorithm first initializes the edge flows $D_{(u,v)}$, for each edge $(u,v)$ in the CFG, as
@@
 \]
-Afterwards, for each vertex $u\in G$, in a post-order traversal of the spanning tree,
+This initialization represents only the trivially known edge flows, which are the
+probes themselves (the highlighted edges in red).
+
+Afterwards, the algorithm consists in propagating the known edge flows, infering
+the symbolic expressions in the unknown edge flows.
+This is achieved by performing a bottom-up propagation from the trivially known
+edge flows (in red) by applying the Kirchhoff's first law, i.e.,
+for each vertex with a single unknown edge flow, we compute the difference between
+the known incoming edge flows by the known outgoing edge flows.
+Formally,
+for each vertex $u\in G$, in a post-order traversal of the spanning tree,
 we can use the Kirchhoff's first law and apply the following operations

### 2017-09-15 work.tex
overall changes

--- 2017-09-14 work.tex
@@
 
-    Similarly to previous work~\citep{giusto01,powell09,brandolese11}, the coefficients, $w_i$, are found by linear regression. We create
-    one regressor data point for each of a set of program and input pairs. The programs are given in Table~\ref{tab:kdatasets:training},
-    and each has 1,000 inputs. The response variable is the runtime of the unoptimized program as it processes the input. To ensure
-    statistically sound results, the runtime is measured by repeated execution until the 99\% confidence interval is no larger than 1\% of
+    Similarly to previous work~\citep{giusto01,powell09,brandolese11}, the coefficients, $w_i$, are estimated by linear regression. We create
+    one regressor data point for each of a set of program and input pairs.
+    We use a separate set of programs from those discussed in Sections~\ref{sec:setup} and \ref{sec:results}.
+    %for estimating the architectural settings for the work metric.
+    The response variable is the execution time of the unoptimized program as it processes the input. To ensure
+    statistically sound results, the execution time is measured by repeated execution until the 99\% confidence interval is no larger than 1\% of
     the mean. The explanatory variables are the dynamic instruction counts for each program and input pair. They are recorded by a separate,
@@
 
-    Figure~\ref{fig:cost-model} shows the correlation of the model with runtime on instances of test benchmarks. The correlation
-    coefficient is \FIXME{XXX}, and the mean absolute percentage error (MAPE) is \FIXME{XXX}. Figure~\ref{fig:motivation-work-efficiency}
-    shows a similar plot to Figure~\FIXME{\ref{fig:???}}, of work efficiency versus speedup for 500 different binaries of \texttt{susan\_c}.
-Here the new metric correlates much better than either performance or IPC.
+    Figure~\ref{fig:work-vs-O0time} shows the correlation of the work metric ($\Delta W$) with runtime on instances of the unoptimized \texttt{susan\_c} benchmark.
+    Each point represents the execution of a single input, from a total of 1,000 inputs.
+    The correlation coefficient is $0.99$, and the mean absolute percentage error (MAPE) is only about 3.5\%.
+    This result suggests that the new metric correlates much better than either execution time or IPC.
+
+    %Figure~\ref{fig:motivation-work-efficiency} shows a similar plot to Figure~\FIXME{\ref{fig:motivation-metric}},
+    %of work efficiency versus speedup for 500 different binaries of \texttt{susan\_c}.
+    %Here the new metric correlates much better than either execution time or IPC,
+    %exhibiting a correlation coefficient of $0.93$ with speedup.
 
@@
         \centering
-        \includegraphics[width=0.6\linewidth]{figs/cost-model.pdf}
+        \includegraphics[width=0.5\textwidth]{figs/work-correlation-with-O0_susan_c.pdf}
         \caption{
-            Linear model for work metric fitted from empirical data.
-            \red{zw: bigger font size.}
+            Relationship between work metric ($\Delta W$) and the execution time
+            of the unoptimized version of the \texttt{susan\_c} benchmark over the
+            1,000 inputs.
+            Each point represents the execution of a single input.
+            They have a correlation coefficient of $0.99$.
         }
-        \label{fig:cost-model}
+        \label{fig:work-vs-O0time}
     \end{figure}
 
-    \begin{figure}[t]
-        \centering
-        \includegraphics[width=0.5\textwidth]{figs/motivation-work-efficiency.pdf}
-        \caption{Relationship between work efficiency vs speedup for 500 different binaries of \texttt{susan\_c}.
-                Each point represents averages over a subset of all inputs.}
-        \label{fig:motivation-work-efficiency}
-    \end{figure}
+    %\begin{figure}[t]
+    %    \centering
+    %    \includegraphics[width=0.6\linewidth]{figs/cost-model.pdf}
+    %    \caption{
+    %        Linear model for work metric fitted from empirical data.
+    %        \red{zw: bigger font size.}
+    %    }
+    %    \label{fig:cost-model}
+    %\end{figure}
+
+    %\begin{figure}[t]
+    %    \centering
+    %    \includegraphics[width=0.5\textwidth]{figs/motivation-work-efficiency.pdf}
+    %    \caption{Relationship between work efficiency \textit{vs} speedup for 500 different binaries of \texttt{susan\_c}.
+    %            Each point represents the average over a subset of all inputs.
+    %            The correlation coefficient is $0.93$.}
+    %    \label{fig:motivation-work-efficiency}
+    %\end{figure}

### 2017-09-15 prof.tex
some of the changes to the structure of the Relaxation Section

@@
 
-In contrast to the naive instrumentation where each basic block records only its own amount of work, with the optimal profiling the work
+In contrast to the na\"ive instrumentation where each basic block records only its own amount of work, with the optimal profiling the work
 counter increments needs to take other basic blocks into account. Figure~\ref{fig:cfg-example} shows an example CFG, where the instrumented
@@
 
-Optimal instrumentation significantly reduces the profiling overhead when compared to naive instrumentation, from an average overhead of
-79\% to 13\%. In some extreme cases though optimal instrumentation can result in overheads of up to 60\% (see benchmark \texttt{adpcm\_d}
-in Figure~\ref{fig:overhead-O3}). To further reduce the overhead in these cases, we propose a relaxation strategy that offers a trade-off
-between accuracy and overhead. While in traditional basic block profiling the aim is to accurately measure the frequency counts of every
-single block, in our case we only care about the total amount of work. Our insight was that if the work associated with a probe is small,
-then ignoring the probe will have little effect on the total work estimation. By removing such probes, especially if they are frequently
+Optimal instrumentation significantly reduces the profiling overhead when compared to na\"ive instrumentation, from an average overhead of
+79\% to 13\%. In some extreme cases, optimal instrumentation can result in overheads of up to 60\% (see benchmark \texttt{adpcm\_d}
+in Figure~\ref{fig:overhead-O3}). To further reduce the overhead in these cases, we propose two relaxation strategies that offer a trade-off
+between accuracy and overhead. Our insight was that if the work associated with a probe is small,
+then ignoring the probe will have little effect on the total work estimation.
+By removing such probes, especially if they are frequently
 executed, we can reduce the instrumentation overhead dramatically.
+To contrast with optimal profiling alone,
+while the optimal placement of probes tries to reduce overhead by placing probes in edges that are less likely to be executed,
+the relaxation strategies focus on removing probes that are more likely to be executed but add little to the work metric.
 
 Our relaxation strategy operates on selection of probes and weights as decided by the optimal algorithm. Removing low weight probes is
-easy, but we need a way to guarantee that the error in our work estimation will not grow too big. Our conservative approach, decides on an
-upper bound for the error and applies it recursively on every loop of the CFG. Specifically, our algorithm extracts all subgraphs
+easy, but we need a way to guarantee that the error in our work estimation will not grow too big.
+To this end, we first propose a more conservative approach that
+calculates induced error from removing a probe by considering
+    the worst case of any execution path through the program,
+and then we propose a more aggressive approach that
+    computes induced error based on the relative change of the work measurement that removing a probe would make to the total amount work for the whole program.
+We discuss each one of them in detail.
+
+\subsubsection{\WCRelaxTitle Relaxation}
+
+Our conservative approach, decides on an upper bound for the error and applies it recursively on every loop of the CFG. Specifically, our algorithm extracts all subgraphs
 representing loops and transforms them into DAGs (Directed Acyclic Graphs) by removing the backedge and moving inner loops into separate
@@
 
-\subsection{Whole Program Relaxation}
+\subsubsection{\WPRelaxTitle Relaxation}
 
-The proposed relaxation strategy provides hard guarantees that error will remain below the chosen threshold but in many cases these
-guarantees are too conservative. Our worst case scenario approach assumes that each loop will always perform the minimum amount of work
-possible which might be orders of magnitude less than the work actually performed. In such cases, we remove too few probes and have little
-effect on the instrumentation overhead, while the actual relative error in our work estimation is negligible. To avoid this problem, we
-propose an adapted version of the relaxation algorithm that operates on the whole program.
+The \WCRelaxLower relaxation provides hard guarantees that error will remain below the chosen threshold but in many cases these
+guarantees are too conservative. It assumes that each loop will always execute the path with the minimum amount of work
+possible, which might be orders of magnitude less than the work actually performed. In such cases, we remove too few probes and have little
+effect on the profiling overhead, while the actual relative error in our work estimation is negligible. To avoid this problem, we
+propose a more aggressive approach that operates on the whole program.
 
-\textit{Whole program relaxation} works by using block-frequency profiling from previous executions. With this information, we are able to
+The \WPRelaxLower relaxation works by using block-frequency profiling from previous executions. With this information, we are able to
 compute the error of removing a given probe in terms of the whole program's execution, and then select a subset of all the probes to be
@@
 
-For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we model the whole program relaxation as the following 0-1 Knapsack
-problem:
+For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, similar to the \WCRelaxLower relaxation,
+we model the \WPRelaxLower relaxation as the following 0-1 Knapsack problem:
 \begin{gather*}
@@
 error threshold, and $\varepsilon(P_i)$ is the error of removing probe $P_i$ relative to the total work.
-%if $\Delta W$ is the work
-%value for the whole program's execution, computed from the basic block frequencies profiled from previous executions, the error for a
-%given probe $P_i$ is
-%\[
-%\varepsilon(P_i) = \frac{\omega(P_i)f(P_i)}{\Delta W}.
-%\]
 
-Contrary to the per DAG relaxation, the error introduced by whole program relaxation is not guaranteed to be bounded by the error threshold
-$M$. Still, we expect it to remain close to the threshold as long as the program's behavior does not deviate significantly from the
-behavior captured in profile we used.
+Contrary to the per DAG approach, the error introduced by \WPRelaxLower relaxation is not guaranteed to be bounded by the threshold $M$.
+Still, we expect it to remain close to the threshold as long as the program's behavior does not deviate significantly from the behavior
+captured in profile we used.

### 2017-09-15 related.tex
some of the changes to the structure of the Relaxation Section

@@
 
-Donald Knuth~\cite{knuth73} introduced an optimal algorithm for profiling block frequency, inserting much fewer probes than naive
+Donald Knuth~\cite{knuth73} introduced an optimal algorithm for profiling block frequency, inserting much fewer probes than na\"ive
 instrumentation approaches that were used in practice~\cite{knuth71}. Forman~\cite{forman81} proposed further overhead reductions by

### 2017-09-15 results.tex
overall changes
some of the changes to the structure of the Relaxation Section

@@
-\section{Experimental Results}
+\section{Experimental Results}\label{sec:results}
 
@@
 %our work metric. Finally we present the results of the online {\itercomp}.
+
+Profiling strategies used throughout this section 
+for measuring either the actual speedup or the work efficiency metric.
+\begin{itemize}[leftmargin=3mm]
+\item \textbf{\OracleRM} measures the actual speedup over the unoptimized version of the program for any given input.
+In our experiments, this oracle is allowed to execute each input at least twice so the speedup can be computed.
+\item \textbf{\OraclePP} measures the work efficiency metric with a \textit{magically perfect} non-intrusive profiling.
+  Because this version simulates a \textit{perfect} profiling, with zero overhead,
+  it allows us to isolate and evaluate the actual effectiveness of the work efficiency metric.
+\item \textbf{\OptProf} corresponds to the work profiling using the optimal placement of the probes.
+  This profiling strategy requires a single execution of the program for any given input.
+\item \textbf{\WCRelax-\textit{N}\%} measures the work efficiency metric using the work profiling with a \textit{N}\% thresheld for
+the \WCRelaxLower relaxation.
+  Similar to the \OptProf, this profiling strategy also requires a single execution of the program for any given input.
+\item \textbf{\WPRelax-\textit{N}\%} is similar to the \WCRelax, but it applies the \WPRelaxLower relaxation instead.
+\end{itemize}
 
@@
 Most of the improvement achieved by both strategies comes from removing only two probes from the hot loop. The two relaxed probes were
-placed in branches with a high probability of being taken (above 55\% probability), but with a small contribution to the total amount of
+placed in branches with a high probability of being taken, but with a small contribution to the total amount of
 work measured in DAG for this loop (both probes with an static error of about 1.3\%). Because these probes were placed in basic blocks with
@@
   it allows us isolate and evaluate the effectiveness work-efficiency metric.
-\item \textbf{Real-OP} corresponds to the online {\itercomp} as it would be applied in a real-world online scenario.
+\item \textbf{Opt-Prof} corresponds to the online {\itercomp} as it would be applied in a real-world online scenario.
   The program is executed only once with each input.

### 2017-09-15 prof.tex
minor changes

@@
 Our relaxation strategy operates on selection of probes and weights as decided by the optimal algorithm. Removing low weight probes is
-easy, but we need a way to guarantee that the error in our work estimation will not grow too big.
+easy, but we need a way to guarantee that the error in our work estimation will not grow too large.
 To this end, we first propose a more conservative approach that
@@
 
-For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$, similar to the \WCRelaxLower relaxation,
-we model the \WPRelaxLower relaxation as the following 0-1 Knapsack problem:
+For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$,
+we also model the \WPRelaxLower relaxation as a 0-1 Knapsack problem,
+very similar to the \WCRelaxLower relaxation but with a different definition for the relative error,
+that is:
 \begin{gather*}

### 2017-09-15 conclusions.tex
conclusions rewrite

--- 2017-09-06 conclusions.tex
@@
-\section{Conclusion and Future Work}
+\section{Conclusions}
+This paper introduced a novel efficiency metric which enables us to compare in terms of performance versions of an application compiled in
+different ways even when they perform different amounts of work. Our approach automatically estimates the work done by each basic block of
+the application and inserts instrumentation code to increment a global work counter every time a block is executed. We optimize the
+placement of the instrumentation probes to reduce the runtime overhead. On top of it, we implement two strategies for removing probes from
+the application to control the trade-off between runtime overhead and work estimation accuracy.
 
-Our proposed technique for relaxed instrumentation can also be applied for similar profiling scenarios, such as profiling empirical computational complexity~\cite{goldsmith07,zaparanuks12,coppa14}.
+We use this efficiency metric to enable for the first time true online iterative compilation. Instead of comparing different optimizations
+sequences offline by running the application repeatedly with different optimizations applied but the same input, we can now compare runs
+using different inputs as they are provided by the user. We evaluated our online iterative compilation methodology using the KDataSets
+benchmark suite. Experimental results show that we deliver \red{80\%} of the speedup achieved by an offline oracle while not having to 
+execute any input more than once. Our probe optimization and removal strategies bring the instrumentation overhead down to only \red{4\%}
+on average, making our approach suitable for integration into software products.
+
+%This paper has presented a novel work efficiency metric to enable one to compare the performance of any two different compiled versions of
+%a program without executing a given input more than once. This ability allows us to implement, for the first time, a true online iterative
+%compilation  to search for the best compilation options  based on actual user inputs seen in the deployment environment.
+
+%We develop a compiler-based, low-overhead instrumentation mechanism to gather the require data to evaluate the work metric. Our mechanism
+%does this without any human involvement. We show that by carefully selecting where to place the probes in the source code, the
+%instrumentation overhead can be negligible. We give two probe optimization strategies, allowing developers to balance the instrumentation
+%overhead and the performance error.
+
+%We  evaluate our approach using the KDataSets benchmark suit. Experimental results show that our approach delivers \red{80\%} of the
+%speedup achieved by an offline  oracle, but without executing any input more than once. We show that our probe optimization strategy is
+%highly effective, which brings the instrumentation overhead down to only \red{4\%} on average, making our approach practical for regular
+%use.
+
+%Our proposed technique for relaxed instrumentation can also be applied for similar profiling scenarios, such as profiling empirical
+%computational complexity~\cite{goldsmith07,zaparanuks12,coppa14}.

### 2017-09-15 prof.tex
relaxation

@@
 
-Our conservative approach, decides on an upper bound for the error and applies it recursively on every loop of the CFG. Specifically, our algorithm extracts all subgraphs
-representing loops and transforms them into DAGs (Directed Acyclic Graphs) by removing the backedge and moving inner loops into separate
-DAGs. An example of this is shown in Figure~\ref{fig:cfg-relax-example}, where the CFG is partitioned into two DAGs.
+In order to guarantee the upper bound for the error, this conservative approach
+decomposes the CFGs of the program into \textit{directed acyclic graphs} (DAGs).
+Then, it remove probes within each DAG while
+keeping the total error per DAG bounded by the threshold.
+
+By constraining the percentage error of every DAG below a given threshold,
+it is guaranteed that the dynamic error of the work metric
+will always be bounded by the threshold.
+This constraint is ensured by computing the error of removing each probe relative
+to the path with the minimum amount of work possible.
+
+The \WCRelaxLower relaxation starts by extracting DAGs from the CFG.
+First, the algorithm extracts all the subgraphs that represent a loop or the outer most region of the function.
+Afterwards, these subgraphs are transformed into DAGs by ignoring the backedge and also by considering that any loop within the subgraph is never executed, i.e., only the headers of the inner loops are actually included into the DAG
+An example of this is shown in Figure~\ref{fig:cfg-relax-example}, where the CFG is partitioned into two DAGs.
 
@@
 %\end{lstlisting}
-
+Finally, the probe removal happens by solving an optimization model.
 For every DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we calculate $m$, the minimum amount of work that can be performed in the

### 2017-09-15 results.tex
minor stuff

@@
 each benchmark, we compute the average overhead over all its 1,000 inputs.
-
 Figure~\ref{fig:overhead-O3} shows the runtime overhead imposed by the work profiling using each of the instrumentation strategies over
-their non-instrumented counterparts. The WPO relaxation was performed using profiling information for every specific input, which provides
-the perfect information to perform the WPO relaxation. This means that the experiments show its best performance enabled by having perfect
-profiling information.
+their non-instrumented counterparts.
+%The WPO relaxation was performed using profiling information for every specific input, which provides
+%the perfect information to perform the WPO relaxation. This means that the experiments show its best performance enabled by having perfect
+%profiling information.
 
@@
 Although the error bound is not guaranteed by the WPO relaxation,
-its dynamic error was always below the 5\% thresholds due to the use
-of perfect profiling information.
+its dynamic error was always below the 5\% thresholds.
+%due to the use of perfect profiling information.
 

### 2017-09-15 infra.tex
overall changes
infra agrees with figure 4

@@
     \includegraphics[width=\linewidth]{figs/infra-diagram}
-    \caption{Overview of the online \itercomp mechanism. \red{zw: mark the boundary of online and offline.}}
+    \caption{Overview of the online \itercomp mechanism.}
     \label{fig:infra-diagram}
@@
 
-This work focuses on the highlighted components of Figure~\ref{fig:infra-diagram}. The first one, \textit{Work Instrumentation} \blue{is
-performed once at offline compile time}. It allows measuring the amount of work performed. Conceptually, we use the cost model of
+This work focuses on the highlighted component of Figure~\ref{fig:infra-diagram}, the \textit{Work Instrumentation}. It is performed once
+and offline, right after the IR is generated. Its purpose is to measure the amount of work performed. Conceptually, we use the cost model of
 Section~\ref{sec:metric} to estimate a cost for each basic block of the IR and then add instrumentation to increase a global work counter
-by that cost every time a basic block is executed. The measured work should depend only on the program and the input, not the applied
-optimizations, so we instrument the unoptimized IR before any optimizations modify it. We present a detailed description of work
-instrumentation and how to keep it lightweight in Section~\ref{chap:instr}.
+by that cost every time a basic block is executed. The measured work represents the computational effort of the unoptimised program and it
+should depend only on the program and the input, not the applied optimizations.
+For that reason, we instrument the unoptimized IR before any optimizations modify it.
+We present a detailed description of work
+instrumentation and how to keep it lightweight in Section~\ref{sec:prof}.
 
-\textit{Optimization Selector} is responsible for selecting which optimization sequence to use for the next execution of the program. Each
-sequence is used for multiple executions of the program in order to gather enough work efficiency measurements. When we are able to
+Another component, the \textit{Search Driver}, is responsible for guiding the optimization process. Each selected optimization sequence
+is used for multiple executions of the program in order to gather enough work efficiency measurements. When we are able to
 estimate this work efficiency with a confidence interval narrower than a certain threshold, we can start evaluating a different
 optimization sequence. We call the number of executions performed using the same optimization sequence the \textit{Input-Window Size}.
-
-%\FIXME{WHAT IS THE OPTIMIZATION EVALUATOR? WHY IS IT HIGHLIGHTED? DO WE HAVE ANYTHING TO SAY ABOUT IT?}
 

### 2017-09-15 prof.tex
minor in prof

@@
 In contrast to the na\"ive instrumentation where each basic block records only its own amount of work, with the optimal profiling the work
-counter increments needs to take other basic blocks into account. Figure~\ref{fig:cfg-example} shows an example CFG, where the instrumented
-blocks are marked with red. In this example, executing block $B_5$ implies not executing $B_6$. To compute the amount of work associated
+counter increments need to take other basic blocks into account. Figure~\ref{fig:cfg-example} shows an example CFG, where the instrumented
+blocks are marked with red. In this example, reaching probe $P_2$ implies not only executing block $B_5$ but also not executing $B_6$.
+
+To compute the amount of work associated
 with each instrumentation probe, we first need to determine how reaching a probe relates with executing or not a basic block. Starting from

### 2017-09-15 work.tex
work minor
Work

@@
 
-    Similarly to previous work~\citep{giusto01,powell09,brandolese11}, the coefficients, $w_i$, are estimated by linear regression. We create
-    one regressor data point for each of a set of program and input pairs.
-    We use a separate set of programs from those discussed in Sections~\ref{sec:setup} and \ref{sec:results}.
+    Similarly to previous work~\citep{giusto01,powell09,brandolese11}, the coefficients, $w_i$, are estimated by linear regression. We
+    create one regressor data point for each of a set of program and input pairs. We use a separate set of programs from those discussed in
+    Sections~\ref{sec:setup} and \ref{sec:results}.
     %for estimating the architectural settings for the work metric.
-    The response variable is the execution time of the unoptimized program as it processes the input. To ensure
-    statistically sound results, the execution time is measured by repeated execution until the 99\% confidence interval is no larger than 1\% of
-    the mean. The explanatory variables are the dynamic instruction counts for each program and input pair. They are recorded by a separate,
-    instrumented run of the program on the input.
+    The response variable is the execution time of the unoptimized program as it processes the input. To ensure statistically sound
+    results, the execution time is measured by repeated execution until the 99\% confidence interval is no larger than 1\% of the mean. The
+    explanatory variables are the dynamic instruction counts for each program and input pair. They are recorded by a separate, instrumented
+    run of the program on the input.
 
-    Figure~\ref{fig:work-vs-O0time} shows the correlation of the work metric ($\Delta W$) with runtime on instances of the unoptimized \texttt{susan\_c} benchmark.
-    Each point represents the execution of a single input, from a total of 1,000 inputs.
-    The correlation coefficient is $0.99$, and the mean absolute percentage error (MAPE) is only about 3.5\%.
-    This result suggests that the new metric correlates much better than either execution time or IPC.
+    Figure~\ref{fig:work-vs-O0time} shows the correlation of the work metric ($\Delta W$) with runtime on instances of the unoptimized
+    \texttt{susan\_c} benchmark. Each point represents the execution of a single input, from a total of 1,000 inputs. The correlation
+    coefficient is $0.99$, and the mean absolute percentage error (MAPE) is only about 3.5\%. This result suggests that the new efficiency
+    metric correlates much better with speedup than either execution time or IPC.
 

### 2017-09-15 results.tex
overall changes in results

@@
 
-Profiling strategies used throughout this section 
-for measuring either the actual speedup or the work efficiency metric.
+In this section we discuss our experimental evaluation of the work profiling strategies and
+the effectiveness of the work efficiency metric on guiding true online {\itercomp}.
+We use the following naming convension for the profiling strategies considered
+throughout this section:
 \begin{itemize}[leftmargin=3mm]
@@
 In this experiment we evaluate the runtime overhead introduced by the work efficiency profiling. We compile the program using the
-{\flagstype -O3}. We then measure the wall-clock time of the instrumented and the native versions to calculate the runtime overhead. For
+\texttt{-O3}. We then measure the wall-clock time of the instrumented and the native versions to calculate the runtime overhead. For
 each benchmark, we compute the average overhead over all its 1,000 inputs.
-Figure~\ref{fig:overhead-O3} shows the runtime overhead imposed by the work profiling using each of the instrumentation strategies over
+Figure~\ref{fig:overhead-O3} shows the runtime overhead imposed by the work profiling using each of the proposed work profiling strategies over
 their non-instrumented counterparts.
@@
 
-While the optimal profiling has a maximum overhead of near 60\%, the work profiling with a relaxation of 5\% threshold has an overhead of
-less than 16\% for all benchmarks, with the WPO relaxation reaching an average overhead of only 5\%, compared to an average of 12\% for the
-optimal profiling.
+While the \OptProf has a maximum overhead of near 60\%, the \WCRelax-\textit{5}\% has an overhead of
+less than 16\% for all benchmarks, with the \WPRelax-\textit{5}\% reaches an average overhead of only 5\%,
+compared to an average overhead of 12\% for the optimal profiling.
 
@@
 
+%Mean values...
+%Real-R5: 30.43\%
+%WPO-R5:  54.97\%
+%adpcm_d:
+%Real-R5: 75.51\%
+%WPO-R5:  79.07\%
+
 The most critical case, amongst the evaluated benchmarks, happens with the
@@
 instrumentation.
-With a threshold of only 5\%, the relaxation reduces
-the overhead by a factor of $4.5\times$ over the optimal instrumentation,
-and the WPO relaxation by a factor of $4.8\times$.
+With a threshold of only 5\%, the \WCRelaxLower relaxation reduces
+the overhead by about 75\% over the \OptProf,
+and the \WPRelax-5\% reduces it by about 80\%.
 This benchmark has a single hot function
@@
 We suggest some reasons for these abnormal regressions.
-By analyzing these cases, we noticed that several of the frequently executed
+By analyzing these cases, we noticed that some of the frequently executed
 probes could not be removed by the relaxation strategies with small thresholds.
@@
 Although relaxation does not directly alter the CFG, it is worth mentioning that
-many optimization passes make use of LLVM's built-in target-based cost model,
+there are optimizations that make use of LLVM's built-in target-based cost model,
 which means that adding or removing instructions from basic blocks may affect
-decisions during transformations of the code.
-For example, some of the optimizations that make use of this built-in cost model
-and also alter the CFG are: loop-unrolling, function inlining, and
-simplifications of the CFG.
-These changes in the program's CFGs may affect other compiler optimizations and
-also possibly in hardware performance, for example affecting the instruction
-caching.
+decisions during transformations of the code, which may end up changing the CFG
+in different ways depending on the instrumentation applied.
+%For example, some of the optimizations that make use of this built-in cost model
+%and also alter the CFG are: loop-unrolling, function inlining, and
+%simplifications of the CFG.
+%These changes in the program's CFGs may affect other compiler optimizations and
+%also possibly in hardware performance, for example affecting the instruction
+%caching.
 
@@
 obtained while incurring very little dynamic error.
-This result confirms the overly conservative aspect of the relaxation algorithm
-applied per DAG, while the WPO relaxation is significantly less conservative.
-Although the error bound is not guaranteed by the WPO relaxation,
+This result confirms the overly conservative aspect of the \WCRelaxLower relaxation,
+while the \WPRelaxLower relaxation is significantly less conservative.
+Although the error bound is not guaranteed by the \WPRelaxLower relaxation,
 its dynamic error was always below the 5\% thresholds.
@@
 
-\subsubsection{\Itercomp Strategies}
-In this experiment, we use evaluate our \itercomp approach against an oracle. We consider five \itercomp strategies:
-%When selecting the best optimization sequence, they are ranked by the average of their performance measurement, using the WP metric, except for the Oracle-RM which uses the real speedup over \texttt{-O0}.
-\begin{itemize}[leftmargin=3mm]
-\item \textbf{Oracle-RM} executes the program twice for each input, measuring the real speedup for each optimization sequence, and then uses the real speedup over {\flagstype -O0} for guiding online {\itercomp}.
-This oracle provides a reference results for assessing the effectiveness work-efficiency metric.
-\item \textbf{Oracle-PP} represents an oracle with a \textit{perfect} non-intrusive profiling.
-  It uses the work-efficiency metric for guiding online {\itercomp}.
-  Because this version simulates a \textit{perfect} profiling, with zero overhead,
-  it allows us isolate and evaluate the effectiveness work-efficiency metric.
-\item \textbf{Opt-Prof} corresponds to the online {\itercomp} as it would be applied in a real-world online scenario.
-  The program is executed only once with each input.
-  It uses the optimal work profiling for computing the work efficiency.
-\item \textbf{Real-R5} is similar to the {Real-OP}, but with a relaxation of 5\% in the work profiling.
-\item \textbf{Real-WPO-R5} is similar to the {Real-OP}, but with a whole-program relaxation of 5\% in the work profiling.
-%\item \textbf{Real-IPC} uses IPC for guiding the online {\itercomp}.
-\end{itemize}
+In this section we evaluate the effectiveness of our work efficiency metric.
+We perform online {\itercomp} using all five profiling strategies, namely
+\OracleRM, \OraclePP,\OptProf,\WCRelax-5\%, and \WPRelax-5\%.
+Because the \OracleRM measures actual speedup, it provides reference results for
+assessing the effectiveness of the work-efficiency metric.
+For the relaxation strategies, we evaluate only those with a 5\% threshold, since
+they offered better overhead reductions for little dynamic errors in the measurement.
 
@@
 
-The comparison between Oracle-RM and Oracle-PP is crucial for validating the
+The comparison between \OracleRM and \OraclePP is crucial for validating the
 effectiveness of the work efficiency metric in guiding online {\itercomp},
@@
 {\itercomp} in real-world scenarios.
-For the relaxation strategies, we evaluate only those with a 5\% threshold, since
-they offered better overhead reductions for little dynamic errors in the measurement.
-
-%Figure~\ref{fig:window-size} shows the average input-window size for each benchmark and configuration.
-%It shows that we can have a statistically sound measurement of the performance metric using just a small number of inputs.
-
-%\begin{figure*}[t]
-%    \centering
-%    \includegraphics[width=\textwidth]{figs/window-size.pdf}
-%    \caption{Average input-window sizes observed during the online {\itercomp}.}
-%    \label{fig:window-size}
-%\end{figure*}
 
@@
 degrades the performance over \texttt{-O3} (with statistical significance).
-
-\subsubsection{Results}
-
 %When measuring the wall-clock time for each input, to reduce noise, we execute
@@
 is very close to the performance of the best optimization sequence selected with
-the Oracle-RM, where Oracle-PP achieves on average about 80\% of the performance
-improvement obtained by the Oracle-RM.
+the \OracleRM, where \OraclePP achieves on average about 80\% of the performance
+improvement obtained by the \OracleRM.
 This result is important for demonstrating that our work efficiency metric has
@@
 
-The Real-OP achieves 4\% improvement on average, which represents 45\% of the
-performance improvement obtained by the Oracle-RM.
-This difference to Oracle-PP is explained by the real (intrusive) profiling used
-in the Real-OP configuration.
+The \OptProf achieves 4\% improvement on average, which represents 45\% of the
+performance improvement obtained by the \OracleRM.
+This difference to \OraclePP is explained by the real (intrusive) profiling used
+in the \OptProf configuration.
 The profiling affects the search in two key ways:
@@
 for the instructions.
-However, even in the Real-OP configuration, which has a highly intrusive instrumentation,
+However, even in the \OptProf configuration, which has a highly intrusive instrumentation,
 it is still possible to obtain performance improvements in a realistic online
@@
 The relaxation has the potential to reduce both effects of the profiling.
-Both the Real-R5 and Real-WPO-R5 achieve about 60\% of the
-performance improvement obtained by the Oracle-RM.
+Both the \WCRelax-5\% and \WPRelax-5\% achieve about 60\% of the
+performance improvement obtained by the \OracleRM.

### 2017-09-15 prof.tex
prof

@@
 
-Afterwards, the algorithm consists in propagating the known edge flows, infering
+Afterwards, the algorithm propagates the known edge flows, inferring
 the symbolic expressions in the unknown edge flows.
@@
 Optimal instrumentation significantly reduces the profiling overhead when compared to na\"ive instrumentation, from an average overhead of
-79\% to 13\%. In some extreme cases, optimal instrumentation can result in overheads of up to 60\% (see benchmark \texttt{adpcm\_d}
-in Figure~\ref{fig:overhead-O3}). To further reduce the overhead in these cases, we propose two relaxation strategies that offer a trade-off
-between accuracy and overhead. Our insight was that if the work associated with a probe is small,
-then ignoring the probe will have little effect on the total work estimation.
-By removing such probes, especially if they are frequently
-executed, we can reduce the instrumentation overhead dramatically.
-To contrast with optimal profiling alone,
-while the optimal placement of probes tries to reduce overhead by placing probes in edges that are less likely to be executed,
-the relaxation strategies focus on removing probes that are more likely to be executed but add little to the work metric.
+79\% to 13\%. In some extreme cases, optimal instrumentation can result in overheads of up to 60\% (see benchmark \texttt{adpcm\_d} in
+Figure~\ref{fig:overhead-O3}). To further reduce the overhead in these cases, we propose two relaxation strategies that offer a trade-off
+between accuracy and overhead. Our insight was that if the work associated with a probe is small, then ignoring the probe will have little
+effect on the total work estimation. By removing such probes, especially if they are frequently executed, we can reduce the instrumentation
+overhead dramatically. To contrast with optimal profiling alone, while the optimal placement of probes tries to reduce overhead by placing
+probes in edges that are less likely to be executed, the relaxation strategies focus on removing probes that are more likely to be executed
+but add little to the work metric.
 
-Our relaxation strategy operates on selection of probes and weights as decided by the optimal algorithm. Removing low weight probes is
-easy, but we need a way to guarantee that the error in our work estimation will not grow too large.
-To this end, we first propose a more conservative approach that
-calculates induced error from removing a probe by considering
-    the worst case of any execution path through the program,
-and then we propose a more aggressive approach that
-    computes induced error based on the relative change of the work measurement that removing a probe would make to the total amount work for the whole program.
-We discuss each one of them in detail.
+To select which probes to remove we consider the probes in a control flow region.
+The saving for removing a probe is the frequency of the block the probe is in.
+Removing a probe adds some error term.
+We solve the 0-1 Knapsack optimization problem, removing probes to maximize the total savings while keeping the accumulated error below a threshold:
+\begin{gather*}
+    \textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
+    \textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
+    x_i\in\{0,1\}, i\in\{0,\ldots,k\}
+\end{gather*}
+where the set of probes in the region being considered is $\{P_0, P_1, \ldots, P_k\}$, $f(P_i)$ is the execution frequency of probe $P_i$ and $x_i$ denotes the probes selected for removal, $\varepsilon(P_i)$ is the error for removing $P_i$ and $M$ is the maximum error threshold.
+
+For our experiments, we implemented two solvers for the 0-1 Knapsack problems: an optimal but potentially slow brute-force solver and a
+greedy heuristic based on sorting the items~\citep{dantzig57}. We use the brute-force solver with a small number of probes, less than
+\FIXME{30}, and the greedy heuristic otherwise.
+
+We have two relaxation strategies which differ in the control flow regions they tackle and how they calculate the error terms. The first is
+a more conservative approach that calculates induced error from removing a probe by considering the worst case of any execution path
+through the program. The second is a more aggressive approach that computes induced error based on the relative change of the work
+measurement that removing a probe would make to the total amount work for the whole program. We discuss each one of them in detail.
 
@@
 This constraint is ensured by computing the error of removing each probe relative
-to the path with the minimum amount of work possible.
+to the path with the minimum amount of work possible through the DAG.
 
@@
 An example of this is shown in Figure~\ref{fig:cfg-relax-example}, where the CFG is partitioned into two DAGs.
-
-%The relaxation starts by extracting DAGs from the CFG.
-%First, the algorithm extracts all the subgraphs that represent a loop or the outer most region of the function.
-%Afterwards, these subgraphs are transformed into DAGs by ignoring the backedge
-%and also by considering that any loop within the subgraph is never executed,
-%i.e., only the headers of the inner loops are actually included into the DAG.
-%Figure~\ref{fig:cfg-relax-example} shows a CFG partitioned into two DAGs (consider
-%only basic blocks and edges completely inside the yellow and green boundaries).
 
@@
 
-%\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
-%// Input: CFG
-%relaxInstrumentation(G) {
-%  for loop in G:
-%     DAG = colapseInnerLoops(loop)
-%     relaxInstrumentedDAG(DAG)
-%  DAG = colapseInnerLoops(G)
-%  relaxInstrumentedDAG(DAG)
-%}
-%\end{lstlisting}
-Finally, the probe removal happens by solving an optimization model.
-For every DAG with a set of probes $\{P_0, P_1, \ldots, P_k\}$, we calculate $m$, the minimum amount of work that can be performed in the
-DAG. Given this, the worst case relative error caused by removing a probe $P_i$ is $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$. Reducing the
-instrumentation overhead as much as possible while keeping the total worst case error below a threshold can be modeled as a 0-1 Knapsack
-problem:
-\begin{gather*}
-\textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
-\textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
-x_i\in\{0,1\}, i\in\{0,\ldots,k\}
-\end{gather*}
-where $f(P_i)$ is the execution frequency of probe $P_i$ and $x_i$ denotes the probes selected for removal. By constraining the worst case
+Finally, the error term is determined by calculating $m$, the minimum amount of work that can be performed in the
+DAG. Given this, the worst case relative error caused by removing a probe $P_i$ is $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$. 
+By constraining the worst case
 relative error in every single DAG of the CFG, we guarantee that the final error introduced by relaxation will always be below the chosen
 threshold.
-% Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold, as demonstrated by Proposition~\ref{prop:relax-bound}.
-%
-% \begin{prop}\label{prop:relax-bound}
-% Let $n_i$ be the number of times a given DAG $i$ is executed, $r_i$ be the total relaxation (amount of work removed) in DAG $i$, and $m_i$ be its minimum amount of work.
-% If $\frac{r_i}{m_i} \leq M$ for every $i$,
-% then the final error of the relaxation will always be bounded by the same threshold.
-% \end{prop}
-% \begin{proof}
-% We can model the overall error of the relaxation as:
-% \[
-% 1 - \frac{n_1(m_1 - r_1) + n_2(m_2 - r_2) + \ldots + n_k(m_k - r_k) + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
-% \]
-% That is,
-% %\begin{equation*}
-% \begin{gather*}
-%  1 - \frac{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} + \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} = \\
-%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
-% \end{gather*}
-% %\end{equation*}
-% If $\frac{r_j}{m_j}$ is the maximum ratio $\frac{r_i}{m_i}$ for every $i$, then
-% \begin{equation*}
-% \begin{aligned}
-%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} &\leq\\
-%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j + c} &\leq\\
-%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &
-% \end{aligned}
-% \end{equation*}
-% If $N = max\{n_i$ for every $i\}$, then
-% \begin{equation*}
-% \begin{aligned}
-%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &\leq\\
-%  \frac{Nr_j + Nr_j + \ldots + Nr_j}{Nm_j + Nm_j + \ldots + Nm_j} &=\\
-%  \frac{Nkr_j}{Nkm_j} = \frac{r_j}{m_j} &\leq M
-% \end{aligned}
-% \end{equation*}
-% \end{proof}
-
-%\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
-%// Input: CFG
-%relaxInstrumentedDAG(DAG){
-%   P = ProbesIn(DAG)
-%   m = minWork(DAG,P)
-%   K = createKnapsackModel(P,m)
-%   Bag = solveKnapsack(K)
-%   for B in (P-Bag):
-%     removeProbe(B)
-%}
-%\end{lstlisting}
-
-%The necessary block-frequency information for optimising both the placement of
-%probes can be acquired from profiles of previous executions of the program or by
-%a static heuristic of the CFG during compilation.
-
-For our experiments, we implemented two solvers for the 0-1 Knapsack problem: an optimal but potentially slow brute-force solver and a
-greedy heuristic based on sorting the items~\citep{dantzig57}. We use the brute-force solver for DAGs with a small number of probes, the
-greedy heuristic otherwise.
 
@@
 effect on the profiling overhead, while the actual relative error in our work estimation is negligible. To avoid this problem, we
-propose a more aggressive approach that operates on the whole program.
+propose a more aggressive approach that operates on the probes from whole program at once.
 
@@
 compute the error of removing a given probe in terms of the whole program's execution, and then select a subset of all the probes to be
-removed.
-
-For a program with a set of probes $\{P_0, P_1, \ldots, P_k\}$,
-we also model the \WPRelaxLower relaxation as a 0-1 Knapsack problem,
-very similar to the \WCRelaxLower relaxation but with a different definition for the relative error,
-that is:
-\begin{gather*}
-\textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
-\textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
-x_i\in\{0,1\}, i\in\{0,\ldots,k\}
-\end{gather*}
-where $f(P_i)$ is the execution frequency of the instrumented basic block $P_i$, $x_i$ denotes the probes selected for removal, $M$ is the
-error threshold, and $\varepsilon(P_i)$ is the error of removing probe $P_i$ relative to the total work.
+removed. $\varepsilon(P_i)$ is the error of removing probe $P_i$ relative to the total work.
 

### 2017-09-16 related.tex
minor changes to related work

@@
 \subsection{{\IterComp}}
-Multiple different adaptive compilation techniques have been proposed for finding a near optimal combination of optimization
-settings~\cite{agakov06,hoste2008cole,kulkarni04,stephenson03} by probabilistically searching the space of possible combinations. But with
+Multiple adaptive compilation techniques have been proposed for finding a near optimal combination of optimization
+settings~\cite{agakov06,hoste08,kulkarni04,stephenson03} by probabilistically searching the space of possible combinations. But with
 all of them there is uncertainty about whether the quality of their results depends on the data set used. Chen~\etal~\cite{chen10,chen12a}
 examined this issue and showed that indeed the best optimization options differ across inputs. Still, in most cases it is possible to find
-a combination of optimizations that produces good results across all inputs, by using a few tens of inputs to evaluate our optimizations
-on. While a positive result, this is still impractical. Producing tens of realistic input sets might be difficult. Even if not, each
-combination of optimization options has to be evaluated on every input set, increasing the time \itercomp takes by at least an order of
+an optimization sequence that produces good results across all inputs, by evaluating the optimizations on a representative set inputs.
+While a positive result, this is still impractical. Producing a set of realistic inputs might be difficult. Even if not, each
+optimization sequence has to be evaluated on every input, increasing the time \itercomp takes by at least an order of
 magnitude.
 
-Recent research~\cite{chen12b,fang15} has used real data for performing online {\itercomp} on distributed data center applications. Each
+Recent research~\cite{chen12b,fang15} has used real data for performing {\itercomp} on distributed data center applications. Each
 worker receives a subset of the input dataset to evaluate a small set of optimization settings on. The best such setting from each round

### 2017-09-16 conclusions.tex
minor

@@
 \section{Conclusions}
-This paper introduced a novel efficiency metric which enables us to compare in terms of performance versions of an application compiled in
-different ways even when they perform different amounts of work. Our approach automatically estimates the work done by each basic block of
-the application and inserts instrumentation code to increment a global work counter every time a block is executed. We optimize the
+This paper has introduced a novel efficiency metric which enables us to compare in terms of performance versions of an application compiled
+in different ways even when they perform different amounts of work. Our approach automatically estimates the work done by each basic block
+of the application and inserts instrumentation code to increment a global work counter every time a block is executed. We optimize the
 placement of the instrumentation probes to reduce the runtime overhead. On top of it, we implement two strategies for removing probes from

### 2017-09-16 related.tex
minor

--- 2017-09-15 related.tex
@@
 \subsection{{\IterComp}}
-Multiple different adaptive compilation techniques have been proposed for finding a near optimal combination of optimization
-settings~\cite{agakov06,hoste2008cole,kulkarni04,stephenson03} by probabilistically searching the space of possible combinations. But with
-all of them there is uncertainty about whether the quality of their results depends on the data set used. Chen~\etal~\cite{chen10,chen12a}
-examined this issue and showed that indeed the best optimization options differ across inputs. Still, in most cases it is possible to find
-a combination of optimizations that produces good results across all inputs, by using a few tens of inputs to evaluate our optimizations
-on. While a positive result, this is still impractical. Producing tens of realistic input sets might be difficult. Even if not, each
-combination of optimization options has to be evaluated on every input set, increasing the time \itercomp takes by at least an order of
-magnitude.
+Various adaptive compilation techniques have been proposed for finding near optimal compilation
+settings~\cite{agakov06,kulkarni04,stephenson03}. These approaches evaluate the performance of candidate compiler options through repeated
+executions using the same input. All these approaches implicitly assume the found compiler settings work for other unseen datasets.
+Chen~\etal~\cite{chen10,chen12a} show that this assumption is not true, and the best optimization options can differ across inputs. Still,
+in most cases it is possible to find a combination of optimizations that produces good results across all inputs, by using a few tens of
+inputs to evaluate our optimizations on. While a positive result, this is still impractical. Producing tens of realistic input sets might
+be difficult. Even if not, each combination of optimization options has to be evaluated on every input set, increasing the time \itercomp
+takes by at least an order of magnitude.
 

### 2017-09-16 infra.tex

@@
 side, we select an optimization sequence and then compile the instrumented IR to binary using this sequence. Next time the user runs the
-application, we measure the work and the execution time. We accumulate multiple such measurements until we can estimate the efficiency
-metric P of this optimization sequence with high confidence. Finally, we use this information to help us select the next optimization
-sequence.
+application, we measure the work and the execution time. We accumulate multiple such measurements, executing each input only once, until we
+can estimate the efficiency metric P of this optimization sequence with high confidence. Finally, we use this information to help us select
+the next optimization sequence.
 
@@
 Section~\ref{sec:metric} to estimate a cost for each basic block of the IR and then add instrumentation to increase a global work counter
-by that cost every time a basic block is executed. The measured work represents the computational effort of the unoptimised program and it
+by that cost every time a basic block is executed. The measured work represents the computational effort of the unoptimized program and it
 should depend only on the program and the input, not the applied optimizations.

### 2017-09-16 related.tex

@@
 settings~\cite{agakov06,kulkarni04,stephenson03}. These approaches evaluate the performance of candidate compiler options through repeated
-executions using the same input. All these approaches implicitly assume the found compiler settings work for other unseen datasets.
-Chen~\etal~\cite{chen10,chen12a} show that this assumption is not true, and the best optimization options can differ across inputs. Still,
-in most cases it is possible to find a combination of optimizations that produces good results across all inputs, by using a few tens of
-inputs to evaluate our optimizations on. While a positive result, this is still impractical. Producing tens of realistic input sets might
-be difficult. Even if not, each combination of optimization options has to be evaluated on every input set, increasing the time \itercomp
-takes by at least an order of magnitude.
+executions using the same input. All these approaches implicitly assume the best-performing compiler settings found from a small set of
+inputs will work well on unseen datasets. Chen~\etal~\cite{chen10,chen12a} show that the best optimization options can differ across
+inputs. Their results suggest that, in most cases it is possible to find a combination of optimizations that produces good results across
+all inputs, by using a few tens of inputs to evaluate our optimizations on. While a positive result, this is still impractical. Producing
+tens of realistic input sets might be difficult. Even if not, each combination of optimization options has to be evaluated on every input
+set, increasing the time \itercomp takes by at least an order of magnitude.
+%=======
+%Multiple adaptive compilation techniques have been proposed for finding a near optimal combination of optimization
+%settings~\cite{agakov06,hoste08,kulkarni04,stephenson03} by probabilistically searching the space of possible combinations. But with
+%all of them there is uncertainty about whether the quality of their results depends on the data set used. Chen~\etal~\cite{chen10,chen12a}
+%examined this issue and showed that indeed the best optimization options differ across inputs. Still, in most cases it is possible to find
+%an optimization sequence that produces good results across all inputs, by evaluating the optimizations on a representative set inputs.
+%While a positive result, this is still impractical. Producing a set of realistic inputs might be difficult. Even if not, each
+%optimization sequence has to be evaluated on every input, increasing the time \itercomp takes by at least an order of
+%magnitude.
+%>>>>>>> origin/master
 
-Recent research~\cite{chen12b,fang15} has used real data for performing online {\itercomp} on distributed data center applications. Each
+Recent research~\cite{chen12b,fang15} has used real data for performing {\itercomp} on distributed data center applications. Each
 worker receives a subset of the input dataset to evaluate a small set of optimization settings on. The best such setting from each round

### 2017-09-16 prof.tex
FIXEdME
make formulae penetrable

@@
 Formally,
-for each vertex $u\in G$, in a post-order traversal of the spanning tree,
-we can use the Kirchhoff's first law and apply the following operations
-in a symbolic fashion:
+for each vertex $u\in G$, following a post-order traversal of the spanning tree,
+the Kirchhoff's first law is applied as the following symbolic operations:
+First, sum all the out-going edge flows
 \[
-\Sigma^+_u = \sum_{v\in N^+(u)} D_{(u,v)}
+%\Sigma^+_u = \sum_{v\in N^+(u)} D_{(u,v)}
+\mathrm{SumOut}(u): \sum_{v\in N^+(u)} D_{(u,v)}
 \]
+and then sum all the in-coming edge flows
 \[
-\Sigma^-_u = \sum_{v\in N^-(u)} D_{(v,u)}
+%\Sigma^-_u = \sum_{v\in N^-(u)} D_{(v,u)}
+\mathrm{SumIn}(u): \sum_{v\in N^-(u)} D_{(v,u)}
 \]
+For all out-going neighbor $v\in N^+(u)$,
 \[
-\forall v\in N^+(u):  D_{(u,v)} \gets
+%\forall v\in N^+(u):  D_{(u,v)} \gets
+D_{(u,v)} \gets
 \begin{cases}
-    D_{(u,v)} & \quad \text{if $D_{(u,v)}\neq 0$}\\
-    \Sigma^-_u - \Sigma^+_u       & \quad \text{otherwise}
+    D_{(u,v)} & \!\!\!\quad \text{if $D_{(u,v)}\neq 0$}\\
+    \mathrm{SumIn}(u) - \mathrm{SumOut}(u)       & \!\!\!\quad \text{otherwise}
 \end{cases}
 \]
+For all in-coming neighbor $v\in N^-(u)$,
 \[
-\forall v\in N^-(u):  D_{(v,u)} \gets
+%\forall v\in N^-(u):  D_{(v,u)} \gets
+D_{(v,u)} \gets
 \begin{cases}
-    D_{(v,u)} & \quad \text{if $D_{(v,u)}\neq 0$}\\
-    \Sigma^+_u - \Sigma^-_u       & \quad \text{otherwise}
+    D_{(v,u)} & \!\!\!\quad \text{if $D_{(v,u)}\neq 0$}\\
+    \mathrm{SumOut}(u) - \mathrm{SumIn}(u)       & \!\!\!\quad \text{otherwise}
 \end{cases}
@@
 In our case, $\kappa_P(x)$ will belong to the set $\{-1,0,1\}$.
-Then, for all probes $P$:
+Once all the edge flows are known, then, 
+for all probes $P$, the amount of work associated
+with probe $P$ is:
 \[
-\omega(P) = \sum_{u\in G} \kappa_P(\Sigma^+_u)w(u)
+%\omega(P) = \sum_{u\in G} \kappa_P(\Sigma^+_u)w(u)
+\omega(P) = \sum_{u\in G} \kappa_P(\mathrm{SumOut}(u))w(u)
 \]
@@
 greedy heuristic based on sorting the items~\citep{dantzig57}. We use the brute-force solver with a small number of probes, less than
-\FIXME{30}, and the greedy heuristic otherwise.
+20, and the greedy heuristic otherwise.
 
@@
 effect on the profiling overhead, while the actual relative error in our work estimation is negligible. To avoid this problem, we
-propose a more aggressive approach that operates on the probes from whole program at once.
+propose a more aggressive approach that operates on the probes from the whole program at once.
 

### 2017-09-16 results.tex
done

@@
 each benchmark, we compute the average overhead over all its 1,000 inputs. 
-Figure~\ref{fig:overhead-O3} shows the runtime overhead imposed by the work profiling using each of the proposed work profiling strategies over
-their non-instrumented counterparts.
+Figure~\ref{fig:overhead-O3} shows the runtime overhead introduced by profiling for each of the proposed strategies.
 %The WPO relaxation was performed using profiling information for every specific input, which provides
@@
 
-While the \OptProf has a maximum overhead of near 60\%, the \WCRelax-\textit{5}\% has an overhead of
-less than 16\% for all benchmarks, with the \WPRelax-\textit{5}\% reaches an average overhead of only 5\%,
-compared to an average overhead of 12\% for the optimal profiling.
+The overhead for \OptProf typically remains below or around 10\% but can reach up to 59\% for \texttt{adpcm\_d}, which would be
+unacceptable. \WCRelax-\textit{5}\% on the other hand manages to eliminate most of the overhead, with a worst
+case of 16\% and an average of 8\%. \WPRelax-\textit{5}\% further reduces the cost of profiling, 5\% overhead on average and 13\%
+worst case. For both \WCRelax and \WPRelax setting the threshold to 2\% makes it too difficult to remove probes. As a result, overhead
+remain high.
 
@@
 
-The most critical case, amongst the evaluated benchmarks, happens with the
-\texttt{adpcm\_d} program, with an overhead of about 59\% for the optimal
-instrumentation.
-With a threshold of only 5\%, the \WCRelaxLower relaxation reduces
-the overhead by about 75\% over the \OptProf,
-and the \WPRelax-5\% reduces it by about 80\%.
-This benchmark has a single hot function
-which consists mainly of a single hot loop with several branches inside it. %, as depicted by Figure~\ref{fig:adpcm_d-cfg-instr}.
-Most of the improvement achieved by both strategies comes from removing only two probes from the hot loop. The two relaxed probes were
-placed in branches with a high probability of being taken, but with a small contribution to the total amount of
-work measured in DAG for this loop (both probes with an static error of about 1.3\%). Because these probes were placed in basic blocks with
-a considerable execution frequency, removing the probes resulted in a significant reduction in runtime overhead.
+The benchmark where \OptProf records its worst result, \texttt{adpcm\_d}, has a single hot function which consists mainly of a single hot
+loop with several branches. This leads to numerous short basic blocks, many of them with probes. Most of the improvement achieved by both
+relaxation strategies comes from removing only two probes from the hot loop. They were initially placed in branches with a high probability
+of being taken but with a small work contribution. Removing either of the probes introduces error of about 1.3\% at the loop level but
+eliminates three quarters of the overhead.
 
-A few of the benchmarks present abnormal regressions in their execution time.
-This increase in the profiling overhead is counter-intuitive because the
-relaxation algorithms only reduces the number of instrumented probes,
-without changing their placement.
-We suggest some reasons for these abnormal regressions.
-By analyzing these cases, we noticed that some of the frequently executed
-probes could not be removed by the relaxation strategies with small thresholds.
-Another important aspect to point out is that we also observed
-differences among the final CFGs after applying the \texttt{-O3} optimization.
-Although relaxation does not directly alter the CFG, it is worth mentioning that
-there are optimizations that make use of LLVM's built-in target-based cost model,
-which means that adding or removing instructions from basic blocks may affect
-decisions during transformations of the code, which may end up changing the CFG
-in different ways depending on the instrumentation applied.
+For a few of the benchmarks, relaxation unexpectedly introduces additional overhead. This is counter-intuitive since relaxation leads to
+less computational work. In these cases, relaxation did not manage to remove high frequency probes, so we did not benefit significantly.
+On the other, the probes that were removed affected how \texttt{-O3} optimized the surrounding code, in some cases blocking useful
+optimizations that were applied originally. This resulted in slower code.
 %For example, some of the optimizations that make use of this built-in cost model
@@
 
-Finally, as shown in Figure~\ref{fig:error-O3}, these improvements can be
-obtained while incurring very little dynamic error.
-This result confirms the overly conservative aspect of the \WCRelaxLower relaxation,
-while the \WPRelaxLower relaxation is significantly less conservative.
-Although the error bound is not guaranteed by the \WPRelaxLower relaxation,
-its dynamic error was always below the 5\% thresholds.
-%due to the use of perfect profiling information.
+In Figure~\ref{fig:error-O3}, we see the error in our work estimation introduced by relaxation. The error for the WC-Relax strategies is
+too low, always below 1\% and usually much lower, regardless of threshold. This confirms that WC-Relax is too conservative. WP-Relax brings
+us closer to our target threshold. Despite not having hard guarantees for the error under WP-Relax, in all cases the error was below the
+threshold.
 
@@
 
-In all configurations, the same optimization sequence is used for multiple inputs, using a dynamic input-window size, as explained in
+For all strategies, the same optimization sequence is used for multiple inputs, using a dynamic input-window size, as explained in
 Section~\ref{sec:oic-infra}. The average performance over the input window provides an estimate for the overall performance of the

### 2017-09-16 results.tex
Spelling

@@
 the effectiveness of the work efficiency metric on guiding true online {\itercomp}.
-We use the following naming convension for the profiling strategies considered
+We use the following naming convention for the profiling strategies considered
 throughout this section:
@@
   This profiling strategy requires a single execution of the program for any given input.
-\item \textbf{\WCRelax-\textit{N}\%} measures the work efficiency metric using the work profiling with a \textit{N}\% thresheld for
+\item \textbf{\WCRelax-\textit{N}\%} measures the work efficiency metric using the work profiling with a \textit{N}\% threshold for
 the \WCRelaxLower relaxation.

### 2017-09-16 related.tex
done with related work

@@
 Various adaptive compilation techniques have been proposed for finding near optimal compilation
-settings~\cite{agakov06,kulkarni04,stephenson03}. These approaches evaluate the performance of candidate compiler options through repeated
-executions using the same input. All these approaches implicitly assume the best-performing compiler settings found from a small set of
-inputs will work well on unseen datasets. Chen~\etal~\cite{chen10,chen12a} show that the best optimization options can differ across
-inputs. Their results suggest that, in most cases it is possible to find a combination of optimizations that produces good results across
-all inputs, by using a few tens of inputs to evaluate our optimizations on. While a positive result, this is still impractical. Producing
-tens of realistic input sets might be difficult. Even if not, each combination of optimization options has to be evaluated on every input
-set, increasing the time \itercomp takes by at least an order of magnitude.
+settings~\cite{agakov06,kulkarni04,stephenson03,hoste08}. These approaches compare the performance of candidate compiler options by first
+compile the program with different options and then profiling the different compiled versions on the same program input. All these
+approaches implicitly assume that the best-performing compiler settings found from a small set of inputs will work well on unseen datasets.
+Chen~\etal~\cite{chen10,chen12a} show that the best optimization options can differ across inputs. Their results suggest that, in most
+cases it is possible to find a single set of optimizations that produces good results across all inputs; but achieving this requires
+developers provide a few tens of representative inputs to be used by the iterative compilation search process. We argue that providing
+representative inputs is difficult in practice, as this require developers to foresee all possible user inputs. Even this is possible,
+current approaches all require to evaluate each candidate compiler option on every single input. Doing so incurs significant overhead,
+which thus prevents iterative compilation to be adopted at a larger scale.
+
 %=======
@@
 
-Recent research~\cite{chen12b,fang15} has used real data for performing {\itercomp} on distributed data center applications. Each
-worker receives a subset of the input dataset to evaluate a small set of optimization settings on. The best such setting from each round
-is used for subsequent executions of the same code. By testing new settings and re-evaluating old ones on new datasets this approach makes
-it more likely to select an optimization setting that works well across inputs. Like in~\cite{chen10} every setting is tested on the same
-set of inputs, wasting time and energy. This approach only works well with MapReduce-like workloads, since it relies on the MapReduce
-framework for repeating the same computation multiple times without side-effects or corrupting the input.
+Recent studies~\cite{chen12b,fang15} have used real data to perform {\itercomp} on distributed data center applications. Each worker
+receives a subset of the input dataset to evaluate a small set of optimization settings on. The best such setting from each round is used
+for subsequent executions of the same code. By testing new settings and re-evaluating old ones on new datasets this approach makes it more
+likely to select an optimization setting that works well across inputs. Like in~\cite{chen10} every setting is tested on the same set of
+inputs, wasting time and energy. This approach only works well with MapReduce-like workloads, since it relies on the MapReduce framework
+for repeating the same computation multiple times without side-effects or corrupting the input.
 
-Fursin~\etal~\cite{fursin07} attempted to find a more general approach for online \itercomp. They proposed instructions per cycle (IPC) as
-a metric for comparing runs of the same program with different optimization settings and different inputs: application binaries displaying
-higher IPC are more efficient than lower IPC binaries. Their results show that IPC seems promising as a robust metric for {\itercomp}.
-However, some optimization techniques affect IPC in the opposite way, making the program more efficient while lowering IPC. In the context
-of multithreaded application, IPC cannot be used to estimate performance~\cite{alameldeen06,eyerman08}. Aalameldeen and
-Wood~\cite{alameldeen06} suggest a work metric for quantifying efficiency, but choosing a unit of work is program specific and challenging.
+Fursin~\etal~\cite{fursin07} attempted to find a general approach for online \itercomp. They proposed to use instructions per cycle (IPC)
+as a metric for comparing different runs of the same program with different optimization settings and different inputs. Their hypothesis is
+that application binaries displaying higher IPC are more efficient than lower IPC binaries. However, many optimization techniques affect
+IPC in the opposite way, making the program more efficient while lowering IPC. We have also shown in Section~\ref{sec:motivation}, IPC is
+not strongly correlate to speedup. Therefore, it is not a good metric for \itercomp. Furthermore, IPC cannot be used to estimate
+performance for multi-threaded applications~\cite{alameldeen06,eyerman08} due to the side-effect of thread contention. Aalameldeen and
+Wood~\cite{alameldeen06} also suggest that relying on IPC can be harmful, and work metrics like time per work unit or work units per time
+should be used. However, prior works all rely on developers to provide such metrics. This paper presents the first automatic approach to
+derive such work metrics from arbitrary code without developer intervention.
 
@@
 
-Donald Knuth~\cite{knuth73} introduced an optimal algorithm for profiling block frequency, inserting much fewer probes than na\"ive
-instrumentation approaches that were used in practice~\cite{knuth71}. Forman~\cite{forman81} proposed further overhead reductions by
-placing the probes in basic blocks that are less likely to be executed based on static heuristics. Ball and Larus~\cite{ball94} offered a
-detailed discussion comparing two approaches for optimally profiling block frequency, namely, placing probes on the vertices or the edges
-of the control-flow graph (CFG). They show that the edge-based approach produces optimal placement of probes. Theirs is the algorithm we
-used for our initial selection of probes before relaxation.
+The seminal work of Donald Knuth~\cite{knuth73} introduced an optimal algorithm for profiling block frequency, inserting much fewer probes
+than the na\"ive instrumentation approaches that were used in practice~\cite{knuth71}. Forman~\cite{forman81} proposed an approach to
+further reduce the instrumentation overhead by placing probes in basic blocks that are less likely to be executed based on static
+heuristics. Ball and Larus~\cite{ball94} offered a detailed discussion to compare these two approaches for optimally profiling block
+frequency, namely, placing probes on the vertices or the edges of the control-flow graph (CFG). They show that the edge-based approach
+produces better placement of probes. Our work built upon these prior works. We use the edge-based approach to place our probes before
+applying relaxation.

### 2017-09-16 results.tex
Formatting
Results

@@
 worst case. For both \WCRelax and \WPRelax setting the threshold to 2\% makes it too difficult to remove probes. As a result, overhead
-remain high.
+remains high.
 
@@
 loop with several branches. This leads to numerous short basic blocks, many of them with probes. Most of the improvement achieved by both
-relaxation strategies comes from removing only two probes from the hot loop. They were initially placed in branches with a high probability
+relaxation strategies comes from removing only two probes from the hot loop. They were initially placed in branches with high probabilities
 of being taken but with a small work contribution. Removing either of the probes introduces error of about 1.3\% at the loop level but
@@
 On the other, the probes that were removed affected how \texttt{-O3} optimized the surrounding code, in some cases blocking useful
-optimizations that were applied originally. This resulted in slower code.
+optimizations that were applied originally. This resulted in slightly slower code.
 %For example, some of the optimizations that make use of this built-in cost model
@@
 
-In Figure~\ref{fig:error-O3}, we see the error in our work estimation introduced by relaxation. The error for the WC-Relax strategies is
-too low, always below 1\% and usually much lower, regardless of threshold. This confirms that WC-Relax is too conservative. WP-Relax brings
+In Figure~\ref{fig:error-O3}, we see the error in our work estimation introduced by relaxation.
+The error for the WC-Relax strategies is low, always below 1\% and usually much lower, regardless of threshold.
+This confirms that WC-Relax is very conservative.
+WP-Relax brings
 us closer to our target threshold. Despite not having hard guarantees for the error under WP-Relax, in all cases the error was below the
@@
 We perform online {\itercomp} using all five profiling strategies, namely
-\OracleRM, \OraclePP,\OptProf,\WCRelax-5\%, and \WPRelax-5\%.
+\OracleRM, \OraclePP, \OptProf, \WCRelax-5\%, and \WPRelax-5\%.
 Because the \OracleRM measures actual speedup, it provides reference results for
@@
 For the relaxation strategies, we evaluate only those with a 5\% threshold, since
-they offered better overhead reductions for little dynamic errors in the measurement.
+they offered better overhead reductions for little dynamic error in the measurement.
 
@@
 while the other configurations demonstrate the viability of applying online
-{\itercomp} in real-world scenarios.
+{\itercomp} in real-world scenarios. \OracleRM manages an average speedup of 9.3\% over \texttt{-O3}. This is modest by typical \itercomp
+standards because search is a relatively small random search. The search driver is not the focus of this work, and we expect higher speedups from more complex drivers, like genetic algorithms.
 
@@
 for the instructions.
-However, even in the \OptProf configuration, which has a highly intrusive instrumentation,
+However, even in the \OptProf configuration
 it is still possible to obtain performance improvements in a realistic online
@@
 by the user, by it also tends to improve the quality of the search.
-This improvement has to do with the aforementioned effects of an intrusive
-instrumentation on the search.
+This improvement has to do with the effects of 
+instrumentation on the search mentioned earlier.
 The relaxation has the potential to reduce both effects of the profiling.

### 2017-09-16 abstract.tex
fix values

@@
 
-    Our online iterative compilation method yields programs that get 80\% of the speedup achievable by an offline oracle,
+    Our online iterative compilation method yields programs that get 60\% of the speedup achievable by an offline oracle,
     but without executing any input more than once, which the oracle cannot do.

### 2017-09-16 conclusions.tex
fix values

@@
 using different inputs as they are provided by the user. We evaluated our online iterative compilation methodology using the KDataSets
-benchmark suite. Experimental results show that we deliver \red{80\%} of the speedup achieved by an offline oracle while not having to
-execute any input more than once. Our probe optimization and removal strategies bring the instrumentation overhead down to only \red{4\%}
+benchmark suite. Experimental results show that we deliver 60\% of the speedup achieved by an offline oracle while not having to
+execute any input more than once. Our probe optimization and removal strategies bring the instrumentation overhead down to only 5\%
 on average, making our approach suitable for integration into software products.

### 2017-09-16 prof.tex
minor change
Spelling
minor comment
minor changes

@@
 \centering {
-  \includegraphics[scale=0.75]{figs/cfg-example.pdf}\\\vspace{1ex}
+  \includegraphics[scale=0.8]{figs/cfg-example.pdf}\\\vspace{1ex}
   \resizebox{0.45\textwidth}{!}{
@@
 \]
-For all out-going neighbor $v\in N^+(u)$,
+For all out-going neighbors $v\in N^+(u)$,
 \[
@@
 \begin{cases}
-    D_{(u,v)} & \!\!\!\quad \text{if $D_{(u,v)}\neq 0$}\\
-    \mathrm{SumIn}(u) - \mathrm{SumOut}(u)       & \!\!\!\quad \text{otherwise}
+    \mathrm{SumIn}(u) - \mathrm{SumOut}(u) & \!\!\!\quad \text{if $D_{(u,v)} = 0$}\\
+    D_{(u,v)}       & \!\!\!\quad \text{otherwise}
 \end{cases}
 \]
-For all in-coming neighbor $v\in N^-(u)$,
+For all in-coming neighbors $v\in N^-(u)$,
 \[
@@
 \begin{cases}
-    D_{(v,u)} & \!\!\!\quad \text{if $D_{(v,u)}\neq 0$}\\
-    \mathrm{SumOut}(u) - \mathrm{SumIn}(u)       & \!\!\!\quad \text{otherwise}
+    \mathrm{SumOut}(u) - \mathrm{SumIn}(u) & \!\!\!\quad \text{if $D_{(v,u)} = 0$}\\
+    D_{(v,u)}  & \!\!\!\quad \text{otherwise}
 \end{cases}
@@
 \]
+It would be equally valid to define $\omega(P)$ in terms of $\mathrm{SumIn}(u)$
+instead of $\mathrm{SumOut}(u)$,
+since these values are equal after populating all edge flows,
+as it is enforced by the Kirchhoff's law.
 
@@
   \centering
-  \includegraphics[scale=0.75]{figs/cfg-relax-example.pdf}
+  \includegraphics[scale=0.8]{figs/cfg-relax-example.pdf}
   \caption{Example of a CFG containing a loop and its decomposition into DAGs.

### 2017-09-16 results.tex
tiny error

@@
 less computational work. In these cases, relaxation did not manage to remove high frequency probes, so we did not benefit significantly.
-On the other, the probes that were removed affected how \texttt{-O3} optimized the surrounding code, in some cases blocking useful
+On the other hand, the probes that were removed affected how \texttt{-O3} optimized the surrounding code, in some cases blocking useful
 optimizations that were applied originally. This resulted in slightly slower code.

### 2017-09-16 results.tex
10 pages

--- 2017-09-16 results.tex
@@
 the effectiveness of the work efficiency metric on guiding true online {\itercomp}.
-We use the following naming convention for the profiling strategies considered
+We use the following naming convension for the profiling strategies considered
 throughout this section:
@@
   This profiling strategy requires a single execution of the program for any given input.
-\item \textbf{\WCRelax-\textit{N}\%} measures the work efficiency metric using the work profiling with a \textit{N}\% threshold for
+\item \textbf{\WCRelax-\textit{N}\%} measures the work efficiency metric using the work profiling with a \textit{N}\% thresheld for
 the \WCRelaxLower relaxation.
@@
 each benchmark, we compute the average overhead over all its 1,000 inputs.
-Figure~\ref{fig:overhead-O3} shows the runtime overhead introduced by profiling for each of the proposed strategies.
+Figure~\ref{fig:overhead-O3} shows the runtime overhead imposed by the work profiling using each of the proposed work profiling strategies over
+their non-instrumented counterparts.
 %The WPO relaxation was performed using profiling information for every specific input, which provides
@@
 
-The overhead for \OptProf typically remains below or around 10\% but can reach up to 59\% for \texttt{adpcm\_d}, which would be
-unacceptable. \WCRelax-\textit{5}\% on the other hand manages to eliminate most of the overhead, with a worst
-case of 16\% and an average of 8\%. \WPRelax-\textit{5}\% further reduces the cost of profiling, 5\% overhead on average and 13\%
-worst case. For both \WCRelax and \WPRelax setting the threshold to 2\% makes it too difficult to remove probes. As a result, overhead
-remain high.
+While the \OptProf has a maximum overhead of near 60\%, the \WCRelax-\textit{5}\% has an overhead of
+less than 16\% for all benchmarks, with the \WPRelax-\textit{5}\% reaches an average overhead of only 5\%,
+compared to an average overhead of 12\% for the optimal profiling.
 
@@
     \label{fig:overhead-O3}
+    \vspace{-2mm}
 \end{figure*}
@@
     \label{fig:error-O3}
+      \vspace{-2mm}
 \end{figure*}
@@
 
-The benchmark where \OptProf records its worst result, \texttt{adpcm\_d}, has a single hot function which consists mainly of a single hot
-loop with several branches. This leads to numerous short basic blocks, many of them with probes. Most of the improvement achieved by both
-relaxation strategies comes from removing only two probes from the hot loop. They were initially placed in branches with a high probability
-of being taken but with a small work contribution. Removing either of the probes introduces error of about 1.3\% at the loop level but
-eliminates three quarters of the overhead.
+The most critical case, amongst the evaluated benchmarks, happens with the
+\texttt{adpcm\_d} program, with an overhead of about 59\% for the optimal
+instrumentation.
+With a threshold of only 5\%, the \WCRelaxLower relaxation reduces
+the overhead by about 75\% over the \OptProf,
+and the \WPRelax-5\% reduces it by about 80\%.
+This benchmark has a single hot function
+which consists mainly of a single hot loop with several branches inside it. %, as depicted by Figure~\ref{fig:adpcm_d-cfg-instr}.
+Most of the improvement achieved by both strategies comes from removing only two probes from the hot loop. The two relaxed probes were
+placed in branches with a high probability of being taken, but with a small contribution to the total amount of
+work measured in DAG for this loop (both probes with an static error of about 1.3\%). Because these probes were placed in basic blocks with
+a considerable execution frequency, removing the probes resulted in a significant reduction in runtime overhead.
 
-For a few of the benchmarks, relaxation unexpectedly introduces additional overhead. This is counter-intuitive since relaxation leads to
-less computational work. In these cases, relaxation did not manage to remove high frequency probes, so we did not benefit significantly.
-On the other, the probes that were removed affected how \texttt{-O3} optimized the surrounding code, in some cases blocking useful
-optimizations that were applied originally. This resulted in slower code.
+A few of the benchmarks present abnormal regressions in their execution time.
+This increase in the profiling overhead is counter-intuitive because the
+relaxation algorithms only reduces the number of instrumented probes,
+without changing their placement.
+We suggest some reasons for these abnormal regressions.
+By analyzing these cases, we noticed that some of the frequently executed
+probes could not be removed by the relaxation strategies with small thresholds.
+Another important aspect to point out is that we also observed
+differences among the final CFGs after applying the \texttt{-O3} optimization.
+Although relaxation does not directly alter the CFG, it is worth mentioning that
+there are optimizations that make use of LLVM's built-in target-based cost model,
+which means that adding or removing instructions from basic blocks may affect
+decisions during transformations of the code, which may end up changing the CFG
+in different ways depending on the instrumentation applied.
 %For example, some of the optimizations that make use of this built-in cost model
@@
 
-In Figure~\ref{fig:error-O3}, we see the error in our work estimation introduced by relaxation. The error for the WC-Relax strategies is
-too low, always below 1\% and usually much lower, regardless of threshold. This confirms that WC-Relax is too conservative. WP-Relax brings
-us closer to our target threshold. Despite not having hard guarantees for the error under WP-Relax, in all cases the error was below the
-threshold.
+Finally, as shown in Figure~\ref{fig:error-O3}, these improvements can be
+obtained while incurring very little dynamic error.
+This result confirms the overly conservative aspect of the \WCRelaxLower relaxation,
+while the \WPRelaxLower relaxation is significantly less conservative.
+Although the error bound is not guaranteed by the \WPRelaxLower relaxation,
+its dynamic error was always below the 5\% thresholds.
+%due to the use of perfect profiling information.
 
@@
     \label{fig:speedups}
+          \vspace{-2mm}
 \end{figure*}

### 2017-09-16 infra.tex
final

@@
 
-Although {\itercomp} had been originally proposed as an \textit{offline} optimization strategy, it can be adapted to work in
-\textit{online} scenarios. Instead of selecting the best optimization sequence at development time, a first version of the program is
-shipped bundled with the {\itercomp} mechanism. Our mechanism produces new application binaries using different optimization sequences.
-When the user starts the application, we load and execute one of these binaries to evaluate the corresponding sequence online. Our
-mechanism operates on the application's Intermediate Representation (IR) produced by LLVM. This saves us the effort of repeatedly parsing
-the code through the frontend and removes the need for shipping the full application code.
+Although {\itercomp} had been originally proposed as an \textit{offline} optimization strategy, this work shows that this powerful
+technique can be adapted to work in \textit{online} scenarios. Instead of selecting the best optimization sequence at development time, a
+first version of the program is shipped bundled with the {\itercomp} mechanism. Our mechanism produces new application binaries using
+different optimization sequences. When the user starts the application, we load and execute one of these binaries to evaluate the
+corresponding sequence online. Our mechanism operates on the application's Intermediate Representation (IR) produced by LLVM. This saves us
+the effort of repeatedly parsing the code through the frontend and removes the need for shipping the full application code.
 

### 2017-09-16 prof.tex
10 pages
minor
final

@@
 
-\paragraph{Populating the edge flows:}
-Intuitively, if all the edge flows are known for the complement of a spanning tree then at any leaf of the spanning tree there is only one
-unknown edge flow. This unknown edge flow can be calculated by Kirchhoff's first law\cite{knuth73,ball94},
-also known as the principle of \textit{conservation of flow}, which states that
-the amount of flow into a vertex equals the amount of output flow.
-This process repeats as a bottom-up propagation until all the unknown edge flows have been calculated. This algorithm can be formally defined as a post-order traversal on the
-spanning tree. Let $G$ be the CFG. This algorithm first initializes the edge flows $D_{(u,v)}$, for each edge $(u,v)$ in the CFG, as
-follows:
+\paragraph{Populating the edge flows.}
+Intuitively, if all the edge flows are known for the complement of a spanning tree, then at any leaf of the spanning tree there is only one
+unknown edge flow. This unknown edge flow can be calculated by Kirchhoff's first law~\cite{knuth73,ball94}, which is also known as the
+principle of \textit{conservation of flow}. Specifically, this theory states that the amount of flow into a vertex equals the amount of
+output flow. To populate the edge flows, we repeat this process as a bottom-up propagation until all the unknown edge flows have been
+calculated. This algorithm can be formally defined as a post-order traversal on the spanning tree. Let $G$ be the CFG. This algorithm first
+initializes the edge flows $D_{(u,v)}$, for each edge $(u,v)$ in the CFG, as follows:
 \[
@@
 \]
-This initialization represents only the trivially known edge flows, which are the
-probes themselves (the highlighted edges in red).
+This initialization represents only the trivially known edge flows, which are the probes themselves, e.g. the highlighted edges in
+Figure~\ref{fig:cfg-example}.
 
@@
 
-\paragraph{Composing the aggregated value of the probes:}
+\paragraph{Composing the aggregated value of the probes.}
 If $x$ is a symbolic expression, then $\kappa_P(x)$ represents the coefficient
@@
            The DAGs are the subgraphs within the dashed boundaries.}
+  \vspace{-4mm}
   \label{fig:cfg-relax-example}

### 2017-09-16 related.tex
final

@@
 approaches implicitly assume that the best-performing compiler settings found from a small set of inputs will work well on unseen datasets.
-Chen~\etal~\cite{chen10,chen12a} show that the best optimization options can differ across inputs. Their results suggest that, in most
-cases it is possible to find a single set of optimizations that produces good results across all inputs; but achieving this requires
+Chen~\etal~\cite{chen10,chen12a} show that the best optimization options, however, can differ across inputs. Their results suggest that, in
+most cases it is possible to find a single set of optimizations that produces good results across all inputs; but achieving this requires
 developers provide a few tens of representative inputs to be used by the iterative compilation search process. We argue that providing
-representative inputs is difficult in practice, as this require developers to foresee all possible user inputs. Even this is possible,
-current approaches all require to evaluate each candidate compiler option on every single input. Doing so incurs significant overhead,
-which thus prevents iterative compilation to be adopted at a larger scale.
+representative inputs is difficult in practice, as users may not use the applications as the developer expects. Even this is possible,
+current approaches all require to evaluate all candidate compiler options on the same input. Doing so across a range of inputs would incur
+significant overhead, which thus prevents iterative compilation to be adopted at a larger scale.
 
@@
 
-Recent studies~\cite{chen12b,fang15} have used real data to perform {\itercomp} on distributed data center applications. Each worker
-receives a subset of the input dataset to evaluate a small set of optimization settings on. The best such setting from each round is used
-for subsequent executions of the same code. By testing new settings and re-evaluating old ones on new datasets this approach makes it more
-likely to select an optimization setting that works well across inputs. Like in~\cite{chen10} every setting is tested on the same set of
-inputs, wasting time and energy. This approach only works well with MapReduce-like workloads, since it relies on the MapReduce framework
-for repeating the same computation multiple times without side-effects or corrupting the input.
+Recent studies~\cite{chen12b,fang15} have used user inputs to perform {\itercomp} in distributed data centers. Using these approaches, each
+compilation worker receives a subset of the input dataset to evaluate a small set of optimization settings on. The best such setting from
+each round is used for subsequent executions of the same code. The best-found compilation strategy is then re-fined over time, by testing
+new settings and re-evaluating old ones on new datasets. Like in~\cite{chen10} every setting is tested on the same set of inputs, wasting
+time and energy. This approach only works well with MapReduce-like workloads, since it relies on the framework for repeating the same
+computation multiple times without causing side-effects.
 
@@
 IPC in the opposite way, making the program more efficient while lowering IPC. We have also shown in Section~\ref{sec:motivation}, IPC is
-not strongly correlate to speedup. Therefore, it is not a good metric for \itercomp. Furthermore, IPC cannot be used to estimate
-performance for multi-threaded applications~\cite{alameldeen06,eyerman08} due to the side-effect of thread contention. Aalameldeen and
-Wood~\cite{alameldeen06} also suggest that relying on IPC can be harmful, and work metrics like time per work unit or work units per time
-should be used. However, prior works all rely on developers to provide such metrics. This paper presents the first automatic approach to
-derive such work metrics from arbitrary code without developer intervention.
+not strongly correlate to speedup. Furthermore, IPC cannot be used to estimate performance for multi-threaded
+applications~\cite{alameldeen06,eyerman08} due to the side-effect of thread contention. Aalameldeen and Wood~\cite{alameldeen06} also
+suggest that relying on IPC can be harmful, and work metrics like time per work unit or work units per time should be used. However, prior
+works all rely on developers to provide such metrics. This paper presents the first automatic approach to derive a such work metric from
+arbitrary code without developer intervention.
 

### 2017-10-19 abstract.tex
minor reviews in abstract

@@
     %optimizations, and there is no baseline against which to define speedup.
-    Unless we find a way to compare optimizations even when inputs are executed only once, we will be unable to use real user
-    input for iterative compilation, users will not have the best optimized programs, and performance will suffer.
+    Unless we find a way to compare optimizations even when inputs are restricted to execute only once, we will be unable to apply 
+    iterative compilation in online scenarios with real user inputs, users will not have the best optimized programs, and performance will suffer.
 

### 2017-10-19 results.tex
minor
final
minor reviews in abstract

@@
     \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.}
+    \vspace{-3mm}
     \label{fig:overhead-O3}
-    \vspace{-2mm}
 \end{figure*}
@@
     \caption{Dynamic error of the work profiling averaged over the 1000 inputs, after relaxing the number of probes.}
+    \vspace{-3mm}
     \label{fig:error-O3}
-      \vspace{-2mm}
 \end{figure*}
@@
     \label{fig:speedups}
-          \vspace{-2mm}
 \end{figure*}

### 2017-10-21 rebuttal.md
Rebuttal

#Review#107A
A.1 The metric is in some ways coarse, but has a strong correlation coefficient to runtime (0.99) and mean absolute error of 3.5% (Fig3). It is enough for iterative compilation (Fig9).

A.2 We used all benchmarks in KDataSets. Some were for the cost model, the others for evaluating our methodology. This standard model
training/evaluation practice. We will make this clear in the paper.

A.3 We will publish the cost model.

#Review#107B

The paper is **not** about *reducing overhead in iterative compilation*. It is about **online** iterative compilation. We will clarify this in the paper.

B.1.Offline iterative compilation is done ahead of time in the development environment. Online iterative compilation is done *live* on users' machines. Offline requires developers to choose representative inputs, online does not. Instead, online optimization targets the actual ways users use the program. Because it is live on users' machines, inputs cannot be run twice. We will clarify this.

B.2.Online iterative compilation has **never** been done before in the general case. Some work has done online optimization where the developer provides a work or throughput metric (e.g. #requests/second assuming each request does the same work). No one has solved the general, automatic case without developer involvement. We are the first in the world to do this. Many papers have suggested it is possible. None have done it and their suggestions would not work. We will clarify this.

Yes, there has been work on input sensitivity. We agree that programs' behavior varies over inputs. That is why online iterative compilation is better, it optimizes according to the inputs the user actual presents. Input sensitive iterative compilation (via multi-versioning etc) is entirely orthogonal to online/offline. Offline requires the developer to choose representative inputs, online does not. Online input sensitive iterative compilation will match the users's inputs better. Until now, online has been impossible. We will clarify this.

The papers, Fursin etal., Ding etal., and Chen etal., are about offline iterative compilation, their approaches do not work in the online case.

That runtimes of programs with different input sets are not directly comparable is not a problem for **offline** iterative compilation, it breaks online iterative compilation. Throughput could be used, but there is no such general metric. Developers would have to provide one (e.g. #requests/second - and ensure each request comprises the same work). Our work is the first general and automatic approach.

IPC has been suggested by other papers as a possible approach to achieve online iterative compilation. None has tried it. We have shown that IPC does not work (Fig2). Our example with NOPs is only the most egregious case, to show logically why IPC is no good. In fact, it is not the cause of IPC's poor correlation in Fig2. Trading high CPI instructions for low CPI instructions (e.g. replace divide by $2^k$ with shift) also shows IPC is no good. Filtering NOPs would not fix the other cases. IPC is only good for architectural optimization. Our observation is back up by paper[17]. We will clarify this and change the example from NOPs.

Our work is not about the *overhead of replay executions*. It is about online iterative compilation where inputs cannot be repeated.

By side effects, we mean changes to external states that may not be repeated (e.g., modifying a database, file system, network activity, interacting with external third-party services). Mitigating these effects would be very cumbersome, if it is possible at all. It is different from the scenario faced by ATLAS and not about caches or noisy environments. We will clarify.

Regarding the distinction between online/offline from Fig4, please see B.1,B.2.

Regarding the shipping of the program to the iterative compilation mechanism: We only mean that the users get the program from somewhere and run it inside the iterative compilation framework. The code needs to be re-compilable, so LLVM bitcode is suitable. Application source code is not required. It is poorly worded, we will fix.

Our work is not meant only for data centers. It is generally applicable. We will clarify.

There do not need to be many alternate binaries stored. It depends on the search strategy. In the smallest case, only one binary exists at any time. Many are produced over time. We will clarify.

    For more on autotuning on the data center, see this paper from CGO16

    Dehao Chen, David Xinliang Li, and Tipp Moseley. 2016. AutoFDO: automatic feedback-directed optimization for warehouse-scale applications.

    The authors claim the proposed strategy is suitable for data center environments. But the evaluation is done on a single Intel system
    with sequential applications.

B.6
We agree that the benchmark suite used does not represent typical data center workloads. Still, we believe that our approach is perfectly suited to data center scenarios and there is no reason it would not work. In a future work, we plan to evaluate our mechanism under such an environment to better support our claim.

    The performance improvement numbers are modest. But this is probably because only 500 sequences are selected out of a possible 200
    factorial sequences in LLVM.
Yes, the performance improvements are modest. We choose a simple, repeatable, random search to allow easy comparison with other metrics. Better searches (e.g.genetic algorithms) often do much better, but the search technique is orthogonal to our work and not our focus.

    What was the basis for choosing the subset of programs from KDataSets? This needs to be clearly justified.

Please see response to A.2 regarding benchmark partitions.

    Minor comments:

    Speedup on X-axes made it difficult for me to interpret the results

Figure 9 reports average speedups over O3 optimization level as a % improvement whereas Figure 1 reports improvement as speedup. % improvement and speedup are not the same. We will clarify.

    Why is the Oracle-RM performing worse than other methods in some instances (e.g., qsort)?

    Section 4 has only one subsection. Consider revising.

    page 10, line 1: correlate -> correlated

We will make all the minor corrections you suggested. Thanks for the feedback.

#Review#107C

    Overall merit
    -------------
    2. Weak reject

    Reviewer expertise
    ------------------
    3. Knowledgeable

    Paper summary ------------- 
    The paper presents a new online iterative optimization strategy. The main new contributions of this work
    are a new work efficiency metrics that makes it possible to compare different program versions (compiled with different optimization
    sessions) across different inputs. The paper also includes an optimization of basic block profiling that is based on the idea of
    removing profiling probes if the amount of work they contribute is low but their execution count might be high.

    Comments for author
    -------------------

    I found the new efficiency metric interesting – making it possible to compare different versions of a program across different inputs
    is a novel contribution. However, some of the discussion around the efficiency metric in the paper was confusing. For example, Section
    2 states that execution time is not correlated to speedup – how is speedup exactly computed here – isn’t it based on execution time?
    Section 3 then introduces the work metrics and shows in Figure 2 that it is correlated to execution time. Earlier we established that
    execution is not correlated with speedup so does that mean the new work metric also is not correlated with speedup? The new work is
    however, positioned to approximate speedup – so these analyses are confusing – it may just be a result of the description in the paper
    but the authors should clarify.

C.1:
Thank you for the feedback, we will clarify this confusion in the paper.

In section 2, we mean that execution time alone does not correlate to the actual speedup, when we compare the program execution time obtained from different inputs, as different inputs would entail different amounts of work. In section 3, we show that work metric obtain from a given input indeed correlates with the unoptimized execution time for *the same input*, regardless of the optimization used to compile the program.

    My main problem with the paper is that it introduces a new online iterative optimization strategy without properly evaluating it. The
    paper offers Figure 9 to show that the final optimization sequence chosen by the new online strategy reaches about 60% of the
    performance that can be achieved with offline iterative optimization.

    This evaluation doesn’t show all that much, it shows me that the proposed new strategy “has the potential to produce good results”.
    This is not enough I also need to know what it takes (i.e., the cost ) of using online optimization versus starting out with a fixed
    version that was produced with less optimally selected input in an offline iterative optimization.

    Specially, to compare online versus offline iterative optimization, I need to know the accumulated performance penalty from running
    suboptimal versions in the online setting, until I find the final optimization strategy. The authors don’t want to get into evaluating
    different search strategies – that is fine – but one could always pick a fixed search strategy that is used in both online and offline
    and then compare. To assess the potential of an online techniques I need to understand how quickly it converges or in other words how
    long the iterative optimization process continues. There is also the questions as to whether there are any advantages over offline
    iterative optimization, once the online iterative optimization sequence concludes – the actual inputs may continue to evolve .

C.2
Regarding the cost of online iterative compilation, we have focused our analysis on the profiling overhead required for guiding the search, but not the time spent on searching for the optimal version. We target scenarios where the program to be optimized will be repeatedly executed, so that the search overhead can be amortised by the improved performance. There are works on using active learning and genetic search to reduce the search overhead, which are orthogonal to our work.

Nonetheless, we will provide results on the search overhead, and refer to those complementary techniques.

    In any realistic setting, one would want to start out with the best statically compiled version  - can this strategy provide any benefits over that scenario?

C.3
Indeed it would interesting to investigate a search scheme that leverages from starting with the best statically compiled version of the program. However, unfortunately we do not have data for this experiment, as we assumed that -O3 would be the initial baseline. We will investigate it in the future.

    We need a comparison to offline iterative optimization to understand whether this approach is viable.

C.4
Prior works on offline iterative compilation are all based on replays or requires the profiling to be performed on the same input. These assumptions are not suitable for the scenarios we target (see also response to B.1). We will, however, provide a comparison against offline iterative compilation.

    The limited evaluation included is further limited by the fact that the benchmarks programs are very small and not representative of realistic datacenter applications.

C.5 Please refer to response B.6 regarding the benchmarks we used.

#Review#107D

    Overall merit
    -------------
    2. Weak reject

    Reviewer expertise
    ------------------
    3. Knowledgeable

    Paper summary
    -------------
    Paper proposes online iterative compilation technique, which is composed of
    multiple components.
    1. Work metric to approximate performance.
    2. Relaxation methods that reduces runtime profiling overhead but penalizes
    estimation accuracy.

    Comments for author

    It is difficult to find what authors are newly contributing on top of
    existing works.
    Authors use work efficiency metric to guide optimization selection. While
    authors claim that they introduce a new "work" metric, this metric does not seem
    to be critical because "work efficiency" metric is what authors actually use to
    guide the optimization selection.
    e.g. if authors defined work efficiency as 1 / t, where t is execution time, it
    would have resulted in same conclusions.
    Although not very clear from the paper, if authors compute "work" metric
    differently for every optimization, then authors are defining the metric wrong,
    as authors highlight that work metric is defined as the amount of time taken by
    the "unoptimized" program.

D.1
Indeed, the work metric does not depend on optimizations but it does depend on the input. More input to process or more complex processing result in more work. Each measurement is done using not only a different program version but also a different input, so the work is different for each measurement. Our work efficiency metric takes this work and the runtime to calculate the amount of work done in a unit of time. Since both of them change, the work efficiency is not similar to the inverse of the runtime.

    The methods described for reducing instrument (or profiling) overhead at runtime
    are also similar to the prior works, and does not seem to contribute any
    novelty.
    In fact, many of the implementation details in the paper could just cite the
    previous work, and authors should have focused on the new parts.

D.2
We will make the contributions more explicit in section 5. As we highlight in the introduction, the main contributions regarding the profiling technique are the two relaxation strategies. Apart from the relaxation strategies, however, the optimal work profiling is not a direct implementation of the classic block-frequency profiling, thus we judged important to also describe it in more detail.

    Another problem is that there are many inconsistencies on authors' claims. For
    example, authors argue that online real inputs are important because they could
    be very different from small set of inputs given by programmer.
    However, some of the methods used by authors' do require those pre-knowledges.
    For example, whole-program relaxation requires block-frequency profiling. Since
    the instrumentation itself is done once at offline, block-frequency profiling
    must be also done before online deployment. This profiling will be again
    dependent on those small set of inputs given by programmer, which contradicts
    what authors were aiming for.

D.3
It is possible to use static heuristics for estimating block frequencies. In fact, these static heuristics are used in the results for the worst-case relaxation, where no prior profiling is required. For the whole-program profiling, however, we want to show the results for the ideal case by using actual block-frequency profiling. These block-frequency profiling can be integrated with the online framework, for example, by interleaving a sequence of executions with one that only performs the block-frequency profiling.

    Minor typos, etc:
    - Page 6, second to last paragraph:
      into the DAG An example of this is
      -> period missing between "DAG" and "An".

    - Page 9, last paragraph:
      user, by it also
      -> "by" should be "but"

### 2017-10-21 rebuttal.md
minor changes

@@
 #Review#107A
-A.1 The metric is in some ways coarse, but has a strong correlation coefficient to runtime (0.99) and mean absolute error of 3.5% (Fig3). It is enough for iterative compilation (Fig9).
+A.1 The metric is in some ways coarse, but has a strong correlation coefficient to runtime (0.99) and a mean absolute error of 3.5% (Fig3). It is enough for iterative compilation (Fig9).
 
 A.2 We used all benchmarks in KDataSets. Some were for the cost model, the others for evaluating our methodology. This standard model
-training/evaluation practice. We will make this clear in the paper.
+training/evaluation practice to evaluate the generality of the model. We will make this clear in the paper.
 
@@
 No one has solved the general, automatic case without developer involvement.
-We are the first in the world to do this.
-Many papers have suggested it is possible.
-None have done it and their suggestions would not work.
+We are the first to do this.
 We will clarify this.
 
-Yes, there has been work on input sensitivity.
+B.3 Yes, there has been work on input sensitivity.
 We agree that programs' behavior varies over inputs.
@@
 
-IPC has been suggested by other papers as a possible approach to achieve online iterative compilation.
+B.4 IPC has been suggested by other papers as a possible approach to achieve online iterative compilation.
 None has tried it.
@@
 Filtering NOPs would not fix the other cases.
-IPC is only good for architectural optimization.
-Our observation is back up by paper[17].
+IPC is only good for architectural optimization where the program is invariant.
+Our observation is back up by paper [17].
 We will clarify this and change the example from NOPs.
 
-Our work is not about the *overhead of replay executions*. It is about online iterative compilation where inputs cannot be repeated.
+B.5 Our work is not about the *overhead of replay executions*. It is about online iterative compilation where inputs cannot be repeated. See also B.1 and B.2.
 
-By side effects, we mean changes to external states that may not be repeated (e.g., modifying a database, file system, network activity, interacting with external third-party services).
+B.6 By side effects, we mean changes to external states that may not be repeated (e.g., modifying a database, file system, network activity, interacting with external third-party services).
 Mitigating these effects would be very cumbersome, if it is possible at all.
@@
 
-Regarding the distinction between online/offline from Fig4, please see B.1,B.2.
+B.7 Regarding the distinction between online/offline from Fig4, please see B.1 and B.2.
 
-Regarding the shipping of the program to the iterative compilation mechanism: We only mean that the users get the program from somewhere and run it inside the iterative compilation framework.
+B.8 Regarding the shipping of the program to the iterative compilation mechanism: We only mean that the users get the program from somewhere and run it inside the iterative compilation framework.
 The code needs to be re-compilable, so LLVM bitcode is suitable.
@@
 
-Our work is not meant only for data centers. It is generally applicable. We will clarify.
+B.9 Our work is not meant only for data centers. It is generally applicable. We will clarify.
 
-There do not need to be many alternate binaries stored. It depends on the search strategy. In the smallest case, only one binary exists at any time. Many are produced over time. We will clarify.
+B.10 There do not need to be many alternate binaries stored. It depends on the search strategy. In the smallest case, only one binary exists at any time. Many are produced over time. We will clarify.
 

### 2017-10-21 rebuttal.md
rebuttal

--- 2017-10-21 rebuttal.md
@@
 
-    For more on autotuning on the data center, see this paper from CGO16
+The AutoFDO paper is not about iterative compilation.
+It repeatedly profiles and feeds profiles to an FDO compiler, but this is not iterative compilation search (despite a section with that title) -- theirs has no search for performance efficiency.
+We will discuss in related work.
 
-    Dehao Chen, David Xinliang Li, and Tipp Moseley. 2016. AutoFDO: automatic feedback-directed optimization for warehouse-scale applications.
+Yes, our evaluation was not done in a data center.
+However, it is entirely general and should work from mobile systems to data centers.
+We did not mean to suggest that we had evaluated it in all domains.
+We will clarify.
 
-    The authors claim the proposed strategy is suitable for data center environments. But the evaluation is done on a single Intel system
-    with sequential applications.
-
-B.6
-We agree that the benchmark suite used does not represent typical data center workloads.
-Still, we believe that our approach is perfectly suited to data center scenarios and there is no reason it would not work.
-In a future work, we plan to evaluate our mechanism under such an environment to better support our claim.
-
-    The performance improvement numbers are modest. But this is probably because only 500 sequences are selected out of a possible 200
-    factorial sequences in LLVM.
-Yes, the performance improvements are modest.
-We choose a simple, repeatable, random search to allow easy comparison with other metrics.
+Yes, the performance improvements are modest because of the small, random search.
+We choose it to allow easy comparison with other metrics.
 Better searches (e.g.genetic algorithms) often do much better, but the search technique is orthogonal to our work and not our focus.
-
-    What was the basis for choosing the subset of programs from KDataSets? This needs to be clearly justified.
 
@@
 
-    Minor comments:
-
-    Speedup on X-axes made it difficult for me to interpret the results
-
 Figure 9 reports average speedups over O3 optimization level as a % improvement whereas Figure 1 reports improvement as speedup. % improvement and speedup are not the same. We will clarify.
 
-    Why is the Oracle-RM performing worse than other methods in some instances (e.g., qsort)?
-
-    Section 4 has only one subsection. Consider revising.
-
-    page 10, line 1: correlate -> correlated
-
-We will make all the minor corrections you suggested. Thanks for the feedback.
+Regarding Oracle-RM performing worse than other methods, these measurements are within the noise floor.  We will run more times increase the run count to get statistically significant comparisons.
 

### 2017-10-21 rebuttal.md

@@
 
-B.1.Offline iterative compilation is done ahead of time in the development environment.
+B.1 Offline iterative compilation is done ahead of time in the development environment.
 Online iterative compilation is done *live* on users' machines.
@@
 
-B.2.Online iterative compilation has **never** been done before in the general case.
+B.2 Online iterative compilation has **never** been done before in the general case.
 Some work has done online optimization where the developer provides a work or throughput metric (e.g.
@@
 
-Our work is not meant only for data centers. It is generally applicable. We will clarify.
+B.9 Our work is not meant only for data centers. It is generally applicable. We will clarify.
 

### 2017-10-21 rebuttal.md
Rebuttal

@@
-#Review#107A
+# Review#107A
 A.1 The metric is in some ways coarse, but has a strong correlation coefficient to runtime (0.99) and a mean absolute error of 3.5% (Fig3). It is enough for iterative compilation (Fig9).
@@
 
-#Review#107B
+# Review#107B
 
@@
 
-#Review#107C
+# Review#107C
 
-    Overall merit
-    -------------
-    2. Weak reject
+C.1 Thank you for the feedback, we will clarify this confusion in the paper.
 
-    Reviewer expertise
-    ------------------
-    3. Knowledgeable
+In Section 2 we mean that execution time does not correlate to efficiency unless all inputs have the same work. We will reword.
 
-    Paper summary ------------- 
-    The paper presents a new online iterative optimization strategy. The main new contributions of this work
-    are a new work efficiency metrics that makes it possible to compare different program versions (compiled with different optimization
-    sessions) across different inputs. The paper also includes an optimization of basic block profiling that is based on the idea of
-    removing profiling probes if the amount of work they contribute is low but their execution count might be high.
+In Section 3 we show the work metric correlates with unoptimized execution time for *the same input*, regardless of optimization. We will clarify.
 
-    Comments for author
-    -------------------
+We think you are asking for a "total lifetime cost", i.e.
+including costs of poorly performing program versions in the search.
+This is a very good idea which we did not think of.
+However, random search is ill suited for this, a well designed GA would be better.
+We were not considering the search itself as interesting.
+We will show total lifetime cost, and in the future apply better search.
 
-    I found the new efficiency metric interesting – making it possible to compare different versions of a program across different inputs
-    is a novel contribution. However, some of the discussion around the efficiency metric in the paper was confusing. For example, Section
-    2 states that execution time is not correlated to speedup – how is speedup exactly computed here – isn’t it based on execution time?
-    Section 3 then introduces the work metrics and shows in Figure 2 that it is correlated to execution time. Earlier we established that
-    execution is not correlated with speedup so does that mean the new work metric also is not correlated with speedup? The new work is
-    however, positioned to approximate speedup – so these analyses are confusing – it may just be a result of the description in the paper
-    but the authors should clarify.
+Starting with the best statically compiled version is a very good idea. We don't know the quantitive benefit. We will consider this in future work when improving the search technique.
 
-C.1:
-Thank you for the feedback, we will clarify this confusion in the paper.
+We do compare offline iterative compilation, OracleRM, it runs each input twice per optimization. Our approach is gets close to this.
 
-In section 2, we mean that execution time alone does not correlate to the actual speedup, when we compare the program execution time obtained from different inputs, as different inputs would entail different amounts of work.
-In section 3, we show that work metric obtain from a given input indeed correlates with the unoptimized execution time for *the same input*, regardless of the optimization used to compile the program.
+Regarding benchmarks not being for data centers, please see B.9.
 
-    My main problem with the paper is that it introduces a new online iterative optimization strategy without properly evaluating it. The
-    paper offers Figure 9 to show that the final optimization sequence chosen by the new online strategy reaches about 60% of the
-    performance that can be achieved with offline iterative optimization.
+# Review#107D
 
-    This evaluation doesn’t show all that much, it shows me that the proposed new strategy “has the potential to produce good results”.
-    This is not enough I also need to know what it takes (i.e., the cost ) of using online optimization versus starting out with a fixed
-    version that was produced with less optimally selected input in an offline iterative optimization.
+For novelty, please see B.2.
 
-    Specially, to compare online versus offline iterative optimization, I need to know the accumulated performance penalty from running
-    suboptimal versions in the online setting, until I find the final optimization strategy. The authors don’t want to get into evaluating
-    different search strategies – that is fine – but one could always pick a fixed search strategy that is used in both online and offline
-    and then compare. To assess the potential of an online techniques I need to understand how quickly it converges or in other words how
-    long the iterative optimization process continues. There is also the questions as to whether there are any advantages over offline
-    iterative optimization, once the online iterative optimization sequence concludes – the actual inputs may continue to evolve .
+Yes, work efficiency is the metric. It is work/time. Time is easy, we need a work metric.
 
-C.2
-Regarding the cost of online iterative compilation, we have focused our analysis on the profiling overhead required for guiding the search, but not the time spent on searching for the optimal version.
-We target scenarios where the program to be optimized will be repeatedly executed, so that the search overhead can be amortised by the improved performance.
-There are works on using active learning and genetic search to reduce the search overhead, which are orthogonal to our work.
+If work efficiency were $1/t$: consider comparing efficiency of two sorting algorithms this way.
+If program A sorts a 2 element list and program B sorts a million, $t_A<t_B$ but we do not know which is more efficient.
+We will clarify in the paper.
 
-Nonetheless, we will provide results on the search overhead, and refer to those complementary techniques.
+Our work metric does not change because of the optimization. We will clarify.
 
-    In any realistic setting, one would want to start out with the best statically compiled version  - can this strategy provide any benefits over that scenario?
+The only similarity in our work is to classic block-frequency profiling. We extend with relaxation which is entirely novel. We will rebalance the writing.
 
-C.3
-Indeed it would interesting to investigate a search scheme that leverages from starting with the best statically compiled version of the program.
-However, unfortunately we do not have data for this experiment, as we assumed that -O3 would be the initial baseline.
-We will investigate it in the future.
+The profiling that drives the whole program relaxation is done *online* - by profiling the first few inputs (not programmer provided).
+It could be periodically updated.
+We regard this as somewhat orthogonal.
+No pre-knowledge is required.
+We will clarify.
 
-    We need a comparison to offline iterative optimization to understand whether this approach is viable.
-
-C.4
-Prior works on offline iterative compilation are all based on replays or requires the profiling to be performed on the same input.
-These assumptions are not suitable for the scenarios we target (see also response to B.1).
-We will, however, provide a comparison against offline iterative compilation.
-
-    The limited evaluation included is further limited by the fact that the benchmarks programs are very small and not representative of realistic datacenter applications.
-
-C.5 Please refer to response B.6 regarding the benchmarks we used.
-
-#Review#107D
-
-    Overall merit
-    -------------
-    2. Weak reject
-
-    Reviewer expertise
-    ------------------
-    3. Knowledgeable
-
-    Paper summary
-    -------------
-    Paper proposes online iterative compilation technique, which is composed of
-    multiple components.
-    1. Work metric to approximate performance.
-    2. Relaxation methods that reduces runtime profiling overhead but penalizes
-    estimation accuracy.
-
-    Comments for author
-
-    It is difficult to find what authors are newly contributing on top of
-    existing works.
-    Authors use work efficiency metric to guide optimization selection. While
-    authors claim that they introduce a new "work" metric, this metric does not seem
-    to be critical because "work efficiency" metric is what authors actually use to
-    guide the optimization selection.
-    e.g. if authors defined work efficiency as 1 / t, where t is execution time, it
-    would have resulted in same conclusions.
-    Although not very clear from the paper, if authors compute "work" metric
-    differently for every optimization, then authors are defining the metric wrong,
-    as authors highlight that work metric is defined as the amount of time taken by
-    the "unoptimized" program.
-
-D.1
-Indeed, the work metric does not depend on optimizations but it does depend on the input.
-More input to process or more complex processing result in more work.
-Each measurement is done using not only a different program version but also a different input, so the work is different for each measurement.
-Our work efficiency metric takes this work and the runtime to calculate the amount of work done in a unit of time.
-Since both of them change, the work efficiency is not similar to the inverse of the runtime.
-
-    The methods described for reducing instrument (or profiling) overhead at runtime
-    are also similar to the prior works, and does not seem to contribute any
-    novelty.
-    In fact, many of the implementation details in the paper could just cite the
-    previous work, and authors should have focused on the new parts.
-
-D.2
-We will make the contributions more explicit in section 5.
-As we highlight in the introduction, the main contributions regarding the profiling technique are the two relaxation strategies.
-Apart from the relaxation strategies, however, the optimal work profiling is not a direct implementation of the classic block-frequency profiling, thus we judged important to also describe it in more detail.
-
-    Another problem is that there are many inconsistencies on authors' claims. For
-    example, authors argue that online real inputs are important because they could
-    be very different from small set of inputs given by programmer.
-    However, some of the methods used by authors' do require those pre-knowledges.
-    For example, whole-program relaxation requires block-frequency profiling. Since
-    the instrumentation itself is done once at offline, block-frequency profiling
-    must be also done before online deployment. This profiling will be again
-    dependent on those small set of inputs given by programmer, which contradicts
-    what authors were aiming for.
-
-D.3
-It is possible to use static heuristics for estimating block frequencies.
-In fact, these static heuristics are used in the results for the worst-case relaxation, where no prior profiling is required.
-For the whole-program profiling, however, we want to show the results for the ideal case by using actual block-frequency profiling.
-These block-frequency profiling can be integrated with the online framework, for example, by interleaving a sequence of executions with one that only performs the block-frequency profiling.
-
-    Minor typos, etc:
-    - Page 6, second to last paragraph:
-      into the DAG An example of this is
-      -> period missing between "DAG" and "An".
-
-    - Page 9, last paragraph:
-      user, by it also
-      -> "by" should be "but"
+Thank for catching the typos.

### 2017-10-21 rebuttal.md
rebuttal

@@
 Input sensitive iterative compilation (via multi-versioning etc) is entirely orthogonal to online/offline.
-Offline requires the developer to choose representative inputs, online does not.
 Online input sensitive iterative compilation will match the users's inputs better.
-Until now, online has been impossible.
 We will clarify this.
 
-The papers, Fursin etal., Ding etal., and Chen etal., are about offline iterative compilation, their approaches do not work in the online case.
+The papers, Fursin etal., Ding etal., and Chen etal., use offline iterative compilation, their approaches do not work in the online case.
 
-That runtimes of programs with different input sets are not directly comparable is not a problem for **offline** iterative compilation, it breaks online iterative compilation.
+That runtimes of programs with different input sets are not directly comparable is not a problem for **offline** iterative compilation, but it breaks online iterative compilation.
 Throughput could be used, but there is no such general metric.
@@
 IPC is only good for architectural optimization where the program is invariant.
-Our observation is back up by paper [17].
 We will clarify this and change the example from NOPs.
@@
 Yes, the performance improvements are modest because of the small, random search.
-We choose it to allow easy comparison with other metrics.
+We chose it to allow easy comparison with other metrics.
 Better searches (e.g.genetic algorithms) often do much better, but the search technique is orthogonal to our work and not our focus.
@@
 
-Regarding Oracle-RM performing worse than other methods, these measurements are within the noise floor.  We will run more times increase the run count to get statistically significant comparisons.
+Regarding Oracle-RM performing worse than other methods, these measurements are within the noise floor.  We will increase the run count to get statistically significant comparisons.
 
@@
 This is a very good idea which we did not think of.
-However, random search is ill suited for this, a well designed GA would be better.
+However, random search is ill-suited for this, a well designed GA would be better.
 We were not considering the search itself as interesting.
@@
 
+**HJL rewrite - lists of different sizes**
 If work efficiency were $1/t$: consider comparing efficiency of two sorting algorithms this way.

### 2017-10-21 rebuttal.md
add labels

@@
 We agree that programs' behavior varies over inputs.
-That is why online iterative compilation is better, it optimizes according to the inputs the user actual presents.
+That is why online iterative compilation is better, it optimizes according to the inputs the actual user presents.
 Input sensitive iterative compilation (via multi-versioning etc) is entirely orthogonal to online/offline.
@@
 
-That runtimes of programs with different input sets are not directly comparable is not a problem for **offline** iterative compilation, but it breaks online iterative compilation.
+B.4 That runtimes of programs with different input sets are not directly comparable is not a problem for **offline** iterative compilation, but it breaks online iterative compilation.
 Throughput could be used, but there is no such general metric.
@@
 
-B.4 IPC has been suggested by other papers as a possible approach to achieve online iterative compilation.
+IPC has been suggested by other papers as a possible approach to achieve online iterative compilation.
 None has tried it.
@@
 
-There do not need to be many alternate binaries stored. It depends on the search strategy. In the smallest case, only one binary exists at any time. Many are produced over time. We will clarify.
+B.10 There do not need to be many alternate binaries stored. It depends on the search strategy. In the smallest case, only one binary exists at any time. Many are produced over time. We will clarify.
 
-The AutoFDO paper is not about iterative compilation.
+B.11 The AutoFDO paper is not about iterative compilation.
 It repeatedly profiles and feeds profiles to an FDO compiler, but this is not iterative compilation search (despite a section with that title) -- theirs has no search for performance efficiency.
@@
 
-Yes, our evaluation was not done in a data center.
+B.12 Yes, our evaluation was not done in a data center.
 However, it is entirely general and should work from mobile systems to data centers.
@@
 
-Yes, the performance improvements are modest because of the small, random search.
+B.13 Yes, the performance improvements are modest because of the small, random search.
 We chose it to allow easy comparison with other metrics.
@@
 
-Please see response to A.2 regarding benchmark partitions.
+B.14 Please see response to A.2 regarding benchmark partitions.
 
-Figure 9 reports average speedups over O3 optimization level as a % improvement whereas Figure 1 reports improvement as speedup. % improvement and speedup are not the same. We will clarify.
+B.15 Figure 9 reports average speedups over O3 optimization level as a % improvement whereas Figure 1 reports improvement as speedup. % improvement and speedup are not the same. We will clarify.
 
-Regarding Oracle-RM performing worse than other methods, these measurements are within the noise floor.  We will increase the run count to get statistically significant comparisons.
+B.16 Regarding Oracle-RM performing worse than other methods, these measurements are within the noise floor.  We will increase the run count to get statistically significant comparisons.
 
@@
 
-We think you are asking for a "total lifetime cost", i.e.
+C.2 We think you are asking for a "total lifetime cost", i.e.
 including costs of poorly performing program versions in the search.
@@
 
-Starting with the best statically compiled version is a very good idea. We don't know the quantitive benefit. We will consider this in future work when improving the search technique.
+C.3 Starting with the best statically compiled version is a very good idea. We don't know the quantitive benefit. We will consider this in future work when improving the search technique.
 
-We do compare offline iterative compilation, OracleRM, it runs each input twice per optimization. Our approach is gets close to this.
+C.4 We do compare offline iterative compilation, OracleRM, it runs each input twice per optimization. Our approach is gets close to this.
 
-Regarding benchmarks not being for data centers, please see B.9.
+C.5 Regarding benchmarks not being for data centers, please see B.9.
 
@@
 
-For novelty, please see B.2.
+D.1 For novelty, please see B.2.
 
-Yes, work efficiency is the metric. It is work/time. Time is easy, we need a work metric.
+D.2 Yes, work efficiency is the metric. It is work/time. Time is easy, we need a work metric.
 
@@
 
-The only similarity in our work is to classic block-frequency profiling. We extend with relaxation which is entirely novel. We will rebalance the writing.
+D.3 The only similarity in our work is to classic block-frequency profiling. We extend with relaxation which is entirely novel. We will rebalance the writing.
 
-The profiling that drives the whole program relaxation is done *online* - by profiling the first few inputs (not programmer provided).
+D.4 The profiling that drives the whole program relaxation is done *online* - by profiling the first few inputs (not programmer provided).
 It could be periodically updated.
@@
 
-Thank for catching the typos.
+D.5 Thank for catching the typos.

### 2017-10-21 rebuttal.md
Rebuttal shrunk

--- 2017-10-21 rebuttal.md
@@
 # Review#107A
-A.1 The metric is in some ways coarse, but has a strong correlation coefficient to runtime (0.99) and a mean absolute error of 3.5% (Fig3). It is enough for iterative compilation (Fig9).
+A.1.The metric is coarse, but has a strong correlation coefficient to runtime (0.99) and a mean absolute error of 3.5% (Figure.3). It is enough for iterative compilation (Figure.9).
 
-A.2 We used all benchmarks in KDataSets. Some were for the cost model, the others for evaluating our methodology. This standard model
-training/evaluation practice to evaluate the generality of the model. We will make this clear in the paper.
+A.2.We used all KDataSets benchmarks: some are for the cost model, others for methodology evaluation. This is standard model
+training/evaluation practice to evaluate the generality of the model. We will clarify.
 
-A.3 We will publish the cost model.
+A.3.We will publish the cost model.
 
@@
 
-The paper is **not** about *reducing overhead in iterative compilation*. It is about **online** iterative compilation. We will clarify this in the paper.
+The paper is **not** about *reducing overhead in iterative compilation*. It is about **online** iterative compilation. We will clarify in the paper.
 
-B.1 Offline iterative compilation is done ahead of time in the development environment.
+B.1.Offline iterative compilation happens ahead-of-time in the development environment.
 Online iterative compilation is done *live* on users' machines.
 Offline requires developers to choose representative inputs, online does not.
-Instead, online optimization targets the actual ways users use the program.
-Because it is live on users' machines, inputs cannot be run twice.
-We will clarify this.
+Instead, online optimization targets the actual user inputs.
+Being live on users' machines, inputs cannot run twice.
+We will clarify.
 
-B.2 Online iterative compilation has **never** been done before in the general case.
+B.2.Online iterative compilation has **never** been done before in the general case.
 Some work has done online optimization where the developer provides a work or throughput metric (e.g.
@@
 We are the first to do this.
-We will clarify this.
+We will clarify.
 
-B.3 Yes, there has been work on input sensitivity.
-We agree that programs' behavior varies over inputs.
-That is why online iterative compilation is better, it optimizes according to the inputs the user actual presents.
-Input sensitive iterative compilation (via multi-versioning etc) is entirely orthogonal to online/offline.
-Online input sensitive iterative compilation will match the users's inputs better.
-We will clarify this.
+B.3.Yes, there is work on input sensitivity.
+We agree, programs' behavior varies over inputs.
+That is why online iterative compilation is better, it optimizes according to actual user inputs.
+Input-sensitive iterative compilation (via multi-versioning etc) is entirely orthogonal to online/offline.
+Online input-sensitive iterative compilation will match the users's inputs better.
+We will clarify.
 
-The papers, Fursin etal., Ding etal., and Chen etal., use offline iterative compilation, their approaches do not work in the online case.
+The papers, Fursin etal.,Ding etal.,and Chen etal., use offline iterative compilation, they do not handle the online case.
 
-That runtimes of programs with different input sets are not directly comparable is not a problem for **offline** iterative compilation, but it breaks online iterative compilation.
-Throughput could be used, but there is no such general metric.
-Developers would have to provide one (e.g.
-#requests/second - and ensure each request comprises the same work).
-Our work is the first general and automatic approach.
+B.4.That runtimes on different inputs are incomparable is okay for **offline** iterative compilation, but breaks online iterative compilation.
+Throughput could be used, but developers would have to define it (e.g.
+#requests/second - and ensure work/request uniformity).
+Ours is the first general, automatic approach.
 
-B.4 IPC has been suggested by other papers as a possible approach to achieve online iterative compilation.
+Other papers suggested IPC for online iterative compilation.
 None has tried it.
-We have shown that IPC does not work (Fig2).
-Our example with NOPs is only the most egregious case, to show logically why IPC is no good.
-In fact, it is not the cause of IPC's poor correlation in Fig2.
+We show that IPC fails (Figure.2).
+Our NOPs example is the most egregious case, showing why IPC is no good.
+It is not the cause of IPC's poor correlation in Figure.2.
 Trading high CPI instructions for low CPI instructions (e.g.
@@
 Filtering NOPs would not fix the other cases.
-IPC is only good for architectural optimization where the program is invariant.
-We will clarify this and change the example from NOPs.
+IPC is good for architectural optimization with invariant programs.
+We will clarify and change the example from NOPs.
 
-B.5 Our work is not about the *overhead of replay executions*. It is about online iterative compilation where inputs cannot be repeated. See also B.1 and B.2.
+B.5.Our work is not about the *overhead of replay executions*. It is about online iterative compilation where inputs are unrepeatable. See B.1-B.2.
 
-B.6 By side effects, we mean changes to external states that may not be repeated (e.g., modifying a database, file system, network activity, interacting with external third-party services).
-Mitigating these effects would be very cumbersome, if it is possible at all.
-It is different from the scenario faced by ATLAS and not about caches or noisy environments.
+B.6.By side effects, we mean unrepeatable changes to external states (e.g., database modification, file modification, network activity, calling external third-party services).
+Mitigating these would be hard or impossible.
+It is different from ATLAS' scenario and not about caches or noisy environments.
 We will clarify.
 
-B.7 Regarding the distinction between online/offline from Fig4, please see B.1 and B.2.
+B.7.Regarding the distinction between online/offline from Figure.4, please see B.1-B.2.
 
-B.8 Regarding the shipping of the program to the iterative compilation mechanism: We only mean that the users get the program from somewhere and run it inside the iterative compilation framework.
-The code needs to be re-compilable, so LLVM bitcode is suitable.
-Application source code is not required.
+B.8.Regarding shipping the program to the iterative compilation mechanism: We only mean that program is run inside the iterative compilation framework.
 It is poorly worded, we will fix.
+The code must be re-compilable, so LLVM bitcode is suitable.
+Application source-code is not required.
 
-B.9 Our work is not meant only for data centers. It is generally applicable. We will clarify.
+B.9.Our work is not data-centers specific. It is generally applicable. We will clarify.
 
-There do not need to be many alternate binaries stored. It depends on the search strategy. In the smallest case, only one binary exists at any time. Many are produced over time. We will clarify.
+B.10.Storing many alternate binaries is unnecessary. It is search strategy dependent. In ours, one binary exists at any time. We will clarify.
 
-The AutoFDO paper is not about iterative compilation.
-It repeatedly profiles and feeds profiles to an FDO compiler, but this is not iterative compilation search (despite a section with that title) -- theirs has no search for performance efficiency.
+B.11.AutoFDO is not about iterative compilation.
+It repeatedly feeds profiles to a FDO compiler.
+That is not iterative compilation search (despite such a section title)--they have no search.
 We will discuss in related work.
 
-Yes, our evaluation was not done in a data center.
-However, it is entirely general and should work from mobile systems to data centers.
-We did not mean to suggest that we had evaluated it in all domains.
-We will clarify.
+B.12.Yes, we did not evaluate on data-centers. We meant it is general from mobile systems to data-centers. We did not mean to suggest that we evaluated in all domains. We will clarify.
 
-Yes, the performance improvements are modest because of the small, random search.
-We chose it to allow easy comparison with other metrics.
-Better searches (e.g.genetic algorithms) often do much better, but the search technique is orthogonal to our work and not our focus.
+B.13.Yes, performance improvements are modest because of the small, random search.
+We chose it for easy comparison with other metrics.
+Better searches (e.g.
+GAs) do much better.
+This is orthogonal to our work and not our focus.
 
-Please see response to A.2 regarding benchmark partitions.
+B.14.See A.2 regarding benchmark partitions.
 
-Figure 9 reports average speedups over O3 optimization level as a % improvement whereas Figure 1 reports improvement as speedup. % improvement and speedup are not the same. We will clarify.
+B.15.Figure.9 reports average speedups over -O3 as a %-improvement whereas Figure.1 reports improvement as speedup. %-improvement and speedup are not the same. We will clarify.
 
-Regarding Oracle-RM performing worse than other methods, these measurements are within the noise floor.  We will increase the run count to get statistically significant comparisons.
+B.16.Regarding Oracle-RM performing worse, these measurements are within the noise floor.  We will run more for statistically significant comparisons.
 
@@
 
-C.1 Thank you for the feedback, we will clarify this confusion in the paper.
+C.1.Thank you for the feedback, we will clarify confusion in the paper.
 
@@
 
-We think you are asking for a "total lifetime cost", i.e.
-including costs of poorly performing program versions in the search.
-This is a very good idea which we did not think of.
-However, random search is ill-suited for this, a well designed GA would be better.
-We were not considering the search itself as interesting.
-We will show total lifetime cost, and in the future apply better search.
+C.2.Regarding "total lifetime cost", i.e.
+costs of poorly performing program versions in the search.
+This is a good idea, thank-you.
+However, random search is ill-suited for this, unlike a well designed GA.
+We were not considering the search interesting.
+We will show total lifetime cost, and in future apply better search.
 
-Starting with the best statically compiled version is a very good idea. We don't know the quantitive benefit. We will consider this in future work when improving the search technique.
+C.3.Starting with the best statically compiled version is a good idea. We don't know the quantitive benefit. We will consider this in future work when improving the search.
 
-We do compare offline iterative compilation, OracleRM, it runs each input twice per optimization. Our approach is gets close to this.
+C.4.We do compare offline iterative compilation; OracleRM runs each input twice per optimization. Our online approach gets close.
 
-Regarding benchmarks not being for data centers, please see B.9.
+C.5.Regarding benchmarks for data-centers, please see B.9-B.12.
 
@@
 
-For novelty, please see B.2.
+D.1.For novelty, please see B.2.
 
-Yes, work efficiency is the metric. It is work/time. Time is easy, we need a work metric.
+D.2.Yes, work efficiency is the metric. It is work/time. Time is easy, we need a work metric.
 
-**HJL rewrite - lists of different sizes**
-If work efficiency were $1/t$: consider comparing efficiency of two sorting algorithms this way.
-If program A sorts a 2 element list and program B sorts a million, $t_A<t_B$ but we do not know which is more efficient.
-We will clarify in the paper.
+If work efficiency were $1/t$: consider comparing efficiency of sorting algorithms this way.If programs, A and B sort lists of different sizes, $t_A<t_B$ does not say which is more efficient.
+We will clarify.
 
@@
 
-The only similarity in our work is to classic block-frequency profiling. We extend with relaxation which is entirely novel. We will rebalance the writing.
+D.3.The only similarity in our work is to classic block-frequency profiling. We extend with relaxation which is entirely novel. We will rebalance the writing.
 
-The profiling that drives the whole program relaxation is done *online* - by profiling the first few inputs (not programmer provided).
+D.4.The profiling that drives the whole program relaxation is done *online* - by profiling the first few inputs (not programmer provided).
 It could be periodically updated.
@@
 
-Thank for catching the typos.
+D.5.Thanks for typos.

### 2017-10-22 rebuttal.md
minor changes

@@
 B.2.Online iterative compilation has **never** been done before in the general case.
-Some work has done online optimization where the developer provides a work or throughput metric (e.g.
-#requests/second assuming each request does the same work).
-No one has solved the general, automatic case without developer involvement.
+Some work has done online autotuning/optimization where the developer provides a work or throughput metric (e.g.
+#requests/second assuming all requests have equal amount of work).
+No one has solved the general case without developer involvement.
 We are the first to do this.
@@
 That is why online iterative compilation is better, it optimizes according to actual user inputs.
-Input-sensitive iterative compilation (via multi-versioning etc) is entirely orthogonal to online/offline.
+Input-sensitive iterative compilation (via multi-versioning, etc) is entirely orthogonal to online/offline.
 Online input-sensitive iterative compilation will match the users's inputs better.
@@
 
-The papers, Fursin etal.,Ding etal.,and Chen etal., use offline iterative compilation, they do not handle the online case.
+The papers, Fursin etal.,Ding etal.,and Chen etal., use offline strategies, they do not handle the online case.
 
@@
 
-Other papers suggested IPC for online iterative compilation.
-None has tried it.
+Other papers suggested IPC for similar problems.
+None has tried it for online iterative compilation.
 We show that IPC fails (Figure.2).
@@
 
-B.6.By side effects, we mean unrepeatable changes to external states (e.g., database modification, file modification, network activity, calling external third-party services).
+B.6.By side effects, we mean unrepeatable changes to external states (e.g., files or database modification, network activity, calling external third-party services).
 Mitigating these would be hard or impossible.
@@
 
-B.15.Figure.9 reports average speedups over -O3 as a %-improvement whereas Figure.1 reports improvement as speedup. %-improvement and speedup are not the same. We will clarify.
+B.15.Figure.9 reports average speedups over -O3 as a %-improvement whereas Figure.1 reports a histogram of speedup per input. We will clarify.
 
@@
 
-In Section 2 we mean that execution time does not correlate to efficiency unless all inputs have the same work. We will reword.
+In Section 2 we mean that runtime does not correlate to true speedup unless comparing inputs with equal amount of work.
 
-In Section 3 we show the work metric correlates with unoptimized execution time for *the same input*, regardless of optimization. We will clarify.
+In Section 3 we show the work metric correlates with unoptimized execution time for *the same input*, regardless of optimization.
 
@@
 
-C.4.We do compare offline iterative compilation; OracleRM runs each input twice per optimization. Our online approach gets close.
+C.4.We do compare an offline strategy; OracleRM runs each input twice per optimization. Our online approach gets close.
 
@@
 
-Our work metric does not change because of the optimization. We will clarify.
+Our work metric does not change across optimization. We will clarify.
 
-D.3.The only similarity in our work is to classic block-frequency profiling. We extend with relaxation which is entirely novel. We will rebalance the writing.
+D.3.The only similarity in our work is to classic block-frequency profiling. However, we extend with relaxation which is entirely novel. We will rebalance the writing.
 
-D.4.The profiling that drives the whole program relaxation is done *online* - by profiling the first few inputs (not programmer provided).
+D.4.The profiling that drives the whole-program relaxation is done *online* - by profiling the first few inputs (not programmer provided).
 It could be periodically updated.
+But no pre-knowledge is required.
 We regard this as somewhat orthogonal.
-No pre-knowledge is required.
 We will clarify.

### 2017-10-22 rebuttal.md
minor changes

@@
 
-The paper is **not** about *reducing overhead in iterative compilation*. It is about **online** iterative compilation. We will clarify in the paper.
+The paper is **not** about *reducing overhead in iterative compilation*. It is about enabling **online** iterative compilation. We will clarify in the paper.
 
@@
 
-If work efficiency were $1/t$: consider comparing efficiency of sorting algorithms this way.If programs, A and B sort lists of different sizes, $t_A<t_B$ does not say which is more efficient.
+If work efficiency were $1/t$: consider comparing efficiency of sorting algorithms this way.
+If programs, A and B sort lists of different sizes, $t_A<t_B$ does not say which is more efficient.
 We will clarify.

### 2017-10-22 rebuttal.md
minor fixes

@@
 Some work has done online autotuning/optimization where the developer provides a work or throughput metric (e.g.
-#requests/second assuming all requests have equal amount of work).
+#requests/second assuming all requests have an equal amount of work).
 No one has solved the general case without developer involvement.
@@
 Input-sensitive iterative compilation (via multi-versioning, etc) is entirely orthogonal to online/offline.
-Online input-sensitive iterative compilation will match the users's inputs better.
+Online input-sensitive iterative compilation will match the users' inputs better.
 We will clarify.
@@
 B.11.AutoFDO is not about iterative compilation.
-It repeatedly feeds profiles to a FDO compiler.
+It repeatedly feeds profiles to an FDO compiler.
 That is not iterative compilation search (despite such a section title)--they have no search.
@@
 
-In Section 2 we mean that runtime does not correlate to true speedup unless comparing inputs with equal amount of work.
+In Section 2 we mean that runtime does not correlate to true speedup unless comparing inputs with an equal amount of work.
 
@@
 This is a good idea, thank-you.
-However, random search is ill-suited for this, unlike a well designed GA.
+However, random search is ill-suited for this, unlike a well-designed GA.
 We were not considering the search interesting.
@@
 D.2.Yes, work efficiency is the metric. It is work/time. Time is easy, we need a work metric.
+The amount of work changes with the input but does not change across optimizations.
 
-If work efficiency were $1/t$: consider comparing efficiency of sorting algorithms this way.
-If programs, A and B sort lists of different sizes, $t_A<t_B$ does not say which is more efficient.
+If efficiency were simply $1/t$: consider comparing the efficiency of sorting algorithms this way. If programs, A and B sort lists of different sizes, $t_A<t_B$ does not say which is more efficient.
+
 We will clarify.
-
-Our work metric does not change across optimization. We will clarify.
 

### 2017-10-22 rebuttal.md
fix minor errors

@@
 
-In Section 2 we mean that runtime does not correlate to true speedup unless comparing inputs with an equal amount of work.
+In Section 2 we mean that runtime does not correlate to true speedup unless we compare inputs with an equal amount of work.
 
@@
 
-If efficiency were simply $1/t$: consider comparing the efficiency of sorting algorithms this way. If programs, A and B sort lists of different sizes, $t_A<t_B$ does not say which is more efficient.
+If efficiency was simply $1/t$: consider comparing the efficiency of sorting algorithms this way. If programs, A and B sort lists of different sizes, $t_A<t_B$ does not say which is more efficient.
 

### 2017-11-02 motivation.tex
done with related work
Alternative motivation section. Old one saved in another file

--- 2017-09-15 motivation.tex
@@
-\section{Motivation}
+% vim: set tabstop=4 shiftwidth=4 expandtab : 
+\section{Motivation \label{sec:motivation}}
 
-    In this section we first motivate online iterative compilation by showing the deleterious effects of choosing the wrong inputs for the
-    offline case. Then we go on to show the unsuitability of previously proposed program version comparison metrics.
+    % Representative inputs are important. Example.
+    % Artificial representative input sets are hard to create
+    % Representative input sets created out of real data are also hard to create
+    % Just using real data online is not enough. Efficiency metrics are needed
+    % We do not have a proper efficiency metric. Example.
+    % The paper shows how to construct one.
 
-    \begin{figure}[t!]
-        \centering
-        \includegraphics[width=0.5\textwidth]{figs/motivation-online.pdf}
-        \caption{
-            Histograms of speedup over \texttt{-O3} for each of the 1,000 inputs of \texttt{susan\_c}. The green histogram shows the
-            speedup when using the optimization sequence selected by offline \itercomp, the blue one is the speedup using the optimization
-            sequence selected by an online-like approach. \FIXME{Beware Daltonism!}
-        }
-        \label{fig:motivation-online}
-    \end{figure}
+    In this section we first motivate online iterative compilation by showing the deleterious effects of using non-representative input
+    sets and that it is hard to create representative ones. Then we go on to examine what stops us from just using real inputs online and
+    what we need to overcome the problem.
 
-    We performed offline iterative compilation on \texttt{susan\_c} to find the optimization settings that maximize the average performance
-    over five randomly chosen inputs from a set of 1,000 provided by the \textit{KDataSets} benchmark suite~\cite{chen10,chen12a}.
-    The green histogram in Figure~\ref{fig:motivation-online} shows for each of the 1,0000 inputs the speedup versus \texttt{-O3} of that
-    chosen optimization setting. The dark green line shows the average speedup for all those inputs. The five training inputs are
-    all among the top performing inputs. On the other hand, the chosen optimization does poorly on other inputs, often causing
-    a slowdown and yielding only a slight speedup of 4\% on average.
+    \paragraph{Iterative compilation needs representative inputs.} \FIXME{Fig.XXX} shows what happens when we apply iterative compilation
+    on \FIXME{susan\_c} using the wrong inputs. We selected \FIXME{50} of the 1,000 inputs of \textit{KDataSets}~\cite{chen10,chen12a} for
+    \FIXME{susan\_c} and we used them to evaluate different optimizations. The red bars show the average speedup over \texttt{-O3} achieved
+    by the optimization sequence selected by iterative compilation. On the left is the average over the \FIXME{50} inputs, on the right the
+    average over all 1,000 inputs. The selected sequence works great for the inputs we used, but it fails miserably for the rest, slowing
+    down the benchmark by \FIXME{5\%} on average.
 
-    The blue bars in Figure~\ref{fig:motivation-online} show a different approach. Here the iterative compilation proceeds by evaluating
-    each optimization on five inputs. Each input is evaluated only once during the search, but the search is provided with oracle
-    knowledge about the unoptimized runtime for the inputs, giving a perfect speedup metric. The environment of the search constantly
-    changes in the same way as for online iterative compilation. The best optimization for this case never degrades performance compared to
-    \texttt{-O3} and is on average 13\% better than the offline approach. The online case, considering truly representative inputs, leads
-    to superior performance, if the speedup caused by optimizing can be estimated.
+    The blue bars, on the other hand, show the ideal case where good optimizations are selected based on their average effect on all
+    inputs, not just a subset. Not only is the average speedup over all 1,000 inputs considerable, \FIXME{13\%} over \texttt{-O3}, but it
+    also works well for the \FIXME{50} input subset. The speedup for the subset is only \FIXME{1\%} less of what we got when we optimized
+    specifically for that subset.
 
-    \begin{figure}[t]
-        \centering
-        %\begin{tabular}{cc}
-        %\includegraphics[width=0.25\textwidth]{figs/motivation-metric-ipc.pdf} &
-        %\includegraphics[width=0.25\textwidth]{figs/motivation-metric-runtime.pdf}
-        %\end{tabular}
-        \includegraphics[width=0.5\textwidth]{figs/motivation-metric.pdf}
-        \caption{
-            Relationship between IPC and execution time \textit{vs} speedup over the unoptimized program
-            for 500 different binaries of \texttt{susan\_c}.
-            Each point represents the average over a subset of all inputs.
-            The IPC and execution time show correlation coefficients of $-0.14$ and $-0.16$, respectively.
-        }
-        \label{fig:motivation-metric}
-    \end{figure}
+    Making iterative compilation work in this example is not a matter of handling input sensitivity, we were able to find an optimization
+    sequence that works well both on average and under most inputs. It is a matter of finding a set of inputs that represents the full
+    range of inputs the application is likely to process. If we do, we can improve average performance by \FIXME{13\%}. If not, we might
+    hurt average performance instead.
 
-    Prior to this paper there were two proposed metrics to correlate with program efficiency. The first is simply to use runtime, ignoring
-    the work aspect. This functions well when the inputs all require similar effort. When that is not the case it bears little or no
-    correlation to efficiency. The second approach is \textit{instructions per cycle}~(IPC), suggested by~\citep{fursin07}. While IPC seems
-    promising in the sense that faster instruction execution is a good thing, this is belied by the simple observation that adding
-    redundant NOPs to a program will increase IPC while also slowing the program down. Figure~\ref{fig:motivation-metric} shows the
-    relationship between execution time, IPC and speedup for different optimizations of \texttt{susan\_c}, averaged across a subset of the
-    inputs. Neither IPC nor execution time correlate well, and so neither can be used for iterative compilation. We need a work efficiency
-    metric which can approximate speedup instead, using only a single execution.
+    \paragraph{Representative input sets are hard to create.} For developers to create artificial input sets, they first need to know how
+    their application is actually being used. This is not always easy or even possible. Then, they need to create data that will trigger a
+    similar behavior and this needs a deep knowledge of the application and considerable effort. This will have to be a continuous work in
+    progress, to handle changes in the way the application is used.
 
-    In the remainder of this paper we present our work efficiency metric and demonstrate its utility for iterative compilation.
+    Collecting real input data online and processing them again offline to evaluate optimizations is not a solution either. It requires
+    manual modifications to the application to save the data, more modifications to avoid side effects while reusing data, and even more
+    modifications to ensure that the application does the same work every time the same data are reused, despite changes in the rest of
+    the system state. This is significant engineering work. Even if this was not the case, many applications are just not able to save
+    data for privacy or storage overhead reasons.
+
+    \paragraph{Using real inputs online is not enough.} We could overcome these problems by just evaluating optimization sequences online,
+    while the application is being used. This gives us real data but creates a new problem. Iterative compilation finds good optimizations
+    by measuring the efficiency of the resulting binary in terms of speedup over another baseline binary. This requires processing the same
+    input in the exact same way twice, one with the optimized and one with the baseline binary, measuring the runtimes, and calculating the
+    speedup. When done online, we cannot process the same input twice, each time we process whatever the input happens to be. We need
+    another efficiency metric that only requires a single use of an optimized binary.
+
+    \paragraph{Existing efficiency metrics are not up to the task.} A application-specific ones~\cite{alameldeen06,coppa14} may work well,
+    but the developer has to choose the appropriate metric and then manually modify the application to calculate it. Existing
+    application-agnostic ones do not work. We run each of the \FIXME{32} \textit{KDataSets} benchmarks 1,000 times, each time with a
+    different input \textit{and} a different optimized binary. We selected three performance metrics that can be easily measured in a
+    single run and can potentially be used as efficiency metrics: runtime, inverse runtime, and
+    \textit{Instructions Per Cycle}~(IPC)~\citep{fursin07}.
+
+    \FIXME{Fig.XXX} shows how well these three candidates for an efficiency metric correlate with the metric we would actually like to have,
+    speedup over \texttt{-O0}. For all benchmarks, the correlation coefficient~\cite{Bishop:2006:PRM:1162264} is below \FIXME{0.20} for
+    runtime and below \FIXME{0.35} for inverse runtime. Both have obvious reasons for being inappropriate efficiency metrics. Doubling the
+    amount of data to be processed will cause the runtime to rise and the inverse runtime to fall, even if we use the same binary.
+    Efficiency has nothing to do with it. IPC also correlates poorly with performance speedup, the correlation coefficient being just
+    \FIXME{0.15}. While it is a common efficiency metric in computer architecture, it works there because it is used with invariant binaries.
+    In contrast, any compiler optimization that trades high latency instructions for multiple low latency instructions can reduce IPC while
+    improving efficiency.
+
+    No existing metric can give us an application-agnostic measure of efficiency with a single execution. The rest of the paper will show
+    how to create one and how to use it for iterative compilation.

### 2017-11-02 motivation.tex
changes in motivation

@@
 
-    \paragraph{Using real inputs online is not enough.} We could overcome these problems by just evaluating optimization sequences online,
-    while the application is being used. This gives us real data but creates a new problem. Iterative compilation finds good optimizations
-    by measuring the efficiency of the resulting binary in terms of speedup over another baseline binary. This requires processing the same
-    input in the exact same way twice, one with the optimized and one with the baseline binary, measuring the runtimes, and calculating the
-    speedup. When done online, we cannot process the same input twice, each time we process whatever the input happens to be. We need
-    another efficiency metric that only requires a single use of an optimized binary.
+    \paragraph{Using real inputs online is not enough.} We could overcome these problems by just evaluating optimization sequences online
+    while the application is being used. This gives us live data but creates a new problem. Optimizations sequences are compared based on
+    how much faster the program becomes. When all of them are tested against the same inputs, it is just a matter of comparing runtimes.
+    A shorter runtime for processing the same data means a more efficient binary. With live data, we cannot do this, each binary processes
+    whatever the input happens to be. We need another efficiency metric that only requires a single use of an optimized binary with a
+    single input.
 
-    \paragraph{Existing efficiency metrics are not up to the task.} A application-specific ones~\cite{alameldeen06,coppa14} may work well,
-    but the developer has to choose the appropriate metric and then manually modify the application to calculate it. Existing
-    application-agnostic ones do not work. We run each of the \FIXME{32} \textit{KDataSets} benchmarks 1,000 times, each time with a
-    different input \textit{and} a different optimized binary. We selected three performance metrics that can be easily measured in a
-    single run and can potentially be used as efficiency metrics: runtime, inverse runtime, and
-    \textit{Instructions Per Cycle}~(IPC)~\citep{fursin07}.
+    \paragraph{Existing efficiency metrics are not up to the task.} Application-specific efficiency metrics~\cite{alameldeen06,coppa14} may
+    work well, but the developer has to choose the appropriate metric and then manually modify the application to calculate it. Speedup
+    over a common baseline binary is a good application-agnostic metric, equating efficiency with how much faster a given input is
+    processed. It would be ideal, if we did not have to use the same input twice, one with the baseline and one with the optimized binary.
 
-    \FIXME{Fig.XXX} shows how well these three candidates for an efficiency metric correlate with the metric we would actually like to have,
-    speedup over \texttt{-O0}. For all benchmarks, the correlation coefficient~\cite{Bishop:2006:PRM:1162264} is below \FIXME{0.20} for
-    runtime and below \FIXME{0.35} for inverse runtime. Both have obvious reasons for being inappropriate efficiency metrics. Doubling the
-    amount of data to be processed will cause the runtime to rise and the inverse runtime to fall, even if we use the same binary.
-    Efficiency has nothing to do with it. IPC also correlates poorly with performance speedup, the correlation coefficient being just
-    \FIXME{0.15}. While it is a common efficiency metric in computer architecture, it works there because it is used with invariant binaries.
-    In contrast, any compiler optimization that trades high latency instructions for multiple low latency instructions can reduce IPC while
-    improving efficiency.
+    Typical single-run performance metrics, like runtime or Instructions Per Cycle~(IPC), are not fit for purpose. To show this, we ran
+    each of the \FIXME{32} \textit{KDataSets} benchmarks 1,000 times, each time with a different input \textit{and} a different optimized
+    binary, and we measured three metrics: runtime, inverse runtime, and IPC. \FIXME{Fig.XXX} shows the correlation coefficient between
+    them and a reasonable efficiency metric, speedup over \texttt{-O0}. For all benchmarks, the correlation coefficient is below
+    \FIXME{0.20} for runtime and below \FIXME{0.35} for inverse runtime. Both have obvious reasons for being inappropriate efficiency
+    metrics. Doubling the amount of data to be processed will cause the runtime to rise and the inverse runtime to fall, even if we use the same binary.
+    Efficiency has nothing to do with it. IPC also correlates poorly with efficiency, the correlation coefficient being just \FIXME{0.15}.
+    While it is used often in computer architecture, it works there because it is used with invariant binaries. In contrast, any compiler
+    optimization that trades high latency instructions for multiple low latency instructions can reduce IPC while improving efficiency.
 

### 2017-11-07 work.tex
fix mult operator
minor
Merged work and profiling. Many other changes

--- 2017-09-15 work.tex
@@
-\section{Work Efficiency Metric} \label{sec:metric}
+\section{Work Efficiency} \label{sec:work}
 
-    In this section we define work efficiency, a metric for comparing different optimized versions of a program when executing with
-    different inputs. We define it as the ratio between the amount of \textit{work}, $\Delta W$, performed during a period of time, $\Delta
-    t$.
-
+    In this section, we introduce \textit{work efficiency}, a novel performance metric for comparing the quality of different optimized
+    versions of a program, even when we run them only once with a different input each time. Similarly to existing one-shot performance
+    metrics, work efficiency is simply the amount of \textit{work done per unit of time}: 
     \[
@@
     \]
+    where $\Delta W$ is the amount of work performed during a period of time $\Delta t$.
 
-    The main challenge is to precisely define what represents \textit{work}. We define the work done by a program on an input to be the
-    time taken by the unoptimized program to process that input. We estimate the runtime of each instruction type and then work is the sum
-    of the runtime for each instruction type multiplied by its dynamic count.
+    With similar metrics, what exactly is work depends on the application and is defined by the developer: SQL queries executed,
+    transactions completed, bytes processed, etc. In our case, we need a similar but application-agnostic metric which correlates with what
+    the developer considers useful work and is simple enough to be calculated online.
 
-    \[ \Delta W = \sum_i w_i.n_i \]
+    Section~\ref{subsec:workmetric} describes this new work metric and the design choices we made, Section~\ref{subsec:prof} details how
+    we measure work, and Section~\ref{subsec:relaxed} explores ways to measure it without interfering with the normal operation of the
+    program.
 
-    Where $i$ is the instruction type; $w_i$ is the estimated runtime for instruction type $i$; and $n_i$ is the number of observed
-    executions of instruction type $i$.
+    \subsection{Work Metric} \label{subsec:workmetric}
 
-    Similarly to previous work~\citep{giusto01,powell09,brandolese11}, the coefficients, $w_i$, are estimated by linear regression. We create
-    one regressor data point for each of a set of program and input pairs.
-    We use a separate set of programs from those discussed in Sections~\ref{sec:setup} and \ref{sec:results}.
-    %for estimating the architectural settings for the work metric.
-    The response variable is the execution time of the unoptimized program as it processes the input. To ensure
-    statistically sound results, the execution time is measured by repeated execution until the 99\% confidence interval is no larger than 1\% of
-    the mean. The explanatory variables are the dynamic instruction counts for each program and input pair. They are recorded by a separate,
-    instrumented run of the program on the input.
+    To build a reasonable work metric, our approach starts by assuming that each executed instruction contributes a certain amount towards
+    the total useful work. To keep the measurement simple and minimize the online overhead, we assume that the contribution depends only on
+    the type of instruction. Additionally, we would expect two program versions processing the same data in the same way to do the same
+    work. If measured using the actual instructions executed for the two versions, we would get a different amount of work. Instead, we
+    increase the work total based on the instructions that would be executed with the unoptimized code, regardless of the version we use.
+    The unoptimized version is convenient as a common baseline, since we start with it anyway and it is as close as possible to the
+    programmer's intention.
 
-    Figure~\ref{fig:work-vs-O0time} shows the correlation of the work metric ($\Delta W$) with runtime on instances of the unoptimized \texttt{susan\_c} benchmark.
-    Each point represents the execution of a single input, from a total of 1,000 inputs.
-    The correlation coefficient is $0.99$, and the mean absolute percentage error (MAPE) is only about 3.5\%.
-    This result suggests that the new metric correlates much better than either execution time or IPC.
-
-    %Figure~\ref{fig:motivation-work-efficiency} shows a similar plot to Figure~\FIXME{\ref{fig:motivation-metric}},
-    %of work efficiency versus speedup for 500 different binaries of \texttt{susan\_c}.
-    %Here the new metric correlates much better than either execution time or IPC,
-    %exhibiting a correlation coefficient of $0.93$ with speedup.
+    We selected the work contribution of each instruction type so that total work becomes an estimator of the runtime of the unoptimized
+    program. Our choice to associate work with unoptimized runtime is largely arbitrary but instinctively reasonable. Unoptimized time is
+    close to how the programmer understands the effort put by the computer. Using it as the work metric makes work efficiency (work per unit
+    of time) an estimator of the speedup of the optimized version over the unoptimized version. While other ways of setting the instruction
+    contributions might be better, our results in Section~\ref{sec:results} show that this simple definition of work produces good results.
+    Using the experimental setup described in Section~\ref{sec:setup}, we run the unoptimized versions of the \FIXME{7} benchmarks listed
+    in \FIXME{XXX} with all 1,000 available inputs. For each run, we recorded execution time and dynamic counts for all instruction types.
+    Similarly to prior work~\cite{giusto01,powell09,brandolese11}, we then applied linear regression to find the contribution of each
+    instruction type that minimizes the error between measured runtime and work.\footnote{The resulting regression model uses \FIXME{70}
+    different instruction types which make it too long to present here.}
 
@@
         \caption{
-            Relationship between work metric ($\Delta W$) and the execution time
-            of the unoptimized version of the \texttt{susan\_c} benchmark over the
-            1,000 inputs.
-            Each point represents the execution of a single input.
-            They have a correlation coefficient of $0.99$.
+            Relationship between work metric ($\Delta W$) and the execution time of the unoptimized version of the \texttt{susan\_c}
+            benchmark over the 1,000 inputs. Each point represents the execution of a single input. They have a correlation coefficient of
+            $0.99$.
         }
@@
 
-    %\begin{figure}[t]
-    %    \centering
-    %    \includegraphics[width=0.6\linewidth]{figs/cost-model.pdf}
-    %    \caption{
-    %        Linear model for work metric fitted from empirical data.
-    %        \red{zw: bigger font size.}
-    %    }
-    %    \label{fig:cost-model}
-    %\end{figure}
+    Figure~\ref{fig:work-vs-O0time} shows the correlation of the work metric ($\Delta W$) with runtime for the unoptimized version of one
+    benchmark that was not used for training the linear model, \texttt{susan\_c}. Each point represents the execution of a single input
+    from a total of 1,000 inputs. The correlation coefficient is $0.99$, and the mean absolute percentage error (MAPE) is only about 3.5\%.
+    \FIXME{More benchmarks and only coefficients?}.
 
-    %\begin{figure}[t]
-    %    \centering
-    %    \includegraphics[width=0.5\textwidth]{figs/motivation-work-efficiency.pdf}
-    %    \caption{Relationship between work efficiency \textit{vs} speedup for 500 different binaries of \texttt{susan\_c}.
-    %            Each point represents the average over a subset of all inputs.
-    %            The correlation coefficient is $0.93$.}
-    %    \label{fig:motivation-work-efficiency}
-    %\end{figure}
+    \subsection{Work Profiling}\label{subsec:prof}
+
+    Total work is just the sum of the cost of all instructions that would be executed in the unoptimized binary. A na\"{\i}ve but
+    straightforward way to measure work online is to instrument the unoptimized code with a global work counter and instructions for
+    incrementing this counter after each instruction of the original code. Since the work contribution of each instruction depends only on
+    its type, it can be determined at compile time. Taking this one step further, we can calculate work contributions at the basic block
+    level and insert only a single increment instruction in each block. Although easy to implement, the na\"{\i}ve approach introduces a
+    significant overhead.
+
+    Previous work on basic block profiling has proposed ways for instrumenting the code with as little overhead as possible without losing
+    any profiling information~\cite{knuth73,ball94}. The underlying idea is that we can still calculate precisely the execution counts of
+    all basic blocks if we choose a spanning tree of the function Control Flow Graph (CFG) and only instrument the edges of the CFG
+    \textit{not} in the tree~\cite{nahapetian73,forman81}. To make the instrumentation probe placement optimal, we choose the maximum
+    spanning tree, with the edge weights representing edge frequency estimates. This maximises the number of high frequency edges that will
+    not be instrumented, minimizing the instrumentation cost~\cite{forman81,ball94}.
+
+    \begin{figure}[t]
+    \centering {
+      \includegraphics[scale=0.8]{figs/cfg-example.pdf}\\\vspace{1ex}
+      \resizebox{0.45\textwidth}{!}{
+      %\scalebox{0.8}{
+         \begin{minipage}{0.5\textwidth}
+         Instrumented value for each probe $P_i$:
+         \begin{align*}
+         \omega(P_0) &= w(B_0) + w(B_1) + w(B_4) + w(B_6) + w(B_7) + w(B_8) + w(B_9)\\
+         \omega(P_1) &= w(B_2) - w(B_1) - w(B_4) - w(B_6) - w(B_7) - w(B_8)\\
+         \omega(P_2) &= w(B_5) - w(B_6)\\
+         \omega(P_3) &= w(B_3) - w(B_4) - w(B_6) - w(B_7)
+         \end{align*}
+         \end{minipage}
+      }
+    }
+      \caption{Example of a CFG with its maximum spanning tree in black. Basic blocks and edges highlighted in red are instrumented.
+        Instrumenting them is enough for calculating the total work performed by the whole CFG.}
+      \label{fig:cfg-example}
+    \end{figure}
+
+    In contrast to the na\"ive instrumentation where each basic block records only its own amount of work, with optimal probe placement 
+    the work counter increments need to take other basic blocks into account. Figure~\ref{fig:cfg-example} shows an example CFG, where the
+    instrumented blocks are marked with red. In this example, reaching probe $P_2$ implies not only executing block $B_5$ but also not
+    executing $B_6$.
+
+    To compute the amount of work associated with each instrumentation probe, we first need to determine how reaching a probe relates with
+    executing or not a basic block. Starting from instrumented edges, we use Kirchhoff's first law~\cite{knuth73,ball94} and a post-order
+    traversal of the spanning tree to build symbolic expressions of edge frequencies as linear functions of probe counts, as seen in
+    Figure~\ref{fig:cfg-example}. This part is identical to how edge and block frequencies are calculated in prior work for basic block
+    profiling.
+
+    %Simply put, the sum of the frequencies of the edges entering a block and the sum of the frequencies
+    %exiting the same block must match. Starting from the leaf nodes of the spanning tree, where by definition all edges except one should
+    %be instrumented, and working upwards, we should be able to calculate the frequencies of all nodes and edges as a function of the probe
+    %of the probe counts.
+
+    If a block's frequency expression contains a positive term for a probe, this means that the block is executed when we reach the probe,
+    so the block's cost should be added at that probe. Conversely, if the expression contains a negative term for a probe, then the block
+    is executed when we do not reach the probe, so the block's cost should be subtracted. In our example CFG, the symbolic expression for
+    the basic block $B_8$ is $P_0 - P_1$, so the amount of work of $B_8$, denoted by $w(B_8)$, is incremented in probe $P_0$ and
+    decremented in $P_1$.
+
+    \subsection{Relaxed Instrumentation}\label{subsec:relaxed}
+
+    Optimal instrumentation significantly reduces the profiling overhead when compared to na\"ive instrumentation, from an average overhead
+    of 79\% to 13\%. Still, 13\% slowdown might be more than what the user is ready to tolerate, while for isolated case the overhead might
+    be as high as 60\%. We need to reduce overhead further to make the work metric practical.
+
+    We propose two novel relaxation strategies that offer a trade-off between accuracy and overhead. Our insight was that, unlike basic
+    block profiling, not all probes are of equal importance. The work associated with some is insignificant compared to the total work
+    done by the program, so ignoring them will have little negative impact on the accuracy of measuring work. By removing such probes,
+    especially if they are frequently executed, we can reduce the instrumentation overhead dramatically. These two strategies are
+    particular to work profiling and reduce the overhead much more than what previous research in basic block profiling has achieved.
+
+    To select which probes to remove we consider the probes in a control flow region. The runtime benefit of removing a probe is
+    proportional to the frequency of the block the probe is in, while the error introduced is proportional to both the work contribution of
+    the probe and its frequency. We solve the 0-1 Knapsack optimization problem, removing probes to maximize the total savings while
+    keeping the accumulated error below a threshold:
+    \begin{gather*}
+        \textrm{max.}\quad\sum_{i=0}^{k} f(P_i)x_i,\quad
+        \textrm{s.t.}\quad\sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M \\
+        x_i\in\{0,1\}, i\in\{0,\ldots,k\}
+    \end{gather*}
+    where the set of probes in the region being considered is $\{P_0, P_1, \ldots, P_k\}$, $f(P_i)$ is the execution frequency of probe
+    $P_i$ and $x_i$ denotes the probes selected for removal, $\varepsilon(P_i)$ is the error for removing $P_i$ and $M$ is the maximum
+    error threshold.
+
+    For our experiments, we implemented two solvers for the 0-1 Knapsack problems: an optimal but potentially slow brute-force solver and a
+    greedy heuristic based on sorting the items~\cite{dantzig57}. We use the brute-force solver with a small number of probes, less than
+    20, and the greedy heuristic otherwise.
+
+    We have two relaxation strategies which differ in the control flow regions they tackle and how they calculate the error terms. The
+    first is a more conservative approach that calculates induced error from removing a probe by considering the worst case of any
+    execution path through the program. The second is a more aggressive approach that computes induced error based on the relative change
+    of the work measurement that removing a probe would make to the total amount work for the whole program. We discuss each one of them
+    in detail.
+
+    \subsubsection{\WCRelaxTitle Relaxation}
+
+    In order to offer hard guarantees for the error upper bound, this conservative approach first decomposes the CFGs of the program into
+    subgraphs where the maximum possible work contributed by each probe, as well as the minimum possible amount of total work for the whole
+    subgraph, can be calculated. Then, it removes probes within each subgraph while keeping the maximum possible error rate for that
+    subgraph bounded by the threshold. With the error rate of each subgraph below the threshold, it is guaranteed that the dynamic error of
+    the work metric for the whole program will always be bounded by the threshold.
+
+    \begin{figure}[t]
+      \centering
+      \includegraphics[scale=0.8]{figs/cfg-relax-example.pdf}
+      \caption{Example of a CFG containing a loop and its decomposition into DAGs.
+               The DAGs are the subgraphs within the dashed boundaries.}
+      \vspace{-4mm}
+      \label{fig:cfg-relax-example}
+    \end{figure}
+
+    The \WCRelaxLower relaxation starts by extracting directed acyclical graphs (DAGs) from the CFG. The algorithm extracts subgraphs that
+    represent loops or the outer most region of the function. The subgraphs are transformed into DAGs by ignoring the backedge and moving
+    inner loops to their own subgraphs. An example of this is shown in Figure~\ref{fig:cfg-relax-example}, where the CFG is partitioned
+    into two DAGs.
+
+    Since there are no loops in the DAG, we can determine $m$, the minimum amount of work that can be done in a single execution of the
+    DAG. For the same reason, the maximum amount of work that a probe can contribute is when it is reached once. Given this, the worst case
+    relative error caused by removing a probe $P_i$ is $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$. By repeating the same process for
+    all DAGs of the CFG, we guarantee that the final error introduced by relaxation will always be below the chosen threshold.
+
+    \subsubsection{\WPRelaxTitle Relaxation}
+
+    The \WCRelaxLower relaxation error guarantees are hard but also too conservative. It assumes that each loop will always execute the
+    path with the minimum amount of work possible, which might be orders of magnitude less than the work actually performed. When this is
+    the case, we end up with a relative error much lower than intended and a profiling overhead unnecessarily high. Our second, more
+    aggressive, strategy avoids this problem, by estimating the error and overhead contributions of each probe in relation to the whole
+    program.
+
+    The \WPRelaxLower relaxation does this by using block-frequency profiling from previous executions. With this information, we are able
+    to estimate the probe frequencies and the relative error of removing a given probe in terms of total program work. We select a subset
+    of all the probes to be removed, while keeping the total relative error (over total program work) below the threshold.
+
+    Contrary to the per DAG approach, the error introduced by \WPRelaxLower relaxation is not guaranteed to be bounded by the threshold
+    $M$. Still, we expect it to remain close to the threshold as long as the program's behavior does not deviate significantly from the
+    behavior captured in the profile we used.

### 2017-11-08 infra.tex
infrastructure rewrites

@@
 \section{Online {\IterComp} Overview} \label{sec:oic-infra}
-
-Although {\itercomp} had been originally proposed as an \textit{offline} optimization strategy, this work shows that this powerful
-technique can be adapted to work in \textit{online} scenarios. Instead of selecting the best optimization sequence at development time, a
-first version of the program is shipped bundled with the {\itercomp} mechanism. Our mechanism produces new application binaries using
-different optimization sequences. When the user starts the application, we load and execute one of these binaries to evaluate the
-corresponding sequence online. Our mechanism operates on the application's Intermediate Representation (IR) produced by LLVM. This saves us
-the effort of repeatedly parsing the code through the frontend and removes the need for shipping the full application code.
 
@@
 
-Figure~\ref{fig:infra-diagram} shows an overview of how our online {\itercomp} mechanism works. On the developer's side, the program is
-pre-compiled to IR without optimizations and is instrumented to measure the amount of work performed and the wall-clock time. On the user's
-side, we select an optimization sequence and then compile the instrumented IR to binary using this sequence. Next time the user runs the
-application, we measure the work and the execution time. We accumulate multiple such measurements, executing each input only once, until we
-can estimate the efficiency metric P of this optimization sequence with high confidence. Finally, we use this information to help us select
-the next optimization sequence.
+    Although {\itercomp} had been originally proposed as an \textit{offline} optimization strategy, this work shows that this powerful
+    technique can be adapted to work in \textit{online} scenarios. Figure~\ref{fig:infra-diagram} shows an overview of how our online
+    {\itercomp} mechanism works. On the developers' side, we would traditionally produce an application binary optimized for the inputs
+    provided by them. In our approach instead, the program is pre-compiled to LLVM Intermediate Representation (IR) without optimizations,
+    is instrumented to measure the amount of work performed and the wall-clock time, and is bundled with the {\itercomp} mechanism.
+    On the user's side, this mechanism selects an optimization sequence and compiles the instrumented IR to binary using this sequence.
+    Next time the user runs the application, we load and execute the latest binary and measure its work and execution time. We accumulate
+    multiple such measurements, executing each input only once, until we can estimate the work efficiency of this optimization sequence
+    with high confidence. Finally, we use this information to help us select the next optimization sequence.
 
-This work focuses on the highlighted component of Figure~\ref{fig:infra-diagram}, the \textit{Work Instrumentation}. It is performed once
-and offline, right after the IR is generated. Its purpose is to measure the amount of work performed. Conceptually, we use the cost model of
-Section~\ref{sec:metric} to estimate a cost for each basic block of the IR and then add instrumentation to increase a global work counter
-by that cost every time a basic block is executed. The measured work represents the computational effort of the unoptimized program and it
-should depend only on the program and the input, not the applied optimizations.
-For that reason, we instrument the unoptimized IR before any optimizations modify it.
-We present a detailed description of work
-instrumentation and how to keep it lightweight in Section~\ref{sec:prof}.
+    In the previous Section, we examined the highlighted component of Figure~\ref{fig:infra-diagram}, the \textit{Work Instrumentation}.
+    We perform it once and offline, right after the IR is generated before applying optimizations. Its purpose is to measure the amount of
+    work performed. The instrumentation creates a global work counter and adds code to a subset of the basic blocks to increment the counter
+    each time a block is executed. We calculate the work contribution of each basic block according to the model of Section~\ref{subsec:workmetric}.
+    We then select the subset of instrumented basic blocks and the work increments for each one, as discussed in Sections~\ref{subsec:prof}
+    and \ref{subsec:relaxed}.
 
@@
     optimization sequence. We call the number of executions performed using the same optimization sequence the \textit{Input-Window Size}.
+    \FIXME{MORE. How does the new search work?}
 

### 2017-11-09 cgo18mscr.tex
changes to the result section; and other minor changes
Merged work and profiling. Many other changes
Switched template to PLDI 2018

@@
 \PassOptionsToPackage{table}{xcolor}
-
-\documentclass[sigplan,10pt]{acmart}
-\input{premable}
+\PassOptionsToPackage{draft}{hyperref}
+\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\input{preamble}
 
@@
 
-\author{  }
+\author{Rodrigo C. O. Rocha}
+\affiliation{
+  \institution{University of Edinburgh, UK}
+}
+\email{r.rocha@ed.ac.uk}
 
-%\author{Rodrigo C. O. Rocha}
-%\affiliation{%
-%  \institution{University of Edinburgh, UK}
-%}
-%\email{r.rocha@ed.ac.uk}
+\author{Pavlos Petoumenos}
+\affiliation{
+  \institution{University of Edinburgh, UK}
+}
+\email{ppetoume@inf.ed.ac.uk}
 
-%\author{Pavlos Petoumenos}
-%\affiliation{%
-%  \institution{University of Edinburgh, UK}
-%}
-%\email{ppetoume@inf.ed.ac.uk}
+\author{Lu\'is F. W. G\'oes}
+\affiliation{
+  \institution{PUC Minas, Brazil}
+}
+\email{lfwgoes@pucminas.br}
 
-%\author{Lu\'is F. W. G\'oes}
-%\affiliation{%
-%  \institution{PUC Minas, Brazil}
-%}
-%\email{lfwgoes@pucminas.br}
+\author{Murray Cole}
+\affiliation{
+  \institution{University of Edinburgh, UK}
+}
+\email{mic@inf.ed.ac.uk}
 
-%\author{Murray Cole}
-%\affiliation{
-%  \institution{University of Edinburgh, UK}
-%}
-%\email{mic@inf.ed.ac.uk}
+\author{Zheng Wang}
+\affiliation{
+  \institution{Lancaster University, UK}
+}
+\email{z.wang@lancaster.ac.uk}
 
-%\author{Zheng Wang}
-%\affiliation{%
-%  \institution{Lancaster University, UK}
-%}
-%\email{z.wang@lancaster.ac.uk}
+\author{Hugh Leather}
+\affiliation{
+  \institution{University of Edinburgh, UK}
+}
+\email{hleather@inf.ed.ac.uk}
 
-%\author{Hugh Leather}
-%\affiliation{%
-%  \institution{University of Edinburgh, UK}
-%}
-%\email{hleather@inf.ed.ac.uk}
-
-%\renewcommand{\shortauthors}{R. Rocha et al.}
+\renewcommand{\shortauthors}{R. Rocha et al.}
 
@@
 %
-%\begin{CCSXML}
-%<ccs2012>
-%<concept>
-%<concept_id>10011007.10011006.10011041</concept_id>
-%<concept_desc>Software and its engineering~Compilers</concept_desc>
-%<concept_significance>500</concept_significance>
-%</concept>
-%<concept>
-%<concept_id>10011007.10011006.10011041.10011043</concept_id>
-%<concept_desc>Software and its engineering~Retargetable compilers</concept_desc>
-%<concept_significance>500</concept_significance>
-%</concept>
-%</ccs2012>
-%\end{CCSXML}
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10011007.10011006.10011041</concept_id>
+<concept_desc>Software and its engineering~Compilers</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011006.10011041.10011043</concept_id>
+<concept_desc>Software and its engineering~Retargetable compilers</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
 
-%\ccsdesc[500]{Software and its engineering~Compilers}
-%\ccsdesc[500]{Software and its engineering~Retargetable compilers}
+\ccsdesc[500]{Software and its engineering~Compilers}
+\ccsdesc[500]{Software and its engineering~Retargetable compilers}
 \keywords{{\IterComp}, Profiling} %, Relaxed Instrumentation}
@@
 \input{infra}
-\input{prof}
+%\input{prof}
 \input{setup}

### 2017-11-09 work.tex
minor changes in work

@@
     In this section, we introduce \textit{work efficiency}, a novel performance metric for comparing the quality of different optimized
-    versions of a program, even when we run them only once with a different input each time. Similarly to existing one-shot performance
-    metrics, work efficiency is simply the amount of \textit{work done per unit of time}: 
+    versions of a program, even when we evaluate them with a different input each time. Similarly to existing one-shot performance metrics,
+    work efficiency is simply the amount of \textit{work done per unit of time}: 
     \[
@@
 
-    In contrast to the na\"ive instrumentation where each basic block records only its own amount of work, with optimal probe placement 
-    the work counter increments need to take other basic blocks into account. Figure~\ref{fig:cfg-example} shows an example CFG, where the
-    instrumented blocks are marked with red. In this example, reaching probe $P_2$ implies not only executing block $B_5$ but also not
-    executing $B_6$.
+    With na\"ive instrumentation each basic block records only its own amount of work. With optimal probe placement, reaching a probe
+    implies not only executing the containing block but also executing or not executing other uninstrumented blocks, so work counter
+    increments have to reflect this. Figure~\ref{fig:cfg-example} shows an example CFG, where the instrumented blocks are marked with red.
+    In this example, reaching probe $P_2$ implies not only executing block $B_5$ but also not executing $B_6$.
 

### 2017-11-13 motivation.tex
updates to the motivation section.

@@
 
-    \paragraph{Iterative compilation needs representative inputs.} \FIXME{Fig.XXX} shows what happens when we apply iterative compilation
-    on \FIXME{susan\_c} using the wrong inputs. We selected \FIXME{50} of the 1,000 inputs of \textit{KDataSets}~\cite{chen10,chen12a} for
-    \FIXME{susan\_c} and we used them to evaluate different optimizations. The red bars show the average speedup over \texttt{-O3} achieved
-    by the optimization sequence selected by iterative compilation. On the left is the average over the \FIXME{50} inputs, on the right the
-    average over all 1,000 inputs. The selected sequence works great for the inputs we used, but it fails miserably for the rest, slowing
-    down the benchmark by \FIXME{5\%} on average.
+    \paragraph{Iterative compilation needs representative inputs.} 
+    In order to highlight what happens when we apply iterative compilation using the wrong set of inputs, we selected two distinct sets of
+    10 out of the 1,000 inputs of the of \textit{KDataSets}~\cite{chen10,chen12a} for the \texttt{tiff2bw} benchmark.
+    The first set represents the training set, i.e., the set of inputs that will be used for selecting the best optimization sequence by
+    iterative compilation, and the second set is a specific set of inputs which is misrepresented by the training set.
+    After performing iterative compilation on the first set, the selected optimization sequence has an average speedup of 1.22 over -O3 for
+    the training set itself, but it produces a slowdown of 0.91 for the second set of inputs.
+    However, if the iterative compilation was performed on the second set itself, it would be able to find a optimization sequence that
+    produces a speedup of \FIXME{1.02} over -O3.
 
-    The blue bars, on the other hand, show the ideal case where good optimizations are selected based on their average effect on all
-    inputs, not just a subset. Not only is the average speedup over all 1,000 inputs considerable, \FIXME{13\%} over \texttt{-O3}, but it
-    also works well for the \FIXME{50} input subset. The speedup for the subset is only \FIXME{1\%} less of what we got when we optimized
-    specifically for that subset.
+    %Even with larger set of training inputs, the lack of representative inputs for iterative compilation can still hurt performance.
+    %For the same \textttt{tiff2bw} benchmark, we were able to partition the whole dataset into 
+    %I.C. on 85% set has speedup 1.06 on the set itself and 0.975 slowdown on the 15% set.
+    %I.C. on 15% set has speedup 1.03.
 
-    Making iterative compilation work in this example is not a matter of handling input sensitivity, we were able to find an optimization
-    sequence that works well both on average and under most inputs. It is a matter of finding a set of inputs that represents the full
-    range of inputs the application is likely to process. If we do, we can improve average performance by \FIXME{13\%}. If not, we might
-    hurt average performance instead.
+%%    \paragraph{Iterative compilation needs representative inputs.} \FIXME{Fig.XXX} shows what happens when we apply iterative compilation
+%%    on \FIXME{susan\_c} using the wrong inputs. We selected \FIXME{50} of the 1,000 inputs of \textit{KDataSets}~\cite{chen10,chen12a} for
+%%    \FIXME{susan\_c} and we used them to evaluate different optimizations. The red bars show the average speedup over \texttt{-O3} achieved
+%%    by the optimization sequence selected by iterative compilation. On the left is the average over the \FIXME{50} inputs, on the right the
+%%    average over all 1,000 inputs. The selected sequence works great for the inputs we used, but it fails miserably for the rest, slowing
+%%    down the benchmark by \FIXME{5\%} on average. 
+%%
+%%    The blue bars, on the other hand, show the ideal case where good optimizations are selected based on their average effect on all
+%%    inputs, not just a subset. Not only is the average speedup over all 1,000 inputs considerable, \FIXME{13\%} over \texttt{-O3}, but it
+%%    also works well for the \FIXME{50} input subset. The speedup for the subset is only \FIXME{1\%} less of what we got when we optimized
+%%    specifically for that subset.
+%%
+%%    Making iterative compilation work in this example is not a matter of handling input sensitivity, we were able to find an optimization
+%%    sequence that works well both on average and under most inputs. It is a matter of finding a set of inputs that represents the full
+%%    range of inputs the application is likely to process. If we do, we can improve average performance by \FIXME{13\%}. If not, we might
+%%    hurt average performance instead.
 
@@
 
-    \paragraph{Using real inputs online is not enough.} We could overcome these problems by just evaluating optimization sequences online
-    while the application is being used. This gives us live data but creates a new problem. Optimizations sequences are compared based on
-    how much faster the program becomes. When all of them are tested against the same inputs, it is just a matter of comparing runtimes.
-    A shorter runtime for processing the same data means a more efficient binary. With live data, we cannot do this, each binary processes
+    %\paragraph{Using real inputs online is not enough.} We could overcome these problems by just evaluating optimization sequences online
+    %while the application is being used. This gives us live data but creates a new problem. Optimization sequences are compared based on
+    %how much faster the program becomes. When all of them are tested against the same inputs, it is just a matter of comparing runtimes.
+    %A shorter runtime for processing the same data means a more efficient binary. With live data, we cannot do this, each binary processes
+    %whatever the input happens to be. We need another efficiency metric that only requires a single use of an optimized binary with a
+    %single input.
+
+    \paragraph{Challenges of using real inputs online.} We could overcome the problems of representative inputs by just evaluating
+    optimization sequences online while the application is being used. 
+    %Optimization sequences are compared based on how much faster the program becomes.
+    This gives us live data but creates a new problems.
+    With live data, we cannot just compare runtimes. 
+    When all of the optimization sequences are tested against the same inputs, it is just a matter of comparing runtimes.
+    A shorter runtime for processing the same data means a more efficient binary.
+    With live data, we cannot do this, each binary processes
     whatever the input happens to be. We need another efficiency metric that only requires a single use of an optimized binary with a

### 2017-11-13 abstract.tex
abstract

--- 2017-09-16 abstract.tex
@@
 
-    Iterative compilation can dramatically improve the performance of a program by searching for the best optimization.
-    However, it relies upon repeatedly evaluating optimizations on predefined representative inputs.
-    If the developers fail to guess what is a representative input or usage patterns change over time,
-    then the wrong optimizations may be selected.
-    Searching the optimization space with actual user inputs, during deployment, has previously seemed infeasible.
-    Program side effects and difficulties capturing the system state mean that any given input can
-    only be executed once. %Since each input may entail a different amount of work, the search cannot use runtime to compare
-    %optimizations, and there is no baseline against which to define speedup.
-    Unless we find a way to compare optimizations even when inputs are executed only once, we will be unable to use real user
-    input for iterative compilation, users will not have the best optimized programs, and performance will suffer.
+    Iterative compilation can dramatically improve the performance of a program by searching for the best optimization. However, it relies
+    upon repeatedly evaluating optimizations on predefined representative inputs. If the developers fail to guess what is a representative
+    input or usage patterns change over time, then the wrong optimizations may be selected. Searching the optimization space with actual
+    user inputs, during deployment, has previously seemed infeasible. Program side effects and difficulties capturing the system state mean
+    that any given input can only be executed once, online. Existing performance metrics cannot be used to compare optimizations evaluated
+    under different inputs. Unless we devise a new metric that can be used in such a way, we will be unable to apply iterative compilation
+    online with real live inputs, users will not have properly optimized programs, and performance will suffer.
 
-    We present metrics and techniques which, for the first time,  enable the use of true {\em online iterative compilation}.
-    We compare optimizations by a new work efficiency metric, which requires only a single execution with any one input.
-    We develop an instrumented, low overhead, mechanism for gathering the required data, using optimal probe placement in the source
-    code. We show that overhead can be further reduced by removing high-cost probes whose absence introduces only small errors to the
-    work metric, without significantly affecting the outcome of the iterative compilation.
-    We give two variants for probe removal optimization: the first offers strong
+    We present metrics and techniques which, for the first time, enable the use of true {\em online iterative compilation}. We compare
+    optimizations by a new work efficiency metric, which requires only a single execution with any one input. We develop an instrumented,
+    low overhead, mechanism for gathering the required data, using optimal probe placement in the source code. We show that overhead can be
+    further reduced by removing high-cost probes whose absence introduces only small errors to the work metric, without significantly
+    affecting the outcome of the iterative compilation. We give two variants for probe removal optimization: the first offers strong
     guarantees about the maximum possible error, while the second targets the average whole program error.
 
-    Our online iterative compilation method yields programs that get 60\% of the speedup achievable by an offline oracle,
-    but without executing any input more than once, which the oracle cannot do.
-    Through probe removal we bring the instrumentation overhead down to only 5\% on average, making our
-    approach suitable for real-world scenarios. %We show practical online iterative compilation for the first time,
-    %optimizing programs according to the range of inputs users actually present.
+    Our online iterative compilation method yields programs that get 60\% of the speedup achievable by an oracle that depends on reusing
+    inputs, something that cannot be done online. Through probe removal we bring the instrumentation overhead down to only 5\% on average,
+    making our approach suitable for real-world scenarios. %We show practical online iterative compilation for the first time, optimizing
+    %programs according to the range of inputs users actually present.
 \end{abstract}

### 2017-11-13 work.tex
minor updates to text

@@
     Section~\ref{subsec:workmetric} describes this new work metric and the design choices we made, Section~\ref{subsec:prof} details how
-    we measure work, and Section~\ref{subsec:relaxed} explores ways to measure it without interfering with the normal operation of the
-    program.
+    we measure work, and Section~\ref{subsec:relaxed} explores ways to measure it with reduced interference on %without interfering with
+    the normal operation of the program.
 
@@
 
-    To build a reasonable work metric, our approach starts by assuming that each executed instruction contributes a certain amount towards
+    To build a reasonable work metric, our approach starts by assuming that each executed instruction\footnote{In this paper, we consider instructions at the level of the intermediate representation of the LLVM compiler.} contributes a certain amount towards
     the total useful work. To keep the measurement simple and minimize the online overhead, we assume that the contribution depends only on
@@
     work. If measured using the actual instructions executed for the two versions, we would get a different amount of work. Instead, we
-    increase the work total based on the instructions that would be executed with the unoptimized code, regardless of the version we use.
+    increase the total work based on the instructions that would be executed with the unoptimized code, regardless of the version we use.
     The unoptimized version is convenient as a common baseline, since we start with it anyway and it is as close as possible to the
@@
     We selected the work contribution of each instruction type so that total work becomes an estimator of the runtime of the unoptimized
-    program. Our choice to associate work with unoptimized runtime is largely arbitrary but instinctively reasonable. Unoptimized time is
+    program.
+    %Our choice to associate work with unoptimized runtime is largely arbitrary but instinctively reasonable.
+    Our choice to associate work with unoptimized runtime is largely for practical reasons but it is also instinctively reasonable.
+    Unoptimized time is
     close to how the programmer understands the effort put by the computer. Using it as the work metric makes work efficiency (work per unit
@@
     Using the experimental setup described in Section~\ref{sec:setup}, we run the unoptimized versions of the \FIXME{7} benchmarks listed
-    in \FIXME{XXX} with all 1,000 available inputs. For each run, we recorded execution time and dynamic counts for all instruction types.
+    in \FIXME{XXX} with all 1,000 available inputs. For each run, we recorded execution time and dynamic counters for all instruction types\footnote{
+    Dynamic instruction counts were obtained by using the classic block-frequency profiling.}.
     Similarly to prior work~\cite{giusto01,powell09,brandolese11}, we then applied linear regression to find the contribution of each
-    instruction type that minimizes the error between measured runtime and work.\footnote{The resulting regression model uses \FIXME{70}
-    different instruction types which make it too long to present here.}
+    instruction type that minimizes the error between measured runtime and work.
+    %\footnote{The resulting regression model uses \FIXME{70}
+    %different instruction types which make it too long to present here.}
 

### 2017-11-14 motivation.tex
adds corr_coeff fig for all benchmarks
updates speedup and adds figures

@@
     However, if the iterative compilation was performed on the second set itself, it would be able to find a optimization sequence that
-    produces a speedup of \FIXME{1.02} over -O3.
+    produces a speedup of 1.07 over -O3.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.7\linewidth]{figs/motivation_wrong_inputs.pdf}
+    \caption{Iterative compilation performed on the wrong set of inputs compared to the ideal case for the \texttt{tiff2bw} benchmark.}
+    \label{fig:motivation_wrong_inputs}
+\end{figure}
 
@@
     how to create one and how to use it for iterative compilation.
+
+\begin{figure*}[t]
+    \centering
+    \includegraphics[width=\textwidth]{figs/corr_coeff.pdf}
+    \caption{Correlation coefficient.}
+    \label{fig:corr_coeff}
+\end{figure*}

### 2017-11-14 work.tex
updates speedup and adds figures

@@
 
-    To build a reasonable work metric, our approach starts by assuming that each executed instruction\footnote{In this paper, we consider instructions at the level of the intermediate representation of the LLVM compiler.} contributes a certain amount towards
+    To build a reasonable work metric, our approach starts by assuming that each executed instruction\footnote{In this paper, we consider instructions at the level of the Intermediate Representation (IR) of the LLVM compiler.} contributes a certain amount towards
     the total useful work. To keep the measurement simple and minimize the online overhead, we assume that the contribution depends only on

### 2017-11-14 motivation.tex
adds fall-back results of I.C. with IPC and 1/Runtime
adds text about offline I.C. with IPC in motivation

@@
 
-    Typical single-run performance metrics, like runtime or Instructions Per Cycle~(IPC), are not fit for purpose. To show this, we ran
+    Typical single-run performance metrics, like runtime or Instructions Per Cycle~(IPC), are not fit for purpose.
+    \FIXME{Evidence1:}
+    To show this, we performed iterative compilation guided by IPC for the same set of 10 inputs
+    discussed in Figure~\ref{fig:motivation_wrong_inputs} and compare with the previous approach which optimizes directly for execution time.
+    The best optimization sequence found in Figure~\ref{fig:motivation_wrong_inputs}
+    has an average speedup of 1.7 over \texttt{-O3} and an average IPC of 2.59.
+    However, iterative compilation guided by IPC is able to find an optimization
+    sequence that produces a higher IPC of 2.72 but an average slowdown of 0.98.
+    This result agrees with the known fact that higher IPC does not necessarily translate to lower execution time when comparing different versions of the program.
+    \FIXME{Evidence2:}
+    To show this, we ran
     each of the \FIXME{32} \textit{KDataSets} benchmarks 1,000 times, each time with a different input \textit{and} a different optimized
@@
     \centering
-    \includegraphics[width=\textwidth]{figs/corr_coeff.pdf}
-    \caption{Correlation coefficient.}
-    \label{fig:corr_coeff}
+    \includegraphics[width=\textwidth]{figs/motivation-speedups.pdf}
+    \caption{\FIXME{[Fall-back Result]} Performance improvement of iterative compilation guided by IPC and runtime.}
+    \label{fig:motivation-speedups}
 \end{figure*}

### 2017-11-14 related.tex
minor reviews in abstract
related work, small changes

@@
 settings~\cite{agakov06,kulkarni04,stephenson03,hoste08}. These approaches compare the performance of candidate compiler options by first
-compile the program with different options and then profiling the different compiled versions on the same program input. All these
+compiling the program with different options and then profiling the different compiled versions on the same program input. All these
 approaches implicitly assume that the best-performing compiler settings found from a small set of inputs will work well on unseen datasets.
+
 Chen~\etal~\cite{chen10,chen12a} show that the best optimization options, however, can differ across inputs. Their results suggest that, in
@@
 developers provide a few tens of representative inputs to be used by the iterative compilation search process. We argue that providing
-representative inputs is difficult in practice, as users may not use the applications as the developer expects. Even this is possible,
+representative inputs is difficult in practice, as users may not use the applications as the developer expects. Even if this is possible,
 current approaches all require to evaluate all candidate compiler options on the same input. Doing so across a range of inputs would incur
@@
 
-%=======
-%Multiple adaptive compilation techniques have been proposed for finding a near optimal combination of optimization
-%settings~\cite{agakov06,hoste08,kulkarni04,stephenson03} by probabilistically searching the space of possible combinations. But with
-%all of them there is uncertainty about whether the quality of their results depends on the data set used. Chen~\etal~\cite{chen10,chen12a}
-%examined this issue and showed that indeed the best optimization options differ across inputs. Still, in most cases it is possible to find
-%an optimization sequence that produces good results across all inputs, by evaluating the optimizations on a representative set inputs.
-%While a positive result, this is still impractical. Producing a set of realistic inputs might be difficult. Even if not, each
-%optimization sequence has to be evaluated on every input, increasing the time \itercomp takes by at least an order of
-%magnitude.
-%>>>>>>> origin/master
-
 Recent studies~\cite{chen12b,fang15} have used user inputs to perform {\itercomp} in distributed data centers. Using these approaches, each
 compilation worker receives a subset of the input dataset to evaluate a small set of optimization settings on. The best such setting from
-each round is used for subsequent executions of the same code. The best-found compilation strategy is then re-fined over time, by testing
+each round is used for subsequent executions of the same code. The best-found compilation strategy is then refined over time, by testing
 new settings and re-evaluating old ones on new datasets. Like in~\cite{chen10} every setting is tested on the same set of inputs, wasting
@@
 that application binaries displaying higher IPC are more efficient than lower IPC binaries. However, many optimization techniques affect
-IPC in the opposite way, making the program more efficient while lowering IPC. We have also shown in Section~\ref{sec:motivation}, IPC is
-not strongly correlate to speedup. Furthermore, IPC cannot be used to estimate performance for multi-threaded
-applications~\cite{alameldeen06,eyerman08} due to the side-effect of thread contention. Aalameldeen and Wood~\cite{alameldeen06} also
-suggest that relying on IPC can be harmful, and work metrics like time per work unit or work units per time should be used. However, prior
-works all rely on developers to provide such metrics. This paper presents the first automatic approach to derive a such work metric from
-arbitrary code without developer intervention.
+IPC in the opposite way, making the program more efficient while lowering IPC. The authors in~\cite{fursin07} acknowledge that this limits
+the applicability of IPC for \itercomp. In the context of multithreaded applications, IPC cannot be used to compare the performance of
+different runs even when they use the same intput~\cite{alameldeen06,eyerman08}. Work-related metrics are shown to be more appropriate, but prior work
+relied on application-specific work metrics. This paper presents the first automatic and application agnostic approach for deriving a work metric.
+
+%PP: Discuss input sensitivity?
 

### 2017-11-14 motivation.tex
small udpdates to motivation

@@
     \paragraph{Iterative compilation needs representative inputs.} 
-    In order to highlight what happens when we apply iterative compilation using the wrong set of inputs, we selected two distinct sets of
+    Figure~\ref{fig:motivation_wrong_inputs} highlights what happens when we apply iterative compilation using the wrong set of inputs, we selected two distinct sets of
     10 out of the 1,000 inputs of the of \textit{KDataSets}~\cite{chen10,chen12a} for the \texttt{tiff2bw} benchmark.
     The first set represents the training set, i.e., the set of inputs that will be used for selecting the best optimization sequence by
-    iterative compilation, and the second set is a specific set of inputs which is misrepresented by the training set.
-    After performing iterative compilation on the first set, the selected optimization sequence has an average speedup of 1.22 over -O3 for
-    the training set itself, but it produces a slowdown of 0.91 for the second set of inputs.
-    However, if the iterative compilation was performed on the second set itself, it would be able to find a optimization sequence that
-    produces a speedup of 1.07 over -O3.
+    iterative compilation, and the second set is a specific set of inputs which is misrepresented by the training set, let us call it the production set.
+    After performing iterative compilation on the training set, the selected optimization sequence has an average of 22\% improvement over -O3 for
+    the training set itself, but it produces a slowdown of about 9\% for the production set.
+    However, in the ideal case, if the iterative compilation was performed on the production set itself, it would be able to find an optimization sequence that
+    results on a 7\% improvement over -O3, this is an improvement of about 17\% over
+    the result of iterative compilation on the wrong set of inputs.
 
@@
     \includegraphics[width=0.7\linewidth]{figs/motivation_wrong_inputs.pdf}
-    \caption{Iterative compilation performed on the wrong set of inputs compared to the ideal case for the \texttt{tiff2bw} benchmark.}
+    \caption{Performance results for the \texttt{tiff2bw} benchmark, comparing optimization sequences selected by iterative compilation when performed on the wrong set of inputs and the ideal case using the same set of inputs.
+    This result supports the fact that iterative compilation needs representative inputs.}
     \label{fig:motivation_wrong_inputs}

### 2017-11-14 motivation.tex
minor changes

@@
 
-\begin{figure}[t]
+\begin{figure}[ht]
     \centering
@@
     This result agrees with the known fact that higher IPC does not necessarily translate to lower execution time when comparing different versions of the program.
-    \FIXME{Evidence2:}
-    To show this, we ran
-    each of the \FIXME{32} \textit{KDataSets} benchmarks 1,000 times, each time with a different input \textit{and} a different optimized
-    binary, and we measured three metrics: runtime, inverse runtime, and IPC. \FIXME{Fig.XXX} shows the correlation coefficient between
-    them and a reasonable efficiency metric, speedup over \texttt{-O0}. For all benchmarks, the correlation coefficient is below
-    \FIXME{0.20} for runtime and below \FIXME{0.35} for inverse runtime. Both have obvious reasons for being inappropriate efficiency
-    metrics. Doubling the amount of data to be processed will cause the runtime to rise and the inverse runtime to fall, even if we use the same binary.
-    Efficiency has nothing to do with it. IPC also correlates poorly with efficiency, the correlation coefficient being just \FIXME{0.15}.
-    While it is used often in computer architecture, it works there because it is used with invariant binaries. In contrast, any compiler
-    optimization that trades high latency instructions for multiple low latency instructions can reduce IPC while improving efficiency.
+%     \FIXME{Evidence2:}
+%    To show this, we ran
+%    each of the \FIXME{32} \textit{KDataSets} benchmarks 1,000 times, each time with a different input \textit{and} a different optimized
+%    binary, and we measured three metrics: runtime, inverse runtime, and IPC. \FIXME{Fig.XXX} shows the correlation coefficient between
+%    them and a reasonable efficiency metric, speedup over \texttt{-O0}. For all benchmarks, the correlation coefficient is below
+%    \FIXME{0.20} for runtime and below \FIXME{0.35} for inverse runtime. Both have obvious reasons for being inappropriate efficiency
+%    metrics. Doubling the amount of data to be processed will cause the runtime to rise and the inverse runtime to fall, even if we use the same binary.
+%    Efficiency has nothing to do with it. IPC also correlates poorly with efficiency, the correlation coefficient being just \FIXME{0.15}.
+%    While it is used often in computer architecture, it works there because it is used with invariant binaries. In contrast, any compiler
+%    optimization that trades high latency instructions for multiple low latency instructions can reduce IPC while improving efficiency.
 
@@
 
-\begin{figure*}[t]
-    \centering
-    \includegraphics[width=\textwidth]{figs/motivation-speedups.pdf}
-    \caption{\FIXME{[Fall-back Result]} Performance improvement of iterative compilation guided by IPC and runtime.}
-    \label{fig:motivation-speedups}
-\end{figure*}
+%\begin{figure*}[t]
+%    \centering
+%    \includegraphics[width=\textwidth]{figs/motivation-speedups.pdf}
+%    \caption{\FIXME{[Fall-back Result]} Performance improvement of iterative compilation guided by IPC and runtime.}
+%    \label{fig:motivation-speedups}
+%\end{figure*}

### 2017-11-14 work.tex
minor changes

@@
 
-    \begin{figure}[t]
-        \centering
-        \includegraphics[width=0.5\textwidth]{figs/work-correlation-with-O0_susan_c.pdf}
-        \caption{
-            Relationship between work metric ($\Delta W$) and the execution time of the unoptimized version of the \texttt{susan\_c}
-            benchmark over the 1,000 inputs. Each point represents the execution of a single input. They have a correlation coefficient of
-            $0.99$.
-        }
-        \label{fig:work-vs-O0time}
-    \end{figure}
+%%    \begin{figure}[t]
+%%        \centering
+%%        \includegraphics[width=0.5\textwidth]{figs/work-correlation-with-O0_susan_c.pdf}
+%%        \caption{
+%%            Relationship between work metric ($\Delta W$) and the execution time of the unoptimized version of the \texttt{susan\_c}
+%%            benchmark over the 1,000 inputs. Each point represents the execution of a single input. They have a correlation coefficient of
+%%            $0.99$.
+%%        }
+%%        \label{fig:work-vs-O0time}
+%%    \end{figure}
 
@@
     \FIXME{More benchmarks and only coefficients?}.
+
+\begin{figure*}[t]
+    \centering
+    \includegraphics[width=\textwidth]{figs/corr_coeff.pdf}
+    \caption{Correlation coefficients between the work metric ($\Delta W$) and the execution time of the unoptimized version for each benchmark.}
+    \label{fig:motivation-speedups}
+\end{figure*}
 

### 2017-11-15 work.tex
removed fixme. Also setup

@@
     contributions might be better, our results in Section~\ref{sec:results} show that this simple definition of work produces good results.
-    Using the experimental setup described in Section~\ref{sec:setup}, we run the unoptimized versions of the \FIXME{7} benchmarks listed
-    in \FIXME{XXX} with all 1,000 available inputs. For each run, we recorded execution time and dynamic counters for all instruction types\footnote{
+    Using the experimental setup described in Section~\ref{sec:setup}, we run the unoptimized versions of the 10 benchmarks listed in
+    Table~\ref{tab:kdatasets:training} with all 1,000 available inputs. For each run, we recorded execution time and dynamic counters for all instruction types\footnote{
     Dynamic instruction counts were obtained by using the classic block-frequency profiling.}.
@@
 
-    Figure~\ref{fig:work-vs-O0time} shows the correlation of the work metric ($\Delta W$) with runtime for the unoptimized version of one
-    benchmark that was not used for training the linear model, \texttt{susan\_c}. Each point represents the execution of a single input
-    from a total of 1,000 inputs. The correlation coefficient is $0.99$, and the mean absolute percentage error (MAPE) is only about 3.5\%.
-    \FIXME{More benchmarks and only coefficients?}.
+    Figure~\ref{fig:work-vs-O0time} shows the correlation coefficient between the work metric ($\Delta W$) and the runtime of the
+    unoptimized version for 14 benchmarks that were not used for training the linear model. The work metric correlates very strongly with
+    runtime for almost all the benchmarks, the correlation coefficient being typically higher than $0.98$. The exception is the benchmarks
+    based on \texttt{tiff}. The coefficient for them ranges from $0.47$ to $0.96$. Despite that, in most cases work is an excellent proxy
+    for unoptimized runtime. For the rest, our Section~\ref{sec:results} evaluation shows that our work metric is still useful for quantifying
+    performance.
 

### 2017-11-15 work.tex
Minor error

@@
 
-    Figure~\ref{fig:work-vs-O0time} shows the correlation coefficient between the work metric ($\Delta W$) and the runtime of the
+    Figure~\ref{fig:motivation-speedups} shows the correlation coefficient between the work metric ($\Delta W$) and the runtime of the
     unoptimized version for 14 benchmarks that were not used for training the linear model. The work metric correlates very strongly with

### 2017-11-15 results.tex
minor comments
instrumentation evaluation rewrite

@@
 \section{Experimental Results}\label{sec:results}
-
-%In this section we discuss our experimental evaluation. Afterwards, we discuss our results concerning the instrumentations for profiling
-%our work metric. Finally we present the results of the online {\itercomp}.
 
@@
 
-\subsection{Evaluation of the Instrumentation}
-
-In this experiment we evaluate the runtime overhead introduced by the work efficiency profiling. We compile the program using the
-\texttt{-O3}. We then measure the wall-clock time of the instrumented and the native versions to calculate the runtime overhead. For
-each benchmark, we compute the average overhead over all its 1,000 inputs.
-Figure~\ref{fig:overhead-O3} shows the runtime overhead imposed by the work profiling using each of the proposed work profiling strategies over
-their non-instrumented counterparts.
-%The WPO relaxation was performed using profiling information for every specific input, which provides
-%the perfect information to perform the WPO relaxation. This means that the experiments show its best performance enabled by having perfect
-%profiling information.
-
-%\begin{figure}[t]
-%    \centering
-%    \includegraphics[width=0.395\textwidth]{figs/adpcm_d-cfg-instr.pdf}
-%    \caption{CFG of the function that contains the hot loop of the {\flagstype adpcm\_d} benchmark.}
-%    \label{fig:adpcm_d-cfg-instr}
-%\end{figure}
-
-While the \OptProf has a maximum overhead of near 60\%, the \WCRelax-\textit{5}\% has an overhead of
-less than 16\% for all benchmarks, with the \WPRelax-\textit{5}\% reaches an average overhead of only 5\%,
-compared to an average overhead of 12\% for the optimal profiling.
-
 \begin{figure*}[t!]
@@
     \includegraphics[width=\textwidth]{figs/overhead-O3.pdf}
-    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.}
+    \caption{Instrumentation overhead for each benchmark averaged over all inputs, when compiled with {\flagstype -O3}.}
     \vspace{-3mm}
@@
 \end{figure*}
+
 \begin{figure*}[t!]
@@
 
-The most critical case, amongst the evaluated benchmarks, happens with the
-\texttt{adpcm\_d} program, with an overhead of about 59\% for the optimal
-instrumentation.
-With a threshold of only 5\%, the \WCRelaxLower relaxation reduces
-the overhead by about 75\% over the \OptProf,
-and the \WPRelax-5\% reduces it by about 80\%.
-This benchmark has a single hot function
-which consists mainly of a single hot loop with several branches inside it. %, as depicted by Figure~\ref{fig:adpcm_d-cfg-instr}.
-Most of the improvement achieved by both strategies comes from removing only two probes from the hot loop. The two relaxed probes were
-placed in branches with a high probability of being taken, but with a small contribution to the total amount of
-work measured in DAG for this loop (both probes with an static error of about 1.3\%). Because these probes were placed in basic blocks with
-a considerable execution frequency, removing the probes resulted in a significant reduction in runtime overhead.
+\subsection{Evaluation of the Instrumentation}
 
-A few of the benchmarks present abnormal regressions in their execution time.
-This increase in the profiling overhead is counter-intuitive because the
-relaxation algorithms only reduces the number of instrumented probes,
-without changing their placement.
-We suggest some reasons for these abnormal regressions.
-By analyzing these cases, we noticed that some of the frequently executed
-probes could not be removed by the relaxation strategies with small thresholds.
-Another important aspect to point out is that we also observed
-differences among the final CFGs after applying the \texttt{-O3} optimization.
-Although relaxation does not directly alter the CFG, it is worth mentioning that
-there are optimizations that make use of LLVM's built-in target-based cost model,
-which means that adding or removing instructions from basic blocks may affect
-decisions during transformations of the code, which may end up changing the CFG
-in different ways depending on the instrumentation applied.
-%For example, some of the optimizations that make use of this built-in cost model
-%and also alter the CFG are: loop-unrolling, function inlining, and
-%simplifications of the CFG.
-%These changes in the program's CFGs may affect other compiler optimizations and
-%also possibly in hardware performance, for example affecting the instruction
-%caching.
+First, we evaluate the runtime overhead introduced by the work efficiency profiling. A high overhead would impact the user's experience
+and would make our approach impractical. We take the original and instrumented versions of each benchmark, we compile them using the
+baseline \texttt{-O3} setting, and execute them with all 1,000 inputs. The difference in their runtimes, for the same input, is the
+instrumentation overhead. Figure~\ref{fig:overhead-O3} shows the average overhead for each benchmark and work profiling strategy.
 
-Finally, as shown in Figure~\ref{fig:error-O3}, these improvements can be
-obtained while incurring very little dynamic error.
-This result confirms the overly conservative aspect of the \WCRelaxLower relaxation,
-while the \WPRelaxLower relaxation is significantly less conservative.
-Although the error bound is not guaranteed by the \WPRelaxLower relaxation,
-its dynamic error was always below the 5\% thresholds.
-%due to the use of perfect profiling information.
+\OptProf typically slows the application down by 10\%, noticeably affecting the user experience, while in the worst case its overhead
+reaches 59\%. It is clearly unsuitable for online work profiling. \WCRelax and \WPRelax fare better, especially when used with a higher
+5\% threshold. \WPRelax-\textit{5}\%, in particular, has an average overhead of only 5\%, 13\% in its worst case.
+
+For almost all the profiling strategies, their worst case is \texttt{adpcm\_d}. Still, compared to the 59\% overhead of the \OptProf,
+\WCRelax-\textit{5}\% and \WPRelax-\textit{5}\% incur only about a quarter of that overhead. This benchmark has a single hot function
+consisting mainly of a single hot loop with several branches inside it. Most of the overhead comes from two probes which are placed in
+very frequently executed blocks but contribute little to the total work of the loop. The maximum possible error caused by removing either
+of the probes, based on the analysis of the DAG, is about 1.3\%. Both relaxation strategies identify this opportunity and remove the probes.
+
+Some benchmarks display unexpectedly higher overheads under the relaxation strategies. This is counter-intuitive because relaxation only
+reduces the number of instrumented probes without changing their placement. By analyzing these cases, we noticed that we could not remove
+the most frequently executed probes, so relaxation had little positive impact. At the same time, we removed some probes which were
+infrequently executed but were taken into consideration by the LLVM optimization heuristics. In some cases, removed probes caused different
+and suboptimal decisions to be made, increasing the runtime.
+
+Figure~\ref{fig:error-O3} shows that profiling with relaxation incurs very little dynamic error. It also confirms that the \WCRelaxLower
+relaxation is overly conservative, while the \WPRelaxLower relaxation achieves a better trade-off. Although the error bound is not
+guaranteed by the \WPRelaxLower relaxation, its dynamic error is always below the 5\% threshold. For both relaxation strategies the 2\%
+threshold is too conservative to be useful. The rest of the paper only evaluates the 5\% threshold.
 
@@
 	         The speedups reported for each benchmark represents the average speedup across their complete 1000 input datasets.}
+
+    \FIXME{Sort  the bars according to speedup, i.e. Oracle should be the last bars.}
     \label{fig:speedups}
@@
 assessing the effectiveness of the work-efficiency metric.
-For the relaxation strategies, we evaluate only those with a 5\% threshold, since
-they offered better overhead reductions for little dynamic errors in the measurement.
 

### 2017-11-15 results.tex
rest of results

@@
 relaxation is overly conservative, while the \WPRelaxLower relaxation achieves a better trade-off. Although the error bound is not
-guaranteed by the \WPRelaxLower relaxation, its dynamic error is always below the 5\% threshold. For both relaxation strategies the 2\%
-threshold is too conservative to be useful. The rest of the paper only evaluates the 5\% threshold.
+guaranteed by the \WPRelaxLower relaxation, its dynamic error is always below 5\%. For both relaxation strategies the 2\% threshold is too
+conservative to be useful. The rest of the paper only evaluates the 5\% threshold.
 
@@
 
-In this section we evaluate the effectiveness of our work efficiency metric.
-We perform online {\itercomp} using all five profiling strategies, namely
-\OracleRM, \OraclePP,\OptProf,\WCRelax-5\%, and \WPRelax-5\%.
-Because the \OracleRM measures actual speedup, it provides reference results for
-assessing the effectiveness of the work-efficiency metric.
+In this section we evaluate the effectiveness of our work efficiency metric. We perform online {\itercomp} using all five profiling
+strategies, namely \OracleRM, \OraclePP, \OptProf, \WCRelax-5\%, and \WPRelax-5\%. We evaluate every optimization sequence on multiple
+inputs, using a dynamic input-window size, as explained in Section~\ref{sec:oic-infra}. We use the average work efficiency over the input
+window to rank different optimization sequences and we select the one with the highest work efficiency.
 
-For all strategies, the same optimization sequence is used for multiple inputs, using a dynamic input-window size, as explained in
-Section~\ref{sec:oic-infra}. The average performance over the input window provides an estimate for the overall performance of the
-optimization sequence across distinct inputs. The optimization sequences are ranked based on their average performance, and the best
-optimization sequence is selected.
+\OracleRM makes iterative compilation decisions based on the actual speedup over the unoptimized version. It does not depend on any kind of
+instrumentation or work estimation, so it acts as the baseline against which the rest of the profiling strategies are compared. \OraclePP
+is also not affected by instrumentation but uses work efficiency instead of speedup. We use it to decide whether our work metric can be
+used for iterative compilation on principle. The other configurations demonstrate the viability of applying online {\itercomp} in
+real-world scenarios.
 
-The comparison between \OracleRM and \OraclePP is crucial for validating the
-effectiveness of the work efficiency metric in guiding online {\itercomp},
-while the other configurations demonstrate the viability of applying online
-{\itercomp} in real-world scenarios.
+We evaluate the quality of the selected optimization sequences by comparing their average speedup over \texttt{-O3} across all 1,000 inputs
+of each benchmark.
+% None of the configurations degrades the performance over \texttt{-O3} (with statistical significance). \FIXME{PP: According to the figure there are slowdowns!!!}
+Figure~\ref{fig:speedups} shows these speedups for all test benchmarks. In most cases, \OraclePP is very close to \OracleRM, achieving on
+average 80\% of its speedup. This demonstrates that it is a valid choice to drive iterative compilation decision using work efficiency. The
+few cases where the two strategies produce noticeably different results are for the \texttt{tiff} benchmarks. The weaker correlation between
+work and unoptimized runtime for these benchmarks is to blame for the lower speedups achieved by \OraclePP.
 
-In order to evaluate the quality of the final optimization sequences selected by each configuration of the online {\itercomp}, we compare
-their speedup over the \texttt{-O3} optimization across all the 1,000 inputs of the benchmark being optimized. None of the configurations
-degrades the performance over \texttt{-O3} (with statistical significance).
-%When measuring the wall-clock time for each input, to reduce noise, we execute
-%the same input until we have a statistically sound measurement, i.e. we execute
-%until we have an interval no larger than 1\% with 99\% confidence.
-Figure~\ref{fig:speedups} shows these average speedups for all test benchmarks.
-This figure shows that the best optimization sequence selected with the Oracle-PP
-is very close to the performance of the best optimization sequence selected with
-the \OracleRM, where \OraclePP achieves on average about 80\% of the performance
-improvement obtained by the \OracleRM.
-This result is important for demonstrating that our work efficiency metric has
-the potential to produce good results in real-world online scenarios.
+The \OptProf achieves 4\% improvement on average, which represents a little bit more that half of the speedup of \OraclePP. The culprit is
+the one thing that is different between the two strategies, the profiling instrumentation. It affects the search in two key ways:
+\textit{(i.)} the overhead incurred by profiling affects the execution time and, through it, the work efficiency metric,
+\textit{(ii.)} the inserted instrumentation code occasionally changes how optimizations are applied, e.g., due to the use of a global
+variable or in decisions taken based on a cost-model for the instructions. With the \OptProf configuration, instrumentation is very
+intrusive and this translates into lower speedups or even slowdowns for five benchmarks. Just measuring work with traditional profiling
+approaches is not enough for online iterative compilation.
 
-The \OptProf achieves 4\% improvement on average, which represents 45\% of the
-performance improvement obtained by the \OracleRM.
-This difference to \OraclePP is explained by the real (intrusive) profiling used
-in the \OptProf configuration.
-The profiling affects the search in two key ways:
-\textit{(i.)} the overhead incurred by the profiling affects the execution time,
-and therefore also affects the work efficiency metric,
-\textit{(ii.)} the instrumentation code may affect many of the optimizations, e.g.,
-due to the use of a global variable or in decisions taken based on a cost-model
-for the instructions.
-However, even in the \OptProf configuration, which has a highly intrusive instrumentation,
-it is still possible to obtain performance improvements in a realistic online
-{\itercomp}.
-
-Finally, the evaluation also indicates that the use of relaxation algorithms
-is beneficial not only for its reduction in overhead, which is directly observed
-by the user, by it also tends to improve the quality of the search.
-This improvement has to do with the aforementioned effects of an intrusive
-instrumentation on the search.
-The relaxation has the potential to reduce both effects of the profiling.
-Both the \WCRelax-5\% and \WPRelax-5\% achieve about 60\% of the
-performance improvement obtained by the \OracleRM.
+This is where our two relaxation algorithms become particularly important. Beyond just reducing the user perceived overhead of iterative
+compilation, they also improve the search itself by keeping the behavior of the instrumented binary as close as possible to that of the
+uninstrumented one. Relaxation reduces the deleterious effects of instrumentation on both the runtime and how optimizations are applied.
+As a result, \WCRelax-5\% and \WPRelax-5\% are very close to the behavior of \OraclePP and they achieve about 60\% of the performance
+improvement obtained by \OracleRM.

### 2017-11-15 motivation.tex
updates to motivation.

@@
     \paragraph{Iterative compilation needs representative inputs.} 
+%%    Figure~\ref{fig:motivation_wrong_inputs} highlights what happens when we apply iterative compilation using the wrong set of inputs, we selected two distinct sets of
+%%    10 out of the 1,000 inputs of the of \textit{KDataSets}~\cite{chen10,chen12a} for the \texttt{tiff2bw} benchmark.
+%%    The first set represents the training set, i.e., the set of inputs that will be used for selecting the best optimization sequence by
+%%    iterative compilation, and the second set is a specific set of inputs which is misrepresented by the training set, let us call it the production set.
+%%    After performing iterative compilation on the training set, the selected optimization sequence has an average of 22\% improvement over -O3 for
+%%    the training set itself, but it produces a slowdown of about 9\% for the production set.
+%%    However, in the ideal case, if the iterative compilation was performed on the production set itself, it would be able to find an optimization sequence that
+%%    results on a 7\% improvement over -O3, this is an improvement of about 17\% over
+%%    the result of iterative compilation on the wrong set of inputs.
+
     Figure~\ref{fig:motivation_wrong_inputs} highlights what happens when we apply iterative compilation using the wrong set of inputs, we selected two distinct sets of
     10 out of the 1,000 inputs of the of \textit{KDataSets}~\cite{chen10,chen12a} for the \texttt{tiff2bw} benchmark.
-    The first set represents the training set, i.e., the set of inputs that will be used for selecting the best optimization sequence by
-    iterative compilation, and the second set is a specific set of inputs which is misrepresented by the training set, let us call it the production set.
-    After performing iterative compilation on the training set, the selected optimization sequence has an average of 22\% improvement over -O3 for
-    the training set itself, but it produces a slowdown of about 9\% for the production set.
-    However, in the ideal case, if the iterative compilation was performed on the production set itself, it would be able to find an optimization sequence that
-    results on a 7\% improvement over -O3, this is an improvement of about 17\% over
-    the result of iterative compilation on the wrong set of inputs.
+    We perform iterative compilation in one of the set of inputs and then evaluate the selected optimization on both set of inputs.
+    The goal is to show that if iterative compilation is performed on a training set which misrepresents the user's real workload,
+    the search can end up selecting a optimization sequence that results in poor performance for the user's real workload.
+    After performing iterative compilation on the input set \#1, the selected optimization sequence, refered as optimization \#1, has an average of 22\% improvement over \texttt{-O3} for
+    the input set \#1 itself, but it produces a slowdown of about 9\% for the input set \#2.
+    On the other hand, if iterative compilation is performed on the input set \#2, the selected optimization sequence, refered as optimization \#2, is able to produce an average improvement of 7\% over \texttt{-O3} for the input set \#2 itself, but resulting in an improvement of only 5\% for the input set \#1.
+    In both cases, performing iterative compilation on the ideal set of inputs provides an improvement of over 15\% compared to the other option.
+    This result shows that when iterative compilation is performed on representative inputs it results in the most profitable optimization sequence for the user's real workload.
 
@@
     \centering
-    \includegraphics[width=0.7\linewidth]{figs/motivation_wrong_inputs.pdf}
+    \includegraphics[width=0.7\linewidth]{figs/motivation_two_inputs.pdf}
     \caption{Performance results for the \texttt{tiff2bw} benchmark, comparing optimization sequences selected by iterative compilation when performed on the wrong set of inputs and the ideal case using the same set of inputs.

### 2017-11-16 motivation.tex
adds some text for the new experiments

@@
     With live data, we cannot do this, each binary processes
-    whatever the input happens to be. We need another efficiency metric that only requires a single use of an optimized binary with a
+    whatever the input happens to be. We need another efficiency metric that only requires a single execution of an optimized binary with a
     single input.
@@
     Typical single-run performance metrics, like runtime or Instructions Per Cycle~(IPC), are not fit for purpose.
-    \FIXME{Evidence1:}
-    To show this, we performed iterative compilation guided by IPC for the same set of 10 inputs
-    discussed in Figure~\ref{fig:motivation_wrong_inputs} and compare with the previous approach which optimizes directly for execution time.
-    The best optimization sequence found in Figure~\ref{fig:motivation_wrong_inputs}
-    has an average speedup of 1.7 over \texttt{-O3} and an average IPC of 2.59.
-    However, iterative compilation guided by IPC is able to find an optimization
-    sequence that produces a higher IPC of 2.72 but an average slowdown of 0.98.
+    To show this, we performed iterative compilation executing each optimization sequence with a different input and then measuring both runtime and IPC.
+    The experiments were performed in the set of inputs \#2 discussed in Figure~\ref{fig:motivation_wrong_inputs}.
+    For the iterative compilation guided by runtime, it resulted on an average slowdown of 0.97 compared to \texttt{-O3}.
+    For the iterative compilation guided by IPC, it resulted in an average slowdown of 0.51 with an IPC of 3.25., while the optimization expected with an average speedup of 1.07 has an average IPC of 2.31.
     This result agrees with the known fact that higher IPC does not necessarily translate to lower execution time when comparing different versions of the program.
+
 %     \FIXME{Evidence2:}

### 2017-11-16 results.tex
adds some text for the new experiments

@@
 improvement obtained by \OracleRM.
+
+\subsubsection{Analysis of the cost during the online search}
+
+We also evaluate the actual cost perceived by the user when applying online iterative compilation.
+To this end we used the architecture described in Section~\ref{sec:oic-infra} with the search based on genetic algorithm.
+The results suggest that the proposed approach can be profitable in real-world online scenarios.
+For the \WCRelax-5\% configuration, an average speedup of XX was observed during the search, in the worst case with an slowdown of XX.
+For the \WPRelax-5\% configuration, an average speedup of XX was observed during the search, in the worst case with an slowdown of XX.

### 2017-11-16 work.tex
detailed the differences of work profiling and block frequency profiling

@@
     traversal of the spanning tree to build symbolic expressions of edge frequencies as linear functions of probe counts, as seen in
-    Figure~\ref{fig:cfg-example}. This part is identical to how edge and block frequencies are calculated in prior work for basic block
-    profiling.
+    Figure~\ref{fig:cfg-example}. This part is similar to how edge and block frequencies are calculated in prior work for basic block
+    profiling, with a few key differences:
+    (i)   work profiling does not have to store the execution count of each basic block, only the total work which makes it cheaper to profile;
+    (ii)  it does not require a post-processing of the recorded profiling for populating the edge flows;
+    (iii) the propagation of the edge flows happens in a symbolic fashion during the instrumentation of the code, as depicted in Figure~\ref{fig:cfg-example}.
+    This symbolic edge flows are used to derive the frequency expression of the basic blocks, which is then used to derive the amount of work
+    recorded by each probe.
 

### 2017-11-16 work.tex
fix minor typo

@@
     (iii) the propagation of the edge flows happens in a symbolic fashion during the instrumentation of the code, as depicted in Figure~\ref{fig:cfg-example}.
-    This symbolic edge flows are used to derive the frequency expression of the basic blocks, which is then used to derive the amount of work
+    These symbolic edge flows are used to derive the frequency expression of the basic blocks, which is then used to derive the amount of work
     recorded by each probe.

### 2017-11-16 work.tex
minor fix

--- 2017-11-16 work.tex
@@
     (iii) the propagation of the edge flows happens in a symbolic fashion during the instrumentation of the code, as depicted in Figure~\ref{fig:cfg-example}.
-    This symbolic edge flows are used to derive the frequency expression of the basic blocks, which is then used to derive the amount of work
+    These symbolic edge flows are used to derive the frequency expressions of the basic blocks, which are then used to derive the amount of work
     recorded by each probe.

### 2017-11-16 motivation.tex
motivation

@@
     their application is actually being used. This is not always easy or even possible. Then, they need to create data that will trigger a
-    similar behavior and this needs a deep knowledge of the application and considerable effort. This will have to be a continuous work in
+    similar behavior as when the application is used by users. This needs a deep knowledge of the potential user behavior and
+    considerable effort. This will have to be a continuous work in
     progress, to handle changes in the way the application is used.
@@
     modifications to ensure that the application does the same work every time the same data are reused, despite changes in the rest of
-    the system state. This is significant engineering work. Even if this was not the case, many applications are just not able to save
+    the system state. \FIXME{ZW: This is a bit vague. I think we need an example to illustrate why the application may not perform the same
+    work if the same input is used.}
+    This is significant engineering work. Even if this was not the case, many applications are just not able to save
     data for privacy or storage overhead reasons.
@@
     With live data, we cannot do this, each binary processes
-    whatever the input happens to be. We need another efficiency metric that only requires a single execution of an optimized binary with a
-    single input.
+    whatever the input happens to be.
+    As an example, considering a sorting algorithm. To measure the efficiency of an optimization for the algorithm, we need to look at not just
+    how long does the algorithm run but also how many items it processes per run. Simply comparing the running time of two invocations does not
+    tell us which program version is faster, as the two invocations may not process the same number of items. As a result, we need another efficiency metric
+    that only requires a single execution of an optimized binary with a single input.
 
-    \paragraph{Existing efficiency metrics are not up to the task.} Application-specific efficiency metrics~\cite{alameldeen06,coppa14} may
-    work well, but the developer has to choose the appropriate metric and then manually modify the application to calculate it. Speedup
+    \paragraph{Existing efficiency metrics are not up to the task.} Application-specific efficiency metrics such as the number of transactions per second may
+    work well, but the developer has to choose the appropriate metric and then manually modify the application to calculate it~\cite{alameldeen06,coppa14}. Speedup
     over a common baseline binary is a good application-agnostic metric, equating efficiency with how much faster a given input is
@@
     For the iterative compilation guided by runtime, it resulted on an average slowdown of 0.97 compared to \texttt{-O3}.
-    For the iterative compilation guided by IPC, it resulted in an average slowdown of 0.51 with an IPC of 3.25., while the optimization expected with an average speedup of 1.07 has an average IPC of 2.31.
-    This result agrees with the known fact that higher IPC does not necessarily translate to lower execution time when comparing different versions of the program.
+    For the iterative compilation guided by IPC, it resulted in an average slowdown of 0.51 with an IPC of 3.25, while the optimization expected with an average speedup of 1.07 has an average IPC of 2.31.
+    This result agrees with the known fact that higher IPC does not necessarily translate to lower execution time when comparing \emph{different versions} of the program.
 

### 2017-11-16 motivation.tex
motivation

--- 2017-11-16 motivation.tex
@@
 
+    \begin{figure}[t]
+        \centering
+        \includegraphics[width=0.7\linewidth]{figs/motivation_two_inputs.pdf}
+        \caption{Average speedup over \texttt{-O3} for the \texttt{tiff2bw} benchmark when optimized with two different optimization sequences and
+        evaluated on two distinct input sets. Each sequence was selected by iterative compilation on only one of the input sets. Processing
+        inputs with a version optimized for a different input set produces suboptimal results, supporting that iterative compilation needs
+        representative inputs.}
+        \label{fig:motivation_wrong_inputs}
+    \end{figure}
+
     \paragraph{Iterative compilation needs representative inputs.} 
@@
 
-    Figure~\ref{fig:motivation_wrong_inputs} highlights what happens when we apply iterative compilation using the wrong set of inputs, we selected two distinct sets of
-    10 out of the 1,000 inputs of the of \textit{KDataSets}~\cite{chen10,chen12a} for the \texttt{tiff2bw} benchmark.
-    We perform iterative compilation in one of the set of inputs and then evaluate the selected optimization on both set of inputs.
-    The goal is to show that if iterative compilation is performed on a training set which misrepresents the user's real workload,
-    the search can end up selecting a optimization sequence that results in poor performance for the user's real workload.
-    After performing iterative compilation on the input set \#1, the selected optimization sequence, refered as optimization \#1, has an average of 22\% improvement over \texttt{-O3} for
-    the input set \#1 itself, but it produces a slowdown of about 9\% for the input set \#2.
-    On the other hand, if iterative compilation is performed on the input set \#2, the selected optimization sequence, refered as optimization \#2, is able to produce an average improvement of 7\% over \texttt{-O3} for the input set \#2 itself, but resulting in an improvement of only 5\% for the input set \#1.
-    In both cases, performing iterative compilation on the ideal set of inputs provides an improvement of over 15\% compared to the other option.
-    This result shows that when iterative compilation is performed on representative inputs it results in the most profitable optimization sequence for the user's real workload.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=0.7\linewidth]{figs/motivation_two_inputs.pdf}
-    \caption{Performance results for the \texttt{tiff2bw} benchmark, comparing optimization sequences selected by iterative compilation when performed on the wrong set of inputs and the ideal case using the same set of inputs.
-    This result supports the fact that iterative compilation needs representative inputs.}
-    \label{fig:motivation_wrong_inputs}
-\end{figure}
-
-    %Even with larger set of training inputs, the lack of representative inputs for iterative compilation can still hurt performance.
-    %For the same \textttt{tiff2bw} benchmark, we were able to partition the whole dataset into 
-    %I.C. on 85% set has speedup 1.06 on the set itself and 0.975 slowdown on the 15% set.
-    %I.C. on 15% set has speedup 1.03.
-
-%%    \paragraph{Iterative compilation needs representative inputs.} \FIXME{Fig.XXX} shows what happens when we apply iterative compilation
-%%    on \FIXME{susan\_c} using the wrong inputs. We selected \FIXME{50} of the 1,000 inputs of \textit{KDataSets}~\cite{chen10,chen12a} for
-%%    \FIXME{susan\_c} and we used them to evaluate different optimizations. The red bars show the average speedup over \texttt{-O3} achieved
-%%    by the optimization sequence selected by iterative compilation. On the left is the average over the \FIXME{50} inputs, on the right the
-%%    average over all 1,000 inputs. The selected sequence works great for the inputs we used, but it fails miserably for the rest, slowing
-%%    down the benchmark by \FIXME{5\%} on average. 
-%%
-%%    The blue bars, on the other hand, show the ideal case where good optimizations are selected based on their average effect on all
-%%    inputs, not just a subset. Not only is the average speedup over all 1,000 inputs considerable, \FIXME{13\%} over \texttt{-O3}, but it
-%%    also works well for the \FIXME{50} input subset. The speedup for the subset is only \FIXME{1\%} less of what we got when we optimized
-%%    specifically for that subset.
-%%
-%%    Making iterative compilation work in this example is not a matter of handling input sensitivity, we were able to find an optimization
-%%    sequence that works well both on average and under most inputs. It is a matter of finding a set of inputs that represents the full
-%%    range of inputs the application is likely to process. If we do, we can improve average performance by \FIXME{13\%}. If not, we might
-%%    hurt average performance instead.
+    Figure~\ref{fig:motivation_wrong_inputs} highlights what happens when we apply iterative compilation using the wrong set of inputs. We
+    selected two distinct subsets of the \textit{KDataSets}~\cite{chen10,chen12a} inputs for the \texttt{tiff2bw} benchmark. Each subset
+    has ten inputs. Using iterative compilation we find the best optimization sequence for each subset separately.
+    Figure~\ref{fig:motivation_wrong_inputs} shows the speedup over \texttt{-O3} achieved by each sequence when using either of the two
+    subsets. Optimization \#1, which was selected using input set \#1, speeds up \texttt{tiff2bw} by 22\% on average when processing set
+    \#1, but hurts performance when used with set \#2. Similarly, using optimization \#2 on input set \#1 only improves performance by 5\%,
+    much less than what optimization \#1 achieves. In both cases, optimizing the benchmark for the same set it will be evaluated on
+    improves performance by 15\% compared to optimizing for some other set. To make iterative compilation profitable, applying it on a
+    inputs similar to the ones the application will be used with is absolutely necessary.
 
@@
     modifications to ensure that the application does the same work every time the same data are reused, despite changes in the rest of
-    the system state. This is significant engineering work. Even if this was not the case, many applications are just not able to save
-    data for privacy or storage overhead reasons.
+    the system state. \FIXME{ZW: This is a bit vague. I think we need an example to illustrate why the application may not perform the same
+    work if the same input is used.}
+    This is significant engineering work. Even if this was not the case, many applications are just not able to save
+    data for privacy or storage overhead reasons. \FIXME{Refs?}
 
-    %\paragraph{Using real inputs online is not enough.} We could overcome these problems by just evaluating optimization sequences online
-    %while the application is being used. This gives us live data but creates a new problem. Optimization sequences are compared based on
-    %how much faster the program becomes. When all of them are tested against the same inputs, it is just a matter of comparing runtimes.
-    %A shorter runtime for processing the same data means a more efficient binary. With live data, we cannot do this, each binary processes
-    %whatever the input happens to be. We need another efficiency metric that only requires a single use of an optimized binary with a
-    %single input.
+    \paragraph{Using live inputs online is not enough.} We could overcome the problem of representative inputs by just evaluating
+    optimization sequences online while the application is being used. This gives us live data but creates a bigger problem, that we cannot
+    compare different optimizations just by runtime anymore. When optimizations are tested against the same inputs, determining their
+    relative efficiency is just a matter of comparing runtimes. A shorter runtime for the same work means a more efficient binary.
+    With live data, we cannot do this, the binary processes whatever the input happens to be. Larger inputs require more work and will
+    likely lead to longer execution times, regardless of optimizations.
 
-    \paragraph{Challenges of using real inputs online.} We could overcome the problems of representative inputs by just evaluating
-    optimization sequences online while the application is being used. 
-    %Optimization sequences are compared based on how much faster the program becomes.
-    This gives us live data but creates a new problems.
-    With live data, we cannot just compare runtimes. 
-    When all of the optimization sequences are tested against the same inputs, it is just a matter of comparing runtimes.
-    A shorter runtime for processing the same data means a more efficient binary.
-    With live data, we cannot do this, each binary processes
-    whatever the input happens to be. We need another efficiency metric that only requires a single execution of an optimized binary with a
-    single input.
+    As an example, consider a sorting algorithm. To measure the efficiency of an optimization for the algorithm, we need to look at not just
+    how long the algorithm runs but also how many items it processes per run. Just comparing the runtimes of two invocations does not
+    provide any information about efficiency, as the two invocations may not process the same number of items. We need another efficiency metric
+    that only requires a single execution of an optimized binary with a single input.
 
@@
     work well, but the developer has to choose the appropriate metric and then manually modify the application to calculate it. Speedup
-    over a common baseline binary is a good application-agnostic metric, equating efficiency with how much faster a given input is
-    processed. It would be ideal, if we did not have to use the same input twice, one with the baseline and one with the optimized binary.
+    over a common baseline binary is an adequate application-agnostic metric but it cannot be used online. It requires measuring runtime
+    for each input twice, one with the baseline and one with the optimized binary.
 
-    Typical single-run performance metrics, like runtime or Instructions Per Cycle~(IPC), are not fit for purpose.
-    To show this, we performed iterative compilation executing each optimization sequence with a different input and then measuring both runtime and IPC.
-    The experiments were performed in the set of inputs \#2 discussed in Figure~\ref{fig:motivation_wrong_inputs}.
-    For the iterative compilation guided by runtime, it resulted on an average slowdown of 0.97 compared to \texttt{-O3}.
-    For the iterative compilation guided by IPC, it resulted in an average slowdown of 0.51 with an IPC of 3.25., while the optimization expected with an average speedup of 1.07 has an average IPC of 2.31.
-    This result agrees with the known fact that higher IPC does not necessarily translate to lower execution time when comparing different versions of the program.
+    Typical single-run performance metrics, like runtime or Instructions Per Cycle~(IPC), are not fit for purpose. To show this, we
+    performed iterative compilation on \texttt{tiff2bw}, evaluating each optimization sequence on a different input and using either
+    runtime or IPC as our efficiency metric. We selected our inputs from input set \#2 of Figure~\ref{fig:motivation_wrong_inputs}. When
+    iterative compilation is driven by runtime, the selected optimization slows down the benchmark by 3\% compared to \texttt{-O3}. When
+    driven by IPC, it is even worse with an average slowdown of 49\%. Its very high IPC of 3.25 does not indicate high performance. The
+    actual best optimization over the whole input set improves performance by 7\% while only having an average IPC of 2.31.
 
@@
 %    optimization that trades high latency instructions for multiple low latency instructions can reduce IPC while improving efficiency.
-
-    No existing metric can give us an application-agnostic measure of efficiency with a single execution. The rest of the paper will show
-    how to create one and how to use it for iterative compilation.
+    No existing metric can give us an application-agnostic measure of efficiency with a single execution. Without one, online iterative
+    compilation is impossible. Without doing it online, iterative compilation will always be limited by the quality of the input set.
+    The rest of the paper will show how to create an appropriate efficiency metric and how to use it for true online iterative
+    compilation.
 

### 2017-11-16 motivation.tex
WORK

--- 2017-11-16 motivation.tex
@@
     This gives us live data but creates a new problems.
-    With live data, we cannot just compare runtimes.
-    When all of the optimization sequences are tested against the same inputs, it is just a matter of comparing runtimes.
+    With live data, we cannot just compare runtime.
+    When all of the optimization sequences are tested against the same inputs, it is just a matter of comparing runtime.
     A shorter runtime for processing the same data means a more efficient binary.
@@
     As an example, considering a sorting algorithm. To measure the efficiency of an optimization for the algorithm, we need to look at not just
-    how long does the algorithm run but also how many items it processes per run. Simply comparing the running time of two invocations does not
-    tell us which program version is faster, as the two invocations may not process the same number of items. As a result, we need another efficiency metric
+    how long does the algorithm run but also how many items it processes per run. Simply comparing the running time of two invocations of two program versions does not
+    tell us which program version is faster, as the two invocations may not process the same number of items in a production environment. As a result, we need another efficiency metric
     that only requires a single execution of an optimized binary with a single input.

### 2017-11-16 work.tex
WORK

@@
     With similar metrics, what exactly is work depends on the application and is defined by the developer: SQL queries executed,
-    transactions completed, bytes processed, etc. In our case, we need a similar but application-agnostic metric which correlates with what
+    transactions completed, bytes processed, etc. In our case, we need a similar but \emph{application-agnostic} metric which correlates with what
     the developer considers useful work and is simple enough to be calculated online.
@@
     Table~\ref{tab:kdatasets:training} with all 1,000 available inputs. For each run, we recorded execution time and dynamic counters for all instruction types\footnote{
-    Dynamic instruction counts were obtained by using the classic block-frequency profiling.}.
+    Dynamic instruction counts were obtained by using the classic block-frequency profiling~\cite{knuth73}.}.
     Similarly to prior work~\cite{giusto01,powell09,brandolese11}, we then applied linear regression to find the contribution of each

### 2017-11-16 results.tex
results

@@
 \begin{itemize}[leftmargin=3mm]
-\item \textbf{\OracleRM} measures the actual speedup over the unoptimized version of the program for any given input.
-In our experiments, this oracle is allowed to execute each input at least twice so the speedup can be computed.
+\item \textbf{\OracleRM} measures the actual speedup over the unoptimized version of the program for any given input. In our experiments,
+    this oracle is allowed to execute each input \emph{at least twice} so the speedup can be computed.
 \item \textbf{\OraclePP} measures the work efficiency metric with a \textit{magically perfect} non-intrusive profiling.
@@
   it allows us to isolate and evaluate the actual effectiveness of the work efficiency metric.
-\item \textbf{\OptProf} corresponds to the work profiling using the optimal placement of the probes.
-  This profiling strategy requires a single execution of the program for any given input.
+\item \textbf{\OptProf} corresponds to the work profiling using the optimal placement of the probes. This profiling strategy requires
+    \emph{a single execution} of the program for any given input.
 \item \textbf{\WCRelax-\textit{N}\%} measures the work efficiency metric using the work profiling with a \textit{N}\% threshold for
@@
 	         The speedups reported for each benchmark represents the average speedup across their complete 1000 input datasets.}
-
-    \FIXME{Sort  the bars according to speedup, i.e. Oracle should be the last bars.}
     \label{fig:speedups}

### 2017-11-16 work.tex
applied Hugh's comments on work

@@
+\begin{figure*}[t]
+    \centering
+    \includegraphics[width=\textwidth]{figs/corr_coeff.pdf}
+    \caption{Correlation coefficients between the work metric ($\Delta W$) and the execution time of the unoptimized version for each benchmark.}
+    \label{fig:motivation-speedups}
+\end{figure*}
+
 \section{Work Efficiency} \label{sec:work}
@@
     With similar metrics, what exactly is work depends on the application and is defined by the developer: SQL queries executed,
-    transactions completed, bytes processed, etc. In our case, we need a similar but \emph{application-agnostic} metric which correlates with what
+    transactions completed, bytes processed, etc. In our case, we need an \emph{application-agnostic} metric which correlates with what
     the developer considers useful work and is simple enough to be calculated online.
@@
     program.
-    %Our choice to associate work with unoptimized runtime is largely arbitrary but instinctively reasonable.
-    Our choice to associate work with unoptimized runtime is largely for practical reasons but it is also instinctively reasonable.
+    Our choice to associate work with unoptimized runtime was instinctively reasonable.
     Unoptimized time is
     close to how the programmer understands the effort put by the computer. Using it as the work metric makes work efficiency (work per unit
-    of time) an estimator of the speedup of the optimized version over the unoptimized version. While other ways of setting the instruction
-    contributions might be better, our results in Section~\ref{sec:results} show that this simple definition of work produces good results.
+    of time) an estimator of the speedup of the optimized version over the unoptimized version. This speedup cannot be measured otherwise without
+    running the program more than once on each input. While other ways of setting the instruction contributions are possible, our evaluation in
+    Section~\ref{sec:results} shows that this simple definition of work produces good results.
     Using the experimental setup described in Section~\ref{sec:setup}, we run the unoptimized versions of the 10 benchmarks listed in
@@
     runtime for almost all the benchmarks, the correlation coefficient being typically higher than $0.98$. The exception is the benchmarks
-    based on \texttt{tiff}. The coefficient for them ranges from $0.47$ to $0.96$. Despite that, in most cases work is an excellent proxy
-    for unoptimized runtime. For the rest, our Section~\ref{sec:results} evaluation shows that our work metric is still useful for quantifying
-    performance.
-
-\begin{figure*}[t]
-    \centering
-    \includegraphics[width=\textwidth]{figs/corr_coeff.pdf}
-    \caption{Correlation coefficients between the work metric ($\Delta W$) and the execution time of the unoptimized version for each benchmark.}
-    \label{fig:motivation-speedups}
-\end{figure*}
+    based on \texttt{tiff}. The coefficient for them ranges from $0.47$ to $0.96$. Despite that, our work metric is still an excellent proxy
+    for unoptimized runtime and, as seen in Section~\ref{sec:results}, it is useful for quantifying performance.
 
@@
     Total work is just the sum of the cost of all instructions that would be executed in the unoptimized binary. A na\"{\i}ve but
-    straightforward way to measure work online is to instrument the unoptimized code with a global work counter and instructions for
-    incrementing this counter after each instruction of the original code. Since the work contribution of each instruction depends only on
-    its type, it can be determined at compile time. Taking this one step further, we can calculate work contributions at the basic block
+    straightforward way to measure work online is to instrument the code, before optimization, with a global work counter and instructions for
+    incrementing this counter after each instruction of the original code. In our simplifying scheme, the work contribution of each instruction depends only on
+    its type, so it can be determined at compile time. Taking this one step further, we can calculate work contributions at the basic block
     level and insert only a single increment instruction in each block. Although easy to implement, the na\"{\i}ve approach introduces a

### 2017-11-16 infra.tex
infra minor

@@
     {\itercomp} mechanism works. On the developers' side, we would traditionally produce an application binary optimized for the inputs
-    provided by them. In our approach instead, the program is pre-compiled to LLVM Intermediate Representation (IR) without optimizations,
-    is instrumented to measure the amount of work performed and the wall-clock time, and is bundled with the {\itercomp} mechanism.
-    On the user's side, this mechanism selects an optimization sequence and compiles the instrumented IR to binary using this sequence.
+    provided by them. In our approach instead, the program is pre-compiled to LLVM Intermediate Representation (IR) without optimizations
+    and is instrumented to measure the amount of work performed and the wall-clock time.
+    On the user's side, the online iterative compilation mechanism selects an optimization sequence and compiles the instrumented IR to binary using this sequence.
     Next time the user runs the application, we load and execute the latest binary and measure its work and execution time. We accumulate
@@
     estimate this work efficiency with a confidence interval narrower than a certain threshold, we can start evaluating a different
-    optimization sequence. We call the number of executions performed using the same optimization sequence the \textit{Input-Window Size}.
-    \FIXME{MORE. How does the new search work?}
+    optimization sequence.
 
@@
     the context of data centers~\citep{armbrust10,chen12b}. The proposed \itercomp infrastructure is very well suited to such scenarios. In
-    particular, we can use periods of idleness or underutilization to select the next few optimization sequences to try and compile the
+    particular, we can use periods of idleness or underutilization to select the next few optimization sequences to try and to compile the
     corresponding binaries. This approach effectively hides the compilation overhead from the user. During peak periods, we only have to select

### 2017-11-16 results.tex
minor results

@@
 infrequently executed but were taken into consideration by the LLVM optimization heuristics. In some cases, removed probes caused different
-and suboptimal decisions to be made, increasing the runtime.
+and suboptimal decisions to be made, very slightly increasing the runtime.
 
@@
 strategies, namely \OracleRM, \OraclePP, \OptProf, \WCRelax-5\%, and \WPRelax-5\%. We evaluate every optimization sequence on multiple
-inputs, using a dynamic input-window size, as explained in Section~\ref{sec:oic-infra}. We use the average work efficiency over the input
-window to rank different optimization sequences and we select the one with the highest work efficiency.
+inputs. We use work efficiency to rank different optimization sequences and we select the one with the highest work efficiency.
 

### 2017-11-16 related.tex
related work. added two of grigori's papers

@@
 
-Fursin~\etal~\cite{fursin07} attempted to find a general approach for online \itercomp. They proposed to use instructions per cycle (IPC)
+Older works attempted to drive \itercomp with execution time information~\cite{fursin:hal-01257279,fursin:inria-00001054}. They create
+multiple versions of the hottest procedures, optimized and unoptimized, and at runtime randomly and dynamically select which one to use.
+For each invocation, they record the time spent in the procedure and the version used. By comparing the average time required for each
+invocation, they are able to evaluate the relative performance of different optimizations. The limitations of this approach are that it
+depends on the amount of work performed by the procedure not changing dramatically from invocation to invocation and the number of
+invocations being sufficiently high to collect multiple data points for each version. These drawbacks keep it from being a general solution
+to the problem of iterative compilation.
+
+Fursin~\etal~\cite{fursin07} attempted to find such a general approach. They proposed to use instructions per cycle (IPC)
 as a metric for comparing different runs of the same program with different optimization settings and different inputs. Their hypothesis is
@@
 IPC in the opposite way, making the program more efficient while lowering IPC. The authors in~\cite{fursin07} acknowledge that this limits
-the applicability of IPC for \itercomp. In the context of multithreaded applications, IPC cannot be used to compare the performance of
+the applicability of IPC for \itercomp. Our experiments in Section~\ref{sec:motivation} also show that IPC is not an appropriate metric
+for quantifying efficiency for iterative compilation. In the context of multithreaded applications, IPC cannot be used to compare the performance of
 different runs even when they use the same intput~\cite{alameldeen06,eyerman08}. Work-related metrics are shown to be more appropriate, but prior work

### 2017-11-16 results.tex
updates plot
updates result

@@
 To this end we used the architecture described in Section~\ref{sec:oic-infra} with the search based on genetic algorithm.
-The results suggest that the proposed approach can be profitable in real-world online scenarios.
+Our results provide evidence that the proposed approach is profitable in real-world online scenarios.
+Figure~\ref{fig:deployment_cost} shows, for each benchmark and configuration, the profitability of the final optimization sequence
+selected by the search (as in Figure~\ref{fig:speedups}) but also the average profitability/cost experienced by the user throughout the search process (labeled as [Live]),
+which takes into account the overhead of the work profiling.
+
+A Search Driver based on genetic algorithm is able to focus the search on promising optimization sequences, reducing the risk of
+executing sub-optimal versions of the program as the search evolves through generations.
+Although the user may experience slowdown during the optimization process, due to overheads or by executing sub-optimal versions of the program, after the first generation,
+the optimization sequences produced tend to overcome these costs, with their benefits surpassing the profiling overheads.
+
 For the \WCRelax-5\% configuration, an average speedup of XX was observed during the search, in the worst case with an slowdown of XX.
 For the \WPRelax-5\% configuration, an average speedup of XX was observed during the search, in the worst case with an slowdown of XX.
+
+%\begin{figure*}[t]
+%  \centering
+%  \subfigure[Opt-Profiling]{\includegraphics[width=0.3\textwidth]{figs/Opt-Profiling-deployment.pdf}}
+%  \subfigure[WC-Relax-5]{\includegraphics[width=0.3\textwidth]{figs/WC-Relax-5-deployment.pdf}}
+%  \subfigure[WP-Relax-5]{\includegraphics[width=0.3\textwidth]{figs/WP-Relax-5-deployment.pdf}}
+%  \caption{deployment cost.}
+%  \label{fig:deployment_cost}
+%\end{figure*}

### 2017-11-16 results.tex
updates to result section

@@
 \end{figure*}
-
-%Mean values...
-%Real-R5: 30.43\%
-%WPO-R5:  54.97\%
-%adpcm_d:
-%Real-R5: 75.51\%
-%WPO-R5:  79.07\%
 
@@
 
-\subsubsection{Analysis of the cost during the online search}
+\subsubsection{Evaluation of the Costs During the Online Search}
 
@@
 
-A Search Driver based on genetic algorithm is able to focus the search on promising optimization sequences, reducing the risk of
+With online iterative optimization, although the 
+user may experience slowdown during the search process by executing sub-optimal versions of the program,
+a Search Driver based on genetic algorithm is able to focus the search on promising optimization sequences, reducing the risk of
 executing sub-optimal versions of the program as the search evolves through generations.
-Although the user may experience slowdown during the optimization process, due to overheads or by executing sub-optimal versions of the program, after the first generation,
-the optimization sequences produced tend to overcome these costs, with their benefits surpassing the profiling overheads.
+After the first generation, the optimization sequences produced tend to overcome this issue, with their benefits surpassing the profiling overheads.
 

### 2017-11-16 related.tex
Mars et al in related work

@@
 as a metric for comparing different runs of the same program with different optimization settings and different inputs. Their hypothesis is
-that application binaries displaying higher IPC are more efficient than lower IPC binaries. However, many optimization techniques affect
+that application binaries displaying higher IPC are more efficient than lower IPC binaries. Similarly, Mars~\etal~\cite{Mars:2009:SBO:1545006.1545068}
+propose using IPC for online adaptation. During the learning/monitoring phase, a small number of competing versions of a hot function are executed, each
+for the same amount of time. The one with the highest number of retired instructions is selected for use.
+
+However, many optimization techniques affect
 IPC in the opposite way, making the program more efficient while lowering IPC. The authors in~\cite{fursin07} acknowledge that this limits
 the applicability of IPC for \itercomp. Our experiments in Section~\ref{sec:motivation} also show that IPC is not an appropriate metric
-for quantifying efficiency for iterative compilation. In the context of multithreaded applications, IPC cannot be used to compare the performance of
-different runs even when they use the same intput~\cite{alameldeen06,eyerman08}. Work-related metrics are shown to be more appropriate, but prior work
+for quantifying efficiency when the binary is changing. In the context of multithreaded applications, IPC cannot be used to compare the performance of
+different runs even when they process the same input~\cite{alameldeen06,eyerman08}. Work-related metrics are shown to be more appropriate, but prior work
 relied on application-specific work metrics. This paper presents the first automatic and application agnostic approach for deriving a work metric.
-
-%PP: Discuss input sensitivity?
 

### 2017-11-16 related.tex
related work

@@
 
-Older works attempted to drive \itercomp with execution time information~\cite{fursin:hal-01257279,fursin:inria-00001054}. They create
-multiple versions of the hottest procedures, optimized and unoptimized, and at runtime randomly and dynamically select which one to use.
+Some approaches attempt to drive \itercomp with execution time information~\cite{fursin:hal-01257279,fursin:inria-00001054,tiwari2011online}. They create
+multiple versions of the hottest procedures, optimized and unoptimized, and they evaluate them at runtime on different invocations.
 For each invocation, they record the time spent in the procedure and the version used. By comparing the average time required for each
-invocation, they are able to evaluate the relative performance of different optimizations. The limitations of this approach are that it
-depends on the amount of work performed by the procedure not changing dramatically from invocation to invocation and the number of
+invocation, they are able to evaluate the relative performance of different optimizations. The limitations of such approaches are that they
+depend on the amount of work performed by the procedure not changing dramatically from invocation to invocation and the number of
 invocations being sufficiently high to collect multiple data points for each version. These drawbacks keep it from being a general solution

### 2017-11-16 motivation.tex
updates to motivation and results

--- 2017-11-16 motivation.tex
@@
     performed iterative compilation on \texttt{tiff2bw}, evaluating each optimization sequence on a different input and using either
-    runtime or IPC as our efficiency metric. We selected our inputs from input set \#2 of Figure~\ref{fig:motivation_wrong_inputs}. When
+    runtime or IPC as our efficiency metric. We selected a set of ten inputs from the KDataSets. When
     iterative compilation is driven by runtime, the selected optimization slows down the benchmark by 3\% compared to \texttt{-O3}. When
@@
     actual best optimization over the whole input set improves performance by 7\% while only having an average IPC of 2.31.
+
+\begin{table}[h]
+\centering
+\label{tab:motivation_metrics}
+\begin{tabular}{|l|l|l|}
+\hline
+\multicolumn{1}{|c|}{Metric} & \multicolumn{1}{c|}{Speedup/Slowdown} & \multicolumn{1}{c|}{IPC} \\ \hline
+Oracle (Speedup)             & 7\% speedup                                 & 2.31                      \\ \hline
+IPC                          & 49\% slowdown                                & 3.25                      \\ \hline
+1/Runtime                    & 3\% slowdown                                  & \multicolumn{1}{c|}{-}   \\ \hline
+\end{tabular}
+\caption{}
+\end{table}
+
+%    \begin{figure}[t]
+%        \centering
+%        \includegraphics[width=0.7\linewidth]{figs/motivation_metrics.pdf}
+%        \caption{Average speedup over \texttt{-O3} for the \texttt{tiff2bw} benchmark when optimized with two different optimization sequences and
+%        evaluated on two distinct input sets. Each sequence was selected by iterative compilation on only one of the input sets. Processing
+%        inputs with a version optimized for a different input set produces suboptimal results, supporting that iterative compilation needs
+%        representative inputs.}
+%        \label{fig:motivation_metrics}
+%    \end{figure}
 

### 2017-11-16 results.tex
updates to motivation and results

@@
 
-\subsubsection{Evaluation of the Costs During the Online Search}
+\subsubsection{Evaluating Costs During the Online Search}
+
+\begin{figure*}[t]
+  \centering
+  \subfigure[Opt-Profiling]{\includegraphics[width=0.3\textwidth]{figs/Opt-Profiling-deployment.pdf}}
+  \subfigure[WC-Relax-5\%]{\includegraphics[width=0.3\textwidth]{figs/WC-Relax-5-deployment.pdf}}
+  \subfigure[WP-Relax-5\%]{\includegraphics[width=0.3\textwidth]{figs/WP-Relax-5-deployment.pdf}}
+  \caption{.}
+  \label{fig:deployment_cost}
+\end{figure*}
 
@@
 For the \WPRelax-5\% configuration, an average speedup of XX was observed during the search, in the worst case with an slowdown of XX.
-
-%\begin{figure*}[t]
-%  \centering
-%  \subfigure[Opt-Profiling]{\includegraphics[width=0.3\textwidth]{figs/Opt-Profiling-deployment.pdf}}
-%  \subfigure[WC-Relax-5]{\includegraphics[width=0.3\textwidth]{figs/WC-Relax-5-deployment.pdf}}
-%  \subfigure[WP-Relax-5]{\includegraphics[width=0.3\textwidth]{figs/WP-Relax-5-deployment.pdf}}
-%  \caption{deployment cost.}
-%  \label{fig:deployment_cost}
-%\end{figure*}

### 2017-11-17 motivation.tex
overall changes

@@
     actual best optimization over the whole input set improves performance by 7\% while only having an average IPC of 2.31.
+   Higher IPC does not necessarily translate to better performance,
+ different optimizations affect IPC in different ways.
+   Similarly, a shorter runtime may be due to inputs with fewer work for the program to process.
 
@@
 \end{tabular}
-\caption{}
+\caption{Comparison of iterative compilation guided by different metrics.
+The oracle is guided by the average speedup over \texttt{-O3} on all inputs.
+For IPC and runtime, each optimization was evaluated on a different input.
+}
+\vspace{-2ex}
 \end{table}

### 2017-11-17 results.tex
overall changes

@@
 
-\subsubsection{Evaluating Costs During the Online Search}
+\subsubsection{Cost Evaluation During the Online Search}
 
@@
   \centering
-  \subfigure[Opt-Profiling]{\includegraphics[width=0.3\textwidth]{figs/Opt-Profiling-deployment.pdf}}
-  \subfigure[WC-Relax-5\%]{\includegraphics[width=0.3\textwidth]{figs/WC-Relax-5-deployment.pdf}}
-  \subfigure[WP-Relax-5\%]{\includegraphics[width=0.3\textwidth]{figs/WP-Relax-5-deployment.pdf}}
-  \caption{.}
+  \subfigure[\OptProf]{\includegraphics[width=0.3\textwidth]{figs/Opt-Profiling-deployment.pdf}}
+  \subfigure[\WCRelax-5\%]{\includegraphics[width=0.3\textwidth]{figs/WC-Relax-5-deployment.pdf}}
+  \subfigure[\WPRelax-5\%]{\includegraphics[width=0.3\textwidth]{figs/WP-Relax-5-deployment.pdf}}
+  \caption{Evaluation of the profitability/cost experienced by the user throughout the search process (labeled as [Live]) and of the best optimization sequence selected by the search.
+These results provide evidence that the proposed approach is profitable in real-world online scenarios.}
   \label{fig:deployment_cost}
@@
 We also evaluate the actual cost perceived by the user when applying online iterative compilation.
-To this end we used the architecture described in Section~\ref{sec:oic-infra} with the search based on genetic algorithm.
+To this end we used the architecture described in Section~\ref{sec:oic-infra} with the search based on genetic algorithm~\cite{knijnenburg02,cooper99,kulkarni04}.
 Our results provide evidence that the proposed approach is profitable in real-world online scenarios.
@@
 After the first generation, the optimization sequences produced tend to overcome this issue, with their benefits surpassing the profiling overheads.
-
-For the \WCRelax-5\% configuration, an average speedup of XX was observed during the search, in the worst case with an slowdown of XX.
-For the \WPRelax-5\% configuration, an average speedup of XX was observed during the search, in the worst case with an slowdown of XX.
+We evaluate the deployment cost for the three configurations that use work profiling, namely \OptProf, \WCRelax-5\%, and \WPRelax-5\%.
+In all three configurations, the online search has been shown profitable on average.
+The wider search space is also able to improve the performance gains over \texttt{-O3}.

### 2017-11-17 motivation.tex
minor changes

@@
     actual best optimization over the whole input set improves performance by 7\% while only having an average IPC of 2.31.
-   Higher IPC does not necessarily translate to better performance,
+   Higher IPC does not necessarily mean better performance,
  different optimizations affect IPC in different ways.
-   Similarly, a shorter runtime may be due to inputs with fewer work for the program to process.
+   Similarly, a shorter runtime may be due to inputs with fewer work to process.
 
@@
 }
-\vspace{-2ex}
 \end{table}

### 2017-11-17 results.tex
minor changes

@@
 We also evaluate the actual cost perceived by the user when applying online iterative compilation.
-To this end we used the architecture described in Section~\ref{sec:oic-infra} with the search based on genetic algorithm~\cite{knijnenburg02,cooper99,kulkarni04}.
+To this end we used the architecture described in Section~\ref{sec:oic-infra} with the search based on genetic algorithm~\cite{knijnenburg02,kulkarni04}. %cooper99
 Our results provide evidence that the proposed approach is profitable in real-world online scenarios.

### 2017-11-17 motivation.tex
minor formats

@@
 
-\begin{table}[h]
+\begin{table}[t]
 \centering
@@
 \hline
-\multicolumn{1}{|c|}{Metric} & \multicolumn{1}{c|}{Speedup/Slowdown} & \multicolumn{1}{c|}{IPC} \\ \hline
-Oracle (Speedup)             & 7\% speedup                                 & 2.31                      \\ \hline
+\textbf{Metric} & \textbf{Speedup/Slowdown} & \textbf{IPC} \\ \hline
+\rowcolor{Gray} Oracle (Speedup)             & 7\% speedup                                 & 2.31                      \\ \hline
 IPC                          & 49\% slowdown                                & 3.25                      \\ \hline
-1/Runtime                    & 3\% slowdown                                  & \multicolumn{1}{c|}{-}   \\ \hline
+\rowcolor{Gray}1/Runtime                    & 3\% slowdown                                  & {-}   \\ \hline
 \end{tabular}

### 2017-11-17 motivation.tex
minor

@@
     actual best optimization over the whole input set improves performance by 7\% while only having an average IPC of 2.31.
-   Higher IPC does not necessarily mean better performance,
- different optimizations affect IPC in different ways.
-   Similarly, a shorter runtime may be due to inputs with fewer work to process.
+    When the binary changes, higher IPC does not necessarily mean better performance, different optimizations affect IPC in different ways.
+    Similarly, when the input changes, shorter runtimes do not mean faster execution, it may just be that processing the input requires less work.
 
@@
 \hline
-\textbf{Metric} & \textbf{Speedup/Slowdown} & \textbf{IPC} \\ \hline
+\textbf{Metric} & \textbf{Average Speedup/Slowdown} & \textbf{IPC} \\ \hline
 \rowcolor{Gray} Oracle (Speedup)             & 7\% speedup                                 & 2.31                      \\ \hline

### 2017-11-17 results.tex
minor formats
minor

--- 2017-11-17 results.tex
@@
 of each benchmark.
-% None of the configurations degrades the performance over \texttt{-O3} (with statistical significance). \FIXME{PP: According to the figure there are slowdowns!!!}
 Figure~\ref{fig:speedups} shows these speedups for all test benchmarks. In most cases, \OraclePP is very close to \OracleRM, achieving on
@@
   \subfigure[\OptProf]{\includegraphics[width=0.3\textwidth]{figs/Opt-Profiling-deployment.pdf}}
+  \hfill
   \subfigure[\WCRelax-5\%]{\includegraphics[width=0.3\textwidth]{figs/WC-Relax-5-deployment.pdf}}
+  \hfill
   \subfigure[\WPRelax-5\%]{\includegraphics[width=0.3\textwidth]{figs/WP-Relax-5-deployment.pdf}}
-  \caption{Evaluation of the profitability/cost experienced by the user throughout the search process (labeled as [Live]) and of the best optimization sequence selected by the search.
-These results provide evidence that the proposed approach is profitable in real-world online scenarios.}
+  \caption{Speedup over \texttt{-O3} for the best optimization selected by the search, as well as the average speedup as experienced by the
+    user throughout the search process ([Live]). The results show that the proposed approach is profitable in real-world online scenarios.}
   \label{fig:deployment_cost}
@@
 
-We also evaluate the actual cost perceived by the user when applying online iterative compilation.
-To this end we used the architecture described in Section~\ref{sec:oic-infra} with the search based on genetic algorithm~\cite{knijnenburg02,cooper99,kulkarni04}.
-Our results provide evidence that the proposed approach is profitable in real-world online scenarios.
-Figure~\ref{fig:deployment_cost} shows, for each benchmark and configuration, the profitability of the final optimization sequence
-selected by the search (as in Figure~\ref{fig:speedups}) but also the average profitability/cost experienced by the user throughout the search process (labeled as [Live]),
+A typical argument against online adaptive techniques is that the overhead of management and the slowdown due to testing suboptimal
+configuration might overshadow the benefits of the technique. In this subsection, we establish that online iterative compilation is
+profitable even when factoring in the online costs. We modify the architecture described in Section~\ref{sec:oic-infra} to make it more
+realistic by using genetic search~\cite{knijnenburg02,kulkarni04} over the whole space of possible optimizations.
+
+Figure~\ref{fig:deployment_cost} shows, for each benchmark and configuration, the speedup over \texttt{-O3} of the best optimization sequence
+evaluated by the search but also the average profitability/cost experienced by the user throughout the search process (labeled as [Live]),
 which takes into account the overhead of the work profiling.
-
 With online iterative optimization, although the 

### 2017-11-17 results.tex
last results subsec

@@
 
-\subsubsection{Cost Evaluation During the Online Search}
-
 \begin{figure*}[t]
@@
 
+\subsubsection{Cost Evaluation During the Online Search}
+
 A typical argument against online adaptive techniques is that the overhead of management and the slowdown due to testing suboptimal
@@
 evaluated by the search but also the average profitability/cost experienced by the user throughout the search process (labeled as [Live]),
-which takes into account the overhead of the work profiling.
-With online iterative optimization, although the 
-user may experience slowdown during the search process by executing sub-optimal versions of the program,
-a Search Driver based on genetic algorithm is able to focus the search on promising optimization sequences, reducing the risk of
-executing sub-optimal versions of the program as the search evolves through generations.
-After the first generation, the optimization sequences produced tend to overcome this issue, with their benefits surpassing the profiling overheads.
-We evaluate the deployment cost for the three configurations that use work profiling, namely \OptProf, \WCRelax-5\%, and \WPRelax-5\%.
-In all three configurations, the online search has been shown profitable on average.
-The wider search space is also able to improve the performance gains over \texttt{-O3}.
+which takes into account the overhead of the work profiling. We evaluate three configurations that use work profiling, namely \OptProf,
+\WCRelax-5\%, and \WPRelax-5\%. By widening the search space, we are able to improve performance significantly compared to the results in
+Figure~\ref{fig:speedups}. In terms of the overall speedup throughout the search, all three configurations are profitable on average. 
+Although there is a slight risk we select a suboptimal version and the user experiences slowdown, our genetic search focuses on promising
+optimization sequences, reducing the risk of selecting a suboptimal version of the program as the genetic population evolves. For all three
+configurations, after the first generation, the optimization sequences we evaluate tend to overcome this issue, with their benefits
+surpassing the profiling overheads. This is particularly true for \WPRelax-5\%, where the small overhead combined with the large speedup
+lead to a 7.3\% overall speedup.

### 2017-11-17 motivation.tex
minor motivation

@@
     much less than what optimization \#1 achieves. In both cases, optimizing the benchmark for the same set it will be evaluated on
-    improves performance by 15\% compared to optimizing for some other set. To make iterative compilation profitable, applying it on a
+    improves performance by 15\% compared to optimizing for some other set. To make iterative compilation profitable, applying it on
     inputs similar to the ones the application will be used with is absolutely necessary.
@@
     similar behavior as when the application is used by users. This needs a deep knowledge of the potential user behavior and
-    considerable effort. This will have to be a continuous work in
-    progress, to handle changes in the way the application is used.
+    considerable effort. This will have to be a continuous work in progress, to handle changes in the way the application is used. In some
+    areas, such as mobile computing, where developers do not traditionally optimize their applications or provide inputs for performance
+    testing, depending on them to provide inputs is a suboptimal choice.
 

### 2017-11-17 related.tex
work and related minor

@@
 most cases it is possible to find a single set of optimizations that produces good results across all inputs; but achieving this requires
-developers provide a few tens of representative inputs to be used by the iterative compilation search process. We argue that providing
+developers provide a large number of representative inputs to be used by the iterative compilation search process. We argue that providing
 representative inputs is difficult in practice, as users may not use the applications as the developer expects. Even if this is possible,
@@
 invocation, they are able to evaluate the relative performance of different optimizations. The limitations of such approaches are that they
-depend on the amount of work performed by the procedure not changing dramatically from invocation to invocation and the number of
+depend on the amount of work performed by the procedure not changing from invocation to invocation and the number of
 invocations being sufficiently high to collect multiple data points for each version. These drawbacks keep it from being a general solution

### 2017-11-17 work.tex
minor comments
overall changes
minor changes
work and related minor

@@
     \centering
-    \includegraphics[width=\textwidth]{figs/corr_coeff.pdf}
+    \includegraphics[width=0.95\textwidth]{figs/corr_coeff.pdf}
     \caption{Correlation coefficients between the work metric ($\Delta W$) and the execution time of the unoptimized version for each benchmark.}
@@
     Optimal instrumentation significantly reduces the profiling overhead when compared to na\"ive instrumentation, from an average overhead
-    of 79\% to 13\%. Still, 13\% slowdown might be more than what the user is ready to tolerate, while for isolated case the overhead might
+    of 79\% to 13\%. Still, 13\% slowdown might be more than what the user is ready to tolerate, while for isolated cases the overhead might
     be as high as 60\%. We need to reduce overhead further to make the work metric practical.
@@
 
-    The \WCRelaxLower relaxation error guarantees are hard but also too conservative. It assumes that each loop will always execute the
+    The \WCRelaxLower relaxation error guarantee is hard but also too conservative. It assumes that each loop will always execute the
     path with the minimum amount of work possible, which might be orders of magnitude less than the work actually performed. When this is

### 2017-11-17 motivation.tex
motivation minor
motivation fix

@@
     areas, such as mobile computing, where developers do not traditionally optimize their applications or provide inputs for performance
-    testing, depending on them to provide inputs is a suboptimal choice.
+    testing, depending on them to provide inputs is unrealistic.
 
@@
 \hline
-\textbf{Metric} & \textbf{Average Speedup/Slowdown} & \textbf{IPC} \\ \hline
+\textbf{Metric} & \textbf{Speedup/Slowdown} & \textbf{IPC} \\ \hline
 \rowcolor{Gray} Oracle (Speedup)             & 7\% speedup                                 & 2.31                      \\ \hline

### 2017-11-17 results.tex
Exp.Setup pointer to online search results
search is now subsection
tiny

@@
 
-\subsubsection{Cost Evaluation During the Online Search}
+\subsection{Overall Benefit of Online Search}\label{subsec:search}
 
@@
 surpassing the profiling overheads. This is particularly true for \WPRelax-5\%, where the small overhead combined with the large speedup
-lead to a 7.3\% overall speedup.
+leads to a 7.3\% overall speedup.
